### 2023-02-22 analysis/spatiotemporal-transmission-risk.Rmd
iwhod abstract analysis and preprocessing xovis

---
title: "Spatiotemporal transmission risk"
author: "Nicolas Banholzer"
date: "2022-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(LaplacesDemon)
library(rriskDistributions)
library(tidybayes)
library(lubridate)
library(sf)
library(raster)
library(terra)
library(ggridges)
library(ggplot2)

source("../utils/distr.r")
source("../utils/trans_risk.r")
source("../utils/spatial.r")
source("../utils/plotting.R")
```

## Assumptions

### Spatial emission

* Ko et al. (2004): exponential decrease on an airplane, but they are more strongly ventilated
* Abkarian et al. (2020) deduce that, for the respiratory jet generated by typical speaking, the concentration of pathogen is diminished to approximately 3% of its initial value at a distance of 2 m.
* SSPH report: distance depends on size of aerosols and initial speed, which depends on the activity. For speaking, it will be around 1-2m, for coughing or sneezing it can be up to 7-8m
* Wang et al. (2021) and other studies: viral concentration is highest closest to the source

--> Assume an exponential decrease in viral concentration with distance to source relative to the source (relative concentration = 1).

```{r}
rMuIc <- function(n) {
  rgamma(n, shape = 10, scale = .2)
}

hist(rMuIc(1e3))

dIc <- function(distance, mu) { 
  1 - pexp(distance, 1 / mu)
}

expand.grid(distance = seq(0, 10, .1), rate = rMuIc(1e3)) %>%
  mutate(dens_inf = dIc(distance, rate)) %>%
  ggplot(aes(x = distance, y = dens_inf)) +
  stat_lineribbon() +
  scale_fill_brewer() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) 
```

### Infectious quanta

* Catanzaro (1982): 249 quanta/h 
* Nardell (1991): 12.7 quanta/h
* Riley (1962): 1.25 quanta/h but patients already started on chemotheraphy and also few highly infectious patients
* Escombe (2008): 8.2 quanta/h but wide range, including an extreme outlier of 226 quanta/h
* Andrews (2014): 0.44 to 5.69 quanta/h

--> The distribution should roughly reflect the following quantiles
* <1: 10%
* <5: 50%
* <10: 75%
* <50: 95%
* <100: 99%

```{r}
q_a <- c(1, 5, 10, 50, 100, 200)
p_a <- c(.1, .5, .75, .95, .99, 1)
q_range <- seq(0, 200, .1)
p_q_range <- ptrunc(q_range, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
p_a_selected <- ptrunc(q_a, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
lnorm_par <- inv_lnorm_moments(5, 10)
p_q_range_lnorm <- plnorm(q_range, meanlog = lnorm_par$meanlog, sdlog = lnorm_par$sdlog)

ggplot(mapping = aes(x = q_range)) +
  geom_line(mapping = aes(y = p_q_range), color = "black") +
  geom_line(mapping = aes(y = p_q_range_lnorm), color = "red") +
  geom_point(mapping = aes(x = q_a, y = p_a_selected), color = "black") +
  geom_point(mapping = aes(x = q_a, y = p_a), color = "blue") +
  scale_x_log10(breaks = q_a) +
  scale_y_continuous(breaks = p_a)

rq <- function(n) {
  rtrunc(n, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
}

dq <- function(x) {
  dtrunc(x, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
}

rq_distr <- data.frame(quanta = seq(0, 100, .1)) %>% mutate(prob = dq(quanta))

ggplot(rq_distr, aes(x = quanta, y = prob)) +
  geom_line()
```

### Average volume of exhaled gas

* Richardson et al. (2014): 0.1 - 1.17 l/s per person based on Feher (2017)

```{r}
rV <- function(n) {
  runif(n, 0.1, 0.17)
}
```

### Loss rate due to inactivation

* Nardell (2016) reviews Loudon et al. (1969) and Lever et al. (2000) reporting contradictory findings of half-lives between 20m to 6h. Exact viability in the air is unknown.
* The finding by Lever et al. would suggest that sustained contact is required for transmission.
* The viability in the air probably depends a lot on the conditions, in particular ventilation and UV.

--> TODO: Check how high the ventilation rate is compared to half-lives as to whether we can ignore the half-lives.

### Loss rate due to deposition onto surfaces

* Wang et al. (2021): The larger the size of the particle, the less long it will reside in the air. Environmental factors such as temperature, humidity and ventilation also play an important role.

### CO2 levels in outdoor air

* https://gml.noaa.gov/dv/data/index.php?site=CPT&parameter_name=Carbon%2BDioxide,%20accessed%2012%20November%202019

```{r}
co2_outdoor <- 412.5
```

### CO2 levels in exhaled air

* Rudnick and Milton (2003): 40,000

```{r}
co2_exhaled <- 40000
```

### CO2 generation rate

* Persily et al.: see Table 5, around 0.003 to 0.005

```{r}
rG <- function(n) {
  runif(n, 0.0041, 0.0055)
}
```

### Unmasked TB patients

* Masked TB patients: TB suspects
* Unmasked TB patients: Poisson(mu = TB suspects)

```{r}
rTBunmasked <- function(n, lambda) {
  rpois(n, lambda)
}
```

## Data

### Clinical

```{r}
# clinical data
clinXov <- readRDS("../data-clean/Masi/merged-clinical-xovis.rds")

yesTB <- clinXov %>%
  filter(tb_suspect=="yes") %>%
  group_by(patient_id) %>%
  slice(1) %>%
  ungroup()

nTBmasked <- nrow(yesTB)

noTB <- clinXov %>%
  dplyr::filter(ifelse(is.na(tb_suspect), T, ifelse(tb_suspect=="no", T, F))) %>%
  group_by(person_id) %>%
  slice(1) %>%
  ungroup()

# Time period (for the first day)
time_period <- seq(min(as.POSIXct(format(clinXov$time, "%Y-%m-%d %H:%M"))), 
                   max(as.POSIXct(format(clinXov$time, "%Y-%m-%d %H:%M"))), 
                   by = "min")
```

### Co2

```{r}
# co2
co2 <- readRDS("../data-clean/Masi/environmental-data/co2-levels.rds") %>%
  filter(date == "2021-10-25") 
#' assume waiting area has mean co2 of waiting room and registration
#' passage has co2 of watiing room only
#' tb room has co2 of tb room only
co2 <- co2 %>%
  filter(location %in% c("waiting room", "registration")) %>%
  group_by(date_time) %>%
  summarize(co2 = mean(co2)) %>%
  ungroup() %>%
  mutate(location = "waiting room") %>%
  rbind(co2 %>%
          filter(location == "waiting room") %>%
          mutate(location = "passage") %>%
          dplyr::select(date_time, location, co2)) %>%
  rbind(co2 %>%
          filter(location == "TB room") %>%
          mutate(location = "tb room") %>%
          dplyr::select(date_time, location, co2)) 
ggplot(co2, aes(x = date_time, y = co2, color = location)) + geom_line()
```

### Room

```{r}
# choose cell size such that the squared cell has 1m3 volume
cellVolume <- 1
cellSize <- sqrt(cellVolume * 1e9 / 3e3)

# waiting room
sh_wr <- vect("../data-raw/Masi/building/waiting_room.shp")
sf_wr <- st_as_sf(sh_wr)
wrRoomSP <- shapeToSpatial(sh_wr, cellSize)
wrVol <- compute_volume(wrRoomSP)
wrCellCoords <- create_coord_df(wrRoomSP)
wrRoomXY <- as.matrix(dplyr::select(wrCellCoords, x, y))

# tb room
sh_tb <- vect("../data-raw/Masi/building/tb_room.shp")
sf_tb <- st_as_sf(sh_tb)
tbRoomSP <- shapeToSpatial(sh_tb, cellSize)
tbVol <- compute_volume(tbRoomSP)
tbCellCoords <- create_coord_df(tbRoomSP)
tbRoomXY <- as.matrix(dplyr::select(tbCellCoords, x, y))

# passage
sh_pa <- vect("../data-raw/Masi/building/passage.shp")
sf_pa <- st_as_sf(sh_pa)
paRoomSP <- shapeToSpatial(sh_pa, cellSize)
paVol <- compute_volume(paRoomSP)
paCellCoords <- create_coord_df(paRoomSP)
paRoomXY <- as.matrix(dplyr::select(paCellCoords, x, y))

# combine
clinicCellCoords <- rbind(
  mutate(wrCellCoords, location = "waiting room"),
  mutate(tbCellCoords, location = "tb room"),
  mutate(paCellCoords, location = "passage")
) %>% mutate(id = 1:nrow(.))
clinicCellCoordsXY <- cbind(clinicCellCoords$x, clinicCellCoords$y)
```

### No. people

```{r}
# number of people in space
nDF <- tibble()
for (tp in time_period) {
  n_tp <- clinXov %>%
    filter(between(time, tp, tp + seconds(59))) %>%
    mutate(location = mapply(function(x,y) clinicCellCoords$location[find_raster(x,y,clinicCellCoordsXY)], x, y)) %>%
    group_by(person_id, location) %>%
    slice(1) %>%
    ungroup() %>%
    group_by(location) %>%
    summarize(n = n()) %>%
    ungroup() 
  for (loc in unique(clinicCellCoords$location)) {
    if (!(loc %in% n_tp$location)) {
      n_tp <- rbind(n_tp, data.frame(location = loc, n = 0))
    } 
  }
  nDF <- rbind(nDF, n_tp %>% mutate(time = tp))
}
nDF <- nDF %>% 
  mutate(time = as.POSIXct(time, origin = as.Date("1970-01-01")))
nDF$n[nDF$location=="waiting room"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"]
nDF$n[nDF$location=="passage"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"]
ggplot(nDF, aes(x = time, y = n, color = location)) + geom_line()
```

### Movement

```{r}
patMov <- clinXov %>%
  mutate(id = map2_int(x, y, find_raster, cellCoordsXY = clinicCellCoordsXY))
```

## Simulation

```{r}
# Settings
set.seed(12345)
nSim <- 100

# Input parameters
inputPar <- data.frame(
  spatial_mu = rMuIc(nSim),
  quanta = rq(nSim),
  V = rV(nSim),
  co2_ex = co2_exhaled,
  co2_out = min(co2$co2),
  co2_gen = rG(nSim),
  unmasked_TB = rTBunmasked(nSim, nTBmasked)
)

# Loop
for (s in 1:nSim) {
  message(sprintf("Simulation %i of %i", s, nSim))

  # parameters
  smu <- inputPar$spatial_mu[s]
  q <- inputPar$quanta[s]
  V <- inputPar$V[s]
  Ca <- inputPar$co2_ex[s]
  Co <- inputPar$co2_out[s]
  G <- inputPar$co2_gen[s]
  unmasked_TB <- inputPar$unmasked_TB[s]

  # transformed parameters
  # quanta emission rate per minute
  ER <- q / 60

  # TB patients
  masked_tb <- yesTB$person_id
  unmasked_tb <- sample(noTB$person_id, size = unmasked_TB)
  TBpatients <- patMov %>%
    filter(person_id %in% c(masked_tb, unmasked_tb)) %>%
    dplyr::select(person_id, observation_id, time, x, y) 
  TBpatients_full <- TBpatients %>%
    group_by(person_id) %>%
    summarize(time = list(seq(min(time), max(time), by = "sec"))) %>%
    ungroup() %>%
    unnest(cols = c(time))
  TBpatients_full <- TBpatients_full %>%
    left_join(TBpatients, by = c("person_id", "time")) %>%
    tidyr::fill(observation_id, x, y, .direction = "down")

  # no TB patients
  noTBpatients <- patMov %>%
    filter(!(person_id %in% c(masked_tb, unmasked_tb))) %>%
    dplyr::select(person_id, time, id) 
  noTBpatients_full <- noTBpatients %>%
    group_by(person_id) %>%
    summarize(time = list(seq(min(time), max(time), by = "sec"))) %>%
    ungroup() %>%
    unnest(cols = c(time))
  noTBpatients_full <- noTBpatients_full %>%
    left_join(noTBpatients, by = c("person_id", "time")) %>%
    tidyr::fill(id, .direction = "down") %>%
    mutate(time = as.POSIXct(format(time, "%Y-%m-%d %H:%M"))) %>%
    group_by(person_id, id, time) %>%
    summarize(minutes = n() / 60) %>%
    ungroup()

  # df to store quanta concentration by minute
  quantaDF <- clinicCellCoords %>%
    mutate(time = time_period[1] - minutes(1),
           I_SP = 0,
           I_WM = 0,
           E_SP = 0,
           E_WM = 0,
           N_SP = 0,
           AER = NA,
           RR = NA,
           N_WM = 0,
           N_SP = 0)

  for (t in 1:length(time_period)) {

    # time point
    tp <- time_period[t] 
    quantaDF_new <- clinicCellCoords %>%
      mutate(time = tp,
             I_SP = 0,
             I_WM = 0,
             E_SP = 0,
             E_WM = 0,
             N_SP = 0,
             N_WM = 0)

    # spatial spread of infectious individuals and quanta emission
    TBpatients_tp <- TBpatients_full %>%
      dplyr::filter(between(time, tp, tp + minutes(1)))
    if (nrow(TBpatients_tp) > 0) {
      TBpatients_tp <- TBpatients_tp %>%
        nest(data = -person_id)
      # iterate over patients
      for (p in 1:nrow(TBpatients_tp)) {
        # duration and location
        duration <- nrow(TBpatients_tp$data[[p]])
        xy <- TBpatients_tp$data[[p]] %>% dplyr::select(x, y) %>% slice(1) %>% unlist()
        # generated quanta by cell
        locationID <- find_raster(xy[1], xy[2], clinicCellCoordsXY)
        locationName <- clinicCellCoords$location[locationID]
        locationCellCoords <- filter(clinicCellCoords, location == locationName)
        locationShape <- if(locationName == "waiting room") {sf_wr} else if(locationName == "tb room") {sf_tb} else {sf_pa}
        cellInReach <- numeric(nrow(quantaDF_new))
        cellDistance <- numeric(nrow(quantaDF_new))
        isIinCell <- ifelse(quantaDF_new$location==locationName, 1, 0 )
        for (cell in 1:nrow(quantaDF_new)) {
          # if cell is in room/location
          if (quantaDF_new$location[cell] != locationName) {
            cellInReach[cell] <- 0
            cellDistance[cell] <- -999
          } else {
            # can aerosols travel to this cell?
            straight_line <- st_linestring(rbind(xy, c(quantaDF_new$x[cell], quantaDF_new$y[cell]))) 
            intersects <- st_difference(straight_line, locationShape)
            cellInReach[cell] <- ifelse(length(intersects) == 0, 1, 0)
            cellDistance[cell] <- convert_dist(st_length(straight_line))
          }
        }
        # proportion of infectious individuals per cell per minute
        distance_dens <- cellInReach * dIc(cellDistance, smu)
        quantaDF_new$I_SP <- quantaDF_new$I_SP + distance_dens / sum(distance_dens) * (duration / 60)
        quantaDF_new$I_WM <- quantaDF_new$I_WM + isIinCell / sum(isIinCell)
      }
      # quanta emission
      quantaDF_new$E_SP <- quantaDF_new$E_SP + ER * quantaDF_new$I_SP
      quantaDF_new$E_WM <- quantaDF_new$E_WM + ER * quantaDF_new$I_WM
    }

    # compute air exchange rate and quanta removal
    Ct <- co2 %>%
      group_by(location) %>%
      # if CO2 is missing, take the last observed value
      filter(date_time <= tp) %>%
      arrange(desc(date_time)) %>%
      slice(1) %>%
      mutate(date_time = tp) %>%
      ungroup() %>%
      # add number of people
      left_join(filter(nDF, time == tp), by = c("date_time" = "time", "location")) %>%
      # add volume
      mutate(vol = ifelse(location == "tb room", tbVol, wrVol + paVol)) %>%
      # compute AER and RR per minute
      mutate(AER = compute_AER(G, Co, co2, n, vol),
             RR = compute_RR(AER) / 60) 
    quantaDF_new <- quantaDF_new %>% 
      left_join(Ct %>% dplyr::select(location, AER, RR), by = "location")

    # previous quanta concentration
    N0_SP <- quantaDF %>% 
      filter(time == tp - minutes(1)) %>%
      dplyr::select(N_SP) %>% unlist
    N0_WM <- quantaDF %>% 
      filter(time == tp - minutes(1)) %>%
      dplyr::select(N_WM) %>% unlist

    # compute quanta concentration
    quantaDF_new$N_SP <- pmap_dbl(list(E = quantaDF_new$E_SP, 
                                       RR = quantaDF_new$RR, 
                                       N0 = N0_SP,
                                       V = cellVolume), compute_Nt)
    quantaDF_new$N_WM <- pmap_dbl(list(E = quantaDF_new$E_WM, 
                                       RR = quantaDF_new$RR, 
                                       N0 = N0_WM,
                                       V = cellVolume), compute_Nt)

    # append new data
    quantaDF <- rbind(quantaDF, quantaDF_new)
  }

  # compute (spatio)temporal transmission risk for each individual
  risk_of_infection <- noTBpatients_full %>%
    left_join(quantaDF, by = c("id", "time")) %>%
    group_by(person_id) %>%
    summarize(exposure_time = sum(minutes),
              P_SP = 1-exp(-sum(N_SP * minutes)),
              P_WM = 1-exp(-sum(N_WM * minutes))) %>%
    ungroup()

  # compute average transmission risk
  sum_I <- data.frame(location = c("waiting room", "passage", "tb room"),
                      I = length(c(masked_tb, unmasked_tb))) 
  mean_n <- nDF %>% 
    group_by(location) %>%
    summarize(n = median(n) + 1) %>%
    ungroup()
  mean_f <- co2 %>%
    group_by(location) %>%
    summarize(f = (mean(co2) - Co) / Ca) %>%
    ungroup()
  risk_of_infection_av <- noTBpatients_full %>%
    left_join(quantaDF %>% dplyr::select(id, time, location), by = c("id", "time")) %>%
    group_by(person_id, location) %>%
    summarize(minutes = sum(minutes)) %>%
    ungroup() %>%
    left_join(sum_I, by = "location") %>%
    left_join(mean_n, by = "location") %>%
    left_join(mean_f, by = "location") %>%
    mutate(mu = f * I * ER * minutes / n) %>%
    group_by(person_id) %>%
    summarize(P_SS = 1 - exp(-sum(mu))) %>%
    ungroup()
  risk_of_infection <- risk_of_infection %>%
    left_join(risk_of_infection_av, by = "person_id")

  # save simulation
  saveRDS(quantaDF, paste0("../simulations/2021-10-25/sim-quanta", s, ".rds"))
  saveRDS(risk_of_infection, paste0("../simulations/2021-10-25/sim-transRisk", s, ".rds"))
}
```

## Evaluation

### TB risk

```{r}
# files
transRiskFiles <- list.files("../simulations/2021-10-25", full.names = T)
transRiskFiles <- transRiskFiles[grepl("transRisk", transRiskFiles)]

# read files
transRiskDF <- data.frame(file = transRiskFiles) %>%
  mutate(simulation = stringi::stri_extract(basename(file), regex = "\\d{1,4}"), 
         data = lapply(file, readRDS)) %>%
  unnest(cols = c(data))

# QQ Plot
transRiskDF_Q <- transRiskDF %>%
  dplyr::select(simulation, P_SP, P_WM) %>%
  melt("simulation") %>%
  group_by(simulation, variable) %>%
  summarize(quantile = seq(0.05, .95, 0.05),
            value = quantile(value, seq(0.05, 0.95, 0.05))) %>%
  ungroup()

transRiskDF_Q_av <- transRiskDF_Q %>%
  group_by(variable, quantile) %>%
  summarize(value = mean(value)) %>%
  ungroup()

transRiskDF_Q_av %>%
  ggplot(aes(x = quantile, y = value, color = variable)) +
  geom_line()

# average transmission risk across patients
average_tr <- transRiskDF %>%
  dplyr::select(simulation, P_SP, P_WM) %>%
  mutate(P_SP = P_SP * 100,
         P_WM = P_WM * 100) %>%
  melt("simulation") %>%
  group_by(simulation, variable) %>%
  summarize(`Mean` = mean(value),
            `Standard deviation` = sd(value),
            `Median` = median(value),
            `Tail risk (95%-Quantile)` = quantile(value, .95)) %>%
  ungroup()

figA <- average_tr %>%
  rename(model = variable) %>%
  mutate(model = ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling")) %>%
  melt(c("simulation", "model")) %>%
  ggplot(aes(x = value, y = model, fill = model, color = model)) +
  stat_dots(quantiles = 100) +
  facet_wrap(~ variable, scales = "free_x") +
  scale_color_manual(values = cbPalette[2:3]) +
  scale_fill_manual(values = cbPalette[2:3]) +
  labs(x = "Risk of infection (%)", 
       title = "Fig. 1a: Risk of infection across patients",
       subtitle = "Colored dots show individual simulation results.") +
  theme_bw2() +
  theme(legend.position = "top", axis.title.y = element_blank(), axis.text.y = element_blank(), legend.title = element_blank(),
        legend.justification='left', legend.direction='horizontal')

figA

average_tr %>%
  dplyr::select(-simulation) %>%
  group_by(variable) %>%
  median_qi() %>%
  mutate_if(is.numeric, round, 2)

figB <- transRiskDF %>%
  ggplot(aes(x = P_SP, y = P_WM)) +
  geom_point(shape = 1) +
  geom_abline(mapping = aes(intercept = 0, slope = 1), color = "red", linetype = "dashed") +
  annotate("richtext", x = 0.75, y = 0.82, 
           label = "<span style = 'background-color:#fff'>Perfect correlation</span>", 
           size = 8 / cm(1), color = "red", angle = 22, label.color = NA) +
  geom_smooth(method = "lm") +
  annotate("richtext", x = 0.85, y = 0.55, 
           label = "<span style = 'background-color:#fff'>Empirical correlation</span>", 
           size = 8 / cm(1), color = "blue", angle = 16, label.color = NA) +
  labs(x = "Risk of infection (%) with spatiotemporal modeling", 
       y = "Risk of infection (%) with temporal modeling",
       title = "Fig. 1b: Patient-specific risk of infection",
       subtitle = "One dot per patient and simulation. Dots off (on) the diagonal imply a (no) differing risk of infection by modeling.") +
  scale_x_continuous(expand = c(0,0), limits = c(0,1), label = function(x) x * 100) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1), label = function(x) x * 100) +
  theme_bw2() +
  theme(plot.margin = unit(c(5.5, 7.5, 5.5, 5.5), "points"))

figB

ggsave("../results/model-comparison.png", grid.arrange(figA, figB, ncol = 1), width = 16 / cm(1), height = 16 / cm(1))

# visualization of spatial approach
transRriskSP_pl <- transRiskDF %>%
  mutate(P_SP = P_SP * 100) %>%
  ggplot(aes(x = P_SP, y = simulation)) + # , color = ..x..
  stat_dots(quantiles = 20, color = "black") +
  #scale_color_viridis() +
  labs(y = "Simulation", x = "Risk of infection (%)",
       title = "Figure 1: Risk of infection", subtitle = "Quantile distribution of patient-specific risk of infection (%) for each simulation") +
  scale_x_sqrt(limits = c(0,NA), expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  theme_bw2() +
  theme(legend.position = "none")

transRriskSP_pl

# low, medium, high risk
transRiskDF_cat <- transRiskDF %>%
  dplyr::select(simulation, P_SP, P_WM) %>%
  melt("simulation") %>%
  group_by(simulation, variable) %>%
  summarize(`Very low` = sum(value <= .01),
            Low = sum(value > .01 & value <= .05),
            Medium = sum(value > .05 & value <= .1),
            High = sum(value > .1 & value <= .5),
            `Very high` = sum(value > .5)) %>%
  ungroup()

figB <- transRiskDF_cat %>%
  rename(model = variable) %>%
  mutate(model = factor(ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling"),
                        levels = c("Spatiotemporal modeling", "Temporal modeling"))) %>%
  melt(c("simulation", "model")) %>%
  group_by(model, variable) %>%
  summarize(value = mean(value)) %>%
  ungroup() %>%
  ggplot(aes(y = variable, fill = model, x = value)) +
  geom_bar(stat = "identity", position = position_dodge2()) +
  scale_x_log10(expand = c(0,0)) + 
  scale_fill_manual(values = cbPalette[c(2,1)]) +
  labs(title = "B | Average number of visitors by risk of infection",
       subtitle = "Number of visitors with very high (>50%), high (10-50%), medium (5-10%), low (1-5%), and very low (<1%) risk of infection.",
       x = "Number of visitors") +
  theme_bw2() +
  theme(legend.position = "top", legend.justification='left', legend.direction='horizontal',
        legend.title = element_blank(), legend.key.width = unit(2.5, "cm"),
        axis.title.y = element_blank(),
        plot.title.position = "plot")

figB

transRiskDF_cat_alt <- transRiskDF %>%
  dplyr::select(simulation, P_SP, P_WM) %>%
  melt("simulation") %>%
  group_by(simulation, variable) %>%
  summarize(Low = sum(value <= .1),
            Medium = sum(value > .1 & value <= .5),
            High = sum(value > .5)) %>%
  ungroup()

figB_alt <- transRiskDF_cat_alt %>%
  rename(model = variable) %>%
  mutate(model = factor(ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling"),
                        levels = c("Spatiotemporal modeling", "Temporal modeling"))) %>%
  melt(c("simulation", "model")) %>%
  group_by(model, variable) %>%
  summarize(value = mean(value)) %>%
  ungroup() %>%
  ggplot(aes(y = variable, fill = model, x = value)) +
  geom_bar(stat = "identity", position = position_dodge2()) +
  scale_x_log10(expand = expansion(mult = c(0,.01))) + 
  scale_fill_manual(values = cbPalette[c(2,1)], breaks = c("Temporal modeling", "Spatiotemporal modeling")) +
  labs(title = "B | Estimated number of visitors by risk of infection",
       subtitle = "Average number of visitors with high (>50%), medium (10-50%), and low (<10%) risk of infection.",
       x = "Number of visitors") +
  theme_bw2() +
  theme(legend.position = c(0.675, 0.825), legend.justification='left', legend.direction='vertical',
        legend.title = element_blank(), legend.key.width = unit(1.5, "cm"),
        axis.title.y = element_blank(),
        plot.title.position = "plot")

figB_alt
```

### Quanta

```{r}
# files
quantaFiles <- list.files("../simulations/2021-10-25", full.names = T)
quantaFiles <- quantaFiles[grepl("quanta", quantaFiles)]

# read files
quantaDF <- data.frame(file = quantaFiles) %>%
  mutate(simulation = stringi::stri_extract(basename(file), regex = "\\d{1,4}"), 
         data = lapply(file, readRDS)) %>%
  unnest(cols = c(data))

# average quanta concentration
quantaDF_av <- quantaDF %>%
  filter(location == "waiting room") %>%
  # mutate(daytime = ifelse(time <= "2021-10-25 11:00:00", "Morning", 
  #                         ifelse(time <= "2021-10-25 13:00:00", "Noon", "Afternoon"))) %>%
  mutate(daytime = ifelse(time <= "2021-10-25 12:00:00", "Morning", "Afternoon")) %>%
  group_by(daytime, id) %>%
  summarize(N_SP = mean(N_SP)) %>%
  ungroup()

wrRoomSP_df <- fortify(wrRoomSP) %>%
  left_join(quantaDF_av %>% mutate(id = as.character(id)), by = "id")

figA <- wrRoomSP_df %>%
  # mutate(daytime = factor(daytime, c("Morning", "Noon", "Afternoon"))) %>%
  mutate(daytime = factor(daytime, c("Morning", "Afternoon"))) %>%
  ggplot(aes(x = lat, y = long, group = group, fill = N_SP)) +
  geom_polygon() +
  facet_wrap(~ daytime) +
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA) +
  labs(title = "A | Estimated spatiotemporal quanta concentration in the clinic by daytime",
       # subtitle = "Quanta concentration (quanta/m3) in the morning (7:30 to 11am), noon (11am to 1pm), and afternoon (1 to 4pm).") +
       subtitle = "Average quanta concentration (quanta/m3) in the morning (7:30 to 12am) and afternoon (12am to 4pm).") +
  geom_path(color = "white", size = 0.2) +
  theme_bw2() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        legend.position = "top", legend.justification='left', legend.direction='horizontal',
        legend.title = element_blank(), legend.key.width = unit(2.5, "cm"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())

figA

ggsave("../results/model-comparison.png", grid.arrange(figA, figB_alt, ncol = 1), width = 16 / cm(1), height = 16 / cm(1))
```

### 2023-02-22 preprocessing/match_clinical-xovis.R

@@ -1,0 +1,117 @@
+#### Libraries ####
+
+library(tidyverse)
+
+#### Helper ####
+
+
+#### Masi ####
+
+# Data
+
+clinical_masi <- readRDS("data-clean/Masi/clinical-data/clinical-data.rds")
+xovis_masi <- readRDS("data-clean/Masi/xovis/patients-time-location.rds")
+# env_masi <- readRDS("data-clean/Masi/environmental-data/co2-levels.rds")
+
+
+# Only first entry from clinical data
+
+clinical_masi <- clinical_masi %>%
+  group_by(patient_id, date) %>%
+  slice(1) %>%
+  ungroup()
+
+# summary xovis
+
+# xovis_reg_masi <- xovis_masi %>%
+#   group_by(PersonID) %>%
+#   summarise(Enter_Date = min(Time),
+#             Exit_Date = max(Time),
+#             InE = any(InE),
+#             Height = max(Height),
+#             TimepointEnteringtoRegistry = min(TimepointEnteringtoRegistry, na.rm = T)) %>%
+#   ungroup() %>%
+#   filter(Height >= 1400 & Height <= 2100,
+#          InE,
+#          !is.infinite(TimepointEnteringtoRegistry))
+
+xovis_reg_masi <- xovis_masi %>%
+  group_by(person_id) %>%
+  summarise(min_time = min(time),
+            max_time = max(time),
+            time_registration = min(time_registration, na.rm = T)) %>%
+  ungroup() 
+
+# pre-filter clinical data
+clinical_masi <- clinical_masi %>%
+  filter(date_time >= min(xovis_reg_masi$time_registration),
+         date_time <= max(xovis_reg_masi$time_registration))
+
+# matching clinical and xovis data
+#' match the date of first event in the clinical data 
+#' to the person that last came to the Entrance 
+#' with a tolerance of 15min
+
+merge_masi <- merge(xovis_reg_masi %>% select(person_id, time_registration),
+                    clinical_masi %>% select(patient_id, date_time, tb_suspect),
+                    id = NULL) %>%
+  mutate(timediff = as.numeric(difftime(date_time, time_registration), units = "secs")) %>%
+  filter(timediff <= 15 * 60,
[…]

### 2023-02-22 preprocessing/prep_clinical-data.R

@@ -1,0 +1,31 @@
+#### Libraries ####
+
+library(tidyverse)
+library(lubridate)
+
+
+#### Helper ####
+
+
+#### Data ####
+
+# Masi
+
+df_pat_time_masi <- read.csv("data-raw/Masi/clinical-data/patient-time.csv") %>%
+  rename(date = action_date) %>%
+  mutate(time = parse_date_time(paste(date, time), '%m/%d/%Y %I:%M:%S %p'),
+         date = as.Date(time)) %>%
+  filter(!is.na(time))
+
+df_tb_suspects <- read.csv("data-raw/Masi/clinical-data/tb-suspects.csv") %>%
+  select(patient_id, result_txt) %>%
+  rename(result = result_txt)
+
+df <- left_join(df_pat_time_masi, df_tb_suspects) %>%
+  rename(date_time = time) %>%
+  mutate(tb_suspect = ifelse(is.na(result), "no", "yes"),
+         time = format(date_time, "%H:%M:%S"),
+         date_time = as.POSIXct(as.character(date_time))) %>%
+  select(patient_id, date, time, date_time, tb_suspect)
+
+saveRDS(df, "data-clean/Masi/clinical-data/clinical-data.rds")

### 2023-02-22 preprocessing/prep_environmental-data.R

@@ -1,0 +1,21 @@
+#### Libraries ####
+
+library(tidyverse)
+library(haven)
+
+
+#### Masi ####
+
+df <- read_dta("data-raw/Masi/environmental-data/co2_data.dta") %>%
+  mutate(date = Date,
+         time = format(new_time, "%H:%M:%S"),
+         date_time = as.POSIXct(paste(Date, time)),
+         location = Monitor) %>%
+  rename(temperature = Temperature,
+         humidity = Humidity,
+         co2 = CO2) %>%
+  select(date, time, date_time, location, co2, temperature, humidity) 
+
+
+saveRDS(df, "data-clean/Masi/environmental-data/co2-levels.rds")
+  

### 2023-02-22 preprocessing/prep_room.R

@@ -1,0 +1,116 @@
+library(raster)
+library(tidyverse)
+library(terra)
+
+# read.data <- function(x) {
+# 
+#   message(sprintf("Reading file %s", x))
+# 
+#   data <- read.table(x, header = F, sep = ",", na.strings = "NONE", blank.lines.skip = F)
+#   data <- data[-1, ]
+#   if (ncol(data) > 22) {
+#     data <- data[ ,-ncol(data)]
+#   }
+#   colnames(data) <- data[1,]
+#   data <- data[-1, ]
+# 
+#   data <- data %>%
+#     filter(!is.na(PersonID)) %>%
+#     rename(LastX = `LastX `) %>%
+#     mutate(across(c(Time, ObservedSince, TimepointEnteringtoRegistry,TimepointclosestdistancetoRe),
+#                   ~ paste(Date, .x)),
+#            Date = as.Date(Date, format = "%d.%m.%Y"),
+#            Timeframe = factor(Timeframe, levels = c("Morning", "Midday", "Afternoon")),
+#            across(c(Time, ObservedSince, TimepointEnteringtoRegistry,TimepointclosestdistancetoRe),
+#                   ~ strptime(.x, format = "%d.%m.%Y %H:%M:%S")),
+#            across(c(LastX, LastY, Height, DistanceNextPerson, TotalMovement,
+#                     TotalPersons, TotalSitting, TotalWZ, TotalE, TotalTB,
+#                     RankEnteringtoRegistry, ClosestDistancetoRegistry),
+#                   as.numeric),
+#            across(c(RankEnteringtoRegistry, ClosestDistancetoRegistry),
+#                   ~ ifelse(.x == 0, NA, .x)),
+#            across(c(InWZ, InE, InTB), as.logical),
+#            PersonID = paste0(Date, "_ID:", PersonID))
+# 
+#   return(data)
+# 
+# }
+# 
+# 
+# #### Masi ####
+# 
+# # Data
+# 
+# masi_files <- list.files("data-raw/Masi/xovis", pattern = "Coordinates", full.names = T)
+# masi_files <- masi_files[-1]
+# df_masi <- map_dfr(masi_files, read.data) 
+# xy <- df_masi %>% dplyr::select(LastX, LastY) %>% na.omit %>% distinct() %>% as.matrix() 
+# 
+# plot(xy)
[…]

### 2023-02-22 preprocessing/prep_xovis.R

@@ -1,0 +1,481 @@
+#### Libraries ####
+
+library(tidyverse)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(gridExtra)
+library(VennDiagram)
+source("utils/spatial.r")
+source("utils/plotting.r")
+
+#### Parameters ####
+max_height_ratio <- 2
+min_height_ratio <- 0.5
+min_standing_height <- 1.4
+same_height_ratio_lower <- 0.95
+same_height_ratio_upper <- 1.05
+max_long_time <- 10 * 60 # maximum time (in sec) to look ahead for possible matches
+max_short_time <- 10 # maximum time (in sec) to look for short matches of people moving
+max_mov_distance <- 3 # maximum distance (in m) to look for someone moving
+max_sisa_distance <- 1 # maximum distance (in m) to look for someone who is sitting down or standing up
+min_duration <- 5 # minimum duration (in min) to stay in the clinic
+
+#### Data ####
+
+# xovis
+
+masi <- read.csv("data-raw/Masi/xovis/2021_10_25_coordinates_only.csv", header = F) %>%
+  set_names(c("obs_id", "time", "x", "y", "height")) %>%
+  mutate(time = as.POSIXct(paste0("2021-10-25 ", time), format = "%Y-%m-%d %H:%M:%S"))
+
+# roomplan
+clinic <- vect("data-raw/Masi/building/clinic-vector.gpkg")
+clinic_sf <- sf::st_as_sf(clinic, crs = NA)
+st_crs(clinic_sf) <- NA
+clinic_df <- fortify(clinic_sf)
+
+
+# find polygon
+masi$point <- map2(masi$x, masi$y, function(xi,yi) sfheaders::sf_point(c(xi,yi)))
+masi$poly = map(masi$point, sf::st_within, y = clinic_sf)
+masi$is_entrance <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 1), T, F))
+masi$is_waitingroom <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 2), T, F))
+masi$is_passage <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 3), T, F))
+masi$is_tbroom <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 4), T, F))
+masi$is_reception <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 5), T, F))
[…]

### 2023-02-22 utils/distr.r

@@ -1,0 +1,12 @@
+lnorm_moments <- function(meanlog, sdlog) {
+  m <- exp(meanlog + (1/2)*sdlog^2)
+  s <- exp(meanlog + (1/2)*sdlog^2)*sqrt(exp(sdlog^2) - 1)
+  return(list(mean = m, sd = s))
+}
+
+
+inv_lnorm_moments <- function(mean, sd) {
+  m <- log(mean) - 0.5 * log((sd/mean)^2 + 1)
+  s <- sqrt(log((sd/mean)^2 + 1))
+  return(list(meanlog = m, sdlog = s))
+}

### 2023-02-22 utils/plotting.R

@@ -1,0 +1,93 @@
+require(ggpattern)
+require(gridExtra)
+require(GGally)
+library(ggtext)
+
+# ggplot theme for Nature style
+text_size = 8
+update_geom_defaults("text", list(size = text_size))
+theme_nature <- function () { 
+  theme_classic(base_size = text_size, base_family = 'sans') %+replace% 
+    theme(
+      axis.text = element_text(size = text_size),
+      axis.title = element_text(size = text_size),
+      plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0.5, margin = ggplot2::margin(0, 0, 5, 0))
+    )
+}
+
+theme_bw2 <- function () { 
+  theme_bw(base_size = text_size, base_family = 'sans') %+replace% 
+    theme(
+      axis.text = element_text(size = text_size),
+      axis.title = element_text(size = text_size),
+      plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 5, 0))
+    )
+}
+
+# Correlation plot 
+# Helpful code:
+# https://stackoverflow.com/questions/44961437/how-to-include-density-coloring-in-pairwise-correlation-scatter-plot
+# https://stackoverflow.com/questions/37889222/change-colors-in-ggpairs-now-that-params-is-deprecated?answertab=votes#tab-top
+contours <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    stat_density2d() +
+    scale_x_continuous(breaks = seq(-.6,0,.2)) +
+    scale_y_continuous(breaks = seq(-.6,0,.2))
+  p
+}
+
+# Plot for correlation
+contours <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    stat_density2d() #+
+  #scale_x_continuous(breaks = seq(-.8,.4,.4)) +
+  #scale_y_continuous(breaks = seq(-.8,.4,.4))
+  p
+}
+hex <- function(data, mapping, ...){
+  p <- ggplot(data = data, mapping = mapping) + 
+    geom_hex() #+
+  #scale_x_continuous(breaks = seq(-.8,.4,.4)) +
+  #scale_y_continuous(breaks = seq(-.8,.4,.4))
+  p
+}
[…]

### 2023-02-22 utils/spatial.r

@@ -1,0 +1,185 @@
+#' Compute euclidean distance in two dimensions
+#' 
+#' @param x1 x coordinate of first point
+#' @param x2 x coordinate of second point
+#' @param y1 y coordinate of first point
+#' @param y2 y coordinate of second point
+#' 
+
+euclidean <- function(x1, x2, y1, y2) {
+  sqrt((x1-x2)^2) + sqrt((y1-y2)^2)
+}
+
+
+#' Convert distance from px to m
+#' 
+#' @param x distance (in px)
+#' 
+
+convert_dist <- function(x) {
+  x / 1e3
+}
+
+
+#' Convert shape to rastered spatial polygon data frame
+#' 
+#' @param sh shape
+#' @param cell_size length/width of squared cell (in px)
+#' 
+
+shapeToSpatial <- function(sh, cell_size) {
+  ng <- sf::st_make_grid(sh, cellsize = cell_size) 
+  ng <- sf::st_sf(ng)
+  ng <- sf::st_intersection(ng, sf::st_geometry(sf::st_as_sf(sh)))
+  ng_types <- vapply(sf::st_geometry(ng), function(x) {class(x)[2]}, "")
+  ng <- ng[grepl("*POLYGON", ng_types), ]
+  roomSP <- sf::as_Spatial(st_sf(ng))
+  return(roomSP)
+}
+
+
+#' Create data frame with coordinates (x,y,id) from spatial polygon data frame
+#' 
+#' @param spat_poly_df spatial polygon data frame
+#' 
+
+create_coord_df <- function(spat_poly_df) {
+  cellCoords <- data.frame(coordinates(spat_poly_df)) 
+  colnames(cellCoords) <- c("x", "y")
+  cellCoords$id <- 1:nrow(cellCoords)
+  return(cellCoords)
+}
+
+
+#' Compute volume from spatial polygon data frame 
+#' 
+#' @param spat_poly_df spatial polygon data frame
+#' @param H height of the room
+
+compute_volume <- function(spat_poly_df, H = 3) {
+  A <- convert_dist(convert_dist(sum(area(spat_poly_df))))
+  V <- A * H
+  return(V)
+}
+
+
+#' Find raster cell id for data point based on euclidean distance
+#' 
+#' @param x x coordinate of data point
+#' @param y y coordinate of data point
+#' @param cellCoordsXY matrix of x and y coordinates of cells
+#' 
+
+find_raster <- function(x, y, cellCoordsXY) {
+  # assigns cell id by closest distance to cell center coordinates
[…]

### 2023-02-22 utils/trans_risk.r

@@ -1,0 +1,69 @@
+#' Compute the quanta removal rate
+#' 
+#' 
+#' @param AER air exchange rate per hour
+#' @param k rate per hour for particle deposition on surfaces
+#' @param lambda viral inactivation rate per hour
+#' 
+
+compute_RR <- function(AER, k = 0, lambda = 0) {
+  return(AER + k + lambda)
+}
+
+
+#' Compute air exchange rate per hour
+#' 
+#' @param G CO2 generation rate in L/s per person
+#' @param Co CO2 concentration in outdoor air (in ppm)
+#' @param C CO2 concentration (in ppm)
+#' @param V volume of ventilated space (in m3)
+#' @param n number of people in ventilated space (in 1/m3) 
+#' 
+
+compute_AER <- function(G, Co, C, n, V) {
+  # add 1 person for the person working behind the registry (and such that n is not 0)
+  n <- n + 1
+  # excess CO2 in ppm
+  Ce <- C - Co
+  # TODO: is there are more elegant way to ensure that the denominator is not 0?
+  Ce <- ifelse(Ce == 0, 1, Ce) 
+  # excess CO2 in l/m3 (1l/m3 = 1,000ppm)
+  Ce <- Ce / 1000
+  # ventilation rate (in l/s per person)
+  Q <- G / Ce
+  # air exchange rate (in s)
+  AER <- Q * n / V
+  # air exchange rate (in h)
+  AER <- 3600 * AER
+  return( AER )
+}
+
+
+#' Compute quanta concentration (quanta / m3)
+#' 
+#' 
+#' @param E quanta emission (quanta emission rate times the number of infectious)
+#' @param I number of infectious people in space
+#' @param RR quanta removal rate
+#' @param N0 initial quanta concentration
+#' @param V volume of space (in 1/m3) 
+#' 
+#' @references Buonanno G, Stabile L, Morawska L. Estimation of airborne viral emission:
+#' Quanta emission rate of SARS-CoV-2 for infection risk assessment. 
+#' Environment international. 2020;141:105794.
+
+compute_Nt <- function(E, RR, N0, V) {
+  return( E / (RR * V) + (N0 + E / (RR * V)) * exp(-RR) )
+}
+
+
+#' Compute risk of infection
+#' 
+#' 
+#' @param N quanta concentration (quanta / m3)
+#' @param t duration of exposure (in minutes)
+#' 
+
+compute_P <- function(N, t = 1) {
+  1 - exp(-N * t)
+} 

### 2023-03-03 preprocessing/prep_room.R
first complete code preprocessing

@@ -86,26 +86,33 @@ sa_wr_bottom_left_pt <- line_point(wr[3,], sa_wr_width, rotation, start = F)
 sa_wr <- parallelogram(sa_wr_bottom_left_pt, wr[3,], sa_wr_width, sa_wr_breadth, rotation)
 
 # TB room entrance
-er_tb_width <- 850
-er_tb_breadth <- 1100
-er_tb_bottom_left_pt <- line_point(tb[3, ], er_tb_width, rotation, start = F)
-er_tb <- parallelogram(er_tb_bottom_left_pt, tb[3,], er_tb_width, er_tb_breadth, rotation)
+# er_tb_width <- 850
+# er_tb_breadth <- 1100
+# er_tb_bottom_left_pt <- line_point(tb[3, ], er_tb_width, rotation, start = F)
+# er_tb <- parallelogram(er_tb_bottom_left_pt, tb[3,], er_tb_width, er_tb_breadth, rotation)
 
 # seating area passage left to TB room
-sa_pa_leftTB_width <- tb_width - er_tb_width
-sa_pa_leftTB_breadth <- er_tb_breadth
-sa_pa_leftTB_bottom_right_pt <- line_point(pa[1,], sa_pa_leftTB_width, rotation)
-sa_pa_leftTB <- parallelogram(pa[1, ], er_tb_bottom_left_pt, sa_pa_leftTB_width, sa_pa_leftTB_breadth)
+sa_pa_leftTB_width <- 3250
+sa_pa_leftTB_breadth <- 1200
+sa_pa_leftTB_bottom_left_pt <- line_point(tb[3,], sa_pa_leftTB_width, rotation, start = F)
+sa_pa_leftTB <- parallelogram(sa_pa_leftTB_bottom_left_pt, tb[3, ], sa_pa_leftTB_width, sa_pa_leftTB_breadth)
 
 # seating area passage right to TB room
-sa_pa_rightTB_width <- 2300
-sa_pa_rightTB_breadth <- er_tb_breadth
+sa_pa_rightTB_width <- 2450
+sa_pa_rightTB_breadth <- sa_pa_leftTB_breadth
 sa_pa_rightTB_bottom_right_pt <- line_point(tb[3,], sa_pa_rightTB_width, rotation)
 sa_pa_rightTB <- parallelogram(tb[3,], sa_pa_rightTB_bottom_right_pt, sa_pa_rightTB_width, sa_pa_rightTB_breadth, rotation)
[…]

### 2023-03-03 preprocessing/prep_xovis.R

@@ -1,500 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-library(reshape2)
-library(lubridate)
-library(sf)
-library(sfheaders)
-library(raster)
-library(terra)
-library(gridExtra)
-library(VennDiagram)
-source("utils/spatial.r")
-source("utils/plotting.r")
-
-#### Parameters ####
-max_height_ratio <- 2
-min_height_ratio <- 0.5
-min_standing_height <- 1.4
-same_height_ratio_lower <- 0.9
-same_height_ratio_upper <- 1.1
-max_long_time <- 10 * 60 # maximum time (in sec) to look ahead for possible matches
-max_mov_time <- 10 # maximum time (in sec) to look for short matches of people moving
-max_sit_time <- 5 * 60 # maximum time to wait until someone is standing up again
-max_mov_distance <- 3 # maximum distance (in m) to look for someone moving
-max_sisa_distance <- 1 # maximum distance (in m) to look for someone who is sitting down or standing up
-min_duration <- 5 # minimum duration (in min) to stay in the clinic
-
-#### Data ####
-
-# xovis
-
-masi <- read.csv("data-raw/Masi/xovis/2021_10_25_coordinates_only.csv", header = F) %>%
-  set_names(c("obs_id", "time", "x", "y", "height")) %>%
-  mutate(time = as.POSIXct(paste0("2021-10-25 ", time), format = "%Y-%m-%d %H:%M:%S"))
-
-# roomplan
-clinic <- vect("data-raw/Masi/building/clinic-vector.gpkg")
-clinic_sf <- sf::st_as_sf(clinic, crs = NA)
-st_crs(clinic_sf) <- NA
-clinic_df <- fortify(clinic_sf)
-
-
-# find polygon
-masi$point <- map2(masi$x, masi$y, function(xi,yi) sfheaders::sf_point(c(xi,yi)))
-masi$poly = map(masi$point, sf::st_within, y = clinic_sf)
-masi$is_entrance <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 1), T, F))
-masi$is_waitingroom <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 2), T, F))
-masi$is_passage <- map_lgl(masi$poly, function(p) ifelse(any(unlist(p) == 3), T, F))
[…]

### 2023-03-03 preprocessing/prep_xovis_matching.R

@@ -1,0 +1,559 @@
+#### Libraries ####
+
+library(tidyverse)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(gridExtra)
+library(VennDiagram)
+source("utils/spatial.r")
+source("utils/plotting.r")
+
+
+#' Algorith to match observation IDs
+#' 
+#' @param date date %Y-%m-%d
+#' @param max_time maximum time (in seconds) to look ahead for matches
+#' @param min_standing_height minimum standing height
+#' @param max_mov_time maximum time (in seconds) to look ahead for people moving
+#' @param max_sit_time maximum time (in seconds) to look ahead for people sitting
+#' @param max_tbroom_time maximum time (in seconds) to look ahead for someone being in the TB room
+#' @param max_mov_distance maximum distance someone can make while moving
+#' @param max_sit_distance maximum distance someone can make while sitting
+#' @param max_heightdiff maximum height difference for a person not changing her posture
+#' @param min_duration minimum time (in minutes) to stay in the clinic to be considered
+
+match_xovis <- function(date = "2021-10-25", 
+                        max_time = 30 * 60, 
+                        min_standing_height = 1400, 
+                        max_mov_time = c(3, 5, 5, 5, 10, 10, 15, 15, 30, 60), 
+                        max_sit_time = c(5, 10, 20, 30, 45, 60, 90, 150, 300, 600),
+                        max_tbroom_time = 300,
+                        max_mov_distance = c(0.5, 0.5, 1, 2, 2, 3, 3, 3.5, 4, 4),
+                        max_sit_distance = c(0.25, 0.25, 0.5, 0.5, 0.5), 
+                        max_heightdiff = c(50, 50, 75, 100, 100, 125, 125, 150, 175, 200), 
+                        min_duration = 5 * 60
+                        ) {
+  
+  #### Data ####
+  
+  # xovis
+  file <- paste0("data-raw/Masi/xovis/", date, "-with_roominfo.rds")
+  save_dir <- paste0("data-clean/Masi/patient-tracking-data/", date, "/")
+  if (!dir.exists(save_dir)) {
+    dir.create(save_dir)
+  }
+  masi <- readRDS(file) 
+  
+  n_oid <- length(unique(masi$obs_id))
+  
+  # remove very short tracks
+  
+  masi <- masi %>%
+    group_by(obs_id) %>%
+    filter(! ( as.numeric(difftime(last(time), first(time), units = "secs")) < 10  &
[…]

### 2023-03-03 preprocessing/prep_xovis_room.R

@@ -1,0 +1,55 @@
+#### Libraries ####
+
+library(tidyverse)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(VennDiagram)
+source("utils/spatial.r")
+
+# roomplan
+clinic <- vect("data-raw/Masi/building/clinic-vector.gpkg")
+clinic_sf <- sf::st_as_sf(clinic, crs = NA)
+st_crs(clinic_sf) <- NA
+clinic_df <- fortify(clinic_sf)
+
+# xovis raw data files
+files <- list.files("data-raw/Masi/xovis/Study_Data_Masi_2021/export", full.names = T)
+
+for (f in files) {
+  # file info
+  file_name <- basename(f)
+  date <- gsub("_exported_data[[:punct:]]csv", "", file_name)
+  date <- gsub("_", "-", date)
+  file_name <- paste0(date, "-", "with_roominfo")
+  
+  
+  # read data
+  df <- read.csv(f, header = T) %>% 
+    set_names(c("obs_id", "time", "x", "y", "height"))  %>%
+    mutate(time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M:%S"))
+  
+  # add room info
+  df$point <- map2(df$x, df$y, function(xi,yi) sfheaders::sf_point(c(xi,yi)))
+  df$poly = map(df$point, sf::st_within, y = clinic_sf)
+  df$is_waitingroom <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 2), T, F))
+  df$is_tbroom <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 3), T, F))
+  df$is_passage <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 4), T, F))
+  df$is_reception <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 5), T, F))
+  df$is_entrance <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 6), T, F))
+  df$is_inside_reception <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 8), T, F))
+  df$is_seat = map_lgl(df$poly, function(p) ifelse(any(unlist(p) %in% c(7, 9, 10, 11)), T, F) ) 
[…]

### 2023-03-03 utils/spatial.r

@@ -83,211 +83,3 @@ find_raster <- function(x, y, cellCoordsXY) {
   return(closest_id)
 }
 
-
-#' Plot individual tracks from patients with linked observations
-#' 
-#' @param df data frame with columns patient_id, obs_id, time, x, y, and height
-#' 
-
-plot_single_track <- function(df) {
-  
-  # Descriptives
-  descr <- df %>%
-    mutate(dt = as.numeric(difftime(lead(time), time, units = "secs"))) %>%
-    summarize(
-      Total = as.numeric(difftime(last(time), first(time), units = "secs")),
-      `Time in waiting room` = sum(dt[is_waitingroom], na.rm = T),
-      `Time in passage` = sum(dt[is_passage], na.rm = T),
-      `Time in TB room` = sum(dt[is_tbroom], na.rm = T),
-      `Time in seating area` = sum(dt[is_seating], na.rm = T),
-      `Time in care room` = sum(dt[is_in_room1 | is_in_room2], na.rm = T),
-      `Time at reception` = sum(dt[is_reception], na.rm = T)
-    ) %>%
-    gather() %>%
-    mutate(value = ifelse(key == "Time at reception", paste0(value, "sec"), paste0(format(round(value / 60, 1), nsmall = 1), "min"))) %>%
-    set_names(c("Variable", "Duration"))
-  
-  tt <- ttheme_default(base_size = 6)
-  
-  # Start and end of patient track
-  se <- df %>%
-    group_by(patient_id) %>%
-    arrange(time) %>%
-    slice(c(1, n())) %>%
-    mutate(type = c("start", "end")) %>%
-    ungroup() 
-  
-  # Start and end of each observation
-  se_obs <- df %>%
-    group_by(patient_id, obs_id) %>%
-    arrange(time) %>%
-    slice(1, n()) %>%
-    ungroup() 
-  
-  # Start and end of each observation
-  tab <- se_obs %>%
-    mutate(height = round(height / 10, digits = 0),
-           time = format(time, "%H:%M:%S")) %>%
-    dplyr::select(obs_id, time, x, y, height) %>%
-    mutate_all(as.character)
-  
-  # Start and end of each observation in plot
-  se_obs <- se_obs %>%
-    slice(c(-1, -n())) 
-  
-  # Plot 
-  pl <- ggplot() +
-    geom_sf(data = clinic_df, linewidth = 1, fill = NA) +
[…]

### 2023-03-09 analysis/spatiotemporal-transmission-risk.Rmd
Update preprocessing, descriptive plots, and utils

@@
 library(reshape2)
-library(LaplacesDemon)
 library(rriskDistributions)
@@
 library(ggridges)
+library(grid)
 library(ggplot2)
+library(ggrepel)
 
@@
 source("../utils/plotting.R")
+
+# plot with and height of input and data graphs
+aw <- 12
+ah <- 8
 ```
+
+## Data
+
+### Clinical
+
+```{r}
+clinical_dta <- readRDS("../data-clean/Masi/clinical-data/clinical-data.rds")
+share_tb <- clinical_dta %>%
+  group_by(date, patient_id) %>%
+  summarize(tb = any(tb_suspect == "yes")) %>%
+  ungroup() %>%
+  group_by(date) %>%
+  summarize(n = n(),
+            n_tb = sum(tb)) %>%
+  arrange(date) %>%
+  mutate(prop_tb = n_tb / n * 100,
+         date = format(date, format = "%b %d"),
+         date = factor(date, levels = unique(date))) %>%
+  ungroup() 
+share_tb_pl <- share_tb %>%
+  ggplot(aes(x = date, y = n_tb)) +
+  geom_bar(stat = "identity", fill = "grey", color = "black", width = .5) +
+  annotate("text", x = "Nov 09", y = mean(share_tb$n_tb), label = paste0("Mean = ",round(mean(share_tb$n_tb), 1)), color = "darkred", size = 8 / cm(1), vjust = -1) +
+  geom_hline(aes(yintercept = mean(share_tb$n_tb)), color = "darkred") +
+  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
+  scale_x_discrete(breaks = c("Oct 11", "Oct 18", "Oct 25", "Nov 02", "Nov 09", "Nov 16")) +
+  labs(y = "No. of TB patients") +
+  theme_bw2() +
+  theme(axis.title.x = element_blank(),
+        plot.margin = margin(5.5, 10, 5.5, 5.5))
+save_plot(share_tb_pl, pdf_file = "../results/data/share-tb-patients.pdf", w = aw, h = ah)
+```
+
+### CO2
+
+```{r}
+co2_pl <- readRDS("../data-clean/Masi/environmental-data/co2-temp-humid.rds") %>%
+  filter(location %in% c("waiting room", "tb room")) %>%
+  mutate(location = factor(ifelse(location == "tb room", "TB room", "Waiting room"), levels = c("Waiting room", "TB room")),
+         time = as.POSIXct(time, format = "%H:%M", tz = "CET")) %>%
+  ggplot(aes(x = time, y = co2, color = location, fill = location)) +
+  stat_lineribbon(alpha = .25, size = .5) +
+  scale_x_datetime(expand = c(0,0), date_breaks = "2 hours", date_labels = "%H:%M") +
+  scale_y_continuous(expand = c(0,0), breaks = scales::breaks_width(width = 100)) +
+  scale_fill_brewer(palette = "Accent") +
+  scale_color_brewer(palette = "Accent") +
+  labs(y = expression("CO"[2]*" (ppm)"), fill = "Quantile", color = "Quantile") +
+  theme_bw2() +
+  theme(legend.position = c(.8,.95), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        plot.margin = margin(5.5, 10, 5.5, 5.5), axis.title.x = element_blank(), legend.title = element_blank()) 
+save_plot(co2_pl, pdf_file = "../results/data/co2-levels.pdf", w = aw, h = ah)
+```
+
+### Building
+
+```{r}
+# background image
+clinic_img <- png::readPNG("../data-raw/Masi/building/clinic_clipped.png")
+```
+
+<!-- ```{r} -->
+<!-- # choose cell size such that the squared cell has 1m3 volume -->
+<!-- cellVolume <- 1 -->
+<!-- cellSize <- sqrt(cellVolume * 1e9 / 3e3)  -->
+
+<!-- # waiting room -->
+<!-- sh_wr <- vect("../data-raw/Masi/building/waiting_room.shp") -->
+<!-- sf_wr <- st_as_sf(sh_wr) -->
+<!-- wrRoomSP <- shapeToSpatial(sh_wr, cellSize) -->
+<!-- wrVol <- compute_volume(wrRoomSP) -->
+<!-- wrCellCoords <- create_coord_df(wrRoomSP) -->
+<!-- wrRoomXY <- as.matrix(dplyr::select(wrCellCoords, x, y)) -->
+
+<!-- # tb room -->
+<!-- sh_tb <- vect("../data-raw/Masi/building/tb_room.shp") -->
+<!-- sf_tb <- st_as_sf(sh_tb) -->
+<!-- tbRoomSP <- shapeToSpatial(sh_tb, cellSize) -->
+<!-- tbVol <- compute_volume(tbRoomSP) -->
+<!-- tbCellCoords <- create_coord_df(tbRoomSP) -->
+<!-- tbRoomXY <- as.matrix(dplyr::select(tbCellCoords, x, y)) -->
+
+<!-- # passage -->
+<!-- sh_pa <- vect("../data-raw/Masi/building/passage.shp") -->
+<!-- sf_pa <- st_as_sf(sh_pa) -->
+<!-- paRoomSP <- shapeToSpatial(sh_pa, cellSize) -->
+<!-- paVol <- compute_volume(paRoomSP) -->
+<!-- paCellCoords <- create_coord_df(paRoomSP) -->
+<!-- paRoomXY <- as.matrix(dplyr::select(paCellCoords, x, y)) -->
+
+<!-- # combine -->
+<!-- clinicCellCoords <- rbind( -->
+<!--   mutate(wrCellCoords, location = "waiting room"), -->
+<!--   mutate(tbCellCoords, location = "tb room"), -->
+<!--   mutate(paCellCoords, location = "passage") -->
+<!-- ) %>% mutate(id = 1:nrow(.)) -->
+<!-- clinicCellCoordsXY <- cbind(clinicCellCoords$x, clinicCellCoords$y) -->
+<!-- ``` -->
+
+### No. people
+
+```{r}
+files <- list.files("../data-clean/Masi/patient-tracking-data/", full.names = T)
+no_people <- do.call(rbind, lapply(files, function(x) readRDS(paste0(x, "/patient-id-data.rds")))) %>%
+  mutate(date = format(time, format = "%Y-%m-%d"),
+         time = as.POSIXct(time, format = "%H:%M", tz = "CET"),
+         daytime = ifelse(time <= "2021-10-25 12:00:00", "Morning", "Afternoon"),
+         daytime = factor(daytime, levels = c("Morning", "Afternoon")))
+
+# over time
+no_people_time_pl <- no_people %>%
+  group_by(date, time) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  ggplot(aes(x = time, y = n)) +
+  geom_line() +
+  #stat_lineribbon(alpha = .25, size = .5) +
+  scale_x_datetime(expand = c(0,0), date_breaks = "2 hours", date_labels = "%H:%M") +
+  scale_y_continuous(expand = c(0,0), limits = c(0, NA), breaks = scales::breaks_width(5)) +
+  scale_fill_brewer(palette = "OrRd") +
+  labs(y = "No. of clinical attendees") +
+  theme_bw2() +
+  theme(axis.title.x = element_blank())
+  #labs(y = "No. of clinical attendees", fill = "Quantile", color = "Quantile") +
+  # theme(legend.position = c(.8,.9), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+  #       plot.margin = margin(5.5, 10, 5.5, 5.5)) +
+  # guides(fill = guide_legend(title.position = "top"))
+save_plot(no_people_time_pl, pdf_file = "../results/data/no-people-over-time.pdf", w = aw, h = ah)
+
+# Spatial
+no_people_spatial_pl <- no_people %>%
+  dplyr::select(x, y) %>%
+  adespatial::rotation(angle = -8*pi/180) %>%
+  as.data.frame() %>%
+  set_names(c("x", "y")) %>%
+  mutate(x = -x) %>%
+  ggplot(aes(x = x, y = y)) +
+  geom_bin2d(binwidth = 100) +
+  ggpubr::background_image(clinic_img) +
+  annotate("text", x = -13200, y = -5300, label = "Entrance", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
+  annotate("text", x = -13400, y = -2300, label = "Waiting room", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
+  annotate("text", x = -5000, y = -4100, label = "Registration", size = 8 / cm(1), fontface = 2, angle = 90, hjust=0,vjust=0) +
+  annotate("text", x = 2600, y = -2700, label = "TB room", size = 8 / cm(1), fontface = 2) +
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = c(1,5,10,50,100), limits = c(0,100)) +
+  scale_y_continuous(expand = expansion(mult = c(0,.18))) +
+  scale_x_continuous(expand = expansion(mult = c(0,0.025))) +
+  labs(fill = "Density") +
+  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
+        legend.position = c(0.825,0.075), legend.direction = "horizontal", legend.title = element_blank(), #legend.title = element_text(vjust = .75),
+        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 3, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+save_plot(no_people_spatial_pl, pdf_file = "../results/data/no-people-spatial.pdf", w = aw, h = ah)
+```
+
+<!-- ```{r} -->
+<!-- # number of people in space -->
+<!-- nDF <- tibble() -->
+<!-- for (tp in time_period) { -->
+<!--   n_tp <- clinXov %>% -->
+<!--     filter(between(time, tp, tp + seconds(59))) %>% -->
+<!--     mutate(location = mapply(function(x,y) clinicCellCoords$location[find_raster(x,y,clinicCellCoordsXY)], x, y)) %>% -->
+<!--     group_by(person_id, location) %>% -->
+<!--     slice(1) %>% -->
+<!--     ungroup() %>% -->
+<!--     group_by(location) %>% -->
+<!--     summarize(n = n()) %>% -->
+<!--     ungroup()  -->
+<!--   for (loc in unique(clinicCellCoords$location)) { -->
+<!--     if (!(loc %in% n_tp$location)) { -->
+<!--       n_tp <- rbind(n_tp, data.frame(location = loc, n = 0)) -->
+<!--     }  -->
+<!--   } -->
+<!--   nDF <- rbind(nDF, n_tp %>% mutate(time = tp)) -->
+<!-- } -->
+<!-- nDF <- nDF %>%  -->
+<!--   mutate(time = as.POSIXct(time, origin = as.Date("1970-01-01"))) -->
+<!-- nDF$n[nDF$location=="waiting room"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"] -->
+<!-- nDF$n[nDF$location=="passage"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"] -->
+<!-- ggplot(nDF, aes(x = time, y = n, color = location)) + geom_line() -->
+<!-- ``` -->
+
+<!-- ### Movement -->
+
+<!-- ```{r} -->
+<!-- patMov <- clinXov %>% -->
+<!--   mutate(id = map2_int(x, y, find_raster, cellCoordsXY = clinicCellCoordsXY)) -->
+<!-- ``` -->
 
@@
 
+```{r}
+n_samples <- 1000
+set.seed(12345)
+```
+
 ### Spatial emission
 
-* Ko et al. (2004): exponential decrease on an airplane, but they are more strongly ventilated
-* Abkarian et al. (2020) deduce that, for the respiratory jet generated by typical speaking, the concentration of pathogen is diminished to approximately 3% of its initial value at a distance of 2 m.
-* SSPH report: distance depends on size of aerosols and initial speed, which depends on the activity. For speaking, it will be around 1-2m, for coughing or sneezing it can be up to 7-8m
-* Wang et al. (2021) and other studies: viral concentration is highest closest to the source
+* Wang (2021): 
+  * Travel distance of particles depends on particle size, activity, and environmental factors
+  * Smaller aerosols can travel further, stay in the air for a longer time, and typically contain higher viral load
+  * The concentration of exhaled aerosols is highest close to the source and decreases with distance
+  * Short-range aerosol transmission difficult to prevent with ventilation
+* Chen (2020): Short-range aerosols dominate over most distance up to 2m, droplets only play a role at very close distance <0.5
+* Ko (2004): exponential decrease of potential infections on an airplance (high ventilation) with distance to infector
 
@@
 ```{r}
-rMuIc <- function(n) {
-  rgamma(n, shape = 10, scale = .2)
-}
-
-hist(rMuIc(1e3))
-
-dIc <- function(distance, mu) { 
-  1 - pexp(distance, 1 / mu)
-}
-
-expand.grid(distance = seq(0, 10, .1), rate = rMuIc(1e3)) %>%
-  mutate(dens_inf = dIc(distance, rate)) %>%
-  ggplot(aes(x = distance, y = dens_inf)) +
-  stat_lineribbon() +
-  scale_fill_brewer() +
+relQC_pl <- expand.grid(distance = seq(0, 10, .1), 
+                           mu = rMuIc(n_samples)) %>%
+  mutate(relQC = dIc(distance, mu) * 100) %>%
+  ggplot(aes(x = distance, y = relQC)) +
+  stat_lineribbon(alpha = .5, size = .5) +
   scale_x_continuous(expand = c(0,0)) +
-  scale_y_continuous(expand = c(0,0)) 
+  scale_y_continuous(expand = c(0,0)) +
+  scale_fill_brewer(palette = "OrRd") +
+  labs(x = "Distance (m)", y = "Relative quanta concentration (%)", fill = "Quantile") +
+  theme_bw2() +
+  theme(legend.position = c(.8,.9), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        plot.margin = margin(5.5, 10, 5.5, 5.5)) +
+  guides(fill = guide_legend(title.position = "top"))
+save_plot(relQC_pl, pdf_file = "../results/inputs/spatial-relative-quanta-concentration.pdf", w = aw, h = ah)
 ```
@@
 
-* Catanzaro (1982): 249 quanta/h 
-* Nardell (1991): 12.7 quanta/h
-* Riley (1962): 1.25 quanta/h but patients already started on chemotheraphy and also few highly infectious patients
-* Escombe (2008): 8.2 quanta/h but wide range, including an extreme outlier of 226 quanta/h
-* Andrews (2014): 0.44 to 5.69 quanta/h
+* Riley (1962): 130 patients, q: 1.25
+* Escombe (2008): 117 patients, q: 8.2; but high values above 100 are possible
+* Nardell (1991) : 1 patients, q: 12.5
+* Andrews (2014) : 571 patients, q: 0.89
 
---> The distribution should roughly reflect the following quantiles
-* <1: 10%
-* <5: 50%
-* <10: 75%
-* <50: 95%
-* <100: 99%
+--> Take the mean and standard deviation of these values and use a student-t distribution with small degrees of freedom to consider outliers
 
 ```{r}
-q_a <- c(1, 5, 10, 50, 100, 200)
-p_a <- c(.1, .5, .75, .95, .99, 1)
-q_range <- seq(0, 200, .1)
-p_q_range <- ptrunc(q_range, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
-p_a_selected <- ptrunc(q_a, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
-lnorm_par <- inv_lnorm_moments(5, 10)
-p_q_range_lnorm <- plnorm(q_range, meanlog = lnorm_par$meanlog, sdlog = lnorm_par$sdlog)
+q <- c(1.25, 8.2, 12.5, 0.89)
+n_q <- c(130, 117, 1, 571)
+mean_q <- sum(n_q * q) / sum(n_q)
+sd_q <- sqrt( sum( (n_q / sum(n_q) * (q - mean_q) ^ 2 ) ) )
 
-ggplot(mapping = aes(x = q_range)) +
-  geom_line(mapping = aes(y = p_q_range), color = "black") +
-  geom_line(mapping = aes(y = p_q_range_lnorm), color = "red") +
-  geom_point(mapping = aes(x = q_a, y = p_a_selected), color = "black") +
-  geom_point(mapping = aes(x = q_a, y = p_a), color = "blue") +
-  scale_x_log10(breaks = q_a) +
-  scale_y_continuous(breaks = p_a)
+qLiterature <- data.frame(q = q, 
+                          y = .01, 
+                          lab = c("Riley", "Escombe", "Nardell", "Andrews"))
 
-rq <- function(n) {
-  rtrunc(n, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
-}
-
-dq <- function(x) {
-  dtrunc(x, spec = "st", a = 0, b = 300, mu = 4, sigma = 4, nu = 1)
-}
-
-rq_distr <- data.frame(quanta = seq(0, 100, .1)) %>% mutate(prob = dq(quanta))
-
-ggplot(rq_distr, aes(x = quanta, y = prob)) +
-  geom_line()
+q_pl <- data.frame(q = seq(0, 40, .1)) %>%
+  mutate(density = dq(q)) %>%
+  ggplot(aes(x = q, y = density)) +
+  geom_line() +
+  geom_vline(mapping = aes(xintercept = 2), linetype = "dashed", color = "darkred") +
+  annotate("text", x = 3, y = 0.175, hjust = 0, label = "Mean = 2.0", size = 8 / cm(1), color = "darkred") +
+  geom_point(data = qLiterature, mapping = aes(x = q, y = y, color = lab, shape = lab)) +
+  geom_text_repel(data = qLiterature, mapping = aes(x = q, y = y, color = lab, label = lab), size = 8 / cm(1)) +
+  annotate("text", x = 30, y = 0.175, label = expression("q ~ Student-t("*mu*"=2.0,"*sigma*"=2.5"*","*nu*"=1)"), size = 8 / cm(1)) +
+  scale_x_continuous(expand = c(0,0), limits = c(0, 40)) +
+  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
+  scale_color_brewer(palette = "Dark2", direction = -1) +
+  labs(x = "Quanta emission (log q/h)", y = "Density") +
+  theme_bw2() +
+  theme(plot.margin = margin(5.5, 10, 5.5, 5.5), legend.position = "none")
+save_plot(q_pl, pdf_file = "../results/inputs/quanta-emission.pdf", w = aw, h = ah)
 ```
 
-### Average volume of exhaled gas
+### Unmasked TB patients
 
-* Richardson et al. (2014): 0.1 - 1.17 l/s per person based on Feher (2017)
+* Berhanu (2023): 
+  * study group: clinic attendees living with HIV, contact with TB patients in past year, or had TB in the past 2 years
+  * finding: 55% of  with a sputum result positive for M. tuberculosis did not have a positive TB symptom screen
+* Hamada (2018): The WHO 4-question symptom screen (cough, fever, weight loss, and night sweats) misses up to half the TB cases among people living with human immunodeficiency virus (HIV) on antiretroviral therapy (ART)
+--> Unmasked TB patients: Poisson(mu = No. of TB suspects)
 
 ```{r}
-rV <- function(n) {
-  runif(n, 0.1, 0.17)
-}
+unmaskedTB <- data.frame(n_tb = share_tb$n_tb) %>%
+  mutate(unmasked_tb = lapply(n_tb, function(x) rTBunmasked(n = n_samples, lambda = x))) %>%
+  unnest(unmasked_tb) %>%
+  group_by(unmasked_tb) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  mutate(prop_unmasked_tb = n / sum(n)) 
+mean_unmaskedTB <- sum(unmaskedTB$unmasked_tb * unmaskedTB$prop_unmasked_tb)
+unmaskedTB_pl <- unmaskedTB %>%
+  ggplot(aes(y = prop_unmasked_tb, x = unmasked_tb)) +
+  geom_bar(stat="identity", fill = "grey", color = "black") +
+  geom_vline(mapping = aes(xintercept = mean_unmaskedTB), linetype = "dashed", color = "darkred") +
+  annotate("text", y = 0.25, x = mean_unmaskedTB, hjust = -.2, label = paste0("Mean = ", round(mean_unmaskedTB, 1)), size = 8 / cm(1), color = "darkred") +
+  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
+  scale_x_continuous(expand = c(0,0)) +
+  labs(x = "No. of unmasked TB patients", y = "Probability (%)") +
+  theme_bw2()
+save_plot(unmaskedTB_pl, pdf_file = "../results/inputs/unmasked-tb-patients.pdf", w = aw, h = ah)
 ```
 
-### Loss rate due to inactivation
+### Ventilation
+
+Average volume of exhaled gas:
+* Richardson et al. (2014): 0.1 - 0.17 l/s per person based on Feher (2017)
+--> Use Uniform(0.1, 0.17)
+
+CO2 levels in outdoor air:
+* https://gml.noaa.gov/dv/data/index.php?site=CPT&parameter_name=Carbon%2BDioxide,%20accessed%2012%20November%202019
+--> Set to 4125.5
+
+CO2 levels in exhaled air:
+* Rudnick and Milton (2003): 40,000
+--> Set to 40,000
+
+CO2 generation rate
+* Persily et al.: see Table 5, around 0.003 to 0.005
+--> Use Uniform(0.003, 0.005)
+
+### Inactivation
 
@@
 
-### Loss rate due to deposition onto surfaces
+### Deposition onto surfaces
 
 * Wang et al. (2021): The larger the size of the particle, the less long it will reside in the air. Environmental factors such as temperature, humidity and ventilation also play an important role.
-
-### CO2 levels in outdoor air
-
-* https://gml.noaa.gov/dv/data/index.php?site=CPT&parameter_name=Carbon%2BDioxide,%20accessed%2012%20November%202019
-
-```{r}
-co2_outdoor <- 412.5
-```
-
-### CO2 levels in exhaled air
-
-* Rudnick and Milton (2003): 40,000
-
-```{r}
-co2_exhaled <- 40000
-```
-
-### CO2 generation rate
-
-* Persily et al.: see Table 5, around 0.003 to 0.005
-
-```{r}
-rG <- function(n) {
-  runif(n, 0.0041, 0.0055)
-}
-```
-
-### Unmasked TB patients
-
-* Masked TB patients: TB suspects
-* Unmasked TB patients: Poisson(mu = TB suspects)
-
-```{r}
-rTBunmasked <- function(n, lambda) {
-  rpois(n, lambda)
-}
-```
-
-## Data
-
-### Clinical
-
-```{r}
-# clinical data
-clinXov <- readRDS("../data-clean/Masi/merged-clinical-xovis.rds")
-
-yesTB <- clinXov %>%
-  filter(tb_suspect=="yes") %>%
-  group_by(patient_id) %>%
-  slice(1) %>%
-  ungroup()
-
-nTBmasked <- nrow(yesTB)
-
-noTB <- clinXov %>%
-  dplyr::filter(ifelse(is.na(tb_suspect), T, ifelse(tb_suspect=="no", T, F))) %>%
-  group_by(person_id) %>%
-  slice(1) %>%
-  ungroup()
-
-# Time period (for the first day)
-time_period <- seq(min(as.POSIXct(format(clinXov$time, "%Y-%m-%d %H:%M"))), 
-                   max(as.POSIXct(format(clinXov$time, "%Y-%m-%d %H:%M"))), 
-                   by = "min")
-```
-
-### Co2
-
-```{r}
-# co2
-co2 <- readRDS("../data-clean/Masi/environmental-data/co2-levels.rds") %>%
-  filter(date == "2021-10-25") 
-#' assume waiting area has mean co2 of waiting room and registration
-#' passage has co2 of watiing room only
-#' tb room has co2 of tb room only
-co2 <- co2 %>%
-  filter(location %in% c("waiting room", "registration")) %>%
-  group_by(date_time) %>%
-  summarize(co2 = mean(co2)) %>%
-  ungroup() %>%
-  mutate(location = "waiting room") %>%
-  rbind(co2 %>%
-          filter(location == "waiting room") %>%
-          mutate(location = "passage") %>%
-          dplyr::select(date_time, location, co2)) %>%
-  rbind(co2 %>%
-          filter(location == "TB room") %>%
-          mutate(location = "tb room") %>%
-          dplyr::select(date_time, location, co2)) 
-ggplot(co2, aes(x = date_time, y = co2, color = location)) + geom_line()
-```
-
-### Room
-
-```{r}
-# choose cell size such that the squared cell has 1m3 volume
-cellVolume <- 1
-cellSize <- sqrt(cellVolume * 1e9 / 3e3)
-
-# waiting room
-sh_wr <- vect("../data-raw/Masi/building/waiting_room.shp")
-sf_wr <- st_as_sf(sh_wr)
-wrRoomSP <- shapeToSpatial(sh_wr, cellSize)
-wrVol <- compute_volume(wrRoomSP)
-wrCellCoords <- create_coord_df(wrRoomSP)
-wrRoomXY <- as.matrix(dplyr::select(wrCellCoords, x, y))
-
-# tb room
-sh_tb <- vect("../data-raw/Masi/building/tb_room.shp")
-sf_tb <- st_as_sf(sh_tb)
-tbRoomSP <- shapeToSpatial(sh_tb, cellSize)
-tbVol <- compute_volume(tbRoomSP)
-tbCellCoords <- create_coord_df(tbRoomSP)
-tbRoomXY <- as.matrix(dplyr::select(tbCellCoords, x, y))
-
-# passage
-sh_pa <- vect("../data-raw/Masi/building/passage.shp")
-sf_pa <- st_as_sf(sh_pa)
-paRoomSP <- shapeToSpatial(sh_pa, cellSize)
-paVol <- compute_volume(paRoomSP)
-paCellCoords <- create_coord_df(paRoomSP)
-paRoomXY <- as.matrix(dplyr::select(paCellCoords, x, y))
-
-# combine
-clinicCellCoords <- rbind(
-  mutate(wrCellCoords, location = "waiting room"),
-  mutate(tbCellCoords, location = "tb room"),
-  mutate(paCellCoords, location = "passage")
-) %>% mutate(id = 1:nrow(.))
-clinicCellCoordsXY <- cbind(clinicCellCoords$x, clinicCellCoords$y)
-```
-
-### No. people
-
-```{r}
-# number of people in space
-nDF <- tibble()
-for (tp in time_period) {
-  n_tp <- clinXov %>%
-    filter(between(time, tp, tp + seconds(59))) %>%
-    mutate(location = mapply(function(x,y) clinicCellCoords$location[find_raster(x,y,clinicCellCoordsXY)], x, y)) %>%
-    group_by(person_id, location) %>%
-    slice(1) %>%
-    ungroup() %>%
-    group_by(location) %>%
-    summarize(n = n()) %>%
-    ungroup() 
-  for (loc in unique(clinicCellCoords$location)) {
-    if (!(loc %in% n_tp$location)) {
-      n_tp <- rbind(n_tp, data.frame(location = loc, n = 0))
-    } 
-  }
-  nDF <- rbind(nDF, n_tp %>% mutate(time = tp))
-}
-nDF <- nDF %>% 
-  mutate(time = as.POSIXct(time, origin = as.Date("1970-01-01")))
-nDF$n[nDF$location=="waiting room"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"]
-nDF$n[nDF$location=="passage"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"]
-ggplot(nDF, aes(x = time, y = n, color = location)) + geom_line()
-```
-
-### Movement
-
-```{r}
-patMov <- clinXov %>%
-  mutate(id = map2_int(x, y, find_raster, cellCoordsXY = clinicCellCoordsXY))
-```
 

### 2023-03-09 preprocessing/match_clinical-xovis.R

@@ -1,117 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-
-#### Helper ####
-
-
-#### Masi ####
-
-# Data
-
-clinical_masi <- readRDS("data-clean/Masi/clinical-data/clinical-data.rds")
-xovis_masi <- readRDS("data-clean/Masi/xovis/patients-time-location.rds")
-# env_masi <- readRDS("data-clean/Masi/environmental-data/co2-levels.rds")
-
-
-# Only first entry from clinical data
-
-clinical_masi <- clinical_masi %>%
-  group_by(patient_id, date) %>%
-  slice(1) %>%
-  ungroup()
-
-# summary xovis
-
-# xovis_reg_masi <- xovis_masi %>%
-#   group_by(PersonID) %>%
-#   summarise(Enter_Date = min(Time),
-#             Exit_Date = max(Time),
-#             InE = any(InE),
-#             Height = max(Height),
-#             TimepointEnteringtoRegistry = min(TimepointEnteringtoRegistry, na.rm = T)) %>%
-#   ungroup() %>%
-#   filter(Height >= 1400 & Height <= 2100,
-#          InE,
-#          !is.infinite(TimepointEnteringtoRegistry))
-
-xovis_reg_masi <- xovis_masi %>%
-  group_by(person_id) %>%
-  summarise(min_time = min(time),
-            max_time = max(time),
-            time_registration = min(time_registration, na.rm = T)) %>%
-  ungroup() 
-
-# pre-filter clinical data
-clinical_masi <- clinical_masi %>%
-  filter(date_time >= min(xovis_reg_masi$time_registration),
-         date_time <= max(xovis_reg_masi$time_registration))
-
-# matching clinical and xovis data
-#' match the date of first event in the clinical data 
-#' to the person that last came to the Entrance 
-#' with a tolerance of 15min
-
-merge_masi <- merge(xovis_reg_masi %>% select(person_id, time_registration),
-                    clinical_masi %>% select(patient_id, date_time, tb_suspect),
-                    id = NULL) %>%
-  mutate(timediff = as.numeric(difftime(date_time, time_registration), units = "secs")) %>%
-  filter(timediff <= 15 * 60,
[…]

### 2023-03-09 preprocessing/prep_environmental-data.R

@@ -6,7 +6,8 @@ library(haven)
 
 #### Masi ####
 
-df <- read_dta("data-raw/Masi/environmental-data/co2_data.dta") %>%
+# load data
+masi <- read_dta("data-raw/Masi/environmental-data/co2_data.dta") %>%
   mutate(date = Date,
          time = format(new_time, "%H:%M:%S"),
          date_time = as.POSIXct(paste(Date, time)),
@@ -16,6 +17,30 @@ df <- read_dta("data-raw/Masi/environmental-data/co2_data.dta") %>%
          co2 = CO2) %>%
   select(date, time, date_time, location, co2, temperature, humidity) 
 
+# waiting room is the average of waiting room and registration
+wr <- masi %>%
+  filter(location %in% c("registration", "waiting room")) %>%
+  group_by(date, time, date_time) %>%
+  summarize(across(c(co2, temperature, humidity), mean)) %>%
+  ungroup() %>%
+  mutate(location = "waiting room")
 
-saveRDS(df, "data-clean/Masi/environmental-data/co2-levels.rds")
+# assume: co2 level in passage is the same as in the waiting room
+#' TODO: check if this assumptions is valid; 
+#' maybe not because the door to the passage from the waiting room may be closed
+
+pa <- wr %>%
+  filter(location == "waiting room") %>%
+  mutate(location = "passage") 
+
+# tb room
+tb <- masi %>%
+  filter(location == "TB room") %>%
+  mutate(location = "tb room")
+
+# combine 
+masi_prep <- rbind(wr, pa, tb)
+
+
+saveRDS(masi_prep, "data-clean/Masi/environmental-data/co2-temp-humid.rds")
   

### 2023-03-09 preprocessing/prep_room.R

@@ -9,7 +9,7 @@ masi <- do.call(rbind, lapply(files, function(x) read.csv(x, header = T) %>% set
 
 pl <- ggplot() + 
   geom_bin2d(data = masi, mapping = aes(x = x, y = y), binwidth = 100) +
-  scale_fill_viridis_c()
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = c(1,10,100,1000,10000), limits = c(0,10000)) 
 
 pl
 
@@ -41,31 +41,39 @@ line_point <- function(pt, l, r, start = T) {
 }
 
 # outline
-width <- 24000
-breadth <- 8500
-rotation <- 9.5
-bottom_left_pt <- c(-6700, -9000)
+width <- 22750
+breadth <- 8250
+rotation <- 9
+bottom_left_pt <- c(-6200, -8600)
 bottom_right_pt <- line_point(bottom_left_pt, width, rotation)
 outline <- parallelogram(bottom_left_pt, bottom_right_pt, width, breadth, rotation)
 
 # waiting room
-wr_width <- 10700
+wr_width <- 10000
 wr_breadth <- 5700
 wr_top_right_pt <- outline[3,]
 wr_bottom_right_pt <- line_point(wr_top_right_pt, wr_width, rotation, start = F)
 wr <- parallelogram(wr_bottom_right_pt, wr_top_right_pt, wr_width, -wr_breadth, rotation)
 
+# sputum room
+sp_width <- 5000
+sp_breadth <- 1250
+sp_bottom_left_pt <- outline[1,]
+sp_bottom_right_pt <- line_point(sp_bottom_left_pt, tb_width, rotation)
+sp <- parallelogram(sp_bottom_left_pt, sp_bottom_right_pt, sp_width, sp_breadth, rotation)
+
 # TB room
-tb_width <- 5550
-tb_breadth <- 6500
-tb_bottom_left_pt <- outline[1,]
+tb_width <- sp_width
+tb_breadth <- 5000
+tb_bottom_left_pt <- sp[4,]
 tb_bottom_right_pt <- line_point(tb_bottom_left_pt, tb_width, rotation)
 tb <- parallelogram(tb_bottom_left_pt, tb_bottom_right_pt, tb_width, tb_breadth, rotation)
 
 # passage
 pa_width <- width - wr_width
[…]

### 2023-03-09 preprocessing/prep_xovis_matching.R

@@ -51,7 +51,6 @@ match_xovis <- function(date = "2021-10-25",
   n_oid <- length(unique(masi$obs_id))
   
   # remove very short tracks
-  
   masi <- masi %>%
     group_by(obs_id) %>%
     filter(! ( as.numeric(difftime(last(time), first(time), units = "secs")) < 10  &

### 2023-03-09 utils/spatial.r

@@ -1,3 +1,7 @@
+require(sf)
+require(terra)
+require(raster)
+
 #' Compute euclidean distance in two dimensions
 #' 
 #' @param x1 x coordinate of first point
@@ -44,7 +48,7 @@ shapeToSpatial <- function(sh, cell_size) {
 #' 
 
 create_coord_df <- function(spat_poly_df) {
-  cellCoords <- data.frame(coordinates(spat_poly_df)) 
+  cellCoords <- data.frame(raster::coordinates(spat_poly_df)) 
   colnames(cellCoords) <- c("x", "y")
   cellCoords$id <- 1:nrow(cellCoords)
   return(cellCoords)
@@ -57,7 +61,7 @@ create_coord_df <- function(spat_poly_df) {
 #' @param H height of the room
 
 compute_volume <- function(spat_poly_df, H = 3) {
-  A <- convert_dist(convert_dist(sum(area(spat_poly_df))))
+  A <- convert_dist(convert_dist(sum(raster::area(spat_poly_df))))
   V <- A * H
   return(V)
 }

### 2023-03-09 utils/trans_risk.r

@@ -26,7 +26,7 @@ compute_AER <- function(G, Co, C, n, V) {
   # excess CO2 in ppm
   Ce <- C - Co
   # TODO: is there are more elegant way to ensure that the denominator is not 0?
-  Ce <- ifelse(Ce == 0, 1, Ce) 
+  Ce <- ifelse(Ce < 0, 1, Ce) 
   # excess CO2 in l/m3 (1l/m3 = 1,000ppm)
   Ce <- Ce / 1000
   # ventilation rate (in l/s per person)
@@ -67,3 +67,68 @@ compute_Nt <- function(E, RR, N0, V) {
 compute_P <- function(N, t = 1) {
   1 - exp(-N * t)
 } 
+
+
+#' Relative viral concentration by distance to infector
+#' 
+#' @param distance distance in m
+#' @param mu mean of the exponential decaying distribution
+#' 
+
+dIc <- function(distance, mu) { 
+  1 - pexp(distance, 1 / mu)
+}
+
+#' Mean parameter of exponentially decaying distribution for the relative viral concentration by distance to infector
+#' 
+#' @param n number of samples
+
+rMuIc <- function(n) {
+  rgamma(n, shape = 10, scale = .2)
+}
+
+
+#' Quanta emission rate
+#' 
+#' @param n number of samples
+#' @param x quanta rate
+
+rq <- function(n, disease = "TB") {
+  LaplacesDemon::rtrunc(n, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
+}
+
+dq <- function(x) {
+  LaplacesDemon::dtrunc(x, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
+}
+
+
+#' Average volume of exhaled gas
+#' 
+#' @param n number of samples
+#' 
+
+
+rV <- function(n) {
+  runif(n, 0.1, 0.17)
+}
+
+
+#' CO2 generation rate
+#' 
+#' @param n number of samples
+#' 
+
+rG <- function(n) {
+  runif(n, 0.003, 0.005)
+}
+
+
+#' Unmasked TB Patients
+#' 
+#' @param n number of samples
+#' @param mu expected number of unmasked TB patients
+#' 
+
+rTBunmasked <- function(n, lambda) {
+  rpois(n, lambda)
+}

### 2023-03-09 analysis/spatiotemporal_model.R
spatiotemporal model in parallel workflow

@@ -1,0 +1,292 @@
+# Libraries
+
+library(tidyverse)
+options(dplyr.summarise.inform = FALSE)
+library(parallel)
+library(foreach)
+
+source("utils/trans_risk.R")
+source("utils/spatial.R")
+
+
+#' Prepare data for spatiotemporal model
+#' 
+#' @param clinic name of clinic
+#' @param sel_date selected date of patient movement data
+#' @param seed seed number
+#' @param nSim number of simulations
+#' @param cellVolume desired volume of cells to compute quanta concentration (in m3)
+#' @param co2_exhaled exhaled CO2 per person (default: 40,000 ppm)
+#' @param co2_outdoor CO2 level in outdoor air (default: 400 ppm)
+#' @param ... additional arguments include the distributions below
+#' @param rTBunmasked number of unmasked TB patients distribution
+#' @param rMuIc mean of spatial emission weight distribution
+#' @param dIc  spatial emission weight distribution
+#' @param rq quanta emission rate distribution
+#' @param rV average volume of exhaled gas distribution
+#' @param rG CO2 generation rate distribution
+
+
+prep_data <- function(
+    clinic, 
+    sel_date,
+    seed = 12345,
+    nSim = 1000,
+    min_duration = 5,
+    cellVolume = 1,
+    co2_exhaled = 40000,
+    co2_outdoor = 400,
+    ...) {
+  
+  #### Data ####
+  
+  # Settings
+  set.seed(seed)
+  
+  # patient movement data
+  patMov <- readRDS(paste("data-clean",clinic,"patient-tracking-data",sel_date,"patient-id-data.rds",sep="/")) %>%
+    group_by(patient_id) %>%
+    filter(as.numeric(difftime(last(time), first(time), units = "mins")) > min_duration) %>%
+    ungroup() %>%
+    mutate(location = ifelse(is_waitingroom|is_passage, "waiting room", ifelse(is_tbroom, "tb room", NA)),
+           hhmm = format(time, format = "%H:%M")) %>%
+    filter(!is.na(location)) 
+  patIDs <- unique(patMov$patient_id)
+  
+  # CO2 data
+  co2 <- readRDS(paste("data-clean", clinic, "environmental-data", "co2-temp-humid.rds", sep = "/")) %>%
+    filter(date == sel_date)
[…]

### 2023-03-14 analysis/spatiotemporal-transmission-risk.Rmd
update spatiotemporal modeling workflow

@@
 # plot with and height of input and data graphs
-aw <- 12
-ah <- 8
+width_descr <- 10.5 / cm(1)
+height_descr <- 7.5 / cm(1)
+text_descr <- 12
 ```
@@
 
+```{r}
+selected_dates <- as.Date(list.files("../data-clean/Masi/patient-tracking-data")[-1])
+```
+
 ### Clinical
@@
 ```{r}
-clinical_dta <- readRDS("../data-clean/Masi/clinical-data/clinical-data.rds")
+clinical_dta <- readRDS("../data-clean/Masi/clinical-data/clinical-data.rds") %>%
+  filter(date %in% selected_dates)
+
+# number of registered patients
+clinical_dta %>%
+  group_by(patient_id) %>%
+  slice(1) %>%
+  nrow()
+
+# share of TB patients
 share_tb <- clinical_dta  %>%
@@
   geom_bar(stat = "identity", fill = "grey", color = "black", width = .5) +
-  annotate("text", x = "Nov 09", y = mean(share_tb$n_tb), label = paste0("Mean = ",round(mean(share_tb$n_tb), 1)), color = "darkred", size = 8 / cm(1), vjust = -1) +
+  annotate("text", x = "Nov 10", y = mean(share_tb$n_tb), label = paste0("Mean = ",round(mean(share_tb$n_tb), 1)), color = "darkred", size = 8 / cm(1), vjust = -1) +
   geom_hline(aes(yintercept = mean(share_tb$n_tb)), color = "darkred") +
   scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
-  scale_x_discrete(breaks = c("Oct 11", "Oct 18", "Oct 25", "Nov 02", "Nov 09", "Nov 16")) +
-  labs(y = "No. of TB patients") +
-  theme_bw2() +
+  scale_x_discrete(breaks = format(selected_dates, format = "%b %d")[c(1, 4, 8, 12)]) +
+  labs(y = "No. of masked TB patients") +
+  theme_classic() +
   theme(axis.title.x = element_blank(),
-        plot.margin = margin(5.5, 10, 5.5, 5.5))
-save_plot(share_tb_pl, pdf_file = "../results/data/share-tb-patients.pdf", w = aw, h = ah)
+        plot.margin = margin(5.5, 5.5, 5.5, 5.5),
+        text = element_text(size = text_descr))
+ggsave(plot = share_tb_pl, filename = "../doc/IWHOD/presentation/figures/share-tb-patients.png", w = width_descr, h = height_descr)
 ```
@@
 co2_pl <- readRDS("../data-clean/Masi/environmental-data/co2-temp-humid.rds") %>%
-  filter(location %in% c("waiting room", "tb room")) %>%
+  filter(location %in% c("waiting room", "tb room"),
+         date %in% selected_dates) %>%
   mutate(location = factor(ifelse(location == "tb room", "TB room", "Waiting room"), levels = c("Waiting room", "TB room")),
@@
   ggplot(aes(x = time, y = co2, color = location, fill = location)) +
-  stat_lineribbon(alpha = .25, size = .5) +
+  stat_lineribbon(alpha = .25, size = .75) +
   scale_x_datetime(expand = c(0,0), date_breaks = "2 hours", date_labels = "%H:%M") +
   scale_y_continuous(expand = c(0,0), breaks = scales::breaks_width(width = 100)) +
-  scale_fill_brewer(palette = "Accent") +
-  scale_color_brewer(palette = "Accent") +
+  scale_fill_manual(values = wes_palette("Moonrise2")) +
+  scale_color_manual(values = wes_palette("Moonrise2")) +
   labs(y = expression("CO"[2]*" (ppm)"), fill = "Quantile", color = "Quantile") +
-  theme_bw2() +
-  theme(legend.position = c(.8,.95), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
-        plot.margin = margin(5.5, 10, 5.5, 5.5), axis.title.x = element_blank(), legend.title = element_blank()) 
-save_plot(co2_pl, pdf_file = "../results/data/co2-levels.pdf", w = aw, h = ah)
+  theme_classic() +
+  theme(legend.position = c(.65,.95), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        plot.margin = margin(5.5, 5.5, 5.5, 5.5), axis.title.x = element_blank(), legend.title = element_blank(),
+        text = element_text(size = text_descr)) 
+ggsave(plot = co2_pl, filename = "../doc/IWHOD/presentation/figures/co2-levels.png", width = width_descr, height = height_descr)
 ```
@@
 clinic_img <- png::readPNG("../data-raw/Masi/building/clinic_clipped.png")
-```
 
-<!-- ```{r} -->
-<!-- # choose cell size such that the squared cell has 1m3 volume -->
-<!-- cellVolume <- 1 -->
-<!-- cellSize <- sqrt(cellVolume * 1e9 / 3e3)  -->
+# building
+cellSize <- sqrt(1 * 1e9 / 3e3) 
+building <- terra::vect(paste("../data-raw", "Masi", "building", "clinic-vector.gpkg", sep = "/"))
+building_sf <- sf::st_as_sf(building, crs = NA)
+sf::st_crs(building_sf) <- NA
+waiting_room <- shapeToSpatial(building_sf$geometry[2], cellSize)
+waiting_room_df <- fortify(waiting_room)
+passage <- shapeToSpatial(building_sf$geometry[4], cellSize)
+passage_df <- fortify(passage)
+tb_room <- shapeToSpatial(building_sf$geometry[3], cellSize)
+tb_room_df <- fortify(tb_room)
 
-<!-- # waiting room -->
-<!-- sh_wr <- vect("../data-raw/Masi/building/waiting_room.shp") -->
-<!-- sf_wr <- st_as_sf(sh_wr) -->
-<!-- wrRoomSP <- shapeToSpatial(sh_wr, cellSize) -->
-<!-- wrVol <- compute_volume(wrRoomSP) -->
-<!-- wrCellCoords <- create_coord_df(wrRoomSP) -->
-<!-- wrRoomXY <- as.matrix(dplyr::select(wrCellCoords, x, y)) -->
+# test
+waiting_room_df$value <- rnorm(nrow(waiting_room_df))
+passage_df$value <- rnorm(nrow(passage_df))
+tb_room_df$value <- rnorm(nrow(tb_room_df))
 
-<!-- # tb room -->
-<!-- sh_tb <- vect("../data-raw/Masi/building/tb_room.shp") -->
-<!-- sf_tb <- st_as_sf(sh_tb) -->
-<!-- tbRoomSP <- shapeToSpatial(sh_tb, cellSize) -->
-<!-- tbVol <- compute_volume(tbRoomSP) -->
-<!-- tbCellCoords <- create_coord_df(tbRoomSP) -->
-<!-- tbRoomXY <- as.matrix(dplyr::select(tbCellCoords, x, y)) -->
-
-<!-- # passage -->
-<!-- sh_pa <- vect("../data-raw/Masi/building/passage.shp") -->
-<!-- sf_pa <- st_as_sf(sh_pa) -->
-<!-- paRoomSP <- shapeToSpatial(sh_pa, cellSize) -->
-<!-- paVol <- compute_volume(paRoomSP) -->
-<!-- paCellCoords <- create_coord_df(paRoomSP) -->
-<!-- paRoomXY <- as.matrix(dplyr::select(paCellCoords, x, y)) -->
-
-<!-- # combine -->
-<!-- clinicCellCoords <- rbind( -->
-<!--   mutate(wrCellCoords, location = "waiting room"), -->
-<!--   mutate(tbCellCoords, location = "tb room"), -->
-<!--   mutate(paCellCoords, location = "passage") -->
-<!-- ) %>% mutate(id = 1:nrow(.)) -->
-<!-- clinicCellCoordsXY <- cbind(clinicCellCoords$x, clinicCellCoords$y) -->
-<!-- ``` -->
-
-### No. people
-
-```{r}
-files <- list.files("../data-clean/Masi/patient-tracking-data/", full.names = T)
-no_people <- do.call(rbind, lapply(files, function(x) readRDS(paste0(x, "/patient-id-data.rds")))) %>%
-  mutate(date = format(time, format = "%Y-%m-%d"),
-         time = as.POSIXct(time, format = "%H:%M", tz = "CET"),
-         daytime = ifelse(time <= "2021-10-25 12:00:00", "Morning", "Afternoon"),
-         daytime = factor(daytime, levels = c("Morning", "Afternoon")))
-
-# over time
-no_people_time_pl <- no_people %>%
-  group_by(date, time) %>%
-  summarize(n = n()) %>%
-  ungroup() %>%
-  ggplot(aes(x = time, y = n)) +
-  geom_line() +
-  #stat_lineribbon(alpha = .25, size = .5) +
-  scale_x_datetime(expand = c(0,0), date_breaks = "2 hours", date_labels = "%H:%M") +
-  scale_y_continuous(expand = c(0,0), limits = c(0, NA), breaks = scales::breaks_width(5)) +
-  scale_fill_brewer(palette = "OrRd") +
-  labs(y = "No. of clinical attendees") +
-  theme_bw2() +
-  theme(axis.title.x = element_blank())
-  #labs(y = "No. of clinical attendees", fill = "Quantile", color = "Quantile") +
-  # theme(legend.position = c(.8,.9), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
-  #       plot.margin = margin(5.5, 10, 5.5, 5.5)) +
-  # guides(fill = guide_legend(title.position = "top"))
-save_plot(no_people_time_pl, pdf_file = "../results/data/no-people-over-time.pdf", w = aw, h = ah)
-
-# Spatial
-no_people_spatial_pl <- no_people %>%
-  dplyr::select(x, y) %>%
-  adespatial::rotation(angle = -8*pi/180) %>%
-  as.data.frame() %>%
-  set_names(c("x", "y")) %>%
-  mutate(x = -x) %>%
-  ggplot(aes(x = x, y = y)) +
-  geom_bin2d(binwidth = 100) +
+ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
+  geom_polygon(data = waiting_room_df) +
+  geom_polygon(data = passage_df) +
+  geom_polygon(data = tb_room_df) +
   ggpubr::background_image(clinic_img) +
@@
   annotate("text", x = 2600, y = -2700, label = "TB room", size = 8 / cm(1), fontface = 2) +
-  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = c(1,5,10,50,100), limits = c(0,100)) +
-  scale_y_continuous(expand = expansion(mult = c(0,.18))) +
-  scale_x_continuous(expand = expansion(mult = c(0,0.025))) +
+  scale_y_continuous(expand = expansion(add=c(2500, 4000))) +
+  scale_x_continuous(expand = expansion(add=c(400, 500))) +
   labs(fill = "Density") +
@@
         panel.grid.minor = element_blank())
-save_plot(no_people_spatial_pl, pdf_file = "../results/data/no-people-spatial.pdf", w = aw, h = ah)
 ```
 
-<!-- ```{r} -->
-<!-- # number of people in space -->
-<!-- nDF <- tibble() -->
-<!-- for (tp in time_period) { -->
-<!--   n_tp <- clinXov %>% -->
-<!--     filter(between(time, tp, tp + seconds(59))) %>% -->
-<!--     mutate(location = mapply(function(x,y) clinicCellCoords$location[find_raster(x,y,clinicCellCoordsXY)], x, y)) %>% -->
-<!--     group_by(person_id, location) %>% -->
-<!--     slice(1) %>% -->
-<!--     ungroup() %>% -->
-<!--     group_by(location) %>% -->
-<!--     summarize(n = n()) %>% -->
-<!--     ungroup()  -->
-<!--   for (loc in unique(clinicCellCoords$location)) { -->
-<!--     if (!(loc %in% n_tp$location)) { -->
-<!--       n_tp <- rbind(n_tp, data.frame(location = loc, n = 0)) -->
-<!--     }  -->
-<!--   } -->
-<!--   nDF <- rbind(nDF, n_tp %>% mutate(time = tp)) -->
-<!-- } -->
-<!-- nDF <- nDF %>%  -->
-<!--   mutate(time = as.POSIXct(time, origin = as.Date("1970-01-01"))) -->
-<!-- nDF$n[nDF$location=="waiting room"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"] -->
-<!-- nDF$n[nDF$location=="passage"] <- nDF$n[nDF$location=="waiting room"] + nDF$n[nDF$location=="passage"] -->
-<!-- ggplot(nDF, aes(x = time, y = n, color = location)) + geom_line() -->
-<!-- ``` -->
+### No. people
 
-<!-- ### Movement -->
+```{r}
+files <- list.files("../data-clean/Masi/patient-tracking-data/", full.names = T)[-1]
+no_people <- do.call(rbind, lapply(files, function(x) readRDS(paste0(x, "/patient-id-data_matched-filtered.rds")))) %>%
+  mutate(date = as.Date(format(time, format = "%Y-%m-%d")),
+         time_min = as.POSIXct(format(time, format = "%H:%M"), format = "%H:%M"))
 
-<!-- ```{r} -->
-<!-- patMov <- clinXov %>% -->
-<!--   mutate(id = map2_int(x, y, find_raster, cellCoordsXY = clinicCellCoordsXY)) -->
-<!-- ``` -->
+# number of clinical attendees
+no_people %>%
+  group_by(date, patient_id) %>%
+  slice(1) %>%
+  nrow()
+
+# over time
+no_people_time <- no_people %>%
+  group_by(date, time_min, time) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  group_by(date, time_min) %>%
+  summarize(n = mean(n)) %>%
+  ungroup()
+
+no_people_time_pl <- no_people_time %>%
+  ggplot(aes(x = time_min, y = n)) +
+  stat_lineribbon(alpha = .25, size = .5) +
+  scale_x_datetime(expand = c(0,0), date_breaks = "2 hours", date_labels = "%H:%M") +
+  scale_y_continuous(expand = c(0,0), limits = c(0, NA), breaks = scales::breaks_width(5)) +
+  scale_fill_brewer(palette = "OrRd") +
+  labs(y = "No. of clinical attendees", fill = "Quantile", color = "Quantile") +
+  theme_bw2() +
+  theme(axis.title.x = element_blank(), legend.position = c(.8,.9), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        plot.margin = margin(5.5, 10, 5.5, 5.5)) +
+  guides(fill = guide_legend(title.position = "top"))
+no_people_time_pl
+
+# Spatial
+no_people_spatial_pl <- no_people %>%
+  ggplot(aes(x = x, y = y)) +
+  geom_bin2d(binwidth = 100) +
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log", breaks = c(1, 10, 100, 1000, 10000)) +
+    ggpubr::background_image(clinic_img) +
+    annotate("text", x = -13200, y = -5800, label = "Entrance", size = text_descr / cm(1), hjust=0,vjust=0) +
+    annotate("text", x = -14000, y = -2600, label = "Waiting room", size = text_descr / cm(1), hjust=0,vjust=0) +
+    annotate("text", x = -5800, y = -4500, label = "Registration", size = text_descr / cm(1), angle = 90, hjust=0,vjust=0) +
+    annotate("text", x = 3000, y = -2700, label = "TB room", size = text_descr / cm(1)) +
+    scale_y_continuous(expand = expansion(mult=c(0.05, 0.15))) +
+    scale_x_continuous(expand = expansion(mult=c(0.03, 0))) +
+    labs(fill = "Density") +
+    theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.3, "cm"),
+          legend.position = c(0.825,0.075), legend.direction = "horizontal",legend.title = element_text(vjust = .75),
+          panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank()) +
+  guides(fill = guide_legend(title.position = "top", label.position = "bottom"))
+ggsave(plot=no_people_spatial_pl, filename = "../doc/IWHOD/presentation/figures/no-people-spatial.png", width = width_descr, height = height_descr)
+```
 
@@
 
-## Simulation
+## Evaluation
 
 ```{r}
-# Settings
-set.seed(12345)
-nSim <- 100
+sim_results <- list.files("../simulations/Masi/2021-10-25/", full.names = T)
+sim_results <- do.call(rbind, lapply(sim_results, readRDS))
+```
 
-# Input parameters
-inputPar <- data.frame(
-  spatial_mu = rMuIc(nSim),
-  quanta = rq(nSim),
-  V = rV(nSim),
-  co2_ex = co2_exhaled,
-  co2_out = min(co2$co2),
-  co2_gen = rG(nSim),
-  unmasked_TB = rTBunmasked(nSim, nTBmasked)
-)
+### Transmission risk
 
-# Loop
-for (s in 1:nSim) {
-  message(sprintf("Simulation %i of %i", s, nSim))
-
-  # parameters
-  smu <- inputPar$spatial_mu[s]
-  q <- inputPar$quanta[s]
-  V <- inputPar$V[s]
-  Ca <- inputPar$co2_ex[s]
-  Co <- inputPar$co2_out[s]
-  G <- inputPar$co2_gen[s]
-  unmasked_TB <- inputPar$unmasked_TB[s]
-
-  # transformed parameters
-  # quanta emission rate per minute
-  ER <- q / 60
-
-  # TB patients
-  masked_tb <- yesTB$person_id
-  unmasked_tb <- sample(noTB$person_id, size = unmasked_TB)
-  TBpatients <- patMov %>%
-    filter(person_id %in% c(masked_tb, unmasked_tb)) %>%
-    dplyr::select(person_id, observation_id, time, x, y) 
-  TBpatients_full <- TBpatients %>%
-    group_by(person_id) %>%
-    summarize(time = list(seq(min(time), max(time), by = "sec"))) %>%
-    ungroup() %>%
-    unnest(cols = c(time))
-  TBpatients_full <- TBpatients_full %>%
-    left_join(TBpatients, by = c("person_id", "time")) %>%
-    tidyr::fill(observation_id, x, y, .direction = "down")
-
-  # no TB patients
-  noTBpatients <- patMov %>%
-    filter(!(person_id %in% c(masked_tb, unmasked_tb))) %>%
-    dplyr::select(person_id, time, id) 
-  noTBpatients_full <- noTBpatients %>%
-    group_by(person_id) %>%
-    summarize(time = list(seq(min(time), max(time), by = "sec"))) %>%
-    ungroup() %>%
-    unnest(cols = c(time))
-  noTBpatients_full <- noTBpatients_full %>%
-    left_join(noTBpatients, by = c("person_id", "time")) %>%
-    tidyr::fill(id, .direction = "down") %>%
-    mutate(time = as.POSIXct(format(time, "%Y-%m-%d %H:%M"))) %>%
-    group_by(person_id, id, time) %>%
-    summarize(minutes = n() / 60) %>%
+```{r}
+# individual patient transmission risk by simulation
+transRisk <- sim_results %>%
+  group_by(patient_id, sim) %>%
+  slice(1) %>%
   ungroup()
 
-  # df to store quanta concentration by minute
-  quantaDF <- clinicCellCoords %>%
-    mutate(time = time_period[1] - minutes(1),
-           I_SP = 0,
-           I_WM = 0,
-           E_SP = 0,
-           E_WM = 0,
-           N_SP = 0,
-           AER = NA,
-           RR = NA,
-           N_WM = 0,
-           N_SP = 0)
-
-  for (t in 1:length(time_period)) {
-
-    # time point
-    tp <- time_period[t] 
-    quantaDF_new <- clinicCellCoords %>%
-      mutate(time = tp,
-             I_SP = 0,
-             I_WM = 0,
-             E_SP = 0,
-             E_WM = 0,
-             N_SP = 0,
-             N_WM = 0)
-
-    # spatial spread of infectious individuals and quanta emission
-    TBpatients_tp <- TBpatients_full %>%
-      dplyr::filter(between(time, tp, tp + minutes(1)))
-    if (nrow(TBpatients_tp) > 0) {
-      TBpatients_tp <- TBpatients_tp %>%
-        nest(data = -person_id)
-      # iterate over patients
-      for (p in 1:nrow(TBpatients_tp)) {
-        # duration and location
-        duration <- nrow(TBpatients_tp$data[[p]])
-        xy <- TBpatients_tp$data[[p]] %>% dplyr::select(x, y) %>% slice(1) %>% unlist()
-        # generated quanta by cell
-        locationID <- find_raster(xy[1], xy[2], clinicCellCoordsXY)
-        locationName <- clinicCellCoords$location[locationID]
-        locationCellCoords <- filter(clinicCellCoords, location == locationName)
-        locationShape <- if(locationName == "waiting room") {sf_wr} else if(locationName == "tb room") {sf_tb} else {sf_pa}
-        cellInReach <- numeric(nrow(quantaDF_new))
-        cellDistance <- numeric(nrow(quantaDF_new))
-        isIinCell <- ifelse(quantaDF_new$location==locationName, 1, 0 )
-        for (cell in 1:nrow(quantaDF_new)) {
-          # if cell is in room/location
-          if (quantaDF_new$location[cell] != locationName) {
-            cellInReach[cell] <- 0
-            cellDistance[cell] <- -999
-          } else {
-            # can aerosols travel to this cell?
-            straight_line <- st_linestring(rbind(xy, c(quantaDF_new$x[cell], quantaDF_new$y[cell]))) 
-            intersects <- st_difference(straight_line, locationShape)
-            cellInReach[cell] <- ifelse(length(intersects) == 0, 1, 0)
-            cellDistance[cell] <- convert_dist(st_length(straight_line))
-          }
-        }
-        # proportion of infectious individuals per cell per minute
-        distance_dens <- cellInReach * dIc(cellDistance, smu)
-        quantaDF_new$I_SP <- quantaDF_new$I_SP + distance_dens / sum(distance_dens) * (duration / 60)
-        quantaDF_new$I_WM <- quantaDF_new$I_WM + isIinCell / sum(isIinCell)
-      }
-      # quanta emission
-      quantaDF_new$E_SP <- quantaDF_new$E_SP + ER * quantaDF_new$I_SP
-      quantaDF_new$E_WM <- quantaDF_new$E_WM + ER * quantaDF_new$I_WM
-    }
-
-    # compute air exchange rate and quanta removal
-    Ct <- co2 %>%
-      group_by(location) %>%
-      # if CO2 is missing, take the last observed value
-      filter(date_time <= tp) %>%
-      arrange(desc(date_time)) %>%
-      slice(1) %>%
-      mutate(date_time = tp) %>%
-      ungroup() %>%
-      # add number of people
-      left_join(filter(nDF, time == tp), by = c("date_time" = "time", "location")) %>%
-      # add volume
-      mutate(vol = ifelse(location == "tb room", tbVol, wrVol + paVol)) %>%
-      # compute AER and RR per minute
-      mutate(AER = compute_AER(G, Co, co2, n, vol),
-             RR = compute_RR(AER) / 60) 
-    quantaDF_new <- quantaDF_new %>% 
-      left_join(Ct %>% dplyr::select(location, AER, RR), by = "location")
-
-    # previous quanta concentration
-    N0_SP <- quantaDF %>% 
-      filter(time == tp - minutes(1)) %>%
-      dplyr::select(N_SP) %>% unlist
-    N0_WM <- quantaDF %>% 
-      filter(time == tp - minutes(1)) %>%
-      dplyr::select(N_WM) %>% unlist
-
-    # compute quanta concentration
-    quantaDF_new$N_SP <- pmap_dbl(list(E = quantaDF_new$E_SP, 
-                                       RR = quantaDF_new$RR, 
-                                       N0 = N0_SP,
-                                       V = cellVolume), compute_Nt)
-    quantaDF_new$N_WM <- pmap_dbl(list(E = quantaDF_new$E_WM, 
-                                       RR = quantaDF_new$RR, 
-                                       N0 = N0_WM,
-                                       V = cellVolume), compute_Nt)
-
-    # append new data
-    quantaDF <- rbind(quantaDF, quantaDF_new)
-  }
-
-  # compute (spatio)temporal transmission risk for each individual
-  risk_of_infection <- noTBpatients_full %>%
-    left_join(quantaDF, by = c("id", "time")) %>%
-    group_by(person_id) %>%
-    summarize(exposure_time = sum(minutes),
-              P_SP = 1-exp(-sum(N_SP * minutes)),
-              P_WM = 1-exp(-sum(N_WM * minutes))) %>%
+# Average transmission risk across patients
+transRisk_av <- transRisk %>%
+  group_by(sim) %>%
+  summarize(across(c(P_N, P_N_ns), mean)) %>%
   ungroup()
 
-  # compute average transmission risk
-  sum_I <- data.frame(location = c("waiting room", "passage", "tb room"),
-                      I = length(c(masked_tb, unmasked_tb))) 
-  mean_n <- nDF %>% 
-    group_by(location) %>%
-    summarize(n = median(n) + 1) %>%
+mean_risk <- round(mean(transRisk_av$P_N * 100), 1)
+q25_risk <- round(quantile(transRisk_av$P_N * 100, .25), 1)
+q75_risk <- round(quantile(transRisk_av$P_N * 100, .75), 1)
+message(sprintf("Spatiotemporal transmission risk across clinical attendee: %s percent (IQR: %s to %s)", mean_risk, q25_risk, q75_risk))
+
+# individual transmission risk across simulations
+transRisk_indv <- transRisk %>%
+  group_by(patient_id) %>%
+  summarize(across(c(P_N, P_N_ns), list(mean = mean, median = median,q25 = function(x) quantile(x, .25), q75 = function(x) quantile(x, .75)))) %>%
   ungroup()
-  mean_f <- co2 %>%
-    group_by(location) %>%
-    summarize(f = (mean(co2) - Co) / Ca) %>%
+
+transRisk_indv_pl <- transRisk_indv %>%
+  mutate(patient_id = factor(patient_id)) %>%
+  ggplot(aes(x = reorder(patient_id, -P_N_mean), y = P_N_mean)) +
+  geom_point(shape = 4, alpha = .5) +
+  geom_hline(aes(yintercept = mean_risk / 100), color = "darkred") +
+  geom_hline(aes(yintercept = q25_risk / 100), color = "darkred", linetype = "dashed") +
+  geom_hline(aes(yintercept = q75_risk / 100), color = "darkred", linetype = "dashed") +
+  annotate("text", x = 375, y = 0.0075, label = paste0("Average across patients = ", mean_risk, "% (IQR:", q25_risk, "-", q75_risk,")"),
+           size = 8 / cm(1), color = "darkred") +
+  scale_y_continuous(labels = function(x) x * 100, expand = expansion(mult = c(0., .05)), breaks = c(seq(0, 0.05, 0.01), 0.004)) +
+  labs(y = "Mean risk of infection (%)", x = "Clinic attendee") +
+  theme_bw2() +
+  theme(axis.text.x = element_blank())
+save_plot(transRisk_indv_pl, pdf_file = "../results/modeling/risk-of-infection.pdf", w = aw, h = ah)
+
+# high risk of infection plot
+group_labs <- c("Small (1-5%)", "Medium (5-10%)", "High (>10%)")
+transRisk_qq <- transRisk %>%
+  mutate(risk_cat = base::cut(P_N * 100, breaks = c(-Inf, 1, 5, 10, Inf), labels = c("negligible risk", group_labs))) %>%
+  group_by(sim, risk_cat) %>%
+  summarize(n = n()) %>%
   ungroup()
-  risk_of_infection_av <- noTBpatients_full %>%
-    left_join(quantaDF %>% dplyr::select(id, time, location), by = c("id", "time")) %>%
-    group_by(person_id, location) %>%
-    summarize(minutes = sum(minutes)) %>%
+transRisk_MinorVsHigh <- transRisk_qq %>%
+  mutate(risk_cat = ifelse(risk_cat == "negligible risk", "low", "high")) %>%
+  group_by(sim, risk_cat) %>%
+  summarize(n = sum(n)) %>%
   ungroup() %>%
-    left_join(sum_I, by = "location") %>%
-    left_join(mean_n, by = "location") %>%
-    left_join(mean_f, by = "location") %>%
-    mutate(mu = f * I * ER * minutes / n) %>%
-    group_by(person_id) %>%
-    summarize(P_SS = 1 - exp(-sum(mu))) %>%
-    ungroup()
-  risk_of_infection <- risk_of_infection %>%
-    left_join(risk_of_infection_av, by = "person_id")
-
-  # save simulation
-  saveRDS(quantaDF, paste0("../simulations/2021-10-25/sim-quanta", s, ".rds"))
-  saveRDS(risk_of_infection, paste0("../simulations/2021-10-25/sim-transRisk", s, ".rds"))
-}
+  group_by(sim) %>%
+  mutate(share_n = n / sum(n) * 100) %>%
+  ungroup() %>%
+  filter(risk_cat == "high")
+strMinorVsHigh <- paste0("Total risk attendees: ", 
+                         round_k(mean(transRisk_MinorVsHigh$n), 0),
+                         " \u00B1 ",
+                         round_k(sd(transRisk_MinorVsHigh$n), 0),
+                         " (",
+                         round_k(mean(transRisk_MinorVsHigh$share_n), 1),
+                         "% \u00B1 ",
+                         round_k(sd(transRisk_MinorVsHigh$share_n), 1),
+                         "%)")
+transRisk_high <- transRisk_qq %>%
+  dplyr::filter(risk_cat %in% group_labs) 
+transRisk_qq_pl <- transRisk_high %>%
+  mutate(risk_cat = factor(risk_cat, group_labs)) %>%
+  ggplot(aes(x = risk_cat, y = n, fill = risk_cat)) +
+  stat_boxplot(geom ='errorbar', width = .2) +
+  geom_boxplot(width = .2, outlier.shape = NA) +
+  annotate("text", x = .5, y = 18.5, label = strMinorVsHigh, size = 8 / cm(1), hjust = 0, vjust = 0) +
+  scale_y_continuous(sec.axis = sec_axis(~ . / nrow(transRisk_indv), labels = scales::percent, name = "Proportion of attendees"),
+                     expand = c(0,0), limits = c(0,20)) +
+  scale_fill_manual(values = wes_palette("Darjeeling2")) +
+  labs(x = "Risk of infection", y = "Number of attendees") +
+  theme_bw2() +
+  theme(legend.position = "none")
+save_plot(transRisk_qq_pl, pdf_file = "../results/modeling/high-risk-of-infection.pdf", w = aw, h = ah)
 ```
 
-## Evaluation
-
-### TB risk
+### Quanta concentration
 
 ```{r}
-# files
-transRiskFiles <- list.files("../simulations/2021-10-25", full.names = T)
-transRiskFiles <- transRiskFiles[grepl("transRisk", transRiskFiles)]
-
-# read files
-transRiskDF <- data.frame(file = transRiskFiles) %>%
-  mutate(simulation = stringi::stri_extract(basename(file), regex = "\\d{1,4}"), 
-         data = lapply(file, readRDS)) %>%
-  unnest(cols = c(data))
-
-# QQ Plot
-transRiskDF_Q <- transRiskDF %>%
-  dplyr::select(simulation, P_SP, P_WM) %>%
-  melt("simulation") %>%
-  group_by(simulation, variable) %>%
-  summarize(quantile = seq(0.05, .95, 0.05),
-            value = quantile(value, seq(0.05, 0.95, 0.05))) %>%
+# average quanta concentration
+quantaDF_av <- sim_results %>%
+  mutate(daytime = ifelse(hhmm <= "12:00", "Morning", "Afternoon"),
+         daytime = factor(daytime, levels = c("Morning", "Afternoon"))) %>%
+  group_by(daytime, id, location) %>%
+  summarize(N = mean(N)) %>%
   ungroup()
 
-transRiskDF_Q_av <- transRiskDF_Q %>%
-  group_by(variable, quantile) %>%
-  summarize(value = mean(value)) %>%
-  ungroup()
+a_ac_grid = -9
+quantaDF_av_wr <- quantaDF_av %>%
+  group_by(daytime) %>%
+  head(nrow(waiting_room_df)) %>%
+  mutate(id = as.character(id)) %>%
+  ungroup() %>%
+  left_join(waiting_room_df, by = c("id")) %>%
+  rename(x = long, y = lat) %>%
+  rotate_xy(a_ac_grid)
+quantaDF_av_pa <- quantaDF_av %>%
+  group_by(daytime) %>%
+  tail(nrow(passage_df)) %>%
+  mutate(id = as.character(1:n())) %>%
+  ungroup() %>%
+  left_join(passage_df, by = c("id")) %>%
+  rename(x = long, y = lat) %>%
+  rotate_xy(a_ac_grid)
+quantaDF_av_tb <- quantaDF_av %>%
+  filter(location == "tb room") %>%
+  group_by(daytime) %>%
+  mutate(id = as.character(1:n())) %>%
+  ungroup() %>%
+  left_join(tb_room_df, by = c("id")) %>%
+  rename(x = long, y = lat) %>%
+  rotate_xy(a_ac_grid)
 
-transRiskDF_Q_av %>%
-  ggplot(aes(x = quantile, y = value, color = variable)) +
-  geom_line()
+qc_pl <- ggplot(mapping = aes(x = x, y = y, group = group, fill = N)) +
+  geom_polygon(data = quantaDF_av_wr) +
+  geom_polygon(data = quantaDF_av_pa) +
+  geom_polygon(data = quantaDF_av_tb) +
+  coord_fixed() +
+  #scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = seq(1,6,1)*1e-6, limits = c(0,100), na.value = NA) +
+  facet_wrap(~daytime, ncol = 1)
+qc_pl
+qc_pl <- spatial_ggpl(qc_pl)
 
-# average transmission risk across patients
-average_tr <- transRiskDF %>%
-  dplyr::select(simulation, P_SP, P_WM) %>%
-  mutate(P_SP = P_SP * 100,
-         P_WM = P_WM * 100) %>%
-  melt("simulation") %>%
-  group_by(simulation, variable) %>%
-  summarize(`Mean` = mean(value),
-            `Standard deviation` = sd(value),
-            `Median` = median(value),
-            `Tail risk (95%-Quantile)` = quantile(value, .95)) %>%
-  ungroup()
-
-figA <- average_tr %>%
-  rename(model = variable) %>%
-  mutate(model = ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling")) %>%
-  melt(c("simulation", "model")) %>%
-  ggplot(aes(x = value, y = model, fill = model, color = model)) +
-  stat_dots(quantiles = 100) +
-  facet_wrap(~ variable, scales = "free_x") +
-  scale_color_manual(values = cbPalette[2:3]) +
-  scale_fill_manual(values = cbPalette[2:3]) +
-  labs(x = "Risk of infection (%)", 
-       title = "Fig. 1a: Risk of infection across patients",
-       subtitle = "Colored dots show individual simulation results.") +
-  theme_bw2() +
-  theme(legend.position = "top", axis.title.y = element_blank(), axis.text.y = element_blank(), legend.title = element_blank(),
-        legend.justification='left', legend.direction='horizontal')
-
-figA
-
-average_tr %>%
-  dplyr::select(-simulation) %>%
-  group_by(variable) %>%
-  median_qi() %>%
-  mutate_if(is.numeric, round, 2)
-
-figB <- transRiskDF %>%
-  ggplot(aes(x = P_SP, y = P_WM)) +
-  geom_point(shape = 1) +
-  geom_abline(mapping = aes(intercept = 0, slope = 1), color = "red", linetype = "dashed") +
-  annotate("richtext", x = 0.75, y = 0.82, 
-           label = "<span style = 'background-color:#fff'>Perfect correlation</span>", 
-           size = 8 / cm(1), color = "red", angle = 22, label.color = NA) +
-  geom_smooth(method = "lm") +
-  annotate("richtext", x = 0.85, y = 0.55, 
-           label = "<span style = 'background-color:#fff'>Empirical correlation</span>", 
-           size = 8 / cm(1), color = "blue", angle = 16, label.color = NA) +
-  labs(x = "Risk of infection (%) with spatiotemporal modeling", 
-       y = "Risk of infection (%) with temporal modeling",
-       title = "Fig. 1b: Patient-specific risk of infection",
-       subtitle = "One dot per patient and simulation. Dots off (on) the diagonal imply a (no) differing risk of infection by modeling.") +
-  scale_x_continuous(expand = c(0,0), limits = c(0,1), label = function(x) x * 100) +
-  scale_y_continuous(expand = c(0,0), limits = c(0,1), label = function(x) x * 100) +
-  theme_bw2() +
-  theme(plot.margin = unit(c(5.5, 7.5, 5.5, 5.5), "points"))
-
-figB
-
-ggsave("../results/model-comparison.png", grid.arrange(figA, figB, ncol = 1), width = 16 / cm(1), height = 16 / cm(1))
-
-# visualization of spatial approach
-transRriskSP_pl <- transRiskDF %>%
-  mutate(P_SP = P_SP * 100) %>%
-  ggplot(aes(x = P_SP, y = simulation)) + # , color = ..x..
-  stat_dots(quantiles = 20, color = "black") +
-  #scale_color_viridis() +
-  labs(y = "Simulation", x = "Risk of infection (%)",
-       title = "Figure 1: Risk of infection", subtitle = "Quantile distribution of patient-specific risk of infection (%) for each simulation") +
-  scale_x_sqrt(limits = c(0,NA), expand = c(0,0)) +
-  scale_y_discrete(expand = c(0,0)) +
-  theme_bw2() +
-  theme(legend.position = "none")
-
-transRriskSP_pl
-
-# low, medium, high risk
-transRiskDF_cat <- transRiskDF %>%
-  dplyr::select(simulation, P_SP, P_WM) %>%
-  melt("simulation") %>%
-  group_by(simulation, variable) %>%
-  summarize(`Very low` = sum(value <= .01),
-            Low = sum(value > .01 & value <= .05),
-            Medium = sum(value > .05 & value <= .1),
-            High = sum(value > .1 & value <= .5),
-            `Very high` = sum(value > .5)) %>%
-  ungroup()
-
-figB <- transRiskDF_cat %>%
-  rename(model = variable) %>%
-  mutate(model = factor(ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling"),
-                        levels = c("Spatiotemporal modeling", "Temporal modeling"))) %>%
-  melt(c("simulation", "model")) %>%
-  group_by(model, variable) %>%
-  summarize(value = mean(value)) %>%
-  ungroup() %>%
-  ggplot(aes(y = variable, fill = model, x = value)) +
-  geom_bar(stat = "identity", position = position_dodge2()) +
-  scale_x_log10(expand = c(0,0)) + 
-  scale_fill_manual(values = cbPalette[c(2,1)]) +
-  labs(title = "B | Average number of visitors by risk of infection",
-       subtitle = "Number of visitors with very high (>50%), high (10-50%), medium (5-10%), low (1-5%), and very low (<1%) risk of infection.",
-       x = "Number of visitors") +
-  theme_bw2() +
-  theme(legend.position = "top", legend.justification='left', legend.direction='horizontal',
-        legend.title = element_blank(), legend.key.width = unit(2.5, "cm"),
-        axis.title.y = element_blank(),
-        plot.title.position = "plot")
-
-figB
-
-transRiskDF_cat_alt <- transRiskDF %>%
-  dplyr::select(simulation, P_SP, P_WM) %>%
-  melt("simulation") %>%
-  group_by(simulation, variable) %>%
-  summarize(Low = sum(value <= .1),
-            Medium = sum(value > .1 & value <= .5),
-            High = sum(value > .5)) %>%
-  ungroup()
-
-figB_alt <- transRiskDF_cat_alt %>%
-  rename(model = variable) %>%
-  mutate(model = factor(ifelse(model == "P_SP", "Spatiotemporal modeling", "Temporal modeling"),
-                        levels = c("Spatiotemporal modeling", "Temporal modeling"))) %>%
-  melt(c("simulation", "model")) %>%
-  group_by(model, variable) %>%
-  summarize(value = mean(value)) %>%
-  ungroup() %>%
-  ggplot(aes(y = variable, fill = model, x = value)) +
-  geom_bar(stat = "identity", position = position_dodge2()) +
-  scale_x_log10(expand = expansion(mult = c(0,.01))) + 
-  scale_fill_manual(values = cbPalette[c(2,1)], breaks = c("Temporal modeling", "Spatiotemporal modeling")) +
-  labs(title = "B | Estimated number of visitors by risk of infection",
-       subtitle = "Average number of visitors with high (>50%), medium (10-50%), and low (<10%) risk of infection.",
-       x = "Number of visitors") +
-  theme_bw2() +
-  theme(legend.position = c(0.675, 0.825), legend.justification='left', legend.direction='vertical',
-        legend.title = element_blank(), legend.key.width = unit(1.5, "cm"),
-        axis.title.y = element_blank(),
-        plot.title.position = "plot")
-
-figB_alt
+qc_pl
+save_plot(qc_pl, pdf_file = "../results/modeling/quanta-concentration.pdf", w = aw, h = 2*ah)
 ```
-
-### Quanta
-
-```{r}
-# files
-quantaFiles <- list.files("../simulations/2021-10-25", full.names = T)
-quantaFiles <- quantaFiles[grepl("quanta", quantaFiles)]
-
-# read files
-quantaDF <- data.frame(file = quantaFiles) %>%
-  mutate(simulation = stringi::stri_extract(basename(file), regex = "\\d{1,4}"), 
-         data = lapply(file, readRDS)) %>%
-  unnest(cols = c(data))
-
-# average quanta concentration
-quantaDF_av <- quantaDF %>%
-  filter(location == "waiting room") %>%
-  # mutate(daytime = ifelse(time <= "2021-10-25 11:00:00", "Morning", 
-  #                         ifelse(time <= "2021-10-25 13:00:00", "Noon", "Afternoon"))) %>%
-  mutate(daytime = ifelse(time <= "2021-10-25 12:00:00", "Morning", "Afternoon")) %>%
-  group_by(daytime, id) %>%
-  summarize(N_SP = mean(N_SP)) %>%
-  ungroup()
-
-wrRoomSP_df <- fortify(wrRoomSP) %>%
-  left_join(quantaDF_av %>% mutate(id = as.character(id)), by = "id")
-
-figA <- wrRoomSP_df %>%
-  # mutate(daytime = factor(daytime, c("Morning", "Noon", "Afternoon"))) %>%
-  mutate(daytime = factor(daytime, c("Morning", "Afternoon"))) %>%
-  ggplot(aes(x = lat, y = long, group = group, fill = N_SP)) +
-  geom_polygon() +
-  facet_wrap(~ daytime) +
-  scale_fill_gradient(low = "yellow", high = "red", na.value = NA) +
-  labs(title = "A | Estimated spatiotemporal quanta concentration in the clinic by daytime",
-       # subtitle = "Quanta concentration (quanta/m3) in the morning (7:30 to 11am), noon (11am to 1pm), and afternoon (1 to 4pm).") +
-       subtitle = "Average quanta concentration (quanta/m3) in the morning (7:30 to 12am) and afternoon (12am to 4pm).") +
-  geom_path(color = "white", size = 0.2) +
-  theme_bw2() +
-  theme(axis.title = element_blank(), axis.text = element_blank(),
-        legend.position = "top", legend.justification='left', legend.direction='horizontal',
-        legend.title = element_blank(), legend.key.width = unit(2.5, "cm"),
-        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-
-figA
-
-ggsave("../results/model-comparison.png", grid.arrange(figA, figB_alt, ncol = 1), width = 16 / cm(1), height = 16 / cm(1))
-```

### 2023-03-14 analysis/simSpatiotempModel.R

@@ -39,7 +39,6 @@ prep_data <- function(
     sel_date,
     seed = 12345,
     nSim = 1000,
-    min_duration = 5,
     cellVolume = 1,
     co2_exhaled = 40000,
     co2_outdoor = 400,
@@ -51,10 +50,7 @@ prep_data <- function(
   set.seed(seed)
   
   # patient movement data
-  patMov <- readRDS(paste("data-clean",clinic,"patient-tracking-data",sel_date,"patient-id-data.rds",sep="/")) %>%
-    group_by(patient_id) %>%
-    filter(as.numeric(difftime(last(time), first(time), units = "mins")) > min_duration) %>%
-    ungroup() %>%
+  patMov <- readRDS(paste("data-clean",clinic,"patient-tracking-data",sel_date,"patient-id-data_matched-filtered.rds",sep="/")) %>%
     mutate(location = ifelse(is_waitingroom|is_passage, "waiting room", ifelse(is_tbroom, "tb room", NA)),
            hhmm = format(time, format = "%H:%M")) %>%
     filter(!is.na(location)) 
@@ -230,7 +226,7 @@ spatiotemporal_model <- function(time_period, patMov, co2_and_n, room_coords, ce
   
   #### Risk of infection ####
   risk_of_infection <- patMov %>%
-    dplyr::select(patient_id, location, id, hhmm) %>%
+    dplyr::select(patient_id, location, id,  hhmm) %>%
     filter(!(patient_id %in% tb_ids)) %>%
     left_join(quantaDF %>% 
                 mutate(hhmm = format(time, format = "%H:%M")) %>%
@@ -238,7 +234,7 @@ spatiotemporal_model <- function(time_period, patMov, co2_and_n, room_coords, ce
               by = c("id", "location", "hhmm")) %>%
     group_by(patient_id) %>%
     mutate(exposure_time = n(),
-           across(c(N, N_ns), ~ 1-exp(-sum(.x, na.rm = T)), .names = "P_{.col}")) %>%
+           across(c(N, N_ns), ~ compute_P(sum(.x, na.rm = T)), .names = "P_{.col}")) %>%
     ungroup()
   
   return(risk_of_infection)
@@ -262,7 +258,7 @@ message(sprintf("Info: Cluster using %i cores", foreach::getDoParWorkers()))
 for(cl in clinics) {
   
   message(sprintf("Clinic: %s", cl))
[…]

### 2023-03-14 preprocessing/data_check_xovis.R

@@ -1,0 +1,31 @@
+library(tidyverse)
+
+files <- list.files("data-raw/Masi/xovis/Study_Data_Masi_2021/export", full.names = T)
+file_names <- basename(files)
+dates <- gsub("_exported_data[[:punct:]]csv", "", file_names)
+dates <- gsub("_", "-", dates)
+
+i = 21
+print(dates[i])
+masi <- read.csv(files[i], header = T) %>% 
+  set_names(c("obs_id", "time", "x", "y", "height")) %>%
+  mutate(time = as.POSIXct(paste(dates[i], time)))
+min(masi$time); max(masi$time)
+length(unique(masi$obs_id))
+dt <- masi %>%
+  group_by(obs_id) %>%
+  mutate(dt = as.numeric(difftime(time, lag(time), units = "secs"))) %>%
+  ungroup() %>%
+  na.omit()
+summary(dt$dt)
+sum(dt$dt>1)
+sum(dt$dt<1)
+plot(unique(sort(dt$time)))
+bad_ones <- filter(dt, dt > 1 | dt < 1) 
+bad_ones %>% View()
+bad_ids <- unique(bad_ones$obs_id)
+
+saveRDS(masi, paste0("data-raw/Masi/xovis/", dates[i], ".rds"))
+saveRDS(masi %>% filter(!obs_id %in% bad_ids), paste0("data-raw/Masi/xovis/", dates[i], ".rds"))
+
+  

### 2023-03-14 preprocessing/match_clinical-tracking.R

@@ -1,0 +1,97 @@
+#### Libraries ####
+
+library(tidyverse)
+
+#### Matching function ####
+
+#' Matching clinical records to patient tracking data
+#' 
+#' @param clinic clinic
+#' @param sel_date date of tracking data
+#' @param min_time_in_cl minimum time spent in clinic in order for the id to be considered a clinical attendee
+#' @param min_time_at_re minimum time a track should have stayed at the reception
+#' @param min_duration_after_re minimum time a patient should have been in the clinic after she was at the reception
+#' @param max_tolerance maximum delay of clinical record after a patient was at the reception
+
+match_clinical_tracking <- function(clinic, sel_date, min_time_in_cl = 5 * 60, min_time_at_re = 5, min_duration_after_re = 5 * 60, max_tolerance = 15 * 60) {
+  
+  # directory of patient tracking data
+  save_str <- paste0("data-clean/", clinic, "/patient-tracking-data/", sel_date, "/")
+  
+  # clinical data
+  clinical <- readRDS(paste0("data-clean/", clinic, "/clinical-data/clinical-data.rds")) %>%
+    filter(date == as.Date(sel_date)) %>%
+    group_by(patient_id) %>%
+    slice(1) %>%
+    ungroup() %>%
+    rename(clinical_patient_id = patient_id,
+           clinical_record_time = date_time)
+  
+  # tracking data
+  tracking <- readRDS(paste0(save_str, "/patient-id-data.rds")) %>%
+    group_by(patient_id) %>%
+    filter(as.numeric(difftime(last(time), first(time), units = "secs")) > min_time_in_cl) %>%
+    ungroup()
+  
+  # tracking: time at reception
+  track_time_at_re <- tracking %>%
+    group_by(patient_id) %>%
+    filter(sum(is_reception) > min_time_at_re) %>%
+    mutate(time_at_re = time[is_reception][1]) %>%
+    filter(as.numeric(difftime(last(time), time_at_re, units = "secs")) > min_duration_after_re) %>%
+    slice(1) %>%
+    ungroup() 
+  
+  # matching
+  merged <- merge(track_time_at_re %>% dplyr::select(patient_id, time_at_re),
+                  clinical %>% dplyr::select(clinical_patient_id, clinical_record_time, tb_suspect),
+                  id = NULL) %>%
[…]

### 2023-03-14 preprocessing/prep_room.R

@@ -4,11 +4,13 @@ library(terra)
 source("utils/spatial.r")
 
 # Data 
-files <- list.files("data-raw/Masi/xovis/Study_Data_Masi_2021/export", full.names = T)
-masi <- do.call(rbind, lapply(files, function(x) read.csv(x, header = T) %>% set_names(c("obs_id", "time", "x", "y", "height")))) 
+files <- list.files("data-raw/Masi/xovis/", full.names = T)
+files <- files[grepl("rds", files)]
+masi <- do.call(rbind, lapply(files, readRDS))
+masi_rot <- rotate_xy(masi, a = -9)
 
 pl <- ggplot() + 
-  geom_bin2d(data = masi, mapping = aes(x = x, y = y), binwidth = 100) +
+  geom_bin2d(data = masi_rot, mapping = aes(x = x, y = y), binwidth = 100) +
   scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = c(1,10,100,1000,10000), limits = c(0,10000)) 
 
 pl
@@ -41,57 +43,58 @@ line_point <- function(pt, l, r, start = T) {
 }
 
 # outline
-width <- 22750
+width <- 23000
 breadth <- 8250
-rotation <- 9
-bottom_left_pt <- c(-6200, -8600)
+rotation <- 0
+bottom_left_pt <- c(-15800, -7500)
 bottom_right_pt <- line_point(bottom_left_pt, width, rotation)
 outline <- parallelogram(bottom_left_pt, bottom_right_pt, width, breadth, rotation)
 
 # waiting room
-wr_width <- 10000
+wr_width <- 9750 + 800
 wr_breadth <- 5700
-wr_top_right_pt <- outline[3,]
-wr_bottom_right_pt <- line_point(wr_top_right_pt, wr_width, rotation, start = F)
-wr <- parallelogram(wr_bottom_right_pt, wr_top_right_pt, wr_width, -wr_breadth, rotation)
+wr_top_left_pt <- outline[4,]
+wr_top_right_pt <- line_point(wr_top_left_pt, wr_width, rotation)
+wr <- parallelogram(wr_top_left_pt, wr_top_right_pt, wr_width, -wr_breadth, rotation)
 
 # sputum room
-sp_width <- 5000
[…]

### 2023-03-14 preprocessing/prep_xovis_matching.R

@@ -1,6 +1,7 @@
 #### Libraries ####
 
 library(tidyverse)
+options(dplyr.summarise.inform = FALSE)
 library(reshape2)
 library(lubridate)
 library(sf)
@@ -9,12 +10,16 @@ library(raster)
 library(terra)
 library(gridExtra)
 library(VennDiagram)
+library(parallel)
+library(foreach)
 source("utils/spatial.r")
 source("utils/plotting.r")
 
 
 #' Algorith to match observation IDs
 #' 
+#' @param location string
+#' @param rooms fortified dataframe of spatial polygons of the rooms in the building
 #' @param date date %Y-%m-%d
 #' @param max_time maximum time (in seconds) to look ahead for matches
 #' @param min_standing_height minimum standing height
@@ -26,7 +31,9 @@ source("utils/plotting.r")
 #' @param max_heightdiff maximum height difference for a person not changing her posture
 #' @param min_duration minimum time (in minutes) to stay in the clinic to be considered
 
-match_xovis <- function(date = "2021-10-25", 
+match_xovis <- function(location,
+                        rooms,
+                        date, 
                         max_time = 30 * 60, 
                         min_standing_height = 1400, 
                         max_mov_time = c(3, 5, 5, 5, 10, 10, 15, 15, 30, 60), 
@@ -41,8 +48,8 @@ match_xovis <- function(date = "2021-10-25",
   #### Data ####
   
   # xovis
-  file <- paste0("data-raw/Masi/xovis/", date, "-with_roominfo.rds")
-  save_dir <- paste0("data-clean/Masi/patient-tracking-data/", date, "/")
+  file <- paste0("data-raw/", location, "/xovis/", date, ".rds")
+  save_dir <- paste0("data-clean/", location, "/patient-tracking-data/", date, "/")
   if (!dir.exists(save_dir)) {
     dir.create(save_dir)
   }
@@ -61,12 +68,6 @@ match_xovis <- function(date = "2021-10-25",
   
   message(sprintf("%i of %i (%i percent) observation IDs removed; %i remaining", 
                   n_oid-n_oid_filt, n_oid, round(100 - 100 * n_oid_filt / n_oid), n_oid_filt))
-  
-  # roomplan
[…]

### 2023-03-14 preprocessing/prep_xovis_room.R

@@ -17,20 +17,17 @@ st_crs(clinic_sf) <- NA
 clinic_df <- fortify(clinic_sf)
 
 # xovis raw data files
-files <- list.files("data-raw/Masi/xovis/Study_Data_Masi_2021/export", full.names = T)
+files <- list.files("data-raw/Masi/xovis/", full.names = T)
+files <- files[grepl("rds", files)]
 
 for (f in files) {
   # file info
   file_name <- basename(f)
-  date <- gsub("_exported_data[[:punct:]]csv", "", file_name)
-  date <- gsub("_", "-", date)
-  file_name <- paste0(date, "-", "with_roominfo")
-  
-  
+  message(sprintf("File: %s", f))
+
   # read data
-  df <- read.csv(f, header = T) %>% 
-    set_names(c("obs_id", "time", "x", "y", "height"))  %>%
-    mutate(time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M:%S"))
+  df <- readRDS(f) %>% 
+    rotate_xy(a = -9)
   
   # add room info
   df$point <- map2(df$x, df$y, function(xi,yi) sfheaders::sf_point(c(xi,yi)))
@@ -46,7 +43,7 @@ for (f in files) {
   df$is_in_room_right <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 13), T, F) ) 
   df$is_exit <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) %in% c(6, 14:15)), T, F)) # entrance and corridors are considered as exits
  
-  saveRDS(df, file = paste0("data-raw/Masi/xovis/", file_name, ".rds")) 
+  saveRDS(df, file = paste0("data-raw/Masi/xovis/", file_name)) 
 }
 
 

### 2023-03-14 utils/plotting.R

@@ -1,7 +1,11 @@
 require(ggpattern)
 require(gridExtra)
-require(GGally)
+library(GGally)
 library(ggtext)
+library(wesanderson)
+
+# round format
+round_k <- function(x, k = 0) format(round(x, k), nsmall = k)
 
 # ggplot theme for Nature style
 text_size = 8

### 2023-03-14 utils/spatial.r

@@ -87,3 +87,20 @@ find_raster <- function(x, y, cellCoordsXY) {
   return(closest_id)
 }
 
+
+#' Rotate x and y date by a specific angle
+#' 
+#' @param df data frame with x and y coordinate
+#' @param a angle
+#' 
+
+rotate_xy <- function(df, a) {
+  xy <- df %>%
+    dplyr::select(x, y) %>%
+    adespatial::rotation(a*pi/180) %>%
+    as.data.frame() %>%
+    set_names(c("x", "y")) %>%
+    mutate(x = -x)
+  df <- cbind(df %>% dplyr::select(-x,-y), xy)
+  return(df)
+}

### 2023-03-14 utils/trans_risk.r

@@ -61,11 +61,12 @@ compute_Nt <- function(E, RR, N0, V) {
 #' 
 #' 
 #' @param N quanta concentration (quanta / m3)
-#' @param t duration of exposure (in minutes)
+#' @param t duration of exposure (in seconds)
+#' @param p is the breathing rate (in L/seconds)
 #' 
 
-compute_P <- function(N, t = 1) {
-  1 - exp(-N * t)
+compute_P <- function(N, t = 1, p = 8.0 / 60) {
+  1 - exp(-N * t * p)
 } 
 
 

### 2023-03-20 analysis/spatiotemporal-transmission-risk.Rmd
data, workflow and minor fixes

@@
 ```{r}
-selected_dates <- as.Date(list.files("../data-clean/Masi/patient-tracking-data")[-1])
+selected_dates <- list.files("../simulations/Masi/")
+selected_dates <- as.Date(selected_dates[!grepl("old", selected_dates)])
+```
+
+```{r}
+# overview
+files_by_date <- data.frame(
+  directory = list.files("../data-clean/Masi/combined-data", full.names = T)
+) %>%
+  mutate(date = basename(directory),
+         files_in_dir = map(directory, list.files),
+         Clinical = map_lgl(files_in_dir, function(x) any(grepl("clinical", x))),
+         CO2 = map_lgl(files_in_dir, function(x) any(grepl("environmental", x))),
+         Tracking = map_lgl(files_in_dir, function(x) any(grepl("tracking", x))))
+
+files_by_date %>%
+  dplyr::select(date, Clinical, CO2, Tracking) %>%
+  reshape2::melt("date") %>%
+  ggplot(aes(x = variable, y = date, fill = value)) +
+  geom_tile() +
+  labs(x = "Dataset", fill = "Available") +
+  scale_fill_manual(values = wes_palette("Chevalier1")[c(2,1)]) +
+  theme_classic() +
+  theme(axis.title.y = element_blank())
+ggsave(filename = "../insights/data-check-dataset-availability.pdf", width = 16 / cm(1), height = 21 / cm(1))
 ```
@@
   geom_bar(stat = "identity", fill = "grey", color = "black", width = .5) +
-  annotate("text", x = "Nov 10", y = mean(share_tb$n_tb), label = paste0("Mean = ",round(mean(share_tb$n_tb), 1)), color = "darkred", size = 8 / cm(1), vjust = -1) +
+  annotate("text", x = "Nov 12", y = mean(share_tb$n_tb), label = paste0("Mean = ",round(mean(share_tb$n_tb), 1)), color = "darkred", size = 8 / cm(1), vjust = -1) +
   geom_hline(aes(yintercept = mean(share_tb$n_tb)), color = "darkred") +
@@
   scale_x_discrete(breaks = format(selected_dates, format = "%b %d")[c(1, 4, 8, 12)]) +
-  labs(y = "No. of masked TB patients") +
+  labs(y = "Number of masked TB patients") +
   theme_classic() +
@@
         text = element_text(size = text_descr))
+share_tb_pl
 ggsave(plot = share_tb_pl, filename = "../doc/IWHOD/presentation/figures/share-tb-patients.png", w = width_descr, h = height_descr)
@@
         text = element_text(size = text_descr)) 
+co2_pl
 ggsave(plot = co2_pl, filename = "../doc/IWHOD/presentation/figures/co2-levels.png", width = width_descr, height = height_descr)
@@
 # test
-waiting_room_df$value <- rnorm(nrow(waiting_room_df))
-passage_df$value <- rnorm(nrow(passage_df))
-tb_room_df$value <- rnorm(nrow(tb_room_df))
-
-ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
-  geom_polygon(data = waiting_room_df) +
-  geom_polygon(data = passage_df) +
-  geom_polygon(data = tb_room_df) +
-  ggpubr::background_image(clinic_img) +
-  annotate("text", x = -13200, y = -5300, label = "Entrance", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
-  annotate("text", x = -13400, y = -2300, label = "Waiting room", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
-  annotate("text", x = -5000, y = -4100, label = "Registration", size = 8 / cm(1), fontface = 2, angle = 90, hjust=0,vjust=0) +
-  annotate("text", x = 2600, y = -2700, label = "TB room", size = 8 / cm(1), fontface = 2) +
-  scale_y_continuous(expand = expansion(add=c(2500, 4000))) +
-  scale_x_continuous(expand = expansion(add=c(400, 500))) +
-  labs(fill = "Density") +
-  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
-        legend.position = c(0.825,0.075), legend.direction = "horizontal", legend.title = element_blank(), #legend.title = element_text(vjust = .75),
-        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 3, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank())
+# waiting_room_df$value <- rnorm(nrow(waiting_room_df))
+# passage_df$value <- rnorm(nrow(passage_df))
+# tb_room_df$value <- rnorm(nrow(tb_room_df))
+# 
+# ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
+#   geom_polygon(data = waiting_room_df) +
+#   geom_polygon(data = passage_df) +
+#   geom_polygon(data = tb_room_df) +
+#   ggpubr::background_image(clinic_img) +
+#   annotate("text", x = -13200, y = -5300, label = "Entrance", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
+#   annotate("text", x = -13400, y = -2300, label = "Waiting room", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
+#   annotate("text", x = -5000, y = -4100, label = "Registration", size = 8 / cm(1), fontface = 2, angle = 90, hjust=0,vjust=0) +
+#   annotate("text", x = 2600, y = -2700, label = "TB room", size = 8 / cm(1), fontface = 2) +
+#   scale_y_continuous(expand = expansion(add=c(2500, 4000))) +
+#   scale_x_continuous(expand = expansion(add=c(400, 500))) +
+#   labs(fill = "Density") +
+#   theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
+#         legend.position = c(0.825,0.075), legend.direction = "horizontal", legend.title = element_blank(), #legend.title = element_text(vjust = .75),
+#         panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 3, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
+#         panel.grid.minor = element_blank())
 ```
@@
 ```{r}
-files <- list.files("../data-clean/Masi/patient-tracking-data/", full.names = T)[-1]
-no_people <- do.call(rbind, lapply(files, function(x) readRDS(paste0(x, "/patient-id-data_matched-filtered.rds")))) %>%
+no_people <- do.call(rbind, lapply(selected_dates, function(x) readRDS(paste0("../data-clean/Masi/combined-data/", x, "/tracking-linked-clinical-data.rds")))) %>%
   mutate(date = as.Date(format(time, format = "%Y-%m-%d")),
@@
   geom_bin2d(binwidth = 100) +
-  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log", breaks = c(1, 10, 100, 1000, 10000)) +
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log", breaks = c(1, 10, 100, 1000, 10000), labels = function(x) signif(x / nrow(no_people_time) * 60 * 24, 1)) +
     ggpubr::background_image(clinic_img) +
@@
     annotate("text", x = 3000, y = -2700, label = "TB room", size = text_descr / cm(1)) +
-    scale_y_continuous(expand = expansion(mult=c(0.05, 0.15))) +
+    scale_y_continuous(expand = expansion(mult=c(0.05, 0.215))) +
     scale_x_continuous(expand = expansion(mult=c(0.03, 0))) +
-    labs(fill = "Density") +
+    labs(fill = "Density (#tracks/day)") +
     theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.3, "cm"),
@@
   guides(fill = guide_legend(title.position = "top", label.position = "bottom"))
+no_people_spatial_pl
 ggsave(plot=no_people_spatial_pl, filename = "../doc/IWHOD/presentation/figures/no-people-spatial.png", width = width_descr, height = height_descr)
@@
   stat_lineribbon(alpha = .5, size = .5) +
-  scale_x_continuous(expand = c(0,0)) +
-  scale_y_continuous(expand = c(0,0)) +
+  scale_x_continuous(expand = c(0,0), limits = c(0, 10), breaks = seq(0, 10, 2)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0,100)) +
   scale_fill_brewer(palette = "OrRd") +
-  labs(x = "Distance (m)", y = "Relative quanta concentration (%)", fill = "Quantile") +
-  theme_bw2() +
-  theme(legend.position = c(.8,.9), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
-        plot.margin = margin(5.5, 10, 5.5, 5.5)) +
-  guides(fill = guide_legend(title.position = "top"))
-save_plot(relQC_pl, pdf_file = "../results/inputs/spatial-relative-quanta-concentration.pdf", w = aw, h = ah)
+  labs(x = "Distance (m)", y = "Rel. prob. (%)", fill = "Quantile") +
+  theme_classic() +
+  theme(legend.position = c(.9,.85), legend.direction = "vertical", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        legend.spacing.y = unit(.1, "cm"),
+        plot.margin = margin(5.5, 10, 5.5, 5.5),
+        text = element_text(size = text_descr),
+        legend.title = element_blank()) +
+  guides(fill = guide_legend(title.position = "top", byrow = T))
+ggsave(plot = relQC_pl, filename = "../doc/IWHOD/presentation/figures/spatial-relative-quanta-concentration.png", width = width_descr * 0.66, height = height_descr * 0.66)
 ```
@@
 
-```{r}
-sim_results <- list.files("../simulations/Masi/2021-10-25/", full.names = T)
-sim_results <- do.call(rbind, lapply(sim_results, readRDS))
-```
-
 ### Transmission risk
@@
 ```{r}
+# transmission risk 
+sim_files <- sapply("../simulations/Masi", list.files, full.names = T) 
+sim_files <- unlist(sapply(sim_files, list.files, full.names = T))
+tr_files <- sim_files[grepl("TR", sim_files)]
+
+average_transRisk <- function(file) {
+  # quanta concentration and transmission risk
+  qctr <- readRDS(file)
+
   # individual patient transmission risk by simulation
-transRisk <- sim_results %>%
-  group_by(patient_id, sim) %>%
+  tr <- qctr %>%
+    group_by(patient_id, date, sim) %>%
     slice(1) %>%
@@
 
-# Average transmission risk across patients
+  return(tr)
+}
+
+transRisk <- parallel::mclapply(tr_files, average_transRisk, mc.cores = 6)
+transRisk <- do.call(rbind, transRisk)
+
 transRisk_av <- transRisk %>%
-  group_by(sim) %>%
+    group_by(date, sim) %>%
     summarize(across(c(P_N, P_N_ns), mean)) %>%
@@
 
-mean_risk <- round(mean(transRisk_av$P_N * 100), 1)
-q25_risk <- round(quantile(transRisk_av$P_N * 100, .25), 1)
-q75_risk <- round(quantile(transRisk_av$P_N * 100, .75), 1)
-message(sprintf("Spatiotemporal transmission risk across clinical attendee: %s percent (IQR: %s to %s)", mean_risk, q25_risk, q75_risk))
-
-# individual transmission risk across simulations
-transRisk_indv <- transRisk %>%
-  group_by(patient_id) %>%
-  summarize(across(c(P_N, P_N_ns), list(mean = mean, median = median,q25 = function(x) quantile(x, .25), q75 = function(x) quantile(x, .75)))) %>%
+transRisk_av_daily <- transRisk_av %>%
+  group_by(date) %>%
+  summarize(across(c(P_N, P_N_ns), list(m = mean, s = sd))) %>%
   ungroup()
 
-transRisk_indv_pl <- transRisk_indv %>%
-  mutate(patient_id = factor(patient_id)) %>%
-  ggplot(aes(x = reorder(patient_id, -P_N_mean), y = P_N_mean)) +
-  geom_point(shape = 4, alpha = .5) +
-  geom_hline(aes(yintercept = mean_risk / 100), color = "darkred") +
-  geom_hline(aes(yintercept = q25_risk / 100), color = "darkred", linetype = "dashed") +
-  geom_hline(aes(yintercept = q75_risk / 100), color = "darkred", linetype = "dashed") +
-  annotate("text", x = 375, y = 0.0075, label = paste0("Average across patients = ", mean_risk, "% (IQR:", q25_risk, "-", q75_risk,")"),
-           size = 8 / cm(1), color = "darkred") +
-  scale_y_continuous(labels = function(x) x * 100, expand = expansion(mult = c(0., .05)), breaks = c(seq(0, 0.05, 0.01), 0.004)) +
-  labs(y = "Mean risk of infection (%)", x = "Clinic attendee") +
-  theme_bw2() +
-  theme(axis.text.x = element_blank())
-save_plot(transRisk_indv_pl, pdf_file = "../results/modeling/risk-of-infection.pdf", w = aw, h = ah)
+mean_risk_daily <- round(mean(transRisk_av_daily$P_N_m * 100), 2)
+range_daily <- round(range(transRisk_av_daily$P_N_s * 100), 2)
+message(sprintf("Spatiotemporal transmission risk across clinical attendee: %s (range: %s percent to %s percent)", mean_risk_daily, range_daily[1], range_daily[2]))
 
-# high risk of infection plot
-group_labs <- c("Small (1-5%)", "Medium (5-10%)", "High (>10%)")
-transRisk_qq <- transRisk %>%
-  mutate(risk_cat = base::cut(P_N * 100, breaks = c(-Inf, 1, 5, 10, Inf), labels = c("negligible risk", group_labs))) %>%
-  group_by(sim, risk_cat) %>%
-  summarize(n = n()) %>%
-  ungroup()
-transRisk_MinorVsHigh <- transRisk_qq %>%
-  mutate(risk_cat = ifelse(risk_cat == "negligible risk", "low", "high")) %>%
-  group_by(sim, risk_cat) %>%
-  summarize(n = sum(n)) %>%
+# high risk patients
+transRisk_high <- transRisk %>%
+  mutate(high_risk = ifelse(P_N > 0.01, 1, 0),
+         date = as.Date(date)) %>%
+  group_by(date, sim) %>%
+  summarize(n_risk = sum(high_risk)) %>%
   ungroup() %>%
   group_by(sim) %>%
-  mutate(share_n = n / sum(n) * 100) %>%
-  ungroup() %>%
-  filter(risk_cat == "high")
-strMinorVsHigh <- paste0("Total risk attendees: ", 
-                         round_k(mean(transRisk_MinorVsHigh$n), 0),
-                         " \u00B1 ",
-                         round_k(sd(transRisk_MinorVsHigh$n), 0),
-                         " (",
-                         round_k(mean(transRisk_MinorVsHigh$share_n), 1),
-                         "% \u00B1 ",
-                         round_k(sd(transRisk_MinorVsHigh$share_n), 1),
-                         "%)")
-transRisk_high <- transRisk_qq %>%
-  dplyr::filter(risk_cat %in% group_labs) 
-transRisk_qq_pl <- transRisk_high %>%
-  mutate(risk_cat = factor(risk_cat, group_labs)) %>%
-  ggplot(aes(x = risk_cat, y = n, fill = risk_cat)) +
-  stat_boxplot(geom ='errorbar', width = .2) +
-  geom_boxplot(width = .2, outlier.shape = NA) +
-  annotate("text", x = .5, y = 18.5, label = strMinorVsHigh, size = 8 / cm(1), hjust = 0, vjust = 0) +
-  scale_y_continuous(sec.axis = sec_axis(~ . / nrow(transRisk_indv), labels = scales::percent, name = "Proportion of attendees"),
-                     expand = c(0,0), limits = c(0,20)) +
-  scale_fill_manual(values = wes_palette("Darjeeling2")) +
-  labs(x = "Risk of infection", y = "Number of attendees") +
-  theme_bw2() +
-  theme(legend.position = "none")
-save_plot(transRisk_qq_pl, pdf_file = "../results/modeling/high-risk-of-infection.pdf", w = aw, h = ah)
+  arrange(desc(date)) %>%
+  mutate(date_letter = factor(LETTERS[2:(n()+1)], levels = LETTERS[1:(length(selected_dates)+1)])) %>%
+  ungroup()
+
+transRisk_high_daily <- transRisk_high %>%
+  group_by(date_letter) %>%
+  summarize(n_risk_m = mean(n_risk),
+            n_risk_s = sd(n_risk),
+            n_risk_l = quantile(n_risk, 0.025),
+            n_risk_u = quantile(n_risk, 0.975)) %>%
+  ungroup()
+
+transRisk_high_daily_av <- data.frame(
+  n_risk_m = mean(transRisk_high_daily$n_risk_m),
+  n_risk_s = sd(transRisk_high_daily$n_risk_m)) %>%
+  mutate(n_risk_u = n_risk_m + qnorm(.975) * n_risk_s,
+         n_risk_l = n_risk_m + qnorm(.025) * n_risk_s,
+         date_letter = factor(LETTERS[1], levels = LETTERS[1:(length(selected_dates)+1)]),
+         n_risk_l = ifelse(n_risk_l < 0, 0, n_risk_l))
+
+ylabs <- c("Average", rev(format(as.Date(selected_dates), format = "%b %d")))
+
+highRiskPat_pl <- ggplot() +
+  geom_vline(aes(xintercept = mean(transRisk_high_daily$n_risk_m)), linetype = "dotted", color = "darkred") +
+  stat_pointinterval(data = transRisk_high, mapping = aes(x = n_risk, y = date_letter, slab_color = after_stat(level)), 
+                    point_interval = "mean_qi", .width = c(.5, .9), point_size = 2, fill = "white", shape = 21, point_color = "darkred") +
+  geom_errorbar(data = transRisk_high_daily_av, mapping = aes(xmin = n_risk_l, xmax = n_risk_u, y = date_letter),
+                width = .2, color = "darkred") +
+  geom_point(data = transRisk_high_daily_av, mapping = aes(x = n_risk_m, y = date_letter),
+             shape = 23, color = "darkred", fill = "white", size = 4) +
+  scale_color_manual(values = scales::brewer_pal()(3)[-1], aesthetics = "slab_color") +
+  scale_x_continuous(expand = c(0.01,0)) +
+  scale_y_discrete(breaks = LETTERS[1:(nrow(transRisk_high_daily)+1)], labels = ylabs) +
+  labs(x = "Number of attendees") +
+  theme_classic() +
+  theme(text = element_text(size = text_descr),
+        axis.title.y = element_blank(), 
+        legend.position = "none")
+
+tab_data <- rbind(transRisk_high_daily, transRisk_high_daily_av) %>%
+  mutate(meanCI = paste0(add_ws(round(n_risk_m, 0)), " [", add_ws(round(n_risk_l, 0), n = 1), ", ", add_ws(round(n_risk_u, 0)), "]"))
+tab_base <- ggplot(data = tab_data, aes(y = date_letter)) +
+    ylab(NULL) + xlab("  ") + 
+    theme(plot.title = element_text(hjust = 0, face = 2), ## centering title on text
+        axis.text.x=element_text(color="white"), ## need text to be printed so it stays aligned with figure but white so it's invisible
+        axis.line=element_blank(),
+        axis.text.y=element_blank(),axis.ticks=element_blank(),
+        axis.title.y=element_blank(),legend.position="none",
+        panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
+        panel.grid.minor=element_blank(),plot.background=element_blank())
+summary_tab <- tab_base +
+  geom_text(aes(x = 0, label = meanCI), size = text_descr / cm(1)) +
+  ggtitle("Mean (95%-UI)")
+
+highRiskPat_pl_wTable <- grid.arrange(highRiskPat_pl, summary_tab, ncol = 2, heights = height_descr, widths = width_descr * c(0.7,.3))
+ggsave(plot = highRiskPat_pl_wTable, filename = "../doc/IWHOD/presentation/figures/high-risk-patients.png", width = width_descr * 1.5, height = height_descr * 1.5)
 ```
@@
 ```{r}
+# files
+qc_files <- sim_files[grepl("QC", sim_files)]
+
+average_quantaConc <- function(file) {
+  # quanta concentration and transmission risk
+  qctr <- readRDS(file)
+
   # average quanta concentration
-quantaDF_av <- sim_results %>%
-  mutate(daytime = ifelse(hhmm <= "12:00", "Morning", "Afternoon"),
+  quantaDF_av <- qctr %>%
+    mutate(daytime = ifelse(format(time, "%H:%M") <= "12:00", "Morning", "Afternoon"),
            daytime = factor(daytime, levels = c("Morning", "Afternoon"))) %>%
-  group_by(daytime, id, location) %>%
-  summarize(N = mean(N)) %>%
+    group_by(date, daytime, id, location) %>%
+    summarize(N = mean(N, na.rm = T)) %>%
     ungroup()
 
-a_ac_grid = -9
-quantaDF_av_wr <- quantaDF_av %>%
+  return(quantaDF_av)
+}
+
+quantaConc <- parallel::mclapply(qc_files, average_quantaConc, mc.cores = 6)
+quantaConc <- do.call(rbind, quantaConc)
+
+quantaConc_av <- quantaConc %>%
+  group_by(date, daytime, location, id) %>%
+  summarize(N = mean(N, na.rm = T)) %>%
+  ungroup() %>%
+  group_by(daytime, location, id) %>%
+  summarize(N = mean(N, na.rm = T)) %>%
+  ungroup()
+
+quantaConc_av_wr <- quantaConc_av %>%
+  mutate(id = as.character(id)) %>%
+  dplyr::filter(location == "waiting room",
+                id %in% unique(waiting_room_df$id)) 
+quantaConc_av_wr <- left_join(waiting_room_df, quantaConc_av_wr, by = c("id"))
+n_pa_ids <- max(as.numeric(passage_df$id))
+quantaConc_av_pa <- quantaConc_av %>%
   group_by(daytime) %>%
-  head(nrow(waiting_room_df)) %>%
-  mutate(id = as.character(id)) %>%
-  ungroup() %>%
-  left_join(waiting_room_df, by = c("id")) %>%
-  rename(x = long, y = lat) %>%
-  rotate_xy(a_ac_grid)
-quantaDF_av_pa <- quantaDF_av %>%
-  group_by(daytime) %>%
-  tail(nrow(passage_df)) %>%
+  arrange(id) %>%
+  slice((n()-n_pa_ids+1):n()) %>%
   mutate(id = as.character(1:n())) %>%
-  ungroup() %>%
-  left_join(passage_df, by = c("id")) %>%
-  rename(x = long, y = lat) %>%
-  rotate_xy(a_ac_grid)
-quantaDF_av_tb <- quantaDF_av %>%
+  ungroup() 
+quantaConc_av_pa <- left_join(passage_df, quantaConc_av_pa, by = c("id"))
+quantaConc_av_tb <- quantaConc_av %>%
   filter(location == "tb room") %>%
@@
   mutate(id = as.character(1:n())) %>%
-  ungroup() %>%
-  left_join(tb_room_df, by = c("id")) %>%
-  rename(x = long, y = lat) %>%
-  rotate_xy(a_ac_grid)
+  ungroup() 
+quantaConc_av_tb <- left_join(tb_room_df, quantaConc_av_tb, by = c("id"))
 
-qc_pl <- ggplot(mapping = aes(x = x, y = y, group = group, fill = N)) +
-  geom_polygon(data = quantaDF_av_wr) +
-  geom_polygon(data = quantaDF_av_pa) +
-  geom_polygon(data = quantaDF_av_tb) +
-  coord_fixed() +
-  #scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[c(1,3,5,7,9)], breaks = seq(1,6,1)*1e-6, limits = c(0,100), na.value = NA) +
-  facet_wrap(~daytime, ncol = 1)
+entrance_lab <- textGrob(label = "Entrance", x = 0.1, y = 0.165, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+wr_lab <- textGrob(label = "Waiting room", x = 0.05, y = 0.45, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+tb_lab <- textGrob(label = "TB room", x = 0.75, y = 0.4, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(size = text_descr / cm(1)), just = c("left"), rot = 90)
+
+qc_pl <- ggplot(mapping = aes(x = long, y = lat, group = group, fill = N)) +
+  geom_polygon(data = quantaConc_av_wr) +
+  geom_polygon(data = quantaConc_av_pa) +
+  geom_polygon(data = quantaConc_av_tb) +
+  ggpubr::background_image(clinic_img) +
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log1p", labels = function(x) x * 1000, breaks = c(.1, .5, 1, .15, 2)/1000) +
+  scale_y_continuous(expand = expansion(add=c(2500, 3250))) +
+  scale_x_continuous(expand = expansion(add=c(400, 500))) +
+  facet_wrap(~daytime, ncol = 1) +
+  annotation_custom(entrance_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  labs(fill = expression("Density (10"^-3*"quanta/m"^3*")")) +
+  theme_classic() +
+  theme(text = element_text(size = text_descr),
+        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
+        legend.position = c(0.825,0.05), legend.direction = "horizontal", legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
+        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank()) +
+  guides(fill = guide_legend(title.position = "top"))
 qc_pl
-qc_pl <- spatial_ggpl(qc_pl)
-
-qc_pl
-save_plot(qc_pl, pdf_file = "../results/modeling/quanta-concentration.pdf", w = aw, h = 2*ah)
+ggsave(plot = qc_pl, filename = "../doc/IWHOD/presentation/figures/quanta-concentration.png", width = width_descr * 1.5, height = height_descr * 2)
 ```

### 2023-03-20 analysis/simSpatiotempModel.R

@@ -3,7 +3,6 @@
 library(tidyverse)
 options(dplyr.summarise.inform = FALSE)
 library(parallel)
-library(foreach)
 
 source("utils/trans_risk.R")
 source("utils/spatial.R")
@@ -13,7 +12,8 @@ source("utils/spatial.R")
 args <- commandArgs(trailingOnly = TRUE)
 clinics <- ifelse(args[1]=="Both", c("Masi", "Ocean"), args[1])
 no_sim <- as.numeric(args[2])
-no_cores <- as.numeric(args[3])
+start_sim <- as.numeric(args[3])
+no_cores <- as.numeric(args[4])
 
 
 #' Prepare data for spatiotemporal model
@@ -50,16 +50,17 @@ prep_data <- function(
   set.seed(seed)
   
   # patient movement data
-  patMov <- readRDS(paste("data-clean",clinic,"patient-tracking-data",sel_date,"patient-id-data_matched-filtered.rds",sep="/")) %>%
+  patMov <- readRDS(paste("data-clean", clinic, "combined-data", sel_date, "tracking-linked-clinical-data.rds", sep = "/")) %>%
     mutate(location = ifelse(is_waitingroom|is_passage, "waiting room", ifelse(is_tbroom, "tb room", NA)),
            hhmm = format(time, format = "%H:%M")) %>%
     filter(!is.na(location)) 
   patIDs <- unique(patMov$patient_id)
   
+  # time period
+  time_period <- seq(lubridate::floor_date(min(patMov$time), unit = "minutes"), lubridate::floor_date(max(patMov$time), unit = "minutes"), by = "1 min")
+  
   # CO2 data
-  co2 <- readRDS(paste("data-clean", clinic, "environmental-data", "co2-temp-humid.rds", sep = "/")) %>%
-    filter(date == sel_date)
-  time_period <- seq(min(co2$date_time), max(co2$date_time), by = "1 min")
+  co2 <- readRDS(paste("data-clean", clinic, "combined-data", sel_date, "environmental-data.rds", sep = "/")) 
   
   # room data
   cellSize <- sqrt(cellVolume * 1e9 / 3e3) 
@@ -106,9 +107,8 @@ prep_data <- function(
   patIDS_TBmasked <- unique(patMOV_TBmasked$patient_id)
   
   # unmasked TB patients 
-  nTBmasked <- readRDS(paste0("data-clean/", clinic, "/clinical-data/clinical-data.rds")) %>%
[…]

### 2023-03-20 preprocessing/match_clinical-tracking.R

@@ -15,12 +15,8 @@ library(tidyverse)
 
 match_clinical_tracking <- function(clinic, sel_date, min_time_in_cl = 5 * 60, min_time_at_re = 5, min_duration_after_re = 5 * 60, max_tolerance = 15 * 60) {
   
-  # directory of patient tracking data
-  save_str <- paste0("data-clean/", clinic, "/patient-tracking-data/", sel_date, "/")
-  
   # clinical data
-  clinical <- readRDS(paste0("data-clean/", clinic, "/clinical-data/clinical-data.rds")) %>%
-    filter(date == as.Date(sel_date)) %>%
+  clinical <- readRDS(paste0("data-clean/", clinic, "/combined-data/", sel_date, "/clinical-data.rds")) %>%
     group_by(patient_id) %>%
     slice(1) %>%
     ungroup() %>%
@@ -28,7 +24,7 @@ match_clinical_tracking <- function(clinic, sel_date, min_time_in_cl = 5 * 60, m
            clinical_record_time = date_time)
   
   # tracking data
-  tracking <- readRDS(paste0(save_str, "/patient-id-data.rds")) %>%
+  tracking <- readRDS(paste0("data-clean/", clinic, "/patient-tracking-data/", sel_date, "/", "/patient-id-data.rds")) %>%
     group_by(patient_id) %>%
     filter(as.numeric(difftime(last(time), first(time), units = "secs")) > min_time_in_cl) %>%
     ungroup()
@@ -69,24 +65,27 @@ match_clinical_tracking <- function(clinic, sel_date, min_time_in_cl = 5 * 60, m
     scale_y_continuous(expand = c(0, 0)) +
     theme_bw() +
     theme(text = element_text(size = 8))
-  save_plot(report_pl, pdf_file = paste0(save_str, "clinical-tracking-matching-report.pdf"), w = 16, h = 8)
+  ggsave(plot = report_pl, filename = paste0("data-clean/", clinic, "/patient-tracking-data/", sel_date, "/", "clinical-tracking-matching-report.pdf"), width = 16 / cm(1), height = 8 / cm(1))
   
   # append tracking data
-  tracking <- tracking %>%
+  tracking_linked_clinical <- tracking %>%
     left_join(merged %>% dplyr::select(patient_id, clinical_patient_id, tb_suspect), by = "patient_id") 
[…]

### 2023-03-20 preprocessing/prep_clinical-data.R

@@ -18,7 +18,7 @@ df_pat_time_masi <- read.csv("data-raw/Masi/clinical-data/patient-time.csv") %>%
   filter(!is.na(time))
 
 df_tb_suspects <- read.csv("data-raw/Masi/clinical-data/tb-suspects.csv") %>%
-  select(patient_id, result_txt) %>%
+  dplyr::select(patient_id, result_txt) %>%
   rename(result = result_txt)
 
 df <- left_join(df_pat_time_masi, df_tb_suspects) %>%
@@ -26,6 +26,17 @@ df <- left_join(df_pat_time_masi, df_tb_suspects) %>%
   mutate(tb_suspect = ifelse(is.na(result), "no", "yes"),
          time = format(date_time, "%H:%M:%S"),
          date_time = as.POSIXct(as.character(date_time))) %>%
-  select(patient_id, date, time, date_time, tb_suspect)
+  dplyr::select(patient_id, date, time, date_time, tb_suspect)
 
 saveRDS(df, "data-clean/Masi/clinical-data/clinical-data.rds")
+
+selected_dates <- unique(df$date)
+save_dir <- paste0("data-clean/Masi/combined-data/", selected_dates)
+for (i in 1:length(selected_dates)) {
+  if (!dir.exists(save_dir[i])) {
+    dir.create(save_dir[i])
+  }
+  df_sub <- df %>%
+    dplyr::filter(date == as.Date(selected_dates[i])) 
+  saveRDS(df_sub, paste0(save_dir[i], "/clinical-data.rds"))
+}

### 2023-03-20 preprocessing/prep_environmental-data.R

@@ -1,46 +1,137 @@
 #### Libraries ####
 
 library(tidyverse)
-library(haven)
-
 
 #### Masi ####
 
 # load data
-masi <- read_dta("data-raw/Masi/environmental-data/co2_data.dta") %>%
-  mutate(date = Date,
-         time = format(new_time, "%H:%M:%S"),
-         date_time = as.POSIXct(paste(Date, time)),
-         location = Monitor) %>%
-  rename(temperature = Temperature,
+masi <- readxl::read_xlsx("data-raw/Masi/environmental-data/co2_data follow-up_2.xlsx", 
+                          sheet = "all_in_one") %>%
+  rename(date = Date, 
+         time = Time,
          humidity = Humidity,
-         co2 = CO2) %>%
-  select(date, time, date_time, location, co2, temperature, humidity) 
+         temperature = Temperature,
+         co2 = CO2,
+         location = Monitor) %>%
+  mutate(exact_time = format(time, format = "%H:%M:%S"),
+         time = format(time, format = "%H:%M"),
+         exact_date_time = as.POSIXct(paste(date, exact_time), format = "%Y-%m-%d %H:%M:%S"),
+         date_time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M"),
+         date = as.Date(date),
+         location = ifelse(location == "TB room", "tb room", location),
+         across(c(temperature, humidity, co2), as.numeric)) %>%
+  dplyr::select(date, exact_date_time, date_time, location, co2, temperature, humidity)
+
+# check
+masi %>%
+  ggplot(aes(x = date_time, y = co2, color = location)) +
+  geom_line() +
+  facet_wrap(~ date, scales = "free_x") +
+  scale_x_datetime(date_labels = "%H:%M")
+masi %>%
+  group_by(location, date, date_time) %>%
+  summarize(n = n()) %>%
+  ggplot(aes(x = n)) + geom_histogram()
+# masi %>%
+#   group_by(location, date, date_time) %>%
+#   mutate(n = n()) %>%
+#   filter(n > 1) %>%
+#   arrange(date_time) %>%
+#   View()
+
+#' TODO: check reason for double entries
+#' for now: use the first timestamp
+
+masi_double_entries <- masi %>%
+  group_by(location, date, date_time) %>%
+  arrange(exact_date_time) %>%
+  mutate(id = 1:n()) %>%
[…]

### 2023-03-20 utils/plotting.R

@@ -7,6 +7,12 @@ library(wesanderson)
 # round format
 round_k <- function(x, k = 0) format(round(x, k), nsmall = k)
 
+# add white space
+add_ws <- function(x, n = 2) {
+  ifelse(nchar(x) < n, paste0(" ",x), x)
+}
+
+
 # ggplot theme for Nature style
 text_size = 8
 update_geom_defaults("text", list(size = text_size))

### 2023-03-20 utils/trans_risk.r

@@ -26,7 +26,7 @@ compute_AER <- function(G, Co, C, n, V) {
   # excess CO2 in ppm
   Ce <- C - Co
   # TODO: is there are more elegant way to ensure that the denominator is not 0?
-  Ce <- ifelse(Ce < 0, 1, Ce) 
+  Ce <- ifelse(Ce < 1, 1, Ce) 
   # excess CO2 in l/m3 (1l/m3 = 1,000ppm)
   Ce <- Ce / 1000
   # ventilation rate (in l/s per person)

### 2023-03-20 preprocessing/linking-patient-data-app/app.R
Shiny app for tracking

@@ -1,0 +1,251 @@
+#### Libraries ###
+library(shiny)
+library(tidyverse)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(gridExtra)
+source("../../utils/spatial.R")
+
+#### Data ####
+# tracking data
+# tracking_dat <- readRDS("data-raw/Masi/xovis/2021-10-25.rds")
+# test <- slice(tracking_dat, 1:1000)
+# saveRDS(test, file = "test.rds")
+
+# building
+building <- terra::vect(paste0("../../data-raw/", "Masi", "/building/clinic-vector.gpkg"))
+building_sf <- sf::st_as_sf(building, crs = NA)
+st_crs(building_sf) <- NA
+building_df <- fortify(building_sf)
+building_df$col_group <- c(NA, "Clinic", "Clinic", "Clinic", NA, "Exit", "Seating area", "Entrance reception",
+                           "Seating area", "Seating area", "Seating area", "Entrance treatment room", "Entrance treatment room",
+                           "Exit", "Exit", NA)
+
+# building plot
+building_pl <- ggplot(building_df) +
+  geom_sf(mapping = aes(fill = col_group), linewidth = .5, alpha = 0.2) +
+  annotate("text", x = -13500, y = -1500, label = "Waiting room", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = -14000, y = -2000, label = "Seating area", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = -5000, y = -1000, label = "Seating area", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = 200, y = -1000, label = "Seating area", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = 2700, y = -1000, label = "Seating area", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = -11500, y = -6500, label = "Entrance / Main exit", hjust = 0, size = 10 / cm(1), vjust = 0) +
+  annotate("text", x = -6000, y = -4700, label = "Reception", hjust = 0, size = 10 / cm(1), angle = 90, vjust = 0) +
+  annotate("text", x = -4000, y = -7000, label = "Reception\nroom", hjust = 0, size = 10 / cm(1), vjust = 0) +
[…]

### 2023-03-29 analysis/spatiotemporal-transmission-risk.Rmd
minor changes to spatiotemporal model

@@
 tb_room_df <- fortify(tb_room)
-
-# test
-# waiting_room_df$value <- rnorm(nrow(waiting_room_df))
-# passage_df$value <- rnorm(nrow(passage_df))
-# tb_room_df$value <- rnorm(nrow(tb_room_df))
-# 
-# ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
-#   geom_polygon(data = waiting_room_df) +
-#   geom_polygon(data = passage_df) +
-#   geom_polygon(data = tb_room_df) +
-#   ggpubr::background_image(clinic_img) +
-#   annotate("text", x = -13200, y = -5300, label = "Entrance", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
-#   annotate("text", x = -13400, y = -2300, label = "Waiting room", size = 8 / cm(1), fontface = 2, hjust=0,vjust=0) +
-#   annotate("text", x = -5000, y = -4100, label = "Registration", size = 8 / cm(1), fontface = 2, angle = 90, hjust=0,vjust=0) +
-#   annotate("text", x = 2600, y = -2700, label = "TB room", size = 8 / cm(1), fontface = 2) +
-#   scale_y_continuous(expand = expansion(add=c(2500, 4000))) +
-#   scale_x_continuous(expand = expansion(add=c(400, 500))) +
-#   labs(fill = "Density") +
-#   theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
-#         legend.position = c(0.825,0.075), legend.direction = "horizontal", legend.title = element_blank(), #legend.title = element_text(vjust = .75),
-#         panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 3, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
-#         panel.grid.minor = element_blank())
 ```
@@
 ggsave(plot = qc_pl, filename = "../doc/IWHOD/presentation/figures/quanta-concentration.png", width = width_descr * 1.5, height = height_descr * 2)
+
+# for abstract
+highRiskPat_pl_abs <- highRiskPat_pl +
+  labs(title = "1a | Estimated no. of attendees with high-risk of MtB infection",
+       subtitle = "Number (mean and 95%-limits) of high-risk clinical attendees (defined as P>1%)") +
+  theme(plot.title = element_text(face = 2),
+        plot.title.position = "plot")
+
+summary_tab_abs <- summary_tab +
+  labs(title = " ",
+       subtitle = " ")
+
+qc_pl_abs <- qc_pl +
+  labs(title = "1b | Estimated quanta concentration in the clinic by daytime",
+        subtitle = "Average concentration in the morning (7:30 to 12am) and afternoon (12am to 4pm)") +
+  theme(plot.title = element_text(face = 2))
+
+comb_abs_pl <- grid.arrange(grid.arrange(highRiskPat_pl_abs, summary_tab_abs, ncol = 2, widths = 16 * c(0.7,.3)), qc_pl_abs, ncol = 1, heights = 25 / cm(1) * c(0.35, 0.65))
+ggsave(plot = comb_abs_pl, filename = "../doc/Union/abstract/figure.png", width = 16 / cm(1), height = 25 / cm(1))
 ```

### 2023-03-29 preprocessing/data_check_xovis.R
automatic pre matching and improved manual matching shiny app

@@ -5,7 +5,7 @@ file_names <- basename(files)
 dates <- gsub("_exported_data[[:punct:]]csv", "", file_names)
 dates <- gsub("_", "-", dates)
 
-i = 21
+i = 1
 print(dates[i])
 masi <- read.csv(files[i], header = T) %>% 
   set_names(c("obs_id", "time", "x", "y", "height")) %>%
@@ -21,11 +21,29 @@ summary(dt$dt)
 sum(dt$dt>1)
 sum(dt$dt<1)
 plot(unique(sort(dt$time)))
-bad_ones <- filter(dt, dt > 1 | dt < 1) 
+bad_ones <- filter(dt, dt > 2 | dt < 0) 
 bad_ones %>% View()
 bad_ids <- unique(bad_ones$obs_id)
 
 saveRDS(masi, paste0("data-raw/Masi/xovis/", dates[i], ".rds"))
 saveRDS(masi %>% filter(!obs_id %in% bad_ids), paste0("data-raw/Masi/xovis/", dates[i], ".rds"))
 
-  
+
+# automated
+file_dates <- c(paste0("2021-10-", c("13", "14", "15", "18", "19", "20", "22", "25")), 
+                paste0("2021-11-", c("02", "04", "05", "10", "12", "18")))
+files_aut <- paste0("data-raw/Masi/xovis/Study_Data_Masi_2021/export/", gsub("-", "_", file_dates), "_exported_data.csv")
+
+for (i in 1:length(files_aut)) {
+  masi <- read.csv(files_aut[i], header = T) %>% 
+    set_names(c("obs_id", "time", "x", "y", "height")) %>%
+    mutate(time = as.POSIXct(paste(file_dates[i], time)))
+  dt <- masi %>%
+    group_by(obs_id) %>%
+    mutate(dt = as.numeric(difftime(time, lag(time), units = "secs"))) %>%
+    ungroup() %>%
+    na.omit()
+  bad_ones <- filter(dt, dt > 2 | dt < 0) 
+  bad_ids <- unique(bad_ones$obs_id)
+  saveRDS(masi %>% filter(!obs_id %in% bad_ids), paste0("data-raw/Masi/xovis/", file_dates[i], ".rds"))
+}

### 2023-03-29 preprocessing/linking-patient-data-app/app.R

@@ -1,5 +1,6 @@
 #### Libraries ###
 library(shiny)
+library(shinyalert)
 library(tidyverse)
 library(reshape2)
 library(lubridate)
@@ -9,6 +10,7 @@ library(raster)
 library(terra)
 library(gridExtra)
 source("../../utils/spatial.R")
+options(shiny.maxRequestSize=10*1024^2)
 
 #### Data ####
 # tracking data
@@ -44,25 +46,34 @@ building_pl <- ggplot(building_df) +
   annotate("text", x = -12000, y = 900, label = "Emergency exit", hjust = 0, size = 10 / cm(1), angle = 90, vjust = 0) +
   annotate("text", x = -7000, y = 900, label = "Exit", hjust = 0, size = 10 / cm(1), angle = 90, vjust = 0) +
   scale_fill_discrete(na.translate = F) +
+  scale_x_continuous(breaks = seq(-20000, 20000, 1000)) +
+  scale_y_continuous(breaks = seq(-20000, 20000, 1000)) +
   guides(fill = "none")
 building_pl
 
 #### Functions ####
 
+standing_height <- function(x, min_height = 1500) {
+  quantile(x[x>1500], .95)
+}
+
 filter_oids <- function(df, oid, max_timediff, max_distance) {
   
   # current id
   df_i <- df %>%
-    filter(obs_id == oid) %>%
+    filter(patient_id == oid) %>%
     slice(n())
   
   # other potential ids
   df_other <- df %>%
-    filter(time > df_i$time[1])
+    filter(is.na(tracking_end)) %>%
+    group_by(patient_id) %>%
+    filter(first(time) > df_i$time[1]) %>%
+    ungroup()
   
   # other ids filtered for maximum timediff and distance
   df_other_feat <- df_other %>%
-    group_by(obs_id) %>%
+    group_by(patient_id) %>%
     slice(1) %>%
     ungroup() %>%
     merge(df_i, suffixes = c("_other", "_i"), by = NULL) %>%
@@ -73,21 +84,27 @@ filter_oids <- function(df, oid, max_timediff, max_distance) {
   
   # paths of other ids
   df_other <- df_other %>%
-    filter(obs_id %in% df_other_feat$obs_id_other)
+    filter(patient_id %in% df_other_feat$obs_id_other) %>%
+    left_join(df_other_feat %>% 
+                rename(patient_id = patient_id_other) %>%
[…]

### 2023-03-29 preprocessing/prep_automatic_pre_matching.R

@@ -1,0 +1,204 @@
+#### Libraries ####
+
+library(tidyverse)
+options(dplyr.summarise.inform = FALSE)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(parallel)
+library(foreach)
+source("utils/spatial.r")
+
+
+#' Algorith to match observation IDs
+#' 
+#' @param location string
+#' @param date date %Y-%m-%d
+#' @param max_time maximum time (in seconds) to look ahead for matches
+#' @param min_standing_height minimum standing height
+#' @param max_mov_time maximum time (in seconds) to look ahead for people moving
+#' @param max_sit_time maximum time (in seconds) to look ahead for people sitting
+#' @param max_mov_distance maximum distance someone can make while moving
+#' @param max_sit_distance maximum distance someone can make while sitting
+#' @param max_mov_heightdiff maximum height difference when moving
+#' @param min_duration minimum duration of a track, otherwise filtered
+#' @param min_total_distance minimum total distance of a track, otherwise filtered
+
+match_xovis <- function(location,
+                        date, 
+                        max_time = 30 * 60, 
+                        min_standing_height = 1500, 
+                        max_mov_time = 10, 
+                        max_mov_distance = 3,
+                        max_sit_time = 30,
+                        max_sit_distance =  0.5,
+                        max_mov_heightdiff = 100,
+                        min_duration = 5,
+                        min_total_distance = 1
+) {
+  
+  #### Data ####
+  
+  # xovis
+  file <- paste0("data-raw/", location, "/xovis/", date, ".rds")
+  save_dir <- paste0("data-raw/", location, "/patient-tracking-data/", date, ".rds")
+  masi <- readRDS(file) 
+  
+  # pre filter
+  masi <- masi %>%
+    group_by(obs_id) %>%
+    mutate(duration = as.numeric(difftime(last(time), first(time), units = "secs")),
+           distance = convert_dist( euclidean(first(x), last(x), first(y), last(y)) ),
+           first_entered = first(is_entrance),
+           has_exited = any(is_exit)) %>%
+    ungroup() %>%
+    filter(!(duration <= min_duration & distance <= min_total_distance & !first_entered & !has_exited) ) %>%
+    dplyr::select(-distance, -duration)
+  
[…]

### 2023-03-29 preprocessing/prep_xovis_matching.R

@@ -1,597 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-options(dplyr.summarise.inform = FALSE)
-library(reshape2)
-library(lubridate)
-library(sf)
-library(sfheaders)
-library(raster)
-library(terra)
-library(gridExtra)
-library(VennDiagram)
-library(parallel)
-library(foreach)
-source("utils/spatial.r")
-source("utils/plotting.r")
-
-
-#' Algorith to match observation IDs
-#' 
-#' @param location string
-#' @param rooms fortified dataframe of spatial polygons of the rooms in the building
-#' @param date date %Y-%m-%d
-#' @param max_time maximum time (in seconds) to look ahead for matches
-#' @param min_standing_height minimum standing height
-#' @param max_mov_time maximum time (in seconds) to look ahead for people moving
-#' @param max_sit_time maximum time (in seconds) to look ahead for people sitting
-#' @param max_tbroom_time maximum time (in seconds) to look ahead for someone being in the TB room
-#' @param max_mov_distance maximum distance someone can make while moving
-#' @param max_sit_distance maximum distance someone can make while sitting
-#' @param max_heightdiff maximum height difference for a person not changing her posture
-#' @param min_duration minimum time (in minutes) to stay in the clinic to be considered
-
-match_xovis <- function(location,
-                        rooms,
-                        date, 
-                        max_time = 30 * 60, 
-                        min_standing_height = 1400, 
-                        max_mov_time = c(3, 5, 5, 5, 10, 10, 15, 15, 30, 60), 
-                        max_sit_time = c(5, 10, 20, 30, 45, 60, 90, 150, 300, 600),
-                        max_tbroom_time = 300,
-                        max_mov_distance = c(0.5, 0.5, 1, 2, 2, 3, 3, 3.5, 4, 4),
-                        max_sit_distance = c(0.25, 0.25, 0.5, 0.5, 0.5), 
-                        max_heightdiff = c(50, 50, 75, 100, 100, 125, 125, 150, 175, 200), 
-                        min_duration = 5 * 60
-                        ) {
-  
-  #### Data ####
-  
-  # xovis
-  file <- paste0("data-raw/", location, "/xovis/", date, ".rds")
-  save_dir <- paste0("data-clean/", location, "/patient-tracking-data/", date, "/")
-  if (!dir.exists(save_dir)) {
-    dir.create(save_dir)
-  }
-  masi <- readRDS(file) 
-  
[…]

### 2023-03-29 preprocessing/prep_xovis_room.R

@@ -8,6 +8,7 @@ library(sfheaders)
 library(raster)
 library(terra)
 library(VennDiagram)
+library(parallel)
 source("utils/spatial.r")
 
 # roomplan
@@ -30,8 +31,8 @@ for (f in files) {
     rotate_xy(a = -9)
   
   # add room info
-  df$point <- map2(df$x, df$y, function(xi,yi) sfheaders::sf_point(c(xi,yi)))
-  df$poly = map(df$point, sf::st_within, y = clinic_sf)
+  df$point <- mcmapply(function(xi,yi) sfheaders::sf_point(c(xi,yi)), df$x, df$y, mc.cores = 6)
+  df$poly = mclapply(df$point, sf::st_within, y = clinic_sf, mc.cores = 6)
   df$is_waitingroom <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 2), T, F))
   df$is_tbroom <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 3), T, F))
   df$is_passage <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 4), T, F))
@@ -43,6 +44,9 @@ for (f in files) {
   df$is_in_room_right <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) == 13), T, F) ) 
   df$is_exit <- map_lgl(df$poly, function(p) ifelse(any(unlist(p) %in% c(6, 14:15)), T, F)) # entrance and corridors are considered as exits
  
+  df <- df %>% 
+    dplyr::select(-point, -poly)
+  
   saveRDS(df, file = paste0("data-raw/Masi/xovis/", file_name)) 
 }
 

### 2023-04-20 analysis/spatiotemporal-transmission-risk.Rmd
preparing and testing spatiotemporal model with diffusion

@@
 height_descr <- 7.5 / cm(1)
-text_descr <- 12
+text_descr <- 10
 ```
@@
 # for abstract
-highRiskPat_pl_abs <- highRiskPat_pl +
-  labs(title = "1a | Estimated no. of attendees with high-risk of MtB infection",
-       subtitle = "Number (mean and 95%-limits) of high-risk clinical attendees (defined as P>1%)") +
-  theme(plot.title = element_text(face = 2),
-        plot.title.position = "plot")
-
-summary_tab_abs <- summary_tab +
-  labs(title = " ",
-       subtitle = " ")
+# highRiskPat_pl_abs <- highRiskPat_pl +
+#   labs(title = "1a | Estimated no. of attendees with high-risk of MtB infection",
+#        subtitle = "Number (mean and 95%-limits) of high-risk clinical attendees (defined as P>1%)") +
+#   theme(plot.title = element_text(face = 2),
+#         plot.title.position = "plot")
+# 
+# summary_tab_abs <- summary_tab +
+#   labs(title = " ",
+#        subtitle = " ")
 
@@
   theme(plot.title = element_text(face = 2))
+```
 
-comb_abs_pl <- grid.arrange(grid.arrange(highRiskPat_pl_abs, summary_tab_abs, ncol = 2, widths = 16 * c(0.7,.3)), qc_pl_abs, ncol = 1, heights = 25 / cm(1) * c(0.35, 0.65))
-ggsave(plot = comb_abs_pl, filename = "../doc/Union/abstract/figure.png", width = 16 / cm(1), height = 25 / cm(1))
+x### Track examples
+
+```{r}
+examples <- no_people %>%
+  group_by(patient_id) %>%
+  mutate(first_entered = first(is_entrance),
+         last_exited = last(is_entrance|is_exit),
+         duration = as.numeric(difftime(last(time), first(time), units = "mins"))) %>%
+  ungroup() %>%
+  filter(first_entered,
+         last_exited,
+         duration > 10) %>%
+  ungroup()
+
+n_distinct(examples$patient_id)
+
+examples_se <- examples %>%
+  group_by(patient_id) %>%
+  slice(1, n()) %>%
+  mutate(type = factor(c("Start", "End"), levels=c("Start", "End"))) %>%
+  ungroup()
+
+examples_pl <- examples %>%
+  ggplot(aes(x = x, y = y, group = patient_id)) +
+  geom_path(color = "darkred", alpha = .5) +
+  geom_point(data = examples_se, mapping = aes(shape = type)) +
+  ggpubr::background_image(clinic_img) +
+  scale_y_continuous(expand = expansion(add=c(2000, 3250))) +
+  scale_x_continuous(expand = expansion(add=c(800, 0))) +
+  scale_shape_manual(values = c(1,13)) +
+  annotation_custom(entrance_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
+  labs(title = "1a | Subset of patient movement data",
+       subtitle = "Examples of trackings from 40 patients in the clinic") + 
+  theme_classic() +
+  theme(text = element_text(size = text_descr),
+        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.3, "cm"),
+        legend.position = c(0.825,0.075), legend.direction = "horizontal",
+        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(), legend.title = element_blank()) +
+  theme(plot.title = element_text(face = 2))
+
+examples_pl
+
+comb_abs_pl <- grid.arrange(grid.arrange(examples_pl, ggplot() + theme_classic() + theme(axis.line = element_blank()), ncol = 1, heights = c(8.5,6.5)), 
+                            qc_pl_abs, ncol = 2, widths = 26.5 / cm(1) * c(0.475, 0.525))
+ggsave(plot = comb_abs_pl, filename = "../doc/Union/abstract/figure.png", width = 26.5 / cm(1), height = 15 / cm(1))
 ```

### 2023-04-20 preprocessing/prep_environmental-data.R

@@ -4,134 +4,162 @@ library(tidyverse)
 
 #### Masi ####
 
+masi <- do.call(rbind,
+              lapply(files, function(f) {
+                read.table(f, sep = "\t", header = F, skip = 33) %>%
+                  set_names(c("date", "time", "humidity", "temperature", "co2", "ext")) %>%
+                  dplyr::select(-ext) %>%
+                  mutate(file = basename(f))
+              })) %>%
+  mutate(date_created = as.Date(stringi::stri_extract(file, regex = "\\d{8}"), "%Y%m%d"),
+         date = as.Date(date, format = "%m/%d/%Y"),
+         sensor = as.integer(gsub("-", "", stringi::stri_extract(file, regex = "-\\d{1}"))),
+         date_time = parse_date_time(paste(date, time), '%Y-%m-%d %I:%M:%S %p'),
+         across(c(humidity, temperature, co2), as.numeric)) 
+
+masi %>% 
+  filter(sensor == 1) %>%
+  ggplot(aes(x = date_time, y = co2)) +
+  geom_line() +
+  facet_wrap(~ date, scales = "free_x") +
+  scale_x_datetime(date_labels = "%H:%M")
+
 # load data
-masi <- readxl::read_xlsx("data-raw/Masi/environmental-data/co2_data follow-up_2.xlsx", 
-                          sheet = "all_in_one") %>%
-  rename(date = Date, 
-         time = Time,
-         humidity = Humidity,
-         temperature = Temperature,
-         co2 = CO2,
-         location = Monitor) %>%
-  mutate(exact_time = format(time, format = "%H:%M:%S"),
-         time = format(time, format = "%H:%M"),
-         exact_date_time = as.POSIXct(paste(date, exact_time), format = "%Y-%m-%d %H:%M:%S"),
-         date_time = as.POSIXct(paste(date, time), format = "%Y-%m-%d %H:%M"),
-         date = as.Date(date),
-         location = ifelse(location == "TB room", "tb room", location),
-         across(c(temperature, humidity, co2), as.numeric)) %>%
-  dplyr::select(date, exact_date_time, date_time, location, co2, temperature, humidity)
-
-# check
-masi %>%
-  ggplot(aes(x = date_time, y = co2, color = location)) +
-  geom_line() +
[…]

### 2023-04-20 tests/spatiotemporal-quanta-concentration-example.R

@@ -1,0 +1,125 @@
+#' Room:
+#' We assume a room with length = 10m, breadth = 6m, and height = 3m.
+#' We divide the room into a grid with cell length = 0.1m, i.e. 100 x 60 matrix.
+#' 
+#' Time:
+#' We run our simulation for 1h with a time step of 1s.
+#' 
+#' Quanta emission:
+#' There are two quanta emission sources:
+#' 1. One in the center in the first five minutes.
+#' 2. One near the top-right edge between minute 30 and 35.
+#' 
+#' The quanta emission rate is 4 quanta/hour.
+#' We assume an exponential decrease with distance to source of emission.
+#' 
+#' Quanta removal:
+#' Quanta is removed only through air exchange.
+#' We assume a constant air exchange rate of 3 air changes per hour.
+#' 
+#' Quanta diffusion:
+#' We assume radial diffusion at a constant rate of 0.05m2/min. 
+#' 
+
+
+#### Libraries ####
+library(tidyverse)
+source("utils/trans_risk.r")
+source("utils/spatial.R")
+
+
+#### Setup ####
+
+# room
+Lx <- 10
+Ly <- 6
+Lz <- 3
+V <- 10 * 6 * 3
+Vxy <- 1
+dx <- dy <- sqrt(Vxy/Lz)
+nx <- Lx / dx
+ny <- Ly / dy
+
+# initial quanta concentration
+C0 <- matrix(0, nrow = ny, ncol = nx)
+
+# time
+TT <- 60 * 60
+dt <- 1
+
+# quanta emission rate
+q <- 4 / 3600
+
+# weights to source
+W1 <- matrix(NA, nrow = ny, ncol = nx)
+for (i in 1:ny) {
+  for (j in 1:nx) {
+    W1[i,j] <- euclidean(floor(ny/2), i, floor(nx/2), j) * dx
+  }
+}
+W1 <- apply(W1, c(1,2), dIc, mu = 1)
+W1 <- W1 / sum(W1)
+
+W2 <- matrix(NA, nrow = ny, ncol = nx)
+for (i in 1:ny) {
+  for (j in 1:nx) {
+    W2[i,j] <- euclidean(ny, i, nx, j) * dx
+  }
+}
+
+# weights
+W2 <- apply(W2, c(1,2), dIc, mu = 1)
+W2 <- W2 / sum(W1)
+
+# quanta removal rate
+aer <- 3 / 3600
+
+# quanta diffusion constant
[…]

### 2023-04-20 utils/trans_risk.r

@@ -1,13 +1,44 @@
-#' Compute the quanta removal rate
+#' Compute quanta removal
 #' 
-#' 
-#' @param AER air exchange rate per hour
+#' @param C N x M matrix of quanta concentration per grid cell
+#' @param aer air exchange rate per hour
 #' @param k rate per hour for particle deposition on surfaces
 #' @param lambda viral inactivation rate per hour
+#' @param dt time step (in s)
 #' 
 
-compute_RR <- function(AER, k = 0, lambda = 0) {
-  return(AER + k + lambda)
+removal <- function(C, aer, k = 0, lambda = 0, dt = 1) {
+  
+  # quanta removal rate
+  rr <- aer + k + lambda
+  
+  # removal
+  qr <- function(x) x * exp(- rr * dt)
+  
+  # quanta concentration after removal
+  C_rq <- apply(C, c(1,2), qr)
+  
+  return(C_rq)
+}
+
+
+#' Compute quanta emission
+#' 
+#' @param C N x M matrix of quanta concentration per grid cell
+#' @param W N x M matrix with weights for quanta emission (in m)
+#' @param qer quanta emission rate (in quanta/s)
+#' @param dt time step (in s)
+#' 
+
+emission <- function(C, W, qer, dt = 1) {
+  
+  # emitted quanta 
+  eq <- qer * dt
+  
+  # quanta concentration after emission
+  C_eq <- C + W * eq
+  
+  return(C_eq)
 }
 
 
@@ -20,7 +51,7 @@ compute_RR <- function(AER, k = 0, lambda = 0) {
 #' @param n number of people in ventilated space (in 1/m3) 
 #' 
 
-compute_AER <- function(G, Co, C, n, V) {
+compute_aer <- function(G, Co, C, n, V) {
   # add 1 person for the person working behind the registry (and such that n is not 0)
   n <- n + 1
   # excess CO2 in ppm
@@ -32,41 +63,21 @@ compute_AER <- function(G, Co, C, n, V) {
   # ventilation rate (in l/s per person)
   Q <- G / Ce
   # air exchange rate (in s)
-  AER <- Q * n / V
+  aer <- Q * n / V
   # air exchange rate (in h)
-  AER <- 3600 * AER
-  return( AER )
+  aer <- 3600 * aer
+  return( aer )
 }
-
-
-#' Compute quanta concentration (quanta / m3)
-#' 
-#' 
-#' @param E quanta emission (quanta emission rate times the number of infectious)
[…]

### 2023-07-18 doc/paper/manuscript.tex
Spatiotemporal model description

\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
\documentclass[fleqn,11pt]{wlscirep}

\usepackage{import}
\usepackage{main}

\renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}

% word count
% \newcommand{\maincount}[1]{%
%   \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
%   \input{#1-words.sum}%
% }

% \newcommand{\abstractcount}[1]{%
%   \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
%   \input{#1-words.sum}%
% }

\begin{document}

\doublespacing

\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mtb} transmission in a South African primary care clinic}}
% author list
\author[1$\ddag$]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[1]{Kathrin Zürcher}
\author[2]{Robin Wood}
\author[1]{Matthias Egger}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}

\affil[*]{Corresponding author: lukas.fenner@unibe.ch }

\vspace{1em}

% \begin{information}\normalfont
% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners

% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections

% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)

% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references

% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.

% \vspace{1em}

% \noindent\textbf{Funding}

% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland. NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05. ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498). \medskip

% \noindent\textbf{Contributions}

% \noindent Conception and design: NB, LF. Epidemiological and environmental data collection: NB, PJ, TS, LF. Laboratory data collection: PB, LFu. Additional data collection: TH. Statistical analysis: NB, KZ. Genomic analysis: LB, LFu. Paper draft: NB, LF, ME. All authors reviewed and approved the final version of the manuscript.

% \par
% \end{information}

%TC:newcounter abst Words in abstract
%TC:envir abstract [] abst
\begin{abstract}\normalfont
\noindent\textbf{Background:} ... \medskip
\noindent\textbf{Methods and Findings:} ... 
\medskip %\vspace{-1.3em}
\noindent\textbf{Conclusions:} ...

\par
\end{abstract}

%TC:ignore

\flushbottom
\maketitle
\setcounter{page}{1}
\thispagestyle{fancy}

\vspace{2em}

%\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)

\vspace{0.5em}

\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne transmission, spatiotemporal modeling, Wells-Riley model
% maximum of 3-5 keywords
\newpage

\sloppy
\raggedbottom
%TC:endignore

\newpage

%TC:break main
\section{Introduction}

...

\newpage

\section{Methods}

...

\newpage

\section{Results}

...

\FloatBarrier

\newpage

\section{Discussion}

% summary

...

\newpage

%TC:ignore
\section*{Acknowledgements}
...

\bibliography{references.bib}
%TC:endignore

\end{document}

### 2023-07-18 doc/paper/supplements.tex
Spatiotemporal model description

\PassOptionsToPackage{top=3cm,left=3cm,right=3cm,bottom=3cm}{geometry}
\documentclass[fleqn,11pt]{wlscirep_supp}

\usepackage[]{minitoc}
\mtcsetdepth{secttoc}{3}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\mtcsettitle{secttoc}{}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
%\usepackage[top=3cm,left=3cm,right=3cm,bottom=3cm]{geometry}% by courtesy of Mico
\usepackage{lmodern}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{colortbl}
\usepackage{siunitx}
\sisetup{
  detect-all,
  detect-weight=true,
  detect-family=true,
  mode=text,
%   detect-inline-family=math,
  group-separator={,},
%   group-minimum-digits={3}			
}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{tabu}
\usepackage{authblk}
\usepackage{mathtools}
\usepackage{overpic}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{fit}
\usepackage{multirow}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{bm}
\usepackage{enumerate}
\usepackage[absolute,overlay%,showboxes
                        ]{textpos}
% \usepackage{caption}
\usepackage[font=small,labelfont=bf,justification=justified]{caption}
\usepackage{subcaption}
\usepackage{xspace}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{placeins}
\usepackage{makecell, booktabs}
\usepackage{eqparbox}
\usepackage{rotating}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{setspace}
%\usepackage{comment}
\usepackage[resetlabels,labeled]{multibib}
\newcites{Supp}{References}
\usepackage[sort&compress]{cleveref}
\Crefname{appendix}{Supplement}{Supplements}

%\usepackage{mathabx}

% Table formatting packages
\usepackage{dcolumn} % align decimal points in tables
\newcolumntype{d}[1]{D{.}{.}{#1}}
\usepackage{booktabs} 
\usepackage[flushleft]{threeparttable}
\usepackage{siunitx} % align on decimal point in tables
\usepackage{lineno}
\usepackage{etoolbox}

\usepackage{lscape}
\usepackage{longtable}
\usepackage{arydshln}

% math
\usepackage{amsmath,amsfonts,amssymb}
% additional math symbols
\DeclareFontFamily{U}{mathb}{}
\DeclareFontShape{U}{mathb}{m}{n}{
  <-5.5> mathb5
  <5.5-6.5> mathb6
  <6.5-7.5> mathb7
  <7.5-8.5> mathb8
  <8.5-9.5> mathb9
  <9.5-11.5> mathb10
  <11.5-> mathb12
}{}
\DeclareSymbolFont{mathb}{U}{mathb}{m}{n}
\DeclareMathSymbol{\ulsh}{3}{mathb}{"E8}
\DeclareMathSymbol{\ursh}{3}{mathb}{"E9}
\DeclareMathSymbol{\dlsh}{3}{mathb}{"EA}
\DeclareMathSymbol{\drsh}{3}{mathb}{"EB}

%% Patch 'normal' math environments:
\newcommand*\linenomathpatch[1]{%
  \cspreto{#1}{\linenomath}%
  \cspreto{#1*}{\linenomath}%
  \csappto{end#1}{\endlinenomath}%
  \csappto{end#1*}{\endlinenomath}%
}

\linenomathpatch{equation}
\linenomathpatch{gather}
\linenomathpatch{multline}
\linenomathpatch{align}
\linenomathpatch{alignat}
\linenomathpatch{flalign}

\linenumbers

% PLOS formatting
\makeatletter %only needed in preamble
\renewcommand\Large{\@setfontsize\Large{18pt}{18}}
\renewcommand\large{\@setfontsize\large{16pt}{18}}
\makeatother

\addto\captionsenglish{\renewcommand{\figurename}{Figure}}

% \usepackage{xstring}
% \usepackage{etoolbox}
% \usepackage{caption}

% \captionsetup{labelfont=bf,tableposition=top}

% \makeatletter
% \newcommand\formatlabel[1]{%
%     \noexpandarg
%     \IfSubStr{#1}{.}{%
%       \StrBefore{#1}{.}[\firstcaption]%
%       \StrBehind{#1}{.}[\secondcaption]%
%       \textbf{\firstcaption.} \secondcaption}{%
%       #1}%
%       }

% \patchcmd{\@caption}{#3}{\formatlabel{#3}}
% \makeatother

\renewcommand*{\Affilfont}{\normalsize\normalfont}
\renewcommand*{\Authfont}{\normalfont}

% referencing of unnumbered materials and methods
\newcounter{methods}
\renewcommand{\themethods}{Materials and methods}

% Track changes
%\usepackage[markup=underlined]{changes}
\makeatletter
\@namedef{Changes@AuthorColor}{magenta}
\colorlet{Changes@Color}{magenta}
\makeatother

%=====================================================================% Declare

\DeclareSIUnit\eur{\officialeuro}
\DeclareSIUnit\M{M}
\DeclareSIUnit\k{k}

% Widebar symbol
% \DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
% \DeclareFontShape{U}{mathx}{m}{n}{<-> mathx10}{}
% \DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
% \DeclareMathAccent{\widebar}{0}{mathx}{"73}

%=====================================================================% New commands (Macros)

% def
\def\sym#1{\ifmmode^{#1}\else\(^{#1}\)\fi}
\definecolor{darkgreen}{rgb}{0.0, 0.5, 0.0}

% new command
\newcommand{\smallsim}{\smallsym{\mathrel}{\sim}}
\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
\newcommand{\specialcellc}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\newcommand\ie{i.\,e.\xspace}
\newcommand\eg{e.\,g.\xspace}
\newcommand{\dd}[1][]{\mathrm{d}#1}
\newcommand{\BK}[1]{{\color{orange}{BK: #1}}}
\newcommand{\figletter}[1]{{{\fontfamily{\sfdefault}\selectfont \textbf{#1}}}}
\newcommand\TODO[1]{{\color{red}#1}}  
\newcommand{\FIX}[1]{{\color{darkgreen}#1}}

% renewcommand
\renewcommand\theadfont{\bfseries}
\renewcommand\theadalign{lc}
\renewcommand\cellalign{tl}

\makeatletter

\newbox\@abstract%
\def\abstitle{\textbf{Abstract}}%
\renewenvironment{abstract}{
  \global\setbox\@abstract\vbox\bgroup%
   \noindent
}{%
   \egroup%
}%

\renewcommand*{\Affilfont}{\normalsize\normalfont}
\renewcommand*{\Authfont}{\normalfont}

\addto\captionsenglish{% Replace "english" with the language you use
  \renewcommand{\contentsname}{List of Texts}
}

\def\@maketitle{%
  \newpage
    {\raggedright\fontsize{18pt}{20pt}\selectfont \@title \par}%
    \vskip 0.5em%
    {\large
      \lineskip .5em%
      \begin{tabular}[t]{l}%
        \raggedright \normalsize\mdseries{\@author} %
      \end{tabular}\par}%
      \vskip 1em
%      \raggedright\Large\abstitle\par
%      \vskip 1em
%    {\unvbox\@abstract\par}%
    \par
  \vskip 0.5em
}

\makeatother

\renewcommand{\thesection}{Text \arabic{section}}
\usepackage{titlesec}
%\titleformat{\section}{\normalfont\Large\bfseries}{Text \thesection.~#1}{1em}{}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

\begin{document}
\doublespacing
\nolinenumbers

\newcommand{\supp}{SI Appendix}

\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
Spatiotemporal modeling of \emph{Mtb} transmission in a South African primary care clinic}}

% author list
\author[1$\ddag$]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[1]{Kathrin Zürcher}
\author[2]{Robin Wood}
\author[1]{Matthias Egger}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}

\affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }

\affil[$\ddag$]{These authors contributed equally to this work.}

%\begin{abstract}\normalfont
%The supplementary material contains (1)~the detailed method, (2)~the simulation-based study, (3)~further descriptives, and (4)~the results from the sensitivity analysis.
%\end{abstract}

\flushbottom
\maketitle
\thispagestyle{empty}

%\newpage

\sloppy
\raggedbottom

\newpage

\appendix

\tableofcontents

\listoffigures

\listoftables

%\listoffigures
%\listoftables

\newpage

\section{Spatiotemporal model}\label{sec:spattemp-model}

\subsection{The Wells-Riley model}

We extend the Wells-Riley model [Wells-Riley], which is an established epidemiological model to estimate the risk of infection. The model estimates the probability of airborne indoor transmission $P$ using the following equation: 
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of disease cases, $S$ is the number of susceptibles, $I$ is the number of infectious individuals in space, $p$ is the breathing rate per person, $q$ is the generation rate of infectious quanta, $t$ is the duration of exposure, and $Q$ is the outdoor air supply rate.

The unknown parameter $q$ is not the actual number of infectious particles in the air. It rather represents the dose of infectious particles that corresponds to a probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection [Rudnick and Milton]. It is an attempt to consider the stochastic behavior of airborne transmission, which depends on the environment, characteristics of the pathogen, and characteristics of the individual. For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person [Wang], increasing the probability of infection.

While considering the stochastic behaviour of airborne transmission, the Wells-Riley model makes two simplifying assumptions. First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and evenly in the indoor airspace. In other words, the exposure to infectious quanta is the same regardless of the locations of the infectious and susceptible individual. Second, the Wells-Riley model assumes steady-state conditions, which means that the quanta concentration and the outdoor air supply rate are constant over time. In other words, the risk of infection for each susceptible depends only on the duration of exposure (the time they spent in the indoor space).

Rudnick and Milton [Rudnick and Milton] modified the Wells-Riley model to relax the steady-state assumption. They use CO$_2$ as a biomarker for exhaled breath to compute the outdoor air supply rate, which is otherwise difficult to measure. Their modified equation is
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{\bar{f}Iqt}{n}\right),
\end{align}
where $f = \frac{CO^{\text{I}}-CO^{\text{O}}}{CO^{\text{A}}}$ is the fraction of indoor air that is exhaled breath, which is computed based on the CO$_2$ level in indoor air $CO^{\text{I}}$, outdoor air $CO^{\text{O}}$, and exhaled breath $CO^{\text{A}}$ (in parts per million [ppm]).

While the extension by Rudnick and Milton resolves the steady-state assumption, it still assumes a well-mixed airspace. Therefore, our aim is to extend the Wells-Riley model further to allow the risk of infection to vary both across time and airspace. We accomplish this by modeling the quanta concentration spatiotemporally. The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points. Formally, we estimate the cumulative risk of infection as 
\begin{equation}
    P = \sum_s \sum_t N_{s,t} \cdot p,
\end{equation}
where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), and $p$ is the breathing rate per person (in $m^3/s$). The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta: 
\begin{itemize}
    \item[\ref{sec:quanta-generation}] Quanta generation: Infectious individuals generate quanta at their location in the indoor space. 
    \item[\ref{sec:quanta-diffusion}] Quanta diffusion: Over time, the infectious quanta diffuse and we approach a well-mixed airspace. 
    \item[\ref{sec:quanta-removal}] Quanta removal: Infectious quanta is removed from the air through (a)~outdoor air exchange, (b)~viral inactivation, and (c)~gravitational settling. 
\end{itemize}
We describe each process in more detail in the following. While we apply our model to \emph{Mtb}, we note that it can be applied to airborne respiratory infections in general.

\subsection{Setup}

We divide the airspace in a 2-dimensional grid, irrespective of the height of the airspace. The cell denotes the unit airspace with $s = 1, \dots, S$, where $(x_s, y_s)$ are the x and y coordinate of cell $s$. In some cases, we may use the (x,y)-notation in place of the cell ID $s$. The quanta concentration will be the same within each grid cell $s$, thus the grid provides a discrete approximation to the spatially varying quanta concentration. The number of grid cells $S = n_x \cdot n_y$ corresponds to the size of the matrix and determines the quality of the approximation as well as the model run time. More grid cells (smaller grid cells) provide a more granular picture of the spatial quanta concentration, at the expense of higher computation time.

\subsection{Notation}

\begin{itemize}
    \item $N_{s,t}$: quanta concentration in airspace unit $s$ at time $t$
    \item $I_{s,t}$: number of infectious individuals in airspace unit $s$ at time $t$
    \item $q$: quanta generation rate (in quanta/s)
    \item $D$: diffusion constant (in m$^2$/s)
    \item $AER$: air change rate (in air changes/s)
    \item $V$: volume of the airspace (in m$^2$)
    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air
    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air
    \item $G$: average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$)
    \item $n_t$: number of individuals in airspace at time $t$
    \item $\lambda$: viral inactivation rate (in quanta/s)
\end{itemize}

\subsection{Quanta generation}\label{sec:quanta-generation}

The number of quanta generated by infectious individuals at time $t$ is 
\begin{align}\label{eq:generation}
    I_t \cdot q ~.
\end{align}
The initial spatial spread of the quanta over the grid is informed prior literature. The concentration of virus-laden aerosols is highest closest to the infectious individual [References], in line with findings showing that the risk of infection correlates with the distance to the infectious individual [References]. Although coughing and sneezing can transport aerosols further away from the individual, these activities are much less frequent than breathing. Therefore, prior work suggests that breathing may contribute more infectious particles to the airspace than coughing or sneezing [Dinkele]. Overall, it seems plausible that the majority of infectious aerosols are generated near the infectious individual and then disperse over time. In line with prior findings, we will assume that virus-laden aerosols reach a distance of 0.50m to 1.0m through exhalation [References]. Since we do not observe the directional flow of the exhaled breadth, we assume that the generated quanta is evenly and radially distributed around the individual.

\subsection{Quanta diffusion}\label{sec:quanta-diffusion}

Absent further knowledge on airflow, we assume that aerosols diffuse uniformly in the indoor air, over time approaching a well-mixed airspace. Initially, the quanta concentration will be highest at the location where it is generated and then disperse radially in (x,y)-direction towards locations where the concentration is lower. The diffusion equation is 
\begin{align}\label{eq:diffusion}
    \frac{\delta N_{s,t}}{\delta t} = D \Delta N_{s,t},
\end{align}
where $\Delta$ is the Laplace operator and $D$ is the diffusion constant (in m$^2$/s). We solve this equation with a standard second-order finite difference model using the Euler method with a 5-point stencil.

The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)
\begin{align}
    K = (0.52 \cdot AER + 8.61e-5) * V^(2/3),
\end{align}
where $V$ is the volume of the airspace (in m$^3$). We assume that $D \approx K$.

The $AER$ can be estimated from indoor CO$_2$ levels and room occupancy. Under steady-state conditions (CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
\begin{align}
    AER = \frac{6e4 \cdot n \cdot G}{V\cdot(CO^{\text{I}}-CO^{\text{O}})},
\end{align}
where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$), $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air (both in parts per million [ppm]) [Battermann]. In a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration. Therefore, we determine $AER$ using a transient mass balance model of the form [Battermann]
\begin{align}
    CO_{t+1}^{\text{I}} = \frac{6e4 \cdot n_t + G}{Q} \cdot \left(1 - \exp(-Q/V \Delta t)\right) + (CO_t^{\text{I}}-CO^{\text{O}}) \cdot \exp(-Q/V \Delta t) + CO^{\text{O}},
\end{align}
where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$. We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) [REF], a quasi-Newton method. The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error. We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$ [Battermann]. The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$

\subsection{Quanta removal}\label{sec:quanta-removal}

Infectious quanta may be removed from the air through outdoor air exchange, viral inactivation, or gravitational settling. We will ignore gravitational settling. Vuorinen et al. [Reference] notes that particles smaller than 20\,$\mu$m rapidly become airborne. This is far above the upper end of the particle size distribution of most respiratory infections, including \emph{Mtb}, which is carried in particles typically smaller than 4.7\,$\mu$m [Fennelly]. The removal rate is thus the same of the outdoor air exchange rate $AER_t$ and the viral inactivation rate $\lambda$, \ie 
\begin{align}\label{eq:removal}
    AER_t + \lambda ~.
\end{align}
Note that $\lambda$ depends on the pathogen and is generally difficult to estimate and likely also depends on environmental factors, particularly relative humidity [Reference].

Considering the generation (\Cref{eq:generation}), diffusion (\Cref{eq:diffusion}), and removal (\Cref{eq:removal}) of quanta together, the quanta concentration at time $t$ in the airspace is computed as
\begin{align}
    \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev. concn.}} + \underbrace{I_t \cdot q}_{\text{generation}})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda)\right)}_{\text{removal}} ~.
\end{align}

\subsection{Illustrative example}

We illustrate our modeling approach with an example. Consider a room with a volume $V$=150m$^3$ (length [L] = 10m, breadth [B] = 5m, height [H] = 3m). We divide the room into a grid of $S = 40 \cdot 20 = 800$ cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m. We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells (\ie initially quanta reach a maximum distance of 0.7m through exhalation corresponding to the diagonal length of two cells). We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(20\pm1,10\pm1)$. We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of also 1\,quanta/h and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$\,quanta/h.

\Cref{fig:toy-example} shows the quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am). The quanta is generated at the location of the infectious individual. The quanta concentration remains high around this location until the individual ceases to generate more quanta at 9am. The quanta diffuses relatively quickly and at 10min past 9am the airspace is well-mixed with quanta. Over time, the removal through outdoor air exchange and viral inactivation brings the concentration in the room back towards 0.

\begin{figure}
    \centering
    \includegraphics{}
    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10$\times$B=\5$\times$H=3m.}
    \label{fig:toy-example}
\end{figure}

\clearpage

\section{Case study}

We divide the clinical day into a morning (7am to 12am) and afternoon period (12am to 5pm) and estimate the $AER$ separately for each period, allowing it to vary by daytime.

Previous studies estimated $q$ for \emph{Mtb} or SARS-CoV-2 using the Wells-Riley model [REFs]. The range of estimates is large, suggesting that the infectiousness of \emph{Mtb} patients varies considerably. We model $q$ with a Student-t distribution with 1 degree of freedom $t_1(\mu = 2, \sigma = 2.5)$ that is truncated at 0 (lower) and 200 (upper). The distribution is shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.

\begin{figure}
    \centering
    \includegraphics{}
    \caption{Caption}
    \label{fig:quanta-distribution}
\end{figure}

A schematic view of the clinic is shown in \Cref{fig:clinic}. We divide the clinic into three airspaces: the waiting room (l x b x h), the corridor (l x b x h), and the TB room (l x b x h). Each room is divided into a grid of cubic cells with a fixed cell length and breath of 0.25m. The volume of the grid cells thus depends on the height of each airspace.

\begin{figure}
    \centering
    \includegraphics{}
    \caption{Caption}
    \label{fig:clinic}
\end{figure}

We compute the quanta concentration at $t = 1, \dots, T$, which is in seconds, corresponding to the frequency of the tracking data. We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{s,0} = 0 \quad \forall s$. In other words, we assume that all quanta from the previous day has been removed from the air at the start of the following day.

\clearpage

\bibliography{references.bib}

\end{document}

### 2023-07-18 models/stm-v2.R
toy example

@@ -30,9 +30,9 @@ stm <- function(c0,
                 roomDim,
                 cellLength = 0.5,
                 pd = 1,
-                q = function() LaplacesDemon::rtrunc(1, "st", a = 0, b = 200, mu = 1, sigma = 0.67, nu = 1), 
-                lambda = function() rgamma(1, shape = 2.25, rate = 1.5),
-                k = function() 0,
+                q = 10, 
+                lambda = 0.5,
+                k = 0,
                 seed = 12345) {
   
   # room dimensions
@@ -64,7 +64,7 @@ stm <- function(c0,
         yb <- min(nr,inf_t$y[i]+pd)
         
         # distribute quanta uniformely
-        quanta_per_sec <- q() / 3600
+        quanta_per_sec <- q / 3600
         n_cells <- prod(dim(ct[[t+1]][yt:yb,xl:xr]))
         ct[[t+1]][yt:yb,xl:xr] <- ct[[t+1]][yt:yb,xl:xr] + quanta_per_sec * dt[t] / n_cells
       }
@@ -87,8 +87,8 @@ stm <- function(c0,
     }
     
     # removal
-    lambda_per_sec <- lambda() / 3600
-    k_per_sec <- k() / 3600
+    lambda_per_sec <- lambda / 3600
+    k_per_sec <- k / 3600
     ct[[t+1]] <- ct[[t+1]] * exp(-(aer_per_sec+lambda_per_sec+k_per_sec) * dt[t])
   }
   

### 2023-07-18 tests/test_stm_v2.R

@@ -1,6 +1,7 @@
 # Libraries and function
 library(tidyverse)
 source("models/stm-v2.R")
+source("utils/plotting.R")
 
 # Setup room
 dz <- 3
@@ -15,7 +16,7 @@ c0 <- matrix(0, nrow = ny, ncol = nx)
 # assume constant air exchange rate
 aer <- data.frame(
   time = seq(as.POSIXct("2023-01-01 08:00:00"), as.POSIXct("2023-01-01 12:00:00"), by = "1 sec"),
-  aer = 3
+  aer = 1
 )
 
 # assume one infectious person in the middle of the room at the beginning staying there for one hour
@@ -27,15 +28,29 @@ inf <- data.frame(
 
 
 # run model assuming no masks (pd = 1)
-ct <- stm(c0, aer, inf, roomDim, gridCellLength)
+ct <- stm(c0 = c0, aer = aer, inf = inf, roomDim = roomDim, cellLength = gridCellLength, pd = 1, q = 10, lambda = 0.5, k = 0, seed = 12345)
 
 # plot results
 ctDF <- do.call(rbind, lapply(seq_along(ct), function(t) reshape2::melt(ct[[t]]) %>% mutate(t = t) ))
 
-ctDF %>%
-  filter(t %in% c(1, 60, 600, 1800, 3600, 3610, 3630, 3660, 3900, 4200, 7200, 14401)) %>%
+t_names <- c("08:00:00am", "08:01:00am", "08:10:00am", "08:30:00am", "09:00:00", "09:00:10", "09:00:30", "09:01:00", "09:05:00", "09:10:00", "10:00:00", "12:00:00")
+t_sel <- c(1, 60, 600, 1800, 3600, 3610, 3630, 3660, 3900, 4200, 7200, 14401)
+names(t_names) <- t_sel
+
+quanta_concn_pl <- ctDF %>%
+  filter(t %in% t_sel) %>%
+  mutate(t = dplyr::recode(t, !!! t_names),
+         value = value * (1 / (gridCellLength^2*3))) %>%
   ggplot(aes(x = Var2, y = Var1, fill = value)) +
   facet_wrap(~ t) +
   geom_tile() +
+  labs(fill = expression("Concentration (quanta/m"^3*")")) +
   scale_fill_viridis_c(limits = c(0, NA)) +
-  theme(legend.position = "top", legend.key.width = unit(2, "cm"))
[…]

### 2023-07-18 utils/plotting.R

@@ -1,21 +1,15 @@
+require(patchwork)
 require(ggpattern)
 require(gridExtra)
-library(GGally)
-library(ggtext)
-library(wesanderson)
-
-# round format
-round_k <- function(x, k = 0) format(round(x, k), nsmall = k)
-
-# add white space
-add_ws <- function(x, n = 2) {
-  ifelse(nchar(x) < n, paste0(" ",x), x)
-}
-
+require(GGally)
+require(Cairo)
+require(gg.layers)
+require(wesanderson) # https://github.com/karthik/wesanderson
 
 # ggplot theme for Nature style
 text_size = 8
 update_geom_defaults("text", list(size = text_size))
+
 theme_nature <- function () { 
   theme_classic(base_size = text_size, base_family = 'sans') %+replace% 
     theme(
@@ -23,6 +17,21 @@ theme_nature <- function () {
       axis.title = element_text(size = text_size),
       plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0.5, margin = ggplot2::margin(0, 0, 5, 0))
     )
+}
+
+theme_custom <- function() {
+  theme_minimal() %+replace% 
+    theme(text = element_text(size = text_size),
+          axis.text = element_text(size = text_size),
+          axis.title = element_text(size = text_size),
+          plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 5, 0)),
+          strip.text = element_text(size = text_size),
+          panel.grid.major = element_blank(),
+          panel.border = element_blank(),
+          axis.line.x = element_line(),
+          axis.line.y = element_line(),
+          axis.ticks =  element_line(),
+          legend.text = element_text(size = 8))
 }
 
 theme_bw2 <- function () { 
@@ -90,7 +99,7 @@ hist_fun <- function(data, mapping) {
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
 
-save_plot <- function(pl, pdf_file = NULL, tikz_file = NULL, w = 8, h = 4) {
+save_plot <- function(pl, pdf_file = NULL, eps_file = NULL, tikz_file = NULL, w = 8, h = 4) {
   print(pl)
   if (!is.null(pdf_file)) {
[…]

### 2023-07-18 utils/trans_risk.r

@@ -11,6 +11,17 @@ dq <- function(x) {
   LaplacesDemon::dtrunc(x, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
 }
 
+#' Viral inactivation rate
+#' 
+
+rlambda <- function(n, disease = "TB") {
+  rgamma(1, shape = 2.25, rate = 1.5)
+}
+
+dlambda <- function(x, disease = "TB") {
+  dgamma(x, shape = 2.25, rate = 1.5)
+}
+
 
 #' Unmasked TB Patients
 #' 

### 2023-07-18 tests/test_stm_v2.R
minor fi

@@ -44,13 +44,14 @@ quanta_concn_pl <- ctDF %>%
   ggplot(aes(x = Var2, y = Var1, fill = value)) +
   facet_wrap(~ t) +
   geom_tile() +
-  labs(fill = expression("Concentration (quanta/m"^3*")")) +
-  scale_fill_viridis_c(limits = c(0, NA)) +
-  scale_x_continuous(expand = c(0,0), limits = c(0,40), labels = function(x) x * 0.25) +
-  scale_y_continuous(expand = c(0,0), limits = c(0,20), labels = function(x) x * 0.25) +
+  labs(fill = expression("Concentration (10"^-2*" quanta/m"^3*")")) +
+  scale_fill_viridis_c(limits = c(0, NA), labels = function(x) x * 100) +
+  scale_x_continuous(expand = c(0,0), limits = c(0,40), labels = function(x) x * 0.25, breaks = seq(0,40,10)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0,20), labels = function(x) x * 0.25, breaks = seq(0,10,5)) +
   theme_bw2() +
   theme(legend.position = "bottom", legend.key.width = unit(2, "cm"),
-        axis.title = element_blank())
+        axis.title = element_blank(), panel.spacing = unit(0.025, "cm"),
+        plot.margin = margin(r = 10))
 
 quanta_concn_pl
 save_plot(quanta_concn_pl, pdf_file = "tests/stm_v2-toy_example.pdf", w = 16, h = 10)

### 2023-07-18 doc/paper/supplements.tex
Updates from Overleaf

@@
 \begin{align}
-    K = (0.52 \cdot AER + 8.61e-5) * V^(2/3),
+    K = (0.52 \cdot AER + 8.61e-5) * V^{\frac{2}{3}},
 \end{align}
@@
     \centering
-    \includegraphics{}
-    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10$\times$B=\5$\times$H=3m.}
+    \includegraphics{../../tests/stm_v2-toy_example.pdf}
+    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10m$\times$B=5m$\times$H=3m.}
     \label{fig:toy-example}

### 2023-07-18 illustrations/priors.Rmd
plots for application

---
title: "Priors"
author: "Nicolas Banholzer"
date: "2023-07-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(LaplacesDemon)
source("../utils/trans_risk.r")
source("../utils/plotting.r")
```

## Quanta

```{r}
q_dat <- data.frame(x = seq(0, 200, 0.1)) %>%
  mutate(y = pq(x))

prev_est <- data.frame(
  study = c("Andrews (2014)", "Escombe (2008)", "Nardell (1991)", "Riley (1962)"),
  x = c(0.89, 8.2, 12.5, 1.25)
) %>%
  mutate(y = pq(x))

q_pl <- ggplot(mapping = aes(x = x, y = y)) +
  geom_line(data = q_dat, direction = "vh") +
  geom_segment(data = prev_est, mapping = aes(xend = x, yend = 0, color = study), linetype = "dotted") +
  geom_segment(data = prev_est, mapping = aes(yend = y, xend = 0, color = study), linetype = "dotted") +
  geom_point(data = prev_est, mapping = aes(color = study), shape = 17, size = 2) +
  geom_text(data = prev_est, mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35)) +
  scale_x_continuous(expand = expansion(mult = c(0,.025)), breaks = c(0, 1, 10, 100, 200), trans = scales::pseudo_log_trans(base = 10)) +
  scale_y_continuous(expand = c(0,0), labels = scales::percent_format(suffix = "")) +
  scale_color_manual(values = wes_palette("Cavalcanti1")) +
  labs(x = "Quanta generation rate (quanta/h)", y = "Cumulative probability (%)") +
  theme_custom() +
  theme(legend.position = "none")

q_pl
save_plot(q_pl, pdf_file = "../illustrations/prior-q.pdf", w = 12, h = 8)
```

## Viral inactivation rate

```{r}
plambda <- function(x, disease = "TB") {
  #LaplacesDemon::ptrunc(x, spec = "st", a = 0, b = 10, mu = 0.5, sigma = 1, nu = 5)
  plnorm(x, meanlog = log(0.5), sdlog = 1)
}

lambda_dat <- data.frame(x = seq(0, 10, .01)) %>%
  mutate(y = plambda(x))

prev_est_vir <- data.frame(
  study = c("Loudon (1969)", "Lever (2000)", "Gannon (2007)"),
  x = c(0.12, 8.3, 0.46)
) %>%
  mutate(y = plambda(x))

lambda_pl <- ggplot(mapping = aes(x = x, y = y)) +
  geom_line(data = lambda_dat, direction = "vh") +
  geom_segment(data = prev_est_vir, mapping = aes(xend = x, yend = 0, color = study), linetype = "dotted") +
  geom_segment(data = prev_est_vir, mapping = aes(yend = y, xend = 0, color = study), linetype = "dotted") +
  geom_point(data = prev_est_vir, mapping = aes(color = study), shape = 17, size = 2) +
  geom_text(data = prev_est_vir %>% dplyr::filter(study != "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35), nudge_y = .02) +
  geom_text(data = prev_est_vir %>% dplyr::filter(study == "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = 0.05) +
  scale_x_continuous(expand = expansion(mult = c(0,.025)), breaks = c(0, 0.5, 1, 2, 5, 10), trans = scales::pseudo_log_trans(base = 10)) +
  scale_y_continuous(expand = expansion(mult = c(0,.025)), labels = scales::percent_format(suffix = "")) +
  scale_color_manual(values = wes_palette("Cavalcanti1")) +
  labs(x = "Viral inactivation rate (quanta/h)", y = "Cumulative probability (%)") +
  theme_custom() +
  theme(legend.position = "none")

lambda_pl
save_plot(lambda_pl, pdf_file = "../illustrations/prior-lambda.pdf", w = 12, h = 8)
```

### 2023-07-18 tests/test_stm_v2.R

@@ -46,11 +46,11 @@ quanta_concn_pl <- ctDF %>%
   geom_tile() +
   labs(fill = expression("Concentration (10"^-2*" quanta/m"^3*")")) +
   scale_fill_viridis_c(limits = c(0, NA), labels = function(x) x * 100) +
-  scale_x_continuous(expand = c(0,0), limits = c(0,40), labels = function(x) x * 0.25, breaks = seq(0,40,10)) +
-  scale_y_continuous(expand = c(0,0), limits = c(0,20), labels = function(x) x * 0.25, breaks = seq(0,10,5)) +
+  scale_x_continuous(expand = c(0,0), limits = c(0,40), labels = function(x) x * 0.25, breaks = seq(0,40,20)) +
+  scale_y_continuous(expand = c(0,0), limits = c(0,20), labels = function(x) x * 0.25, breaks = seq(0,20,10)) +
   theme_bw2() +
   theme(legend.position = "bottom", legend.key.width = unit(2, "cm"),
-        axis.title = element_blank(), panel.spacing = unit(0.025, "cm"),
+        axis.title = element_blank(), panel.spacing = unit(0.5, "cm"),
         plot.margin = margin(r = 10))
 
 quanta_concn_pl

### 2023-07-18 utils/trans_risk.r

@@ -7,19 +7,29 @@ rq <- function(n, disease = "TB") {
   LaplacesDemon::rtrunc(n, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
 }
 
-dq <- function(x) {
+dq <- function(x, disease = "TB") {
   LaplacesDemon::dtrunc(x, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
+}
+
+pq <- function(x, disease = "TB") {
+  LaplacesDemon::ptrunc(x, spec = "st", a = 0, b = 200, mu = 2, sigma = 2.5, nu = 1)
 }
 
 #' Viral inactivation rate
 #' 
+#' @param n number of samples
+#' @param x viral inactivation rate
 
 rlambda <- function(n, disease = "TB") {
-  rgamma(1, shape = 2.25, rate = 1.5)
+  rgamma(n, shape = 2.25, rate = 1.5)
 }
 
 dlambda <- function(x, disease = "TB") {
   dgamma(x, shape = 2.25, rate = 1.5)
+}
+
+plambda <- function(x, disease = "TB") {
+  pgamma(x, shape = 0.21, rate = 0.46)
 }
 
 
@@ -44,6 +54,7 @@ rTBunmasked <- function(n, lambda) {
 #' based on Equation 16, page 7, Batterman et al. (2017): https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5334699/
 #' assume CO2 generation rate based G on Persily et al. (2007): https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5666301/
 #' --> use 0.004 L/s = 0.24 L/min
+
 transient_mass_balance <- function(A, C, n, V, Cr, G, dt) {
   Q <- A * V
   C1hat <- 6 * 10^4 * n * G / Q * (1 - exp(- Q / V * dt)) + (C - Cr) * exp(- Q / V * dt) + Cr

### 2023-07-19 analysis/priors.Rmd
prior illustrations

--- 2023-07-18 illustrations/priors.Rmd
@@
   geom_point(data = prev_est_vir, mapping = aes(color = study), shape = 17, size = 2) +
-  geom_text(data = prev_est_vir %>% dplyr::filter(study != "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35), nudge_y = .02) +
+  geom_text(data = prev_est_vir %>% dplyr::filter(study != "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.13), nudge_y = .02) +
   geom_text(data = prev_est_vir %>% dplyr::filter(study == "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = 0.05) +

### 2023-07-19 utils/trans_risk.r

@@ -1,3 +1,17 @@
+#' Reduction in quanta emission rate through surgical mask wearing
+#' 
+#' @param n number of samples
+#' @param x reduction rate
+
+rrmask <- function(n) {
+  LaplacesDemon::rtrunc(n, spec = "norm", a = 0, b = 1, mean = 0.56, sd = 0.11)
+}
+
+drmask <- function(n) {
+  LaplacesDemon::dtrunc(x, spec = "norm", a = 0, b = 1, mean = 0.56, sd = 0.11)
+}
+
+
 #' Quanta emission rate
 #' 
 #' @param n number of samples
@@ -21,15 +35,15 @@ pq <- function(x, disease = "TB") {
 #' @param x viral inactivation rate
 
 rlambda <- function(n, disease = "TB") {
-  rgamma(n, shape = 2.25, rate = 1.5)
+  rlnorm(x, meanlog = log(1), sdlog = 1)
 }
 
 dlambda <- function(x, disease = "TB") {
-  dgamma(x, shape = 2.25, rate = 1.5)
+  dlnorm(x, meanlog = log(1), sdlog = 1)
 }
 
 plambda <- function(x, disease = "TB") {
-  pgamma(x, shape = 0.21, rate = 0.46)
+  plnorm(x, meanlog = log(1), sdlog = 1)
 }
 
 

### 2023-07-19 doc/paper/supplements.tex
Updates from Overleaf

@@
 It rather represents the dose of infectious particles that corresponds to a probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection [Rudnick and Milton].
-It is an attempt to consider the stochastic behavior of airborne transmission, which depends on the environment, characteristics of the pathogen, and characteristics of the individual.
+It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown.
+Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual.
 For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person [Wang], increasing the probability of infection.
@@
 \begin{align}
-    K = (0.52 \cdot AER + 8.61e-5) * V^{\frac{2}{3}},
+    K = (0.52 \cdot AER + 8.61\cdot10^{-5}) \cdot V^{\frac{2}{3}},
 \end{align}
@@
 \begin{align}
-    AER = \frac{6e4 \cdot n \cdot G}{V\cdot(CO^{\text{I}}-CO^{\text{O}})},
+    AER = \frac{6\cdot10^4 \cdot n \cdot G}{V\cdot(CO^{\text{I}}-CO^{\text{O}})},
 \end{align}
@@
 \begin{align}
-    CO_{t+1}^{\text{I}} = \frac{6e4 \cdot n_t + G}{Q} \cdot \left(1 - \exp(-Q/V \Delta t)\right) + (CO_t^{\text{I}}-CO^{\text{O}}) \cdot \exp(-Q/V \Delta t) + CO^{\text{O}},
+    CO_{t+1}^{\text{I}} = \frac{6\cdot10^4 \cdot n_t + G}{Q} \cdot \left(1 - \exp(-Q/V \Delta t)\right) + (CO_t^{\text{I}}-CO^{\text{O}}) \cdot \exp(-Q/V \Delta t) + CO^{\text{O}},
 \end{align}
@@
 We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$ [Battermann].
-The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$
+The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$.
 
@@
 
-\subsection{Illustrative example}
+\subsection{Illustrative example}\label{sec:example}
 
@@
 We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells (\ie initially quanta reach a maximum distance of 0.7m through exhalation corresponding to the diagonal length of two cells).
-We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(20\pm1,10\pm1)$.
+We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(x_s = 20\pm1, y_s = 10\pm1)$, \ie $N_{s,t} = (I \cdot q) / 9 ~~ \forall s \in (x_s = 20\pm1, y_s = 10\pm1)$.
 We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of also 1\,quanta/h and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$\,quanta/h.
@@
 
-\begin{figure}
+\begin{figure}[!htpb]
     \centering
     \includegraphics{../../tests/stm_v2-toy_example.pdf}
-    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10m$\times$B=5m$\times$H=3m.}
+    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10m $\times$ B=5m $\times$ H=3m.}
     \label{fig:toy-example}
@@
 
-\section{Case study}
+\section{Estimation of infection risk in the primary care clinic}
 
-We divide the clinical day into a morning (7am to 12am) and afternoon period (12am to 5pm) and estimate the $AER$ separately for each period, allowing it to vary by daytime.
+\subsection{Building and setup}
+
+A schematic view of the clinic is shown in \Cref{fig:clinic}.
+Clinical attendees were tracked when entering the clinic, in the waiting room, the corridor, and the TB room.
+If they moved into one of the care rooms along the corridor or into other parts of the clinic they were only tracked once they came back.
+We divide the clinic into three spaces: the waiting room (L=10.55m $\times$ B=5.50m $\times$ H=3.00m), the corridor (L=12.45m $\times$ B=2.20m $\times$ H=2.50m), and the TB room (L=4.75m $\times$ B=3.50m $\times$ H=3.00m).
+We separated the waiting room and the corridor because we observed small differences in the time-varying CO$_2$ levels, but we acknowledge that the two areas could also be evaluated jointly since the doors to the corridor were often most of the time.
+Moreover, the corridor and waiting room are not completely enclosed, but connected to the entrance, exit, and other parts of the clinic.
+Thus, we simplify our modeling analysis when we limit quanta diffusion within these rooms and also note that the calculated air change rate is affected by the separation of the waiting room and corridor (\ie the volume and room occupancy would otherwise be different).
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
+    \caption{Schematic view of the part of the clinic where clinical attendees were tracked.}
+    \label{fig:clinic}
+\end{figure}
+
+As in our illustrative example (\Cref{sec:example}), we divide each room into a grid of cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+Since the rooms have different height, the quanta concentration (quanta/m$^3$) in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
+We assume that quanta generation is confined only to the cell where the infectious individual is located, thereby using a smaller radius as compared to our example due to the fact that mask wearing was compulsory during the time of the study (\ie we assume that mask wearing halves the maximum distance that the quanta reaches through exhalation).
+Nevertheless, we compare our results with the assumption made in our example that the quanta concentration is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
+
+We divide the clinical day into a morning (7am to 12am) and afternoon period (12am to 5pm) and estimate the outdoor air exchange rate separately for each period.
+Thereby, we allow the $AER$ to vary by daytime, considering potential differences in natural ventilation habits in the morning and afternoon.
+
+We update the quanta concentration every second at $t = 1, \dots, T$ (aligned with the frequency of the tracking data).
+We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{s,0} = 0 ~~ \forall s$.
+In other words, we assume that all quanta from the previous day has been removed from the air before the start of the following clinical day.
+
+\subsection{Identifying infectious individuals}
+
+\subsection{Monte-Carlo simulation}
+
+\subsection{Priors for modeling parameters}
 
 Previous studies estimated $q$ for \emph{Mtb} or SARS-CoV-2 using the Wells-Riley model [REFs].
-The range of estimates is large, suggesting that the infectiousness of \emph{Mtb} patients varies considerably.
-We model $q$ with a Student-t distribution with 1 degree of freedom $t_1(\mu = 2, \sigma = 2.5)$ that is truncated at 0 (lower) and 200 (upper).
+Andrews (2014) [REF] reported an estimate of 0.89 quanta/h (range from 0.44 to 5.69), close to the early estimate provided by Riley (1962) [REF] with 1.25 quanta/h.
+On the other end, Escombe (2008) [REF] arrived at an estimate of 8.2 quanta/h, which was lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients.
+Nardell (1991) [REF] also reported a higher estiamte of 13 quanta/h, but only for one index case.
+The wide range of estimates and the findings by Escombe (2008) suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients.
+Therefore, we model $q$ with a Student-t distribution with 1 degree of freedom $t_1(\mu = 2, \sigma = 2.5)$ that is truncated at 0 (lower) and 200 (upper), as in Banholzer (X) [OUR PLOS GLOBAL].
+The Student-t distribution with only 1 degree of freedom considers the possibility of superspreaders that generate large amounts of quanta.
 The distribution is shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.
 
-\begin{figure}
+\begin{figure}[!htpb]
     \centering
-    \includegraphics{}
-    \caption{Caption}
+    \includegraphics{illustrations/prior-q.pdf}
+    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014) [REF], 1.25 by Riley (1962) [REF], 8.2 by Escombe (2008) [REF], and 13 by Nardell (1991) [REF].}
     \label{fig:quanta-distribution}
@@
 
-A schematic view of the clinic is shown in \Cref{fig:clinic}.
-We divide the clinic into three airspaces: the waiting room (l x b x h), the corridor (l x b x h), and the TB room (l x b x h).
-Each room is divided into a grid of cubic cells with a fixed cell length and breath of 0.25m.
-The volume of the grid cells thus depends on the height of each airspace.
+Face masks will reduce the number of quanta that are exhaled into the air, but the extent of the reduction is subject to uncertainty.
+A study with multidrug-resistant TB cases estimated a decrease of 56\% (95\%-CrI, 33\% to 70.5\%) when patients used surgical face masks\cite{Dharmadhikari2012AJRCCM}.
+This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic.
+The estimate is also comparable to the reduction in the particle number concentration during mask wearing in school rooms of 69\% (95\%-CrI 42\% to 86\%)\cite{Banholzer2023PLoSMed}.
+Based on the estimate by Dharmadhikari (2012)\cite{Dharmadhikari2012AJRCCM}, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
 
-\begin{figure}
+The viral inactivation rate for \emph{Mtb} has not been precisely estimated.
+Loudon (1969) [REF] report a half-life for aerosolized TB bacilli of 6h ($\lambda$ = 0.12).
+By contrast, Lever (2000) [REF] reported a half-life of just 5min ($\lambda$ = 8.3).
+The difference may largely be explained by variation in relative humidity, \ie Loudon (1969) studied survival during a relative humidity of about 50\% whereas Lever (2000) studied survival during a relative humidity of about 70\%.
+Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) seems more representative of the environmental conditions in our study.
+Moreover, Gannon (2007) [REF] report a half-life of 1.5 hours ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}.
+In contrast to that, Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30min ($\lambda$ = ), which would be closer to Lever (2000).
+We assume a ...
+.
+The distribution is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
+
+\begin{figure}[!htpb]
     \centering
-    \includegraphics{}
-    \caption{Caption}
-    \label{fig:clinic}
+    \includegraphics{illustrations/prior-lambda.pdf}
+    \caption{Prior distribution function for the viral inactivation rate along with reported estimates in the literature: ... [REFs]}
+    \label{fig:lambda-distribution}
 \end{figure}
-
-We compute the quanta concentration at $t = 1, \dots, T$, which is in seconds, corresponding to the frequency of the tracking data.
-We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{s,0} = 0 \quad \forall s$.
-In other words, we assume that all quanta from the previous day has been removed from the air at the start of the following day.
 

### 2023-07-19 illustrations/priors.Rmd
minor update to illustrations of priors

@@
 plambda <- function(x, disease = "TB") {
-  #LaplacesDemon::ptrunc(x, spec = "st", a = 0, b = 10, mu = 0.5, sigma = 1, nu = 5)
-  plnorm(x, meanlog = log(0.5), sdlog = 1)
+  plnorm(x, meanlog = log(1), sdlog = 1)
 }
@@
 prev_est_vir <- data.frame(
-  study = c("Loudon (1969)", "Lever (2000)", "Gannon (2007)"),
-  x = c(0.12, 8.3, 0.46)
+  study = c("Loudon (1969)", "Lever (2000)", "Gannon (2007)", "Klein (2014)"),
+  x = c(log(2)/6, log(2)/(5/60), log(2) / 1.5, log(2) / (0.5/(log(20/100, 0.5))) )
 ) %>%
@@
   geom_point(data = prev_est_vir, mapping = aes(color = study), shape = 17, size = 2) +
-  geom_text(data = prev_est_vir %>% dplyr::filter(study != "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35), nudge_y = .02) +
-  geom_text(data = prev_est_vir %>% dplyr::filter(study == "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = 0.05) +
+  geom_text(data = dplyr::filter(prev_est_vir, study == "Lever (2000)"), 
+            mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = 0.05) +
+  geom_text(data = dplyr::filter(prev_est_vir, study == "Klein (2014)"), 
+            mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = -0.05, nudge_x = 0.1) +
+  geom_text(data = dplyr::filter(prev_est_vir, study == "Gannon (2007)"), 
+                                 mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = 0.125) +
+  geom_text(data = dplyr::filter(prev_est_vir, study == "Loudon (1969)"), 
+                                 mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = 0.125, nudge_y = 0.025) +
   scale_x_continuous(expand = expansion(mult = c(0,.025)), breaks = c(0, 0.5, 1, 2, 5, 10), trans = scales::pseudo_log_trans(base = 10)) +
@@
   scale_color_manual(values = wes_palette("Cavalcanti1")) +
-  labs(x = "Viral inactivation rate (quanta/h)", y = "Cumulative probability (%)") +
+  labs(x = "Viral inactivation rate", y = "Cumulative probability (%)") +
   theme_custom() +

### 2023-07-19 doc/paper/supplements.tex

@@
 
-We extend the Wells-Riley model [Wells-Riley], which is an established epidemiological model to estimate the risk of infection.
+We extend the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection.
 The model estimates the probability of airborne indoor transmission $P$ using the following equation:
@@
 The unknown parameter $q$ is not the actual number of infectious particles in the air.
-It rather represents the dose of infectious particles that corresponds to a probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection [Rudnick and Milton].
+It rather represents the dose of infectious particles that corresponds to a certain probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection\cite{Rudnick2003IndoorAir}.
 It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown.
 Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual.
-For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person [Wang], increasing the probability of infection.
+For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person\cite{Wang2021Science}, increasing the probability of infection.
 
@@
 
-Rudnick and Milton [Rudnick and Milton] modified the Wells-Riley model to relax the steady-state assumption.
+Rudnick and Milton\cite{Rudnick2003IndoorAir} modified the Wells-Riley model to relax the steady-state assumption.
 They use CO$_2$ as a biomarker for exhaled breath to compute the outdoor air supply rate, which is otherwise difficult to measure.
@@
 Formally, we estimate the cumulative risk of infection as
-\begin{equation}
+\begin{equation}\label{eq:spattemp-P}
     P = \sum_s \sum_t N_{s,t} \cdot p,
@@
 
-\subsection{Setup}
+\subsection{Model setup and notation}
 
@@
 
-\subsection{Notation}
-
+We define the following variables and modeling parameters:
 \begin{itemize}
@@
     \item $n_t$: number of individuals in airspace at time $t$
-    \item $\lambda$: viral inactivation rate (in quanta/s)
+    \item $\lambda$: viral inactivation rate
 \end{itemize}
@@
 The initial spatial spread of the quanta over the grid is informed prior literature.
-The concentration of virus-laden aerosols is highest closest to the infectious individual [References], in line with findings showing that the risk of infection correlates with the distance to the infectious individual [References].
+The concentration of virus-laden aerosols is highest closest to the source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}, in line with findings showing that the risk of infection correlates with the distance to the infectious individual [References].
 Although coughing and sneezing can transport aerosols further away from the individual, these activities are much less frequent than breathing.
-Therefore, prior work suggests that breathing may contribute more infectious particles to the airspace than coughing or sneezing [Dinkele].
+Therefore, prior work suggests that breathing may contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}.
 Overall, it seems plausible that the majority of infectious aerosols are generated near the infectious individual and then disperse over time.
-In line with prior findings, we will assume that virus-laden aerosols reach a distance of 0.50m to 1.0m through exhalation [References].
 Since we do not observe the directional flow of the exhaled breadth, we assume that the generated quanta is evenly and radially distributed around the individual.
@@
 \end{align}
-where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$), $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air (both in parts per million [ppm]) [Battermann].
+where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$), $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air (both in parts per million [ppm])\cite{Batterman2017IJERPH}.
 In a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration.
-Therefore, we determine $AER$ using a transient mass balance model of the form [Battermann]
+Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
 \begin{align}
@@
 where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$.
-We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) [REF], a quasi-Newton method.
+We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) as implemented in the statistical software R\cite{Byrd1995SIAM}, a quasi-Newton method.
 The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
-We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$ [Battermann].
+We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$\cite{Batterman2017IJERPH}.
 The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$.
@@
 We will ignore gravitational settling.
-Vuorinen et al.
-[Reference] notes that particles smaller than 20\,$\mu$m rapidly become airborne.
-This is far above the upper end of the particle size distribution of most respiratory infections, including \emph{Mtb}, which is carried in particles typically smaller than 4.7\,$\mu$m [Fennelly].
+Vuorinen (2020)\cite{Vuorinen2020SafSci} notes that particles smaller than 20\,$\mu$m rapidly become airborne.
+This is far above the upper end of the particle size distribution of most respiratory infections, including \emph{Mtb}, which is carried in particles typically smaller than 4.7\,$\mu$m\cite{Fennelly2020Lancet}.
 The removal rate is thus the same of the outdoor air exchange rate $AER_t$ and the viral inactivation rate $\lambda$, \ie
@@
 Considering the generation (\Cref{eq:generation}), diffusion (\Cref{eq:diffusion}), and removal (\Cref{eq:removal}) of quanta together, the quanta concentration at time $t$ in the airspace is computed as
-\begin{align}
+\begin{align}\label{eq:spattemp-N}
     \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
@@
 We divide the room into a grid of $S = 40 \cdot 20 = 800$ cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
-We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells (\ie initially quanta reach a maximum distance of 0.7m through exhalation corresponding to the diagonal length of two cells).
+We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
 We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(x_s = 20\pm1, y_s = 10\pm1)$, \ie $N_{s,t} = (I \cdot q) / 9 ~~ \forall s \in (x_s = 20\pm1, y_s = 10\pm1)$.
-We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of also 1\,quanta/h and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$\,quanta/h.
+We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of 1 and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$.
 
@@
 Since the rooms have different height, the quanta concentration (quanta/m$^3$) in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
-We assume that quanta generation is confined only to the cell where the infectious individual is located, thereby using a smaller radius as compared to our example due to the fact that mask wearing was compulsory during the time of the study (\ie we assume that mask wearing halves the maximum distance that the quanta reaches through exhalation).
-Nevertheless, we compare our results with the assumption made in our example that the quanta concentration is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
 
@@
 
-\subsection{Identifying infectious individuals}
+\subsection{Data and linkage}
 
@@
 
-\subsection{Priors for modeling parameters}
+We have uncertainty in the data (clinical attendees that are infectious but undiagnosed) and modeling parameters (quanta generation and viral inactivation rate).
+We consider this uncertainty using Monte-Carlo simulation where uncertain parameters are modeled with prior distributions (see \Cref{sec:priors} below).
+In each run of the simulation, we perform the following steps:
+\begin{itemize}
+    \item[\textbf{1.}] Sample uncertain modeling parameters such as the quanta generation rate and the proportion of undiagnosed TB patients among clinical attendees.
+    \item[\textbf{2.}] Sample undiagnosed TB patients and combine with diagnosed TB patients, and then use the tracking data to determine the location and time of where and when quanta was generated by these infectious individuals.
+    \item[\textbf{3.}] For each room (waiting room, corridor, and passage), compute the quanta concentration in each cell $s$ and at time $t$ using \Cref{eq:spattemp-N} from the spatiotemporal model.
+    \item[\textbf{4.}] For each susceptible clinical attendee, compute the risk of infection from the cumulative exposure to quanta over airspace and time using \Cref{eq:spattemp-P}.
+\end{itemize}
 
-Previous studies estimated $q$ for \emph{Mtb} or SARS-CoV-2 using the Wells-Riley model [REFs].
-Andrews (2014) [REF] reported an estimate of 0.89 quanta/h (range from 0.44 to 5.69), close to the early estimate provided by Riley (1962) [REF] with 1.25 quanta/h.
-On the other end, Escombe (2008) [REF] arrived at an estimate of 8.2 quanta/h, which was lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients.
-Nardell (1991) [REF] also reported a higher estiamte of 13 quanta/h, but only for one index case.
-The wide range of estimates and the findings by Escombe (2008) suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients.
+Most clinical attendees could be undiagnosed TB Patients and there is considerable uncertainty in some modeling parameters.
+As a consequence, we need a high number of Monte Carlo runs to reasonably estimate the risk of infection for each individual.
+Therefore, for each study day, we perform 10,000 Monte-Carlo simulations.
+If not stated otherwise, we summarize the results with the mean and 95\% credible interval.
+
+\subsection{Prior choices for modeling parameters}\label{sec:priors}
+
+\subsubsection{Initial spread}
+
+% spatial spread of quanta generation
+
+An analysis of exhalation activities suggest a propagation distance (\ie the maximum distance from the source to the end of the exhaled air before dispersion) of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity.
+Face masks reduce the propagation distance as most exhaled air (and particles) by locking the formation of the jet and leaking air on the sides of the masks\cite{Tang2009RoyalInt,Hui2012PLoSOne}.
+Therefore, we assume that mask wearing confined the quanta generation only to the cell where the infectious individual is located.
+This corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.
+
+To assess the impact of mask wearing, we also model the scenario without mask wearing.
+In this scenario, we assume a propagation distance of 0.7m (the diagonal lengths of two cells), in line with the aforementioned prior findings.
+Since we cannot observe the direction of jet, we assume that all neighbouring cells are affected equally.
+That is, we divide the generated quanta by 9 (the cell where the individual is located and all first-neighbouring cells) and distribute it equally across the cells.
+
+\subsubsection{Quanta generation rate}
+
+% quanta generation
+
+Previous studies estimated $q$ for \emph{Mtb} or SARS-CoV-2 using the Wells-Riley model\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD}.
+Andrews (2014)\cite{Andrews2014JID} reported an estimate of 0.89 quanta/h (range from 0.44 to 5.69), close to the early estimate provided by Riley (1962)\cite{Riley1962ARRD} with 1.25 quanta/h.
+On the other end, Escombe (2008)\cite{Escombe2008PLoSMed} arrived at an estimate of 8.2 quanta/h, which was lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients.
+Nardell (1991)\cite{Nardell1991ARRD} also reported a higher estiamte of 13 quanta/h, but only for one index case.
+The wide range of estimates suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}.
 Therefore, we model $q$ with a Student-t distribution with 1 degree of freedom $t_1(\mu = 2, \sigma = 2.5)$ that is truncated at 0 (lower) and 200 (upper), as in Banholzer (X) [OUR PLOS GLOBAL].
 The Student-t distribution with only 1 degree of freedom considers the possibility of superspreaders that generate large amounts of quanta.
-The distribution is shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.
+The distribution function is shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.
 
@@
 
-Face masks will reduce the number of quanta that are exhaled into the air, but the extent of the reduction is subject to uncertainty.
+% face mask reduction
+
+Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens} and thus lower the generation of infectious quanta.
+The extent of the reduction is subject to uncertainty.
 A study with multidrug-resistant TB cases estimated a decrease of 56\% (95\%-CrI, 33\% to 70.5\%) when patients used surgical face masks\cite{Dharmadhikari2012AJRCCM}.
 This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic.
-The estimate is also comparable to the reduction in the particle number concentration during mask wearing in school rooms of 69\% (95\%-CrI 42\% to 86\%)\cite{Banholzer2023PLoSMed}.
-Based on the estimate by Dharmadhikari (2012)\cite{Dharmadhikari2012AJRCCM}, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
+Therefore,, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
+The effective quanta generation rate is thus $q$ times the reduction due to mask wearing.
+Again, we model what the risk of infection would have been without mask wearing by removing the reduction term from the model.
+
+\subsubsection{Viral inactivation rate}
+
+% viral inactivation rate
 
 The viral inactivation rate for \emph{Mtb} has not been precisely estimated.
-Loudon (1969) [REF] report a half-life for aerosolized TB bacilli of 6h ($\lambda$ = 0.12).
-By contrast, Lever (2000) [REF] reported a half-life of just 5min ($\lambda$ = 8.3).
+Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6h ($\lambda$ = 0.12).
+By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5min ($\lambda$ = 8.3).
 The difference may largely be explained by variation in relative humidity, \ie Loudon (1969) studied survival during a relative humidity of about 50\% whereas Lever (2000) studied survival during a relative humidity of about 70\%.
 Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) seems more representative of the environmental conditions in our study.
-Moreover, Gannon (2007) [REF] report a half-life of 1.5 hours ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}.
-In contrast to that, Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30min ($\lambda$ = ), which would be closer to Lever (2000).
-We assume a ...
-.
-The distribution is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
+Moreover, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5 hours ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}.
+In contrast to that, Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30min ($\lambda$ = 3.2), which would be closer to Lever (2000).
+We choose a middle ground and, therefore, use a Lognormal($\mu$ = $\log$ 1, $\sigma$ = 1) with a median viral inactivation rate of 1.
+The distribution function is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
 
@@
     \includegraphics{illustrations/prior-lambda.pdf}
-    \caption{Prior distribution function for the viral inactivation rate along with reported estimates in the literature: ... [REFs]}
+    \caption{Prior distribution function for the viral inactivation rate along with reported estimates in the literature: 0.12 by Loudon (1969) [REF], 0.46 by Gannon (2007) [REF], 3.2 by Klein (2014) [REF], and 8.3 by Lever (2000) [REF].}
     \label{fig:lambda-distribution}
 \end{figure}
+
+\subsubsection{Number of undiagnosed TB patients}
+
+% proportion of undiagnosed TB patients
+
+In addition to the diagnosed, known TB patients, there may also be some undiagnosed, unknown TB patients among the clinical attendees.
+These unmasked TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats).
+A systematic review concluded that half of TB cases among people living with HIV that are on ART may be missed by the symptom screen\cite{Hamada2018LancetHIV}.
+Similarly, a study universally testing clinical attendees for TB that belong to a risk group (living with HIV, self-reported close contact with TB patients, prior history of TB) finds that 55\% of the attendees without a positive symptom screen were tested positive for TB\cite{Berhanu2023CID}.
+We interpret these findings such that the number of masked TB patients is roughly equal to the number of unmasked TB patients.
+On average, there were X masked TB patients per day attending the clinic.
+Therefore, we model the daily number of unmasked TB patients with a Poisson($\mu$ = X) distribution.
+
+% based on incidence
 

### 2023-08-03 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mtb} transmission in a South African primary care clinic}}
+\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}
 % author list
@@
 
-...
+% TB and routes of transmission
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in South-East Asian and African region [TB report].
+Progress made before the COVID-19 pandemic has stalled or reversed as the number of TB-related deaths have increased between 2019 and 2021 [TB report].
+\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious person [Rieder, Patterson].
+\emph{Mtb} is carried primarily in smaller particles $\leq5\mu$m called aerosols [Fennelly], which can survive in the air for multiple hours [Loudon].
+Airborne transmission is more likely in crowded, poorly ventilated indoor settings [Rieder, Center for Prevention Services, Nardell, Morawska, Wang].
+Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities, such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic [McCreesh 2020].
+A study in South Africa estimates that 4\% to 14\% of TB cases in adults originate from primary care clinics [McCreesh 2022].
+
+% the Wells-Riley model
+The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings [Andrews Integrating 2014, Taylor, Hella], including primary care clinics [Zuercher, McCreesh 2021].
+The Wells-Riley model assumes a well-mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (so-called quanta), the breathing rate per person, and the outdoor air supply rate.
+The quanta generation rate (commonly denoted with the parameter $q$) is unknown, but can be estimated by computing the risk of infection as the ratio of diseased/susceptible cases, and then solving the Wells-Riley equation for the parameter $q$\cite{Nardell1991ARRD,Escombe2008PLoSMed}.
+The Wells-Riley model considers the importance of ventilation by which infectious quanta are removed from the indoor space through outdoor air exchange.
+While the ventilation rate is difficult to estimate, an alternative formulation of the Wells-Riley equation exists that uses indoor CO$_2$ levels as a marker of exhaled-breath exposure\cite{Rudnick2003IndoorAir}, which is related to the ventilation rate.
+
+% limitation of the Wells-Riley model
+A crucial assumption of the Wells-Riley model is that the airspace is well-mixed, which means that there is no spatial variation in the exposure to infectious doses.
+Although it is reasonable to assume that infectious particles disperse in the indoor space over time, the aerosol concentration is at first highest closest to the infectious person\cite{Vuorinen2020SafSci,Chen2020BuildEnv}.
+Therefore, both short- and long-range transmission of \emph{Mtb} (and other pathogens) seem plausible.
+Nevertheless, previous studies show that the risk of infection is associated with proximity to an infectious person [Ko 2004, Kenyon], which is in line with the hypothesis that prolonged close contact is required for transmission [Leung, Brankston2007LancetID, Narasimhan].
+However, proximity to an infectious person is not considered in the Wells-Riley model.
+It would also be difficult to test such a spatiotemporal extension of the Wells-Riley model because the location of the infectious and susceptible individuals are rarely observed.
+
+% what this study adds
+We build on the Wells-Riley equation and develop a spatiotemporal model for the concentration of infectious quanta in the indoor space over time.
+We combine clinical and video tracking data to inform the spatiotemporal location of \emph{Mtb} infectious and susceptible individuals visiting a primary care clinic in South Africa between October and November 2021.
+Furthermore, we measure indoor CO$_2$ levels to model the diffusion and removal of infectious quanta over time.
+Our model allows us to identify spatiotemporal hotspots of quanta inside the clinic and estimate the risk of infection individually for each clinical attendee.
+We compare our modeling approach with an alternative specification that assumes a well-mixed airspace, thereby quantifying the contribution of close contact with infectious attendees to \emph{Mtb} transmission risk.
 
@@
 % summary
+We modeled the spatiotemporal concentration of infectious quanta, considering the location of infectious individuals (quanta generation), spatial dispersion (quanta diffusion), and natural ventilation (quanta removal).
+We found ...
 
-...
+% spatial modeling and airflow
+
+% factors affecting quanta generation
+The generation, dispersion, and survival of infectious particles in the air is subject to considerable uncertainty.
+Therefore, Wells [Wells 1955] introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission.
+A higher concentration of infectious quanta corresponds to a higher probability for an individual to get infected.
+The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
+Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~properties of the specific pathogenic, and the (3)~infectiousness of infected and susceptibility of uninfected TB patients.
+
+(1)~Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate outdoor air supply.
+The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
+This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace [REFs].
+However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic.
+The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models [REFs].
+CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow.
+For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air [Li].
+However, airflow is rarely considered and it may be difficult to assess it in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
+
+(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components [Wang].
+Environmental factors, in particular temperature and humidity, can affect these properties but the effects are often unclear and depending on the pathogen [Songer, Chan, Cox, Fernstrom, Tang].
+Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear [Tang].
+On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity [Lever, Loudon].
+On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity [Sornboot, Matuka].
+
+(3)~Previous studies show considerable variation in the generation of infectious quanta by TB infected patients [Escombe, Andrews].
+We model such variation by using a wide prior distribution for the quanta generation rate.
+Our model could be easily extended to incorporate prior information on the infectiousness of each individual TB patient.
+In general, only patients with active pulmonary TB are able to produce infectious droplets [Rieder].
+Infectiousness could be determines based on the most recent sputum test result.
+Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients [Shaw, Brindle, Grzybowski].
+HIV status could further inform both infectiousness and susceptibility.
+On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum [Brindle, Telzak].
+On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status [Forte, Kwan, Shen], and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
+
+% data and other limitations
+While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations.
+First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors.
+We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (see Appendix), but some measurement error remains.
+Second, we linked video tracking and clinical patient data using the times clinical attendees were at the registration and the entry times in the clinical database.
+False linkages could result when a tracking ID only passed by the registration (without registering) or when multiple IDs were at the registration closely before a clinical entry.
+Third, to facilitate modeling, we considered the waiting room and corridor as separate, although the door between both rooms was usually open (\zref{prep:building} in \supp).
+Fifth, our video sensors could not cover the whole area of the entrance and we did not collect data in the care rooms and other places of the clinic.
+As a result, the time spent in the clinic and the exposure to infectious quanta may be underestimated for some clinical attendees, although we expect the largest exposure to be in the waiting room, which was fully covered.
+
+% conclusions
+We conclude ...
 

### 2023-08-03 doc/paper/supplements.tex

@@
 \title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
-Spatiotemporal modeling of \emph{Mtb} transmission in a South African primary care clinic}}
+Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}
 
@@
 Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual.
-For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person\cite{Wang2021Science}, increasing the probability of infection.
+For example, the type of respiratory activity (\eg breathing, coughing, or sneezing) determines the site in the respiratory tract where the exhaled particles are generated, which in turn influences the particle size distribution\cite{Wei2016AMJIC}.
+Smaller particles reach deeper into the lungs of the susceptible person\cite{Wang2021Science}, which increases the probability of infection.
 
@@
 The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points.
-Formally, we estimate the cumulative risk of infection as
+Formally, we estimate the cumulative risk of infection for susceptible individual $a$ as
 \begin{equation}\label{eq:spattemp-P}
-    P = \sum_s \sum_t N_{s,t} \cdot p,
+    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
 \end{equation}
-where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), and $p$ is the breathing rate per person (in $m^3/s$).
+where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether $a$ was at location $s$ at time $t$, and $p$ is the breathing rate per person (in $m^3/s$).
 The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta:
@@
 
-\subsection{Building and setup}
+\subsection{Building and setup}\label{prep:building}
 
@@
 An analysis of exhalation activities suggest a propagation distance (\ie the maximum distance from the source to the end of the exhaled air before dispersion) of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity.
-Face masks reduce the propagation distance as most exhaled air (and particles) by locking the formation of the jet and leaking air on the sides of the masks\cite{Tang2009RoyalInt,Hui2012PLoSOne}.
+Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting particles (leakage of particles on the sides of the mask)\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}.
 Therefore, we assume that mask wearing confined the quanta generation only to the cell where the infectious individual is located.
@@
     \includegraphics{illustrations/prior-q.pdf}
-    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014) [REF], 1.25 by Riley (1962) [REF], 8.2 by Escombe (2008) [REF], and 13 by Nardell (1991) [REF].}
+    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014)\cite{Andrews2014JID}, 1.25 by Riley (1962)\cite{Riley1962ARRD}, 8.2 by Escombe (2008)\cite{Escombe2008PLoSMed}, and 13 by Nardell (1991)\cite{Nardell1991ARRD}.}
     \label{fig:quanta-distribution}
@@
 
-Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens} and thus lower the generation of infectious quanta.
+Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens,Leung2020NatMed} and should thus lower the generation of infectious quanta.
 The extent of the reduction is subject to uncertainty.
@@
     \includegraphics{illustrations/prior-lambda.pdf}
-    \caption{Prior distribution function for the viral inactivation rate along with reported estimates in the literature: 0.12 by Loudon (1969) [REF], 0.46 by Gannon (2007) [REF], 3.2 by Klein (2014) [REF], and 8.3 by Lever (2000) [REF].}
+    \caption{Prior distribution function for the viral inactivation rate for \emph{Mtb tuberculosis} along with reported estimates in the literature: 0.12 by Loudon (1969)\cite{Loudon1969AMRRD}, 0.46 by Gannon (2007) for \emph{Mtb bovis}\cite{Gannon2007ResVetSci}, 3.2 by Klein (2014)\cite{Klein2014IJMyco}, and 8.3 by Lever (2000)\cite{Lever2000LettersAppliedMicrobio}.}
     \label{fig:lambda-distribution}

### 2023-08-03 doc/paper/manuscript.tex

@@
 % TB and routes of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in South-East Asian and African region [TB report].
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in South-East Asian and African region\cite{WHO2022TBReport}.
 Progress made before the COVID-19 pandemic has stalled or reversed as the number of TB-related deaths have increased between 2019 and 2021 [TB report].
-\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious person [Rieder, Patterson].
-\emph{Mtb} is carried primarily in smaller particles $\leq5\mu$m called aerosols [Fennelly], which can survive in the air for multiple hours [Loudon].
-Airborne transmission is more likely in crowded, poorly ventilated indoor settings [Rieder, Center for Prevention Services, Nardell, Morawska, Wang].
-Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities, such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic [McCreesh 2020].
-A study in South Africa estimates that 4\% to 14\% of TB cases in adults originate from primary care clinics [McCreesh 2022].
+\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious person\cite{Rieder1999,Patterson2021Tuberculosis}.
+\emph{Mtb} is carried primarily in smaller particles $\leq5\mu$m called aerosols\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
+Airborne transmission is more likely in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
+Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities, such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic\cite{McCreesh2020IJTLD}.
+A study in South Africa estimates that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
 % the Wells-Riley model
-The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings [Andrews Integrating 2014, Taylor, Hella], including primary care clinics [Zuercher, McCreesh 2021].
+The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
 The Wells-Riley model assumes a well-mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (so-called quanta), the breathing rate per person, and the outdoor air supply rate.
@@
 Therefore, both short- and long-range transmission of \emph{Mtb} (and other pathogens) seem plausible.
-Nevertheless, previous studies show that the risk of infection is associated with proximity to an infectious person [Ko 2004, Kenyon], which is in line with the hypothesis that prolonged close contact is required for transmission [Leung, Brankston2007LancetID, Narasimhan].
+Nevertheless, previous studies show that the risk of infection is associated with proximity to an infectious person\cite{Ko2004RiskAnal,Kenyon1996NEJM}, which is in line with the hypothesis that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
 However, proximity to an infectious person is not considered in the Wells-Riley model.
@@
 The generation, dispersion, and survival of infectious particles in the air is subject to considerable uncertainty.
-Therefore, Wells [Wells 1955] introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission.
+Therefore, Wells\cite{Wells1955} introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission.
 A higher concentration of infectious quanta corresponds to a higher probability for an individual to get infected.
@@
 The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
-This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace [REFs].
+This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}.
 However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic.
-The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models [REFs].
+The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
 CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow.
-For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air [Li].
+For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}.
 However, airflow is rarely considered and it may be difficult to assess it in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
 
-(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components [Wang].
-Environmental factors, in particular temperature and humidity, can affect these properties but the effects are often unclear and depending on the pathogen [Songer, Chan, Cox, Fernstrom, Tang].
-Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear [Tang].
-On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity [Lever, Loudon].
-On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity [Sornboot, Matuka].
+(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science} [Wang].
+Environmental factors, in particular temperature and humidity, can affect these properties but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
+Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
+On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}.
+On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
 
-(3)~Previous studies show considerable variation in the generation of infectious quanta by TB infected patients [Escombe, Andrews].
+(3)~Previous studies show considerable variation in the generation of infectious quanta by TB infected patients\cite{Escombe2008PLoSMed,Andrews2014JID}.
 We model such variation by using a wide prior distribution for the quanta generation rate.
 Our model could be easily extended to incorporate prior information on the infectiousness of each individual TB patient.
-In general, only patients with active pulmonary TB are able to produce infectious droplets [Rieder].
+In general, only patients with active pulmonary TB are able to produce infectious droplets\cite{Rieder1999}.
 Infectiousness could be determines based on the most recent sputum test result.
-Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients [Shaw, Brindle, Grzybowski].
+Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.
 HIV status could further inform both infectiousness and susceptibility.
-On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum [Brindle, Telzak].
-On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status [Forte, Kwan, Shen], and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
+On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}.
+On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
 

### 2023-08-03 doc/paper/supplements.tex

@@
 The initial spatial spread of the quanta over the grid is informed prior literature.
-The concentration of virus-laden aerosols is highest closest to the source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}, in line with findings showing that the risk of infection correlates with the distance to the infectious individual [References].
+The concentration of virus-laden aerosols is highest closest to the source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}, in line with findings showing that the risk of infection correlates with the distance to the infectious individual\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
 Although coughing and sneezing can transport aerosols further away from the individual, these activities are much less frequent than breathing.

### 2023-08-03 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science} [Wang].
-Environmental factors, in particular temperature and humidity, can affect these properties but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
+(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
+Currently, these properties are only indirectly incorporated in our model by modeling the quanta generation rate as a stochastic parameter.
+It is also challenging to consider these properties because environmental factors, in particular temperature and humidity, can affect particle generation and survival, but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
 Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
@@
 In general, only patients with active pulmonary TB are able to produce infectious droplets\cite{Rieder1999}.
-Infectiousness could be determines based on the most recent sputum test result.
+Infectiousness could be determined based on the most recent sputum test result.
 Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.

### 2023-08-08 doc/paper/manuscript.tex
Updates from Overleaf

@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} ... \medskip
-\noindent\textbf{Methods and Findings:} ... 
-\medskip %\vspace{-1.3em}
+\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} was the leading cause of death from any single infectious disease before the COVID-19 pandemic, and progress made before the pandemic has stalled or reversed.
+TB is mostly transmitted through the airborne route and the risk of transmission is thus higher in crowded indoor settings, especially healthcare facilities where both more infectious and susceptible people gather.
+Modeling the risk of airborne transmission is difficult and existing approaches, such as the Wells-Riley model, assume that the airspace is well-mixed, so that the spatial location of infectious and susceptible individuals is not considered.
+\medskip
+
+\noindent\textbf{Methods:} We develop a spatiotemporal model for estimating the risk of infection that builds upon the Wells-Riley model.
+We apply our model to environmental (CO$_2$ levels), clinical data (patient visits and TB status), and patient movement data (anonymized trackings from video sensors) that were collected for 5~days between October and November 2021 (during the COVID-19 pandemic) in a primary care clinic in Cape Town, South Africa.
+We link tracking and clinical data to identify the spatiotemporal location of TB infectious patients inside the clinic.
+Incorporating prior assumptions about the generation, diffusion, and removal of infectious doses (so-called quanta), we perform Monte Carlo simulation to estimate the personal risk of infection for each clinical attendee.
+We compare our simulation results with a non-spatial variant of our model and assess the correlation with characteristics of the clinical visit (\eg time spent in the clinic and the number of close contacts).
+\medskip
+
+\noindent\textbf{Results:} ...
+
 \noindent\textbf{Conclusions:} ...
@@
 
-...
+\subsection{Study design}
+
+We build on the design of a pilot study in 2019\cite{Zurcher2022JID}, which was described in detail in a study protocol \cite{Zurcher2020BMJ}.
+In this follow-up study, we collected environmental data (indoor CO$_2$ levels), clinical (TB status), and tracking data (patient movements) on five days between October and November 2021 (October 13, 15, 25 and November 4, 5) at a primary care clinic in Cape Town, South Africa.
+Compared to the pilot study, we extended the patient tracking to include the TB treatment room.
+Molecular data was not part of the analysis of this study.
+
+\subsection{Study setting}
+
+The primary care clinic offers both TB and HIV services and reproductive health and childhood immunization services, Monday to Friday, from 7\,am to 4\,pm.
+The clinic is situated within a large settlement of formal and semiformal housing where both TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS} [Wood 2007, Middelkoop 2011].
+We delineated 3~areas within the clinic: the waiting room, the corridor, and the TB treatment room (Supplementary Figure~\zref{fig:floor-plan} in \supp).
+Furthermore, we defined two time periods: morning (7:00\,am to 12:00\,am) and afternoon (12:00\,am to 4:00\,pm).
+
+\subsection{Data}
+
+Supplementary Table~\zref{tab:data-overview} in \supp provides an overview of the collected data, which is described in the following.
+
+\subsubsection{Environmental data}
+
+Four devices monitored indoor CO$_2$ levels (Digital CO2 Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB treatment room (Supplementary Figure 1).
+CO$_2$ concentrations were recorded in parts per million (ppm) at 1-minute intervals (Supplementary Table 1).
+Missing values were linearly imputed, except for October~13 where the corridor's CO$_2$ levels were almost completely missing.
+Here we assumed the same levels as in the waiting room because both rooms had also similar CO$_2$ levels on the other days.
+
+\subsubsection{Clinical data}
+
+We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study period.
+These data included the date and time of arrival for the clinic visit, TB diagnostic results, and date of TB treatment start and (if applicable).
+We defined infectious patients as individuals testing positive for TB during the study or being on TB treatment since $\leq$28\,days.
+Patients who had completed a treatment were not considered infectious anymore.
+
+\subsubsection{Tracking data}
+
+We used an anonymized movement tracking system (Xovis, Zollikofen, Switzerland) to monitor people’s movements (staff members, patients, and other visitors) throughout the clinic (Supplementary Figure~\zref{fig:floor-plan} in \supp).
+The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
+Supplementary Figure \zref{fig:tracking-examples} in \supp shows a small, random sample of the tracks from our study.
+Note that individuals could contribute multiple tracks if they went out of a sensor’s range and subsequently returned or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person).
+To link these tracks, we created a R Shiny tool and manually linked tracks back together that most likely belonged to the same person (Supplementary Text describing and illustrating the tool and automatic pre-matching).
+We defined close contacts as other persons passing within a radius of $\leq$1\,m.
+
+\subsection{Spatiotemporal modeling}
+
+The workflow of our spatiotemporal modeling approach is shown in \Cref{fig:modeling-flow}.
+(1)~We combine patient tracking and clinical data to identify the spatiotemporal location of infectious individuals inside the clinic.
+We record the registration time for patients spending $\geq$5~seconds in the registration area, and then link their registration time with the nearest arrival time recorded in the clinical data, considering a maximum delay of 15~minutes.
+(2)~In addition to the number of TB patients masked from clinical data, we consider a similar number of unmasked TB patients among all other clinical attendees\cite{Berhanu2023CID}.
+Both masked and unmasked TB patients generate infectious quanta at their spatiotemporal location.
+The quanta generation rate $q$ is informed by estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD} and we apply a reduced rate to account for the effect of mask wearing\cite{Dharmadhikari2012AJRCCM}, which was mandatory in the clinic during the study period.
+(3)~We combine patient tracking and environmental data to monitor natural ventilation by computing the outdoor air exchange rate $AER$ from the number of people in the room and indoor CO$_2$ levels\cite{Batterman2017IJERPH}.
+(4)~The $AER$ is related to the diffusion constant $D$ (in m$^2$/s)\cite{Cheng2011EnvSciTech}, which determines the speed at which the infectious quanta spatially diffuses in the indoor air.
+We assume that the infectious quanta diffuses radially from the spot where it was generated.
+(5)~The $AER$ also corresponds to the rate of quanta removal through natural ventilation.
+The total removal further considers the viral inactivation rate $\lambda$, which is informed by prior literature\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+(6)~Quanta generation, diffusion, and removal determine the quanta concentration $N$ at time $t$ in the clinic, which is computed as
+\begin{align}\label{eq:spattemp-N}
+    \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
+concn.}} + \underbrace{I_t \cdot q}_{\text{generation}})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda)\right)}_{\text{removal}} ~.
+\end{align}
+where $\Delta$ is the Laplace operator (second-order differential operator) and $I_t$ is the number of infectious individuals in space.
+Finally, the patient-specific risk of infection depends on the cumulative exposure to infectious quanta during the clinical visit.
+We use the Wells-Riley Poisson relation to estimate the risk of TB transmission for each clinical attendee $a$ as
+\begin{align}
+    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
+\end{align}
+where $P$ is the probability of infection, $\mathbb{I}$ is a binary variable indicating whether attendee $a$ was at location $s$ at time $t$, and $p = 8$\,l/min is the volumetric breathing rate per person.
+The general modeling framework together is described together with an example in detail in Text~\zref{sec:spattemp-model} in \supp.
+
+\begin{figure}
+    \centering
+    \includegraphics{}
+    \caption{Flow chart illustrating the spatiotemporal modeling approach from data (blue) to the modeling (yellow).}
+    \label{fig:modeling-flow}
+\end{figure}
+
+We divide each separate area of the clinic (waiting room, corridor, TB room) into a grid of cubic cells with an area of 0.25m$^2$ and a volume depending on the height of the room (waiting and TB room: 3m; corridor: 2.5m).
+We update the quanta concentration every second and assume that $N_{s,0} = 0 ~~ \forall s$, \ie all quanta from the previous day has been removed from the air before the start of the following clinical day.
+Due to mask wearing, we assume that the initial spread of quanta before diffusion is limited to the cell where the infectious individual is located.
+We model all model parameters with prior distributions and perform Monte-Carlo simulation to estimate the risk of infection.
+In each simulation, we perform the following three steps: (i)~sampling uncertain modeling parameters and undiagnosed TB patients, (ii)~computing the spatiotemporal quanta concentration, and (iii)~computing the cumulative risk of infection for each clinical attendee.
+The model setup, including the specific assumptions and prior distributions, is described in detail in Text~\zref{sec:estimation} in \supp.
+
+\subsection{Statistical analysis}
+
+We report the time-varying CO$_2$ levels, the daily number of registered and masked TB patients, and the number of clinical attendees monitored through patient tracking.
+We perform 10,000 Monte Carlo simulations and show the average quanta concentration by daytime and summarize the risk of infection with the mean and 95\%-credible interval (CrI) across simulations.
+To demonstrate the impact of spatial modeling, we compare our modeling approach with one that assumes a well-mixed airspace (\ie where the initial quanta concentration is the same everywhere in the room regardless of where it was generated).
+Furthermore, we assess the correlation between the personal risk of infection and characteristics of the clinical visit (time spent in each area, number and duration of close contacts).
+To assess the effectiveness of mask wearing, we compare our modeling results with a scenario where there is no reduction in the generated quanta and the initial spread is extended to the first-neighbouring cells of where the infectious individual is located.
+All analyses were performed in R software (version 4.3.1)\cite{RCoreTeam2023}.
+
+\subsection{Ethics statment}
+
+The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 228/2019); the City of Cape Town (Project ID: 8139), South Africa; and the Ethics Committee of the Canton of Bern (2019-02131), Switzerland, approved the study.
 
@@
 
-% spatial modeling and airflow
+% critical appraisal of our results
 
@@
 For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered and it may be difficult to assess it in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
+However, airflow is rarely considered and it may be difficult to assess in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
 

### 2023-08-08 doc/paper/supplements.tex

@@
 
-The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)
+The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)\cite{Cheng2011EnvSciTech}
 \begin{align}
@@
 
-\section{Estimation of infection risk in the primary care clinic}
+\section{Estimation of infection risk in the primary care clinic}\label{sec:estimation}
 
@@
 In other words, we assume that all quanta from the previous day has been removed from the air before the start of the following clinical day.
-
-\subsection{Data and linkage}
 
@@
 This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic.
-Therefore,, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
+Therefore, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
 The effective quanta generation rate is thus $q$ times the reduction due to mask wearing.

### 2023-08-08 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-Supplementary Table~\zref{tab:data-overview} in \supp provides an overview of the collected data, which is described in the following.
-
 \subsubsection{Environmental data}
 
-Four devices monitored indoor CO$_2$ levels (Digital CO2 Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB treatment room (Supplementary Figure 1).
+Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB treatment room (Supplementary Figure 1).
 CO$_2$ concentrations were recorded in parts per million (ppm) at 1-minute intervals (Supplementary Table 1).
@@
 
-\begin{figure}
+\begin{figure}[!htpb]
     \centering
-    \includegraphics{}
-    \caption{Flow chart illustrating the spatiotemporal modeling approach from data (blue) to the modeling (yellow).}
+    \includegraphics{doc/paper/flow-chart.pdf}
+    \caption{Flow chart of the spatiotemporal modeling approach illustrating the steps from data (blue) to modeling (yellow).}
     \label{fig:modeling-flow}

### 2023-08-10 doc/paper/manuscript.tex
Updates from Overleaf

--- 2023-08-03 doc/paper/manuscript.tex
@@
 
-\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}
+\title{\bfseries\LARGE\singlespacing{Spatiotemporal risks of \emph{Mycobacterium tuberculosis} transmission: A modelling study of environmental, clinical, and patient movement data}}
 % author list
@@
 \author[1]{Kathrin Zürcher}
+\author[1]{Matthias Egger}
 \author[2]{Robin Wood}
-\author[1]{Matthias Egger}
 \author[1*]{Lukas Fenner}
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} ... \medskip
-\noindent\textbf{Methods and Findings:} ... 
-\medskip %\vspace{-1.3em}
+\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} was the leading cause of death from any single infectious disease before the COVID-19 pandemic, and progress made in TB infection and control before the pandemic has stalled or reversed.
+TB is strictly airborne and thus the transmission risk is higher in crowded indoor environments, especially healthcare facilities where both more infectious and susceptible people congregate.
+Modeling the risk of airborne transmission is difficult and existing approaches, such as the traditional Wells-Riley transmission model, assume that the airspace is well mixed, so that the spatial location of infectious and susceptible individuals is not taken into account.
+\medskip
+
+\noindent\textbf{Methods:} We developed a spatiotemporal model for estimating the risk of infection based on the Wells-Riley model.
+We applied our model to environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) that were collected for 5~days between October and November 2021 (during the COVID-19 pandemic) in a primary care clinic in Cape Town, South Africa.
+We linked tracking and clinical data to identify the spatiotemporal location of TB infectious patients within the clinic.
+Incorporating prior assumptions about the generation, diffusion, and removal of infectious doses (called quanta), we performed Monte Carlo simulations to estimate the personal risk of infection for each clinic attendee.
+We compare our simulation results with a non-spatial variant of our model and assess the correlation with characteristics of the clinical visit (\eg time spent at the clinic and the number of close contacts).
+\medskip
+
+\noindent\textbf{Results:} ...
+
 \noindent\textbf{Conclusions:} ...
@@
 
-% TB and routes of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in South-East Asian and African region\cite{WHO2022TBReport}.
-Progress made before the COVID-19 pandemic has stalled or reversed as the number of TB-related deaths have increased between 2019 and 2021 [TB report].
-\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious person\cite{Rieder1999,Patterson2021Tuberculosis}.
+% TB and the route of transmission
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South-East Asian and the African region\cite{WHO2022TBReport}.
+Progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed, as the number of TB-related deaths have increased between 2019 and 2021 while the number of reported TB diagnosis decreased [TB report].
+\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
 \emph{Mtb} is carried primarily in smaller particles $\leq5\mu$m called aerosols\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
-Airborne transmission is more likely in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities, such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic\cite{McCreesh2020IJTLD}.
-A study in South Africa estimates that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+Airborne transmission is more likely in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
+Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic\cite{McCreesh2020IJTLD}.
+A study in South Africa estimated that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
 % the Wells-Riley model
-The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
-The Wells-Riley model assumes a well-mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (so-called quanta), the breathing rate per person, and the outdoor air supply rate.
+The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
+The Wells-Riley model assumes a well mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (called quanta), the breathing rate per person, and the outdoor air supply rate.
 The quanta generation rate (commonly denoted with the parameter $q$) is unknown, but can be estimated by computing the risk of infection as the ratio of diseased/susceptible cases, and then solving the Wells-Riley equation for the parameter $q$\cite{Nardell1991ARRD,Escombe2008PLoSMed}.
@@
 % limitation of the Wells-Riley model
-A crucial assumption of the Wells-Riley model is that the airspace is well-mixed, which means that there is no spatial variation in the exposure to infectious doses.
+Traditional transmission modells such as the Wells-Riley model do not take into account spatial variations in the exposure to infectious doses.
 Although it is reasonable to assume that infectious particles disperse in the indoor space over time, the aerosol concentration is at first highest closest to the infectious person\cite{Vuorinen2020SafSci,Chen2020BuildEnv}.
-Therefore, both short- and long-range transmission of \emph{Mtb} (and other pathogens) seem plausible.
-Nevertheless, previous studies show that the risk of infection is associated with proximity to an infectious person\cite{Ko2004RiskAnal,Kenyon1996NEJM}, which is in line with the hypothesis that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
+Therefore, both short- and long-range airborne transmission of \emph{Mtb} (and other respiratory viruses) seem plausible.
+Nevertheless, previous studies showed that the risk of \emph{Mtb} infection is associated with proximity to an infectious person\cite{Ko2004RiskAnal,Kenyon1996NEJM}, which is in line with the hypothesis that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
 However, proximity to an infectious person is not considered in the Wells-Riley model.
@@
 We combine clinical and video tracking data to inform the spatiotemporal location of \emph{Mtb} infectious and susceptible individuals visiting a primary care clinic in South Africa between October and November 2021.
-Furthermore, we measure indoor CO$_2$ levels to model the diffusion and removal of infectious quanta over time.
-Our model allows us to identify spatiotemporal hotspots of quanta inside the clinic and estimate the risk of infection individually for each clinical attendee.
-We compare our modeling approach with an alternative specification that assumes a well-mixed airspace, thereby quantifying the contribution of close contact with infectious attendees to \emph{Mtb} transmission risk.
+Furthermore, we measure indoor CO$_2$ levels to model the diffusion (\ie spatial spread) and removal of infectious quanta over time.
+Our model allowed us to identify spatiotemporal hotspots of quanta within the clinic and to estimate the risk of infection individually for each clinic attendee.
+We compared our spatiotemporal modeling approach with an alternative model that assumed a well mixed airspace, thereby quantifying the contribution of close contact with infectious people to \emph{Mtb} transmission risk.
 
@@
 
-...
+\subsection{Study design}
+
+We build on the design of a pilot study in 2019\cite{Zurcher2022JID}, which was described in detail in a study protocol \cite{Zurcher2020BMJ}.
+In this follow-up study, we collected environmental data (indoor CO$_2$ levels), clinical (TB status), and tracking (patient movements) data on five days between October and November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
+Compared to the pilot study, we used the full patient tracking dataset, reflecting patient movements within the clinic, and included the TB treatment room in addition to the waiting room.
+
+\subsection{Study setting}
+
+The primary care clinic offers both TB and HIV services and other basic clinical services, Monday to Friday, from 7\,am to 4\,pm.
+The clinic is situated within a large settlement of formal and semiformal housing where both TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
+We defined 3~areas within the clinic: the waiting room, the corridor, and the TB treatment room (Supplementary Figure~\zref{fig:floor-plan} in \supp).
+Furthermore, we defined two time periods: morning (7:00\,am to 12:00\,am) and afternoon (12:00\,am to 4:00\,pm).
+
+\subsection{Data}
+
+\subsubsection{Environmental data}
+
+Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB treatment room (Supplementary Figure~\zref{fig:floor-plan} in \supp).
+CO$_2$ concentrations were recorded in parts per million (ppm) at 1-minute intervals.
+Missing values were linearly imputed, except for October~13 where the corridor's CO$_2$ levels were almost completely missing.
+Here we assumed the same levels as in the waiting room because both rooms had also similar CO$_2$ levels on the other days.
+
+\subsubsection{Clinical data}
+
+We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study period.
+These data included the date and time of arrival for the clinic visit, TB diagnostic results, and date of TB treatment start (if applicable).
+We defined infectious patients as individuals testing bacteriologically positive for TB during the study or being on TB treatment in the last $\leq$28\,days after treatment initiation.
+Patients who had recently completed a treatment were neither considered infectious nor susceptible.
+
+\subsubsection{Tracking data}
+
+We used an anonymous personal movement tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (staff members, patients, and other visitors) throughout the clinic (Supplementary Figure~\zref{fig:floor-plan} in \supp).
+The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
+Supplementary Figure \zref{fig:tracking-examples} in \supp~shows a small, random sample of the tracks from our study.
+Note that individuals could contribute multiple tracks if they moved out of the range of a sensor and then returned, or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person).
+To link these tracks, we created a R Shiny tool and manually linked tracks back together that most likely belonged to the same person (Supplementary Text~\zref{sec:setting-and-data} in \supp).
+We defined close contacts as other persons within a radius of $\leq$1\,m.
+
+\subsection{Spatiotemporal modeling}
+
+The workflow of our spatiotemporal modeling approach is shown in \Cref{fig:modeling-flow}.
+(1)~We combined patient tracking and clinical data to identify the spatiotemporal location of infectious individuals inside the clinic.
+We record the registration time for patients spending $\geq$5~seconds in the registration area, and then linked their registration time with the nearest arrival time recorded in the clinical data, considering a maximum delay of 15~minutes.
+(2)~In addition to the number of TB patients as reported in the clinical dataset, we also included a similar number of undiagnosed (unmasked) TB patients among all other clinical attendees\cite{Berhanu2023CID}.
+Both masked and unmasked TB patients generate infectious quanta at their spatiotemporal location.
+The quanta generation rate $q$ was informed by estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD}, and we applied a reduced rate to account for the effect of mask wearing\cite{Dharmadhikari2012AJRCCM}, which was mandatory in the clinic during the study period (3).
+Next we combined patient tracking and environmental data to monitor natural ventilation by calculating the outdoor air exchange rate $AER$ from the number of people in the room and the indoor CO$_2$ level\cite{Batterman2017IJERPH}.
+(4)~The $AER$ is related to the diffusion constant $D$ (in m$^2$/s)\cite{Cheng2011EnvSciTech}, which determines the speed at which the infectious quanta spatially spreads (diffuses) in the indoor air.
+We assumed that the infectious quanta diffuses radially from the spot where it was generated.
+(5)~The $AER$ also corresponds to the rate of quanta removal through natural ventilation.
+The total removal further considers the viral inactivation rate $\lambda$, which was informed by prior literature\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+(6)~Quanta generation, diffusion, and removal determined the quanta concentration $N$ at time $t$ in the clinic, which is computed as
+\begin{align}\label{eq:spattemp-N}
+    \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
+concn.}} + \underbrace{I_t \cdot q}_{\text{generation}})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda)\right)}_{\text{removal}} ~.
+\end{align}
+where $\Delta$ is the Laplace operator (second-order differential operator) and $I_t$ is the number of infectious individuals in space.
+Finally, the patient-specific risk of infection depends on the cumulative exposure to infectious quanta during the clinical visit.
+We use the Wells-Riley Poisson relation to estimate the risk of TB transmission for each clinical attendee $a$ as
+\begin{align}
+    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
+\end{align}
+where $P$ is the probability of infection, $\mathbb{I}$ is a binary variable indicating whether attendee $a$ was at location $s$ at time $t$, and $p = 8$\,l/min is the volumetric breathing rate per person.
+The general modeling framework together is described together with an example in detail in Text~\zref{sec:spattemp-model} in \supp.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/paper/flow-chart.pdf}
+    \caption{Flow chart of the spatiotemporal modeling approach illustrating the steps from data (blue) to modeling (yellow).}
+    \label{fig:modeling-flow}
+\end{figure}
+
+We divide each separate area of the clinic (waiting room, corridor, TB room) into a grid of cubic cells with an area of 0.25m$^2$ and a volume depending on the height of the room (waiting and TB room: 3m; corridor: 2.5m).
+We update the quanta concentration every second and assume that $N_{s,0} = 0 ~~ \forall s$, \ie all quanta from the previous day has been removed from the air before the start of the following clinical day.
+Due to mask wearing, we assume that the initial spread of quanta before diffusion is limited to the cell where the infectious individual is located.
+We model all model parameters with prior distributions and perform Monte-Carlo simulation to estimate the risk of infection.
+In each simulation, we perform the following three steps: (i)~sampling uncertain modeling parameters and undiagnosed TB patients, (ii)~computing the spatiotemporal quanta concentration, and (iii)~computing the cumulative risk of infection for each clinical attendee.
+The model setup, including the specific assumptions and prior distributions, is described in detail in Text~\zref{sec:estimation} in \supp.
+
+\subsection{Statistical analysis}
+
+We analyzed the time-varying CO$_2$ levels, the daily number of registered and masked TB patients, and the number of clinic attendees monitored through patient tracking.
+We perform 10,000 Monte Carlo simulations and show the average quanta concentration by daytime and summarize the risk of infection with the mean and 95\%-credible interval (CrI) across simulations.
+To demonstrate the impact of spatial modeling, we compare our modeling approach with one that assumes a well-mixed airspace (\ie where the initial quanta concentration is the same everywhere in the room regardless of where it was generated).
+Furthermore, we assess the correlation between the personal risk of infection and characteristics of the clinical visit (time spent in each area, number and duration of close contacts).
+To assess the effectiveness of mask wearing, we compare our modeling results with a scenario where there is no reduction in the generated quanta and the initial spread is extended to the first-neighbouring cells of where the infectious individual is located.
+All analyses were performed in R software (version 4.3.1)\cite{RCoreTeam2023}.
+
+\subsection{Ethics statement}
+
+The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 228/2019), the City of Cape Town (Project ID: 8139), South Africa, and the Ethics Committee of the Canton of Bern (KEK/REF: 2019-02131), Switzerland, approved the study.
 
@@
 First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors.
-We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (see Appendix), but some measurement error remains.
+We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (Supplementary Text~\zref{sec:setting-and-data} in \supp), but some measurement error remains.
 Second, we linked video tracking and clinical patient data using the times clinical attendees were at the registration and the entry times in the clinical database.

### 2023-08-10 doc/paper/supplements.tex

@@
 \newpage
+
+\section{Study setting and data}\label{sec:setting-and-data}
+
+\subsection{Building}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/paper/floor_plan_2021.pdf}
+    \caption{Floor plan showing the defined areas inside the clinic and the placement of the video sensors and CO$_2$ devices.}
+    \label{fig:floor-plan}
+\end{figure}
+
+\subsection{Preprocessing of tracking data}
+
+Tracking data required preprocessing before it could be used for spatiotemporal modeling.
+The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database.
+Clinical attendees most likely contributed multiple tracks for several reasons:
+\begin{itemize}
+    \item A person moved outside the range of the video sensors and then re-entered. 
+    \item An individual may temporarily not have been recognized as a person by the tracking software, \eg because the person bowed.
+    \item A person may have been temporarily hidden behind another person.
+    \item Moving objects such as opening doors may have been recognized as a person, adding a few noisy tracks into the data.
+\end{itemize}
+We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person.
+In the first step, we filtered noise (\ie IDs of less than 10s and a total moving distance of less than 2m) and automatically matched IDs that were less than 10s and 3m apart while moving (\ie IDs anywhere in the clinic) or less than 60s and 0.5m apart while sitting (\ie two IDs ending/starting in a seating area of the clinic).
+In the second step, we build on the automatically matched data and performed an additional manual linkage.
+The aim was to link IDs that were further apart but still most likely belonged together.
+Through manual inspection, we could better assess whether a link resulted in a movement that made sense and preclude better alternatives.
+A Shiny application and a matching manual/wiki (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) was developed to assist the students that performed the manual linkage.
+
+The Shiny app is shown in \Cref{fig:shiny-app}.
+In the right panel, a colored floor plan of the clinic is shown with the different areas delineated.
+Seating areas are marked in purple and exit areas are marked in blue.
+Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track.
+The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey.
+Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link.
+Possible links are sorted in ascending order by the time difference.
+
+In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information, \ie the name of the loaded file at the top and the path to the file where the updated data is automatically saved at the bottom, the date of the tracking, the initial number of IDs before any linkages and current number of IDs after linkages, and the number of links made.
+We created labels to denote finished tracks as "Ended" if the ID was continuously tracked throughout the clinic or "Lost" if the ID was somewhere lost in the clinic.
+The number of labels are shown below the number of IDs.
+Below the descriptive information are the tools related to performing linkages.
+First, the currently selected ID (black track) is shown, together with action buttons allowing students to quickly move to the first unfinished ID, the next or previous ID, or the previous or next ID starting at the entrance.
+Second, students could decide whether they want to match forward or backward in time.
+Third, the duration and number of links are shown for the current ID, which for example can alert students about IDs spending an unusually long time in the clinic.
+Fourth, the sliding bars allow students to filter possible links below a certain difference in time, distance, or height.
+Pre-set filters are added atop to quickly switch between common thresholds.
+Fifth, there is a drop-down menu to select IDs for which possible alternative links should be shown (light grey tracks).
+If there is a more plausible alternative, students could select the corresponding ID and check whether it is indeed the better match.
+Sixth, another drop-down menu below selects the ID which should be linked to the current patient ID (colored tracks) and the action "Link IDs" performs the link and updates the floor plan, table, and all other related information.
+If the students realized that the link was erroneous, they could unlink the last/first ID.
+Finally, once no further links could be made, either because the ID left the clinic or was lost, students could finish the track and label it accordingly.
+Ended tracks are removed from the set of possible links whereas lost tracks remain in the set.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/paper/shiny-tool-screenshot.png}
+    \caption{Screenshot of the Shiny app used for manual linking IDs that probably belong together.}
+    \label{fig:shiny-app}
+\end{figure}
+
+\clearpage
 

### 2023-08-10 illustrations/tracking-examples.Rmd
update tracking plots

---
title: "Tracking Examples"
author: "Nicolas Banholzer"
date: "2023-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(grid)
source("../utils/plotting.r")
```

## Data

```{r}
# building
clinic_img <- png::readPNG("../data-raw/Massi/building/clinic_clipped.png")
building <- terra::vect(paste0("../data-raw/", "Massi", "/building/clinic-vector.gpkg"))
building_sf <- sf::st_as_sf(building, crs = NA)
st_crs(building_sf) <- NA
building_df <- fortify(building_sf)

# files
files <- list.files("../data-clean/Massi/2021/patient-tracking", full.names = T)
files <- files[grepl("rds", files)]

# load data
df <- do.call(rbind, lapply(files, function(x) {
  d <- readRDS(x)
  if ("date_time" %in% colnames(d) & "time" %in% colnames(d)) {
    d <- d %>%
      dplyr::select(-time) %>%
      dplyr::select(patient_id, obs_id_new, obs_id, tracking_end, match_type, date_time, everything())
  } else if ("time" %in% colnames(d)) {
    d <- dplyr::rename(d, date_time = time)
  }
  return(d)
})) %>%
  mutate(patient_id = factor(patient_id),
         date = as.Date(date_time))

# subset of tracks
smpl <- df %>%
  filter(tracking_end == "Entered + Exited") %>%
  group_by(patient_id, date) %>%
  filter(n_distinct(obs_id_new) == 1) %>%
  summarize(duration = as.numeric(difftime(last(date_time), first(date_time)))) %>%
  ungroup() %>%
  filter(duration > 5) %>%
  mutate(selected = T) %>%
  filter(patient_id %in% c(8139, 1433, 339, 510, 1003))

df_smpl <- df %>%
  left_join(smpl) %>%
  filter(selected)

# start and end times of subset
df_smpl_se <- df_smpl %>%
  group_by(patient_id) %>%
  slice(1, n()) %>%
  mutate(type = factor(c("Start", "End"), levels=c("Start", "End"))) %>%
  ungroup()
```

## Plot

```{r}
building_pl <- ggplot(building_df) +
  geom_sf(linewidth = .5, alpha = 0, lwd = 0) +
  scale_fill_discrete(na.translate = F) +
  scale_x_continuous(breaks = seq(-20000, 20000, 1000)) +
  scale_y_continuous(breaks = seq(-20000, 20000, 1000)) +
  guides(fill = "none")

text_descr <- 10
en_lab <- textGrob(label = "Entrance", x = 0.1, y = 0.165, gp = gpar(size = text_descr / cm(1)), just = c("left"))
wr_lab <- textGrob(label = "Waiting room", x = 0.05, y = 0.45, gp = gpar(size = text_descr / cm(1)), just = c("left"))
tb_lab <- textGrob(label = "TB room", x = 0.75, y = 0.4, gp = gpar(size = text_descr / cm(1)), just = c("left"))
re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(size = text_descr / cm(1)), just = c("left"), rot = 90)

examples_pl <- building_pl +
  geom_path(data = df_smpl, mapping = aes(x = x, y = y, group = factor(patient_id), color = factor(patient_id)), alpha = .5) +
  geom_point(data = df_smpl_se, mapping = aes(x = x, y = y, group = factor(patient_id), color = factor(patient_id), shape = type)) +
  ggpubr::background_image(clinic_img) +
  #scale_y_continuous(expand = expansion(add=c(2000, 3250))) +
  #scale_x_continuous(expand = expansion(add=c(800, 0))) +
  scale_shape_manual(values = c(1,13)) +
  annotation_custom(en_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  theme_classic() +
  theme(text = element_text(size = text_descr),
        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.3, "cm"),
        legend.position = c(0.825,0.075), legend.direction = "horizontal",
        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), legend.title = element_blank())

examples_pl
```

### 2023-08-10 doc/paper/supplements.tex

@@
 
+\clearpage
+
+\subsection{Tracking data}
+
+\begin{figure}
+    \centering
+    \includegraphics{trac}
+    \caption{Caption}
+    \label{fig:enter-label}
+\end{figure}
+
+\clearpage
+
 \subsection{Preprocessing of tracking data}
 
-Tracking data required preprocessing before it could be used for spatiotemporal modeling.
+Tracking data required preprocessing before it could be fully used for spatiotemporal modeling.
 The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database.
@@
 The Shiny app is shown in \Cref{fig:shiny-app}.
-In the right panel, a colored floor plan of the clinic is shown with the different areas delineated.
-Seating areas are marked in purple and exit areas are marked in blue.
+In the right panel, a floor plan of the clinic is shown in (x,y)-space.
+Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
 Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track.
@@
     \centering
-    \includegraphics{doc/paper/shiny-tool-screenshot.png}
+    \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
     \caption{Screenshot of the Shiny app used for manual linking IDs that probably belong together.}

### 2023-08-14 illustrations/tracking-examples.Rmd
update references

@@
 building_pl <- ggplot(building_df) +
-  geom_sf(linewidth = .5, alpha = 0, lwd = 0) +
+  geom_sf(linewidth = .5, alpha = 0, color = NA) +
   scale_fill_discrete(na.translate = F) +
@@
 examples_pl
+save_plot(examples_pl, pdf_file = "tracking-examples.pdf", w = 16, h = 12)
 ```

### 2023-08-14 doc/paper/supplements.tex
Updates from Overleaf

@@
 
-\begin{figure}
+\begin{figure}[!htpb]
     \centering
-    \includegraphics{trac}
-    \caption{Caption}
-    \label{fig:enter-label}
+    \includegraphics{illustrations/tracking-examples.pdf}
+    \caption{Sample tracks from our video sensors of this study.}
+    \label{fig:tracking-examples}
 \end{figure}
@@
 
+\section{Discussion of factors affecting the risk of airborne infection}\label{sec:depth-discussion}
+
+The generation, dispersion, and survival of infectious particles in the air is subject to considerable uncertainty.
+Therefore, Wells\cite{Wells1955} introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission.
+A higher concentration of infectious quanta corresponds to a higher probability for an individual to get infected.
+The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
+Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~properties of the specific pathogenic, and the (3)~infectiousness of infected and susceptibility of uninfected TB patients.
+
+\subsection{Environmental factors}
+
+Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate outdoor air supply.
+The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
+This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}.
+However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic.
+The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
+CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow.
+For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}.
+However, airflow is rarely considered and it may be difficult to assess in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
+
+\subsection{Pathogen-specific factors}
+
+Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
+Currently, these properties are only indirectly incorporated in our model by modeling the quanta generation rate as a stochastic parameter.
+It is also challenging to consider these properties because environmental factors, in particular temperature and humidity, can affect particle generation and survival, but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
+Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
+On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}.
+On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
+
+\subsection{Patient-specific factors}
+
+Previous studies show considerable variation in the generation of infectious quanta by TB infected patients\cite{Escombe2008PLoSMed,Andrews2014JID}.
+We model such variation by using a wide prior distribution for the quanta generation rate.
+Our model could be easily extended to incorporate prior information on the infectiousness of each individual TB patient.
+In general, only patients with active pulmonary TB are able to produce infectious droplets\cite{Rieder1999}.
+Infectiousness could be determined based on the most recent sputum test result.
+Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.
+HIV status could further inform both infectiousness and susceptibility.
+On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}.
+On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
+
+% data and other limitations
+While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations.
+First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors.
+We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (Supplementary Text~\zref{sec:setting-and-data} in \supp), but some measurement error remains.
+Second, we linked video tracking and clinical patient data using the times clinical attendees were at the registration and the entry times in the clinical database.
+False linkages could result when a tracking ID only passed by the registration (without registering) or when multiple IDs were at the registration closely before a clinical entry.
+Third, to facilitate modeling, we considered the waiting room and corridor as separate, although the door between both rooms was usually open (Supplementary Text~\zref{sec:estimation} in \supp).
+Fifth, our video sensors could not cover the whole area of the entrance and we did not collect data in the care rooms and other places of the clinic.
+As a result, the time spent in the clinic and the exposure to infectious quanta may be underestimated for some clinical attendees, although we expect the largest exposure to be in the waiting room, which was fully covered.
+
+\clearpage
+
 \bibliography{references.bib}

### 2023-08-14 doc/paper/manuscript.tex

@@
 % the Wells-Riley model
-The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
+The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
 The Wells-Riley model assumes a well mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (called quanta), the breathing rate per person, and the outdoor air supply rate.
@@
 The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-Supplementary Figure \zref{fig:tracking-examples} in \supp~shows a small, random sample of the tracks from our study.
+Supplementary Figure~\zref{fig:tracking-examples} in \supp~shows a few sample tracks from our study.
 Note that individuals could contribute multiple tracks if they moved out of the range of a sensor and then returned, or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person).
@@
 
-% factors affecting quanta generation
-The generation, dispersion, and survival of infectious particles in the air is subject to considerable uncertainty.
-Therefore, Wells\cite{Wells1955} introduced the concept of a ``quantum'' (infectious dose) to capture the stochastic nature of airborne transmission.
-A higher concentration of infectious quanta corresponds to a higher probability for an individual to get infected.
-The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
-Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~properties of the specific pathogenic, and the (3)~infectiousness of infected and susceptibility of uninfected TB patients.
+% close contacts
+Our spatiotemporal model can be viewed as a modified version of the original Wells-Riley transmission model\cite{Riley1978AJE}.
+It is motivated by previous findings that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
+As a result, our model assigns a higher risk of infection to clinical attendees that spend more time in close proximity to infectious individuals.
+It could thus be used to assess the effectiveness of interventions targeting a reduction in the number of close contacts.
+Although such assessments only show the potential (modeled) impact of interventions, they can still provide a first estimate about how contact reductions translate into lower transmission risks in crowded indoor settings.
 
-(1)~Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate outdoor air supply.
-The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
-This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}.
-However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic.
-The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
-CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow.
-For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered and it may be difficult to assess in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
-
-(2)~Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
-Currently, these properties are only indirectly incorporated in our model by modeling the quanta generation rate as a stochastic parameter.
-It is also challenging to consider these properties because environmental factors, in particular temperature and humidity, can affect particle generation and survival, but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
-Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
-On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}.
-On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
-
-(3)~Previous studies show considerable variation in the generation of infectious quanta by TB infected patients\cite{Escombe2008PLoSMed,Andrews2014JID}.
-We model such variation by using a wide prior distribution for the quanta generation rate.
-Our model could be easily extended to incorporate prior information on the infectiousness of each individual TB patient.
-In general, only patients with active pulmonary TB are able to produce infectious droplets\cite{Rieder1999}.
-Infectiousness could be determined based on the most recent sputum test result.
-Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.
-HIV status could further inform both infectiousness and susceptibility.
-On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}.
-On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
-
-% data and other limitations
-While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations.
-First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors.
-We invested considerable effort to manually link individual tracking IDs probably belonging to the same attendee back together (Supplementary Text~\zref{sec:setting-and-data} in \supp), but some measurement error remains.
-Second, we linked video tracking and clinical patient data using the times clinical attendees were at the registration and the entry times in the clinical database.
-False linkages could result when a tracking ID only passed by the registration (without registering) or when multiple IDs were at the registration closely before a clinical entry.
-Third, to facilitate modeling, we considered the waiting room and corridor as separate, although the door between both rooms was usually open (\zref{prep:building} in \supp).
-Fifth, our video sensors could not cover the whole area of the entrance and we did not collect data in the care rooms and other places of the clinic.
-As a result, the time spent in the clinic and the exposure to infectious quanta may be underestimated for some clinical attendees, although we expect the largest exposure to be in the waiting room, which was fully covered.
+% more complex models and our relation
+Despite our modeling efforts, transmission events remain difficult to predict.
+Their occurrence depends on many other factors such as airflow, the properties of the pathogen, and variation in infectiousness of the source and susceptibility of the host\cite{Wang2021Science}.
+We discuss these factors in more depth in Supplementary Text~\zref{sec:depth-discussion} in \supp.
+Importantly, there is large stream of literature considering airflow using computational fluid dynamics (CFD) models to simulate the spatial spread of infectious aerosols in indoor environments, \eg Ref.
+\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
+CFD models can also be combined with the Wells-Riley model to estimate the risk of infection\cite{Yan2023BE,Qian2009BE,Li2022SOTTE}, thus are seemingly appealing to our setting as well.
+However, because they are computationally expensive, CFD models are often only used for experiments, short simulations, and static environments.
+They also require detailed knowledge about the indoor space, in particular airflow, to model particle transport.
+By contrast, our setting presents a dynamic environment with frequent patient movements over multiple days in a naturally ventilated space (varying airflow).
+Moreover, we run our model multiple times in a large number of Monte Carlo simulations to consider uncertainty in all modeling parameters and unmasked infectious TB patients.
 

### 2023-12-06 analysis/clinical-data.Rmd
complete preprocessing, first descriptive look into data, updating modeling assumptions

---
title: "Clinical Data"
author: "Nicolas Banholzer"
date: "2023-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)

source("../utils/plotting.r")
```

## Data

```{r}
df <- readRDS("../data-clean/clinical/clinical-data.rds") %>%
  filter(between(month(date), 10, 11))

modeled_days <- list.files("../data-clean/patient-tracking/linked-and-imputed")
modeled_days <- gsub("[[:punct:]]rds", "", modeled_days)
modeled_days <- as.Date(modeled_days)
```

## Descriptives

### Diagnosed TB over time

```{r}
diag_tb <- df %>%
  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
  group_by(date, `Date modeled`) %>%
  summarize(n_infectious = sum(is_infectious)) %>%
  ungroup()

diag_tb %>%
  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
  geom_bar(stat = "identity") +
  scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
  scale_y_continuous(expand = c(0,0)) +
  theme_custom() +
  theme(legend.position = "top")
```

```{r}
diag_tb %>%
  ggplot(aes(x = n_infectious)) +
  geom_histogram()

#' Assumption:
#' Number of diagnosed ~ number of diagnosed TB
#' --> Assume a Dirichlet prior with support according to the counts in the range (1,5).
```

### Suspected TB over time

```{r}
df %>%
  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
  group_by(date, `Date modeled`) %>%
  summarize(n_infectious = sum(was_suspected & !is_infectious)) %>%
  ungroup() %>%
  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
  geom_bar(stat = "identity") +
  scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
  scale_y_continuous(expand = c(0,0)) +
  theme_custom() +
  theme(legend.position = "top")
```

### 2023-12-06 analysis/environmental-data.Rmd
complete preprocessing, first descriptive look into data, updating modeling assumptions

---
title: "Environmental data"
author: "Nicolas Banholzer"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(lubridate)

source("../utils/plotting.r")
```

## Data

```{r}
dates <- sort(unique(cont_df$date))
abbr_dates <- paste(month(dates, label = T, abbr = T), day(dates))
cont_df <- readRDS("../data-clean/environmental/continuous-iaq-data.rds") %>%
  mutate(time = as.POSIXct(format(date_time, format = "%H:%M:%S"), format = "%H:%M:%S"),
         date = factor(date, levels = dates))
aer <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
  mutate(mth = month(date, label = T, abbr= T),
         dy = day(date),
         mthdy = paste(mth, dy),
         mthdy = factor(as.character(mthdy), levels = abbr_dates),
         daytime = factor(daytime, levels = c("Morning", "Afternoon")))
```

## Descriptives

### Summaries

```{r}
cont_df %>%
  group_by(date) %>%
  summarize(across(c(co2, temperature, humidity), .fns = list(mean, sd)))

cont_df %>%
  group_by(date, room) %>%
  summarize(across(c(co2), .fns = list(mean, sd))) %>%
  mutate_if(is.numeric, round) %>%
  ungroup() %>%
  arrange(room)
```

### CO2 over time

```{r}
cont_df %>%
  ggplot(aes(x = time, y = co2, color = date)) +
  geom_step() +
  facet_wrap(~ room) +
  coord_cartesian(expand = F) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
  scale_color_manual(values = wes_palette("Moonrise3")) +
  labs(y = "Number of people in waiting room") +
  theme_custom() +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank())
```

### Air change rate

```{r}
aer %>%
  ggplot(aes(x = mthdy, fill = daytime, y = aer)) +
  geom_bar(stat = "identity", position = position_dodge(width = 1, preserve = "single"), width = .66) +
  facet_wrap(~ room) +
  scale_fill_brewer(palette = 9) +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  theme_custom()
```

### 2023-12-06 analysis/patient-tracking.Rmd
complete preprocessing, first descriptive look into data, updating modeling assumptions

---
title: "Patient movements"
author: "Nicolas Banholzer"
date: "2023-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(lubridate)
library(parallel)
library(reshape2)

source("../utils/plotting.r")
source("../utils/spatial.r")

text_descr <- 8
```

## Data

### Tracking

```{r}
df <- readRDS("../data-clean/patient-tracking/augmented-and-combined-data.rds")
```

### Building

```{r}
source("../helper/building.R")
```

## Descriptives

### Tracking quality

```{r}
df %>%
  group_by(date, patient_id) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(tracking_end) %>%
  unlist() %>%
  table()

df %>%
  filter(tracking_end != "Possible HCW") %>%
  group_by(date, patient_id) %>%
  slice(1) %>%
  ungroup() %>%
  group_by(tracking_end) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(prop = round(n / sum(n) * 100))
```

### Time in clinic

```{r}
time_in_clin <- df %>%
  group_by(date, patient_id) %>%
  summarize(tracking_end = first(tracking_end),
            `Clinic` = n(),
            `Waiting room` = sum(is_waitingroom),
            `TB room` = sum(is_tbroom),
            `Corridor` = sum(is_passage)) %>%
  ungroup() %>%
  mutate(across(c(`Clinic`, `Waiting room`, `TB room`, `Corridor`), ~ .x / 60))

time_in_clin %>%
  ggplot(aes(x = tracking_end, y = total_duration)) +
  geom_boxplot()

time_in_clin %>%
  filter(tracking_end != "Possible HCW") %>%
  dplyr::select(-tracking_end) %>%
  melt(c("date", "patient_id")) %>%
  mutate(variable = factor(variable, levels = c("Clinic", "Waiting room", "TB room", "Corridor"))) %>%
  ggplot(aes(x = variable, fill = variable, y = value)) +
  geom_boxplot() +
  scale_y_sqrt()
```

### Close contacts

```{r}
# number of close contacts <1m for >1min
no_close_contacts <- df %>%
  unnest(cols = closeContacts) %>%
  group_by(date, patient_id, closeContacts, tracking_end) %>%
  summarise(n_time = n(),
            tracking_end = first(tracking_end)) %>%
  ungroup() %>%
  group_by(date, patient_id) %>%
  summarize(n_cc = sum(n_time > 60),
            tracking_end = first(tracking_end)) %>%
  ungroup()

time_close_contacts <- df %>%
  mutate(has_closeContact = ifelse(sapply(closeContacts, length) >= 1, T, F)) %>%
  group_by(date, patient_id) %>%
  summarize(cc_time = sum(has_closeContact) / 60,
            tracking_end = first(tracking_end)) %>%
  ungroup()

close_contact <- no_close_contacts %>%
  left_join(time_close_contacts) %>%
  left_join(time_in_clin)

close_contact %>%
  filter(tracking_end != "Possible HCW") %>%
  ggplot(aes(x = Clinic, y = n_cc)) +
  geom_point()

close_contact %>%
  filter(tracking_end != "Possible HCW") %>%
  ggplot(aes(x = 1, y = cc_time / Clinic)) +
  geom_boxplot() 
```

### Temporal hotspots

```{r}
clinic_occupancy <- df %>%
  group_by(date, time) %>%
  summarize(
    `Clinic` = n(),
    `Waiting room` = sum(is_waitingroom),
    `TB room` = sum(is_tbroom),
    `Corridor` = sum(is_passage)) %>%
  ungroup() %>%
  melt(c("date", "time")) %>%
  mutate(variable = factor(variable, levels = c("Clinic", "Waiting room", "TB room", "Corridor")),
         date = format(date, "%b %d"),
         time = as.POSIXct(format(time, format = "%H:%M:%S"), format = "%H:%M:%S")) %>%
  filter(between(hour(time), 7, 16))

clinic_occupancy %>%
  ggplot(aes(x = time, y = value, color = date)) +
  geom_step() +
  facet_wrap(~ variable) +
  coord_cartesian(expand = F) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
  scale_color_manual(values = wes_palette("Moonrise3")) +
  labs(y = "Number of people in waiting room") +
  theme_custom() +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank())
```

### Spatial hotspots

```{r}
col <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")

plot_spatial <- function(pl) {
  pl +
  ggpubr::background_image(clinic_img) +
  scale_y_continuous(expand = expansion(add=c(1700, 3000)), limits = c(-6500, 2000)) +
  scale_x_continuous(expand = expansion(add=c(400, 500)), limits = c(-16000, 7500)) +
  annotation_custom(entrance_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  theme_custom() +
  theme(text = element_text(size = text_descr),
        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
        legend.position = c(0.825,0.05), legend.direction = "horizontal", 
        legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
        panel.background = element_rect(fill = col[1]), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  guides(fill = guide_legend(title.position = "top"))
}

plot_spatial(
  ggplot(data = df, mapping = aes(x = x, y = y)) +
    geom_bin2d(binwidth = 200) ) +
  labs(fill = "Number of tracks") +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
                    breaks = c(1, 10, 100, 1000, 10000)) 
```

### Example tracks

```{r}
# unique id
df$day_pid <- paste(df$date, df$patient_id)

# random sample tracks
ex_pid <- df %>%
  filter(tracking_end != "Possible HCW") %>%
  group_by(day_pid) %>%
  slice(1) %>%
  ungroup() %>%
  group_by(tracking_end) %>%
  sample_n(3) %>%
  ungroup() %>%
  dplyr::select(day_pid) %>%
  unlist()
ex_df <- df %>% filter(day_pid %in% ex_pid)
ex_df_se <- ex_df %>%
  group_by(day_pid) %>%
  slice(1, n()) %>%
  mutate(type = factor(c("Start", "End"), levels=c("Start", "End"))) %>%
  ungroup()

# manual sample of tracks
ex_pid_man <- c(
  "2021-10-25 4649", # Entered + Exited, including TB room stay
  "2021-10-13 6297", # Lost Exit, only waiting room
  "2021-11-04 12860", # (Entered) + Exited, waiting room and corridor
  "2021-10-13 9346", # Entered + Exited, including TB room
  "2021-11-05 16625", # Entered + (Exited), only waiting room
  "2021-10-25 9526", # Lost Exit, waiting room and corridor
  "2021-11-04 9841", # Entered = (Exited), waiting room and corridor
  "2021-10-15 8644", # Entered + Exited, including TB room
  "2021-10-15 9419" # Lost Enter
)
ex_df_man <- df %>% filter(day_pid %in% ex_pid_man)
ex_df_man_se <- ex_df_man %>%
  group_by(day_pid) %>%
  slice(1, n()) %>%
  mutate(type = factor(c("Start", "End"), levels=c("Start", "End"))) %>%
  ungroup()

# plot
plot_spatial(
  ggplot() +
    geom_path(data = ex_df_man, mapping = aes(x = x, y = y), color = "darkred") +
    geom_point(data = ex_df_man_se, mapping = aes(x = x, y = y, shape = type), size = 2) +
    facet_wrap(~ day_pid) +
    scale_shape_manual(values = c(1,13)) 
) +
  theme(legend.position = "none")
```

### 2023-12-06 helper/building.R

@@ -1,0 +1,22 @@
+# background image
+clinic_img <- png::readPNG("../data-raw/building/clinic_clipped.png")
+
+# building shapes
+cellSize <- sqrt(1 * 1e9 / 3e3) 
+building <- terra::vect(paste("../data-raw", "building", "clinic-vector.gpkg", sep = "/"))
+building_sf <- sf::st_as_sf(building, crs = NA)
+sf::st_crs(building_sf) <- NA
+waiting_room <- shapeToSpatial(building_sf$geometry[2], cellSize)
+passage <- shapeToSpatial(building_sf$geometry[4], cellSize)
+tb_room <- shapeToSpatial(building_sf$geometry[3], cellSize)
+
+# room data frames
+waiting_room_df <- fortify(waiting_room)
+passage_df <- fortify(passage)
+tb_room_df <- fortify(tb_room)
+
+
+entrance_lab <- textGrob(label = "Entrance", x = 0.1, y = 0.165, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+wr_lab <- textGrob(label = "Waiting room", x = 0.05, y = 0.45, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+tb_lab <- textGrob(label = "TB room", x = 0.75, y = 0.4, gp = gpar(size = text_descr / cm(1)), just = c("left"))
+re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(size = text_descr / cm(1)), just = c("left"), rot = 90)

### 2023-12-06 preprocessing/Massi/2021/prep_clinical-data.R

@@ -1,42 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-library(lubridate)
-
-
-#### Helper ####
-
-
-#### Data ####
-
-# Masi
-
-df_pat_time_masi <- read.csv("data-raw/Massi/2021/clinical/patient-time.csv") %>%
-  rename(date = action_date) %>%
-  mutate(time = parse_date_time(paste(date, time), '%m/%d/%Y %I:%M:%S %p'),
-         date = as.Date(time)) %>%
-  filter(!is.na(time))
-
-df_tb_suspects <- read.csv("data-raw/Massi/2021/clinical/tb-suspects.csv") %>%
-  dplyr::select(patient_id, result_txt) %>%
-  rename(result = result_txt)
-
-df <- left_join(df_pat_time_masi, df_tb_suspects) %>%
-  rename(date_time = time) %>%
-  mutate(tb_suspect = ifelse(is.na(result), "no", "yes"),
-         time = format(date_time, "%H:%M:%S"),
-         date_time = as.POSIXct(as.character(date_time))) %>%
-  dplyr::select(patient_id, date, time, date_time, tb_suspect)
-
-saveRDS(df, "data-clean/Masi/clinical-data/clinical-data.rds")
-
-selected_dates <- unique(df$date)
-save_dir <- paste0("data-clean/Masi/combined-data/", selected_dates)
-for (i in 1:length(selected_dates)) {
-  if (!dir.exists(save_dir[i])) {
-    dir.create(save_dir[i])
-  }
-  df_sub <- df %>%
-    dplyr::filter(date == as.Date(selected_dates[i])) 
-  saveRDS(df_sub, paste0(save_dir[i], "/clinical-data.rds"))
-}

### 2023-12-06 preprocessing/Massi/2021/prep_environmental-data.R

@@ -1,288 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-library(lubridate)
-library(haven)
-
-source("utils/trans_risk.r")
-
-#### Backup data ####
-
-files <- list.files("data-raw/Massi/2021/environmental/backup/", full.names = T)
-
-masi2021 <- do.call(rbind,
-              lapply(files, function(f) {
-                read.table(f, sep = "\t", header = F, skip = 33) %>%
-                  set_names(c("date", "time", "humidity", "temperature", "co2", "ext")) %>%
-                  dplyr::select(-ext) %>%
-                  mutate(file = basename(f))
-              })) %>%
-  mutate(date_created = as.Date(stringi::stri_extract(file, regex = "\\d{8}"), "%Y%m%d"),
-         date = as.Date(date, format = "%m/%d/%Y"),
-         sensor = as.integer(gsub("-", "", stringi::stri_extract(file, regex = "-\\d{1}"))),
-         date_time = parse_date_time(paste(date, time), '%Y-%m-%d %I:%M:%S %p', tz = "CET"),
-         date_time = round_date(date_time, unit = "minutes"),
-         date = as.Date(date_time)) %>%
-  filter(sensor > 1) %>%
-  mutate(room = ifelse(sensor == 2, "Waiting room", ifelse(sensor == 3, "Corridor", "TB room")),
-         room = factor(room, levels = c("Waiting room", "Corridor", "TB room"))) %>%
-  dplyr::select(date, date_time, room, co2, temperature, humidity) %>%
-  mutate(across(c(co2, temperature, humidity), as.numeric))
-
-#### Preprocessed data from Kathrin ####
-
-masi2021_prep <- read_dta("data-raw/Massi/2021/environmental/co2_data.dta") %>%
-  rename(date = Date,
-         co2 = CO2,
-         temperature = Temperature,
-         humidity = Humidity,
-         room = location) %>%
-  filter(room > 1) %>%
-  mutate(room = ifelse(room == 2, "Waiting room", "TB room"),
-         room = factor(room, levels = c("Waiting room", "TB room")),
-         date_time = as.POSIXct(paste(date, format(new_time, format = "%H:%M:%S"))),
[…]

### 2023-12-06 preprocessing/Massi/match_clinical-tracking.R

@@ -1,96 +1,0 @@
-#### Libraries ####
-
-library(tidyverse)
-
-#### Matching function ####
-
-#' Matching clinical records to patient tracking data
-#' 
-#' @param clinic clinic
-#' @param sel_date date of tracking data
-#' @param min_time_in_cl minimum time spent in clinic in order for the id to be considered a clinical attendee
-#' @param min_time_at_re minimum time a track should have stayed at the reception
-#' @param min_duration_after_re minimum time a patient should have been in the clinic after she was at the reception
-#' @param max_tolerance maximum delay of clinical record after a patient was at the reception
-
-match_clinical_tracking <- function(clinic, sel_date, min_time_in_cl = 5 * 60, min_time_at_re = 5, min_duration_after_re = 5 * 60, max_tolerance = 15 * 60) {
-  
-  # clinical data
-  clinical <- readRDS(paste0("data-clean/", clinic, "/combined-data/", sel_date, "/clinical-data.rds")) %>%
-    group_by(patient_id) %>%
-    slice(1) %>%
-    ungroup() %>%
-    rename(clinical_patient_id = patient_id,
-           clinical_record_time = date_time)
-  
-  # tracking data
-  tracking <- readRDS(paste0("data-clean/", clinic, "/patient-tracking-data/", sel_date, "/", "/patient-id-data.rds")) %>%
-    group_by(patient_id) %>%
-    filter(as.numeric(difftime(last(time), first(time), units = "secs")) > min_time_in_cl) %>%
-    ungroup()
-  
-  # tracking: time at reception
-  track_time_at_re <- tracking %>%
-    group_by(patient_id) %>%
-    filter(sum(is_reception) > min_time_at_re) %>%
-    mutate(time_at_re = time[is_reception][1]) %>%
-    filter(as.numeric(difftime(last(time), time_at_re, units = "secs")) > min_duration_after_re) %>%
-    slice(1) %>%
-    ungroup() 
-  
-  # matching
-  merged <- merge(track_time_at_re %>% dplyr::select(patient_id, time_at_re),
-                  clinical %>% dplyr::select(clinical_patient_id, clinical_record_time, tb_suspect),
-                  id = NULL) %>%
-    mutate(timediff = as.numeric(difftime(clinical_record_time, time_at_re), units = "secs")) %>%
[…]

### 2023-12-06 preprocessing/_old/check_tracking.R

@@ -1,0 +1,35 @@
+#### Libraries ####
+
+library(tidyverse)
+
+
+#### Data ####
+
+files <- list.files("data-clean/Massi/2021/patient-tracking", full.names = T)
+ex_file <- files[3]
+df <- readRDS(ex_file)
+
+
+#### Check ####
+
+# stats
+statDF <- df %>%
+  group_by(patient_id) %>%
+  summarize(
+    duration = as.numeric(difftime(last(time), first(time), units = "min")),
+    was_at_reception = any(is_reception)
+  )
+
+# good IDs
+statDF_good <- statDF %>%
+  filter(duration > 5,
+         was_at_reception)
+
+# visualize
+df_good <- df %>%
+  filter(patient_id %in% statDF_good$patient_id)
+
+df_good %>%
+  filter(patient_id %in% sample(statDF_good$patient_id, size = 10)) %>%
+  ggplot(aes(x = x, y = y, color = factor(patient_id))) +
+  geom_path()

### 2023-12-06 preprocessing/_old/prep_xovis_matching.R

@@ -1,0 +1,597 @@
+#### Libraries ####
+
+library(tidyverse)
+options(dplyr.summarise.inform = FALSE)
+library(reshape2)
+library(lubridate)
+library(sf)
+library(sfheaders)
+library(raster)
+library(terra)
+library(gridExtra)
+library(VennDiagram)
+library(parallel)
+library(foreach)
+source("utils/spatial.r")
+source("utils/plotting.r")
+
+
+#' Algorith to match observation IDs
+#' 
+#' @param location string
+#' @param rooms fortified dataframe of spatial polygons of the rooms in the building
+#' @param date date %Y-%m-%d
+#' @param max_time maximum time (in seconds) to look ahead for matches
+#' @param min_standing_height minimum standing height
+#' @param max_mov_time maximum time (in seconds) to look ahead for people moving
+#' @param max_sit_time maximum time (in seconds) to look ahead for people sitting
+#' @param max_tbroom_time maximum time (in seconds) to look ahead for someone being in the TB room
+#' @param max_mov_distance maximum distance someone can make while moving
+#' @param max_sit_distance maximum distance someone can make while sitting
+#' @param max_heightdiff maximum height difference for a person not changing her posture
+#' @param min_duration minimum time (in minutes) to stay in the clinic to be considered
+
+match_xovis <- function(location,
+                        rooms,
+                        date, 
+                        max_time = 30 * 60, 
+                        min_standing_height = 1400, 
+                        max_mov_time = c(3, 5, 5, 5, 10, 10, 15, 15, 30, 60), 
+                        max_sit_time = c(5, 10, 20, 30, 45, 60, 90, 150, 300, 600),
+                        max_tbroom_time = 300,
+                        max_mov_distance = c(0.5, 0.5, 1, 2, 2, 3, 3, 3.5, 4, 4),
+                        max_sit_distance = c(0.25, 0.25, 0.5, 0.5, 0.5), 
+                        max_heightdiff = c(50, 50, 75, 100, 100, 125, 125, 150, 175, 200), 
+                        min_duration = 5 * 60
+                        ) {
+  
+  #### Data ####
+  
+  # xovis
+  file <- paste0("data-raw/", location, "/xovis/", date, ".rds")
+  save_dir <- paste0("data-clean/", location, "/patient-tracking-data/", date, "/")
+  if (!dir.exists(save_dir)) {
+    dir.create(save_dir)
+  }
+  masi <- readRDS(file) 
+  
[…]

### 2023-12-06 preprocessing/compute_close-contacts.R

@@ -1,0 +1,25 @@
+#### Libraries ####
+
+library(tidyverse)
+library(lubridate)
+library(parallel)
+
+source("utils/spatial.r")
+
+
+#### Data ####
+
+df <- do.call(rbind, lapply(list.files("data-clean/patient-tracking/linked-and-imputed/", full.names = T), readRDS)) %>%
+  mutate(date = as.Date(time))
+
+#### Functions ####
+
+close_contacts <- function(i, x_i, y_i, t_i, close_dist = 1) {
+  df_i <- df[df$patient_id != i & df$time == t_i, ]
+  d_i <- convert_dist(euclidean(x_i, df_i[["x"]], y_i, df_i[["y"]]))
+  df_i[["patient_id"]][d_i <= close_dist]
+}
+
+df$closeContacts <- mcmapply(close_contacts, df$patient_id, df$x, df$y, df$time, mc.cores = 6)
+
+saveRDS(df, file = "data-clean/patient-tracking/augmented-and-combined-data.rds")

### 2023-12-06 preprocessing/match_clinical-tracking.R

@@ -1,0 +1,118 @@
+#### Libraries ####
+
+library(tidyverse)
+
+#### Matching function ####
+
+#' Matching clinical records to patient tracking data
+#' 
+#' @param sel_date date of tracking data
+#' @param min_time_in_cl minimum time spent in clinic in order for the id to be considered a clinical attendee
+#' @param min_time_at_re minimum time a track should have stayed at the reception
+#' @param max_tolerance minimum and maximum delay of clinical record after a patient was at the reception
+#' @param prio_diagnosed should only diagnosed TB patients be matched?  
+#' @param prio_suspected should only suspected TB patients be matched? 
+
+match_clinical_tracking <- function(
+    sel_date, 
+    min_time_in_cl = 5 * 60, 
+    min_time_at_re = 5, 
+    max_tolerance = c(0, 15 * 60),
+    prio_diagnosed = T,
+    prio_suspected = T) {
+  
+  # clinical data
+  clinical <- readRDS(paste0("data-clean/clinical/clinical-data.rds")) %>%
+    filter(date == sel_date) %>%
+    rename(clinical_patient_id = patient_id,
+           clinical_record_time = date_time) %>%
+    group_by(clinical_patient_id) %>%
+    arrange(clinical_record_time) %>%
+    slice(1) %>%
+    ungroup()
+  
+  if (prio_diagnosed & prio_suspected) {
+    clinical <- filter(clinical, is_infectious | was_suspected)
+  } else if (prio_diagnosed) {
+    clinical <- filter(clinical, is_infectious)
+  } else if (prio_suspected) {
+    clinical <- filter(clinical, was_suspected)
+  }
+  
+  # tracking data
+  tracking <- readRDS(paste0("data-clean/patient-tracking/linked-and-imputed/", sel_date, ".rds")) %>%
+    group_by(patient_id) %>%
+    filter(as.numeric(difftime(last(time), first(time), units = "secs")) > min_time_in_cl) %>%
+    ungroup()
+  
+  # tracking: time at reception
+  tracking_at_re <- tracking %>%
+    group_by(patient_id) %>%
+    mutate(min_time_at_re = zoo::rollapply(is_reception, width = min_time_at_re, FUN = all, align = "right", fill = F)) %>%
+    ungroup() %>%
+    filter(min_time_at_re) %>%
+    group_by(patient_id) %>%
[…]

### 2023-12-06 preprocessing/prep_clinical.R

@@ -1,0 +1,56 @@
+#### Libraries ####
+
+library(tidyverse)
+library(lubridate)
+
+
+#### Data ####
+
+# Masi
+
+df_pat_time_masi <- read.csv("data-raw/clinical/patient-time.csv") %>%
+  rename(date = action_date) %>%
+  mutate(time = parse_date_time(paste(date, time), '%m/%d/%Y %I:%M:%S %p', tz = "CET"),
+         date = as.Date(time)) %>%
+  filter(!is.na(time))
+
+df_tb_pat <- readxl::read_excel("data-raw/clinical/CocT Masiphumelele_clinic data Apr_Nov 2021.xlsx", sheet = "Worksheet 1") %>%
+  rename(age = `Age(Years)`,
+         gender = Gender,
+         tb_treat_start = `Treatment Start Date`,
+         folder_number = `Folder Number`) %>%
+  dplyr::select(age, gender, tb_treat_start, folder_number) %>%
+  mutate(gender = factor(gender, levels = c("Male", "Female")),
+         tb_treat_start = as.Date(tb_treat_start),
+         folder_number = as.integer(folder_number))
+
+df_tb_sus <- read.csv("data-raw/clinical/tb-suspects.csv") %>%
+  mutate(was_suspected = T,
+         sus_tested_positive = ifelse(result_txt %in% c("TB_GENEXP_POS_RIF_S", "TB_GENEXP_POS_RIF_R"), T, F),
+         sus_started_treat = ifelse(care_status_txt %in% c("CARE_STATUS_STARTED", "CARE_STATUS_STARTED_OTHER"), T, F),
+         sus_treat_start = ifelse(sus_started_treat, care_status_date, NA),
+         sus_treat_start = as.Date(sus_treat_start, format = "%d.%m.%y")) %>%
+  dplyr::select(patient_id, was_suspected, sus_tested_positive, sus_started_treat, sus_treat_start)
+
+df <- df_pat_time_masi %>%
+  left_join(df_tb_pat) %>%
+  left_join(df_tb_sus) %>%
+  rename(date_time = time) %>%
+  mutate(diff_treat_start_1 = as.numeric(date - tb_treat_start),
+         diff_treat_start_2 = as.numeric(date - sus_treat_start),
+         is_infectious_1 = ifelse(diff_treat_start_1 <= 28, T, F),
+         is_infectious_2 = ifelse(diff_treat_start_2 <= 28, T, F),
+         is_infectious = is_infectious_1 | is_infectious_2,
[…]

### 2023-12-06 preprocessing/prep_environmental.R

@@ -1,0 +1,208 @@
+#### Libraries ####
+
+library(tidyverse)
+library(lubridate)
+library(haven)
+
+source("utils/trans_risk.r")
+
+#### Data ####
+
+# Backup data
+files <- list.files("data-raw/environmental/backup/", full.names = T)
+
+masi2021 <- do.call(rbind,
+              lapply(files, function(f) {
+                read.table(f, sep = "\t", header = F, skip = 33) %>%
+                  set_names(c("date", "time", "humidity", "temperature", "co2", "ext")) %>%
+                  dplyr::select(-ext) %>%
+                  mutate(file = basename(f))
+              })) %>%
+  mutate(date_created = as.Date(stringi::stri_extract(file, regex = "\\d{8}"), "%Y%m%d"),
+         date = as.Date(date, format = "%m/%d/%Y"),
+         sensor = as.integer(gsub("-", "", stringi::stri_extract(file, regex = "-\\d{1}"))),
+         date_time = parse_date_time(paste(date, time), '%Y-%m-%d %I:%M:%S %p', tz = "CET"),
+         date_time = round_date(date_time, unit = "minutes"),
+         date = as.Date(date_time)) %>%
+  filter(sensor > 1) %>%
+  mutate(room = ifelse(sensor == 2, "Waiting room", ifelse(sensor == 3, "Corridor", "TB room")),
+         room = factor(room, levels = c("Waiting room", "Corridor", "TB room"))) %>%
+  dplyr::select(date, date_time, room, co2, temperature, humidity) %>%
+  mutate(across(c(co2, temperature, humidity), as.numeric))
+
+# Preprocessed data from Kathrin
+masi2021_prep <- read_dta("data-raw/environmental/co2_data.dta") %>%
+  rename(date = Date,
+         co2 = CO2,
+         temperature = Temperature,
+         humidity = Humidity,
+         room = location) %>%
+  filter(room > 1) %>%
+  mutate(room = ifelse(room == 2, "Waiting room", "TB room"),
+         room = factor(room, levels = c("Waiting room", "TB room")),
+         date_time = as.POSIXct(paste(date, format(new_time, format = "%H:%M:%S"))),
+         across(c(co2, temperature, humidity), as.numeric)) %>%
[…]

### 2023-12-06 preprocessing/prep_tracking_activity.R

@@ -1,0 +1,30 @@
+#### Libraries ####
+
+library(tidyverse)
+source("utils/spatial.r")
+
+
+#### Data ####
+
+files <- list.files("data-clean/patient-tracking/matched-clinical/", full.names = T)
+files <- files[grepl("rds", files)]
+df_list <- lapply(files, readRDS)
+
+
+#### Activity ####
+#' Determine movement, assuming that less than 0.25m movement in 1s means the person is still
+#' sitting or standing, i.e. waiting. The 0.25 correspond to the cell size.
+#' 
+
+for (i in 1:length(df_list)) {
+  df <- df_list[[i]] %>%
+    group_by(patient_id) %>%
+    arrange(time) %>%
+    mutate(distance = euclidean(x, lag(x), y, lag(y))) %>%
+    fill(distance, .direction = "up") %>% 
+    ungroup() %>%
+    mutate(distance = convert_dist(distance),
+           activity = ifelse(distance <= .25, "waiting", "walking")) 
+  
+  saveRDS(df, files[[i]])
+}

### 2023-12-06 preprocessing/prep_tracking_impute.R

@@ -1,0 +1,259 @@
+#### Settings ####
+
+# libraries
+library(tidyverse)
+
+# settings
+load_path <- "data-clean/patient-tracking/finished-manual-linking/"
+save_path <- "data-clean/patient-tracking/linked-and-imputed/"
+
+# function to summarize tracking ends
+summarize.tracking_end <- function(df) {
+  df %>%
+    group_by(patient_id) %>%
+    slice(1) %>%
+    ungroup() %>%
+    dplyr::select(tracking_end) %>%
+    unlist() %>%
+    table()
+}
+
+# function to filter noise
+filter.noise <- function(df) {
+  df %>%
+    group_by(patient_id) %>%
+    filter(first(tracking_end) != "Noise") %>%
+    ungroup()
+}
+
+# exclude tracking end
+filter.exclude <- function(df, exclude = "Possible HCW") {
+  df.sub <- df %>%
+    group_by(patient_id) %>%
+    filter(!(first(tracking_end) %in% exclude)) %>%
+    ungroup()
+}
+
+# include tracking end
+filter.include <- function(df, include = "Entered + Exited") {
+  df.sub <- df %>%
+    group_by(patient_id) %>%
+    filter(first(tracking_end) %in% include) %>%
+    ungroup()
+}
+
+# plot duration distribution
+plot.duration <- function(df) {
+  df %>%
+    group_by(patient_id) %>%
+    summarize(duration = difftime(last(time), first(time), units = "min")) %>%
+    ungroup() %>%
+    ggplot(aes(x = duration)) +
+    geom_histogram()
+}
+
+# determine number of children
+table.minHeight <- function(df, min_height = 1400) {
+  df %>%
+    group_by(patient_id) %>%
+    summarize(height = max(height)) %>%
+    ungroup() %>%
+    mutate(below_min_height = ifelse(height <= 1400, "Small", "Large")) %>%
+    dplyr::select(below_min_height) %>%
+    unlist() %>%
+    table()
+} 
+
+# subset relevant variables
+subset <- function(df) {
+  dplyr::select(df, -obs_id_new, -obs_id, -match_type, -is_seat)
+}
+
+#' impute
+#' Assumption: between two linked tracks, we assume that the person stayed at the previous location
+
[…]

### 2023-12-06 tests/test_stm_v2.R

@@ -33,8 +33,10 @@ ct <- stm(c0 = c0, aer = aer, inf = inf, roomDim = roomDim, cellLength = gridCel
 # plot results
 ctDF <- do.call(rbind, lapply(seq_along(ct), function(t) reshape2::melt(ct[[t]]) %>% mutate(t = t) ))
 
-t_names <- c("08:00:00am", "08:01:00am", "08:10:00am", "08:30:00am", "09:00:00", "09:00:10", "09:00:30", "09:01:00", "09:05:00", "09:10:00", "10:00:00", "12:00:00")
-t_sel <- c(1, 60, 600, 1800, 3600, 3610, 3630, 3660, 3900, 4200, 7200, 14401)
+# t_names <- c("08:00:00am", "08:01:00am", "08:10:00am", "08:30:00am", "09:00:00", "09:00:10", "09:00:30", "09:01:00", "09:05:00", "09:10:00", "10:00:00")
+# t_sel <- c(1, 60, 600, 1800, 3600, 3610, 3630, 3660, 3900, 4200, 7200, 14401)
+t_names <- c("08:01:00", "08:30:00", "09:00:00", "09:01:00", "09:05:00", "10:00:00")
+t_sel <- c(60, 1800, 3600, 3660, 3900, 7200)
 names(t_names) <- t_sel
 
 quanta_concn_pl <- ctDF %>%
@@ -42,16 +44,25 @@ quanta_concn_pl <- ctDF %>%
   mutate(t = dplyr::recode(t, !!! t_names),
          value = value * (1 / (gridCellLength^2*3))) %>%
   ggplot(aes(x = Var2, y = Var1, fill = value)) +
-  facet_wrap(~ t) +
+  facet_wrap(~ t, ncol = 3) +
   geom_tile() +
-  labs(fill = expression("Concentration (10"^-2*" quanta/m"^3*")")) +
-  scale_fill_viridis_c(limits = c(0, NA), labels = function(x) x * 100) +
-  scale_x_continuous(expand = c(0,0), limits = c(0,40), labels = function(x) x * 0.25, breaks = seq(0,40,20)) +
-  scale_y_continuous(expand = c(0,0), limits = c(0,20), labels = function(x) x * 0.25, breaks = seq(0,20,10)) +
-  theme_bw2() +
-  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"),
-        axis.title = element_blank(), panel.spacing = unit(0.5, "cm"),
-        plot.margin = margin(r = 10))
[…]

### 2023-12-06 utils/plotting.R

@@ -2,9 +2,14 @@ require(patchwork)
 require(ggpattern)
 require(gridExtra)
 require(GGally)
+require(grid)
 require(Cairo)
 require(gg.layers)
+require(kunstomverse)
 require(wesanderson) # https://github.com/karthik/wesanderson
+
+# default colors
+bluegrey <- c(wes_palette("Royal1")[1], wes_palette("Darjeeling1")[5])
 
 # ggplot theme for Nature style
 text_size = 8

### 2023-12-06 utils/trans_risk.r

@@ -1,60 +1,101 @@
 #' Reduction in quanta emission rate through surgical mask wearing
 #' 
 #' @param n number of samples
-#' @param x reduction rate
+#' 
+#' @details 
+#' Distribution approximated based on the reported mean and confidence interval
+#' by Dharmadhikari for the reduction with face masks worn by patients with
+#' multi-drug resistant TB.
+#' 
+#' @references 
+#' - Dharmadhikari: Dharmadhikari et al. (2012) in Am J Respir Crit Care Med (doi.org/10.1164/rccm.201107-1190OC)
 
 rrmask <- function(n) {
   LaplacesDemon::rtrunc(n, spec = "norm", a = 0, b = 1, mean = 0.56, sd = 0.11)
 }
 
-drmask <- function(n) {
-  LaplacesDemon::dtrunc(x, spec = "norm", a = 0, b = 1, mean = 0.56, sd = 0.11)
-}
-
 
 #' Quanta emission rate
 #' 
+#' @param viral_load RNA viral or CFU bacillary load
+#' @param cf conversion factor to quanta
+#' @param activity physical activity (either waiting/sitting or walking)
+#' @param breathing_weight proportion of breathing (other: speaking)
+#' 
+#' @details 
+#' Mikszewski estimates the quanta generation rate distribution using 
+#' the inhalation rates (IR) for resting, standing, and light activity (slow walking) 
+#' from Adams (1993) and computes the droplet volume emission rate VER as the product
+#' of the IR and the droplet volume concentration Vd, using the droplet 
+#' emission rates from Stadnytskyi. Stadnytskyi estimates the transmission rates
+#' only for speaking, and (I assume) Mikszewski scales them to other respiratory 
+#' activities (breathing, loud speaking), using the multipliers derived 
+#' from the estimates for Vd from Morawska as done in Buonanno. 
+#' Accordingly, we also re-scale the VER reported by Mikszewski by multiplying them with 
+#' IR ratios for different physical activity levels and weight the VER by the proportion
+#' of breathing and speaking.  
+#' 
+#' @references 
[…]

### 2024-01-19 analysis/clinical-data.Rmd
revised model and new simulation function

@@
 modeled_days <- as.Date(modeled_days)
+modeled_days_abbrv <- format(modeled_days, format = "%b %d")
 ```
@@
 
-### Diagnosed TB over time
+### Diagnosed/Suspected TB over time
 
@@
 diag_tb <- df %>%
-  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
-  group_by(date, `Date modeled`) %>%
-  summarize(n_infectious = sum(is_infectious)) %>%
-  ungroup()
+  mutate(date_in_model = format(as.Date(date), format = "%b %d"),
+         date_in_model = ifelse(date_in_model %in% modeled_days_abbrv, date_in_model, "Not modeled"),
+         date_in_model = factor(date_in_model, levels = c(modeled_days_abbrv, "Not modeled"))) %>%
+  group_by(date, date_in_model) %>%
+  summarize(n_infectious = sum(is_infectious),
+            n_suspected = sum(was_suspected & !is_infectious)) %>%
+  ungroup() %>%
+  reshape2::melt(c("date", "date_in_model")) %>%
+  mutate(variable = ifelse(grepl("infectious", variable), "Diagnosed", "Suspected"),
+         variable = factor(variable, levels = c("Suspected", "Diagnosed")))
 
-diag_tb %>%
-  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
-  geom_bar(stat = "identity") +
+mean_diag <- round(median(diag_tb$value[diag_tb$variable=="Diagnosed"]), 2)
+lower_diag <- round(quantile(diag_tb$value[diag_tb$variable=="Diagnosed"], .25), 2)
+upper_diag <- round(quantile(diag_tb$value[diag_tb$variable=="Diagnosed"], .75), 2)
+annon_diag <- paste0("Median diagnosed = ", mean_diag, " (IQR ", lower_diag, " – ", upper_diag, ")")
+
+diag_tb_pl <- diag_tb %>%
+  ggplot(aes(x = date, y = value, fill = date_in_model, alpha = variable)) +
+  geom_bar(position = position_stack(), stat = "identity") +
+  annotate("text", x = as.Date("2021-11-04"), y = 12, label = annon_diag, size = 14/cm(1)) +
   scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
   scale_y_continuous(expand = c(0,0)) +
-  theme_custom() +
-  theme(legend.position = "top")
+  scale_fill_manual(values = c(wes_palette("Moonrise3"), "grey")) +
+  scale_alpha_discrete(breaks = c("Diagnosed", "Suspected"), range = c(.33, 1)) +
+  labs(y = "Number of patients",
+       title = "Daily number of TB patients attending the clinic",
+       subtitle = "Diagnosed and suspected (tested) patients in October and November") +
+  theme_custom(14) +
+  theme(legend.position = "bottom",
+        legend.title = element_blank(),
+        legend.direction = "vertical", 
+        legend.box = "vertical",
+        axis.title.x = element_blank(),
+        axis.text.x = element_blank()) +
+  guides(fill = guide_legend(nrow = 1, order = 2),
+         alpha = guide_legend(nrow = 1, order = 1))
+
+save_plot(diag_tb_pl, pdf_file = "../results/data/diagnosed-tb-patients.pdf", w = 18, h = 12)
 ```
@@
 #' --> Assume a Dirichlet prior with support according to the counts in the range (1,5).
+
+table(diag_tb$n_infectious)
+
+source("../utils/trans_risk.r")
+
+undiag_tb_pl <- data.frame(x = rundiag(1e5)) %>%
+  group_by(x) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  mutate(p = 100 * n / sum(n),
+         p = round(p)) %>%
+  ggplot(aes(x = factor(x), y = p)) +
+  geom_bar(stat = "identity", color = wes_palette("GrandBudapest1")[3], fill = wes_palette("GrandBudapest1")[4], alpha = .5) +
+  labs(x = "Number of undiagnosed TB patients", y = "Probability (%)",
+       title = "Undiagnosed TB patients among clinical attendees",
+       subtitle = "Assuming that no. undiagnosed similar to no. diagnosed",
+       caption = "Note: Distribution derived from empirical clinical data") +
+  coord_cartesian(expand = F) +
+  scale_y_continuous(limits = c(0,NA)) +
+  theme_custom(14) +
+  theme(plot.caption = element_text(size = 10))
+
+undiag_tb_pl
+save_plot(undiag_tb_pl, pdf_file = "../results/data/undiagnosed-tb-patients.pdf", w = 14, h = 12)
 ```
-
-### Suspected TB over time
-
-```{r}
-df %>%
-  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
-  group_by(date, `Date modeled`) %>%
-  summarize(n_infectious = sum(was_suspected & !is_infectious)) %>%
-  ungroup() %>%
-  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
-  geom_bar(stat = "identity") +
-  scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
-  scale_y_continuous(expand = c(0,0)) +
-  theme_custom() +
-  theme(legend.position = "top")
-```

### 2024-01-19 analysis/environmental-data.Rmd
revised model and new simulation function

@@
 ```{r}
-dates <- sort(unique(cont_df$date))
-abbr_dates <- paste(month(dates, label = T, abbr = T), day(dates))
+dates <- as.Date(c("2021-10-13", "2021-10-15", "2021-10-25", "2021-11-04", "2021-11-05"))
+abbr_dates <- format(dates, format = "%b %d")
 cont_df <- readRDS("../data-clean/environmental/continuous-iaq-data.rds") %>%
   mutate(time = as.POSIXct(format(date_time, format = "%H:%M:%S"), format = "%H:%M:%S"),
-         date = factor(date, levels = dates))
+         mthdy = factor(as.character(format(date, format = "%b %d")), levels = abbr_dates))
 aer <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
-  mutate(mth = month(date, label = T, abbr= T),
-         dy = day(date),
-         mthdy = paste(mth, dy),
-         mthdy = factor(as.character(mthdy), levels = abbr_dates),
+  mutate(mthdy = factor(as.character(format(date, format = "%b %d")), levels = abbr_dates),
          daytime = factor(daytime, levels = c("Morning", "Afternoon")))
@@
 ```{r}
-cont_df %>%
-  ggplot(aes(x = time, y = co2, color = date)) +
+co2_pl <- cont_df %>%
+  ggplot(aes(x = time, y = co2, color = mthdy)) +
   geom_step() +
@@
   coord_cartesian(expand = F) +
-  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
+  scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
   scale_color_manual(values = wes_palette("Moonrise3")) +
-  labs(y = "Number of people in waiting room") +
-  theme_custom() +
+  labs(y = expression("CO"[2]*" (parts per million, ppm)"),
+       title = "Carbon dioxide measurements",
+       subtitle = expression("CO"[2]*" levels over time per room")) +
+  theme_custom(14) +
   theme(axis.title.x = element_blank(),
-        legend.title = element_blank())
+        legend.title = element_blank(),
+        legend.position = "bottom")
+
+save_plot(co2_pl, pdf_file = "../results/data/co2-levels-over-time.pdf", w = 18, h = 12)
 ```
@@
 ```{r}
-aer %>%
-  ggplot(aes(x = mthdy, fill = daytime, y = aer)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 1, preserve = "single"), width = .66) +
-  facet_wrap(~ room) +
+aer_pl <- aer %>%
+  ggplot(aes(x = room, fill = daytime, y = aer)) +
+  geom_boxplot(position = position_dodge(width = .85)) +
+  #geom_bar(stat = "identity", position = position_dodge(width = 1, preserve = "single"), width = .66) +
+  #facet_wrap(~ room) +
   scale_fill_brewer(palette = 9) +
-  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
-  theme_custom()
+  scale_y_continuous(expand = expansion(mult = c(0, .05)), limits = c(0, NA)) +
+  labs(y = "Air change rate (changes per hour)",
+       title = "Ventilation conditions in the clinic",
+       subtitle = expression("Air exchange rate as computed from CO"[2])) +
+  theme_custom(14) +
+  theme(axis.title.x = element_blank(),
+        legend.title = element_blank(),
+        legend.position = c(0.8, 0.9))
+
+save_plot(aer_pl, pdf_file = "../results/data/air-changes-per-hour.pdf", w = 11, h = 12)
 ```

### 2024-01-19 analysis/patient-tracking.Rmd
revised model and new simulation function

@@
 source("../utils/spatial.r")
-
-text_descr <- 8
 ```
@@
 ```{r}
-source("../helper/building.R")
+cellSize = 250
+source("../preprocessing/prep_building-rasterize.R")
 ```
@@
 ```{r}
-df %>%
+track_qual <- df %>%
   group_by(date, patient_id) %>%
@@
   table()
+
+sum(track_qual[names(track_qual) != "Possible HCW"])
+track_qual_str <- paste0(names(track_qual), ": ", track_qual, " (", round(track_qual / sum(track_qual) * 100), "%)")
+track_qual_str
 
@@
 
-time_in_clin %>%
-  ggplot(aes(x = tracking_end, y = total_duration)) +
-  geom_boxplot()
-
-time_in_clin %>%
+time_in_clin_pl <- time_in_clin %>%
   filter(tracking_end != "Possible HCW") %>%
@@
   geom_boxplot() +
-  scale_y_sqrt()
+  scale_y_sqrt() +
+  scale_fill_manual(values = wes_palette("Royal2")) +
+  labs(y = "Time (minutes)", title = "Time spent in clinic", subtitle = "Duration of tracked clinical visit") +
+  theme_custom(14) +
+  theme(legend.position = "none",
+        legend.title = element_blank(),
+        axis.title.x = element_blank(),
+        plot.title.position = "plot")
+
+save_plot(time_in_clin_pl, pdf_file = "../results/data/time-in-clinic.pdf", w = 13, h = 12)
 ```
@@
 
-close_contact %>%
+cc_n_pl <- close_contact %>%
   filter(tracking_end != "Possible HCW") %>%
-  ggplot(aes(x = Clinic, y = n_cc)) +
-  geom_point()
+  ggplot(aes(x = n_cc))  +
+  scale_x_continuous(expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0)) +
+  labs(y = "Density", x = "Number of close contacts",
+       title = "Close contacts",
+       subtitle = "Defined as <1m for >1min") +
+  geom_density(fill = wes_palette("FantasticFox1")[2], color = wes_palette("FantasticFox1")[2], alpha = 0.8) +
+  theme_custom(14) +
+  theme(plot.title.position = "plot",
+        plot.margin = margin(r = 11),
+        plot.subtitle = element_text(margin = margin(b = 11)))
+cc_n_pl
 
-close_contact %>%
+cc_t_pl <- close_contact %>%
   filter(tracking_end != "Possible HCW") %>%
-  ggplot(aes(x = 1, y = cc_time / Clinic)) +
-  geom_boxplot() 
+  ggplot(aes(x = cc_time / Clinic * 100) ) +
+  scale_x_continuous(expand = c(0,0), limits = c(0, 100)) +
+  scale_y_continuous(expand = c(0,0)) +
+  labs(y = "Density", x = "Time in close contact (%)", title = "", subtitle = "") +
+  geom_density(fill = wes_palette("Moonrise1")[2], color = wes_palette("Moonrise1")[2], alpha = 0.8) +
+  theme_custom(14) +
+  theme(plot.title.position = "plot",
+        plot.margin = margin(r = 11))
+cc_t_pl
+
+cc_pl <- (cc_n_pl + cc_t_pl) + plot_layout(nrow = 1, heights = c(0.5, .5)) & 
+  plot_annotation(theme = theme(plot.background = element_rect(fill = "transparent", color = "transparent")))
+save_plot(cc_pl, pdf_file = "../results/data/close-contact.pdf", w = 16, h = 12)
 ```
@@
          date = format(date, "%b %d"),
+         date = factor(as.character(date), levels = c("Oct 13", "Oct 15", "Oct 25", "Nov 04", "Nov 05")),
          time = as.POSIXct(format(time, format = "%H:%M:%S"), format = "%H:%M:%S")) %>%
@@
 
-clinic_occupancy %>%
+time_pl <- clinic_occupancy %>%
   ggplot(aes(x = time, y = value, color = date)) +
@@
   coord_cartesian(expand = F) +
-  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
+  scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
   scale_color_manual(values = wes_palette("Moonrise3")) +
-  labs(y = "Number of people in waiting room") +
-  theme_custom() +
+  labs(y = "Number of people", title = "Occupancy over time", subtitle = "Number of active tracks per room over time") +
+  theme_custom(14) +
   theme(axis.title.x = element_blank(),
-        legend.title = element_blank())
+        legend.title = element_blank(),
+        legend.position = "bottom"
+        )
+
+save_plot(time_pl, pdf_file = "../results/data/no-people-over-time.pdf", w = 16, h = 12)
 ```
@@
 ```{r}
-col <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")
+spatial_pl <- plot_spatial(
+  ggplot(data = df, mapping = aes(x = x, y = y)) +
+    geom_bin2d(binwidth = 200),
+  text_descr = 14) +
+  labs(fill = "Number of tracks", title = "Spatial occupancy", subtitle = "Total number of tracks per grid cell") +
+  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
+                    breaks = c(1, 10, 100, 1000, 10000)) +
+  theme(legend.position = c(0.825,0.075), legend.direction = "horizontal") +
+  guides(fill = guide_legend(nrow = 2, title.position = "top"))
 
-plot_spatial <- function(pl) {
-  pl +
-  ggpubr::background_image(clinic_img) +
-  scale_y_continuous(expand = expansion(add=c(1700, 3000)), limits = c(-6500, 2000)) +
-  scale_x_continuous(expand = expansion(add=c(400, 500)), limits = c(-16000, 7500)) +
-  annotation_custom(entrance_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  theme_custom() +
-  theme(text = element_text(size = text_descr),
-        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
-        legend.position = c(0.825,0.05), legend.direction = "horizontal", 
-        legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
-        panel.background = element_rect(fill = col[1]), panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) +
-  guides(fill = guide_legend(title.position = "top"))
-}
-
-plot_spatial(
-  ggplot(data = df, mapping = aes(x = x, y = y)) +
-    geom_bin2d(binwidth = 200) ) +
-  labs(fill = "Number of tracks") +
-  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
-                    breaks = c(1, 10, 100, 1000, 10000)) 
+save_plot(spatial_pl, pdf_file = "../results/data/no-people-spatial.pdf", w = 16, h = 12)
 ```
@@
   "2021-11-05 16625", # Entered + (Exited), only waiting room
-  "2021-10-25 9526", # Lost Exit, waiting room and corridor
-  "2021-11-04 9841", # Entered = (Exited), waiting room and corridor
-  "2021-10-15 8644", # Entered + Exited, including TB room
+  #"2021-10-25 9526", # Lost Exit, waiting room and corridor
+  #"2021-11-04 9841", # Entered = (Exited), waiting room and corridor
+  #"2021-10-15 8644", # Entered + Exited, including TB room
   "2021-10-15 9419" # Lost Enter
@@
 # plot
-plot_spatial(
+ex_tracks <- plot_spatial(
   ggplot() +
@@
     geom_point(data = ex_df_man_se, mapping = aes(x = x, y = y, shape = type), size = 2) +
-    facet_wrap(~ day_pid) +
-    scale_shape_manual(values = c(1,13)) 
+    facet_wrap(~ day_pid, ncol = 3) +
+    scale_shape_manual(values = c(1,13)),
+  text_descr = 10
 ) +
-  theme(legend.position = "none")
+  labs(title = "Sample of invidiual tracks", subtitle = "Illustrative examples of the mixed tracking quality") +
+  theme(legend.position = "none",
+        plot.title = element_text(size = 16),
+        plot.subtitle = element_text(size = 14))
+
+save_plot(ex_tracks, pdf_file = "../results/data/example-patient-tracks.pdf", w = 22, h = 12)
 ```

### 2024-01-19 helper/building.R

@@ -1,22 +1,0 @@
-# background image
-clinic_img <- png::readPNG("../data-raw/building/clinic_clipped.png")
-
-# building shapes
-cellSize <- sqrt(1 * 1e9 / 3e3) 
-building <- terra::vect(paste("../data-raw", "building", "clinic-vector.gpkg", sep = "/"))
-building_sf <- sf::st_as_sf(building, crs = NA)
-sf::st_crs(building_sf) <- NA
-waiting_room <- shapeToSpatial(building_sf$geometry[2], cellSize)
-passage <- shapeToSpatial(building_sf$geometry[4], cellSize)
-tb_room <- shapeToSpatial(building_sf$geometry[3], cellSize)
-
-# room data frames
-waiting_room_df <- fortify(waiting_room)
-passage_df <- fortify(passage)
-tb_room_df <- fortify(tb_room)
-
-
-entrance_lab <- textGrob(label = "Entrance", x = 0.1, y = 0.165, gp = gpar(size = text_descr / cm(1)), just = c("left"))
-wr_lab <- textGrob(label = "Waiting room", x = 0.05, y = 0.45, gp = gpar(size = text_descr / cm(1)), just = c("left"))
-tb_lab <- textGrob(label = "TB room", x = 0.75, y = 0.4, gp = gpar(size = text_descr / cm(1)), just = c("left"))
-re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(size = text_descr / cm(1)), just = c("left"), rot = 90)

### 2024-01-19 models/stm-v2.R

@@ -2,19 +2,17 @@
 #' 
 #' @description Function to compute the quanta concentration at each time point. 
 #' 
-#' @param c0 matrix with initial quanta concentration
-#' @param aer data frame with columns time and aer (air change rate in air changes per hour)
-#' @param inf data frame with columns time, x and y indicating the cell where the infectious person is located
-#' @param roomDim room dimension vector (length x, length y, height)
-#' @param cellLength length of a grid square cell (in m3)
+#' @param time time vector in 1s
+#' @param c0 2d-matrix with initial quanta concentration (in quanta per m3)
+#' @param inf data frame with time, x- and y-coordinates, and generated quanta q
+#' @param cellLength length of a grid square cell (in m)
 #' @param pd propagation distance of emitted quanta (in number of neighboring cells)
-#' @param q random generation distribution function for the emitted quanta (quanta per hour)
-#' @param lambda random generation distribution function for the viral inactivation rate (per hour)
-#' @param k random generation distribution function for the deposition rate due to gravitational settling (per hour)
-#' @param seed seed number
+#' @param vol room volume (in m3)
+#' @param aer air changes per second
+#' @param lambda viral inactivation rate per second
+#' @param deposit deposition rate per second due to gravitational settling
 #' 
-#' @details The default for parameters q, lambda, and k is based on prior research for Mtb. 
-#' The diffusion constant is computed relative to the air change rate, see Cheng (2011).
+#' @details The diffusion constant is computed relative to the air change rate, see Cheng (2011).
 #' 
 #' @references Cheng KC, Acevedo-Bolton V, Jiang RT, Klepeis NE, Ott WR, Fringer OB, Hildemann LM. 
 #' Modeling exposure close to air pollution sources in naturally ventilated residences: 
@@ -24,74 +22,94 @@
[…]

### 2024-01-19 preprocessing/prep_building-rasterize.R

@@ -1,0 +1,87 @@
+# libraries
+library(tidyverse)
+library(png)
+library(terra)
+library(sf)
+library(grid)
+
+if (file.exists("../utils/spatial.r")) {
+  source("../utils/spatial.r")
+  clinic_img <- png::readPNG("../data-raw/building/clinic_clipped.png")
+  building <- terra::vect(paste("../data-raw", "building", "clinic-vector.gpkg", sep = "/"))
+} else {
+  source("utils/spatial.r")
+  clinic_img <- png::readPNG("data-raw/building/clinic_clipped.png")
+  building <- terra::vect(paste("data-raw", "building", "clinic-vector.gpkg", sep = "/"))
+}
+
+building_sf <- sf::st_as_sf(building, crs = NA)
+sf::st_crs(building_sf) <- NA
+
+# rasterize
+
+if (!exists("cellSize")) {
+  stop("No local cellSize variable found")
+} 
+
+rasterize.rooms <- function(cellSize) {
+  waiting_room <- shapeToSpatial(building_sf$geometry[2], cellSize)
+  passage <- shapeToSpatial(building_sf$geometry[4], cellSize)
+  tb_room <- shapeToSpatial(building_sf$geometry[3], cellSize)
+  return(list(waiting_room = waiting_room, corridor = passage, tb_room = tb_room))
+}
+
+rooms <- rasterize.rooms(cellSize)
+waiting_room <- rooms$waiting_room
+wrMat <- sP_to_matrix(waiting_room)
+wrCoord <- create_coord_df(waiting_room)
+corridor <- rooms$corridor
+cdMat <- sP_to_matrix(corridor)
+cdCoord <- create_coord_df(corridor)
+tb_room <- rooms$tb_room
+tbMat <- sP_to_matrix(tb_room)
+tbCoord <- create_coord_df(tb_room)
+roomCoord <- rbind(wrCoord %>% mutate(room = "Waiting room"), 
+                   cdCoord %>% mutate(room = "Corridor"), 
+                   tbCoord %>% mutate(room = "TB room")) 
+
+# room dimensions
+# - waiting room
+dimWR <- c(10.55, 5.7, 3) # (length x width x height in m)
+# - corridor
+dimCD <- c(7.7, 2.2, 2.5)
+# - TB room
+dimTB <- c(4.75, 3.5, 3)
+
+# plotting
+
+plot_spatial <- function(pl, text_descr = 10) {
+  
[…]

### 2024-01-19 preprocessing/prep_simulation-data.R

@@ -1,0 +1,221 @@
+#### Libraries ####
+
+library(tidyverse)
+library(parallel)
+library(lubridate)
+
+source("utils/spatial.r")
+source("utils/trans_risk.r")
+
+
+#### Settings ####
+
+# number of simulations
+n <- 1e5
+
+# cell grid size in px
+cellSize <- 250
+
+# Seed
+seed <- 12345
+
+# building
+source("preprocessing/prep_building-rasterize.R")
+
+# who is assumed to have tb
+#' du: diagnosed and undiagnosed in proportion
+#' ds: diagnosed and suspected
+#' pu: sample among all clinical attendees in proportion to population prevalence
+who_is_tb <- c("du", "ds", "pu")
+
+# path to store cellSize dependend tracking data
+track_file_path <- paste0("simulations/", "/", "tracking-df-", cellSize, "px-")
+
+# dates
+dates <- paste0("2021-", c(10, 10, 10, 11, 11), "-", c("13", "15", "25", "04", "05"))
+daytimes <- c("both", "morning", "both", "both", "both")
+
+# number of cores
+ncores <- 6
+
+
+#### Track data ####
+
+for (i in 1:length(dates)) {
+  
+  # date
+  d <- dates[i]
+  
+  # daytime
+  dytm <- daytimes[i]
+  t_start <- as.POSIXct(paste(d, "08:00:00"))
+  t_end <- as.POSIXct(paste(d, "16:00:00"))
+  tt <- seq(t_start, t_end, by = "1 sec")
+  midday <- as.POSIXct(paste(d, "12:00:00"))
+  midday_idx <- which(tt == midday)
+  
+  # existing track file
+  track_file <- paste0(track_file_path, d, ".rds")
+  
+  # tracking
+  if (!file.exists(track_file)) {
+    
+    # load data
+    track <- readRDS(paste0("data-clean/patient-tracking/matched-clinical/", d, ".rds")) %>%
+      mutate(activity = ifelse(activity == "waiting", 1, 2))
+    
+    # filter daytime
+    if (dytm == "both") {
+      track <- filter(track, between(time, t_start, t_end)) 
+    } else if (dytm == "morning") {
+      track <- filter(track, between(time, t_start, midday)) 
+    } else if (dytm == "afternoon") {
+      track <- filter(track, between(time, midday, t_end)) 
+    } else {
+      stop("Unknown daytime")
+    }
+    
+    # determine cell
+    track$cell_id <- NA
[…]

### 2024-01-19 preprocessing/simulate_qctr.R

@@ -1,0 +1,344 @@
+#### Command line ###
+
+# all command line arguments
+args <- commandArgs(trailingOnly=TRUE)
+
+# start simulation
+#' e.g. if the previous simulation was interrupted
+#' 1 means to start from beginning
+cont_n <- as.numeric(args[1])
+
+# end simulation
+term_n <- as.numeric(args[2])
+
+# month
+mth <- args[3]
+
+# day
+dy <- args[4]
+
+# analysis name
+aname <- args[5]
+adir <- paste0("simulations/", aname, "/")
+if (!dir.exists(adir)) {
+  dir.create(adir)
+}
+
+# grid cell length
+cellSize <- as.numeric(args[6]) # in px
+
+#' without mask wearing
+#' 0: without mask-wearing
+#' 1: with masking
+is_masking <- as.numeric(args[7]) 
+
+#' fixed air chnage rate
+#' >0: fixed AER replacing empirical one(s)
+#' 0: use empirical one(s)
+fixed_aer <- as.numeric(args[8]) 
+
+#' population-level prevalence
+#' du: use diagnosed and undiagnosed TB patients
+#' ds: use diagnosed and suspsected TB patients
+#' pu: use population-level prevalence to sample from undiagnosed TB patients
+#' for population-level prevalence in SA, see Moyo et al, 2022, Lancet ID, doi:10.1016/S1473-3099(22)00149-9
+who_is_tb <- args[9]
+
+#' type of model
+#' temp: temporal model (no spatial diffusion, assuming well-mixed airspace)
+#' spattemp: spatiotemporal model (including spatial diffusion, no well-mixed airspace)
+mod <- args[10]
+
+#' daytime
+#' TODO: Should the number of undiagnosed patients be adpated if the clinic was only open for half a day?
+
+#' single room
+#' clinic: all rooms
+#' war: waiting room
+#' tbr: tb room
+#' cor: corridor
+sel_rooms <- args[11]
+if (sel_rooms == "clinic") {
+  sel_rooms <- c("Waiting room", "TB room", "Corridor")
+} else if (sel_rooms == "war") {
+  sel_rooms <- c("Waiting room")
+} else if (sel_rooms == "tbr") {
+  sel_rooms <- c("TB room")
+} else if (sel_rooms == "cor") {
+  sel_rooms <- c("Corridor")
+} else {
+  stop("Invalid rooms argument.")
+}
+
+
+# file containing simulation parameters
[…]

### 2024-01-19 tests/test_stm_v2.R

@@ -11,51 +11,65 @@ ny <- 20
 dy <- gridCellLength * ny
 dx <- gridCellLength * nx
 roomDim <- c(dx, dy, dz)
-c0 <- matrix(0, nrow = ny, ncol = nx)
+c0 <- matrix(0, nrow = ny, ncol = nx, dimnames = list(as.character(1:ny), as.character(1:nx)))
 
-# assume constant air exchange rate
-aer <- data.frame(
-  time = seq(as.POSIXct("2023-01-01 08:00:00"), as.POSIXct("2023-01-01 12:00:00"), by = "1 sec"),
-  aer = 1
-)
+# time 
+time <- seq(as.POSIXct("2023-01-01 08:00:00"), as.POSIXct("2023-01-01 10:00:00"), by = "1 sec")
+
+# assumptions
+aer <- 3 / 3600
+lambda <- 0 / 3600
+deposit <- 0
 
 # assume one infectious person in the middle of the room at the beginning staying there for one hour
 inf <- data.frame(
-  time = seq(as.POSIXct("2023-01-01 08:00:00"), as.POSIXct("2023-01-01 09:00:00"), by = "1 sec"),
-  x = 20,
-  y = 10
+  t = 1:(length(time)/2),
+  x = "20",
+  y = "10",
+  q = 10 / 3600
 )
 
 
 # run model assuming no masks (pd = 1)
-ct <- stm(c0 = c0, aer = aer, inf = inf, roomDim = roomDim, cellLength = gridCellLength, pd = 1, q = 10, lambda = 0.5, k = 0, seed = 12345)
+ct <- stm(time = time, c0 = c0, aer = aer, inf = inf, vol = prod(roomDim), cellLength = gridCellLength, pd = 0, lambda = lambda, deposit = deposit)
 
 # plot results
-ctDF <- do.call(rbind, lapply(seq_along(ct), function(t) reshape2::melt(ct[[t]]) %>% mutate(t = t) ))
-
-# t_names <- c("08:00:00am", "08:01:00am", "08:10:00am", "08:30:00am", "09:00:00", "09:00:10", "09:00:30", "09:01:00", "09:05:00", "09:10:00", "10:00:00")
-# t_sel <- c(1, 60, 600, 1800, 3600, 3610, 3630, 3660, 3900, 4200, 7200, 14401)
+ct_to_df <- function(t) {
+  data.frame(ct[,,t]) %>% 
+    add_rownames() %>% 
[…]

### 2024-01-19 utils/plotting.R

@@ -24,19 +24,22 @@ theme_nature <- function () {
     )
 }
 
-theme_custom <- function() {
+theme_custom <- function(text_size = 8) {
   theme_minimal() %+replace% 
     theme(text = element_text(size = text_size),
           axis.text = element_text(size = text_size),
           axis.title = element_text(size = text_size),
-          plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 5, 0)),
-          strip.text = element_text(size = text_size),
+          plot.title = element_text(size = text_size + 2, face = "bold", hjust = 0, margin = ggplot2::margin(0, 0, 10, 0)),
+          strip.text = element_text(size = text_size, face = 2, margin = margin(b = 11)),
           panel.grid.major = element_blank(),
           panel.border = element_blank(),
           axis.line.x = element_line(),
           axis.line.y = element_line(),
           axis.ticks =  element_line(),
-          legend.text = element_text(size = 8))
+          legend.text = element_text(size = text_size),
+          legend.title = element_text(size = text_size),
+          panel.spacing = unit(1, "cm"),
+          plot.title.position = "plot")
 }
 
 theme_bw2 <- function () { 

### 2024-01-19 utils/spatial.r

@@ -54,6 +54,27 @@ create_coord_df <- function(spat_poly_df) {
   return(cellCoords)
 }
 
+#' Create matrix from rectangular spatial polygon
+#' 
+#' @param sP spatialPolygons object
+
+sP_to_matrix <- function(sP) {
+  # Assuming you have the spatial_poly as defined in the previous examples
+  coords <- coordinates(sP)
+  
+  # Extract x and y coordinates
+  x_coords <- coords[, 1]
+  y_coords <- coords[, 2]
+  
+  # Determine the number of rows and columns
+  num_rows <- length(unique(y_coords))
+  num_cols <- length(unique(x_coords))
+  
+  # Create a matrix with zeros
+  matrix_from_polygon <- matrix(0, nrow = num_rows, ncol = num_cols, dimnames = list(sort(unique(y_coords), decreasing = T), sort(unique(x_coords))))
+  
+  return(matrix_from_polygon)
+}
 
 #' Compute volume from spatial polygon data frame 
 #' 
@@ -86,7 +107,6 @@ find_raster <- function(x, y, cellCoordsXY) {
   }
   return(closest_id)
 }
-
 
 #' Rotate x and y date by a specific angle
 #' 

### 2024-01-19 utils/trans_risk.r

@@ -1,3 +1,46 @@
+#' Number of undiagnosed TB patients
+#' 
+#' @param n number of samples 
+#' @param empirical sample based on empirical counts (default `True`) or population-level prevalence
+#' @param counts number of counts of TB patients (0, 1, 2, ...) vising the clinic  
+#' @param prev_est prevalence estimates in %, either a vector with mean and sd, or a ISO-3 char for the country (if estimates are available)
+#' 
+#' @details 
+#' Distribution parametrized empirically based on the counts of occurrences of the
+#' daily number of diagnosed TB patients visiting the clinic,
+#' using a multinomial distribution, and
+#' assuming that the number of undiagnosed TB patients is similar in number.
+#' 
+#' Population-level prevalence for South Africa taken from Moyo. 
+#' 
+#' @return if empirical, number of undiagnosed TB patients, otherwise proportion of undiagnosed TB patients
+#' 
+#' @references 
+#' - Moyo: Moyo et al. (2022) in Lancet Infect Dis (doi.org/10.1016/S1473-3099(22)00149-9)
+
+rundiag <- function(n, empirical = T, counts = c(6, 6, 7, 4, 5, 1), prev_est = "RSA") {
+  if (empirical) {
+    BM <- rmultinom(n, size = 1, prob = counts) 
+    n_undiag <- apply(BM, 2, function(col) which(col == 1)) - 1
+    return(n_undiag)
+  } else {
+    if (is.character(prev_est)) {
+      if (prev_est == "RSA") {
+        prev_mean <- 852 / 100000
+        prev_sd <- (1026 - 679) / 100000 / (qnorm(.975) * 2)
+      } else {
+        stop("No prevalence estimates currently included for this country.")
+      }
+    } else {
+      prev_mean <- prev_est[1]
+      prev_sd <- prev_est[2]
+    }
+    p_undiag <- LaplacesDemon::rtrunc(n, "norm", a = 0, b = 1, mean = prev_mean, sd = prev_sd)
+    return(p_undiag)
+  }
+}
+
+
 #' Reduction in quanta emission rate through surgical mask wearing
 #' 
 #' @param n number of samples
@@ -12,6 +55,33 @@
 
 rrmask <- function(n) {
   LaplacesDemon::rtrunc(n, spec = "norm", a = 0, b = 1, mean = 0.56, sd = 0.11)
+}
+
+
+#' Inhalation rate
+#' 
[…]

### 2024-01-24 analysis/patient-tracking.Rmd
minor fixes and qctr analysis draft for scientific meeting presentation

@@
 df <- readRDS("../data-clean/patient-tracking/augmented-and-combined-data.rds")
+
+# filter considered clinical attendees
+df_matched <- do.call(rbind, lapply(unique(df$date), function(d) readRDS(paste0("../data-clean/patient-tracking/matched-clinical/", d, ".rds")))) %>%
+  mutate(date = as.Date(time))
+df_filt <- filter(df, paste(date, patient_id) %in% unique(paste(df_matched$date, df_matched$patient_id)))
 ```
@@
 ## Descriptives
+
+```{r}
+n_distinct(paste(df$date, df$patient_id))
+n_distinct(paste(df_filt$date, df_filt$patient_id))
+```
 
@@
 ```{r}
-time_in_clin <- df %>%
+time_in_clin <- df_filt %>%
   group_by(date, patient_id) %>%
@@
 # number of close contacts <1m for >1min
-no_close_contacts <- df %>%
+no_close_contacts <- df_filt %>%
   unnest(cols = closeContacts) %>%
@@
 
-time_close_contacts <- df %>%
+time_close_contacts <- df_filt %>%
   mutate(has_closeContact = ifelse(sapply(closeContacts, length) >= 1, T, F)) %>%
@@
   geom_step() +
-  facet_wrap(~ variable) +
+  ggh4x::facet_wrap2(~ variable, axes = "x", remove_labels = "all") +
   coord_cartesian(expand = F) +
@@
         legend.title = element_blank(),
-        legend.position = "bottom"
-        )
+        legend.position = "bottom")
 
@@
   labs(fill = "Number of tracks", title = "Spatial occupancy", subtitle = "Total number of tracks per grid cell") +
-  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
-                    breaks = c(1, 10, 100, 1000, 10000)) +
-  theme(legend.position = c(0.825,0.075), legend.direction = "horizontal") +
-  guides(fill = guide_legend(nrow = 2, title.position = "top"))
+  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", labels = scales::comma) +
+  # scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
+  #                   breaks = c(1, 10, 100, 1000, 10000)) +
+  theme(legend.position = c(0.82,0.095), legend.direction = "horizontal",
+        legend.key.height = unit(0.2, "cm"), legend.key.width = unit(.8, "cm")) +
+  guides(fill = guide_colorbar(title.position = "top"))
 

### 2024-01-24 analysis/qctr.Rmd
minor fixes and qctr analysis draft for scientific meeting presentation

---
title: "Quanta concentration and Transmission risk"
author: "Nicolas Banholzer"
date: "2023-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)
library(rstanarm)
library(tidybayes)
library(gganimate)

source("../utils/plotting.r")
source("../utils/tex.r")
source("../models/stm-v2.R")

# building
cellSize = 250
source("../preprocessing/prep_building-rasterize.R")
```

## Illustration

```{r}
# data
mth = 10
dy = 25
sim = 1
param_1025 <- readRDS(paste0("../simulations/", "/", "sim-param-", 2021, "-", mth, "-", dy, ".rds"))
track_1025 <- readRDS(paste0("../simulations/", "/", "tracking-df-", 250, "px-", 2021, "-", mth, "-", dy, ".rds"))
track_undiag <- track_1025 %>%
  dplyr::filter(tracking_end != "Possible HCW") # always consider HCW as non-infectious
track_diag <- dplyr::filter(track_1025, is_infectious)
track_undiag <- dplyr::filter(track_undiag, !is_infectious | is.na(is_infectious))
track_tb_all <- rbind(
  track_diag,
  filter(track_undiag, patient_id %in% param_1025[[paste0("undiagTB","_", "du")]][[sim]])
)
aer_df <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
  filter(month(date) == as.numeric(mth),
         day(date) == as.numeric(dy))

# time
date <- paste0("2021-",mth, "-", dy)
t_start <- as.POSIXct(paste(date, "08:00:00"))
midday <- as.POSIXct(paste(date, "12:00:00"))
t_end <- as.POSIXct(paste(date, "16:00:00"))
tt <- seq.POSIXt(t_start, t_end, by = "1 sec")
tt_morn <- seq.POSIXt(t_start, midday, by = "1 sec")
tt_noon <- seq.POSIXt(midday + 1, t_end, by = "1 sec")
midday_idx <- which(tt == midday)

# simulation
ct_by_room <- list()
for (room in c("Waiting room", "Corridor", "TB room")) {

    print(paste0("- ", room))

    # room setup
    if (room == "Waiting room") {

      # initial quanta concentration matrix
      c0 <- sP_to_matrix(waiting_room)

      # (no) TB patients in room
      track_tb <- filter(track_tb_all, is_waitingroom)

      # room dimension
      roomDim <- dimWR

    } else if (room == "Corridor") {

      # initial quanta concentration matrix
      c0 <- sP_to_matrix(corridor)

      # TB patients in room
      track_tb <- filter(track_tb_all, is_passage)

      # room dimension
      roomDim <- dimCD

    } else if (room == "TB room") {

      # initial quanta concentration matrix
      c0 <- sP_to_matrix(tb_room)

      # TB patients in room
      track_tb <- filter(track_tb_all, is_tbroom)

      # room dimension
      roomDim <- dimTB

    } else {
      stop("Error: Room invalid.")
    }

    # add quanta generation rate per second by activity level
    track_tb <- track_tb %>%
      dplyr::select(time, cell_x, cell_y, activity) %>%
      rename(x = cell_x, y = cell_y) %>%
      mutate(q = ifelse(activity == 1, param_1025$q_wait[param_1025$sim == sim], param_1025$q_walk[param_1025$sim == sim]),
             q = q * param_1025$mask_red[param_1025$sim == sim],
             q = q / 3600)

    # quanta conc in the morning
    print("-- morning")
    track_tb_morn <- filter(track_tb, time <= midday)
    track_tb_morn$t <- as.numeric(difftime(track_tb_morn$time, t_start, units = "sec"))

    ct_by_room[[room]] <- stm(time = tt_morn,
                     c0 = c0, 
                     inf = dplyr::select(track_tb_morn, -time, -activity),
                     cellLength = convert_dist(cellSize),
                     pd = 0,
                     vol = prod(roomDim),
                     aer = aer_df$aer[aer_df$room==room & aer_df$daytime=="Morning"] / 3600,
                     lambda = param_1025$viral_inact[param_1025$sim == sim] / 3600,
                     deposit = 0)
}

# get subset
subset_qc <- function(t) {

  # quanta conc
  qc_wr <- melt(ct_by_room[["Waiting room"]][,,t]) %>%
    rename(y = Var1, x = Var2) %>%
    mutate(room = "Waiting room", across(c(x, y), as.numeric))
  qc_cr <- melt(ct_by_room[["Corridor"]][,,t]) %>%
    rename(y = Var1, x = Var2) %>%
    mutate(room = "Corridor", across(c(x, y), as.numeric))
  qc_tb <- melt(ct_by_room[["TB room"]][,,t]) %>%
    rename(y = Var1, x = Var2) %>%
    mutate(room = "TB room", across(c(x, y), as.numeric))

  # link to building
  qc <- rbind(qc_wr, qc_cr, qc_tb) 
  qc <- left_join(roomCoord, qc, by = c("x", "y", "room")) %>%
    mutate(id = paste0("ID", id),
           t = t)

  return(qc)
}

track_tb_all %>%
  filter(between(time, as.POSIXct(paste(date, "09:41:00")), as.POSIXct(paste(date, "09:45:00")))) %>%
  ggplot(aes(x = x, y = y, color = factor(patient_id))) +
  geom_path()

tt_selected <- seq.POSIXt(as.POSIXct(paste(date, "09:41:00")), as.POSIXct(paste(date, "09:45:00")), by = "1 sec")
tt_selected_idx <- as.numeric(tt_selected - t_start, unit = "secs")
names(tt_selected) <- tt_selected_idx

qc_sub <- do.call(rbind, parallel::mclapply(tt_selected_idx, subset_qc, mc.cores = 4)) %>%
  mutate(t = t - min(tt_selected_idx))

# animated plot
qc_anim_pl <- ggplot(data = left_join(fortify(waiting_room), filter(qc_sub, room == "Waiting room"), by = c("id")),
         mapping = aes(x = long, y = lat, group = group, fill = value)) +
  geom_polygon() +
  coord_cartesian(expand = F) +
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)) +
  theme_custom(14) +
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.position = "bottom",
        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  transition_time(t) +
  labs(subtitle = "Time (secs): {frame_time}", title = "Quanta concentration in the waiting room on Oct 25 between 9:41 and 09:45am",
       fill = expression("Quanta/m"^3*" x 10e"^-3))

animate(qc_anim_pl, duration = 30, width = 1.850877 * 600, height = 600, renderer = gifski_renderer(), nframes = length(tt_selected_idx))
anim_save("../results/modeling/illustrated-quanta-conc-over-time.gif")
```

## Main

```{r}
main_files <- c(
  list.files("../simulations/main/10-13/", full.names = T),
  list.files("../simulations/main/10-15/", full.names = T),
  list.files("../simulations/main/10-25/", full.names = T),
  list.files("../simulations/main/11-04/", full.names = T),
  list.files("../simulations/main/11-05/", full.names = T))
```

### Quanta conc

```{r}
main_qc_files <- main_files[grepl("quanta", main_files)]

mean_qc <- function(files, room, daytime) {
  room_files <- files[grepl(room, files)]
  room_dt_files <- room_files[grepl(daytime, room_files)]
  qc <- simplify2array(lapply(room_dt_files, readRDS))
  mqc <- apply(qc, c(1,2), mean)
  mqc_df <- melt(mqc) %>%
    rename(y = Var1,
           x = Var2) %>%
    mutate(room = room,
           daytime = daytime,
           across(c(x, y), as.numeric))
}

# total quanta concentration by daytime averaged across simulations 
main_qc_wr_mo <- mean_qc(main_qc_files, "Waiting room", "morning")
main_qc_wr_af <- mean_qc(main_qc_files, "Waiting room", "afternoon")
main_qc_cr_mo <- mean_qc(main_qc_files, "Corridor", "morning")
main_qc_cr_af <- mean_qc(main_qc_files, "Corridor", "afternoon")
main_qc_tb_mo <- mean_qc(main_qc_files, "TB room", "morning")
main_qc_tb_af <- mean_qc(main_qc_files, "TB room", "afternoon")

# link to building
main_qc <- rbind(main_qc_wr_mo, main_qc_wr_af, 
                 main_qc_cr_mo, main_qc_cr_af,
                 main_qc_tb_mo, main_qc_tb_af) 
main_qc <- left_join(roomCoord, 
                     main_qc, 
                     by = c("x", "y", "room")) %>%
  mutate(id = paste0("ID", id),
         daytime = factor(ifelse(daytime == "morning", "Morning", "Afternoon"), levels = c("Morning", "Afternoon")))

# plot

mean_quanta_pl <- plot_spatial(
  ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
    facet_wrap(~ daytime) +
    geom_polygon(data = left_join(fortify(waiting_room), filter(main_qc, room == "Waiting room"), by = c("id"))) +
    geom_polygon(data = left_join(fortify(corridor), filter(main_qc, room == "Corridor"), by = c("id"))) +
    geom_polygon(data = left_join(fortify(tb_room), filter(main_qc, room == "TB room"), by = c("id"))) +
    labs(fill = expression("Quanta/m"^3*" x 10e"^-3), title = "Spatial quanta concentration by daytime", subtitle = "Mean concentration in the morning and afternoon") +
    scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)),
  14
  ) +
  theme(legend.position = c(0.915,0.095), legend.direction = "horizontal",
        legend.key.height = unit(0.2, "cm"), legend.key.width = unit(.8, "cm")) +
  guides(fill = guide_colorbar(title.position = "top"))

save_plot(mean_quanta_pl, pdf_file = "../results/modeling/quanta-concentration.pdf", w = 26, h = 12)
```

### Trans risk

```{r}
# files
main_tr_files <- main_files[grepl("risk", main_files)]

# data
main_tr <- do.call(rbind, lapply(main_tr_files, function(f) readRDS(f) %>% mutate(mthdy = stringi::stri_extract(f, regex = "\\d{2}-\\d{2}"), 
                                                                                  sim = stringi::stri_extract(f, regex = "\\d{1,6}"))))

mean_main_tr <- main_tr %>%
  mutate(patient_id = paste(mthdy, patient_id, sep = "-")) %>%
  group_by(patient_id) %>%
  summarize(P_mean = mean(P),
            P_med = median(P)) %>%
  ungroup() %>% 
  reshape2::melt("patient_id") %>%
  mutate(val_col = ifelse(value < .05, "Low", ifelse(value < .1, "Median", "High")),
         val_col = factor(val_col, levels = c("Low", "Median", "High")),
         variable = ifelse(grepl("mean", variable), "Mean risk", "Median risk"),
         variable = factor(variable, levels = c("Mean risk", "Median risk")))

roi_annon <- mean_main_tr %>%
  group_by(variable) %>%
  summarize(mean_P = mean(value),
            lower_P = quantile(value, .025),
            upper_P = quantile(value, .975)) %>%
  ungroup() %>%
  mutate(across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 1)),
         lab = paste0("Mean ", mean_P, "%", " (95%-CrI ", lower_P, "% – ", upper_P, "%)"),
         value = .25, 
         y = 1350)

roi_pl <- ggplot(mean_main_tr, aes(x = value)) +
  geom_histogram(mapping = aes(fill = val_col), breaks = seq(.0, .5, .05)) +
  geom_text(data = roi_annon, mapping = aes(x = value, y = y, label = lab), size = 12 / cm(1)) +
  facet_wrap(~ variable) +
  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2,3,1)]) +
  coord_cartesian(expand = F) +
  scale_x_continuous(breaks = c(0, .05, 0.1, seq(0.2, .5, .1)), labels = function(x) x * 100) +
  scale_y_continuous(labels = scales::comma, breaks = seq(0, 1500, 250)) +
  labs(x = "Risk of infection (%)", y = "Number of patients",
       title = "Risk of infection",
       subtitle = "Mean and median individual risk of infection in clinic") +
  theme_custom(14) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        panel.spacing = unit(2, "cm"))

save_plot(roi_pl, pdf_file = "../results/modeling/risk-of-infection.pdf", w = 26, h = 12)
```

### Association

```{r}
# load tracking and environmental data
track <- readRDS("../data-clean/patient-tracking/augmented-and-combined-data.rds")
aer <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
  mutate(daytime = factor(daytime, levels = c("Morning", "Afternoon")))

# process tracking data
track_daytime <- track %>%
  mutate(h = lubridate::hour(time)) %>%
  group_by(date, patient_id) %>%
  summarize(dur_morning = sum(h < 12),
            dur_afternoon = sum(h >= 12),
            tracking_end = tracking_end[1]) %>%
  ungroup() %>%
  mutate(daytime = ifelse(dur_morning > dur_afternoon, "Morning", "Afternoon"))

time_in_clin <- track %>%
  group_by(date, patient_id) %>%
  summarize(`Clinic` = n(),
            `Waiting room` = sum(is_waitingroom),
            `TB room` = sum(is_tbroom),
            `Corridor` = sum(is_passage)) %>%
  ungroup() %>%
  mutate(across(c(`Clinic`, `Waiting room`, `TB room`, `Corridor`), ~ .x / 60))

no_close_contacts <- track %>%
  unnest(cols = closeContacts) %>%
  group_by(date, patient_id, closeContacts, tracking_end) %>%
  summarise(n_time = n()) %>%
  ungroup() %>%
  group_by(date, patient_id) %>%
  summarize(n_cc = sum(n_time > 60)) %>%
  ungroup()

time_close_contacts <- track %>%
  mutate(has_closeContact = ifelse(sapply(closeContacts, length) >= 1, T, F)) %>%
  group_by(date, patient_id) %>%
  summarize(cc_time = sum(has_closeContact) / 60) %>%
  ungroup()

# merge data
track_ft <- left_join(track_daytime, left_join(left_join(time_in_clin, no_close_contacts), time_close_contacts)) %>%
  mutate(n_cc = ifelse(is.na(n_cc), 0, n_cc)) %>%
  rename(time_in_clinic = `Clinic`) %>%
  dplyr::select(date, patient_id, tracking_end, daytime, time_in_clinic, n_cc, cc_time)
assoc_df <- track_ft %>%
  left_join(aer %>% filter(room == "Waiting room") %>% dplyr::select(date, daytime, aer)) %>%
  mutate(patient_id = paste(lubridate::month(date), lubridate::day(date), patient_id, sep = "-")) %>%
  left_join(reshape2::dcast(mean_main_tr, patient_id ~ variable) %>% dplyr::select(patient_id, `Mean risk`) %>% rename(mean_risk = `Mean risk`))

# model
amod <- stan_betareg(mean_risk ~ aer + time_in_clinic + n_cc + cc_time, 
                     data = assoc_df %>% mutate(n_cc = n_cc + 1, cc_time = cc_time + 1/60,
                                                across(c(aer, time_in_clinic, n_cc, cc_time), log2)), 
                     seed = 12345, cores = 4)

amod_est <- spread_draws(amod, aer, time_in_clinic, n_cc, cc_time) %>%
  reshape2::melt(c(".chain", ".iteration", ".draw")) %>%
  mutate(value = exp(value))

amod_est_pl <- ggplot(amod_est, aes(y = variable, x = value)) +
  geom_vline(aes(xintercept = 1), linetype = "dotted", color = "red") +
  stat_pointinterval() +
  labs(x = "Odds ratio associated with a doubling of the predictor",
       title = "Predictors of infection risk",
       subtitle = "Association of close contact, time in clinic, and air change rate with mean risk of infection",
       caption = "Note: Median as dots, 66%- and 95%-CrI as thick and thin lines.") +
  scale_y_discrete(labels = c("Air change rate", "Time in clinic", "Number of close contacts", "Time in clocse contact")) +
  theme_custom(14) +
  theme(axis.title.y = element_blank())

save_plot(amod_est_pl, pdf_file = "../results/modeling/association-predictors.pdf", w = 20, h = 12)
```

### 2024-01-24 preprocessing/prep_building-rasterize.R

@@ -63,7 +63,7 @@ plot_spatial <- function(pl, text_descr = 10) {
   re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(fontsize = text_descr), just = c("left"), rot = 90)
   col <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")
   
-  pl +
+ pl +
     ggpubr::background_image(clinic_img) +
     scale_y_continuous(expand = expansion(add=c(1700, 3000)), limits = c(-6500, 2000)) +
     scale_x_continuous(expand = expansion(add=c(400, 500)), limits = c(-16000, 7500)) +
@@ -74,11 +74,8 @@ plot_spatial <- function(pl, text_descr = 10) {
     theme_custom(text_descr) +
     theme(text = element_text(size = text_descr),
           axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
-          legend.position = c(0.825,0.05), legend.direction = "horizontal", 
-          legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
           panel.background = element_rect(fill = col[1]), panel.grid.major = element_blank(),
-          panel.grid.minor = element_blank()) +
-    guides(fill = guide_legend(title.position = "top"))
+          panel.grid.minor = element_blank()) 
 }
 
 # room data frames

### 2024-01-24 preprocessing/simulate_qctr.R

@@ -251,7 +251,7 @@ for (sim in cont_n:term_n) {
       dplyr::select(time, cell_x, cell_y, activity) %>%
       rename(x = cell_x, y = cell_y) %>%
       mutate(q = ifelse(activity == 1, sim_param$q_wait[sim_param$sim == sim], sim_param$q_walk[sim_param$sim == sim]),
-             q = ifelse(is_masking == 1, q * sim_param$mask_red[sim_param$sim == sim]),
+             q = ifelse(is_masking == 1, q * sim_param$mask_red[sim_param$sim == sim], q),
              q = q / 3600)
     
     # quanta conc in the morning
@@ -265,7 +265,7 @@ for (sim in cont_n:term_n) {
                      c0 = c0, 
                      inf = dplyr::select(track_tb_morn, -time, -activity),
                      cellLength = convert_dist(cellSize),
-                     pd = 0,
+                     pd = ifelse(is_masking == 1, 0, 1),
                      vol = prod(roomDim),
                      aer = aer_df$aer[aer_df$room==room & aer_df$daytime=="Morning"] / 3600,
                      lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
@@ -302,7 +302,7 @@ for (sim in cont_n:term_n) {
                      c0 = c0_noon, 
                      inf = dplyr::select(track_tb_noon, -time, -activity),
                      cellLength = convert_dist(cellSize),
-                     pd = 0,
+                     pd = ifelse(is_masking == 1, 0, 1),
                      vol = prod(roomDim),
                      aer = aer_df$aer[aer_df$room==room & aer_df$daytime=="Afternoon"] / 3600,
                      lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
@@ -325,11 +325,18 @@ for (sim in cont_n:term_n) {
     
     # compute quanta conc exposure by susc
     ct <- abind::abind(ct_morn, ct_noon, along = 3)
-    track_susc$quanta <- mapply(function(t, x, y) {if (t == 0) 0 else {ct[y,x,t]}}, 
-                                t=track_susc$t, x=as.character(track_susc$cell_x), y=as.character(track_susc$cell_y))
-    track_susc <- track_susc %>%
-      mutate(quanta_conc = inhal_rate * quanta / (convert_dist(cellSize) ^ 2 * convert_dist(cell_height))) %>%
[…]

### 2024-01-24 tests/test_stm_v2.R

@@ -54,11 +54,11 @@ quanta_concn_pl <- ctDF %>%
   ggplot(aes(x = y, y = x, fill = value)) +
   facet_wrap(~ time, ncol = 3) +
   geom_tile() +
-  labs(fill = expression("Quanta/m"^3), x = "x length in m", y = "y width in m",
+  labs(fill = expression("Quanta/m"^3*" x 10e"^-3), x = "x length in m", y = "y width in m",
        title = "Example: Spatiotemporal model",
        subtitle = "Spatiotemporal quanta concentration with 1 infectious individual (centered, 8-9am)",
-       caption = expression("Assumptions: 150m"^3*" room with q = 10 quanta/h and AER = 10 air changes/h")) +
-  scale_fill_gradient(low = "red", high = "yellow", limits = c(0, NA), labels = function(x) round(x / (gridCellLength^2*dz), 2)) +
+       caption = expression("Assumptions: 150m"^3*" room with q = 10 quanta/h and AER = 3 air changes/h")) +
+  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), labels = function(x) round(1e3 * x / (gridCellLength^2*dz), 1)) +
   scale_x_continuous(expand = c(0,0), limits = c(0,40), breaks = seq(0,40,8), labels = function(x) x * .25) +
   scale_y_continuous(expand = c(0,0), limits = c(0,20), breaks = seq(0,20,4), labels = function(x) x * .25) +
   theme_custom() +

### 2024-01-24 models/tm-v1.R
add temporal model

@@ -1,0 +1,85 @@
+#' Spatiotemporal model of the quanta concentration
+#' 
+#' @description Function to compute the quanta concentration at each time point. 
+#' 
+#' @param time time vector in 1s
+#' @param c0 2d-matrix with initial quanta concentration (in quanta per m3)
+#' @param inf data frame with time, x- and y-coordinates, and generated quanta q
+#' @param aer air changes per second
+#' @param lambda viral inactivation rate per second
+#' @param deposit deposition rate per second due to gravitational settling
+#' 
+#' @details The default for parameters q, lambda, and k is based on prior research for Mtb. 
+#' The diffusion constant is computed relative to the air change rate, see Cheng (2011).
+#' 
+#' @references Cheng KC, Acevedo-Bolton V, Jiang RT, Klepeis NE, Ott WR, Fringer OB, Hildemann LM. 
+#' Modeling exposure close to air pollution sources in naturally ventilated residences: 
+#' Association of turbulent diffusion coefficient with air change rate. 
+#' Environmental science & technology. 2011;45(9):4016-22. 
+#' 
+#' @return a list of matrices with the quanta concentration at each time step t_i   
+#' 
+
+tm <- function(time, 
+               c0, 
+               inf, 
+               aer = 3/3600,
+               lambda = 1/3600,
+               deposit = 0) {
+  
+  # room dimensions
+  nr <- nrow(c0)
+  nc <- ncol(c0)
+  n_cells <- nr * nc
+  
+  # initialize 
+  nT <- length(time)
+  time_int <- 1:nT
+  ct <- replicate(nT+2, c0)
+  x_crds <- row.names(c0)
+  y_crds <- colnames(c0)
+  
+  # add start and end without quanta generation
+  if (nrow(inf) == 0 & sum(c0) == 0) {
+    return(ct[,,-1])
+  } else if (nrow(inf) == 0) {
+    inf <- tibble(t = 1, data = list(data.frame(x = y_crds[1], y = x_crds[1], q = 0)))
+    inf$te <- max(time_int) + 2
+  } else {
+    inf0 <- inf[1, ]
+    inf0$t <- 0
+    inf0$q <- 0
[…]

### 2024-03-04 preprocessing/simulate_qctr.R
Fixed suspected model

@@ -181,10 +181,21 @@ for (sim in cont_n:term_n) {
   )
 
   # no TB patient tracks but including HCW
-  track_susc_all <- rbind(
-    filter(track, tracking_end == "Possible HCW"),
-    filter(track_undiag, !(patient_id %in% sim_param[[paste0("undiagTB", "_", who_is_tb)]][[sim]]))
-  )
+  if (who_is_tb == "ds") {
+    track_susc_all <- rbind(
+      filter(track, tracking_end == "Possible HCW"),
+      filter(
+        track,
+        tracking_end != "Possible HCW",
+        !(patient_id %in% pid_diag)
+      )
+    )
+  } else {
+    track_susc_all <- rbind(
+      filter(track, tracking_end == "Possible HCW"),
+      filter(track_undiag, !(patient_id %in% sim_param[[paste0("undiagTB", "_", who_is_tb)]][[sim]]))
+    )
+  }
 
   # initialize
   susc_quanta_exp <- list()

### 2024-03-04 analysis/qctr.Rmd
QCTR results draft

@@
 
-source("../utils/plotting.r")
-source("../utils/tex.r")
-source("../models/stm-v2.R")
+source("utils/plotting.r")
+source("utils/tex.r")
+source("models/stm-v2.R")
 
 # building
-cellSize = 250
-source("../preprocessing/prep_building-rasterize.R")
+cellSize <- 250
+source("preprocessing/prep_building-rasterize.R")
+
+# load tracking and environmental data
+track <- readRDS("data-clean/patient-tracking/augmented-and-combined-data.rds")
+aer <- readRDS("data-clean/environmental/air-change-rate.rds") %>%
+  mutate(daytime = factor(daytime, levels = c("Morning", "Afternoon")))
 ```
@@
 # data
-mth = 10
-dy = 25
-sim = 1
-param_1025 <- readRDS(paste0("../simulations/", "/", "sim-param-", 2021, "-", mth, "-", dy, ".rds"))
-track_1025 <- readRDS(paste0("../simulations/", "/", "tracking-df-", 250, "px-", 2021, "-", mth, "-", dy, ".rds"))
+mth <- 10
+dy <- 25
+sim <- 1
+param_1025 <- readRDS(
+  paste0(
+    "../simulations/", "/", "sim-param-",
+    2021, "-", mth, "-", dy, ".rds"
+  )
+)
+track_1025 <- readRDS(
+  paste0(
+    "../simulations/", "/", "tracking-df-",
+    250, "px-", 2021, "-", mth, "-", dy, ".rds"
+  )
+)
 track_undiag <- track_1025 %>%
-  dplyr::filter(tracking_end != "Possible HCW") # always consider HCW as non-infectious
+  dplyr::filter(tracking_end != "Possible HCW")
 track_diag <- dplyr::filter(track_1025, is_infectious)
-track_undiag <- dplyr::filter(track_undiag, !is_infectious | is.na(is_infectious))
+track_undiag <- dplyr::filter(
+  track_undiag, !is_infectious | is.na(is_infectious)
+)
 track_tb_all <- rbind(
   track_diag,
-  filter(track_undiag, patient_id %in% param_1025[[paste0("undiagTB","_", "du")]][[sim]])
+  filter(
+    track_undiag,
+    patient_id %in% param_1025[[paste0("undiagTB", "_", "du")]][[sim]]
+  )
 )
 aer_df <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
-  filter(month(date) == as.numeric(mth),
-         day(date) == as.numeric(dy))
+  filter(
+    month(date) == as.numeric(mth),
+    day(date) == as.numeric(dy)
+  )
 
 # time
-date <- paste0("2021-",mth, "-", dy)
+date <- paste0("2021-", mth, "-", dy)
 t_start <- as.POSIXct(paste(date, "08:00:00"))
@@
 for (room in c("Waiting room", "Corridor", "TB room")) {
-
   print(paste0("- ", room))
@@
   if (room == "Waiting room") {
-
     # initial quanta concentration matrix
@@
     roomDim <- dimWR
-
   } else if (room == "Corridor") {
-
     # initial quanta concentration matrix
@@
     roomDim <- dimCD
-
   } else if (room == "TB room") {
-
     # initial quanta concentration matrix
@@
     roomDim <- dimTB
-
   } else {
@@
     rename(x = cell_x, y = cell_y) %>%
-      mutate(q = ifelse(activity == 1, param_1025$q_wait[param_1025$sim == sim], param_1025$q_walk[param_1025$sim == sim]),
+    mutate(
+      q = ifelse(activity == 1,
+        param_1025$q_wait[param_1025$sim == sim],
+        param_1025$q_walk[param_1025$sim == sim]
+      ),
       q = q * param_1025$mask_red[param_1025$sim == sim],
-             q = q / 3600)
+      q = q / 3600
+    )
 
@@
   track_tb_morn <- filter(track_tb, time <= midday)
-    track_tb_morn$t <- as.numeric(difftime(track_tb_morn$time, t_start, units = "sec"))
+  track_tb_morn$t <- as.numeric(difftime(
+    track_tb_morn$time, t_start,
+    units = "sec"
+  ))
 
-    ct_by_room[[room]] <- stm(time = tt_morn,
+  ct_by_room[[room]] <- stm(
+    time = tt_morn,
     c0 = c0,
@@
     vol = prod(roomDim),
-                     aer = aer_df$aer[aer_df$room==room & aer_df$daytime=="Morning"] / 3600,
+    aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Morning"] / 3600,
     lambda = param_1025$viral_inact[param_1025$sim == sim] / 3600,
-                     deposit = 0)
+    deposit = 0
+  )
 }
@@
 subset_qc <- function(t) {
-
   # quanta conc
-  qc_wr <- melt(ct_by_room[["Waiting room"]][,,t]) %>%
+  qc_wr <- melt(ct_by_room[["Waiting room"]][, , t]) %>%
     rename(y = Var1, x = Var2) %>%
     mutate(room = "Waiting room", across(c(x, y), as.numeric))
-  qc_cr <- melt(ct_by_room[["Corridor"]][,,t]) %>%
+  qc_cr <- melt(ct_by_room[["Corridor"]][, , t]) %>%
     rename(y = Var1, x = Var2) %>%
     mutate(room = "Corridor", across(c(x, y), as.numeric))
-  qc_tb <- melt(ct_by_room[["TB room"]][,,t]) %>%
+  qc_tb <- melt(ct_by_room[["TB room"]][, , t]) %>%
     rename(y = Var1, x = Var2) %>%
@@
   qc <- left_join(roomCoord, qc, by = c("x", "y", "room")) %>%
-    mutate(id = paste0("ID", id),
-           t = t)
+    mutate(
+      id = paste0("ID", id),
+      t = t
+    )
 
@@
 track_tb_all %>%
-  filter(between(time, as.POSIXct(paste(date, "09:41:00")), as.POSIXct(paste(date, "09:45:00")))) %>%
+  filter(between(
+    time,
+    as.POSIXct(paste(date, "09:41:00")),
+    as.POSIXct(paste(date, "09:45:00"))
+  )) %>%
   ggplot(aes(x = x, y = y, color = factor(patient_id))) +
@@
 
-tt_selected <- seq.POSIXt(as.POSIXct(paste(date, "09:41:00")), as.POSIXct(paste(date, "09:45:00")), by = "1 sec")
+tt_selected <- seq.POSIXt(
+  as.POSIXct(paste(date, "09:41:00")),
+  as.POSIXct(paste(date, "09:45:00")),
+  by = "1 sec"
+)
 tt_selected_idx <- as.numeric(tt_selected - t_start, unit = "secs")
@@
 
-qc_sub <- do.call(rbind, parallel::mclapply(tt_selected_idx, subset_qc, mc.cores = 4)) %>%
+qc_sub <- do.call(
+  rbind,
+  parallel::mclapply(tt_selected_idx, subset_qc, mc.cores = 4)
+) %>%
   mutate(t = t - min(tt_selected_idx))
@@
 # animated plot
-qc_anim_pl <- ggplot(data = left_join(fortify(waiting_room), filter(qc_sub, room == "Waiting room"), by = c("id")),
-         mapping = aes(x = long, y = lat, group = group, fill = value)) +
+qc_anim_pl <- ggplot(
+  data = left_join(
+    fortify(waiting_room),
+    filter(qc_sub, room == "Waiting room"),
+    by = c("id")
+  ),
+  mapping = aes(x = long, y = lat, group = group, fill = value)
+) +
   geom_polygon() +
-  coord_cartesian(expand = F) +
-  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)) +
+  coord_cartesian(expand = FALSE) +
+  scale_fill_gradientn(
+    colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
+    labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)
+  ) +
   theme_custom(14) +
-  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), legend.position = "bottom",
-        panel.background = element_rect(fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  theme(
+    axis.text = element_blank(),
+    axis.title = element_blank(),
+    axis.ticks = element_blank(),
+    legend.position = "bottom",
+    panel.background = element_rect(
+      fill = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")[1]
+    ),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank()
+  ) +
   transition_time(t) +
-  labs(subtitle = "Time (secs): {frame_time}", title = "Quanta concentration in the waiting room on Oct 25 between 9:41 and 09:45am",
-       fill = expression("Quanta/m"^3*" x 10e"^-3))
+  labs(
+    subtitle = "Time (secs): {frame_time}",
+    title = "Quanta concentration in the waiting room on Oct 25 between 9:41 and 09:45am", # nolint
+    fill = expression("Quanta/m"^3 * " x 10e"^-3)
+  )
 
-animate(qc_anim_pl, duration = 30, width = 1.850877 * 600, height = 600, renderer = gifski_renderer(), nframes = length(tt_selected_idx))
-anim_save("../results/modeling/illustrated-quanta-conc-over-time.gif")
+animate(
+  qc_anim_pl,
+  duration = 30,
+  width = 1.850877 * 600,
+  height = 600,
+  renderer = gifski_renderer(),
+  nframes = length(tt_selected_idx)
+)
+
+anim_save("results/modeling/illustrated-quanta-conc-over-time.gif")
 ```
@@
 main_files <- c(
-  list.files("../simulations/main/10-13/", full.names = T),
-  list.files("../simulations/main/10-15/", full.names = T),
-  list.files("../simulations/main/10-25/", full.names = T),
-  list.files("../simulations/main/11-04/", full.names = T),
-  list.files("../simulations/main/11-05/", full.names = T))
+  list.files("simulations/main/10-13/", full.names = T),
+  list.files("simulations/main/10-15/", full.names = T),
+  list.files("simulations/main/10-25/", full.names = T),
+  list.files("simulations/main/11-04/", full.names = T),
+  list.files("simulations/main/11-05/", full.names = T)
+)
 ```
@@
   qc <- simplify2array(lapply(room_dt_files, readRDS))
-  mqc <- apply(qc, c(1,2), mean)
+  mqc <- apply(qc, c(1, 2), mean)
   mqc_df <- melt(mqc) %>%
-    rename(y = Var1,
-           x = Var2) %>%
-    mutate(room = room,
+    rename(
+      y = Var1,
+      x = Var2
+    ) %>%
+    mutate(
+      room = room,
       daytime = daytime,
-           across(c(x, y), as.numeric))
+      across(c(x, y), as.numeric)
+    )
 }
@@
 # link to building
-main_qc <- rbind(main_qc_wr_mo, main_qc_wr_af, 
+main_qc <- rbind(
+  main_qc_wr_mo, main_qc_wr_af,
   main_qc_cr_mo, main_qc_cr_af,
-                 main_qc_tb_mo, main_qc_tb_af) 
+  main_qc_tb_mo, main_qc_tb_af
+)
 main_qc <- left_join(roomCoord,
   main_qc,
-                     by = c("x", "y", "room")) %>%
-  mutate(id = paste0("ID", id),
-         daytime = factor(ifelse(daytime == "morning", "Morning", "Afternoon"), levels = c("Morning", "Afternoon")))
+  by = c("x", "y", "room")
+) %>%
+  mutate(
+    id = paste0("ID", id),
+    daytime = factor(
+      ifelse(daytime == "morning", "Morning", "Afternoon"),
+      levels = c("Morning", "Afternoon")
+    )
+  )
 
 # plot
-
 mean_quanta_pl <- plot_spatial(
   ggplot(mapping = aes(x = long, y = lat, group = group, fill = value)) +
-    facet_wrap(~ daytime) +
-    geom_polygon(data = left_join(fortify(waiting_room), filter(main_qc, room == "Waiting room"), by = c("id"))) +
-    geom_polygon(data = left_join(fortify(corridor), filter(main_qc, room == "Corridor"), by = c("id"))) +
-    geom_polygon(data = left_join(fortify(tb_room), filter(main_qc, room == "TB room"), by = c("id"))) +
-    labs(fill = expression("Quanta/m"^3*" x 10e"^-3), title = "Spatial quanta concentration by daytime", subtitle = "Mean concentration in the morning and afternoon") +
-    scale_fill_gradientn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)),
-  14
+    facet_wrap(~daytime) +
+    geom_polygon(
+      data = left_join(
+        fortify(waiting_room),
+        filter(main_qc, room == "Waiting room"),
+        by = c("id")
+      )
     ) +
-  theme(legend.position = c(0.915,0.095), legend.direction = "horizontal",
-        legend.key.height = unit(0.2, "cm"), legend.key.width = unit(.8, "cm")) +
+    geom_polygon(
+      data = left_join(
+        fortify(corridor),
+        filter(main_qc, room == "Corridor"),
+        by = c("id")
+      )
+    ) +
+    geom_polygon(
+      data = left_join(fortify(tb_room),
+        filter(main_qc, room == "TB room"),
+        by = c("id")
+      )
+    ) +
+    labs(
+      fill = expression("Quanta concentration in quanta/m"^3 * " x 10e"^-3),
+      title = "a"
+    ) +
+    scale_fill_gradientn(
+      colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
+      labels = function(x) round(1e3 * x / (0.25^2 * 3), 1)
+    ),
+  8
+) +
+  theme(
+    legend.position = "bottom",
+    legend.direction = "horizontal",
+    legend.key.height = unit(0.2, "cm"),
+    legend.key.width = unit(2, "cm")
+  ) +
   guides(fill = guide_colorbar(title.position = "top"))
 
-save_plot(mean_quanta_pl, pdf_file = "../results/modeling/quanta-concentration.pdf", w = 26, h = 12)
+mean_quanta_pl
 ```
@@
 ```{r}
+average_trans_risk <- function(files) {
   # files
-main_tr_files <- main_files[grepl("risk", main_files)]
+  tr_files <- files[grepl("risk", files)]
 
   # data
-main_tr <- do.call(rbind, lapply(main_tr_files, function(f) readRDS(f) %>% mutate(mthdy = stringi::stri_extract(f, regex = "\\d{2}-\\d{2}"), 
-                                                                                  sim = stringi::stri_extract(f, regex = "\\d{1,6}"))))
+  tr <- do.call(rbind, lapply(tr_files, function(f) {
+    readRDS(f) %>% mutate(
+      mthdy = stringi::stri_extract(f, regex = "\\d{2}-\\d{2}"),
+      sim = stringi::stri_extract(f, regex = "\\d{1,6}")
+    )
+  }))
 
-mean_main_tr <- main_tr %>%
+  average_tr <- tr %>%
     mutate(patient_id = paste(mthdy, patient_id, sep = "-")) %>%
     group_by(patient_id) %>%
-  summarize(P_mean = mean(P),
-            P_med = median(P)) %>%
+    summarize(
+      P_mean = mean(P),
+      P_med = median(P)
+    ) %>%
     ungroup() %>%
+    left_join(track %>%
+      mutate(
+        mth = lubridate::month(time),
+        dy = lubridate::day(time)
+      ) %>%
+      group_by(patient_id, mth, dy) %>%
+      slice(1) %>%
+      ungroup() %>%
+      mutate(dy = ifelse(nchar(dy) == 1, paste0("0", dy), dy)) %>%
+      mutate(patient_id = paste(
+        mth, dy,
+        patient_id,
+        sep = "-"
+      )) %>%
+      dplyr::select(patient_id, tracking_end)) %>%
+    filter(tracking_end != "Possible HCW") %>%
+    dplyr::select(-tracking_end) %>%
     reshape2::melt("patient_id") %>%
-  mutate(val_col = ifelse(value < .05, "Low", ifelse(value < .1, "Median", "High")),
+    mutate(
+      val_col = ifelse(value < .05, "Low",
+        ifelse(value < .1, "Median", "High")
+      ),
       val_col = factor(val_col, levels = c("Low", "Median", "High")),
-         variable = ifelse(grepl("mean", variable), "Mean risk", "Median risk"),
-         variable = factor(variable, levels = c("Mean risk", "Median risk")))
+      variable = ifelse(
+        grepl("mean", variable),
+        "Mean risk of infection",
+        "Median risk of infection"
+      ),
+      variable = factor(
+        variable,
+        levels = paste(c("Mean", "Median"), "risk of infection")
+      )
+    )
 
-roi_annon <- mean_main_tr %>%
+  return(average_tr)
+}
+
+# average risk
+main_tr <- average_trans_risk(main_files)
+
+# annotation
+annotate_average_trans_risk <- function(tr) {
+  roi_annon <- tr %>%
     group_by(variable) %>%
-  summarize(mean_P = mean(value),
+    summarize(
+      mean_P = mean(value),
       lower_P = quantile(value, .025),
-            upper_P = quantile(value, .975)) %>%
+      upper_P = quantile(value, .975)
+    ) %>%
     ungroup() %>%
     mutate(across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 1)),
-         lab = paste0("Mean ", mean_P, "%", " (95%-CrI ", lower_P, "% – ", upper_P, "%)"),
-         value = .25, 
-         y = 1350)
+      lab = paste0(
+        "Average ", mean_P, "%",
+        " (95%-CrI ", lower_P, "% – ", upper_P, "%)"
+      )
+    )
 
-roi_pl <- ggplot(mean_main_tr, aes(x = value)) +
-  geom_histogram(mapping = aes(fill = val_col), breaks = seq(.0, .5, .05)) +
-  geom_text(data = roi_annon, mapping = aes(x = value, y = y, label = lab), size = 12 / cm(1)) +
-  facet_wrap(~ variable) +
-  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2,3,1)]) +
-  coord_cartesian(expand = F) +
-  scale_x_continuous(breaks = c(0, .05, 0.1, seq(0.2, .5, .1)), labels = function(x) x * 100) +
-  scale_y_continuous(labels = scales::comma, breaks = seq(0, 1500, 250)) +
-  labs(x = "Risk of infection (%)", y = "Number of patients",
-       title = "Risk of infection",
-       subtitle = "Mean and median individual risk of infection in clinic") +
-  theme_custom(14) +
-  theme(legend.position = "bottom",
+  return(roi_annon)
+}
+
+main_annon <- annotate_average_trans_risk(main_tr)
+
+# plot
+roi_pl <- ggplot(main_tr, aes(x = value)) +
+  geom_histogram(
+    mapping = aes(fill = val_col),
+    binwidth = .05,
+    boundary = 1
+  ) +
+  geom_text(
+    data = main_annon,
+    mapping = aes(x = Inf, y = Inf, label = lab),
+    size = 8 / cm(1),
+    hjust = 1, vjust = 2
+  ) +
+  facet_wrap(~variable) +
+  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
+  coord_cartesian(expand = FALSE) +
+  scale_x_continuous(
+    breaks = c(0, .05, 0.1, seq(0.2, .5, .1)),
+    labels = function(x) x * 100
+  ) +
+  scale_y_continuous(
+    labels = scales::comma
+  ) +
+  labs(
+    x = "Risk of infection (%)",
+    y = "Number of clinical attendees",
+    title = "b"
+  ) +
+  theme_custom() +
+  theme(
+    legend.position = "bottom",
     legend.title = element_blank(),
-        panel.spacing = unit(2, "cm"))
+    legend.key.height = unit(.6, "cm"),
+    legend.key.width = unit(.4, "cm"),
+    panel.spacing = unit(2, "cm")
+  )
 
-save_plot(roi_pl, pdf_file = "../results/modeling/risk-of-infection.pdf", w = 26, h = 12)
+roi_pl
 ```
@@
 ```{r}
-# load tracking and environmental data
-track <- readRDS("../data-clean/patient-tracking/augmented-and-combined-data.rds")
-aer <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
-  mutate(daytime = factor(daytime, levels = c("Morning", "Afternoon")))
-
 # process tracking data
@@
   group_by(date, patient_id) %>%
-  summarize(dur_morning = sum(h < 12),
+  summarize(
+    dur_morning = sum(h < 12),
     dur_afternoon = sum(h >= 12),
-            tracking_end = tracking_end[1]) %>%
+    tracking_end = tracking_end[1]
+  ) %>%
   ungroup() %>%
@@
   group_by(date, patient_id) %>%
-  summarize(`Clinic` = n(),
+  summarize(
+    `Clinic` = n(),
     `Waiting room` = sum(is_waitingroom),
     `TB room` = sum(is_tbroom),
-            `Corridor` = sum(is_passage)) %>%
+    `Corridor` = sum(is_passage)
+  ) %>%
   ungroup() %>%
@@
 time_close_contacts <- track %>%
-  mutate(has_closeContact = ifelse(sapply(closeContacts, length) >= 1, T, F)) %>%
+  mutate(has_closeContact = ifelse(
+    sapply(closeContacts, length) >= 1, TRUE, FALSE
+  )) %>%
   group_by(date, patient_id) %>%
@@
 # merge data
-track_ft <- left_join(track_daytime, left_join(left_join(time_in_clin, no_close_contacts), time_close_contacts)) %>%
+track_ft <- left_join(
+  track_daytime,
+  left_join(left_join(
+    time_in_clin,
+    no_close_contacts
+  ), time_close_contacts)
+) %>%
   mutate(n_cc = ifelse(is.na(n_cc), 0, n_cc)) %>%
   rename(time_in_clinic = `Clinic`) %>%
-  dplyr::select(date, patient_id, tracking_end, daytime, time_in_clinic, n_cc, cc_time)
+  dplyr::select(
+    date,
+    patient_id,
+    tracking_end,
+    daytime,
+    time_in_clinic,
+    n_cc,
+    cc_time
+  )
+
 assoc_df <- track_ft %>%
-  left_join(aer %>% filter(room == "Waiting room") %>% dplyr::select(date, daytime, aer)) %>%
-  mutate(patient_id = paste(lubridate::month(date), lubridate::day(date), patient_id, sep = "-")) %>%
-  left_join(reshape2::dcast(mean_main_tr, patient_id ~ variable) %>% dplyr::select(patient_id, `Mean risk`) %>% rename(mean_risk = `Mean risk`))
+  left_join(
+    aer %>%
+      filter(room == "Waiting room") %>%
+      dplyr::select(date, daytime, aer)
+  ) %>%
+  mutate(patient_id = paste(
+    lubridate::month(date),
+    lubridate::day(date),
+    patient_id,
+    sep = "-"
+  )) %>%
+  left_join(
+    reshape2::dcast(main_tr, patient_id ~ variable) %>%
+      dplyr::select(patient_id, `Mean risk of infection`) %>%
+      rename(mean_risk = `Mean risk of infection`)
+  )
 
@@
 amod <- stan_betareg(mean_risk ~ aer + time_in_clinic + n_cc + cc_time,
-                     data = assoc_df %>% mutate(n_cc = n_cc + 1, cc_time = cc_time + 1/60,
-                                                across(c(aer, time_in_clinic, n_cc, cc_time), log2)), 
-                     seed = 12345, cores = 4)
+  data = assoc_df %>% mutate(
+    n_cc = n_cc + 1, cc_time = cc_time + 1 / 60,
+    across(c(aer, time_in_clinic, n_cc, cc_time), log2)
+  ),
+  seed = 12345, cores = 4
+)
 
@@
   geom_vline(aes(xintercept = 1), linetype = "dotted", color = "red") +
-  stat_pointinterval() +
-  labs(x = "Odds ratio associated with a doubling of the predictor",
-       title = "Predictors of infection risk",
-       subtitle = "Association of close contact, time in clinic, and air change rate with mean risk of infection",
-       caption = "Note: Median as dots, 66%- and 95%-CrI as thick and thin lines.") +
-  scale_y_discrete(labels = c("Air change rate", "Time in clinic", "Number of close contacts", "Time in clocse contact")) +
-  theme_custom(14) +
+  stat_interval() +
+  stat_summary(
+    geom = "point",
+    fun = median,
+    size = 2,
+    shape = 21,
+    fill = "white"
+  ) +
+  labs(
+    x = "Odds ratio of the modeled risk of infection for a doubling in the predictor",
+    title = "c",
+    color = "Credible interval"
+  ) +
+  scale_y_discrete(
+    labels = c(
+      "Air change rate",
+      "Time in clinic",
+      "Number of close contacts",
+      "Time in clocse contact"
+    )
+  ) +
+  scale_color_brewer(
+    labels = paste0(c(95, 80, 50), "%")
+  ) +
+  theme_custom() +
   theme(axis.title.y = element_blank())
 
-save_plot(amod_est_pl, pdf_file = "../results/modeling/association-predictors.pdf", w = 20, h = 12)
+amod_est_pl
 ```
+
+```{r main-figure}
+main_fig <- grid.arrange(
+  mean_quanta_pl,
+  roi_pl,
+  amod_est_pl,
+  ncol = 1,
+  heights = c(8, 8, 4)
+)
+
+ggsave(
+  main_fig,
+  filename = "results/modeling/main-figure.png",
+  width = 16 / cm(1),
+  height = 20 / cm(1)
+)
+```
+
+## Additional analyses
+
+```{r additional-analyses}
+# files
+files_nm <- list.files("simulations/nomasks/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
+)
+files_vent <- list.files("simulations/ventilation/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
+)
+files_temp <- list.files("simulations/temporal/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
+)
+files_prev <- list.files("simulations/prevalence/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
+)
+files_susp <- list.files("simulations/suspected/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
+)
+
+# average risk
+nm_tr <- average_trans_risk(files_nm)
+vent_tr <- average_trans_risk(files_vent)
+temp_tr <- average_trans_risk(files_temp)
+prev_tr <- average_trans_risk(files_prev)
+susp_tr <- average_trans_risk(files_susp)
+
+tr_comp <- rbind(
+  nm_tr %>% mutate(scenario = "No mask wearing"),
+  vent_tr %>% mutate(scenario = "Lower ventilation"),
+  prev_tr %>% mutate(scenario = "Prevalence-based model"),
+  susp_tr %>% mutate(scenario = "Suspected-based model")
+) %>%
+  mutate(
+    scenario = factor(scenario, levels = c(
+      "Spatiotemporal model",
+      "Temporal model",
+      "No mask wearing",
+      "Lower ventilation",
+      "Prevalence-based model",
+      "Suspected-based model"
+    ))
+  )
+
+# annotation
+nm_annon <- annotate_average_trans_risk(nm_tr)
+vent_annon <- annotate_average_trans_risk(vent_tr)
+temp_annon <- annotate_average_trans_risk(temp_tr)
+prev_annon <- annotate_average_trans_risk(prev_tr)
+susp_annon <- annotate_average_trans_risk(susp_tr)
+
+annon_comp <- rbind(
+  nm_annon %>% mutate(scenario = "No mask wearing"),
+  vent_annon %>% mutate(scenario = "Lower ventilation"),
+  prev_annon %>% mutate(scenario = "Prevalence-based model"),
+  susp_annon %>% mutate(scenario = "Suspected-based model")
+) %>%
+  mutate(
+    scenario = factor(scenario, levels = c(
+      "No mask wearing",
+      "Lower ventilation",
+      "Prevalence-based model",
+      "Suspected-based model"
+    ))
+  )
+
+# plot mean
+mean_roi_comp_pl <- ggplot(
+  tr_comp %>%
+    filter(variable == "Mean risk of infection"),
+  aes(x = value)
+) +
+  geom_histogram(
+    mapping = aes(fill = val_col),
+    binwidth = .05,
+    boundary = 1
+  ) +
+  geom_text(
+    data = annon_comp %>% filter(variable == "Mean risk of infection"),
+    mapping = aes(x = Inf, y = Inf, label = lab),
+    size = 8 / cm(1),
+    hjust = 1, vjust = 2
+  ) +
+  facet_wrap(~scenario, ncol = 2) +
+  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
+  coord_cartesian(expand = FALSE) +
+  scale_x_continuous(
+    breaks = c(0, .05, 0.1, seq(0.2, .5, .1)),
+    labels = function(x) x * 100
+  ) +
+  scale_y_continuous(
+    labels = scales::comma
+  ) +
+  labs(
+    x = "Risk of infection (%)",
+    y = "Number of clinical attendees",
+    title = "b"
+  ) +
+  theme_custom() +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    legend.key.height = unit(.6, "cm"),
+    legend.key.width = unit(.4, "cm"),
+    panel.spacing = unit(2, "cm")
+  )
+
+save_plot(
+  mean_roi_comp_pl,
+  "results/modeling/mean-roi-comparison.png",
+  w = 16,
+  h = 12
+)
+
+# comparison with temporal model
+mod_comp <- left_join(
+  temp_tr %>% rename(`Temporal model` = value) %>% dplyr::select(-val_col),
+  main_tr %>% mutate(`Spatiotemporal model` = value) %>% dplyr::select(-val_col)
+) %>%
+  mutate(
+    diff = `Spatiotemporal model` - `Temporal model`
+  ) %>%
+  mutate(
+    val_col = ifelse(diff > .005, "Higher",
+      ifelse(diff < -.005, "Lower", "Similar")
+    ),
+    val_col = factor(val_col, levels = c("Lower", "Similar", "Higher"))
+  )
+
+mod_comp_annon <- mod_comp %>%
+  summarise(
+    mean_diff = mean(diff),
+    lower_diff = quantile(diff, .025),
+    upper_diff = quantile(diff, .975)
+  ) %>%
+  mutate(across(c(mean_diff, lower_diff, upper_diff), ~ round_k(100 * .x, 1)),
+    lab = paste0(
+      "Average ", mean_diff, "%",
+      " (95%-CrI ", lower_diff, "% – ", upper_diff, "%)"
+    )
+  )
+
+mod_comp_pl <- mod_comp %>%
+  ggplot(aes(x = diff)) +
+  geom_histogram(
+    mapping = aes(fill = val_col),
+    binwidth = .005,
+    boundary = 1
+  ) +
+  geom_text(
+    data = mod_comp_annon,
+    mapping = aes(x = Inf, y = Inf, label = lab),
+    size = 8 / cm(1),
+    hjust = 1, vjust = 2
+  ) +
+  labs(
+    y = "Number of clinical attendees (log-scale)",
+    x = "Difference between spatiotemporal and temporal model (perc. points)"
+  ) +
+  scale_y_log10(labels = scales::comma) +
+  scale_x_continuous(labels = function(x) scales::percent(x, suffix = "")) +
+  scale_fill_manual(values = wes_palette("Zissou1")[c(1, 3, 5)]) +
+  coord_cartesian(expand = FALSE) +
+  theme_custom() +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    legend.key.height = unit(.6, "cm"),
+    legend.key.width = unit(.4, "cm")
+  )
+
+save_plot(
+  mod_comp_pl,
+  "results/modeling/model-comparison.png",
+  w = 12,
+  h = 8
+)
+```

### 2024-03-04 analysis/clinical-data.Rmd
First batch of plots and results

--- 2023-12-06 analysis/clinical-data.Rmd
@@
 
-source("../utils/plotting.r")
+source("utils/plotting.r")
 ```
@@
 ```{r}
-df <- readRDS("../data-clean/clinical/clinical-data.rds") %>%
+df <- readRDS("data-clean/clinical/clinical-data.rds") %>%
   filter(between(month(date), 10, 11))
 
-modeled_days <- list.files("../data-clean/patient-tracking/linked-and-imputed")
+modeled_days <- list.files("data-clean/patient-tracking/linked-and-imputed")
 modeled_days <- gsub("[[:punct:]]rds", "", modeled_days)
 modeled_days <- as.Date(modeled_days)
+modeled_days_abbrv <- format(modeled_days, format = "%b %d")
 ```
@@
 diag_tb <- df %>%
-  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
-  group_by(date, `Date modeled`) %>%
-  summarize(n_infectious = sum(is_infectious)) %>%
-  ungroup()
+  mutate(
+    date_in_model = format(as.Date(date), format = "%b %d"),
+    date_in_model = ifelse(
+      date_in_model %in% modeled_days_abbrv, date_in_model,
+      "Not modeled"
+    ),
+    date_in_model = factor(
+      date_in_model,
+      levels = c(modeled_days_abbrv, "Not modeled")
+    )
+  ) %>%
+  group_by(date, date_in_model) %>%
+  summarize(
+    n_infectious = sum(is_infectious),
+    n_suspected = sum(was_suspected & !is_infectious)
+  ) %>%
+  ungroup() %>%
+  reshape2::melt(c("date", "date_in_model")) %>%
+  mutate(
+    variable = ifelse(grepl("infectious", variable), "Diagnosed", "Suspected"),
+    variable = factor(variable, levels = c("Suspected", "Diagnosed"))
+  )
 
-diag_tb %>%
-  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
-  geom_bar(stat = "identity") +
+diag_tb_pl <- diag_tb %>%
+  ggplot(aes(x = date, y = value, fill = date_in_model, alpha = variable)) +
+  geom_bar(position = position_stack(), stat = "identity") +
   scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
-  scale_y_continuous(expand = c(0,0)) +
-  theme_custom() +
-  theme(legend.position = "top")
+  scale_y_continuous(expand = c(0, 0)) +
+  scale_fill_manual(values = c(wes_palette("Moonrise3"), "grey")) +
+  scale_alpha_discrete(
+    breaks = c("Diagnosed", "Suspected"),
+    range = c(.33, 1)
+  ) +
+  labs(
+    y = "Number of TB patients attending the clinic"
+  ) +
+  theme_custom(8) +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    legend.direction = "vertical",
+    legend.box = "vertical",
+    axis.title.x = element_blank(),
+    axis.text.x = element_blank(),
+    legend.key.width = unit(.2, "cm"),
+    legend.key.height = unit(.3, "cm")
+  ) +
+  guides(
+    fill = guide_legend(nrow = 1, order = 2),
+    alpha = guide_legend(nrow = 1, order = 1)
+  )
+
+save_plot(
+  diag_tb_pl,
+  pdf_file = "results/data/diagnosed-tb-patients.png",
+  w = 16, h = 12
+)
 ```
@@
 diag_tb %>%
-  ggplot(aes(x = n_infectious)) +
+  filter(variable == "Diagnosed") %>%
+  ggplot(aes(x = value)) +
   geom_histogram()
+
+table(diag_tb$value[diag_tb$variable == "Diagnosed"])
 
@@
 #' --> Assume a Dirichlet prior with support according to the counts in the range (1,5).
+
+source("utils/trans_risk.r")
+
+undiag_tb_pl <- data.frame(x = rundiag(1e5)) %>%
+  group_by(x) %>%
+  summarize(n = n()) %>%
+  ungroup() %>%
+  mutate(
+    p = 100 * n / sum(n),
+    p = round(p)
+  ) %>%
+  ggplot(aes(x = factor(x), y = p)) +
+  geom_bar(
+    stat = "identity",
+    color = wes_palette("GrandBudapest1")[3],
+    fill = wes_palette("GrandBudapest1")[4],
+    alpha = .5
+  ) +
+  labs(
+    x = "Number of undiagnosed TB patients", y = "Probability (%)"
+  ) +
+  coord_cartesian(expand = F) +
+  scale_y_continuous(limits = c(0, NA)) +
+  theme_custom(8) +
+  theme(plot.caption = element_text(size = 10))
+
+save_plot(
+  undiag_tb_pl,
+  pdf_file = "results/inputs/undiagnosed-tb-patients.png",
+  w = 8, h = 6
+)
 ```
-
-### Suspected TB over time
-
-```{r}
-df %>%
-  mutate(`Date modeled` = ifelse(date %in% modeled_days, "Yes", "No")) %>%
-  group_by(date, `Date modeled`) %>%
-  summarize(n_infectious = sum(was_suspected & !is_infectious)) %>%
-  ungroup() %>%
-  ggplot(aes(x = date, y = n_infectious, fill = `Date modeled`)) +
-  geom_bar(stat = "identity") +
-  scale_x_date(date_labels = "%b %d", breaks = modeled_days) +
-  scale_y_continuous(expand = c(0,0)) +
-  theme_custom() +
-  theme(legend.position = "top")
-```

### 2024-03-04 analysis/environmental-data.Rmd
First batch of plots and results

--- 2023-12-06 analysis/environmental-data.Rmd
@@
 ```{r}
-dates <- sort(unique(cont_df$date))
-abbr_dates <- paste(month(dates, label = T, abbr = T), day(dates))
-cont_df <- readRDS("../data-clean/environmental/continuous-iaq-data.rds") %>%
-  mutate(time = as.POSIXct(format(date_time, format = "%H:%M:%S"), format = "%H:%M:%S"),
-         date = factor(date, levels = dates))
-aer <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
-  mutate(mth = month(date, label = T, abbr= T),
-         dy = day(date),
-         mthdy = paste(mth, dy),
-         mthdy = factor(as.character(mthdy), levels = abbr_dates),
-         daytime = factor(daytime, levels = c("Morning", "Afternoon")))
+dates <- as.Date(c(
+  "2021-10-13",
+  "2021-10-15",
+  "2021-10-25",
+  "2021-11-04",
+  "2021-11-05"
+))
+abbr_dates <- format(dates, format = "%b %d")
+cont_df <- readRDS("data-clean/environmental/continuous-iaq-data.rds") %>%
+  mutate(
+    time = as.POSIXct(
+      format(date_time, format = "%H:%M:%S"),
+      format = "%H:%M:%S"
+    ),
+    mthdy = factor(
+      as.character(format(date, format = "%b %d")),
+      levels = abbr_dates
+    )
+  )
+aer <- readRDS("data-clean/environmental/air-change-rate.rds") %>%
+  mutate(
+    mthdy = factor(
+      as.character(format(date, format = "%b %d")),
+      levels = abbr_dates
+    ),
+    daytime = factor(daytime, levels = c("Morning", "Afternoon"))
+  )
 ```
@@
 ```{r}
-cont_df %>%
-  ggplot(aes(x = time, y = co2, color = date)) +
+co2_pl <- cont_df %>%
+  mutate(time = floor_date(time, "10 minutes")) %>%
+  group_by(mthdy, time, room) %>%
+  summarize(co2 = mean(co2)) %>%
+  ungroup() %>%
+  mutate(room = factor(
+    room,
+    levels = c("Waiting room", "Corridor", "TB room")
+  )) %>%
+  ggplot(aes(x = time, y = co2, color = mthdy)) +
   geom_step() +
-  facet_wrap(~ room) +
-  coord_cartesian(expand = F) +
-  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
+  facet_wrap(~room) +
+  coord_cartesian(expand = FALSE) +
+  scale_x_datetime(date_breaks = "4 hours", date_labels = "%H:%M") +
   scale_color_manual(values = wes_palette("Moonrise3")) +
-  labs(y = "Number of people in waiting room") +
-  theme_custom() +
-  theme(axis.title.x = element_blank(),
-        legend.title = element_blank())
+  labs(
+    y = expression("CO"[2] * " (parts per million, ppm)"),
+    title = "a"
+  ) +
+  theme_custom(8) +
+  theme(
+    axis.title.x = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom",
+    panel.spacing = unit(1, "lines"),
+    legend.key.width = unit(0.275, "cm"),
+    legend.key.height = unit(0.2, "cm")
+  )
+
+save_plot(
+  co2_pl,
+  pdf_file = "results/data/co2-levels-over-time.png",
+  w = 10, h = 6
+)
 ```
@@
 ```{r}
-aer %>%
-  ggplot(aes(x = mthdy, fill = daytime, y = aer)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 1, preserve = "single"), width = .66) +
-  facet_wrap(~ room) +
-  scale_fill_brewer(palette = 9) +
-  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
-  theme_custom()
+aer_pl <- aer %>%
+  mutate(room = factor(
+    room,
+    levels = c("Waiting room", "Corridor", "TB room")
+  )) %>%
+  ggplot(aes(x = room, fill = room, y = aer)) +
+  geom_boxplot() +
+  # geom_jitter(
+  #   mapping = aes(color = daytime),
+  #   width = 0.2,
+  #   height = 0,
+  #   shape = 21
+  # ) +
+  scale_x_discrete(labels = c("WR", "TB", "CR")) +
+  scale_fill_manual(values = wes_palette("Royal2")[-1]) +
+  scale_y_continuous(expand = expansion(mult = c(0, .05)), limits = c(0, NA)) +
+  labs(
+    y = "Air changes per hour",
+    title = "b"
+  ) +
+  theme_custom(8) +
+  theme(
+    axis.title.x = element_blank(),
+    legend.title = element_blank(),
+    legend.position = c(0.8, 0.95)
+  ) +
+  guides(fill = "none")
+
+save_plot(
+  aer_pl,
+  pdf_file = "results/data/air-changes-per-hour.png",
+  w = 6, h = 6
+)
 ```

### 2024-03-04 analysis/patient-tracking.Rmd
First batch of plots and results

--- 2023-12-06 analysis/patient-tracking.Rmd
@@
 
-source("../utils/plotting.r")
-source("../utils/spatial.r")
-
-text_descr <- 8
+source("utils/plotting.r")
+source("utils/spatial.r")
 ```
@@
 ```{r}
-df <- readRDS("../data-clean/patient-tracking/augmented-and-combined-data.rds")
+df <- readRDS("data-clean/patient-tracking/augmented-and-combined-data.rds")
+
+# filter considered clinical attendees
+df_matched <- do.call(
+  rbind,
+  lapply(
+    unique(df$date),
+    function(d) {
+      readRDS(paste0(
+        "data-clean/patient-tracking/matched-clinical/",
+        d, ".rds"
+      ))
+    }
+  )
+) %>%
+  mutate(date = as.Date(time))
+df_filt <- filter(
+  df,
+  paste(date, patient_id) %in%
+    unique(paste(df_matched$date, df_matched$patient_id))
+)
 ```
@@
 ```{r}
-source("../helper/building.R")
+cellSize <- 250
+source("preprocessing/prep_building-rasterize.R")
 ```
@@
 
+```{r}
+n_distinct(paste(df$date, df$patient_id))
+n_distinct(paste(df_filt$date, df_filt$patient_id))
+```
+
 ### Tracking quality
@@
   table()
+
+sum(track_qual[names(track_qual) != "Possible HCW"])
+track_qual_str <- paste0(
+  names(track_qual),
+  ": ",
+  track_qual,
+  " (",
+  round(track_qual / sum(track_qual) * 100),
+  "%)"
+)
+track_qual_str
 
@@
 ```{r}
-time_in_clin <- df %>%
+time_in_clin <- df_filt %>%
   group_by(date, patient_id) %>%
-  summarize(tracking_end = first(tracking_end),
+  summarize(
+    tracking_end = first(tracking_end),
     `Clinic` = n(),
@@
     `TB room` = sum(is_tbroom),
-            `Corridor` = sum(is_passage)) %>%
+    `Corridor` = sum(is_passage)
+  ) %>%
   ungroup() %>%
@@
   melt(c("date", "patient_id")) %>%
-  mutate(variable = factor(variable, levels = c("Clinic", "Waiting room", "TB room", "Corridor"))) %>%
+  mutate(variable = factor(
+    variable,
+    levels = c("Clinic", "Waiting room", "TB room", "Corridor")
+  )) %>%
   ggplot(aes(x = variable, fill = variable, y = value)) +
   geom_boxplot() +
-  scale_y_sqrt()
+  scale_y_sqrt(breaks = c(0, 1, 10, 30, 60, 120, 180)) +
+  scale_x_discrete(labels = c("Clinic", "WR", "TB", "CR")) +
+  scale_fill_manual(values = wes_palette("Royal2")) +
+  labs(
+    y = "Time in clinic (minutes)",
+    title = "c"
+  ) +
+  theme_custom(8) +
+  theme(
+    legend.position = "none",
+    legend.title = element_blank(),
+    axis.title.x = element_blank(),
+    plot.title.position = "plot"
+  )
+
+save_plot(
+  time_in_clin_pl,
+  pdf_file = "results/data/time-in-clinic.png", w = 5.3, h = 6
+)
 ```
@@
 # number of close contacts <1m for >1min
-no_close_contacts <- df %>%
+no_close_contacts <- df_filt %>%
   unnest(cols = closeContacts) %>%
   group_by(date, patient_id, closeContacts, tracking_end) %>%
-  summarise(n_time = n(),
-            tracking_end = first(tracking_end)) %>%
+  summarise(
+    n_time = n(),
+    tracking_end = first(tracking_end)
+  ) %>%
   ungroup() %>%
   group_by(date, patient_id) %>%
-  summarize(n_cc = sum(n_time > 60),
-            tracking_end = first(tracking_end)) %>%
+  summarize(
+    n_cc = sum(n_time > 60),
+    tracking_end = first(tracking_end)
+  ) %>%
   ungroup()
 
-time_close_contacts <- df %>%
-  mutate(has_closeContact = ifelse(sapply(closeContacts, length) >= 1, T, F)) %>%
+time_close_contacts <- df_filt %>%
+  mutate(has_closeContact = ifelse(
+    sapply(closeContacts, length) >= 1, T, F
+  )) %>%
   group_by(date, patient_id) %>%
-  summarize(cc_time = sum(has_closeContact) / 60,
-            tracking_end = first(tracking_end)) %>%
+  summarize(
+    cc_time = sum(has_closeContact) / 60,
+    tracking_end = first(tracking_end)
+  ) %>%
   ungroup()
@@
 
-close_contact %>%
+cc_n_pl <- close_contact %>%
   filter(tracking_end != "Possible HCW") %>%
-  ggplot(aes(x = Clinic, y = n_cc)) +
-  geom_point()
+  ggplot(aes(x = n_cc)) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  labs(
+    y = "Density", x = "Number of close contacts",
+    title = "d"
+    # subtitle = "Defined as <1m for >1min"
+  ) +
+  geom_density(fill = wes_palette("FantasticFox1")[2], color = wes_palette("FantasticFox1")[2], alpha = 0.8) +
+  theme_custom(8) +
+  theme(
+    plot.title.position = "plot",
+    plot.margin = margin(r = 11),
+    plot.subtitle = element_text(margin = margin(b = 11))
+  )
 
-close_contact %>%
+save_plot(
+  cc_n_pl,
+  pdf_file = "results/data/number-close-contacts.png", w = 5.3, h = 6
+)
+
+cc_t_pl <- close_contact %>%
   filter(tracking_end != "Possible HCW") %>%
-  ggplot(aes(x = 1, y = cc_time / Clinic)) +
-  geom_boxplot() 
+  ggplot(aes(x = cc_time / Clinic * 100)) +
+  scale_x_continuous(expand = c(0, 0), limits = c(0, 100)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  labs(
+    y = "Density",
+    x = "Time in close contact (%)",
+    title = "e"
+  ) +
+  geom_density(
+    fill = wes_palette("Moonrise1")[2],
+    color = wes_palette("Moonrise1")[2],
+    alpha = 0.8
+  ) +
+  theme_custom(8) +
+  theme(
+    plot.title.position = "plot",
+    plot.margin = margin(r = 11)
+  )
+
+save_plot(
+  cc_t_pl,
+  pdf_file = "results/data/time-close-contacts.png", w = 5.3, h = 6
+)
 ```
@@
     `TB room` = sum(is_tbroom),
-    `Corridor` = sum(is_passage)) %>%
+    `Corridor` = sum(is_passage)
+  ) %>%
   ungroup() %>%
   melt(c("date", "time")) %>%
-  mutate(variable = factor(variable, levels = c("Clinic", "Waiting room", "TB room", "Corridor")),
+  mutate(
+    variable = factor(
+      variable,
+      levels = c("Clinic", "Waiting room", "TB room", "Corridor")
+    ),
     date = format(date, "%b %d"),
-         time = as.POSIXct(format(time, format = "%H:%M:%S"), format = "%H:%M:%S")) %>%
+    date = factor(
+      as.character(date),
+      levels = c("Oct 13", "Oct 15", "Oct 25", "Nov 04", "Nov 05")
+    ),
+    time = as.POSIXct(format(time, format = "%H:%M:%S"), format = "%H:%M:%S")
+  ) %>%
   filter(between(hour(time), 7, 16))
 
-clinic_occupancy %>%
+time_pl <- clinic_occupancy %>%
+  filter(variable == "Clinic") %>%
   ggplot(aes(x = time, y = value, color = date)) +
   geom_step() +
-  facet_wrap(~ variable) +
-  coord_cartesian(expand = F) +
-  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H:%M") +
+  # ggh4x::facet_wrap2(~variable, axes = "x", remove_labels = "all") +
+  coord_cartesian(expand = FALSE) +
+  scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
   scale_color_manual(values = wes_palette("Moonrise3")) +
-  labs(y = "Number of people in waiting room") +
-  theme_custom() +
-  theme(axis.title.x = element_blank(),
-        legend.title = element_blank())
+  labs(
+    y = "Number of people in clinic",
+    title = "f"
+  ) +
+  theme_custom(8) +
+  theme(
+    axis.title.x = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom",
+    legend.key.width = unit(0.275, "cm"),
+    legend.key.height = unit(0.2, "cm")
+  )
+
+save_plot(
+  time_pl,
+  pdf_file = "results/data/no-people-over-time.png",
+  w = 8, h = 6
+)
 ```
@@
 ```{r}
-col <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")
+spatial_pl <- plot_spatial(
+  ggplot(data = df, mapping = aes(x = x, y = y)) +
+    geom_bin2d(binwidth = 200),
+  text_descr = 8
+) +
+  labs(
+    fill = "Number of tracks",
+    title = "g"
+  ) +
+  scale_fill_gradientn(
+    colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
+    trans = "log10",
+    labels = scales::comma,
+    breaks = c(1, 10, 100, 1000, 10000)
+  ) +
+  theme(
+    legend.position = "bottom",
+    legend.direction = "horizontal",
+    legend.key.height = unit(0.2, "cm"),
+    legend.key.width = unit(1, "cm")
+  ) +
+  guides(fill = guide_colorbar(title.vjust = 1))
 
-plot_spatial <- function(pl) {
-  pl +
-  ggpubr::background_image(clinic_img) +
-  scale_y_continuous(expand = expansion(add=c(1700, 3000)), limits = c(-6500, 2000)) +
-  scale_x_continuous(expand = expansion(add=c(400, 500)), limits = c(-16000, 7500)) +
-  annotation_custom(entrance_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(wr_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(tb_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  annotation_custom(re_lab, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
-  theme_custom() +
-  theme(text = element_text(size = text_descr),
-        axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
-        legend.position = c(0.825,0.05), legend.direction = "horizontal", 
-        legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),
-        panel.background = element_rect(fill = col[1]), panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank()) +
-  guides(fill = guide_legend(title.position = "top"))
-}
-
-plot_spatial(
-  ggplot(data = df, mapping = aes(x = x, y = y)) +
-    geom_bin2d(binwidth = 200) ) +
-  labs(fill = "Number of tracks") +
-  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"), trans = "log10", 
-                    breaks = c(1, 10, 100, 1000, 10000)) 
+save_plot(
+  spatial_pl,
+  pdf_file = "results/data/no-people-spatial.png",
+  w = 8, h = 6
+)
 ```
@@
   slice(1, n()) %>%
-  mutate(type = factor(c("Start", "End"), levels=c("Start", "End"))) %>%
+  mutate(type = factor(c("Start", "End"), levels = c("Start", "End"))) %>%
   ungroup()
@@
 # plot
-plot_spatial(
+ex_tracks <- plot_spatial(
   ggplot() +
-    geom_path(data = ex_df_man, mapping = aes(x = x, y = y), color = "darkred") +
-    geom_point(data = ex_df_man_se, mapping = aes(x = x, y = y, shape = type), size = 2) +
-    facet_wrap(~ day_pid) +
-    scale_shape_manual(values = c(1,13)) 
+    geom_path(
+      data = ex_df_man,
+      mapping = aes(x = x, y = y),
+      color = "darkred"
     ) +
-  theme(legend.position = "none")
+    geom_point(
+      data = ex_df_man_se,
+      mapping = aes(x = x, y = y, shape = type),
+      size = 2
+    ) +
+    facet_wrap(~day_pid, ncol = 3) +
+    scale_shape_manual(values = c(1, 13)),
+  text_descr = 8
+) +
+  theme(
+    legend.position = "none"
+  )
+
+save_plot(
+  ex_tracks,
+  pdf_file = "results/data/example-patient-tracks.png",
+  w = 16, h = 12
+)
 ```

### 2024-03-04 illustrations/priors.Rmd
First batch of plots and results

--- 2023-07-18 illustrations/priors.Rmd
@@
 library(LaplacesDemon)
-source("../utils/trans_risk.r")
-source("../utils/plotting.r")
+source("utils/trans_risk.r")
+source("utils/plotting.r")
 ```
 
-## Quanta
+## Generation rate
 
 ```{r}
-q_dat <- data.frame(x = seq(0, 200, 0.1)) %>%
-  mutate(y = pq(x))
+set.seed(12345)
+
+q_dat <- rbind(
+  data.frame(
+    x = rq(bact_load.mtb(1e6), activity = "waiting"),
+    activity = "Waiting"
+  ),
+  data.frame(
+    x = rq(bact_load.mtb(1e6), activity = "walking"),
+    activity = "Walking"
+  )
+)
+
+q_dat %>%
+  group_by(activity) %>%
+  tidybayes::median_qi(x, .width = .5)
+
+q_dat_dens <- q_dat %>%
+  group_by(activity) %>%
+  summarize(dens = list(density(x, adjust = 2)))
 
 prev_est <- data.frame(
-  study = c("Andrews (2014)", "Escombe (2008)", "Nardell (1991)", "Riley (1962)"),
-  x = c(0.89, 8.2, 12.5, 1.25)
+  study = c(
+    "Andrews (2014)",
+    "Escombe (2008)",
+    "Nardell (1991)",
+    "Riley (1962)"
+  ),
+  x = c(0.89, 8.2, 12.5, 1.25),
+  y = c(1, 7, 9, 3) / 1e2
 ) %>%
-  mutate(y = pq(x))
+  mutate(lab = paste0(study, ": ", round(x, 2)))
 
-q_pl <- ggplot(mapping = aes(x = x, y = y)) +
-  geom_line(data = q_dat, direction = "vh") +
-  geom_segment(data = prev_est, mapping = aes(xend = x, yend = 0, color = study), linetype = "dotted") +
-  geom_segment(data = prev_est, mapping = aes(yend = y, xend = 0, color = study), linetype = "dotted") +
-  geom_point(data = prev_est, mapping = aes(color = study), shape = 17, size = 2) +
-  geom_text(data = prev_est, mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35)) +
-  scale_x_continuous(expand = expansion(mult = c(0,.025)), breaks = c(0, 1, 10, 100, 200), trans = scales::pseudo_log_trans(base = 10)) +
-  scale_y_continuous(expand = c(0,0), labels = scales::percent_format(suffix = "")) +
+q_pl <- ggplot() +
+  geom_density(
+    data = q_dat,
+    mapping = aes(x = x, linetype = activity),
+    adjust = 2
+  ) +
+  geom_segment(
+    data = prev_est,
+    mapping = aes(x = x, xend = x, y = 0, yend = y, color = study),
+    linetype = "dotted"
+  ) +
+  geom_point(
+    data = prev_est,
+    mapping = aes(x = x, color = study, y = y),
+    shape = 17, size = 2
+  ) +
+  geom_text(
+    data = prev_est,
+    mapping = aes(x = x, label = lab, color = study, y = y),
+    size = 8 / cm(1), vjust = -2, hjust = 1
+  ) +
+  scale_x_log10(
+    expand = expansion(mult = c(0, .025)),
+    breaks = c(1e-4, 1e-5, 1e-3, 1e-2, 1e-1, 1e0, 1e1, 1e2, 1e3, 1e4, 1e5)
+  ) +
+  scale_y_continuous(
+    expand = c(0, 0),
+    labels = scales::percent_format(suffix = "")
+  ) +
   scale_color_manual(values = wes_palette("Cavalcanti1")) +
-  labs(x = "Quanta generation rate (quanta/h)", y = "Cumulative probability (%)") +
+  labs(
+    x = "Quanta generation rate (quanta/h)",
+    y = "Density"
+  ) +
   theme_custom() +
@@
 
-q_pl
-save_plot(q_pl, pdf_file = "../illustrations/prior-q.pdf", w = 12, h = 8)
+save_plot(
+  q_pl,
+  pdf_file = "results/inputs/quanta-generation-rate.png",
+  w = 16, h = 12
+)
 ```
 
-## Viral inactivation rate
+## Inactivation rate
 
 ```{r}
-plambda <- function(x, disease = "TB") {
-  #LaplacesDemon::ptrunc(x, spec = "st", a = 0, b = 10, mu = 0.5, sigma = 1, nu = 5)
-  plnorm(x, meanlog = log(0.5), sdlog = 1)
-}
+set.seed(12345)
 
-lambda_dat <- data.frame(x = seq(0, 10, .01)) %>%
-  mutate(y = plambda(x))
+lambda_dat <- data.frame(x = rlambda(1e6))
+
+lambda_dat %>%
+  tidybayes::median_qi(x, .width = .5)
 
 prev_est_vir <- data.frame(
-  study = c("Loudon (1969)", "Lever (2000)", "Gannon (2007)"),
-  x = c(0.12, 8.3, 0.46)
+  study = c(
+    "Loudon (1969)",
+    "Lever (2000)",
+    "Gannon (2007)",
+    "Klein (2014)"
+  ),
+  x = c(
+    log(2) / 6,
+    log(2) / (5 / 60),
+    log(2) / 1.5,
+    log(2) / (0.5 / (log(20 / 100, 0.5)))
+  ),
+  y = c(15, 5, 10, 20) / 1e2
 ) %>%
-  mutate(y = plambda(x))
+  mutate(lab = paste0(study, ": ", round(x, 2)))
 
-lambda_pl <- ggplot(mapping = aes(x = x, y = y)) +
-  geom_line(data = lambda_dat, direction = "vh") +
-  geom_segment(data = prev_est_vir, mapping = aes(xend = x, yend = 0, color = study), linetype = "dotted") +
-  geom_segment(data = prev_est_vir, mapping = aes(yend = y, xend = 0, color = study), linetype = "dotted") +
-  geom_point(data = prev_est_vir, mapping = aes(color = study), shape = 17, size = 2) +
-  geom_text(data = prev_est_vir %>% dplyr::filter(study != "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_x = log(1.35), nudge_y = .02) +
-  geom_text(data = prev_est_vir %>% dplyr::filter(study == "Lever (2000)"), mapping = aes(label = study, color = study), size = 8 / cm(1), nudge_y = 0.05) +
-  scale_x_continuous(expand = expansion(mult = c(0,.025)), breaks = c(0, 0.5, 1, 2, 5, 10), trans = scales::pseudo_log_trans(base = 10)) +
-  scale_y_continuous(expand = expansion(mult = c(0,.025)), labels = scales::percent_format(suffix = "")) +
+lambda_pl <- ggplot() +
+  geom_density(
+    data = lambda_dat,
+    mapping = aes(x = x),
+    adjust = 2
+  ) +
+  geom_segment(
+    data = prev_est_vir,
+    mapping = aes(x = x, xend = x, y = 0, yend = y, color = study),
+    linetype = "dotted"
+  ) +
+  geom_point(
+    data = prev_est_vir,
+    mapping = aes(x = x, color = study, y = y),
+    shape = 17, size = 2
+  ) +
+  geom_text(
+    data = prev_est_vir,
+    mapping = aes(x = x, label = lab, color = study, y = y),
+    size = 8 / cm(1), vjust = -2
+  ) +
+  scale_x_log10(
+    expand = expansion(mult = c(0, .025)),
+    breaks = c(1e-4, 1e-5, 1e-3, 1e-2, 1e-1, 1e0, 1e1, 1e2, 1e3, 1e4, 1e5)
+  ) +
+  scale_y_continuous(
+    expand = expansion(mult = c(0, .05)),
+    labels = scales::percent_format(suffix = "")
+  ) +
   scale_color_manual(values = wes_palette("Cavalcanti1")) +
-  labs(x = "Viral inactivation rate (quanta/h)", y = "Cumulative probability (%)") +
+  labs(
+    x = "Bacterial inactivation rate (quanta/h)",
+    y = "Density"
+  ) +
   theme_custom() +
@@
 
-lambda_pl
-save_plot(lambda_pl, pdf_file = "../illustrations/prior-lambda.pdf", w = 12, h = 8)
+save_plot(
+  lambda_pl,
+  pdf_file = "results/inputs/bacterial-inactivation-rate.png",
+  w = 16, h = 12
+)
 ```

### 2024-03-04 tests/test_stm_v2.R

@@ -11,19 +11,28 @@ ny <- 20
 dy <- gridCellLength * ny
 dx <- gridCellLength * nx
 roomDim <- c(dx, dy, dz)
-c0 <- matrix(0, nrow = ny, ncol = nx, dimnames = list(as.character(1:ny), as.character(1:nx)))
+c0 <- matrix(
+  0,
+  nrow = ny,
+  ncol = nx,
+  dimnames = list(as.character(1:ny), as.character(1:nx))
+)
 
-# time 
-time <- seq(as.POSIXct("2023-01-01 08:00:00"), as.POSIXct("2023-01-01 10:00:00"), by = "1 sec")
+# time
+time <- seq(
+  as.POSIXct("2023-01-01 08:00:00"),
+  as.POSIXct("2023-01-01 10:00:00"),
+  by = "1 sec"
+)
 
 # assumptions
 aer <- 3 / 3600
 lambda <- 0 / 3600
 deposit <- 0
 
-# assume one infectious person in the middle of the room at the beginning staying there for one hour
+# one infectious person in room center for 1h
 inf <- data.frame(
-  t = 1:(length(time)/2),
+  t = 1:(length(time) / 2),
   x = "20",
   y = "10",
   q = 10 / 3600
@@ -31,52 +40,83 @@ inf <- data.frame(
 
 
 # run model assuming no masks (pd = 1)
-ct <- stm(time = time, c0 = c0, aer = aer, inf = inf, vol = prod(roomDim), cellLength = gridCellLength, pd = 0, lambda = lambda, deposit = deposit)
+ct <- stm(
+  time = time,
+  c0 = c0,
+  aer = aer,
+  inf = inf,
+  vol = prod(roomDim),
+  cellLength = gridCellLength,
+  pd = 0,
+  lambda = lambda,
+  deposit = deposit
+)
 
 # plot results
 ct_to_df <- function(t) {
-  data.frame(ct[,,t]) %>% 
-    add_rownames() %>% 
+  data.frame(ct[, , t]) %>%
+    add_rownames() %>%
     reshape2::melt("rowname") %>%
     rename(x = rowname, y = variable) %>%
-    mutate(y = gsub("X", "", y),
-           across(c(x, y), as.numeric),
-           time = t)
+    mutate(
+      y = gsub("X", "", y),
+      across(c(x, y), as.numeric),
+      time = t
+    )
 }
[…]

### 2024-03-04 doc/paper/manuscript.tex
Updates from Overleaf

@@
 Although such assessments only show the potential (modeled) impact of interventions, they can still provide a first estimate about how contact reductions translate into lower transmission risks in crowded indoor settings.
+Nevertheless, we acknowledge that transmission events are generally difficult to predict.
+Their occurrence depends on many other factors such as airflow, the properties of the pathogen, and variation in infectiousness of the source and susceptibility of the host\cite{Wang2021Science}.
+We discuss these factors in more depth in Supplementary Text~\zref{sec:depth-discussion} in \supp.
 
 % more complex models and our relation
-Despite our modeling efforts, transmission events remain difficult to predict.
-Their occurrence depends on many other factors such as airflow, the properties of the pathogen, and variation in infectiousness of the source and susceptibility of the host\cite{Wang2021Science}.
-We discuss these factors in more depth in Supplementary Text~\zref{sec:depth-discussion} in \supp.
-Importantly, there is large stream of literature considering airflow using computational fluid dynamics (CFD) models to simulate the spatial spread of infectious aerosols in indoor environments, \eg Ref.
+An important factor that is omitted from our model is airflow.
+Poor ventilation settings can generate an airflow where infectious particles circulate rather than being dispersed and removed from the indoor space\cite{Li2021BuildEnv}.
+There is large stream of literature considering airflow using computational fluid dynamics (CFD) models to simulate the spatial spread of infectious aerosols in indoor environments, \eg Ref.
 \cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.

### 2024-03-04 doc/paper/supplements.tex
Updates from Overleaf

@@
 % data and other limitations
-While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations.
+% While we combined multiple data sources to allow spatiotemporal modeling of the generation, spread, and removal of infectious quanta, our data has also limitations.
 First, clinical attendees could not always be tracked perfectly throughout the clinic using our video sensors.

### 2024-03-04 analysis/environmental-data.Rmd
Update plots

@@
   # ) +
-  scale_x_discrete(labels = c("WR", "TB", "CR")) +
   scale_fill_manual(values = wes_palette("Royal2")[-1]) +

### 2024-03-04 analysis/patient-tracking.Rmd
Update plots

@@
   scale_y_sqrt(breaks = c(0, 1, 10, 30, 60, 120, 180)) +
-  scale_x_discrete(labels = c("Clinic", "WR", "TB", "CR")) +
   scale_fill_manual(values = wes_palette("Royal2")) +
   labs(
-    y = "Time in clinic (minutes)",
-    title = "c"
+    y = "Time in clinic (min)",
+    title = "e"
   ) +
@@
     axis.title.x = element_blank(),
+    axis.text.x = element_text(angle = 45, hjust = 1),
     plot.title.position = "plot"
@@
     y = "Density", x = "Number of close contacts",
-    title = "d"
+    title = "f"
     # subtitle = "Defined as <1m for >1min"
   ) +
-  geom_density(fill = wes_palette("FantasticFox1")[2], color = wes_palette("FantasticFox1")[2], alpha = 0.8) +
+  geom_density(fill = "grey70", color = "grey0", alpha = 0.8) +
   theme_custom(8) +
@@
     x = "Time in close contact (%)",
-    title = "e"
+    title = "g"
   ) +
   geom_density(
-    fill = wes_palette("Moonrise1")[2],
-    color = wes_palette("Moonrise1")[2],
-    alpha = 0.8
+    fill = "grey70", color = "grey0", alpha = 0.8
   ) +
@@
     y = "Number of people in clinic",
-    title = "f"
+    title = "d"
   ) +
@@
     fill = "Number of tracks",
-    title = "g"
+    title = "c"
   ) +

### 2024-03-04 analysis/qctr.Rmd
Update plots

@@
     labs(
-      fill = expression("Quanta concentration in quanta/m"^3 * " x 10e"^-3),
+      fill = expression("Mean quanta concentration (quanta/m"^3 * " x 10e"^-3 * " per sec)"),
       title = "a"
@@
       lab = paste0(
-        "Average ", mean_P, "%",
+        mean_P, "%",
         " (95%-CrI ", lower_P, "% – ", upper_P, "%)"
@@
     x = "Risk of infection (%)",
-    y = "Number of clinical attendees",
-    title = "b"
+    y = "Number of clinical attendees"
   ) +

### 2024-03-04 doc/paper/manuscript.tex

@@
 
-...
+\begin{figure}
+    \centering
+    \includegraphics{results/data/co2-levels-over-time.png}
+    \includegraphics{results/data/air-changes-per-hour.png}
+     \includegraphics{results/data/no-people-spatial.png}
+    \includegraphics{results/data/no-people-over-time.png}
+    \includegraphics{results/data/time-in-clinic.png}
+    \includegraphics{results/data/number-close-contacts.png}
+    \includegraphics{results/data/time-close-contacts.png}
+    \caption{\textbf{Environmental conditions and patient movement characteristics.} \textbf{(a)}~CO$_2$ levels per room of the clinic over time per day.
+\textbf{(b)}~Air change rates per room of the clinic.
+\textbf{(c)}~Total number of tracks recorded per unit space in the clinic over the study period.
+\textbf{(d)}~Number of people in the clinic over time per day.
+\textbf{(e)}~Time spent in the clinic per clinical attendee.
+\textbf{(f)}~Density distribution of the number of close contacts per clinical attendee during their visit.
+\textbf{(g)}~Density distribution of the time spent in close contact per clinical attendee during their visit.}
+    \label{fig:input-data-descriptives}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{results/modeling/main-figure.png}
+    \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Average quanta concentration in the morning and afternoon in the clinic.
+\textbf{(b)}~Mean and median risk of infection per clinical attendee.
+Low: <5\%, Medium: 5-10\%, High: >10\%.
+\textbf{(c)}~Association of mean infection risk with patient movement characteristics and environmental conditions.
+}
+    \label{fig:main-modeling-results}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{results/modeling/mean-roi-comparison.png}
+    \caption{\textbf{Scenario modeling results.} Mean risk of infection per clinical attendee if \textbf{(a)}~masks had not been worn by all clinical attendees and staff, \textbf{(b)}~the ventilation rate had been lower as in the year before the COVID-19 pandemic, \textbf{(c)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(d)}~the number of infectious people in the clinic corresponded to the number of diagnosed TB patients and those suspected of TB due to respiratory symptoms when attending the clinic.
+Low: <5\%, Medium: 5-10\%, High: >10\%.}
+    \label{fig:scenario-results}
+\end{figure}
 

### 2024-03-04 analysis/environmental-data.Rmd
Update plots

@@
     y = expression("CO"[2] * " (parts per million, ppm)"),
-    title = "a"
+    title = "f"
   ) +
@@
 ```{r}
+aer %>%
+  group_by(room) %>%
+  tidybayes::median_qi(aer, .width = .5) %>%
+  mutate_if(is.numeric, round)
+
 aer_pl <- aer %>%
@@
     y = "Air changes per hour",
-    title = "b"
+    title = "g"
   ) +

### 2024-03-04 analysis/patient-tracking.Rmd
Update plots

@@
 
+time_in_clin %>%
+  filter(tracking_end != "Possible HCW") %>%
+  tidybayes::median_qi(`Clinic`, .width = .5) %>%
+  mutate_if(is.numeric, round)
+
 time_in_clin_pl <- time_in_clin %>%
@@
     variable,
-    levels = c("Clinic", "Waiting room", "TB room", "Corridor")
+    levels = c("Clinic", "Waiting room", "Corridor", "TB room")
   )) %>%
@@
     y = "Time in clinic (min)",
-    title = "e"
+    title = "c"
   ) +
@@
 
+no_close_contacts %>%
+  filter(tracking_end != "Possible HCW") %>%
+  tidybayes::median_qi(n_cc, .width = .5) %>%
+  mutate_if(is.numeric, round)
+
 time_close_contacts <- df_filt %>%
@@
     y = "Density", x = "Number of close contacts",
-    title = "f"
+    title = "d"
     # subtitle = "Defined as <1m for >1min"
@@
 
+close_contact %>%
+  filter(tracking_end != "Possible HCW") %>%
+  tidybayes::median_qi(cc_time / Clinic * 100, .width = .5) %>%
+  mutate_if(is.numeric, round)
+
 cc_t_pl <- close_contact %>%
@@
     x = "Time in close contact (%)",
-    title = "g"
+    title = "e"
   ) +
@@
   labs(
-    y = "Number of people in clinic",
-    title = "d"
+    y = "Number of tracks",
+    title = "b"
   ) +
@@
     fill = "Number of tracks",
-    title = "c"
+    title = "a"
   ) +

### 2024-03-04 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
+Over the five study days, 894 patients have been registered in the clinical database, of which seven patients were diagnosed with TB and 28 patients were suspected of TB following symptom screening (\supp~Figure~\zref{fig:clinical-data}).
+Overall, 1,701 unique patient movements were identified after processing the video sensor data.
+Most patient movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
+Clinical attendees spent spend around half an hour in the clinic (median 24 minutes, interquartile range [IQR] 12$-$45), and several attendees spent more than an hour in the clinic, most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
+During their visit, close contacts, defined as an interaction with another attendee within 1m for more than 1min, were frequent (median 5, IQR 2$-$9), and several attendees had more than 10 close contacts (\Cref{fig:input-data-descriptives}d).
+The majority of clinical attendees was in close contact with at least one other attendee for more than half the time of their visit (median 68\%, IQR 41\%$-$87\%, \Cref{fig:input-data-descriptives}e).
+The CO$_2$ levels inside the clinic were low, peaking at only about 600\,ppm in the morning (\Cref{fig:input-data-descriptives}f).
+Based on room size, time-varying occupancy, and CO$_2$ levels, the estimated air change rates indicated that the clinic was well-ventilated, especially the waiting room (median 29 air changes per hour, IQR 9$-$36) and corridor (median 28, IQR 16$-$36), with lower rates in the less frequently occupied TB room (median 6, IQR 5$-$12, \Cref{fig:input-data-descriptives}g).
+
 \begin{figure}

### 2024-03-07 analysis/clinical-data.Rmd
Update plots
Modeling fixes and visual updates

@@
   ) %>%
+  group_by(patient_id, date) %>%
+  slice(1) %>%
+  ungroup() %>%
   group_by(date, date_in_model) %>%
@@
   )
+
+diag_tb %>%
+  filter(date_in_model != "Not modeled") %>%
+  group_by(variable) %>%
+  summarise(sum(value))
 
@@
   pdf_file = "results/data/diagnosed-tb-patients.png",
-  w = 16, h = 12
+  w = 16, h = 10
 )
@@
 
-table(diag_tb$value[diag_tb$variable == "Diagnosed"])
+counts <- table(diag_tb$value[diag_tb$variable == "Diagnosed"])
+counts
+df_counts <- data.frame(
+  x = names(counts),
+  n = as.numeric(counts),
+  p = round(100 * as.numeric(counts) / sum(counts))
+)
 
-#' Assumption:
-#' Number of diagnosed ~ number of diagnosed TB
-#' --> Assume a Dirichlet prior with support according to the counts in the range (1,5).
-
-source("utils/trans_risk.r")
-
-undiag_tb_pl <- data.frame(x = rundiag(1e5)) %>%
-  group_by(x) %>%
-  summarize(n = n()) %>%
-  ungroup() %>%
-  mutate(
-    p = 100 * n / sum(n),
-    p = round(p)
-  ) %>%
+undiag_tb_pl <- df_counts %>%
   ggplot(aes(x = factor(x), y = p)) +
@@
     stat = "identity",
-    color = wes_palette("GrandBudapest1")[3],
-    fill = wes_palette("GrandBudapest1")[4],
-    alpha = .5
+    color = "grey0",
+    fill = "grey70",
+    alpha = .8
+  ) +
+  geom_text(
+    aes(label = paste0("n" = n, " (", p, "%)")),
+    vjust = -1,
+    size = 8 / cm(1)
   ) +
@@
   ) +
-  coord_cartesian(expand = F) +
-  scale_y_continuous(limits = c(0, NA)) +
+  scale_y_continuous(expand = expansion(mult = c(0, .12))) +
   theme_custom(8) +

### 2024-03-07 analysis/patient-tracking.Rmd
Modeling fixes and visual updates

@@
 ) %>%
-  mutate(date = as.Date(time))
+  mutate(
+    date = as.Date(time),
+    hr = lubridate::hour(time)
+  ) %>%
+  filter(between(hr, 8, 15)) %>%
+  filter(!(hr > 11 & date == "2021-10-15")) %>%
+  group_by(date, patient_id) %>%
+  slice(1) %>%
+  ungroup() %>%
+  mutate(patient_id = paste(date, patient_id))
+
 df_filt <- filter(
   df,
-  paste(date, patient_id) %in%
-    unique(paste(df_matched$date, df_matched$patient_id))
+  paste(date, patient_id) %in% df_matched$patient_id
 )
@@
 n_distinct(paste(df_filt$date, df_filt$patient_id))
+df_filt %>%
+  filter(tracking_end != "Possible HCW") %>%
+  mutate(patient_id = paste(date, patient_id)) %>%
+  dplyr::select(patient_id) %>%
+  n_distinct()
 ```

### 2024-03-07 analysis/qctr.Rmd
Modeling fixes and visual updates

@@
 
+# example translation
+source("utils/trans_risk.r")
+IR_sit <- .5 * IR("sitting", "male") + .5 * IR("sitting", "female")
+WR_vol <- convert_dist(cellSize)^2 * dimWR[3]
+round((1 - exp(-max(main_qc_wr_mo$value) * 600 * IR_sit / WR_vol)) * 100, 0)
+round((1 - exp(-max(main_qc_wr_af$value) * 600 * IR_sit / WR_vol)) * 100, 0)
+
 # link to building
@@
       val_col = ifelse(value < .05, "Low",
-        ifelse(value < .1, "Median", "High")
+        ifelse(value < .1, "Medium", "High")
       ),
-      val_col = factor(val_col, levels = c("Low", "Median", "High")),
+      val_col = factor(val_col, levels = c("Low", "Medium", "High")),
       variable = ifelse(
@@
         "Mean risk of infection",
-        "Median risk of infection"
+        "Medium risk of infection"
       ),
@@
         variable,
-        levels = paste(c("Mean", "Median"), "risk of infection")
+        levels = paste(c("Mean", "Medium"), "risk of infection")
       )
@@
 main_tr <- average_trans_risk(main_files)
+
+main_tr %>%
+  filter(variable == "Mean risk of infection") %>%
+  group_by(val_col) %>%
+  summarise(n = n()) %>%
+  ungroup() %>%
+  mutate(prop = round(n / sum(n) * 100, 0))
 
@@
 
+amod_est %>%
+  group_by(variable) %>%
+  summarize(
+    lower = quantile(value, .025),
+    median = median(value),
+    upper = quantile(value, .975)
+  ) %>%
+  mutate(across(c(lower, median, upper), ~ round_k(.x, 2)))
+
 amod_est_pl <- ggplot(amod_est, aes(y = variable, x = value)) +
@@
       "Lower ventilation",
-      "Prevalence-based model",
-      "Suspected-based model"
+      "Suspected-based model",
+      "Prevalence-based model"
     ))
@@
 
+mod_comp %>%
+  filter(variable == "Mean risk of infection") %>%
+  group_by(val_col) %>%
+  summarize(n = n()) %>%
+  mutate(prop = round(n / sum(n) * 100, 0))
+
 mod_comp_annon <- mod_comp %>%
@@
   labs(
-    y = "Number of clinical attendees (log-scale)",
+    y = "Number of clinical attendees",
     x = "Difference between spatiotemporal and temporal model (perc. points)"

### 2024-03-07 illustrations/priors.Rmd
Modeling fixes and visual updates

@@
   group_by(activity) %>%
-  tidybayes::median_qi(x, .width = .5)
+  tidybayes::median_qi(x, .width = .95) %>%
+  mutate_if(is.numeric, round, 3)
 
@@
 lambda_dat %>%
-  tidybayes::median_qi(x, .width = .5)
+  tidybayes::median_qi(x) %>%
+  mutate_if(is.numeric, round, 1)
 

### 2024-03-07 preprocessing/prep_simulation-data.R

@@ -42,10 +42,9 @@ ncores <- 6
 #### Track data ####
 
 for (i in 1:length(dates)) {
-  
   # date
   d <- dates[i]
-  
+
   # daytime
   dytm <- daytimes[i]
   t_start <- as.POSIXct(paste(d, "08:00:00"))
@@ -53,36 +52,38 @@ for (i in 1:length(dates)) {
   tt <- seq(t_start, t_end, by = "1 sec")
   midday <- as.POSIXct(paste(d, "12:00:00"))
   midday_idx <- which(tt == midday)
-  
+
   # existing track file
   track_file <- paste0(track_file_path, d, ".rds")
-  
+
   # tracking
   if (!file.exists(track_file)) {
-    
     # load data
     track <- readRDS(paste0("data-clean/patient-tracking/matched-clinical/", d, ".rds")) %>%
       mutate(activity = ifelse(activity == "waiting", 1, 2))
-    
+
     # filter daytime
     if (dytm == "both") {
-      track <- filter(track, between(time, t_start, t_end)) 
+      track <- filter(track, between(time, t_start, t_end))
     } else if (dytm == "morning") {
-      track <- filter(track, between(time, t_start, midday)) 
+      track <- filter(track, between(time, t_start, midday))
     } else if (dytm == "afternoon") {
-      track <- filter(track, between(time, midday, t_end)) 
+      track <- filter(track, between(time, midday, t_end))
     } else {
       stop("Unknown daytime")
     }
-    
+
     # determine cell
     track$cell_id <- NA
-    track$cell_id[track$is_waitingroom] <- mcmapply(find_raster, track$x[track$is_waitingroom], track$y[track$is_waitingroom], 
-                                                    MoreArgs = list(cellCoordsXY = as.matrix(wrCoord[,c("x","y")])), mc.cores = ncores)
-    track$cell_id[track$is_passage] <- mcmapply(find_raster, track$x[track$is_passage], track$y[track$is_passage], 
-                                                MoreArgs = list(cellCoordsXY = as.matrix(cdCoord[,c("x","y")])), mc.cores = ncores)
-    track$cell_id[track$is_tbroom] <- mcmapply(find_raster, track$x[track$is_tbroom], track$y[track$is_tbroom], 
-                                               MoreArgs = list(cellCoordsXY = as.matrix(tbCoord[,c("x","y")])), mc.cores = ncores)
[…]

### 2024-03-07 utils/trans_risk.r

@@ -1,26 +1,26 @@
 #' Number of undiagnosed TB patients
-#' 
-#' @param n number of samples 
+#'
+#' @param n number of samples
 #' @param empirical sample based on empirical counts (default `True`) or population-level prevalence
-#' @param counts number of counts of TB patients (0, 1, 2, ...) vising the clinic  
+#' @param counts number of counts of TB patients (0, 1, 2, ...) vising the clinic
 #' @param prev_est prevalence estimates in %, either a vector with mean and sd, or a ISO-3 char for the country (if estimates are available)
-#' 
-#' @details 
+#'
+#' @details
 #' Distribution parametrized empirically based on the counts of occurrences of the
 #' daily number of diagnosed TB patients visiting the clinic,
 #' using a multinomial distribution, and
 #' assuming that the number of undiagnosed TB patients is similar in number.
-#' 
-#' Population-level prevalence for South Africa taken from Moyo. 
-#' 
+#'
+#' Population-level prevalence for South Africa taken from Moyo.
+#'
 #' @return if empirical, number of undiagnosed TB patients, otherwise proportion of undiagnosed TB patients
-#' 
-#' @references 
+#'
+#' @references
 #' - Moyo: Moyo et al. (2022) in Lancet Infect Dis (doi.org/10.1016/S1473-3099(22)00149-9)
 
-rundiag <- function(n, empirical = T, counts = c(6, 6, 7, 4, 5, 1), prev_est = "RSA") {
+rundiag <- function(n, empirical = T, counts = c(6, 12, 8, 2, 1), prev_est = "RSA") {
   if (empirical) {
-    BM <- rmultinom(n, size = 1, prob = counts) 
+    BM <- rmultinom(n, size = 1, prob = counts)
     n_undiag <- apply(BM, 2, function(col) which(col == 1)) - 1
     return(n_undiag)
   } else {
@@ -42,123 +42,142 @@ rundiag <- function(n, empirical = T, counts = c(6, 6, 7, 4, 5, 1), prev_est = "
 
 
 #' Reduction in quanta emission rate through surgical mask wearing
-#' 
+#'
 #' @param n number of samples
-#' 
-#' @details 
+#'
+#' @details
 #' Distribution approximated based on the reported mean and confidence interval
[…]

### 2024-03-07 doc/paper/manuscript.tex
Updates from Overleaf

--- 2024-03-04 doc/paper/manuscript.tex
@@
 
-\title{\bfseries\LARGE\singlespacing{Spatiotemporal risks of \emph{Mycobacterium tuberculosis} transmission: A modelling study of environmental, clinical, and patient movement data}}
+\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}}
 % author list
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} was the leading cause of death from any single infectious disease before the COVID-19 pandemic, and progress made in TB infection and control before the pandemic has stalled or reversed.
-TB is strictly airborne and thus the transmission risk is higher in crowded indoor environments, especially healthcare facilities where both more infectious and susceptible people congregate.
-Modeling the risk of airborne transmission is difficult and existing approaches, such as the traditional Wells-Riley transmission model, assume that the airspace is well mixed, so that the spatial location of infectious and susceptible individuals is not taken into account.
-\medskip
+\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} was the leading cause of death from any single infectious disease before the COVID-19 pandemic.
+TB is strictly airborne and thus the transmission risk is higher in crowded indoor settings, especially healthcare facilities.
+Existing approaches to model the risk of airborne transmission typically assume that the airspace is well mixed, thereby ignoring that closer proximity to infectious individuals may increase the risk of transmission.
 
-\noindent\textbf{Methods:} We developed a spatiotemporal model for estimating the risk of infection based on the Wells-Riley model.
-We applied our model to environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) that were collected for 5~days between October and November 2021 (during the COVID-19 pandemic) in a primary care clinic in Cape Town, South Africa.
-We linked tracking and clinical data to identify the spatiotemporal location of TB infectious patients within the clinic.
-Incorporating prior assumptions about the generation, diffusion, and removal of infectious doses (called quanta), we performed Monte Carlo simulations to estimate the personal risk of infection for each clinic attendee.
-We compare our simulation results with a non-spatial variant of our model and assess the correlation with characteristics of the clinical visit (\eg time spent at the clinic and the number of close contacts).
-\medskip
+\noindent\textbf{Methods:} We developed a spatiotemporal extension of the Wells-Riley model to estimate the risk of \emph{Mtb} transmission using environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors).
+The data were collected for five days in October/November 2021 in a primary care clinic in Cape Town, South Africa.
+We matched patient movements with clinical records to identify the spatiotemporal location of TB infectious patients.
+Based on that, we modeled the spatiotemporal concentration of infectious doses and estimated the patient-specific risk of infection using Monte Carlo simulation.
 
-\noindent\textbf{Results:} ...
+\noindent\textbf{Results:} Clinical records registered 894 patients visiting the clinic, four of which were diagnosed with TB.
+Patient movement data suggested 1,438 people visiting the clinic (median stay 25\,min, interquartile range [IQR] 13\,min$-$46\,min).
+CO$_2$ levels were mostly below 600\,ppm, suggesting that the clinic was very well ventilated.
+The modeled mean risk of infection was 7\% (95\%-credible interval [CrI] 0.5\%$-$31.1\%) and the risk was significantly associated with the number of close-contact encounters, time spent in the clinic, and the air change rate.
+Without infection control measures that were implemented during the COVID-19 pandemic, the modeled risk of infection would have been 7.8\% (95\%-CrI 0.4\%$-$41.3\%) without mask wearing and 8.4\%-CrI 0.6\%$-$41.4\%) with lower ventilation.
 
-\noindent\textbf{Conclusions:} ...
+\noindent\textbf{Conclusions:} Risk of \emph{Mtb} transmission in primary care clinic in South Africa is considerable, especially for attendees with many close-contact interactions.
+Infection control measures probably mitigated the risk of \emph{Mtb} transmission.
+Our spatiotemporal model could be used in future work to assess the impact of measures preventing prolonged close contact with potentially infectious patients.
 
@@
 
-\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne transmission, spatiotemporal modeling, Wells-Riley model
+\noindent\textbf{Keywords:} Mycobacterium tuberculosis, patient movements, Wells-Riley model
 % maximum of 3-5 keywords
@@
 % TB and the route of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South-East Asian and the African region\cite{WHO2022TBReport}.
-Progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed, as the number of TB-related deaths have increased between 2019 and 2021 while the number of reported TB diagnosis decreased [TB report].
-\emph{Mycobacterium tuberculosis (Mtb)}, the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
-\emph{Mtb} is carried primarily in smaller particles $\leq5\mu$m called aerosols\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
-Airborne transmission is more likely in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Furthermore, there is a high risk of \emph{Mtb} infection in healthcare facilities such as primary care clinics because both people that are more infectious and susceptible are also more frequently visiting the clinic\cite{McCreesh2020IJTLD}.
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South-East Asian and the African region, and progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed\cite{WHO2022TBReport}.
+\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
+\emph{Mtb} is carried primarily in smaller particles in the size range of $2\mu$m to $5\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
+Airborne survival is greater in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021} and healthcare facilities present a particular high-risk setting because both people that are more infectious and susceptible also more frequently visit the clinic\cite{McCreesh2020IJTLD}.
 A study in South Africa estimated that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
-% the Wells-Riley model
-The traditional Wells-Riley transmission model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
-The Wells-Riley model assumes a well mixed airspace, where the risk of infection is modeled as a function of the number of infectious individuals in space, the generation rate of infectious doses (called quanta), the breathing rate per person, and the outdoor air supply rate.
-The quanta generation rate (commonly denoted with the parameter $q$) is unknown, but can be estimated by computing the risk of infection as the ratio of diseased/susceptible cases, and then solving the Wells-Riley equation for the parameter $q$\cite{Nardell1991ARRD,Escombe2008PLoSMed}.
-The Wells-Riley model considers the importance of ventilation by which infectious quanta are removed from the indoor space through outdoor air exchange.
-While the ventilation rate is difficult to estimate, an alternative formulation of the Wells-Riley equation exists that uses indoor CO$_2$ levels as a marker of exhaled-breath exposure\cite{Rudnick2003IndoorAir}, which is related to the ventilation rate.
-
-% limitation of the Wells-Riley model
-Traditional transmission modells such as the Wells-Riley model do not take into account spatial variations in the exposure to infectious doses.
-Although it is reasonable to assume that infectious particles disperse in the indoor space over time, the aerosol concentration is at first highest closest to the infectious person\cite{Vuorinen2020SafSci,Chen2020BuildEnv}.
-Therefore, both short- and long-range airborne transmission of \emph{Mtb} (and other respiratory viruses) seem plausible.
-Nevertheless, previous studies showed that the risk of \emph{Mtb} infection is associated with proximity to an infectious person\cite{Ko2004RiskAnal,Kenyon1996NEJM}, which is in line with the hypothesis that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
-However, proximity to an infectious person is not considered in the Wells-Riley model.
-It would also be difficult to test such a spatiotemporal extension of the Wells-Riley model because the location of the infectious and susceptible individuals are rarely observed.
+% the Wells-Riley model and its limitations
+The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
+The model assumes estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate.
+The Wells-Riley, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same everywhere in the room.
+However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies showed that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
 
 % what this study adds
-We build on the Wells-Riley equation and develop a spatiotemporal model for the concentration of infectious quanta in the indoor space over time.
-We combine clinical and video tracking data to inform the spatiotemporal location of \emph{Mtb} infectious and susceptible individuals visiting a primary care clinic in South Africa between October and November 2021.
-Furthermore, we measure indoor CO$_2$ levels to model the diffusion (\ie spatial spread) and removal of infectious quanta over time.
-Our model allowed us to identify spatiotemporal hotspots of quanta within the clinic and to estimate the risk of infection individually for each clinic attendee.
-We compared our spatiotemporal modeling approach with an alternative model that assumed a well mixed airspace, thereby quantifying the contribution of close contact with infectious people to \emph{Mtb} transmission risk.
+We build on the Wells-Riley model and develop a spatiotemporal extension to model the concentration of infectious quanta spatially and over time.
+We combine clinical and video sensor data to identify the spatiotemporal location of TB infectious and susceptible patients visiting a primary care clinic in South Africa during five days in October/November 2021.
+Furthermore, we measure indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time.
+Using Monte Carlo simulation, our model allows us to estimate the risk of infection individually for each clinical attendee.
+Furthermore, we estimated the impact of infection control measures which were in place during the COVID-19 pandemic using hypothetical scenarios.
 
@@
 We build on the design of a pilot study in 2019\cite{Zurcher2022JID}, which was described in detail in a study protocol \cite{Zurcher2020BMJ}.
-In this follow-up study, we collected environmental data (indoor CO$_2$ levels), clinical (TB status), and tracking (patient movements) data on five days between October and November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
-Compared to the pilot study, we used the full patient tracking dataset, reflecting patient movements within the clinic, and included the TB treatment room in addition to the waiting room.
+We collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (patient movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
 
@@
 
-The primary care clinic offers both TB and HIV services and other basic clinical services, Monday to Friday, from 7\,am to 4\,pm.
+The primary care clinic offers both TB and HIV services and other basic clinical services, Monday to Friday, from 8\,am to 4\,pm.
 The clinic is situated within a large settlement of formal and semiformal housing where both TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
-We defined 3~areas within the clinic: the waiting room, the corridor, and the TB treatment room (Supplementary Figure~\zref{fig:floor-plan} in \supp).
-Furthermore, we defined two time periods: morning (7:00\,am to 12:00\,am) and afternoon (12:00\,am to 4:00\,pm).
+We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
+The TB room was primarily used for diagnosed TB patients and suspected patients with respiratory symptoms.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
+    \caption{\textbf{Schematic view of the clinic}.
+Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room, and finally exit the clinic through one of the exits in the waiting room or at the end of the corridor.
+Colored icons show the placement of the video sensors to track patient movements and the devices monitoring CO$_2$ levels.}
+    \label{fig:floor-plan}
+\end{figure}
 
@@
 
-Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB treatment room (Supplementary Figure~\zref{fig:floor-plan} in \supp).
-CO$_2$ concentrations were recorded in parts per million (ppm) at 1-minute intervals.
-Missing values were linearly imputed, except for October~13 where the corridor's CO$_2$ levels were almost completely missing.
-Here we assumed the same levels as in the waiting room because both rooms had also similar CO$_2$ levels on the other days.
+Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}).
+CO$_2$ levels were recorded in parts per million (ppm) at 1min intervals.
+Missing values were linearly imputed.
+On October~13, the corridor's CO$_2$ levels were missing and imputed with the waiting room's CO$_2$ levels because their CO$_2$ measurements were similar on the other study days.
 
@@
 We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study period.
-These data included the date and time of arrival for the clinic visit, TB diagnostic results, and date of TB treatment start (if applicable).
-We defined infectious patients as individuals testing bacteriologically positive for TB during the study or being on TB treatment in the last $\leq$28\,days after treatment initiation.
-Patients who had recently completed a treatment were neither considered infectious nor susceptible.
+These data included the date and time of arrival, TB diagnostic results, and date of TB treatment start.
+We defined infectious patients as individuals testing bacteriologically positive for TB during the study or being on TB treatment for less than four weeks.
+Patients who had been treated for more than four weeks or recently completed a treatment were neither considered infectious nor susceptible.
+We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were subsequently not diagnosed with TB.
 
@@
 
-We used an anonymous personal movement tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (staff members, patients, and other visitors) throughout the clinic (Supplementary Figure~\zref{fig:floor-plan} in \supp).
-The resulting date- and timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-Supplementary Figure~\zref{fig:tracking-examples} in \supp~shows a few sample tracks from our study.
-Note that individuals could contribute multiple tracks if they moved out of the range of a sensor and then returned, or if they were briefly lost because the sensor did not recognize them as a person (\eg if bending or hiding behind another person).
-To link these tracks, we created a R Shiny tool and manually linked tracks back together that most likely belonged to the same person (Supplementary Text~\zref{sec:setting-and-data} in \supp).
-We defined close contacts as other persons within a radius of $\leq$1\,m.
+We used an anonymous personal tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinical staff, patients, and other visitors) throughout the clinic at 1s intervals (\Cref{fig:floor-plan}).
+The resulting timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
+\supp Figure~\zref{fig:tracking-examples} shows a few sample tracks from our study.
+Note that individuals could contribute multiple tracks if they moved out of the sensor's range or were briefly not recognized as a person.
+To remedy these issues, we created an R Shiny tool to manually link interrupted tracks back together (\supp Text~\zref{sec:setting-and-data}).
+Furthermore, we labeled tracks to identify movements from clinical staff.
 
-\subsection{Spatiotemporal modeling}
+\subsection{The Wells-Riley equation}
 
-The workflow of our spatiotemporal modeling approach is shown in \Cref{fig:modeling-flow}.
-(1)~We combined patient tracking and clinical data to identify the spatiotemporal location of infectious individuals inside the clinic.
-We record the registration time for patients spending $\geq$5~seconds in the registration area, and then linked their registration time with the nearest arrival time recorded in the clinical data, considering a maximum delay of 15~minutes.
-(2)~In addition to the number of TB patients as reported in the clinical dataset, we also included a similar number of undiagnosed (unmasked) TB patients among all other clinical attendees\cite{Berhanu2023CID}.
-Both masked and unmasked TB patients generate infectious quanta at their spatiotemporal location.
-The quanta generation rate $q$ was informed by estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD}, and we applied a reduced rate to account for the effect of mask wearing\cite{Dharmadhikari2012AJRCCM}, which was mandatory in the clinic during the study period (3).
-Next we combined patient tracking and environmental data to monitor natural ventilation by calculating the outdoor air exchange rate $AER$ from the number of people in the room and the indoor CO$_2$ level\cite{Batterman2017IJERPH}.
-(4)~The $AER$ is related to the diffusion constant $D$ (in m$^2$/s)\cite{Cheng2011EnvSciTech}, which determines the speed at which the infectious quanta spatially spreads (diffuses) in the indoor air.
-We assumed that the infectious quanta diffuses radially from the spot where it was generated.
-(5)~The $AER$ also corresponds to the rate of quanta removal through natural ventilation.
-The total removal further considers the viral inactivation rate $\lambda$, which was informed by prior literature\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-(6)~Quanta generation, diffusion, and removal determined the quanta concentration $N$ at time $t$ in the clinic, which is computed as
+Our spatiotemporal model builds upon the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability) of infection $P$ as
+\begin{align}
+    P = \frac{D}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
+\end{align}
+where $D$ is the number of diseases cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ s$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta s$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ s$^{-1}$).
+The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
+The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
+Our aim is to relax this assumption by using patient movement data to identify the location of infectious patients and consider spatial variation in the quanta concentration.
+
+\subsection{Spatiotemporal modeling approach}
+
+A detailed description of the Wells-Riley model and our spatiotemporal extension is provided in \supp Text~\zref{sec:spattemp-model}.
+In the following, we describe the main components of our modeling approach, which are summarized in \Cref{fig:modeling-flow}.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{doc/paper/flow-chart.pdf}
+    \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
+(1)~Clinical and tracking data is combined to identify the spatiotemporal location of diagnosed (infectious) TB patients.
+(2)~Both diagnosed and randomly sampled undiagnosed  patients among visitors generate infectious quanta.
+(3)~Tracking and environmental data is combined to estimate the air change rate.
+(4)~The air change rate is related to the speed with which quanta diffuses in the indoor space.
+(5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles.
+(6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, based on which the personal risk of infection is computed.}
+    \label{fig:modeling-flow}
+\end{figure}
+
+\subsubsection*{Quanta generation}
+
+(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients inside the clinic.
+We recorded the timestamp when the patient movement was in the registration area and matched this with the registration time in the clinical records.
+We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching.
+Data were matched if there was a patient movement in the registration area followed by a entry in the clinical database, considering a maximum delay of 15 minutes.
+All diagnosed TB patients could be matched with patient movements this way.
+(2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed patients.
+Prior studies suggest that the number of undiagnosed TB patients is proportional to the number of diagnosed TB patients \cite{Berhanu2023CID}.
+Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:count-tb-patients}): 6 (12\%) days with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four visiting TB patients.
+Undiagnosed patients were randomly sampled among all visitors with the same probability.
+Clinical staff were not considered infectious and excluded from the sampling.
+We assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution and considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+Since mask wearing was mandatory for all clinical staff and visitors during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+
+\subsubsection*{Quanta diffusion}
+
+(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rates by daytime using a transient mass balance model\cite{Batterman2017IJERPH}.
+(4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}.
+Since airflow was not monitored in our study, we assume that the quanta diffuse radially from where they were generated (\supp~Figure~\zref{fig:toy-example}).
+
+\subsubsection*{Quanta removal}
+
+(5)~Quanta removal occurs through replacement of contaminated indoor air with fresh outdoor air, inactivation of bacteria in airborne particles, and gravitational settling.
+The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
+The bacterial inactivation rate was modeled using a lognormal prior distribution (\supp~Figure~\zref{}) with a median of 1\,quanta h$^{-1}$ (0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $2\mu$m to $5\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+
+\subsubsection*{Risk of infection}
+
+(6)~Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as 
 \begin{align}\label{eq:spattemp-N}
@@
 \end{align}
-where $\Delta$ is the Laplace operator (second-order differential operator) and $I_t$ is the number of infectious individuals in space.
-Finally, the patient-specific risk of infection depends on the cumulative exposure to infectious quanta during the clinical visit.
-We use the Wells-Riley Poisson relation to estimate the risk of TB transmission for each clinical attendee $a$ as
+where $D$ is the diffusion constant (m$^2$ s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta s$^{-1}$).
+The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
 \begin{align}
-    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
+    P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a\right),
 \end{align}
-where $P$ is the probability of infection, $\mathbb{I}$ is a binary variable indicating whether attendee $a$ was at location $s$ at time $t$, and $p = 8$\,l/min is the volumetric breathing rate per person.
-The general modeling framework together is described together with an example in detail in Text~\zref{sec:spattemp-model} in \supp.
+where $N_{s,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ is indicates whether the attendee was in $c$ at time $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the distance between two patient tracks was more or less than 0.25m.
 
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{doc/paper/flow-chart.pdf}
-    \caption{Flow chart of the spatiotemporal modeling approach illustrating the steps from data (blue) to modeling (yellow).}
-    \label{fig:modeling-flow}
-\end{figure}
+\subsection{Monte Carlo simulation}
 
-We divide each separate area of the clinic (waiting room, corridor, TB room) into a grid of cubic cells with an area of 0.25m$^2$ and a volume depending on the height of the room (waiting and TB room: 3m; corridor: 2.5m).
-We update the quanta concentration every second and assume that $N_{s,0} = 0 ~~ \forall s$, \ie all quanta from the previous day has been removed from the air before the start of the following clinical day.
-Due to mask wearing, we assume that the initial spread of quanta before diffusion is limited to the cell where the infectious individual is located.
-We model all model parameters with prior distributions and perform Monte-Carlo simulation to estimate the risk of infection.
-In each simulation, we perform the following three steps: (i)~sampling uncertain modeling parameters and undiagnosed TB patients, (ii)~computing the spatiotemporal quanta concentration, and (iii)~computing the cumulative risk of infection for each clinical attendee.
-The model setup, including the specific assumptions and prior distributions, is described in detail in Text~\zref{sec:estimation} in \supp.
+The model setup and simulation approach is described in detail in Text~\zref{sec:estimation}, including a comprehensive review of our modeling assumptions and prior distributions.
+To summarize, we simulated the spatiotemporal quanta concentration separately for the waiting room, corridor, and TB room of the clinic.
+Each room was rasterized into a grid of cubic cells with a squared area of 0.25m$^2$.
+The quanta concentration was modeled during clinic hours from 8am to 4pm and updated every second.
+We used a Monte-Carlo simulation approach to consider uncertainty in key modeling parameters.
+In each simulation, we sampled uncertain modeling parameters from their prior distributions and, based on that, computed the spatiotemporal quanta concentration and the risk of infection per clinical attendee.
+
+\subsection{Hypothetical scenarios}
+
+Our study was set during the COVID-19 pandemic when infection control measures were in place.
+To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~lower ventilation, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pilot study of 2019 before the pandemic\cite{Zurcher2022JID}.
+Furthermore, we examined the impact of different assumptions regarding the number infectious patients in space: 1)~instead of undiagnosed TB patients, we considered suspected TB patients as infectious; 2)~random sampling of infectious patients among all visitors in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
+Except for the scenario-specific modifications, all other modeling assumptions and parameters were the same in each Monte Carlo simulation, ensuring comparability of the results.
 
@@
 
-We analyzed the time-varying CO$_2$ levels, the daily number of registered and masked TB patients, and the number of clinic attendees monitored through patient tracking.
-We perform 10,000 Monte Carlo simulations and show the average quanta concentration by daytime and summarize the risk of infection with the mean and 95\%-credible interval (CrI) across simulations.
-To demonstrate the impact of spatial modeling, we compare our modeling approach with one that assumes a well-mixed airspace (\ie where the initial quanta concentration is the same everywhere in the room regardless of where it was generated).
-Furthermore, we assess the correlation between the personal risk of infection and characteristics of the clinical visit (time spent in each area, number and duration of close contacts).
-To assess the effectiveness of mask wearing, we compare our modeling results with a scenario where there is no reduction in the generated quanta and the initial spread is extended to the first-neighbouring cells of where the infectious individual is located.
-All analyses were performed in R software (version 4.3.1)\cite{RCoreTeam2023}.
+Clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates) and patient movement data (spatiotemporal number of tracks, visit time, and close-contact encounters) were assessed descriptively and summarized with the median and interquartile range (IQR).
+Spatiotemporal quanta concentration and risk of infection per clinical attendee were estimated based on 5,000 Monte Carlo simulations.
+The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean/median and 95\%-credible interval (CrI).
+A Bayesian Beta regression model was used to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the air change rate, visit time, and close-contact encounters.
+All analyses were performed in R software (version 4.3.1) and the Bayesian regression model was estimated in the probabilistic programming language Stan (version 2.26.1).
 
@@
 
+During the five study days, 894 patients have been registered in the clinical database, of which four patients were diagnosed with TB and 16 patients were suspected of TB following symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}).
+Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as being from healthcare workers, identifying 1,438 patients or other visitors.
+Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
+Patients spent around half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
+Close-contact encounters (distance <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9), and several patients had more than ten close contacts during their visit (\Cref{fig:input-data-descriptives}d).
+The majority of patients was in close contact with at least one other person for more than half the time of their visit (median 68\%, IQR 41\%$-$87\%, \Cref{fig:input-data-descriptives}e).
+The CO$_2$ levels inside the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\Cref{fig:input-data-descriptives}f) and the corresponding air change rates were high (\Cref{fig:input-data-descriptives}g), especially the densely occupied waiting room (median 29\,h$^{-1}$), IQR 9\,h$^{-1}$$-$36\,h$^{-1}$).
+
 \begin{figure}
     \centering
-    \includegraphics{results/data/co2-levels-over-time.png}
-    \includegraphics{results/data/air-changes-per-hour.png}
      \includegraphics{results/data/no-people-spatial.png}
@@
     \includegraphics{results/data/time-close-contacts.png}
+    \includegraphics{results/data/co2-levels-over-time.png}
+    \includegraphics{results/data/air-changes-per-hour.png}
     \caption{\textbf{Environmental conditions and patient movement characteristics.} \textbf{(a)}~CO$_2$ levels per room of the clinic over time per day.
@@
 \end{figure}
+
+As an illustrative example of our spatiotemporal model, the animation in \supp File 2 shows the spatiotemporally varying quanta concentration as an infectious patient enters the clinic in the morning of October 25 and moves through the waiting room.
+The quanta concentration is higher near the infectious patient, therefore indicating his position and movement.
+It can further be seen that the quanta diffuses fairly quickly due to its relation with the air change rate, which was very high because all doors and windows were typically open in the clinic.
+\Cref{fig:main-modeling-results} shows the results from our main spatiotemporal modeling analysis across study days and simulations.
+The mean quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a).
+A ten minute exposure to the mean quanta concentration in the center of the waiting room would correspond to a 52\% risk of infection in the morning compared to a 16\% risk in the afternoon.
+\Cref{fig:main-modeling-results}b shows the mean and median risk of infection per clinical attendee.
+The attendee-average mean risk of infection was 7\% (95\%-CrI 0.5\%$-$31.1\%) and the median risk was 0.8\% (95\%-CrI 0.0\%$-$11.2\%).
+The mean risk was typically higher than the median risk because the risk of infection was very low in many simulations and only significant in a few ones.
+The majority of attendees (51\%) had a low mean risk of infection <5\%, roughly a third (31\%) had a medium risk between 5\%$-$10\%, and 18\% had a high risk of infection >10\%.
+A comparison of our spatiotemporal modeling results with those from a temporal model assuming a well-mixed airspace shows a notable difference of more than 0.5 percentage points in the mean risk of infection for 168 attendees (12\%, \supp~Figure~\zref{fig:model-comparison}).
+The mean risk of infection from our spatiotemporal model was associated the number of close contacts, time spent in clinic, and the air change rate (\Cref{fig:main-modeling-results}c).
+If the number of close contacts doubled, \eg from five to ten, the odds of infection were 1.4 (95\%-CrI 1.28$-$1.53).
+If the air change rate doubled, \eg from 15 to 30 air changes per hour, the odds of infection were 0.84 (95\%-CrI 0.81$-$0.88).
 
@@
 \end{figure}
+
+To assess the impact of infection control measures during the COVID-19 pandemic, we performed two hypothetical scenarios shown in the top panel of \Cref{fig:scenario-results}.
+If none of the attendees or staff had worn masks, the attendee-average mean risk of infection would have been almost a percentage point higher (7.8\%, 95\%-CrI 0.4\%$-$41.3\%).
+If the ventilation rate had been fixed at 16 air changes per hour (the average in 2019 before the pandemic), the attendee-average mean risk of infection would have been more than percentage point higher (8.4\%-CrI 0.6\%$-$41.4\%).
+In both scenarios, the upper credible interval is roughly ten percentage points higher, indicating that mask wearing and better ventilation considerably impact the tail of the distribution, \ie high risks of infection.
+To assess the impact of modeling assumptions regarding the number of infectious people in space, we performed two hypothetical scenarios shown in the bottom panel of \Cref{fig:scenario-results}.
+Assuming there were no undiagnosed TB patients but all attendees suspected of TB were infectious, the attendee-average mean risk of infection would have been considerably higher (13.9\%, 95\%-CrI 0.0\%$-$45.2\%).
+Assuming there were only undiagnosed TB patients in proportion to the prevalence of TB in the South African population, the attendee-average risk of infection would have been lower (6.3\%, 95\%-CrI 0.8\%$-$16.3\%).
 
@@
 % summary
-We modeled the spatiotemporal concentration of infectious quanta, considering the location of infectious individuals (quanta generation), spatial dispersion (quanta diffusion), and natural ventilation (quanta removal).
-We found ...
+We developed a spatiotemporal model to estimate the concentration of infectious quanta in an indoor space and applied it to patient movement data from a South African primary care clinic in 2021 to simulate the risk of TB infection for clinical attendees.
+We estimated a mean risk of infection of 7\% (95\%-CrI 0.5\%$-$31.1\%) across clinical attendees.
+Clinical attendees with more close-contact encounters had a higher risk of infection.
+The modeled risk of infection would have been higher without the implementation infection control measures during the COVID-19 pandemic such as compulsory face mask wearing and better natural ventilation.
 
 % critical appraisal of our results
+South Africa has a high burden of TB\cite{WHO2022TBReport}.
+A previous modeling study estimated that between 4\% and 14\% of TB transmission in adults in a high TB burden, high HIV prevalence community in South Africa occurred in primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+In line with this, we estimated an average infection risk of between 1\% and 16\% when assuming the number of infectious attendees is proportional to the prevalence of TB in the South African population.
+Another modeling study further suggests a considerable benefit of infection control measures, \eg the rate of TB transmission in clinics could be reduced by about 50\% with compulsory mask wearing and natural ventilation through opening doors and windows\cite{McCreesh2021BMJGlobalHealth}.
+Our modeling results also suggest a notable reduction in the average risk of infection of more than a percentage point.
+Moreover, we estimated a considerable shift in the tail of the risk distribution as infection control measures significantly reduce high infection risks.
 
 % close contacts
-Our spatiotemporal model can be viewed as a modified version of the original Wells-Riley transmission model\cite{Riley1978AJE}.
-It is motivated by previous findings that prolonged close contact is required for transmission\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
-As a result, our model assigns a higher risk of infection to clinical attendees that spend more time in close proximity to infectious individuals.
-It could thus be used to assess the effectiveness of interventions targeting a reduction in the number of close contacts.
-Although such assessments only show the potential (modeled) impact of interventions, they can still provide a first estimate about how contact reductions translate into lower transmission risks in crowded indoor settings.
-Nevertheless, we acknowledge that transmission events are generally difficult to predict.
-Their occurrence depends on many other factors such as airflow, the properties of the pathogen, and variation in infectiousness of the source and susceptibility of the host\cite{Wang2021Science}.
-We discuss these factors in more depth in Supplementary Text~\zref{sec:depth-discussion} in \supp.
+Our spatiotemporal model can be viewed as a modified version of the Wells-Riley transmission model\cite{Riley1978AJE}, which assumes a well-mixed airspace.
+By contrast, our model allows the quanta concentration to vary both over time and space.
+The exposure to infectious quanta therefore depends on the proximity to infectious individuals because the density of infectious particles is initially higher near them\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
+Our modeled risk of TB infection was significantly associated with the number of close-contact encounters and the time spent in the clinic.
+This result is intuitive because there are only few infectious TB patients visiting the clinic, so that more time in the clinic as well as a higher number of close contacts makes exposure to high doses of infectious particles more likely.
+Previous findings already suggest that prolonged close contact may be required for transmission of respiratory infections\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
+However, proving close contact respiratory transmission is difficult and  empirical studies investigating the potential link between contact data and respiratory transmission are scant\cite{Voirin2015ICHE,Vanhems2013PONE}.
 
 % more complex models and our relation
-An important factor that is omitted from our model is airflow.
-Poor ventilation settings can generate an airflow where infectious particles circulate rather than being dispersed and removed from the indoor space\cite{Li2021BuildEnv}.
-There is large stream of literature considering airflow using computational fluid dynamics (CFD) models to simulate the spatial spread of infectious aerosols in indoor environments, \eg Ref.
-\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
-CFD models can also be combined with the Wells-Riley model to estimate the risk of infection\cite{Yan2023BE,Qian2009BE,Li2022SOTTE}, thus are seemingly appealing to our setting as well.
-However, because they are computationally expensive, CFD models are often only used for experiments, short simulations, and static environments.
-They also require detailed knowledge about the indoor space, in particular airflow, to model particle transport.
-By contrast, our setting presents a dynamic environment with frequent patient movements over multiple days in a naturally ventilated space (varying airflow).
-Moreover, we run our model multiple times in a large number of Monte Carlo simulations to consider uncertainty in all modeling parameters and unmasked infectious TB patients.
+We applied our model to TB, although it can easily be extended to model other respiratory infections.
+Airborne respiratory transmission depends on common environmental and patient-specific factors, mainly ventilation and patient movements.
+Other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can also influence the generation, diffusion, and removal of infectious quanta in the space\cite{Wang2021Science}.
+We discuss these factors in more depth in \supp Text~\zref{sec:depth-discussion} in \supp~Text~\zref{sec:depth-discussion}.
+One important factor not considered by our model is airflow, although it has been shown that poor ventilation settings can produce airflows where infectious particles are trapped in the indoor space rather than being removed\cite{Li2021BuildEnv}.
+Computational fluid dynamics (CFD) models can consider airflow when simulating the spatial spread of infectious particles\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
+However, CFD models are computationally expensive and primarily used in experimental settings and static environments, which makes their application to our dynamic setting with patient movements difficult.
+Moreover, our approach requires a large number of Monte Carlo simulations to reflect uncertainty in modeling parameters and assumptions, necessitating computational efficiency.
+
+% limitations
+Our modeling study has several limitations.
+Tracking patient movements with video sensors is challenging and required considerable data processing.
+Despite that, many patient movements were interrupted and could not be linked back together.
+Several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
+Uncertainty in these modeling parameters was often large and could be reduced.
+For example, incorporating additional patient characteristics from the clinical database could indicate variation in infectiousness and susceptibility of patients visiting the clinic.
+It was not possible to validate our modeling results by empirical findings, which could be the subject of a long-term future study testing the practical implementation of our model.
 
 % conclusions
-We conclude ...
+In conclusion, our modeling study using clinical, environment, and patient movement data showed that the risk of TB transmission in a primary care clinic in South Africa in October/November 2021 was considerable, albeit mitigated by infection control measures that were implemented during the COVID-19 pandemic.
+The number of close-contact encounters by clinical attendees was significantly associated with their modeled risk of infection.
+Our spatiotemporal model could thus be used to assess the potential effectiveness of interventions targeting a reduction in the number of close contacts.
 

### 2024-03-07 doc/paper/supplements.tex

@@
 
-The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)\cite{Cheng2011EnvSciTech}
+The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)\cite{Cheng2011EnvSciTech,Foat2020BE}
 \begin{align}
@@
 
+\section{Additional results}\label{sec:additional-results}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/diagnosed-tb-patients.png}
+    \caption{Number of diagnosed TB patients (dark bars) and those suspected of TB (light bars) following symptom screening.
+Modeled study days in color, other clinic days in October and November in grey.}
+    \label{fig:diagnosed-tb-patients}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/modeling/model-comparison.png}
+    \caption{Difference in the mean risk of infection per clinical attendee between the spatiotemporal and temporal model.
+Lower: <$-$0.5, Similar: between $-$0.5 and 0.5, Higher: >0.5 percentage points.}
+    \label{fig:model-comparison}
+\end{figure}
+
+\clearpage
+
 \section{Discussion of factors affecting the risk of airborne infection}\label{sec:depth-discussion}

### 2024-03-21 preprocessing/prep_simulation-data.R
Fixes in simulation and compactness

@@ -101,7 +101,7 @@ for (i in 1:length(dates)) {
     )
     track$cell_height <- 1 / convert_dist(1) * track$cell_height
 
-    # inhalation rate depending on activity
+    # inhalation rate depending on activity in m3/s
     track$inhal_rate <- ifelse(track$activity == 1,
       .5 * IR("sitting", "male") + .5 * IR("sitting", "female"),
       .5 * IR("walking", "male") + .5 * IR("walking", "female")

### 2024-03-21 preprocessing/simulate_qctr.R

@@ -70,12 +70,23 @@ if (sel_rooms == "clinic") {
   stop("Invalid rooms argument.")
 }
 
+# save quanta concentration
+save_quanta <- ifelse(args[12] == "T", TRUE, FALSE)
 
 # file containing simulation parameters
-param_file <- paste0("simulations/", "/", "sim-param-", 2021, "-", mth, "-", dy, ".rds")
+param_file <- paste0(
+  "simulations/", "sim-param-",
+  2021, "-", mth, "-", dy,
+  ".rds"
+)
 
 # path to store cellSize dependend tracking data
-track_file <- paste0("simulations/", "/", "tracking-df-", cellSize, "px-", 2021, "-", mth, "-", dy, ".rds")
+track_file <- paste0(
+  "simulations/", "tracking-df-",
+  cellSize, "px-",
+  2021, "-", mth, "-", dy,
+  ".rds"
+)
 
 # path to file to save simulation results
 sim_path_to_file <- paste0(adir, mth, "-", dy, "/")
@@ -99,9 +110,9 @@ source("models/tm-v1.R")
 #### Time ####
 
 date <- paste0("2021-", mth, "-", dy)
-t_start <- as.POSIXct(paste(date, "08:00:00"))
-midday <- as.POSIXct(paste(date, "12:00:00"))
-t_end <- as.POSIXct(paste(date, "16:00:00"))
+t_start <- as.POSIXct(paste(date, "08:00:00"), tz = "CET")
+midday <- as.POSIXct(paste(date, "12:00:00"), tz = "CET")
+t_end <- as.POSIXct(paste(date, "16:00:00"), tz = "CET")
 tt <- seq.POSIXt(t_start, t_end, by = "1 sec")
 tt_morn <- seq.POSIXt(t_start, midday, by = "1 sec")
 tt_noon <- seq.POSIXt(midday + 1, t_end, by = "1 sec")
@@ -243,15 +254,26 @@ for (sim in cont_n:term_n) {
       dplyr::select(time, cell_x, cell_y, activity) %>%
       rename(x = cell_x, y = cell_y) %>%
       mutate(
-        q = ifelse(activity == 1, sim_param$q_wait[sim_param$sim == sim], sim_param$q_walk[sim_param$sim == sim]),
-        q = ifelse(is_masking == 1, q * sim_param$mask_red[sim_param$sim == sim], q),
+        q = ifelse(
+          activity == 1,
[…]

### 2024-03-21 analysis/patient-tracking.Rmd
Results and prior vis updates

@@
 ```{r}
-track_qual <- df %>%
+track_qual <- df_filt %>%
   group_by(date, patient_id) %>%
@@
 
-sum(track_qual[names(track_qual) != "Possible HCW"])
 track_qual_str <- paste0(
@@
 
-df %>%
-  filter(tracking_end != "Possible HCW") %>%
-  group_by(date, patient_id) %>%
-  slice(1) %>%
+df_filt %>%
+  group_by(date, patient_id, tracking_end) %>%
+  summarise(time_in_clinic = n() / 60) %>%
   ungroup() %>%
   group_by(tracking_end) %>%
-  summarize(n = n()) %>%
-  ungroup() %>%
-  mutate(prop = round(n / sum(n) * 100))
+  tidybayes::mean_qi(time_in_clinic, .width = .95) %>%
+  ungroup()
 ```

### 2024-03-21 analysis/qctr.Rmd
Results and prior vis updates

@@
 
-```{r}
+```{r illustration}
 # data
@@
 
-```{r}
-main_files <- c(
-  list.files("simulations/main/10-13/", full.names = T),
-  list.files("simulations/main/10-15/", full.names = T),
-  list.files("simulations/main/10-25/", full.names = T),
-  list.files("simulations/main/11-04/", full.names = T),
-  list.files("simulations/main/11-05/", full.names = T)
+```{r main-files}
+main_files <- list.files("simulations/main/",
+  full.names = TRUE, pattern = "*.rds$",
+  recursive = TRUE
 )
@@
 source("utils/trans_risk.r")
-IR_sit <- .5 * IR("sitting", "male") + .5 * IR("sitting", "female")
+IR_sit <- (.5 * IR("sitting", "male") + .5 * IR("sitting", "female")) / 3600
 WR_vol <- convert_dist(cellSize)^2 * dimWR[3]
-round((1 - exp(-max(main_qc_wr_mo$value) * 600 * IR_sit / WR_vol)) * 100, 0)
-round((1 - exp(-max(main_qc_wr_af$value) * 600 * IR_sit / WR_vol)) * 100, 0)
+round((1 - exp(-max(main_qc_wr_mo$value) * 600 * IR_sit / WR_vol)) * 100, 3)
+round((1 - exp(-max(main_qc_wr_af$value) * 600 * IR_sit / WR_vol)) * 100, 3)
 
@@
     labs(
-      fill = expression("Mean quanta concentration (quanta/m"^3 * " x 10e"^-3 * " per sec)"),
+      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10e"^-3 * ")"),
       title = "a"
@@
     mutate(
-      val_col = ifelse(value < .05, "Low",
-        ifelse(value < .1, "Medium", "High")
+      val_col = ifelse(value < .001, "Low",
+        ifelse(value < .01, "Medium", "High")
       ),
       val_col = factor(val_col, levels = c("Low", "Medium", "High")),
+      val_ann = 1 - (1 - value)^12,
       variable = ifelse(
@@
       lower_P = quantile(value, .025),
-      upper_P = quantile(value, .975)
+      upper_P = quantile(value, .975),
+      mean_P_ann = mean(val_ann),
+      lower_P_ann = quantile(val_ann, .025),
+      upper_P_ann = quantile(val_ann, .975)
     ) %>%
     ungroup() %>%
-    mutate(across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 1)),
+    mutate(
+      across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 3)),
       lab = paste0(
@@
         " (95%-CrI ", lower_P, "% – ", upper_P, "%)"
+      ),
+      across(c(mean_P_ann, lower_P_ann, upper_P_ann), ~ round_k(100 * .x, 2)),
+      lab_ann = paste0(
+        mean_P_ann, "%",
+        " (95%-CrI ", lower_P_ann, "% – ", upper_P_ann, "%)"
       )
@@
 
+main_annon
+
 # plot
-roi_pl <- ggplot(main_tr, aes(x = value)) +
+roi_pl <- ggplot(
+  data = main_tr %>% filter(variable == "Mean risk of infection"),
+  mapping = aes(x = value)
+) +
   geom_histogram(
     mapping = aes(fill = val_col),
-    binwidth = .05,
+    binwidth = .0005,
     boundary = 1
@@
   geom_text(
-    data = main_annon,
+    data = main_annon %>% filter(variable == "Mean risk of infection"),
     mapping = aes(x = Inf, y = Inf, label = lab),
@@
   ) +
-  facet_wrap(~variable) +
   scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
@@
   scale_x_continuous(
-    breaks = c(0, .05, 0.1, seq(0.2, .5, .1)),
     labels = function(x) x * 100
@@
 assoc_df <- track_ft %>%
-  left_join(
-    aer %>%
-      filter(room == "Waiting room") %>%
-      dplyr::select(date, daytime, aer)
-  ) %>%
+  # left_join(
+  #   aer %>%
+  #     filter(room == "Waiting room") %>%
+  #     dplyr::select(date, daytime, aer)
+  # ) %>%
   mutate(patient_id = paste(
@@
 # model
-amod <- stan_betareg(mean_risk ~ aer + time_in_clinic + n_cc + cc_time,
+amod <- stan_betareg(mean_risk ~ time_in_clinic + n_cc + cc_time,
   data = assoc_df %>% mutate(
     n_cc = n_cc + 1, cc_time = cc_time + 1 / 60,
-    across(c(aer, time_in_clinic, n_cc, cc_time), log2)
+    across(c(time_in_clinic, n_cc, cc_time), log2)
   ),
@@
 
-amod_est <- spread_draws(amod, aer, time_in_clinic, n_cc, cc_time) %>%
+amod_est <- spread_draws(amod, time_in_clinic, n_cc, cc_time) %>%
   reshape2::melt(c(".chain", ".iteration", ".draw")) %>%
@@
     lower = quantile(value, .025),
-    median = median(value),
+    mean = mean(value),
     upper = quantile(value, .975)
   ) %>%
-  mutate(across(c(lower, median, upper), ~ round_k(.x, 2)))
+  mutate(across(c(lower, mean, upper), ~ round_k(.x, 2)))
 
@@
   geom_vline(aes(xintercept = 1), linetype = "dotted", color = "red") +
-  stat_interval() +
+  stat_interval(point_interval = "mean_qi") +
   stat_summary(
     geom = "point",
-    fun = median,
+    fun = mean,
     size = 2,
@@
   labs(
-    x = "Odds ratio of the modeled risk of infection for a doubling in the predictor",
+    x = "OR of modeled risk if 2 x predictor",
     title = "c",
-    color = "Credible interval"
+    color = "CrI"
   ) +
@@
     labels = c(
-      "Air change rate",
+      # "Air change rate",
       "Time in clinic",
-      "Number of close contacts",
-      "Time in clocse contact"
+      "No. close contacts",
+      "Time close contact"
     )
@@
   theme_custom() +
-  theme(axis.title.y = element_blank())
+  theme(
+    axis.title.y = element_blank(),
+    legend.position = "bottom"
+  )
 
@@
   mean_quanta_pl,
-  roi_pl,
-  amod_est_pl,
+  grid.arrange(roi_pl, amod_est_pl, ncol = 2),
   ncol = 1,
-  heights = c(8, 8, 4)
+  heights = c(8, 8)
 )
@@
   width = 16 / cm(1),
-  height = 20 / cm(1)
+  height = 16 / cm(1)
 )

### 2024-04-10 illustrations/priors.Rmd
Results and prior vis updates

@@
     "Nardell (1991)",
-    "Riley (1962)"
+    "Riley (1962)",
+    "Dharmadhikari (2012)"
   ),
-  x = c(0.89, 8.2, 12.5, 1.25),
-  y = c(1, 7, 9, 3) / 1e2
+  x = c(0.89, 8.2, 12.5, 1.25, 138),
+  y = c(1, 7, 9, 3, 12) / 1e2
 ) %>%

### 2024-04-10 analysis/qctr.Rmd
Update results, resolve git conflicts

@@
   paste0(
-    "../simulations/", "/", "sim-param-",
+    "simulations/", "/", "sim-param-",
     2021, "-", mth, "-", dy, ".rds"
@@
   paste0(
-    "../simulations/", "/", "tracking-df-",
+    "simulations/", "/", "tracking-df-",
     250, "px-", 2021, "-", mth, "-", dy, ".rds"
@@
 )
-aer_df <- readRDS("../data-clean/environmental/air-change-rate.rds") %>%
+aer_df <- readRDS("data-clean/environmental/air-change-rate.rds") %>%
   filter(
@@
 
-animate(
+gganimate::animate(
   qc_anim_pl,
@@
 
-anim_save("results/modeling/illustrated-quanta-conc-over-time.gif")
+gganimate::anim_save("results/modeling/illustrated-quanta-conc-over-time.gif")
 ```
@@
 
-```{r main-files}
-main_files <- list.files("simulations/main/",
-  full.names = TRUE, pattern = "*.rds$",
-  recursive = TRUE
-)
-```
-
 ### Quanta conc
@@
 ```{r}
-main_qc_files <- main_files[grepl("quanta", main_files)]
-
-mean_qc <- function(files, room, daytime) {
-  room_files <- files[grepl(room, files)]
-  room_dt_files <- room_files[grepl(daytime, room_files)]
-  qc <- simplify2array(lapply(room_dt_files, readRDS))
-  mqc <- apply(qc, c(1, 2), mean)
-  mqc_df <- melt(mqc) %>%
-    rename(
-      y = Var1,
-      x = Var2
+main_qc <- data.frame(
+  files = list.files("simulations/main/",
+    full.names = TRUE, pattern = "quanta_conc-.*.txt$",
+    recursive = TRUE
+  )
 ) %>%
   mutate(
-      room = room,
-      daytime = daytime,
-      across(c(x, y), as.numeric)
+    room = ifelse(grepl("Waiting", files), "Waiting room",
+      ifelse(grepl("Corridor", files), "Corridor", "TB room")
+    ),
+    daytime = ifelse(grepl("morning", files), "morning", "afternoon")
   )
+
+mean_qc <- function(file) {
+  df <- read.table(file)
+  colnames(df) <- c("sim", "x", "y", "value")
+  df <- df %>%
+    group_by(x, y) %>%
+    summarize(value = mean(value)) %>%
+    ungroup()
+  return(df)
 }
 
-# total quanta concentration by daytime averaged across simulations
-main_qc_wr_mo <- mean_qc(main_qc_files, "Waiting room", "morning")
-main_qc_wr_af <- mean_qc(main_qc_files, "Waiting room", "afternoon")
-main_qc_cr_mo <- mean_qc(main_qc_files, "Corridor", "morning")
-main_qc_cr_af <- mean_qc(main_qc_files, "Corridor", "afternoon")
-main_qc_tb_mo <- mean_qc(main_qc_files, "TB room", "morning")
-main_qc_tb_af <- mean_qc(main_qc_files, "TB room", "afternoon")
+main_qc$value <- lapply(main_qc$files, mean_qc)
+
+main_mean_qc <- main_qc %>%
+  unnest(value) %>%
+  group_by(room, daytime, x, y) %>%
+  summarize(
+    value = mean(value)
+  ) %>%
+  ungroup()
 
@@
 WR_vol <- convert_dist(cellSize)^2 * dimWR[3]
-round((1 - exp(-max(main_qc_wr_mo$value) * 600 * IR_sit / WR_vol)) * 100, 3)
-round((1 - exp(-max(main_qc_wr_af$value) * 600 * IR_sit / WR_vol)) * 100, 3)
+wr_morn <- main_mean_qc %>%
+  filter(room == "Waiting room", daytime == "morning") %>%
+  pull(value) %>%
+  max()
+wr_noon <- main_mean_qc %>%
+  filter(room == "Waiting room", daytime == "afternoon") %>%
+  pull(value) %>%
+  max()
+round((1 - exp(-wr_morn * 3600 * IR_sit / WR_vol)) * 100, 3)
+round((1 - exp(-wr_noon * 3600 * IR_sit / WR_vol)) * 100, 3)
 
 # link to building
-main_qc <- rbind(
-  main_qc_wr_mo, main_qc_wr_af,
-  main_qc_cr_mo, main_qc_cr_af,
-  main_qc_tb_mo, main_qc_tb_af
-)
-main_qc <- left_join(roomCoord,
-  main_qc,
+main_mean_qc <- left_join(roomCoord,
+  main_mean_qc %>% dplyr::select(room, daytime, x, y, value),
   by = c("x", "y", "room")
@@
         fortify(waiting_room),
-        filter(main_qc, room == "Waiting room"),
+        filter(main_mean_qc, room == "Waiting room"),
         by = c("id")
@@
         fortify(corridor),
-        filter(main_qc, room == "Corridor"),
+        filter(main_mean_qc, room == "Corridor"),
         by = c("id")
@@
       data = left_join(fortify(tb_room),
-        filter(main_qc, room == "TB room"),
+        filter(main_mean_qc, room == "TB room"),
         by = c("id")
@@
     legend.key.height = unit(0.2, "cm"),
-    legend.key.width = unit(2, "cm")
-  ) +
-  guides(fill = guide_colorbar(title.position = "top"))
+    legend.key.width = unit(1.5, "cm"),
+    legend.title = element_text(margin = margin(r = 10))
+  )
 
@@
 ```{r}
-average_trans_risk <- function(files) {
-  # files
-  tr_files <- files[grepl("risk", files)]
-
-  # data
-  tr <- do.call(rbind, lapply(tr_files, function(f) {
-    readRDS(f) %>% mutate(
-      mthdy = stringi::stri_extract(f, regex = "\\d{2}-\\d{2}"),
-      sim = stringi::stri_extract(f, regex = "\\d{1,6}")
-    )
-  }))
-
-  average_tr <- tr %>%
-    mutate(patient_id = paste(mthdy, patient_id, sep = "-")) %>%
+average_trans_risk <- function(file) {
+  average_tr <- read.table(file) %>%
+    set_names(c("sim", "patient_id", "P")) %>%
+    mutate(
+      mthdy = stringi::stri_extract(file, regex = "\\d{2}-\\d{2}"),
+      patient_id = paste(mthdy, patient_id, sep = "-")
+    ) %>%
     group_by(patient_id) %>%
@@
       P_mean = mean(P),
-      P_med = median(P)
+      P025 = quantile(P, .025),
+      P975 = quantile(P, .975),
+      P25 = quantile(P, .25),
+      P75 = quantile(P, .75),
+      P20 = quantile(P, .2),
+      P80 = quantile(P, .8)
     ) %>%
@@
     dplyr::select(-tracking_end) %>%
-    reshape2::melt("patient_id") %>%
     mutate(
-      val_col = ifelse(value < .001, "Low",
-        ifelse(value < .01, "Medium", "High")
+      P_col = ifelse(P_mean < .001, "Low",
+        ifelse(P_mean < .01, "Medium", "High")
       ),
-      val_col = factor(val_col, levels = c("Low", "Medium", "High")),
-      val_ann = 1 - (1 - value)^12,
-      variable = ifelse(
-        grepl("mean", variable),
-        "Mean risk of infection",
-        "Medium risk of infection"
-      ),
-      variable = factor(
-        variable,
-        levels = paste(c("Mean", "Medium"), "risk of infection")
+      P_col = factor(P_col, levels = c("Low", "Medium", "High")),
+      P_ann = 1 - (1 - P_mean)^12
     )
-    )
-
   return(average_tr)
@@
 # average risk
-main_tr <- average_trans_risk(main_files)
+main_tr <- data.frame(
+  files = list.files("simulations/main/",
+    full.names = TRUE, pattern = "trans-risk.txt$",
+    recursive = TRUE
+  )
+)
+main_tr$value <- lapply(main_tr$files, average_trans_risk)
+
+main_tr <- main_tr %>% unnest(value)
 
 main_tr %>%
-  filter(variable == "Mean risk of infection") %>%
-  group_by(val_col) %>%
+  group_by(P_col) %>%
   summarise(n = n()) %>%
@@
   roi_annon <- tr %>%
-    group_by(variable) %>%
     summarize(
-      mean_P = mean(value),
-      lower_P = quantile(value, .025),
-      upper_P = quantile(value, .975),
-      mean_P_ann = mean(val_ann),
-      lower_P_ann = quantile(val_ann, .025),
-      upper_P_ann = quantile(val_ann, .975)
+      mean_P = mean(P_mean),
+      lower_P = quantile(P_mean, .025),
+      upper_P = quantile(P_mean, .975),
+      mean_P_ann = mean(P_ann),
+      lower_P_ann = quantile(P_ann, .025),
+      upper_P_ann = quantile(P_ann, .975)
     ) %>%
-    ungroup() %>%
     mutate(
@@
     )
-
   return(roi_annon)
@@
 
-main_annon
+main_annon$lab
+main_annon$lab_ann
 
@@
 roi_pl <- ggplot(
-  data = main_tr %>% filter(variable == "Mean risk of infection"),
-  mapping = aes(x = value)
+  data = main_tr,
+  mapping = aes(x = P_mean)
 ) +
   geom_histogram(
-    mapping = aes(fill = val_col),
+    mapping = aes(fill = P_col),
     binwidth = .0005,
@@
   geom_text(
-    data = main_annon %>% filter(variable == "Mean risk of infection"),
+    data = main_annon,
     mapping = aes(x = Inf, y = Inf, label = lab),
@@
   scale_x_continuous(
-    labels = function(x) x * 100
+    labels = function(x) x * 100,
+    breaks = c(0, 0.001, 0.005, 0.01, 0.015, 0.02),
+    limits = c(0, 0.02)
   ) +
-  scale_y_continuous(
-    labels = scales::comma
+  scale_y_sqrt(
+    labels = scales::comma,
+    breaks = c(0, 25, 100, 250, 500, 1000)
   ) +
@@
     x = "Risk of infection (%)",
-    y = "Number of clinical attendees",
+    y = "Number of people",
     title = "b"
@@
   theme(
-    legend.position = "bottom",
+    legend.position = c(0.7, .3),
     legend.title = element_blank(),
-    legend.key.height = unit(.6, "cm"),
-    legend.key.width = unit(.4, "cm"),
+    legend.key.height = unit(.3, "cm"),
+    legend.key.width = unit(.2, "cm"),
+    legend.direction = "horizontal",
     panel.spacing = unit(2, "cm")
@@
 roi_pl
+```
+
+```{r}
+main_tr <- main_tr %>%
+  mutate(P_unc = P75 - P25)
+
+hr_lu_pt <- main_tr %>%
+  filter(P_unc < .25 / 100, P_mean > 0.25 / 100) %>%
+  head(1)
+
+main_focus_pl <- main_tr %>%
+  ggplot(aes(x = P_unc, y = P_mean)) +
+  geom_point(shape = 4) +
+  annotate(
+    "point",
+    x = hr_lu_pt$P_unc,
+    y = hr_lu_pt$P_mean,
+    size = 7,
+    shape = 1,
+    color = "red"
+  ) +
+  annotate(
+    geom = "segment",
+    x = hr_lu_pt$P_unc + .00002,
+    y = hr_lu_pt$P_mean + .0004,
+    xend = hr_lu_pt$P_unc + .00008,
+    yend = hr_lu_pt$P_mean + .004,
+    arrow = arrow(length = unit(0.1, "cm")),
+    color = "red"
+  ) +
+  annotate(
+    geom = "text",
+    x = hr_lu_pt$P_unc + .00008,
+    y = hr_lu_pt$P_mean + .004,
+    label = "Moderate risk,\nlow uncertainty",
+    hjust = -.1,
+    size = 6.5 / cm(1),
+    color = "red"
+  ) +
+  scale_y_continuous(
+    expand = expansion(mult = c(0.01, 0.01)),
+    limits = c(0, .01),
+    breaks = seq(0, .01, .0025),
+    labels = function(x) round_k(100 * x, 3)
+  ) +
+  scale_x_continuous(
+    expand = expansion(mult = c(0.01, 0.01)),
+    limits = c(0, .002),
+    breaks = seq(0, .002, .0005),
+    labels = function(x) round_k(100 * x, 3)
+  ) +
+  labs(
+    x = "Uncertainty (interquartile range, pp)",
+    y = "Mean risk of infection (%)",
+    title = "c"
+  ) +
+  theme_bw2() +
+  theme(
+    plot.title.position = "plot",
+    plot.title = element_text(size = 10, face = 2),
+  )
+
+main_focus_pl
 ```
@@
   left_join(
-    reshape2::dcast(main_tr, patient_id ~ variable) %>%
-      dplyr::select(patient_id, `Mean risk of infection`) %>%
-      rename(mean_risk = `Mean risk of infection`)
+    main_tr %>%
+      dplyr::select(patient_id, P_mean)
   )
@@
 # model
-amod <- stan_betareg(mean_risk ~ time_in_clinic + n_cc + cc_time,
+amod <- stan_betareg(P_mean ~ time_in_clinic + n_cc + cc_time,
   data = assoc_df %>% mutate(
@@
   labs(
-    x = "OR of modeled risk if 2 x predictor",
-    title = "c",
+    x = "Odds ratio of the modeled risk of infection for a doubling in the predictor",
+    title = "d",
     color = "CrI"
@@
     axis.title.y = element_blank(),
-    legend.position = "bottom"
+    legend.position = "right",
+    legend.key.size = unit(.3, "cm")
   )
@@
   mean_quanta_pl,
-  grid.arrange(roi_pl, amod_est_pl, ncol = 2),
+  grid.arrange(roi_pl, main_focus_pl, ncol = 2),
+  amod_est_pl,
   ncol = 1,
-  heights = c(8, 8)
+  heights = c(10, 6, 4)
 )
@@
 
+## Comparison with temporal model
+
+```{r}
+# files
+files_temp <- list.files("simulations/temporal/",
+  full.names = TRUE, pattern = "trans-risk",
+  recursive = TRUE
+)
+
+temp_tr <- map_dfr(files_temp, average_trans_risk)
+temp_annon <- annotate_average_trans_risk(temp_tr)
+
+# comparison with temporal model
+mod_comp <- left_join(
+  temp_tr %>%
+    dplyr::select(patient_id, P_mean) %>%
+    rename(`Temporal model` = P_mean),
+  main_tr %>%
+    dplyr::select(patient_id, P_mean) %>%
+    mutate(`Spatiotemporal model` = P_mean)
+) %>%
+  mutate(
+    ratio = abs(`Spatiotemporal model` - `Temporal model`) / `Temporal model` * 100
+  )
+
+summary(mod_comp$ratio)
+```
+
 ## Additional analyses
@@
 files_nm <- list.files("simulations/nomasks/",
-  full.names = TRUE, pattern = "*.rds$",
+  full.names = TRUE, pattern = "trans-risk",
   recursive = TRUE
 )
-files_vent <- list.files("simulations/ventilation/",
-  full.names = TRUE, pattern = "*.rds$",
-  recursive = TRUE
-)
-files_temp <- list.files("simulations/temporal/",
-  full.names = TRUE, pattern = "*.rds$",
+files_vent <- list.files("simulations/vent/",
+  full.names = TRUE, pattern = "trans-risk",
   recursive = TRUE
@@
 files_prev <- list.files("simulations/prevalence/",
-  full.names = TRUE, pattern = "*.rds$",
+  full.names = TRUE, pattern = "trans-risk",
   recursive = TRUE
@@
 files_susp <- list.files("simulations/suspected/",
-  full.names = TRUE, pattern = "*.rds$",
+  full.names = TRUE, pattern = "trans-risk",
   recursive = TRUE
@@
 # average risk
-nm_tr <- average_trans_risk(files_nm)
-vent_tr <- average_trans_risk(files_vent)
-temp_tr <- average_trans_risk(files_temp)
-prev_tr <- average_trans_risk(files_prev)
-susp_tr <- average_trans_risk(files_susp)
+nm_tr <- map_dfr(files_nm, average_trans_risk)
+vent_tr <- map_dfr(files_vent, average_trans_risk)
+prev_tr <- map_dfr(files_prev, average_trans_risk)
+susp_tr <- map_dfr(files_susp, average_trans_risk)
 
+nm_tr %>%
+  group_by(P_col) %>%
+  summarise(n = n()) %>%
+  ungroup() %>%
+  mutate(prop = round(n / sum(n) * 100, 0))
+
+vent_tr %>%
+  group_by(P_col) %>%
+  summarise(n = n()) %>%
+  ungroup() %>%
+  mutate(prop = round(n / sum(n) * 100, 0))
+
+susp_tr %>%
+  group_by(P_col) %>%
+  summarise(n = n()) %>%
+  ungroup() %>%
+  mutate(prop = round(n / sum(n) * 100, 0))
+
+main_tr %>%
+  dplyr::select(patient_id, P_mean) %>%
+  rename(`Spatiotemporal model` = P_mean) %>%
+  left_join(
+    nm_tr %>%
+      dplyr::select(patient_id, P_mean) %>%
+      rename(`No mask wearing` = P_mean)
+  ) %>%
+  left_join(
+    vent_tr %>%
+      dplyr::select(patient_id, P_mean) %>%
+      rename(`Lower ventilation` = P_mean)
+  ) %>%
+  mutate(
+    mask_factor = `No mask wearing` / `Spatiotemporal model`,
+    vent_factor = `Lower ventilation` / `Spatiotemporal model`
+  ) %>%
+  mean_qi(mask_factor, vent_factor) %>%
+  gather() %>%
+  head(6) %>%
+  mutate(value = round(as.numeric(value), 2))
+
+# combine
 tr_comp <- rbind(
@@
 vent_annon <- annotate_average_trans_risk(vent_tr)
-temp_annon <- annotate_average_trans_risk(temp_tr)
 prev_annon <- annotate_average_trans_risk(prev_tr)
@@
 mean_roi_comp_pl <- ggplot(
-  tr_comp %>%
-    filter(variable == "Mean risk of infection"),
-  aes(x = value)
+  tr_comp,
+  aes(x = P_mean)
 ) +
   geom_histogram(
-    mapping = aes(fill = val_col),
-    binwidth = .05,
+    mapping = aes(fill = P_col),
+    binwidth = .0005,
     boundary = 1
@@
   geom_text(
-    data = annon_comp %>% filter(variable == "Mean risk of infection"),
+    data = annon_comp,
     mapping = aes(x = Inf, y = Inf, label = lab),
@@
   scale_x_continuous(
-    breaks = c(0, .05, 0.1, seq(0.2, .5, .1)),
-    labels = function(x) x * 100
+    labels = function(x) x * 100,
+    breaks = c(0, 0.001, 0.005, 0.01, 0.015, 0.02),
+    limits = c(0, 0.02)
   ) +
-  scale_y_continuous(
-    labels = scales::comma
+  scale_y_sqrt(
+    labels = scales::comma,
+    breaks = c(0, 25, 100, 250, 500, 1000),
+    limits = c(0, 1000)
   ) +
@@
 )
-
-# comparison with temporal model
-mod_comp <- left_join(
-  temp_tr %>% rename(`Temporal model` = value) %>% dplyr::select(-val_col),
-  main_tr %>% mutate(`Spatiotemporal model` = value) %>% dplyr::select(-val_col)
-) %>%
-  mutate(
-    diff = `Spatiotemporal model` - `Temporal model`
-  ) %>%
-  mutate(
-    val_col = ifelse(diff > .005, "Higher",
-      ifelse(diff < -.005, "Lower", "Similar")
-    ),
-    val_col = factor(val_col, levels = c("Lower", "Similar", "Higher"))
-  )
-
-mod_comp %>%
-  filter(variable == "Mean risk of infection") %>%
-  group_by(val_col) %>%
-  summarize(n = n()) %>%
-  mutate(prop = round(n / sum(n) * 100, 0))
-
-mod_comp_annon <- mod_comp %>%
-  summarise(
-    mean_diff = mean(diff),
-    lower_diff = quantile(diff, .025),
-    upper_diff = quantile(diff, .975)
-  ) %>%
-  mutate(across(c(mean_diff, lower_diff, upper_diff), ~ round_k(100 * .x, 1)),
-    lab = paste0(
-      "Average ", mean_diff, "%",
-      " (95%-CrI ", lower_diff, "% – ", upper_diff, "%)"
-    )
-  )
-
-mod_comp_pl <- mod_comp %>%
-  ggplot(aes(x = diff)) +
-  geom_histogram(
-    mapping = aes(fill = val_col),
-    binwidth = .005,
-    boundary = 1
-  ) +
-  geom_text(
-    data = mod_comp_annon,
-    mapping = aes(x = Inf, y = Inf, label = lab),
-    size = 8 / cm(1),
-    hjust = 1, vjust = 2
-  ) +
-  labs(
-    y = "Number of clinical attendees",
-    x = "Difference between spatiotemporal and temporal model (perc. points)"
-  ) +
-  scale_y_log10(labels = scales::comma) +
-  scale_x_continuous(labels = function(x) scales::percent(x, suffix = "")) +
-  scale_fill_manual(values = wes_palette("Zissou1")[c(1, 3, 5)]) +
-  coord_cartesian(expand = FALSE) +
-  theme_custom() +
-  theme(
-    legend.position = "bottom",
-    legend.title = element_blank(),
-    legend.key.height = unit(.6, "cm"),
-    legend.key.width = unit(.4, "cm")
-  )
-
-save_plot(
-  mod_comp_pl,
-  "results/modeling/model-comparison.png",
-  w = 12,
-  h = 8
-)
 ```

### 2024-04-10 doc/paper/manuscript.tex
Updates from Overleaf

@@
 % author list
-\author[1$\ddag$]{Nicolas Banholzer}
+\author[1]{Nicolas Banholzer}
 \author[2]{Keren Middelkoop}
 \author[2]{Juane Leukes}
+\author[3]{Ernest Weingartner}
+\author[1]{Remo Schmutz}
 \author[1]{Kathrin Zürcher}
@@
 \affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}
+\affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
 
@@
 TB is strictly airborne and thus the transmission risk is higher in crowded indoor settings, especially healthcare facilities.
-Existing approaches to model the risk of airborne transmission typically assume that the airspace is well mixed, thereby ignoring that closer proximity to infectious individuals may increase the risk of transmission.
+Existing approaches to model the risk of airborne transmission typically assume that the airspace is well mixed, thereby not considering that closer proximity to infectious individuals may increase the risk of transmission.
 
-\noindent\textbf{Methods:} We developed a spatiotemporal extension of the Wells-Riley model to estimate the risk of \emph{Mtb} transmission using environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors).
-The data were collected for five days in October/November 2021 in a primary care clinic in Cape Town, South Africa.
+\noindent\textbf{Methods:} We developed a spatiotemporal model of airborne \emph{Mtb} transmission using environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors).
+The data were collected n a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when mask-wearing was compulsory.
 We matched patient movements with clinical records to identify the spatiotemporal location of TB infectious patients.
-Based on that, we modeled the spatiotemporal concentration of infectious doses and estimated the patient-specific risk of infection using Monte Carlo simulation.
+Using Monte Carlo simulation, we modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection with the Wells-Riley equation.
 
-\noindent\textbf{Results:} Clinical records registered 894 patients visiting the clinic, four of which were diagnosed with TB.
-Patient movement data suggested 1,438 people visiting the clinic (median stay 25\,min, interquartile range [IQR] 13\,min$-$46\,min).
-CO$_2$ levels were mostly below 600\,ppm, suggesting that the clinic was very well ventilated.
-The modeled mean risk of infection was 7\% (95\%-credible interval [CrI] 0.5\%$-$31.1\%) and the risk was significantly associated with the number of close-contact encounters, time spent in the clinic, and the air change rate.
-Without infection control measures that were implemented during the COVID-19 pandemic, the modeled risk of infection would have been 7.8\% (95\%-CrI 0.4\%$-$41.3\%) without mask wearing and 8.4\%-CrI 0.6\%$-$41.4\%) with lower ventilation.
+\noindent\textbf{Results:} Clinical records registered 894 patients, four of which were diagnosed with TB.
+Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min).
+CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), suggesting that the clinic was well ventilated.
+The modeled risk of infection per clinic attendee was 0.059\% (95\%-credible interval [CrI] 0.001\%$-$0.402\%).
+The modeled risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask wearing and 1.74-fold (95\%-CrI 0.47$-$5.34).
+Furthermore, the person-specific risk was significantly associated with the number of close-contact encounters and time spent in the clinic.
 
-\noindent\textbf{Conclusions:} Risk of \emph{Mtb} transmission in primary care clinic in South Africa is considerable, especially for attendees with many close-contact interactions.
-Infection control measures probably mitigated the risk of \emph{Mtb} transmission.
-Our spatiotemporal model could be used in future work to assess the impact of measures preventing prolonged close contact with potentially infectious patients.
+\noindent\textbf{Conclusions:} Risk of \emph{Mtb} transmission in the South African clinic was low during the COVID-19 pandemic, especially due to infection control measures.
+Our spatiotemporal modeling could be used to identify high-risk patients and assess the impact of mitigation measures.
 
@@
 
-\noindent\textbf{Keywords:} Mycobacterium tuberculosis, patient movements, Wells-Riley model
+\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, patient tracking, spatiotemporal model
 % maximum of 3-5 keywords
@@
 % TB and the route of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South-East Asian and the African region, and progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed\cite{WHO2022TBReport}.
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South African region, and progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed\cite{WHO2022TBReport}.
 \emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
-\emph{Mtb} is carried primarily in smaller particles in the size range of $2\mu$m to $5\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
-Airborne survival is greater in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021} and healthcare facilities present a particular high-risk setting because both people that are more infectious and susceptible also more frequently visit the clinic\cite{McCreesh2020IJTLD}.
+\emph{Mtb} is carried primarily in smaller particles in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
+Survival of airborne pathogens is greater in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
+As such, healthcare facilities present a particular high-risk setting because both people that are more infectious and susceptible also more frequently visit the clinic\cite{McCreesh2020IJTLD}.
 A study in South Africa estimated that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
@@
 The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
-The model assumes estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate.
-The Wells-Riley, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same everywhere in the room.
+The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate.
+The Wells-Riley model, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same everywhere in the room.
 However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies showed that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
@@
 We build on the Wells-Riley model and develop a spatiotemporal extension to model the concentration of infectious quanta spatially and over time.
-We combine clinical and video sensor data to identify the spatiotemporal location of TB infectious and susceptible patients visiting a primary care clinic in South Africa during five days in October/November 2021.
+We combine clinical and video sensor data to identify the spatiotemporal location of TB patients and susceptible attendees of a primary care clinic in South Africa for five days in October/November 2021.
 Furthermore, we measure indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time.
-Using Monte Carlo simulation, our model allows us to estimate the risk of infection individually for each clinical attendee.
-Furthermore, we estimated the impact of infection control measures which were in place during the COVID-19 pandemic using hypothetical scenarios.
+Using Monte Carlo simulation, we estimate the risk of infection per clinic attendee.
+Finally, we assess the impact of infection control measures that were in place during the study in response to the COVID-19 by modeling the risk of infection under hypothetical scenarios without mask wearing and assuming pre-pandemic ventilation conditions.
 
@@
 The primary care clinic offers both TB and HIV services and other basic clinical services, Monday to Friday, from 8\,am to 4\,pm.
-The clinic is situated within a large settlement of formal and semiformal housing where both TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
+The clinic is situated within a large settlement of formal and semiformal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
 We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
-The TB room was primarily used for diagnosed TB patients and suspected patients with respiratory symptoms.
+The TB room was mainly used for the treatment of diagnosed TB patients, but also for the screening of suspected patients with respiratory symptoms who were eventually not diagnosed with TB.
 
@@
 Missing values were linearly imputed.
-On October~13, the corridor's CO$_2$ levels were missing and imputed with the waiting room's CO$_2$ levels because their CO$_2$ measurements were similar on the other study days.
+On October~13, the corridor's CO$_2$ levels were missing and imputed with the waiting room's CO$_2$ levels because their CO$_2$ measurements were similar on the other days.
 
@@
 
-We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study period.
-These data included the date and time of arrival, TB diagnostic results, and date of TB treatment start.
+We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study.
+Clinical records included the date and time of arrival, TB diagnostic results, and date of TB treatment start.
 We defined infectious patients as individuals testing bacteriologically positive for TB during the study or being on TB treatment for less than four weeks.
-Patients who had been treated for more than four weeks or recently completed a treatment were neither considered infectious nor susceptible.
+Patients who had been treated for more than four weeks or recently completed treatment were neither considered infectious nor susceptible.
 We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were subsequently not diagnosed with TB.
@@
 
-We used an anonymous personal tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinical staff, patients, and other visitors) throughout the clinic at 1s intervals (\Cref{fig:floor-plan}).
+We used an anonymous personal tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1s intervals (\Cref{fig:floor-plan}).
 The resulting timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-\supp Figure~\zref{fig:tracking-examples} shows a few sample tracks from our study.
-Note that individuals could contribute multiple tracks if they moved out of the sensor's range or were briefly not recognized as a person.
+\supp~Figure~\zref{fig:tracking-examples} shows a few sample tracks from our study.
+Note that persons in the clinic could contribute multiple tracks if they moved out of a sensor's range or if they were briefly not recognized as a person.
 To remedy these issues, we created an R Shiny tool to manually link interrupted tracks back together (\supp Text~\zref{sec:setting-and-data}).
-Furthermore, we labeled tracks to identify movements from clinical staff.
+Furthermore, we labeled tracks to identify movements from clinic staff.
 
@@
 
-Our spatiotemporal model builds upon the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability) of infection $P$ as
+Our spatiotemporal model builds upon the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability in \%) of infection $P$ as
 \begin{align}
@@
 \end{align}
-where $D$ is the number of diseases cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ s$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta s$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ s$^{-1}$).
+where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
 The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
-Our aim is to relax this assumption by using patient movement data to identify the location of infectious patients and consider spatial variation in the quanta concentration.
+Instead, our study considers a mixed airspace with spatial variation in the concentration of quanta.
+This is accomplished using patient movement data combined with clinical records to identify the spatiotemporal location of infectious patients who generate quanta.
 
@@
 
-A detailed description of the Wells-Riley model and our spatiotemporal extension is provided in \supp Text~\zref{sec:spattemp-model}.
+A detailed description of the Wells-Riley model and our spatiotemporal extension is provided in \supp~Text~\zref{sec:spattemp-model}.
 In the following, we describe the main components of our modeling approach, which are summarized in \Cref{fig:modeling-flow}.
@@
     \centering
-    \includegraphics{doc/paper/flow-chart.pdf}
+    \includegraphics{doc/model-flow-chart.pdf}
     \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
@@
 (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles.
-(6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, based on which the personal risk of infection is computed.}
+(6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, based on which the personal risk of infection is computed using the Wells-Riley equation.}
     \label{fig:modeling-flow}
@@
 (1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients inside the clinic.
-We recorded the timestamp when the patient movement was in the registration area and matched this with the registration time in the clinical records.
+We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records.
 We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching.
@@
 All diagnosed TB patients could be matched with patient movements this way.
-(2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed patients.
-Prior studies suggest that the number of undiagnosed TB patients is proportional to the number of diagnosed TB patients \cite{Berhanu2023CID}.
-Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:count-tb-patients}): 6 (12\%) days with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four visiting TB patients.
-Undiagnosed patients were randomly sampled among all visitors with the same probability.
-Clinical staff were not considered infectious and excluded from the sampling.
-We assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution and considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
-Since mask wearing was mandatory for all clinical staff and visitors during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+(2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed (subclinical) TB patients.
+Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
+Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
+Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
 
@@
 (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}.
-Since airflow was not monitored in our study, we assume that the quanta diffuse radially from where they were generated (\supp~Figure~\zref{fig:toy-example}).
+Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-The bacterial inactivation rate was modeled using a lognormal prior distribution (\supp~Figure~\zref{}) with a median of 1\,quanta h$^{-1}$ (0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $2\mu$m to $5\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
 
@@
 \end{align}
-where $D$ is the diffusion constant (m$^2$ s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta s$^{-1}$).
+where $D$ is the diffusion constant (m$^2$ s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta\,s$^{-1}$).
 The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
@@
 \end{align}
-where $N_{s,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ is indicates whether the attendee was in $c$ at time $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+where $N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
 We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the distance between two patient tracks was more or less than 0.25m.
@@
 
-The model setup and simulation approach is described in detail in Text~\zref{sec:estimation}, including a comprehensive review of our modeling assumptions and prior distributions.
+Our setup and simulation approach are described in detail in Text~\zref{sec:estimation}, including a comprehensive review of our modeling assumptions and choices of prior distributions.
 To summarize, we simulated the spatiotemporal quanta concentration separately for the waiting room, corridor, and TB room of the clinic.
@@
 We used a Monte-Carlo simulation approach to consider uncertainty in key modeling parameters.
-In each simulation, we sampled uncertain modeling parameters from their prior distributions and, based on that, computed the spatiotemporal quanta concentration and the risk of infection per clinical attendee.
+In each simulation, we sampled uncertain modeling parameters from their prior distributions and, based on that, computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
 
@@
 Our study was set during the COVID-19 pandemic when infection control measures were in place.
-To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~lower ventilation, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pilot study of 2019 before the pandemic\cite{Zurcher2022JID}.
-Furthermore, we examined the impact of different assumptions regarding the number infectious patients in space: 1)~instead of undiagnosed TB patients, we considered suspected TB patients as infectious; 2)~random sampling of infectious patients among all visitors in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
+To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pilot study of 2019 before the pandemic\cite{Zurcher2022JID}.
+Furthermore, we examined the impact of different assumptions regarding the number of infectious patients in the clinic: 1)~instead of undiagnosed TB patients, we considered suspected TB patients as infectious, because these patients may still generate aerosolized \emph{Mtb} even when they were not diagnosed with TB at first presentation\cite{Patterson2024PNAS}; 2)~random sampling of infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}, thereby pretending that the clinical data had not been available.
 Except for the scenario-specific modifications, all other modeling assumptions and parameters were the same in each Monte Carlo simulation, ensuring comparability of the results.
@@
 
-Clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates) and patient movement data (spatiotemporal number of tracks, visit time, and close-contact encounters) were assessed descriptively and summarized with the median and interquartile range (IQR).
-Spatiotemporal quanta concentration and risk of infection per clinical attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean/median and 95\%-credible interval (CrI).
-A Bayesian Beta regression model was used to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the air change rate, visit time, and close-contact encounters.
+Clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and patient movement data (spatiotemporal number of tracks, visit time, and close-contact encounters) were assessed descriptively and summarized with the median and interquartile range (IQR).
+Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
+The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean and 95\%-credible interval (CrI).
+A Bayesian Beta regression model was used to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact, and frequency of close-contact encounters.
 All analyses were performed in R software (version 4.3.1) and the Bayesian regression model was estimated in the probabilistic programming language Stan (version 2.26.1).
@@
 During the five study days, 894 patients have been registered in the clinical database, of which four patients were diagnosed with TB and 16 patients were suspected of TB following symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}).
-Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as being from healthcare workers, identifying 1,438 patients or other visitors.
+Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as being from healthcare workers, identifying 1,438 clinic attendees (patients or other visitors).
 Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
-Patients spent around half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
-Close-contact encounters (distance <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9), and several patients had more than ten close contacts during their visit (\Cref{fig:input-data-descriptives}d).
-The majority of patients was in close contact with at least one other person for more than half the time of their visit (median 68\%, IQR 41\%$-$87\%, \Cref{fig:input-data-descriptives}e).
-The CO$_2$ levels inside the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\Cref{fig:input-data-descriptives}f) and the corresponding air change rates were high (\Cref{fig:input-data-descriptives}g), especially the densely occupied waiting room (median 29\,h$^{-1}$), IQR 9\,h$^{-1}$$-$36\,h$^{-1}$).
+Clinic attendees spent around half an hour inside the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
+Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9), and several clinic attendees had more than ten different close contacts during their visit (\Cref{fig:input-data-descriptives}d).
+The majority of clinic attendees was in close contact with at least one other person for more than half the time of their visit (median 68\%, IQR 41\%$-$87\%, \Cref{fig:input-data-descriptives}e).
+The CO$_2$ levels inside the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\Cref{fig:input-data-descriptives}f) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
 
@@
 \textbf{(d)}~Number of people in the clinic over time per day.
-\textbf{(e)}~Time spent in the clinic per clinical attendee.
-\textbf{(f)}~Density distribution of the number of close contacts per clinical attendee during their visit.
-\textbf{(g)}~Density distribution of the time spent in close contact per clinical attendee during their visit.}
+\textbf{(e)}~Time spent in the clinic per attendee.
+\textbf{(f)}~Density distribution of the number of close contacts per clinic attendee during their visit.
+\textbf{(g)}~Density distribution of the proportion of visit time spent in close contact per clinic attendee.}
     \label{fig:input-data-descriptives}
@@
 
-As an illustrative example of our spatiotemporal model, the animation in \supp File 2 shows the spatiotemporally varying quanta concentration as an infectious patient enters the clinic in the morning of October 25 and moves through the waiting room.
-The quanta concentration is higher near the infectious patient, therefore indicating his position and movement.
-It can further be seen that the quanta diffuses fairly quickly due to its relation with the air change rate, which was very high because all doors and windows were typically open in the clinic.
+\supp~File~1 shows the spatiotemporally varying quanta concentration as an infectious patient enters the clinic in the morning of October 25 and moves through the waiting room.
+It can be seen that the quanta concentration was higher near the infectious patient, therefore indicating his position and movement.
+The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open throughout clinic hours.
 \Cref{fig:main-modeling-results} shows the results from our main spatiotemporal modeling analysis across study days and simulations.
-The mean quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a).
-A ten minute exposure to the mean quanta concentration in the center of the waiting room would correspond to a 52\% risk of infection in the morning compared to a 16\% risk in the afternoon.
-\Cref{fig:main-modeling-results}b shows the mean and median risk of infection per clinical attendee.
-The attendee-average mean risk of infection was 7\% (95\%-CrI 0.5\%$-$31.1\%) and the median risk was 0.8\% (95\%-CrI 0.0\%$-$11.2\%).
-The mean risk was typically higher than the median risk because the risk of infection was very low in many simulations and only significant in a few ones.
-The majority of attendees (51\%) had a low mean risk of infection <5\%, roughly a third (31\%) had a medium risk between 5\%$-$10\%, and 18\% had a high risk of infection >10\%.
-A comparison of our spatiotemporal modeling results with those from a temporal model assuming a well-mixed airspace shows a notable difference of more than 0.5 percentage points in the mean risk of infection for 168 attendees (12\%, \supp~Figure~\zref{fig:model-comparison}).
-The mean risk of infection from our spatiotemporal model was associated the number of close contacts, time spent in clinic, and the air change rate (\Cref{fig:main-modeling-results}c).
-If the number of close contacts doubled, \eg from five to ten, the odds of infection were 1.4 (95\%-CrI 1.28$-$1.53).
-If the air change rate doubled, \eg from 15 to 30 air changes per hour, the odds of infection were 0.84 (95\%-CrI 0.81$-$0.88).
+On average, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a).
+An exposure of one hour to the mean quanta concentration in the center of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
+The mean risk of infection was 0.059\% (95\%-CrI 0.001\%$-$0.402\%) and the majority of clinic attendees (87\%) had a low risk <0.01\% risk of infection (\Cref{fig:main-modeling-results}b).
+To put this into perspective, a patient visiting the clinic twelve times per year would have a cumulative annual risk of infection of 0.7\% (95\%-CrI 0.01\%$-$4.72\%).
+
+Our modeling approach allowed estimating the person-specific risk of infection with uncertainty.
+As shown in \Cref{fig:main-modeling-results}b, it can be used to identify patients with a comparably higher mean risk of infection at low uncertainty.
+The person-specific modeled risk of infection was further associated with movement characteristics (\Cref{fig:main-modeling-results}c).
+If the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30).
+Analogously, if the time spent in the clinic doubled, the odds of infection increased to 1.09 (95\%-CrI 1.01$-$1.18).
+Finally, a comparison of our spatiotemporal modeling results with those from a temporal model assuming a well-mixed airspace showed notable difference in the person-specific risk of infection with a median of 8\% (IQR 4\%$-$15\%).
 
@@
     \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Average quanta concentration in the morning and afternoon in the clinic.
-\textbf{(b)}~Mean and median risk of infection per clinical attendee.
+\textbf{(b)}~Mean risk of infection per clinic attendee.
 Low: <5\%, Medium: 5-10\%, High: >10\%.
-\textbf{(c)}~Association of mean infection risk with patient movement characteristics and environmental conditions.
-}
+\textbf{(c)}~Person-specific mean risk of infection vs uncertainty (IQR in percentage points, pp).
+\textbf{(d)}~Association of person-specific mean risk of infection with patient movement characteristics (mean odds ratio [OR] and 95\%-credible interval [CrI] if the value of the predictor doubles).}
     \label{fig:main-modeling-results}
@@
 To assess the impact of infection control measures during the COVID-19 pandemic, we performed two hypothetical scenarios shown in the top panel of \Cref{fig:scenario-results}.
-If none of the attendees or staff had worn masks, the attendee-average mean risk of infection would have been almost a percentage point higher (7.8\%, 95\%-CrI 0.4\%$-$41.3\%).
-If the ventilation rate had been fixed at 16 air changes per hour (the average in 2019 before the pandemic), the attendee-average mean risk of infection would have been more than percentage point higher (8.4\%-CrI 0.6\%$-$41.4\%).
-In both scenarios, the upper credible interval is roughly ten percentage points higher, indicating that mask wearing and better ventilation considerably impact the tail of the distribution, \ie high risks of infection.
-To assess the impact of modeling assumptions regarding the number of infectious people in space, we performed two hypothetical scenarios shown in the bottom panel of \Cref{fig:scenario-results}.
-Assuming there were no undiagnosed TB patients but all attendees suspected of TB were infectious, the attendee-average mean risk of infection would have been considerably higher (13.9\%, 95\%-CrI 0.0\%$-$45.2\%).
-Assuming there were only undiagnosed TB patients in proportion to the prevalence of TB in the South African population, the attendee-average risk of infection would have been lower (6.3\%, 95\%-CrI 0.8\%$-$16.3\%).
+If none of the clinic attendees or staff had worn masks, the mean risk of infection would have been 0.075\% (95\%-CrI 0.001\%$-$0.513\%), a 1.28-fold (95\%-CrI 1.19$-$1.36) increase, and there would have been four patients with a high risk of infection >1\%, which would translate into a cumulative annual risk of more than 10\%.
+Similarly, if the air change rate had been as before the COVID-19 pandemic, the mean risk of infection would have been (0.097\%, 95\%-CrI 0.001\%$-$0.820\%), a 1.74-fold (95\%-CrI 0.47$-$5.34) increase, and there would have been 27 patients with a high risk of infection >1\%.
+Both scenarios show an increase in the number of clinic attendees with a high risk of infection, indicating that mask wearing and better ventilation considerably impacted the tail of the risk distribution, which can also be seen in the smaller upper limit of the credible intervals.
+
+To assess the impact of modeling assumptions regarding the number of infectious people in the clinic, we performed two hypothetical scenarios shown in the bottom panel of \Cref{fig:scenario-results}.
+Assuming the number of infectious TB patients had been in proportion to the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.079\%, 95\%-CrI 0.003\%$-$0.304\%).
+Assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean and tail risk of infection would have been considerably higher (0.133\%, 95\%-CrI 0.000\%$-$0.819\%), and there would have been 21 clinic attendees with a high risk of infection >1\%.
 
@@
 % summary
-We developed a spatiotemporal model to estimate the concentration of infectious quanta in an indoor space and applied it to patient movement data from a South African primary care clinic in 2021 to simulate the risk of TB infection for clinical attendees.
-We estimated a mean risk of infection of 7\% (95\%-CrI 0.5\%$-$31.1\%) across clinical attendees.
-Clinical attendees with more close-contact encounters had a higher risk of infection.
-The modeled risk of infection would have been higher without the implementation infection control measures during the COVID-19 pandemic such as compulsory face mask wearing and better natural ventilation.
+We developed a spatiotemporal model to estimate the concentration of infectious quanta in an indoor space and applied it to patient movement data from a South African primary care clinic in 2021 to simulate the risk of TB infection for clinic attendees.
+We estimated an average risk of infection of 0.059\% (95\%-CrI 0.001\%$-$0.402\%), which would correspond to a cumulative annual risk of 0.7\% (95\%-CrI 0.01\%$-$4.72\%) if each patient visited the clinic once per month.
+Our model can identify patients with a higher risk of infection and low uncertainty.
+Furthermore, we found that clinic attendees who stayed longer inside the clinic and had more close-contact encounters also had a higher risk of infection.
+The modeled risk of infection would have been higher without the implementation infection control measures during the COVID-19 pandemic such as compulsory face mask wearing and better ventilation conditions.
+We applied our spatiotemporal model to \emph{Mtb} but it can be readily adapted to model the airborne transmission risk of other respiratory infections such as SARS-CoV-2 or influenza.
 
@@
 South Africa has a high burden of TB\cite{WHO2022TBReport}.
+Despite that, we estimated a low risk of infection per clinic attendee in our primary care clinic.
+In our previous study before the COVID-19 pandemic, we estimated a cumulative annual risk of \emph{Mtb} transmission of between 9\% and 29\% for a patient coming to the clinic once per month for one hour.
+Although the assumed visit time is longer than what we observed in this study, the risk difference is still noteworthy and most likely due to the infection control measures that were in place during the SARS-CoV-2 pandemic.
+We estimated that without mask wearing or with pre-pandemic ventilation rates, the risk of infection would on average have been 1.28-fold or 1.74-fold higher, respectively.
+But mask wearing and natural ventilation were not the only measures in place during the pandemic, patient triage and fixed appointments probably also contributed to a further reduction in the risk of \emph{Mtb} transmission in our study.
+Finally, the risk difference could, to a small extent, be related to other factors.
+The risk estimate before the COVID-19 pandemic was based on TB prevalence in the South African population.
+We also estimated a higher mean risk of infection when modeling the risk based on TB prevalence or when considering both diagnosed and suspected TB patients as infectious.
+
+% relation to infection control
 A previous modeling study estimated that between 4\% and 14\% of TB transmission in adults in a high TB burden, high HIV prevalence community in South Africa occurred in primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
-In line with this, we estimated an average infection risk of between 1\% and 16\% when assuming the number of infectious attendees is proportional to the prevalence of TB in the South African population.
-Another modeling study further suggests a considerable benefit of infection control measures, \eg the rate of TB transmission in clinics could be reduced by about 50\% with compulsory mask wearing and natural ventilation through opening doors and windows\cite{McCreesh2021BMJGlobalHealth}.
-Our modeling results also suggest a notable reduction in the average risk of infection of more than a percentage point.
-Moreover, we estimated a considerable shift in the tail of the risk distribution as infection control measures significantly reduce high infection risks.
+Further studies suggest that compulsory face mask wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+Our findings also suggest that the risk of infection is lower with mask wearing, although the risk reduction estimated in our study is less striking.
+The reason is that the effect of mask wearing is dominated by the impact of very high ventilation rates, which determine both the diffusion and removal of infectious quanta in the indoor space.
+The air change rates in the clinic typically ranged from 10 to 40 air changes per hour, which result in a quick diffusion and removal of infectious quanta from the indoor air.
+To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.48\,h$^{-1}$), Switzerland (0.71\,h$^{-1}$), and Tanzania (13.74\,h$^{-1}$)\cite{Banholzer2024PGPH}.
+Tanzania had a much lower risk of \emph{Mtb} transmission in schools than the other countries, which was largely due to better ventilation, and yet its mean air change rate was less than half the rate measured in our study.
+Thus, we note that our study results do \emph{not} indicate that masks are less effective as found in other studies, rather their effectiveness is overshadowed by excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 
@@
 Our spatiotemporal model can be viewed as a modified version of the Wells-Riley transmission model\cite{Riley1978AJE}, which assumes a well-mixed airspace.
-By contrast, our model allows the quanta concentration to vary both over time and space.
-The exposure to infectious quanta therefore depends on the proximity to infectious individuals because the density of infectious particles is initially higher near them\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
-Our modeled risk of TB infection was significantly associated with the number of close-contact encounters and the time spent in the clinic.
-This result is intuitive because there are only few infectious TB patients visiting the clinic, so that more time in the clinic as well as a higher number of close contacts makes exposure to high doses of infectious particles more likely.
-Previous findings already suggest that prolonged close contact may be required for transmission of respiratory infections\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
-However, proving close contact respiratory transmission is difficult and  empirical studies investigating the potential link between contact data and respiratory transmission are scant\cite{Voirin2015ICHE,Vanhems2013PONE}.
+By contrast, our model allows the quanta concentration to vary both over time and space, considering that density of infectious particles is initially higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
+The exposure to infectious quanta therefore depends on the proximity to infectious individuals, which aligns with previous findings suggesting that prolonged close contact may be required for transmission of respiratory infections\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
+In line with this, our modeled risk of infection was significantly associated with the number of close-contact encounters and the time spent in the clinic.
+This is because more close contacts and longer visits make exposure to high doses of infectious particles more likely.
+However, we note that this association is based on modeled results and it is difficult to definitively prove close contact respiratory transmission.
+Empirical studies investigating the potential link between contact data and respiratory transmission are so far scant\cite{Voirin2015ICHE,Vanhems2013PONE}.
 
 % more complex models and our relation
-We applied our model to TB, although it can easily be extended to model other respiratory infections.
-Airborne respiratory transmission depends on common environmental and patient-specific factors, mainly ventilation and patient movements.
-Other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can also influence the generation, diffusion, and removal of infectious quanta in the space\cite{Wang2021Science}.
-We discuss these factors in more depth in \supp Text~\zref{sec:depth-discussion} in \supp~Text~\zref{sec:depth-discussion}.
-One important factor not considered by our model is airflow, although it has been shown that poor ventilation settings can produce airflows where infectious particles are trapped in the indoor space rather than being removed\cite{Li2021BuildEnv}.
-Computational fluid dynamics (CFD) models can consider airflow when simulating the spatial spread of infectious particles\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
-However, CFD models are computationally expensive and primarily used in experimental settings and static environments, which makes their application to our dynamic setting with patient movements difficult.
-Moreover, our approach requires a large number of Monte Carlo simulations to reflect uncertainty in modeling parameters and assumptions, necessitating computational efficiency.
+Airborne respiratory transmission depends on common environmental and patient-specific factors.
+Ventilation conditions and patient movements were considered by our model.
+However, other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see \supp~Text~\zref{sec:depth-discussion} for a detailed discussion of these factors).
+For example, poor ventilation settings can impede the removal of infectious particles\cite{Li2021BuildEnv}, and computational fluid dynamics (CFD) models have been developed to consider airflow as an important factor influencing the spatial spread of airborne pathogens\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
+However, CFD models are computationally expensive and primarily used in experimental settings and static environments, which makes their application to dynamic settings such as ours difficult.
+Moreover, our modeling approach necessitates computational efficiency because a large number of Monte Carlo simulations is required to reflect uncertainty in important modeling parameters.
 
@@
 Our modeling study has several limitations.
-Tracking patient movements with video sensors is challenging and required considerable data processing.
+First, tracking patient movements with video sensors is challenging and required considerable data processing.
 Despite that, many patient movements were interrupted and could not be linked back together.
-Several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-Uncertainty in these modeling parameters was often large and could be reduced.
-For example, incorporating additional patient characteristics from the clinical database could indicate variation in infectiousness and susceptibility of patients visiting the clinic.
-It was not possible to validate our modeling results by empirical findings, which could be the subject of a long-term future study testing the practical implementation of our model.
+Therefore, some clinic attendees may have had longer visits and the risk of infection may have been underestimated.
+Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
+Uncertainty in these modeling parameters was often large and could be reduced in future work, for example, by incorporating additional information on patient characteristics from clinical records that are known to be associated with patient-specific infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
+Third, we could not validate our modeling results with empirical findings.
+A follow-up study could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be infected.
 
 % conclusions
-In conclusion, our modeling study using clinical, environment, and patient movement data showed that the risk of TB transmission in a primary care clinic in South Africa in October/November 2021 was considerable, albeit mitigated by infection control measures that were implemented during the COVID-19 pandemic.
-The number of close-contact encounters by clinical attendees was significantly associated with their modeled risk of infection.
-Our spatiotemporal model could thus be used to assess the potential effectiveness of interventions targeting a reduction in the number of close contacts.
+In conclusion, our modeling study using clinical, environmental, and patient movement data showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa in October/November 2021 was low and mitigated by infection control measures that were implemented during the COVID-19 pandemic.
+The modeled risk of infection was significantly associated with the number of close-contact encounters by clinic attendees during their visit.
+Our spatiotemporal model could thus be used to assess the impact of interventions targeting patient flows that aim to reduce the number of close-contact encounters.
 

### 2024-04-10 doc/paper/supplements.tex
Updates from Overleaf

--- 2024-03-04 doc/paper/supplements.tex
@@
 
-\newcommand{\supp}{SI Appendix}
-
-\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
-Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}
+\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
+Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}}
 
 % author list
-\author[1$\ddag$]{Nicolas Banholzer}
+\author[1]{Nicolas Banholzer}
 \author[2]{Keren Middelkoop}
 \author[2]{Juane Leukes}
+\author[3]{Ernest Weingartner}
+\author[1]{Remo Schmutz}
 \author[1]{Kathrin Zürcher}
+\author[1]{Matthias Egger}
 \author[2]{Robin Wood}
-\author[1]{Matthias Egger}
 \author[1*]{Lukas Fenner}
@@
 \affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}
+\affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
 
 \affil[*]{Corresponding author: lukas.fenner@ispm.unibe.ch }
-
-\affil[$\ddag$]{These authors contributed equally to this work.}
 
@@
 
-\section{Study setting and data}\label{sec:setting-and-data}
+\section{Processing of tracking data}\label{sec:setting-and-data}
 
-\subsection{Building}
+Tracking data required processing before it could be fully used for modeling.
+The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database.
+Visitors of the clinic most likely contributed multiple tracks for several reasons:
+\begin{itemize}
+    \item A person moved outside the range of the video sensors and then re-entered. 
+    \item An individual may temporarily not have been recognized as a person by the tracking software because the person bowed downwards.
+    \item A person may have been temporarily hidden behind another person.
+    \item Moving objects such as opening doors may have been recognized as a person, introducing noise.
+\end{itemize}
+We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person.
+In the first step, we filtered noise (\ie IDs of less than 10\,s and a total moving distance of less than 2\,m), and then automatically matched the remaining tracks that were less than 10\,s and 3\,m apart while moving or less than 60\,s and 0.5\,m apart while sitting.
+In the second step, we build on the automatically matched tracking data and, through manual inspection, linked tracks that were further apart but still most likely of the same person.
+To facilitate the manual linkage, we developed a Shiny application and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) that assisted the students who were performing this linkage.
+
+The Shiny app is shown in \Cref{fig:shiny-app}.
+In the center panel, a floor plan of the clinic is shown in (x,y)-space.
+Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
+Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track.
+The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey.
+
+Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link.
+Possible links are sorted in ascending order by the time difference.
+
+In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information: the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages.
+It then shows the current Pid with shortcuts to move to the next or previous Pid, or the next or previous Pid at the entrance.
+Linkages can be done either forward or backward in time.
+The duration of the current Pid (time in the clinic) is shown together with the number of linked raw tracks of which the current Pid consists.
+The sliding bars allow filtering of possible links according to difference in time, distance, or height, with pre-set filters above to quickly switch between common thresholds.
+The first dropdown menu selects the Oid for which alternative matches should be shown in the floor plan.
+The idea of this feature is to prevent false linkages when there are more plausible alternatives.
+A linkage is made by selecting the Oid in the "Link with IDs" dropdown menu and clicking "Link IDs".
+After that, the app updates the floor plan, table, and information in the panel.
+If an erroneous link was made, it can be revoked by clicking "Unlink last ID" if the last link in time should be revoked or by clicking "Unlink first ID" if the first link in time should be revoked.
+If no further linkages could be made, the track can be finished by entering a label (see below).
+The manually linked dataset is automatically saved to the file shown in the bottom of the left panel.
+
+We created labels to denote finished tracks: 
+\begin{itemize}
+    \item \textbf{Clean}: A clean track throughout the clinic with known entrance and exit.
+    \item \textbf{Staff}: IDs with moving in- and out of the staff room.
+    \item \textbf{Noise}: A short track ($\leq$1\,min) without entrance or exit.
+    \item \textbf{Lost enter}: A long track (>1\,min) without entrance.
+    \item \textbf{Lost exit}: A long track (>1\,min) without exit.
+    \item \textbf{Lost both}: A long track (>1\,min) without entrance and without exit.
+\end{itemize}
 
@@
     \centering
-    \includegraphics{doc/paper/floor_plan_2021.pdf}
-    \caption{Floor plan showing the defined areas inside the clinic and the placement of the video sensors and CO$_2$ devices.}
-    \label{fig:floor-plan}
+    \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
+    \caption{Screenshot of the Shiny app used for manual linking tracking IDs that probably belong together.}
+    \label{fig:shiny-app}
 \end{figure}
 
-\clearpage
-
-\subsection{Tracking data}
+After manual processing of the tracking data, we arrived at 3,319 tracks.
+We then filtered noise and short tracks <5\,min to obtain 1,563 used for analysis: 209 (13\%) "Clean", 125 (8\%) "Staff", 213 (14\%) "Lost enter", 246 (16\%) "Lost exit", and 770 (49\%) "Lost both".
+A sample of our tracks used for analysis, excluding staff, are shown in \Cref{fig:tracking-examples}.
 
@@
     \centering
-    \includegraphics{illustrations/tracking-examples.pdf}
+    \includegraphics{results/data/example-patient-tracks.png}
     \caption{Sample tracks from our video sensors of this study.}
@@
 
-\subsection{Preprocessing of tracking data}
-
-Tracking data required preprocessing before it could be fully used for spatiotemporal modeling.
-The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database.
-Clinical attendees most likely contributed multiple tracks for several reasons:
-\begin{itemize}
-    \item A person moved outside the range of the video sensors and then re-entered. 
-    \item An individual may temporarily not have been recognized as a person by the tracking software, \eg because the person bowed.
-    \item A person may have been temporarily hidden behind another person.
-    \item Moving objects such as opening doors may have been recognized as a person, adding a few noisy tracks into the data.
-\end{itemize}
-We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person.
-In the first step, we filtered noise (\ie IDs of less than 10s and a total moving distance of less than 2m) and automatically matched IDs that were less than 10s and 3m apart while moving (\ie IDs anywhere in the clinic) or less than 60s and 0.5m apart while sitting (\ie two IDs ending/starting in a seating area of the clinic).
-In the second step, we build on the automatically matched data and performed an additional manual linkage.
-The aim was to link IDs that were further apart but still most likely belonged together.
-Through manual inspection, we could better assess whether a link resulted in a movement that made sense and preclude better alternatives.
-A Shiny application and a matching manual/wiki (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) was developed to assist the students that performed the manual linkage.
-
-The Shiny app is shown in \Cref{fig:shiny-app}.
-In the right panel, a floor plan of the clinic is shown in (x,y)-space.
-Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
-Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track.
-The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey.
-Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link.
-Possible links are sorted in ascending order by the time difference.
-
-In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information, \ie the name of the loaded file at the top and the path to the file where the updated data is automatically saved at the bottom, the date of the tracking, the initial number of IDs before any linkages and current number of IDs after linkages, and the number of links made.
-We created labels to denote finished tracks as "Ended" if the ID was continuously tracked throughout the clinic or "Lost" if the ID was somewhere lost in the clinic.
-The number of labels are shown below the number of IDs.
-Below the descriptive information are the tools related to performing linkages.
-First, the currently selected ID (black track) is shown, together with action buttons allowing students to quickly move to the first unfinished ID, the next or previous ID, or the previous or next ID starting at the entrance.
-Second, students could decide whether they want to match forward or backward in time.
-Third, the duration and number of links are shown for the current ID, which for example can alert students about IDs spending an unusually long time in the clinic.
-Fourth, the sliding bars allow students to filter possible links below a certain difference in time, distance, or height.
-Pre-set filters are added atop to quickly switch between common thresholds.
-Fifth, there is a drop-down menu to select IDs for which possible alternative links should be shown (light grey tracks).
-If there is a more plausible alternative, students could select the corresponding ID and check whether it is indeed the better match.
-Sixth, another drop-down menu below selects the ID which should be linked to the current patient ID (colored tracks) and the action "Link IDs" performs the link and updates the floor plan, table, and all other related information.
-If the students realized that the link was erroneous, they could unlink the last/first ID.
-Finally, once no further links could be made, either because the ID left the clinic or was lost, students could finish the track and label it accordingly.
-Ended tracks are removed from the set of possible links whereas lost tracks remain in the set.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
-    \caption{Screenshot of the Shiny app used for manual linking IDs that probably belong together.}
-    \label{fig:shiny-app}
-\end{figure}
-
-\clearpage
-
 \section{Spatiotemporal model}\label{sec:spattemp-model}
 
-\subsection{The Wells-Riley model}
+\subsection{The Wells-Riley modeling framework}
 
-We extend the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection.
-The model estimates the probability of airborne indoor transmission $P$ using the following equation:
+We build upon the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection.
+The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation:
 \begin{align}
-    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
+    P = \frac{D}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
 \end{align}
-where $C$ is the number of disease cases, $S$ is the number of susceptibles, $I$ is the number of infectious individuals in space, $p$ is the breathing rate per person, $q$ is the generation rate of infectious quanta, $t$ is the duration of exposure, and $Q$ is the outdoor air supply rate.
+where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 
@@
 First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and evenly in the indoor airspace.
-In other words, the exposure to infectious quanta is the same regardless of the locations of the infectious and susceptible individual.
+In other words, the exposure to infectious quanta is the same regardless of the location of the infectious and susceptible individual.
 Second, the Wells-Riley model assumes steady-state conditions, which means that the quanta concentration and the outdoor air supply rate are constant over time.
-In other words, the risk of infection for each susceptible depends only on the duration of exposure (the time they spent in the indoor space).
+In other words, the risk of infection for each susceptible only varies with the duration of exposure (the time they spent in the indoor space).
 
@@
 \end{align}
-where $f = \frac{CO^{\text{I}}-CO^{\text{O}}}{CO^{\text{A}}}$ is the fraction of indoor air that is exhaled breath, which is computed based on the CO$_2$ level in indoor air $CO^{\text{I}}$, outdoor air $CO^{\text{O}}$, and exhaled breath $CO^{\text{A}}$ (in parts per million [ppm]).
+where $f = \frac{CO^{\text{I}}-CO^{\text{O}}}{CO^{\text{A}}}$ is the fraction of indoor air that is exhaled breath, which is computed based on the CO$_2$ level in indoor air $CO^{\text{I}}$, outdoor air $CO^{\text{O}}$, and exhaled breath $CO^{\text{A}}$ (in parts per million [ppm], respectively).
 
-While the extension by Rudnick and Milton resolves the steady-state assumption, it still assumes a well-mixed airspace.
-Therefore, our aim is to extend the Wells-Riley model further to allow the risk of infection to vary both across time and airspace.
-We accomplish this by modeling the quanta concentration spatiotemporally.
-The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points.
-Formally, we estimate the cumulative risk of infection for susceptible individual $a$ as
+While the extension by Rudnick and Milton relaxes the steady-state assumption, it still assumes a well-mixed airspace.
+We relax this assumption by modeling the concentration of infectious quanta over space time.
+Then, we compute the cumulative risk of infection for each susceptible individual over space and time as
 \begin{equation}\label{eq:spattemp-P}
-    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
+    P = \sum_s \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
 \end{equation}
-where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether $a$ was at location $s$ at time $t$, and $p$ is the breathing rate per person (in $m^3/s$).
-The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta:
+where $N$ is the quanta concentration in cell airspace $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in cell $c$ at time $t$, and $p$ is the breathing rate.
+
+Modeling the spatiotemporal quanta concentration involves three processes
 \begin{itemize}
-    \item[\ref{sec:quanta-generation}] Quanta generation: Infectious individuals generate quanta at their location in the indoor space. 
-    \item[\ref{sec:quanta-diffusion}] Quanta diffusion: Over time, the infectious quanta diffuse and we approach a well-mixed airspace. 
-    \item[\ref{sec:quanta-removal}] Quanta removal: Infectious quanta is removed from the air through (a)~outdoor air exchange, (b)~viral inactivation, and (c)~gravitational settling. 
+    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Quanta is generated at the infectious individual's location. 
+    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Quanta diffuses until we approach a well-mixed airspace. 
+    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Quanta is removed from the air or becomes inactive (not infectious). 
 \end{itemize}
-We describe each process in more detail in the following. While we apply our model to \emph{Mtb}, we note that it can be applied to airborne respiratory infections in general.
+We describe each process in more detail in the following, but first we introduce some notation.
 
@@
 
-We divide the airspace in a 2-dimensional grid, irrespective of the height of the airspace.
-The cell denotes the unit airspace with $s = 1, \dots, S$, where $(x_s, y_s)$ are the x and y coordinate of cell $s$.
-In some cases, we may use the (x,y)-notation in place of the cell ID $s$.
-The quanta concentration will be the same within each grid cell $s$, thus the grid provides a discrete approximation to the spatially varying quanta concentration.
-The number of grid cells $S = n_x \cdot n_y$ corresponds to the size of the matrix and determines the quality of the approximation as well as the model run time.
-More grid cells (smaller grid cells) provide a more granular picture of the spatial quanta concentration, at the expense of higher computation time.
+We divide the area of the airspace in a 2-dimensional grid of cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$.
+The grid provides a discrete approximation to the continuously varying quanta concentration across the space.
+The number of grid cells $C = n_x \cdot n_y$ determines the spatial resolution and computation time, \ie more cells increase resolution at the expense of more computation time.
+We model the quanta concentration at time $t = 1, \dots T$ (in seconds).
 
@@
 \begin{itemize}
-    \item $N_{s,t}$: quanta concentration in airspace unit $s$ at time $t$
-    \item $I_{s,t}$: number of infectious individuals in airspace unit $s$ at time $t$
-    \item $q$: quanta generation rate (in quanta/s)
-    \item $D$: diffusion constant (in m$^2$/s)
-    \item $AER$: air change rate (in air changes/s)
-    \item $V$: volume of the airspace (in m$^2$)
-    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air
-    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air
-    \item $G$: average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$)
+    \item $N_{c,t}$: quanta concentration in airspace unit $c$ at time $t$
+    \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
+    \item $q$: quanta generation rate (quanta s$^{-1}$)
+    \item $D$: diffusion constant (m$^2$ s$^{-1}$)
+    \item $AER$: air change rate (air changes s$^{-1}$ / quanta s$^{-1}$)
+    \item $V$: volume of the airspace (m$^3$)
+    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air (ppm)
+    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air (ppm)
+    \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
     \item $n_t$: number of individuals in airspace at time $t$
-    \item $\lambda$: viral inactivation rate
+    \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
+    \item $k$: gravitational settling rate (quanta s$^{-1}$)
 \end{itemize}
@@
 
-The number of quanta generated by infectious individuals at time $t$ is 
+The number of quanta generated by infectious individuals at time $t$ in cell $c$ is 
 \begin{align}\label{eq:generation}
-    I_t \cdot q ~.
+    I_{c,t} \cdot q ~.
 \end{align}
-The initial spatial spread of the quanta over the grid is informed prior literature.
-The concentration of virus-laden aerosols is highest closest to the source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}, in line with findings showing that the risk of infection correlates with the distance to the infectious individual\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
-Although coughing and sneezing can transport aerosols further away from the individual, these activities are much less frequent than breathing.
-Therefore, prior work suggests that breathing may contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}.
-Overall, it seems plausible that the majority of infectious aerosols are generated near the infectious individual and then disperse over time.
-Since we do not observe the directional flow of the exhaled breadth, we assume that the generated quanta is evenly and radially distributed around the individual.
+For now, we assume that the initially generated quanta is confined to the cell where the infectious individuals are located, which is not an unreasonable assumption.
+The concentration of virus-laden aerosols is typically highest near the infectious source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}.
+Although coughing and sneezing can expel the aerosols further away from the individual, these activities are much less frequent than breathing, which may thus contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}.
 
@@
 
-Absent further knowledge on airflow, we assume that aerosols diffuse uniformly in the indoor air, over time approaching a well-mixed airspace.
-Initially, the quanta concentration will be highest at the location where it is generated and then disperse radially in (x,y)-direction towards locations where the concentration is lower.
+Absent measurements of airflow, we assume that aerosols diffuse uniformly in the indoor air, approaching a well-mixed airspace over time.
+We assume that the quanta disperse radially in (x,y)-direction towards locations where the concentration is lower.
 The diffusion equation is
 \begin{align}\label{eq:diffusion}
-    \frac{\delta N_{s,t}}{\delta t} = D \Delta N_{s,t},
+    \frac{\delta N_{c,t}}{\delta t} = D \Delta N_{c,t},
 \end{align}
-where $\Delta$ is the Laplace operator and $D$ is the diffusion constant (in m$^2$/s).
-We solve this equation with a standard second-order finite difference model using the Euler method with a 5-point stencil.
+where $\Delta$ is the Laplace operator and $D$ is the diffusion constant. We solve this equation with a standard second-order finite difference model using the Euler method with a 5-point stencil.
 
-The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)\cite{Cheng2011EnvSciTech}
+The diffusion constant $D$ is unknown.
+We approximate it using an empirical relationship between the eddy diffusion coefficient $K$ ($K \approx D$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
 \begin{align}
@@
 \end{align}
-where $V$ is the volume of the airspace (in m$^3$). We assume that $D \approx K$.
+where $V$ is the volume of the airspace (in m$^3$). While this is only an approximation, it nicely relates the air change rate to quanta diffusion.
 
-The $AER$ can be estimated from indoor CO$_2$ levels and room occupancy. Under steady-state conditions (CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
+The $AER$ is not directly observed, but can be estimated from indoor CO$_2$ levels, room occupancy, and room volume.
+Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
 \begin{align}
@@
 \end{align}
-where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$), $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air (both in parts per million [ppm])\cite{Batterman2017IJERPH}.
-In a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration.
-Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
+where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person, $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air\cite{Batterman2017IJERPH}.
+However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration.
+Therefore, we determine $AER$ using a transient mass balance model\cite{Batterman2017IJERPH}
 \begin{align}
@@
 where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$.
-We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) as implemented in the statistical software R\cite{Byrd1995SIAM}, a quasi-Newton method.
+We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS), a quasi-Newton method, as implemented in the statistical software R\cite{Byrd1995SIAM}.
 The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
-We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$\cite{Batterman2017IJERPH}.
-The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$.
+We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [300,600]$\cite{Batterman2017IJERPH}.
+We can compute $AER$ for different time windows, hence $AER_t$.
 
@@
 
-Infectious quanta may be removed from the air through outdoor air exchange, viral inactivation, or gravitational settling.
-We will ignore gravitational settling.
-Vuorinen (2020)\cite{Vuorinen2020SafSci} notes that particles smaller than 20\,$\mu$m rapidly become airborne.
-This is far above the upper end of the particle size distribution of most respiratory infections, including \emph{Mtb}, which is carried in particles typically smaller than 4.7\,$\mu$m\cite{Fennelly2020Lancet}.
-The removal rate is thus the same of the outdoor air exchange rate $AER_t$ and the viral inactivation rate $\lambda$, \ie
+Infectious quanta may be removed from the air through outdoor air exchange, bacterial inactivation, or gravitational settling.
+We will ignore gravitational settling because \emph{Mtb} is carried in small particles between 2\,$\mu$m and 5\,$\mu$m\cite{Fennelly2020Lancet}, which become airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+The removal rate is thus the sum of the outdoor air exchange rate $AER_t$ and the bacterial inactivation rate $\lambda$ for \emph{Mtb}, \ie
 \begin{align}\label{eq:removal}
@@
 \end{align}
-Note that $\lambda$ depends on the pathogen and is generally difficult to estimate and likely also depends on environmental factors, particularly relative humidity [Reference].
+
+\subsection{Spatiotemporal quanta concentration}
 
@@
 We illustrate our modeling approach with an example.
-Consider a room with a volume $V$=150m$^3$ (length [L] = 10m, breadth [B] = 5m, height [H] = 3m).
-We divide the room into a grid of $S = 40 \cdot 20 = 800$ cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+Consider a room with a volume $V$=150m$^3$ (length = 10m, breadth = 5m, height = 3m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$ cells with an area of 0.25m$^2$.
 We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
-We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(x_s = 20\pm1, y_s = 10\pm1)$, \ie $N_{s,t} = (I \cdot q) / 9 ~~ \forall s \in (x_s = 20\pm1, y_s = 10\pm1)$.
-We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of 1 and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$.
+We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle, \ie cell (20,10), which means that the initially generated quanta is $(x_c = 20\pm1, y_c = 10\pm1)$, \ie $N_{c,t} = \frac{1}{9} \cdot I \cdot q$ in cells $c \in (x_s = 20\pm1, y_s = 10\pm1)$, and zero in all other cells.
+We further assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$, a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$ (removal rate of 1\,quanta h$^{-1}$ and corresponding to a diffusion coefficient of 0.007\,m$^2$ s$^{-1}$), and a bacterial inactivation rate of $\lambda = 0.5$\,quanta h$^{-1}$.
 
-\Cref{fig:toy-example} shows the quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am).
-The quanta is generated at the location of the infectious individual.
-The quanta concentration remains high around this location until the individual ceases to generate more quanta at 9am.
-The quanta diffuses relatively quickly and at 10min past 9am the airspace is well-mixed with quanta.
-Over time, the removal through outdoor air exchange and viral inactivation brings the concentration in the room back towards 0.
+\Cref{fig:toy-example} shows the quanta concentration at different time points.
+The quanta concentration increases over time and is higher near the infectious individual.
+After the individual has left the space, the quanta diffuses fairly quickly and to approach a well-mixed already at 9:05am.
+Over time, the removal through outdoor air exchange and bacterial inactivation brings the concentration in the room back towards zero.
 
@@
     \centering
-    \includegraphics{../../tests/stm_v2-toy_example.pdf}
-    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10m $\times$ B=5m $\times$ H=3m.}
+    \includegraphics{tests/stm_v2-toy_example.png}
+    \caption{Spatiotemporal quanta concentration between 8am and 10am following the generation of quanta by an infectious individual between 8am and 9am in the middle of the room.}
     \label{fig:toy-example}
@@
 
-\section{Estimation of infection risk in the primary care clinic}\label{sec:estimation}
+\section{Simulation approach}\label{sec:estimation}
 
-\subsection{Building and setup}\label{prep:building}
+\subsection{Setup}\label{prep:building}
 
-A schematic view of the clinic is shown in \Cref{fig:clinic}.
-Clinical attendees were tracked when entering the clinic, in the waiting room, the corridor, and the TB room.
-If they moved into one of the care rooms along the corridor or into other parts of the clinic they were only tracked once they came back.
-We divide the clinic into three spaces: the waiting room (L=10.55m $\times$ B=5.50m $\times$ H=3.00m), the corridor (L=12.45m $\times$ B=2.20m $\times$ H=2.50m), and the TB room (L=4.75m $\times$ B=3.50m $\times$ H=3.00m).
-We separated the waiting room and the corridor because we observed small differences in the time-varying CO$_2$ levels, but we acknowledge that the two areas could also be evaluated jointly since the doors to the corridor were often most of the time.
-Moreover, the corridor and waiting room are not completely enclosed, but connected to the entrance, exit, and other parts of the clinic.
-Thus, we simplify our modeling analysis when we limit quanta diffusion within these rooms and also note that the calculated air change rate is affected by the separation of the waiting room and corridor (\ie the volume and room occupancy would otherwise be different).
+We divide the clinic into three spaces: the waiting room (length = 10.55m $\times$ breadth = 5.50m $\times$ height = 3.00m), the corridor (length = 12.45m $\times$ breadth = 2.20m $\times$ height = 2.50m), and the TB room (length = 4.75m $\times$ breadth = 3.50m $\times$ height = 3.00m).
+We separate the waiting room and the corridor because separate rectangular rooms simplify modeling and because we observe small differences in the time-varying CO$_2$ levels, suggesting these rooms are somewhat different.
+However, we note that the doors to the corridor were often most of the time, so that the corridor and waiting room were not completely enclosed.
+As such, it is possible that aerosolized \emph{Mtb} (quanta) spreads from the waiting room to the corridor, and vice versa, but this is not considered in our modeling approach.
+We rasterize each room into a grid of cubic cells with a squared area of 0.25m$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+Since the rooms have different height, the quanta concentration in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
 
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{doc/clinic-schematic-annotated-view.pdf}
-    \caption{Schematic view of the part of the clinic where clinical attendees were tracked.}
-    \label{fig:clinic}
-\end{figure}
-
-As in our illustrative example (\Cref{sec:example}), we divide each room into a grid of cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
-Since the rooms have different height, the quanta concentration (quanta/m$^3$) in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
-
-We divide the clinical day into a morning (7am to 12am) and afternoon period (12am to 5pm) and estimate the outdoor air exchange rate separately for each period.
-Thereby, we allow the $AER$ to vary by daytime, considering potential differences in natural ventilation habits in the morning and afternoon.
-
-We update the quanta concentration every second at $t = 1, \dots, T$ (aligned with the frequency of the tracking data).
-We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{s,0} = 0 ~~ \forall s$.
+We model the spatiotemporal quanta concentration during clinic hours from 8am to 4pm and analyze the modeled concentration by daytime, \ie morning (8am to 12am) and afternoon (12am to 4pm).
+Correspondingly, we estimate the outdoor air exchange rate by daytime, thereby considering potential differences in ventilation patterns between the morning and afternoon.
+We update the quanta concentration every second at $t = 1, \dots, T$, which aligns with the frequency of the tracking data.
+We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{c,0} = 0 ~~ \forall c$.
 In other words, we assume that all quanta from the previous day has been removed from the air before the start of the following clinical day.
@@
 
-We have uncertainty in the data (clinical attendees that are infectious but undiagnosed) and modeling parameters (quanta generation and viral inactivation rate).
-We consider this uncertainty using Monte-Carlo simulation where uncertain parameters are modeled with prior distributions (see \Cref{sec:priors} below).
-In each run of the simulation, we perform the following steps:
+We use Monte-Carlo simulation to estimate the risk of infection, thereby considering uncertainty in several modeling parameters. Each simulation consists of the following steps:
 \begin{itemize}
-    \item[\textbf{1.}] Sample uncertain modeling parameters such as the quanta generation rate and the proportion of undiagnosed TB patients among clinical attendees.
-    \item[\textbf{2.}] Sample undiagnosed TB patients and combine with diagnosed TB patients, and then use the tracking data to determine the location and time of where and when quanta was generated by these infectious individuals.
-    \item[\textbf{3.}] For each room (waiting room, corridor, and passage), compute the quanta concentration in each cell $s$ and at time $t$ using \Cref{eq:spattemp-N} from the spatiotemporal model.
-    \item[\textbf{4.}] For each susceptible clinical attendee, compute the risk of infection from the cumulative exposure to quanta over airspace and time using \Cref{eq:spattemp-P}.
+    \item[1.] \textbf{Sample uncertain modeling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, the mask reduction rate, and the bacterial inactivation rate.
+    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, compute the spatiotemporal quanta concentration in each cell of the rooms over time.
+    \item[3.] \textbf{Estimate risk of infection}: Based on their spatiotemporal quanta exposure, estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
 \end{itemize}
 
-Most clinical attendees could be undiagnosed TB Patients and there is considerable uncertainty in some modeling parameters.
-As a consequence, we need a high number of Monte Carlo runs to reasonably estimate the risk of infection for each individual.
-Therefore, for each study day, we perform 10,000 Monte-Carlo simulations.
-If not stated otherwise, we summarize the results with the mean and 95\% credible interval.
+Most clinic attendees could be undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, and bacterial inactivation rate is large.
+To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean or median across these simulations.
 
-\subsection{Prior choices for modeling parameters}\label{sec:priors}
+\subsection{Modeling assumptions and prior distributions}\label{sec:priors}
 
@@
 
-% spatial spread of quanta generation
-
-An analysis of exhalation activities suggest a propagation distance (\ie the maximum distance from the source to the end of the exhaled air before dispersion) of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity.
-Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting particles (leakage of particles on the sides of the mask)\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}.
-Therefore, we assume that mask wearing confined the quanta generation only to the cell where the infectious individual is located.
-This corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.
-
-To assess the impact of mask wearing, we also model the scenario without mask wearing.
-In this scenario, we assume a propagation distance of 0.7m (the diagonal lengths of two cells), in line with the aforementioned prior findings.
-Since we cannot observe the direction of jet, we assume that all neighbouring cells are affected equally.
-That is, we divide the generated quanta by 9 (the cell where the individual is located and all first-neighbouring cells) and distribute it equally across the cells.
+We model the amount of quanta that is generated and how it diffuses.
+What we have not yet discussed is the initial spread of infectious quanta, referred to as the propagation distance, which is the maximum distance from the source to the end of the exhaled air before dispersion.
+An analysis of exhalation activities suggest a propagation distance of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity (\ie breathing, speaking, coughing).
+Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting (leakage of particles on the sides of the mask) particles\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}.
+Therefore, we assume that mask wearing confines the quanta generation only to the cell where the infectious individual is located.
+In our setup, this corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.
+In a hypothetical scenario, we estimate the risk of infection if mask wearing would not have been compulsory.
+Without masks, we assume double the propagation propagation distance, \ie 0.7m (the diagonal lengths of two cells)\cite{Tang2013PLoSOne}.
+Since we could not observe the flow direction of the initially generated quanta, we assume that all neighbouring cells are affected equally.
+That is, we divide the generated quanta equally among the nine affected cells, \ie the cell where the infectious person is located as well as all first-neighbouring cells.
 
@@
 
-% quanta generation
-
-Previous studies estimated $q$ for \emph{Mtb} or SARS-CoV-2 using the Wells-Riley model\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD}.
-Andrews (2014)\cite{Andrews2014JID} reported an estimate of 0.89 quanta/h (range from 0.44 to 5.69), close to the early estimate provided by Riley (1962)\cite{Riley1962ARRD} with 1.25 quanta/h.
-On the other end, Escombe (2008)\cite{Escombe2008PLoSMed} arrived at an estimate of 8.2 quanta/h, which was lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients.
-Nardell (1991)\cite{Nardell1991ARRD} also reported a higher estiamte of 13 quanta/h, but only for one index case.
-The wide range of estimates suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}.
-Therefore, we model $q$ with a Student-t distribution with 1 degree of freedom $t_1(\mu = 2, \sigma = 2.5)$ that is truncated at 0 (lower) and 200 (upper), as in Banholzer (X) [OUR PLOS GLOBAL].
-The Student-t distribution with only 1 degree of freedom considers the possibility of superspreaders that generate large amounts of quanta.
-The distribution function is shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.
+Previous studies estimated the quanta generation rate of \emph{Mtb} using the Wells-Riley model\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}.
+Andrews et al.\cite{Andrews2014JID} reported an estimate of 0.89\,(quanta h$^{-1}$) (range from 0.44 to 5.69), close to the early estimate provided by Riley et al.\cite{Riley1962ARRD} with 1.25\,quanta h$^{-1}$.
+On the other end, Escombe et al.\cite{Escombe2008PLoSMed} arrived at an average estimate of 8.2\,quanta h$^{-1}$, which, however, could be as high as 226\,quanta h$^{-1}$.
+Lower rates were observed in lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients\cite{Escombe2008PLoSMed}.
+Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM} also estimated a high average rate of 138\,quanta h$^{-1}$ in MDR TB patients.
+Nardell et al.\cite{Nardell1991ARRD} also reported a higher estimate of 13\,quanta h$^{-1}$, but only for a single index case.
+The wide range of estimates within and between studies suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}.
+To consider such variation, Mikszewski et al.\cite{Mikszewski2021GF} used the bacterial load distribution in sputum to estimate the quanta generation rate with the approach developed by Buonanno et al.\cite{Buonanno2020EI}.
+We have recently adopted this approach to derive quanta generation rate distributions for different activity levels in a school setting\cite{Banholzer2024PGPH}.
+Here, we follow the same approach to derive a quanta generation rate distribution for sitting and walking activities, assuming 80\% breathing and 20\% speaking during both activities.
+The resulting prior distributions are shown in \Cref{fig:quanta-distribution}.
+The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking.
+The activity is determined based on whether the distance between two patient tracks was more (walking) or less (sitting) than 0.25m.
 
@@
     \centering
-    \includegraphics{illustrations/prior-q.pdf}
-    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014)\cite{Andrews2014JID}, 1.25 by Riley (1962)\cite{Riley1962ARRD}, 8.2 by Escombe (2008)\cite{Escombe2008PLoSMed}, and 13 by Nardell (1991)\cite{Nardell1991ARRD}.}
+    \includegraphics{results/inputs/quanta-generation-rate.png}
+    \caption{Prior distribution (log scale) for the quanta generation rate (quanta h$^{-1}$) by activity level (sitting and walking), along with reported average estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}.}
     \label{fig:quanta-distribution}
@@
 
-% face mask reduction
+\subsubsection{Mask reduction rate}
 
 Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens,Leung2020NatMed} and should thus lower the generation of infectious quanta.
-The extent of the reduction is subject to uncertainty.
-A study with multidrug-resistant TB cases estimated a decrease of 56\% (95\%-CrI, 33\% to 70.5\%) when patients used surgical face masks\cite{Dharmadhikari2012AJRCCM}.
-This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic.
-Therefore, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1.
-The effective quanta generation rate is thus $q$ times the reduction due to mask wearing.
-Again, we model what the risk of infection would have been without mask wearing by removing the reduction term from the model.
+McCreesh et al.
+assumed a reduction by masks of 75\% (95\%-confidence interval (56\% to 85\%) based on the findings of Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM}.
+Based on that, we model the reduction by masks with $\mathrm{Beta}(24.9, 8.3)$ distribution.
+We assume that masks were correctly worn the whole time by all clinic attendees and staff during the COVID-19 pandemic.
 
-\subsubsection{Viral inactivation rate}
+\subsubsection{Bacterial inactivation rate}
 
-% viral inactivation rate
-
-The viral inactivation rate for \emph{Mtb} has not been precisely estimated.
+The bacterial inactivation rate for \emph{Mtb} has not been precisely estimated.
 Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6h ($\lambda$ = 0.12).
 By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5min ($\lambda$ = 8.3).
-The difference may largely be explained by variation in relative humidity, \ie Loudon (1969) studied survival during a relative humidity of about 50\% whereas Lever (2000) studied survival during a relative humidity of about 70\%.
-Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) seems more representative of the environmental conditions in our study.
+The difference could be explained by differences in humidity levels during the study\cite{Lever2000LettersAppliedMicrobio}: Loudon (1969) studied survival during a relative humidity of about 50\% whereas Lever (2000) studied survival during a relative humidity of about 70\%.
+Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) seems more representative of the our study conditions.
 Moreover, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5 hours ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}.
 In contrast to that, Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30min ($\lambda$ = 3.2), which would be closer to Lever (2000).
-We choose a middle ground and, therefore, use a Lognormal($\mu$ = $\log$ 1, $\sigma$ = 1) with a median viral inactivation rate of 1.
-The distribution function is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
+Considering the variation in these estimates, we model $\lambda$ with a $\mathrm{Lognormal}(\mu = \log 1, \sigma = 1)$ distribution, which has a median rate of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$).
+The distribution is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
 
@@
     \centering
-    \includegraphics{illustrations/prior-lambda.pdf}
-    \caption{Prior distribution function for the viral inactivation rate for \emph{Mtb tuberculosis} along with reported estimates in the literature: 0.12 by Loudon (1969)\cite{Loudon1969AMRRD}, 0.46 by Gannon (2007) for \emph{Mtb bovis}\cite{Gannon2007ResVetSci}, 3.2 by Klein (2014)\cite{Klein2014IJMyco}, and 8.3 by Lever (2000)\cite{Lever2000LettersAppliedMicrobio}.}
+    \includegraphics{results/inputs/bacterial-inactivation-rate.png}
+    \caption{Prior distribution (log scale) for the bacterial inactivation rate (quanta h$^{-1}$) of \emph{Mtb} along with reported estimates from the literature\cite{Loudon1969AMRRD,Gannon2007ResVetSci,Klein2014IJ,Lever2000LettersAppliedMicrobioMyco}.}
     \label{fig:lambda-distribution}
@@
 
-% proportion of undiagnosed TB patients
+Undiagnosed (subclinical) TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats), or the diagnosis could not be made at first presentation based on existing tests\cite{Patterson2023Preprint}.
+Previous studies suggest that the number of undiagnosed (subclinical) TB patients is roughly similar to the number of diagnosed TB patients\cite{Berhanu2023CID,Moyo2022LancetID}.
+Therefore, we model the number of undiagnosed TB patients with a Multinomial distribution based on the counts of the daily number of diagnosed TB patients during our study period from October to November 2021 (\Cref{fig:undiagnosed-distribution}.
+We assume that all clinic attendees can be potential undiagnosed TB patients and sample among them with equal probability.
+We further assume that undiagnosed and diagnosed TB patients have the same quanta generation rate.
 
-In addition to the diagnosed, known TB patients, there may also be some undiagnosed, unknown TB patients among the clinical attendees.
-These unmasked TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats).
-A systematic review concluded that half of TB cases among people living with HIV that are on ART may be missed by the symptom screen\cite{Hamada2018LancetHIV}.
-Similarly, a study universally testing clinical attendees for TB that belong to a risk group (living with HIV, self-reported close contact with TB patients, prior history of TB) finds that 55\% of the attendees without a positive symptom screen were tested positive for TB\cite{Berhanu2023CID}.
-We interpret these findings such that the number of masked TB patients is roughly equal to the number of unmasked TB patients.
-On average, there were X masked TB patients per day attending the clinic.
-Therefore, we model the daily number of unmasked TB patients with a Poisson($\mu$ = X) distribution.
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/inputs/undiagnosed-tb-patients.png}
+    \caption{Counts (\%) of the daily number of diagnosed TB patients visiting the clinic according to clinical records.
+The number of diagnosed patients is modeled with a Multinomial distribution based on these counts.}
+    \label{fig:undiagnosed-distribution}
+\end{figure}
 
-% based on incidence
+\subsubsection{Breathing rate}
+
+Adam et al.\cite{Adams1993} reports breathing (inhalation) rates for different activity levels and by sex.
+We use the average breathing rate for two activity levels: sitting with $p = 0.51$\,m$^3$ h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$ h$^{-1}$.
+Again, the inhalation rate for each susceptible clinic attendee is determined based on whether the distance between two tracks was more (walking) or less (sitting) than 0.25m.
 
@@
 The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
-Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~properties of the specific pathogenic, and the (3)~infectiousness of infected and susceptibility of uninfected TB patients.
+Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~physicochemical properties of the pathogen, and the (3)~infectiousness of the source and susceptibility of the host.
 
@@
 
-Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate outdoor air supply.
+Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate quanta diffusion and removal.
 The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
 This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}.
-However, the radial diffusion of quanta assumed by our model may not reflect the actual airflow at the primary care clinic.
+In our spatiotemporal model, we assume a radial diffusion of quanta, which may not reflect the actual airflow at the primary care clinic.
 The spatial dispersion of infectious particles is often simulated with computational fluid dynamics (CFD) models\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv}.
 CFD models can simulate a variety of diffusion patterns, which help researchers understanding the influence of airflow.
-For example, considering airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered and it may be difficult to assess in a naturally ventilated space, such as our primary care clinic, where it depends on the outdoor wind direction and which doors and windows are open.
+For example, measuring airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh outdoor air\cite{Li2021BuildEnv}.
+However, airflow is rarely considered, especially when it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
 
@@
 
-Survival and dispersion of aerosols are also determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
-Currently, these properties are only indirectly incorporated in our model by modeling the quanta generation rate as a stochastic parameter.
-It is also challenging to consider these properties because environmental factors, in particular temperature and humidity, can affect particle generation and survival, but the effects are often unclear and depending on the pathogen\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
-Survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
-On the one hand, a comparison between two studies suggested that a higher survival rate of airborne \emph{Mtb} may be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}.
-On the other hand, airborne \emph{Mtb} was more likely to be detected in healthcare facilitates with larger relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
+Survival and dispersion of aerosols are further determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
+Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modeling framework via the quanta generation rate parameter\cite{Riley1978AJE,Rudnick2003IndoorAir}.
+It is challenging to consider these properties specifically, especially because they can be modified by environmental factors with often unclear, pathogen-specific effects\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
+For example, survival of bacteria in aerosols decreases with temperatures $>$24$^{\circ}$C, but the effects of humidity are unclear\cite{Tang2009Interface}.
+On the one hand, a comparison between two studies suggested that longer airborne survival of \emph{Mtb} could be attributed to lower relative humidity\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio}.
+On the other hand, airborne \emph{Mtb} was more likely detected in healthcare facilitates with higher relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
 

### 2024-04-18 analysis/environmental-data.Rmd
Modeling fixes and visual updates
Update results presentation

@@
 library(tidyverse)
+library(tidybayes)
 library(lubridate)
@@
   labs(
-    y = expression("CO"[2] * " (parts per million, ppm)"),
-    title = "f"
+    y = expression("CO"[2] * " (parts per million, ppm)")
   ) +
@@
     legend.position = "bottom",
-    panel.spacing = unit(1, "lines"),
     legend.key.width = unit(0.275, "cm"),
@@
   pdf_file = "results/data/co2-levels-over-time.png",
-  w = 10, h = 6
+  w = 16, h = 10
 )
+
+cont_df %>%
+  median_qi(co2, .width = .5) %>%
+  mutate_if(is.numeric, round)
 ```
@@
     y = "Air changes per hour",
-    title = "g"
+    title = "d"
   ) +
@@
   pdf_file = "results/data/air-changes-per-hour.png",
-  w = 6, h = 6
+  w = 8, h = 6
 )

### 2024-04-18 analysis/patient-tracking.Rmd
Update results presentation

@@
   melt(c("date", "patient_id")) %>%
+  filter(variable != "Clinic") %>%
   mutate(variable = factor(
     variable,
-    levels = c("Clinic", "Waiting room", "Corridor", "TB room")
+    levels = c("Waiting room", "Corridor", "TB room")
   )) %>%
@@
   geom_boxplot() +
-  scale_y_sqrt(breaks = c(0, 1, 10, 30, 60, 120, 180)) +
+  scale_y_sqrt(breaks = c(0, 10, 30, 60, 120, 180)) +
   scale_fill_manual(values = wes_palette("Royal2")) +
@@
     axis.title.x = element_blank(),
-    axis.text.x = element_text(angle = 45, hjust = 1),
     plot.title.position = "plot"
@@
   time_in_clin_pl,
-  pdf_file = "results/data/time-in-clinic.png", w = 5.3, h = 6
+  pdf_file = "results/data/time-in-clinic.png", w = 8, h = 6
 )
@@
   ungroup() %>%
+  dplyr::select(-Clinic) %>%
   melt(c("date", "time")) %>%
@@
 time_pl <- clinic_occupancy %>%
-  filter(variable == "Clinic") %>%
-  ggplot(aes(x = time, y = value, color = date)) +
-  geom_step() +
+  filter(variable != "Clinic") %>%
+  group_by(time, variable) %>%
+  summarise(
+    mv = mean(value),
+    sv = sd(value)
+  ) %>%
+  ungroup() %>%
+  mutate(
+    lv = mv - sv,
+    uv = mv + sv
+  ) %>%
+  ggplot(aes(x = time)) +
+  geom_line(
+    aes(y = mv, color = variable)
+  ) +
+  geom_ribbon(
+    aes(ymin = lv, ymax = uv, fill = variable),
+    alpha = 0.4
+  ) +
   # ggh4x::facet_wrap2(~variable, axes = "x", remove_labels = "all") +
@@
   scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
-  scale_color_manual(values = wes_palette("Moonrise3")) +
+  scale_fill_manual(values = wes_palette("Royal2")) +
+  scale_color_manual(values = wes_palette("Royal2")) +
   labs(

### 2024-04-18 analysis/qctr.Rmd
Update results presentation

@@
     labs(
-      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10e"^-3 * ")"),
-      title = "a"
+      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10e"^-3 * ")")
     ) +
@@
 
-mean_quanta_pl
+save_plot(
+  mean_quanta_pl,
+  "results/modeling/mean-quanta-concentration.png",
+  w = 16,
+  h = 8
+)
 ```
@@
       ),
-      P_col = factor(P_col, levels = c("Low", "Medium", "High")),
-      P_ann = 1 - (1 - P_mean)^12
+      P_col = factor(P_col, levels = c("Low", "Medium", "High"))
     )
@@
 
-main_tr <- main_tr %>% unnest(value)
-
-main_tr %>%
-  group_by(P_col) %>%
-  summarise(n = n()) %>%
-  ungroup() %>%
-  mutate(prop = round(n / sum(n) * 100, 0))
+main_tr <- main_tr %>%
+  unnest(value) %>%
+  dplyr::select(-files)
 
@@
       mean_P = mean(P_mean),
-      lower_P = quantile(P_mean, .025),
-      upper_P = quantile(P_mean, .975),
-      mean_P_ann = mean(P_ann),
-      lower_P_ann = quantile(P_ann, .025),
-      upper_P_ann = quantile(P_ann, .975)
+      lower_P = quantile(P_mean, .2),
+      upper_P = quantile(P_mean, .8)
     ) %>%
     mutate(
-      across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 3)),
+      across(c(mean_P, lower_P, upper_P), ~ round_k(100 * .x, 2)),
       lab = paste0(
         mean_P, "%",
-        " (95%-CrI ", lower_P, "% – ", upper_P, "%)"
-      ),
-      across(c(mean_P_ann, lower_P_ann, upper_P_ann), ~ round_k(100 * .x, 2)),
-      lab_ann = paste0(
-        mean_P_ann, "%",
-        " (95%-CrI ", lower_P_ann, "% – ", upper_P_ann, "%)"
+        " (80%-CrI ", lower_P, "% – ", upper_P, "%)"
       )
@@
 
-main_annon$lab
-main_annon$lab_ann
-
-# plot
-roi_pl <- ggplot(
-  data = main_tr,
-  mapping = aes(x = P_mean)
-) +
-  geom_histogram(
-    mapping = aes(fill = P_col),
-    binwidth = .0005,
-    boundary = 1
-  ) +
-  geom_text(
-    data = main_annon,
-    mapping = aes(x = Inf, y = Inf, label = lab),
-    size = 8 / cm(1),
-    hjust = 1, vjust = 2
-  ) +
-  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
-  coord_cartesian(expand = FALSE) +
-  scale_x_continuous(
-    labels = function(x) x * 100,
-    breaks = c(0, 0.001, 0.005, 0.01, 0.015, 0.02),
-    limits = c(0, 0.02)
-  ) +
-  scale_y_sqrt(
-    labels = scales::comma,
-    breaks = c(0, 25, 100, 250, 500, 1000)
-  ) +
-  labs(
-    x = "Risk of infection (%)",
-    y = "Number of people",
-    title = "b"
-  ) +
-  theme_custom() +
-  theme(
-    legend.position = c(0.7, .3),
-    legend.title = element_blank(),
-    legend.key.height = unit(.3, "cm"),
-    legend.key.width = unit(.2, "cm"),
-    legend.direction = "horizontal",
-    panel.spacing = unit(2, "cm")
-  )
-
-roi_pl
+main_annon
 ```
@@
 main_tr <- main_tr %>%
-  mutate(P_unc = P75 - P25)
+  mutate(P_unc = P80 - P20)
 
@@
   labs(
-    x = "Uncertainty (interquartile range, pp)",
-    y = "Mean risk of infection (%)",
-    title = "c"
+    x = "Uncertainty (80%-CrI, pp)",
+    y = "Mean risk of infection (%)"
   ) +
@@
     plot.title = element_text(size = 10, face = 2),
+    plot.margin = margin(.5, .5, .5, .5, "cm")
   )
 
-main_focus_pl
+save_plot(
+  main_focus_pl,
+  "results/modeling/patient-risk-focus-plot.png",
+  w = 12,
+  h = 10
+)
 ```
@@
 
+# risk per hour in clinic
+assoc_df %>%
+  filter(tracking_end != "Possible HCW") %>%
+  mutate(
+    rph = 60 * P_mean / time_in_clinic,
+    rpy = 1 - (1 - rph)^12
+  ) %>%
+  mean_qi(rph, rpy, .width = .8, na.rm = TRUE) %>%
+  mutate_if(is.numeric, ~ round_k(100 * .x, 2)) %>%
+  unlist()
+
 # model
@@
   mutate(across(c(lower, mean, upper), ~ round_k(.x, 2)))
-
-amod_est_pl <- ggplot(amod_est, aes(y = variable, x = value)) +
-  geom_vline(aes(xintercept = 1), linetype = "dotted", color = "red") +
-  stat_interval(point_interval = "mean_qi") +
-  stat_summary(
-    geom = "point",
-    fun = mean,
-    size = 2,
-    shape = 21,
-    fill = "white"
-  ) +
-  labs(
-    x = "Odds ratio of the modeled risk of infection for a doubling in the predictor",
-    title = "d",
-    color = "CrI"
-  ) +
-  scale_y_discrete(
-    labels = c(
-      # "Air change rate",
-      "Time in clinic",
-      "No. close contacts",
-      "Time close contact"
-    )
-  ) +
-  scale_color_brewer(
-    labels = paste0(c(95, 80, 50), "%")
-  ) +
-  theme_custom() +
-  theme(
-    axis.title.y = element_blank(),
-    legend.position = "right",
-    legend.key.size = unit(.3, "cm")
-  )
-
-amod_est_pl
-```
-
-```{r main-figure}
-main_fig <- grid.arrange(
-  mean_quanta_pl,
-  grid.arrange(roi_pl, main_focus_pl, ncol = 2),
-  amod_est_pl,
-  ncol = 1,
-  heights = c(10, 6, 4)
-)
-
-ggsave(
-  main_fig,
-  filename = "results/modeling/main-figure.png",
-  width = 16 / cm(1),
-  height = 16 / cm(1)
-)
 ```
@@
 
-## Additional analyses
+## IPC comparison
 
@@
 )
-files_prev <- list.files("simulations/prevalence/",
-  full.names = TRUE, pattern = "trans-risk",
-  recursive = TRUE
-)
-files_susp <- list.files("simulations/suspected/",
+files_nmvent <- list.files("simulations/nomasksvent/",
   full.names = TRUE, pattern = "trans-risk",
@@
 vent_tr <- map_dfr(files_vent, average_trans_risk)
-prev_tr <- map_dfr(files_prev, average_trans_risk)
-susp_tr <- map_dfr(files_susp, average_trans_risk)
+nmvent_tr <- map_dfr(files_nmvent, average_trans_risk)
 
@@
 
-susp_tr %>%
+nmvent_tr %>%
   group_by(P_col) %>%
@@
   ) %>%
+  left_join(
+    nmvent_tr %>%
+      dplyr::select(patient_id, P_mean) %>%
+      rename(`No mask wearing and lower ventilation` = P_mean)
+  ) %>%
   mutate(
     mask_factor = `No mask wearing` / `Spatiotemporal model`,
-    vent_factor = `Lower ventilation` / `Spatiotemporal model`
+    vent_factor = `Lower ventilation` / `Spatiotemporal model`,
+    nmvent_factor = `No mask wearing and lower ventilation` / `Spatiotemporal model`
   ) %>%
-  mean_qi(mask_factor, vent_factor) %>%
+  mean_qi(mask_factor, vent_factor, nmvent_factor) %>%
   gather() %>%
-  head(6) %>%
-  mutate(value = round(as.numeric(value), 2))
+  head(9) %>%
+  mutate(value = round(as.numeric(value), 1))
 
@@
 tr_comp <- rbind(
-  nm_tr %>% mutate(scenario = "No mask wearing"),
-  vent_tr %>% mutate(scenario = "Lower ventilation"),
-  prev_tr %>% mutate(scenario = "Prevalence-based model"),
-  susp_tr %>% mutate(scenario = "Suspected-based model")
+  main_tr %>% mutate(scenario = "a   Mask use and high ventilation"),
+  nm_tr %>% mutate(scenario = "b   No mask use and high ventilation"),
+  vent_tr %>% mutate(scenario = "c   Mask use and normal ventilation"),
+  nmvent_tr %>% mutate(scenario = "d   No mask use and normal ventilation")
 ) %>%
-  mutate(
-    scenario = factor(scenario, levels = c(
-      "Spatiotemporal model",
-      "Temporal model",
-      "No mask wearing",
-      "Lower ventilation",
-      "Prevalence-based model",
-      "Suspected-based model"
-    ))
-  )
+  mutate(scenario = factor(scenario, levels = unique(scenario)))
 
@@
 vent_annon <- annotate_average_trans_risk(vent_tr)
-prev_annon <- annotate_average_trans_risk(prev_tr)
-susp_annon <- annotate_average_trans_risk(susp_tr)
+nmvent_annon <- annotate_average_trans_risk(nmvent_tr)
 
 annon_comp <- rbind(
-  nm_annon %>% mutate(scenario = "No mask wearing"),
-  vent_annon %>% mutate(scenario = "Lower ventilation"),
-  prev_annon %>% mutate(scenario = "Prevalence-based model"),
-  susp_annon %>% mutate(scenario = "Suspected-based model")
+  main_annon %>% mutate(scenario = "a   Mask use and high ventilation"),
+  nm_annon %>% mutate(scenario = "b   No mask use and high ventilation"),
+  vent_annon %>% mutate(scenario = "c   Mask use and normal ventilation"),
+  nmvent_annon %>% mutate(scenario = "d   No mask use and normal ventilation")
 ) %>%
-  mutate(
-    scenario = factor(scenario, levels = c(
-      "No mask wearing",
-      "Lower ventilation",
-      "Suspected-based model",
-      "Prevalence-based model"
-    ))
-  )
+  mutate(scenario = factor(scenario, levels = unique(scenario)))
 
@@
     breaks = c(0, 25, 100, 250, 500, 1000),
-    limits = c(0, 1000)
+    limits = c(0, NA)
   ) +
@@
     legend.key.width = unit(.4, "cm"),
-    panel.spacing = unit(2, "cm")
+    panel.spacing = unit(2, "cm"),
+    strip.text = element_text(hjust = 0)
   )
@@
 ```
+
+## Additional analysis
+
+```{r} 
+files_prev <- list.files("simulations/prevalence/",
+  full.names = TRUE, pattern = "trans-risk",
+  recursive = TRUE
+)
+files_susp <- list.files("simulations/suspected/",
+  full.names = TRUE, pattern = "trans-risk",
+  recursive = TRUE
+)
+
+prev_tr <- map_dfr(files_prev, average_trans_risk)
+susp_tr <- map_dfr(files_susp, average_trans_risk)
+
+tr_assum <- rbind(
+  prev_tr %>% mutate(scenario = "a   Prevalence-based model"),
+  susp_tr %>% mutate(scenario = "b   Suspected-based model")
+) %>%
+  mutate(scenario = factor(scenario, levels = unique(scenario)))
+
+prev_annon <- annotate_average_trans_risk(prev_tr)
+susp_annon <- annotate_average_trans_risk(susp_tr)
+
+annon_assum <- rbind(
+  prev_annon %>% mutate(scenario = "a   Prevalence-based model"),
+  susp_annon %>% mutate(scenario = "b   Suspected-based model")
+) %>%
+  mutate(scenario = factor(scenario, levels = unique(scenario)))
+
+# plot mean
+mean_roi_assum_pl <- ggplot(
+  tr_assum,
+  aes(x = P_mean)
+) +
+  geom_histogram(
+    mapping = aes(fill = P_col),
+    binwidth = .0005,
+    boundary = 1
+  ) +
+  geom_text(
+    data = annon_assum,
+    mapping = aes(x = Inf, y = Inf, label = lab),
+    size = 8 / cm(1),
+    hjust = 1, vjust = 2
+  ) +
+  facet_wrap(~scenario, ncol = 2) +
+  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
+  coord_cartesian(expand = FALSE) +
+  scale_x_continuous(
+    labels = function(x) x * 100,
+    breaks = c(0, 0.001, 0.005, 0.01, 0.015, 0.02),
+    limits = c(0, 0.02)
+  ) +
+  scale_y_sqrt(
+    labels = scales::comma,
+    breaks = c(0, 25, 100, 250, 500, 1000),
+    limits = c(0, NA)
+  ) +
+  labs(
+    x = "Risk of infection (%)",
+    y = "Number of clinical attendees"
+  ) +
+  theme_custom() +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    legend.key.height = unit(.6, "cm"),
+    legend.key.width = unit(.4, "cm"),
+    panel.spacing = unit(2, "cm"),
+    strip.text = element_text(hjust = 0)
+  )
+
+save_plot(
+  mean_roi_assum_pl,
+  "results/modeling/mean-roi-assumptions.png",
+  w = 16,
+  h = 7
+)
+```

### 2024-04-18 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}}
+\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium} tuberculosis transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
 % author list
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} was the leading cause of death from any single infectious disease before the COVID-19 pandemic.
-TB is strictly airborne and thus the transmission risk is higher in crowded indoor settings, especially healthcare facilities.
-Existing approaches to model the risk of airborne transmission typically assume that the airspace is well mixed, thereby not considering that closer proximity to infectious individuals may increase the risk of transmission.
+\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities.
+Existing approaches to modeling the risk of airborne transmission typically assume that the airspace is well mixed, not considering that proximity to infectious individuals may increase the risk of transmission.
+We aimed to develop a  spatiotemporal model of airborne \emph{Mtb} transmission using environmental, clinical, and patient movement data to estimate TB transmission during the COVID-19 pandemic in healthcare facilities and to evaluate the impact of infection control measures.
 
-\noindent\textbf{Methods:} We developed a spatiotemporal model of airborne \emph{Mtb} transmission using environmental (CO$_2$ levels), clinical data (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors).
-The data were collected n a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when mask-wearing was compulsory.
-We matched patient movements with clinical records to identify the spatiotemporal location of TB infectious patients.
-Using Monte Carlo simulation, we modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection with the Wells-Riley equation.
+\noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased.
+We matched patient movements with clinical records to identify the spatiotemporal location of  infectious TB patients.
+We modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection using the Wells-Riley equation.
 
-\noindent\textbf{Results:} Clinical records registered 894 patients, four of which were diagnosed with TB.
+\noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB.
 Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min).
-CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), suggesting that the clinic was well ventilated.
+CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), compatible with high ventilation rates.
 The modeled risk of infection per clinic attendee was 0.059\% (95\%-credible interval [CrI] 0.001\%$-$0.402\%).
-The modeled risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask wearing and 1.74-fold (95\%-CrI 0.47$-$5.34).
-Furthermore, the person-specific risk was significantly associated with the number of close-contact encounters and time spent in the clinic.
+The risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with lower, pre-pandemic ventilation rates.
+Furthermore, the individual risk was significantly associated with the number of close-contact encounters and time spent in the clinic.
 
-\noindent\textbf{Conclusions:} Risk of \emph{Mtb} transmission in the South African clinic was low during the COVID-19 pandemic, especially due to infection control measures.
-Our spatiotemporal modeling could be used to identify high-risk patients and assess the impact of mitigation measures.
+\noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures.
+Our spatiotemporal modeling could be used to identify high-risk areas and evaluate the impact of infection control measures in clinics.
 
@@
 % TB and the route of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South African region, and progress made in TB prevention and control before the COVID-19 pandemic has stalled or reversed\cite{WHO2022TBReport}.
-\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, transmits via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
-\emph{Mtb} is carried primarily in smaller particles in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
-Survival of airborne pathogens is greater in crowded, poorly ventilated indoor settings\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-As such, healthcare facilities present a particular high-risk setting because both people that are more infectious and susceptible also more frequently visit the clinic\cite{McCreesh2020IJTLD}.
-A study in South Africa estimated that 4\% to 14\% of TB cases in adults originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South African region, which is disproportionately affected by the TB burden.
+After two years of COVID-19-related disruption in the progress made in TB prevention and control before the pandemic, the impact of the pandemic is about to reverse \cite{WHO2022TBReport}.
+However, global TB targets have
+been largely missed (REF).
+\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
+\emph{Mtb} is primarily carried in smaller particles in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
+Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
+As such, healthcare facilities present a particularly high-risk setting because both more infectious and more susceptible individuals are also more likely to visit a clinic\cite{McCreesh2020IJTLD}.
+A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
 % the Wells-Riley model and its limitations
-The Wells-Riley model\cite{Riley1978AJE} is frequently used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
+The Wells-Riley model\cite{Riley1978AJE} is widely used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
 The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate.
-The Wells-Riley model, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same everywhere in the room.
-However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies showed that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
+The Wells-Riley model, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same throughout the room.
+However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
 
 % what this study adds
-We build on the Wells-Riley model and develop a spatiotemporal extension to model the concentration of infectious quanta spatially and over time.
-We combine clinical and video sensor data to identify the spatiotemporal location of TB patients and susceptible attendees of a primary care clinic in South Africa for five days in October/November 2021.
-Furthermore, we measure indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time.
-Using Monte Carlo simulation, we estimate the risk of infection per clinic attendee.
-Finally, we assess the impact of infection control measures that were in place during the study in response to the COVID-19 by modeling the risk of infection under hypothetical scenarios without mask wearing and assuming pre-pandemic ventilation conditions.
+We developed a spatiotemporal extension of the Wells-Riley model to model the concentration of infectious quanta in space and time.
+We combined clinical and person tracking data to determine the spatiotemporal location of TB patients and susceptible attendees at a primary care clinic in South Africa for five days in October/November 2021.
+We also measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time.
+We used Monte Carlo simulation to estimate the risk of infection per clinic attendee.
+Finally, we assessed the impact of infection control measures in place during the study in response to the COVID-19 pandemic (mask use and increased ventilation) by modeling the risk of infection under hypothetical pre-pandemic scenarios without universal mask use and pre-pandemic ventilation conditions in the clinic.
 
@@
 
-The primary care clinic offers both TB and HIV services and other basic clinical services, Monday to Friday, from 8\,am to 4\,pm.
-The clinic is situated within a large settlement of formal and semiformal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
+The primary care clinic offers both TB and HIV services as well as other basic clinical services, Monday through Friday, from 8\,am to 4\,pm.
+The clinic is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
 We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
-The TB room was mainly used for the treatment of diagnosed TB patients, but also for the screening of suspected patients with respiratory symptoms who were eventually not diagnosed with TB.
+The TB room was mainly used for the treatment of diagnosed TB patients, but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
 
@@
 
-We extracted clinical data from the electronic patient registry for all patients who visited the clinic during the study.
-Clinical records included the date and time of arrival, TB diagnostic results, and date of TB treatment start.
-We defined infectious patients as individuals testing bacteriologically positive for TB during the study or being on TB treatment for less than four weeks.
-Patients who had been treated for more than four weeks or recently completed treatment were neither considered infectious nor susceptible.
-We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were subsequently not diagnosed with TB.
+We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study.
+Clinical records included the date and time of arrival, TB diagnostic results, and date of TB treatment initiation.
+We defined infectious patients as individuals who tested bacteriologically positive for TB during the study or who had received TB treatment for less than four weeks.
+Patients on treatment for more than four weeks or who had recently completed treatment were considered neither infectious nor susceptible.
+We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were ultimately not diagnosed with TB.
 
-\subsubsection{Tracking data}
+\subsubsection{Person tracking data}
 
-We used an anonymous personal tracking system (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1s intervals (\Cref{fig:floor-plan}).
-The resulting timestamped movement data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-\supp~Figure~\zref{fig:tracking-examples} shows a few sample tracks from our study.
-Note that persons in the clinic could contribute multiple tracks if they moved out of a sensor's range or if they were briefly not recognized as a person.
-To remedy these issues, we created an R Shiny tool to manually link interrupted tracks back together (\supp Text~\zref{sec:setting-and-data}).
-Furthermore, we labeled tracks to identify movements from clinic staff.
+We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1~second intervals (\Cref{fig:floor-plan}).
+The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
+\supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks from our study.
+Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person.
+To address these issues, we created an R Shiny tool to manually reconnect interrupted tracks back together (\supp~Text~\zref{sec:setting-and-data}).
+We also labeled the tracks to identify movements of the clinic staff.
 
@@
 
-Our spatiotemporal model builds upon the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability in \%) of infection $P$ as
+Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability in \%) of infection $P$ as
 \begin{align}
@@
 The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
-Instead, our study considers a mixed airspace with spatial variation in the concentration of quanta.
+Instead, our study considers a mixed airspace with spatial variation in the quanta concentration.
 This is accomplished using patient movement data combined with clinical records to identify the spatiotemporal location of infectious patients who generate quanta.
@@
     \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
-(1)~Clinical and tracking data is combined to identify the spatiotemporal location of diagnosed (infectious) TB patients.
-(2)~Both diagnosed and randomly sampled undiagnosed  patients among visitors generate infectious quanta.
-(3)~Tracking and environmental data is combined to estimate the air change rate.
-(4)~The air change rate is related to the speed with which quanta diffuses in the indoor space.
+(1)~Clinical and person tracking data were combined to determine the spatiotemporal location of diagnosed (infectious) TB patients.
+(2)~Both diagnosed and randomly selected undiagnosed  TB among clinic attendees generate infectious quanta.
+(3)~Tracking and environmental data were combined to estimate the air change rate.
+(4)~Air change rate was related to the speed at which quanta diffuses indoors.
 (5)~The air change rate also determines quanta removal through outdoor air exchange, in addition to inactivation of \emph{Mtb} in airborne particles.
-(6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, based on which the personal risk of infection is computed using the Wells-Riley equation.}
+(6)~Quanta generation, diffusion, and removal determine the spatiotemporal quanta concentration, from which the personal risk of infection is calculated using the Wells-Riley equation.}
     \label{fig:modeling-flow}
@@
 
-(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients inside the clinic.
+(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients within the clinic.
 We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records.
@@
 Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
+Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
 Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
@@
 (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}.
-Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).
+Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Fig.~\zref{fig:toy-example}).
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
 We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
@@
 Our study was set during the COVID-19 pandemic when infection control measures were in place.
-To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pilot study of 2019 before the pandemic\cite{Zurcher2022JID}.
-Furthermore, we examined the impact of different assumptions regarding the number of infectious patients in the clinic: 1)~instead of undiagnosed TB patients, we considered suspected TB patients as infectious, because these patients may still generate aerosolized \emph{Mtb} even when they were not diagnosed with TB at first presentation\cite{Patterson2024PNAS}; 2)~random sampling of infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}, thereby pretending that the clinical data had not been available.
+To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pre-pandemic pilot study in 2019\cite{Zurcher2022JID}.
+Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~instead of undiagnosed TB patients, we considered patients clinically suspected of having TB as infectious, because these patients may still generate aerosolized \emph{Mtb} even when they were not diagnosed with TB at first presentation\cite{Patterson2024PNAS}; 2)~randomly selecting infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}, thereby pretending that the clinical data were not available.
 Except for the scenario-specific modifications, all other modeling assumptions and parameters were the same in each Monte Carlo simulation, ensuring comparability of the results.
@@
 Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean and 95\%-credible interval (CrI).
+The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).
 A Bayesian Beta regression model was used to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact, and frequency of close-contact encounters.
@@
 
-During the five study days, 894 patients have been registered in the clinical database, of which four patients were diagnosed with TB and 16 patients were suspected of TB following symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}).
-Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as being from healthcare workers, identifying 1,438 clinic attendees (patients or other visitors).
+During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Fig.~\zref{fig:diagnosed-tb-patients}).
+Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as healthcare worker, identifying 1,438 clinic attendees (patients or other visitors).
 Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
-Clinic attendees spent around half an hour inside the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
-Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9), and several clinic attendees had more than ten different close contacts during their visit (\Cref{fig:input-data-descriptives}d).
-The majority of clinic attendees was in close contact with at least one other person for more than half the time of their visit (median 68\%, IQR 41\%$-$87\%, \Cref{fig:input-data-descriptives}e).
-The CO$_2$ levels inside the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\Cref{fig:input-data-descriptives}f) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
+Clinic attendees spent about half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
+Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9) and clinic attendees were in close contact with at least one other person for more than half of their visit (median 68\%, IQR 41\%$-$87\%).
+The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
 
@@
     \includegraphics{results/data/time-in-clinic.png}
-    \includegraphics{results/data/number-close-contacts.png}
-    \includegraphics{results/data/time-close-contacts.png}
-    \includegraphics{results/data/co2-levels-over-time.png}
     \includegraphics{results/data/air-changes-per-hour.png}
-    \caption{\textbf{Environmental conditions and patient movement characteristics.} \textbf{(a)}~CO$_2$ levels per room of the clinic over time per day.
-\textbf{(b)}~Air change rates per room of the clinic.
-\textbf{(c)}~Total number of tracks recorded per unit space in the clinic over the study period.
-\textbf{(d)}~Number of people in the clinic over time per day.
-\textbf{(e)}~Time spent in the clinic per attendee.
-\textbf{(f)}~Density distribution of the number of close contacts per clinic attendee during their visit.
-\textbf{(g)}~Density distribution of the proportion of visit time spent in close contact per clinic attendee.}
+    \caption{\textbf{Person movement characteristics and environmental conditions.} \textbf{(a)}~Total number of tracks recorded per unit space in the clinic over the study period.
+\textbf{(b)}~Number of people in the clinic over time per day.
+\textbf{(c)}~Time spent in the clinic per attendee.
+\textbf{(d)}~Air change rates per room of the clinic.}
     \label{fig:input-data-descriptives}
@@
 
-\supp~File~1 shows the spatiotemporally varying quanta concentration as an infectious patient enters the clinic in the morning of October 25 and moves through the waiting room.
-It can be seen that the quanta concentration was higher near the infectious patient, therefore indicating his position and movement.
-The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open throughout clinic hours.
+\supp~File~1 shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
+It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
+The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
 \Cref{fig:main-modeling-results} shows the results from our main spatiotemporal modeling analysis across study days and simulations.
 On average, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a).
-An exposure of one hour to the mean quanta concentration in the center of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
-The mean risk of infection was 0.059\% (95\%-CrI 0.001\%$-$0.402\%) and the majority of clinic attendees (87\%) had a low risk <0.01\% risk of infection (\Cref{fig:main-modeling-results}b).
-To put this into perspective, a patient visiting the clinic twelve times per year would have a cumulative annual risk of infection of 0.7\% (95\%-CrI 0.01\%$-$4.72\%).
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
+The mean risk of infection per visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
+To put this into perspective, a patient with twelve hourly visits per year would have a cumulative annual risk of infection of 1.37\% (80\%-CrI 0.14\%$-$2.74\%).
 
-Our modeling approach allowed estimating the person-specific risk of infection with uncertainty.
-As shown in \Cref{fig:main-modeling-results}b, it can be used to identify patients with a comparably higher mean risk of infection at low uncertainty.
-The person-specific modeled risk of infection was further associated with movement characteristics (\Cref{fig:main-modeling-results}c).
-If the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30).
-Analogously, if the time spent in the clinic doubled, the odds of infection increased to 1.09 (95\%-CrI 1.01$-$1.18).
-Finally, a comparison of our spatiotemporal modeling results with those from a temporal model assuming a well-mixed airspace showed notable difference in the person-specific risk of infection with a median of 8\% (IQR 4\%$-$15\%).
+Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Fig.\zref{fig:main-modeling-results}).
+The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
+When the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30).
+Similarly, doubling the time spent in the clinic increased the odds of infection to 1.09 (95\%-CrI 1.01$-$1.18).
+The time spent in close contact was not significantly associated with the individual risk of infection (1.05, 95\%-CrI, 0.96$-$1.14).
+Finally, a comparison of our spatiotemporal modeling results with those of a temporal model assuming a well-mixed airspace showed a notable difference in the individual risk of infection with a median of 8\% (IQR 4\%$-$15\%).
 
@@
     \includegraphics{results/modeling/main-figure.png}
-    \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Average quanta concentration in the morning and afternoon in the clinic.
-\textbf{(b)}~Mean risk of infection per clinic attendee.
-Low: <5\%, Medium: 5-10\%, High: >10\%.
-\textbf{(c)}~Person-specific mean risk of infection vs uncertainty (IQR in percentage points, pp).
-\textbf{(d)}~Association of person-specific mean risk of infection with patient movement characteristics (mean odds ratio [OR] and 95\%-credible interval [CrI] if the value of the predictor doubles).}
+    \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Modeled average quanta concentration in the morning and afternoon in the clinic.
+\textbf{(b)}~Mean risk of infection per clinic attendee per visit (median time 25\,min, IQR 13\,min$-$46\,min).
+Low: <5\%, Medium: 5-10\%.
+\textbf{(c)}~Individual mean risk of infection vs uncertainty (IQR in percentage points, pp).
+\textbf{(d)}~Association of individual mean risk of infection with patient movement characteristics (mean odds ratio [OR] and 95\%-credible interval [CrI] if the value of the predictor doubles).}
     \label{fig:main-modeling-results}
@@
 
-To assess the impact of infection control measures during the COVID-19 pandemic, we performed two hypothetical scenarios shown in the top panel of \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection would have been 0.075\% (95\%-CrI 0.001\%$-$0.513\%), a 1.28-fold (95\%-CrI 1.19$-$1.36) increase, and there would have been four patients with a high risk of infection >1\%, which would translate into a cumulative annual risk of more than 10\%.
-Similarly, if the air change rate had been as before the COVID-19 pandemic, the mean risk of infection would have been (0.097\%, 95\%-CrI 0.001\%$-$0.820\%), a 1.74-fold (95\%-CrI 0.47$-$5.34) increase, and there would have been 27 patients with a high risk of infection >1\%.
-Both scenarios show an increase in the number of clinic attendees with a high risk of infection, indicating that mask wearing and better ventilation considerably impacted the tail of the risk distribution, which can also be seen in the smaller upper limit of the credible intervals.
+To assess the impact of infection control measures during the COVID-19 pandemic, we ran hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates (\Cref{fig:scenario-results}).
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.07\% (80\%-CrI 0.01\%$-$0.09\%), a 1.3-fold increase (80\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been (0.10\%, 80\%-CrI 0.01\%$-$0.09\%), a 1.7-fold increase (80\%-CrI 0.5$-$5.3).
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been (0.13\%, 95\%-CrI 0.01\%$-$0.13\%), a 2.5-fold increase (80\%-CrI 0.6$-$8.8).
+The scenarios show an increase in the number of clinic attendees with a high risk of infection, suggesting that universal use of mask and better ventilation in the clinic had a considerable impact on the tail of the risk distribution, which can also be seen in the smaller upper limit of the credible intervals.
 
-To assess the impact of modeling assumptions regarding the number of infectious people in the clinic, we performed two hypothetical scenarios shown in the bottom panel of \Cref{fig:scenario-results}.
-Assuming the number of infectious TB patients had been in proportion to the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.079\%, 95\%-CrI 0.003\%$-$0.304\%).
-Assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean and tail risk of infection would have been considerably higher (0.133\%, 95\%-CrI 0.000\%$-$0.819\%), and there would have been 21 clinic attendees with a high risk of infection >1\%.
+To assess the impact of modeling assumptions, we ran scenarios making different assumptions about the number of infectious people in the clinic.
+Assuming the number of infectious TB patients had been in proportion to the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.08\%, 80\%-CrI 0.02\%$-$0.13\%).
+Assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean and tail risk of infection would have been considerably higher (0.14\%, 80\%-CrI <0.01\%$-$0.26\%).
 
@@
     \includegraphics{results/modeling/mean-roi-comparison.png}
-    \caption{\textbf{Scenario modeling results.} Mean risk of infection per clinical attendee if \textbf{(a)}~masks had not been worn by all clinical attendees and staff, \textbf{(b)}~the ventilation rate had been lower as in the year before the COVID-19 pandemic, \textbf{(c)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(d)}~the number of infectious people in the clinic corresponded to the number of diagnosed TB patients and those suspected of TB due to respiratory symptoms when attending the clinic.
-Low: <5\%, Medium: 5-10\%, High: >10\%.}
+    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
+Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
     \label{fig:scenario-results}
@@
 % summary
-We developed a spatiotemporal model to estimate the concentration of infectious quanta in an indoor space and applied it to patient movement data from a South African primary care clinic in 2021 to simulate the risk of TB infection for clinic attendees.
+We developed a spatiotemporal model to estimate the indoor concentration of infectious quanta and consider patient movement data from a South African primary care clinic in 2021 to accurately simulate the risk of Mtb infection for clinic attendees.
 We estimated an average risk of infection of 0.059\% (95\%-CrI 0.001\%$-$0.402\%), which would correspond to a cumulative annual risk of 0.7\% (95\%-CrI 0.01\%$-$4.72\%) if each patient visited the clinic once per month.
 Our model can identify patients with a higher risk of infection and low uncertainty.
-Furthermore, we found that clinic attendees who stayed longer inside the clinic and had more close-contact encounters also had a higher risk of infection.
-The modeled risk of infection would have been higher without the implementation infection control measures during the COVID-19 pandemic such as compulsory face mask wearing and better ventilation conditions.
-We applied our spatiotemporal model to \emph{Mtb} but it can be readily adapted to model the airborne transmission risk of other respiratory infections such as SARS-CoV-2 or influenza.
+We also found that clinic attendees who stayed longer in the clinic and had more close-contact encounters  had a higher risk of infection, and we estimated the potential impact of infection control intervention.
+The modeled risk of infection would have been higher had infection control measures not been implemented during the COVID-19 pandemic such as compulsory face masks and improved ventilation conditions.
+Our spatiotemporal model can be readily adapted to model to other airborne transmission risk of other respiratory infections such as SARS-CoV-2.
 
@@
 South Africa has a high burden of TB\cite{WHO2022TBReport}.
-Despite that, we estimated a low risk of infection per clinic attendee in our primary care clinic.
-In our previous study before the COVID-19 pandemic, we estimated a cumulative annual risk of \emph{Mtb} transmission of between 9\% and 29\% for a patient coming to the clinic once per month for one hour.
+Despite that, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
+In our previous pre-pandemic study before the COVID-19 pandemic, we estimated a cumulative annual risk of \emph{Mtb} transmission of between 9\% and 29\% for a patient coming to the clinic once per month for one hour.
 Although the assumed visit time is longer than what we observed in this study, the risk difference is still noteworthy and most likely due to the infection control measures that were in place during the SARS-CoV-2 pandemic.
@@
 To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.48\,h$^{-1}$), Switzerland (0.71\,h$^{-1}$), and Tanzania (13.74\,h$^{-1}$)\cite{Banholzer2024PGPH}.
-Tanzania had a much lower risk of \emph{Mtb} transmission in schools than the other countries, which was largely due to better ventilation, and yet its mean air change rate was less than half the rate measured in our study.
+Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, which was largely due to better ventilation, and yet its mean air change rate was less than half the rate measured in our study.
 Thus, we note that our study results do \emph{not} indicate that masks are less effective as found in other studies, rather their effectiveness is overshadowed by excellent ventilation conditions in the clinic during the COVID-19 pandemic.
@@
 Our modeling study has several limitations.
-First, tracking patient movements with video sensors is challenging and required considerable data processing.
-Despite that, many patient movements were interrupted and could not be linked back together.
+First, tracking patient movements with video sensors is challenging and required considerable data processing and many patient movements were interrupted and could not be connected back together.
 Therefore, some clinic attendees may have had longer visits and the risk of infection may have been underestimated.
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-Uncertainty in these modeling parameters was often large and could be reduced in future work, for example, by incorporating additional information on patient characteristics from clinical records that are known to be associated with patient-specific infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
+Uncertainty in these modeling parameters was often large and could potentially be reduced in future workby incorporating , for example, additional information on patient characteristics from clinical records that are known to be associated with patient-specific infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
 Third, we could not validate our modeling results with empirical findings.
-A follow-up study could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be infected.
+Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be infected.
 
 % conclusions
-In conclusion, our modeling study using clinical, environmental, and patient movement data showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa in October/November 2021 was low and mitigated by infection control measures that were implemented during the COVID-19 pandemic.
+In conclusion, our novel spatio-temporal transmission model study using clinical, environmental, and patient movement data showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa under pandemic conditions with strict infection control measures in place.
 The modeled risk of infection was significantly associated with the number of close-contact encounters by clinic attendees during their visit.

### 2024-04-18 doc/paper/supplements.tex

@@
     \includegraphics{results/inputs/bacterial-inactivation-rate.png}
-    \caption{Prior distribution (log scale) for the bacterial inactivation rate (quanta h$^{-1}$) of \emph{Mtb} along with reported estimates from the literature\cite{Loudon1969AMRRD,Gannon2007ResVetSci,Klein2014IJ,Lever2000LettersAppliedMicrobioMyco}.}
+    \caption{Prior distribution (log scale) for the bacterial inactivation rate (quanta h$^{-1}$) of \emph{Mtb} along with reported estimates from the literature\cite{Loudon1969AMRRD,Gannon2007ResVetSci,Klein2014IJMyco,Lever2000LettersAppliedMicrobio}.}
     \label{fig:lambda-distribution}
@@
 
-Undiagnosed (subclinical) TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats), or the diagnosis could not be made at first presentation based on existing tests\cite{Patterson2023Preprint}.
+Undiagnosed (subclinical) TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats), or the diagnosis could not be made at first presentation based on existing tests\cite{Patterson2024PNAS}.
 Previous studies suggest that the number of undiagnosed (subclinical) TB patients is roughly similar to the number of diagnosed TB patients\cite{Berhanu2023CID,Moyo2022LancetID}.
@@
 Again, the inhalation rate for each susceptible clinic attendee is determined based on whether the distance between two tracks was more (walking) or less (sitting) than 0.25m.
+
+\clearpage
+
+\section{Additional results}\label{sec:additional-results}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/diagnosed-tb-patients.png}
+    \caption{Number of diagnosed TB patients (dark bars) and those suspected of TB (light bars) following symptom screening.
+Modeled study days in color, other clinic days in October and November in grey.}
+    \label{fig:diagnosed-tb-patients}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/co2-levels-over-time.png}
+    \caption{CO$_2$ levels per room of the clinic over time per day.}
+    \label{fig:co2-levels}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/modeling/mean-roi-assumptions.png}
+    \caption{\textbf{Scenario modeling results.} Mean risk of infection per clinical attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
+Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
+    \label{fig:assumptions-results}
+\end{figure}
 

### 2024-04-18 analysis/patient-tracking.Rmd
Minor changes to visualizations

@@
   pdf_file = "results/data/no-people-over-time.png",
-  w = 8, h = 6
+  w = 8, h = 8
 )
@@
   pdf_file = "results/data/no-people-spatial.png",
-  w = 8, h = 6
+  w = 8, h = 8
 )

### 2024-04-18 preprocessing/prep_building-rasterize.R

@@ -22,7 +22,7 @@ sf::st_crs(building_sf) <- NA
 
 if (!exists("cellSize")) {
   stop("No local cellSize variable found")
-} 
+}
 
 rasterize.rooms <- function(cellSize) {
   waiting_room <- shapeToSpatial(building_sf$geometry[2], cellSize)
@@ -41,9 +41,11 @@ cdCoord <- create_coord_df(corridor)
 tb_room <- rooms$tb_room
 tbMat <- sP_to_matrix(tb_room)
 tbCoord <- create_coord_df(tb_room)
-roomCoord <- rbind(wrCoord %>% mutate(room = "Waiting room"), 
-                   cdCoord %>% mutate(room = "Corridor"), 
-                   tbCoord %>% mutate(room = "TB room")) 
+roomCoord <- rbind(
+  wrCoord %>% mutate(room = "Waiting room"),
+  cdCoord %>% mutate(room = "Corridor"),
+  tbCoord %>% mutate(room = "TB room")
+)
 
 # room dimensions
 # - waiting room
@@ -56,26 +58,29 @@ dimTB <- c(4.75, 3.5, 3)
 # plotting
 
 plot_spatial <- function(pl, text_descr = 10) {
-  
   entrance_lab <- textGrob(label = "Entrance", x = 0.1, y = 0.165, gp = gpar(fontsize = text_descr), just = c("left"))
   wr_lab <- textGrob(label = "Waiting room", x = 0.05, y = 0.45, gp = gpar(fontsize = text_descr), just = c("left"))
   tb_lab <- textGrob(label = "TB room", x = 0.75, y = 0.4, gp = gpar(fontsize = text_descr), just = c("left"))
+  cr_lab <- textGrob(label = "Corridor", x = 0.6, y = 0.615, gp = gpar(fontsize = text_descr), just = c("left"))
   re_lab <- textGrob(label = "Registration", x = 0.44, y = 0.25, gp = gpar(fontsize = text_descr), just = c("left"), rot = 90)
   col <- RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")
-  
- pl +
+
+  pl +
     ggpubr::background_image(clinic_img) +
-    scale_y_continuous(expand = expansion(add=c(1700, 3000)), limits = c(-6500, 2000)) +
-    scale_x_continuous(expand = expansion(add=c(400, 500)), limits = c(-16000, 7500)) +
+    scale_y_continuous(expand = expansion(add = c(1700, 3000)), limits = c(-6500, 2000)) +
[…]

### 2024-04-18 doc/paper/manuscript.tex

@@
 The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-\supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks from our study.
+\supp~Figure~\zref{fig:tracking-examples} shows some sample tracks from our study.
 Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person.
@@
 Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
+Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
 Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
@@
 (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}.
-Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Fig.~\zref{fig:toy-example}).
+Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
 We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
@@
 
-During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Fig.~\zref{fig:diagnosed-tb-patients}).
+During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}).
 Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as healthcare worker, identifying 1,438 clinic attendees (patients or other visitors).
@@
 Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9) and clinic attendees were in close contact with at least one other person for more than half of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
+The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Figure~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
 
@@
     \includegraphics{results/data/air-changes-per-hour.png}
-    \caption{\textbf{Person movement characteristics and environmental conditions.} \textbf{(a)}~Total number of tracks recorded per unit space in the clinic over the study period.
-\textbf{(b)}~Number of people in the clinic over time per day.
-\textbf{(c)}~Time spent in the clinic per attendee.
-\textbf{(d)}~Air change rates per room of the clinic.}
+    \caption{\textbf{Patient movements and environmental conditions in the clinic.} Across clinic days and for the waiting room, corridor and TB room of the clinic: \textbf{(a)}~total number of tracks recorded per unit space, \textbf{(b)}~number of people in the clinic over time, \textbf{(c)}~time per clinic visit, and \textbf{(d)}~air change rates.
+Lines and ribbons show the mean plus/minus standard deviation, respectively.
+Boxplots show the median as lines, IQR as boxes, range as whiskers, and outliers as dots.}
     \label{fig:input-data-descriptives}
@@
 The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
-\Cref{fig:main-modeling-results} shows the results from our main spatiotemporal modeling analysis across study days and simulations.
-On average, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}a).
+Across clinic days, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
 An exposure of one hour to the mean quanta concentration in the middle of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
-The mean risk of infection per visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
+The mean risk of infection per clinic visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
 To put this into perspective, a patient with twelve hourly visits per year would have a cumulative annual risk of infection of 1.37\% (80\%-CrI 0.14\%$-$2.74\%).
 
-Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Fig.\zref{fig:main-modeling-results}).
+Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Figure\zref{fig:patient-risk-focus}).
 The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
@@
     \centering
-    \includegraphics{results/modeling/main-figure.png}
-    \caption{\textbf{Spatiotemporal modeling results.} \textbf{(a)}~Modeled average quanta concentration in the morning and afternoon in the clinic.
-\textbf{(b)}~Mean risk of infection per clinic attendee per visit (median time 25\,min, IQR 13\,min$-$46\,min).
-Low: <5\%, Medium: 5-10\%.
-\textbf{(c)}~Individual mean risk of infection vs uncertainty (IQR in percentage points, pp).
-\textbf{(d)}~Association of individual mean risk of infection with patient movement characteristics (mean odds ratio [OR] and 95\%-credible interval [CrI] if the value of the predictor doubles).}
+    \includegraphics{results/modeling/mean-quanta-concentration.png}
+    \caption{\textbf{Spatiotemporal quanta concentration in the clinic.} Model-estimated average quanta concentration across clinic days in the morning and afternoon in the waiting room, corridor and TB room of the clinic.}
     \label{fig:main-modeling-results}
@@
     \includegraphics{results/modeling/mean-roi-comparison.png}
-    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
+    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
 Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}

### 2024-04-18 doc/paper/supplements.tex

@@
     \includegraphics{results/data/co2-levels-over-time.png}
-    \caption{CO$_2$ levels per room of the clinic over time per day.}
+    \caption{CO$_2$ levels per clinic room and day over time.}
     \label{fig:co2-levels}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{results/modeling/patient-risk-focus-plot.png}
+    \caption{Mean and 80\%-CrI (uncertainty) of the risk of infection per clinic visit.
+Highlighted in red is a visit from a clinic attendee with moderate risk of infection and low estimated uncertainty.}
+    \label{fig:patient-risk-focus}
 \end{figure}
@@
     \includegraphics{results/modeling/mean-roi-assumptions.png}
-    \caption{\textbf{Scenario modeling results.} Mean risk of infection per clinical attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
+    \caption{Mean risk of infection per clinic visit if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
 Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}

### 2024-05-14 analysis/environmental-data.Rmd
Fixed computation of track counts over time and air change rate

@@
 cont_df %>%
-  group_by(date) %>%
-  summarize(across(c(co2, temperature, humidity), .fns = list(mean, sd)))
-
-cont_df %>%
-  group_by(date, room) %>%
-  summarize(across(c(co2), .fns = list(mean, sd))) %>%
-  mutate_if(is.numeric, round) %>%
-  ungroup() %>%
-  arrange(room)
+  summarize(
+    across(c(co2, temperature, humidity),
+      .fns = list(
+        median = median,
+        q25 = function(x) quantile(x, .25),
+        q75 = function(x) quantile(x, .75)
+      )
+    )
+  ) %>%
+  gather() %>%
+  mutate(value = round(value))
 ```
@@
 )
-
-cont_df %>%
-  median_qi(co2, .width = .5) %>%
-  mutate_if(is.numeric, round)
 ```

### 2024-05-14 analysis/patient-tracking.Rmd
Fixed computation of track counts over time and air change rate

@@
 source("utils/spatial.r")
+cellSize <- 250
+source("preprocessing/prep_building-rasterize.R")
 ```
@@
 ```{r}
+# spatial track counts
+tracking_data <- do.call(
+  rbind,
+  lapply(
+    list.files("data-raw/patient-tracking/annotated",
+      full.names = TRUE
+    ),
+    readRDS
+  )
+) %>%
+  rename(date_time = time) %>%
+  mutate(date = as.Date(date_time)) %>%
+  filter(date %in% tracking_dates) %>%
+  filter(between(hour(date_time), 8, 15)) %>%
+  mutate(
+    room = ifelse(is_waitingroom, "Waiting room",
+      ifelse(is_tbroom, "TB room",
+        ifelse(is_passage, "Corridor", NA)
+      )
+    ),
+    room = factor(room, levels = c("Waiting room", "Corridor", "TB room"))
+  )
+
+# track counts over time
+track_count_time <- readRDS("data-clean/patient-tracking/tracks-over-time.rds")
+
+# manually reconnected tracking data
 df <- readRDS("data-clean/patient-tracking/augmented-and-combined-data.rds")
@@
 ```{r}
-clinic_occupancy <- df %>%
-  group_by(date, time) %>%
-  summarize(
-    `Clinic` = n(),
-    `Waiting room` = sum(is_waitingroom),
-    `TB room` = sum(is_tbroom),
-    `Corridor` = sum(is_passage)
-  ) %>%
-  ungroup() %>%
-  dplyr::select(-Clinic) %>%
-  melt(c("date", "time")) %>%
+track_count_time <- track_count_time %>%
   mutate(
-    variable = factor(
-      variable,
-      levels = c("Clinic", "Waiting room", "TB room", "Corridor")
-    ),
     date = format(date, "%b %d"),
@@
     ),
-    time = as.POSIXct(format(time, format = "%H:%M:%S"), format = "%H:%M:%S")
-  ) %>%
-  filter(between(hour(time), 7, 16))
+    time = as.POSIXct(format(date_time, format = "%H:%M:%S"), format = "%H:%M:%S")
+  )
 
-time_pl <- clinic_occupancy %>%
-  filter(variable != "Clinic") %>%
-  group_by(time, variable) %>%
+time_pl <- track_count_time %>%
+  group_by(time, room) %>%
   summarise(
-    mv = mean(value),
-    sv = sd(value)
+    mv = mean(n),
+    sv = sd(n)
   ) %>%
@@
   geom_line(
-    aes(y = mv, color = variable)
+    aes(y = mv, color = room)
   ) +
   geom_ribbon(
-    aes(ymin = lv, ymax = uv, fill = variable),
+    aes(ymin = lv, ymax = uv, fill = room),
     alpha = 0.4
@@
   coord_cartesian(expand = FALSE) +
-  scale_x_datetime(date_breaks = "3 hours", date_labels = "%H:%M") +
+  scale_x_datetime(
+    date_breaks = "2 hours", date_labels = "%H:%M"
+  ) +
   scale_fill_manual(values = wes_palette("Royal2")) +
@@
     legend.key.width = unit(0.275, "cm"),
-    legend.key.height = unit(0.2, "cm")
+    legend.key.height = unit(0.2, "cm"),
+    plot.margin = margin(r = 11)
   )
@@
 spatial_pl <- plot_spatial(
-  ggplot(data = df, mapping = aes(x = x, y = y)) +
+  ggplot(data = tracking_data, mapping = aes(x = x, y = y)) +
     geom_bin2d(binwidth = 200),

### 2024-05-14 analysis/qctr.Rmd
Fixed computation of track counts over time and air change rate

@@
     legend.key.width = unit(.4, "cm"),
-    panel.spacing = unit(2, "cm"),
     strip.text = element_text(hjust = 0)

### 2024-05-14 preprocessing/prep_environmental.R

@@ -3,61 +3,96 @@
 library(tidyverse)
 library(lubridate)
 library(haven)
+library(zoo)
+
+# linear imputation
+lin_imp <- function(x, mg = 600) {
+  na.locf(
+    na.locf(
+      na.approx(x, maxgap = mg, na.rm = FALSE),
+      maxgap = mg, na.rm = FALSE
+    ),
+    fromLast = TRUE, maxgap = mg, na.rm = FALSE
+  )
+}
+
+# coalesce / imputation
+imp_xy <- function(x, y) {
+  if (!is.na(x) & !is.na(y)) {
+    return(mean(c(x, y)))
+  } else if (!is.na(x)) {
+    return(x)
+  } else if (!is.na(y)) {
+    return(y)
+  } else {
+    return(NA)
+  }
+}
 
 source("utils/trans_risk.r")
 
-#### Data ####
+#### Environmental data ####
 
 # Backup data
 files <- list.files("data-raw/environmental/backup/", full.names = T)
 
-masi2021 <- do.call(rbind,
-              lapply(files, function(f) {
-                read.table(f, sep = "\t", header = F, skip = 33) %>%
-                  set_names(c("date", "time", "humidity", "temperature", "co2", "ext")) %>%
-                  dplyr::select(-ext) %>%
-                  mutate(file = basename(f))
-              })) %>%
-  mutate(date_created = as.Date(stringi::stri_extract(file, regex = "\\d{8}"), "%Y%m%d"),
-         date = as.Date(date, format = "%m/%d/%Y"),
-         sensor = as.integer(gsub("-", "", stringi::stri_extract(file, regex = "-\\d{1}"))),
-         date_time = parse_date_time(paste(date, time), '%Y-%m-%d %I:%M:%S %p', tz = "CET"),
-         date_time = round_date(date_time, unit = "minutes"),
-         date = as.Date(date_time)) %>%
+masi2021 <- do.call(
+  rbind,
+  lapply(files, function(f) {
+    read.table(f, sep = "\t", header = F, skip = 33) %>%
+      set_names(c("date", "time", "humidity", "temperature", "co2", "ext")) %>%
+      dplyr::select(-ext) %>%
+      mutate(file = basename(f))
+  })
+) %>%
+  mutate(
[…]

### 2024-05-16 illustrations/priors.Rmd
Update prior figures

@@
   ) +
-  geom_text(
+  geom_label(
     data = prev_est,
     mapping = aes(x = x, label = lab, color = study, y = y),
-    size = 8 / cm(1), vjust = -2, hjust = 1
+    size = 8 / cm(1), vjust = 0, hjust = 1.05,
   ) +
@@
     expand = c(0, 0),
-    labels = scales::percent_format(suffix = "")
   ) +
@@
   ) +
-  geom_text(
+  geom_label(
     data = prev_est_vir,
     mapping = aes(x = x, label = lab, color = study, y = y),
-    size = 8 / cm(1), vjust = -2
+    size = 8 / cm(1), nudge_y = .05
   ) +
@@
   scale_y_continuous(
-    expand = expansion(mult = c(0, .05)),
-    labels = scales::percent_format(suffix = "")
+    expand = expansion(mult = c(0, .05))
   ) +

### 2024-05-16 preprocessing/prep_clinical.R

@@ -9,48 +9,58 @@ library(lubridate)
 # Masi
 
 df_pat_time_masi <- read.csv("data-raw/clinical/patient-time.csv") %>%
-  rename(date = action_date) %>%
-  mutate(time = parse_date_time(paste(date, time), '%m/%d/%Y %I:%M:%S %p', tz = "CET"),
-         date = as.Date(time)) %>%
-  filter(!is.na(time))
+       rename(date = action_date) %>%
+       mutate(
+              time = parse_date_time(paste(date, time), "%m/%d/%Y %I:%M:%S %p", tz = "CET"),
+              date = as.Date(time)
+       ) %>%
+       filter(!is.na(time))
 
 df_tb_pat <- readxl::read_excel("data-raw/clinical/CocT Masiphumelele_clinic data Apr_Nov 2021.xlsx", sheet = "Worksheet 1") %>%
-  rename(age = `Age(Years)`,
-         gender = Gender,
-         tb_treat_start = `Treatment Start Date`,
-         folder_number = `Folder Number`) %>%
-  dplyr::select(age, gender, tb_treat_start, folder_number) %>%
-  mutate(gender = factor(gender, levels = c("Male", "Female")),
-         tb_treat_start = as.Date(tb_treat_start),
-         folder_number = as.integer(folder_number))
+       rename(
+              age = `Age(Years)`,
+              gender = Gender,
+              tb_treat_start = `Treatment Start Date`,
+              folder_number = `Folder Number`
+       ) %>%
+       dplyr::select(age, gender, tb_treat_start, folder_number) %>%
+       mutate(
+              gender = factor(gender, levels = c("Male", "Female")),
+              tb_treat_start = as.Date(tb_treat_start),
+              folder_number = as.integer(folder_number)
+       )
 
 df_tb_sus <- read.csv("data-raw/clinical/tb-suspects.csv") %>%
-  mutate(was_suspected = T,
-         sus_tested_positive = ifelse(result_txt %in% c("TB_GENEXP_POS_RIF_S", "TB_GENEXP_POS_RIF_R"), T, F),
-         sus_started_treat = ifelse(care_status_txt %in% c("CARE_STATUS_STARTED", "CARE_STATUS_STARTED_OTHER"), T, F),
-         sus_treat_start = ifelse(sus_started_treat, care_status_date, NA),
-         sus_treat_start = as.Date(sus_treat_start, format = "%d.%m.%y")) %>%
[…]

### 2024-05-16 doc/paper/manuscript.tex

@@
 
-\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium} tuberculosis transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
+\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
 % author list
@@
 \author[1]{Kathrin Zürcher}
-\author[1]{Matthias Egger}
+\author[1,4,5]{Matthias Egger}
 \author[2]{Robin Wood}
@@
 \affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}
+\affil[4]{Population Health Sciences, University of Bristol, Bristol, United Kingdom}
+\affil[5]{Centre for Infectious Disease Epidemiology and Research, University of Cape Town, Cape Town, South Africa}
 
@@
 \vspace{1em}
+
+\begin{information}
+
+\noindent\noindent\textbf{Supplementary Information:} Includes supplementary text, tables and figures.
+
+\noindent\textbf{Supplementary Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
+
+\end{information}
 
@@
 % %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
-
-% \noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
 
@@
 \begin{abstract}\normalfont
-\noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities.
+The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics.
+We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation.
+We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased.
+We matched person tracks with clinical records to identify the spatiotemporal location of infectious TB patients.
+We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
+Over five days, video sensors detected 1,438 clinic attendees.
+CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
+The estimated risk of infection per clinic attendee was 0.06\% (95\%-credible interval [CrI] 0.001\%$-$0.40\%).
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.28-fold (95\%-CrI 1.19$-$1.36) higher in scenarios without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with pre-pandemic ventilation rates.
+Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
+
+% \noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities.
 Existing approaches to modeling the risk of airborne transmission typically assume that the airspace is well mixed, not considering that proximity to infectious individuals may increase the risk of transmission.
@@
 
-\noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased.
+% \noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased.
 We matched patient movements with clinical records to identify the spatiotemporal location of  infectious TB patients.
@@
 
-\noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB.
+% \noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB.
 Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min).
@@
 
-\noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures.
+% \noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures.
 Our spatiotemporal modeling could be used to identify high-risk areas and evaluate the impact of infection control measures in clinics.
@@
 
-\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, patient tracking, spatiotemporal model
+\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, person tracking, spatiotemporal model
 % maximum of 3-5 keywords
@@
 %TC:break main
-\section{Introduction}
+\section*{Introduction}
 
 % TB and the route of transmission
-Tuberculosis (TB) is one of the leading causes of death worldwide, especially in the South African region, which is disproportionately affected by the TB burden.
-After two years of COVID-19-related disruption in the progress made in TB prevention and control before the pandemic, the impact of the pandemic is about to reverse \cite{WHO2022TBReport}.
-However, global TB targets have
-been largely missed (REF).
+Tuberculosis (TB) remains one of the leading causes of death globally, with the South African region bearing a disproportionate share of the TB burden.
+After two years of COVID-19-related disruption of TB prevention and control measures, the negative impact of the COVID-19 pandemic has been receding, as documented in the World Health Organization (WHO) 2023 global TB report\cite{WHO2023TBReport}.
+However, progress was insufficient to meet global TB targets\cite{WHO2023TBReport}.
+For example, investment in TB research was only about half of the USD 2~billion target\cite{WHO2023TBReport}.
+
 \emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}.
-\emph{Mtb} is primarily carried in smaller particles in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
+\emph{Mtb} is primarily carried in small particles of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}.
 Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-As such, healthcare facilities present a particularly high-risk setting because both more infectious and more susceptible individuals are also more likely to visit a clinic\cite{McCreesh2020IJTLD}.
+Healthcare facilities are high-risk setting because infectious individuals are more likely to visit a clinic\cite{McCreesh2020IJTLD}.
 A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
@@
 % the Wells-Riley model and its limitations
-The Wells-Riley model\cite{Riley1978AJE} is widely used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
-The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate.
-The Wells-Riley model, and variations thereof\cite{Rudnick2003IndoorAir}, assume a well-mixed airspace, which means that the quanta concentration is the same throughout the room.
-However, the quanta concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
+The Wells-Riley model\cite{Riley1961Book} is widely used to estimate the risk of airborne transmission in a variety of indoor settings\cite{Haddrell2024NatCommun,Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR}, including primary care clinics\cite{Zurcher2022JID,McCreesh2021BMJGlobalHealth}.
+The model estimates the risk of infection based on the number of infectious individuals in space, the generation rate of infectious quanta (doses of pathogen-carrying particles), the breathing rate per person, and the outdoor air supply rate, which is often estimated from indoor CO$_2$ levels\cite{Rudnick2003IndoorAir,Richardson2014PONE,Haddrell2024NatCommun}.
+The Wells-Riley model and its variations\cite{Rudnick2003IndoorAir} assume a well-mixed airspace, meaning that the quanta concentration is the same throughout the room.
+However, this concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
 
@@
 We developed a spatiotemporal extension of the Wells-Riley model to model the concentration of infectious quanta in space and time.
-We combined clinical and person tracking data to determine the spatiotemporal location of TB patients and susceptible attendees at a primary care clinic in South Africa for five days in October/November 2021.
-We also measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time.
-We used Monte Carlo simulation to estimate the risk of infection per clinic attendee.
-Finally, we assessed the impact of infection control measures in place during the study in response to the COVID-19 pandemic (mask use and increased ventilation) by modeling the risk of infection under hypothetical pre-pandemic scenarios without universal mask use and pre-pandemic ventilation conditions in the clinic.
+We combined clinical and person-tracking data to determine the spatiotemporal location of TB patients and susceptible attendees at a primary care clinic in South Africa for five days in October/November 2021.
+We measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time and estimated the risk of infection per clinic attendee.
+Finally, we assessed the impact of infection control measures during the study in response to the COVID-19 pandemic.
 
@@
 
-\section{Methods}
+\section*{Methods}
 
-\subsection{Study design}
+\subsection*{Study design and setting}
 
-We build on the design of a pilot study in 2019\cite{Zurcher2022JID}, which was described in detail in a study protocol \cite{Zurcher2020BMJ}.
-We collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (patient movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
-
-\subsection{Study setting}
-
-The primary care clinic offers both TB and HIV services as well as other basic clinical services, Monday through Friday, from 8\,am to 4\,pm.
-The clinic is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
+Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person tracks) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
+The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm.
+It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
 We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
-The TB room was mainly used for the treatment of diagnosed TB patients, but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
+The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
 
@@
     \caption{\textbf{Schematic view of the clinic}.
-Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room, and finally exit the clinic through one of the exits in the waiting room or at the end of the corridor.
-Colored icons show the placement of the video sensors to track patient movements and the devices monitoring CO$_2$ levels.}
+Patients typically enter the clinic through the entrance on the bottom left, register at the reception, then wait in the waiting room or along the corridor until they see a doctor in one of the care rooms or the TB room.
+They exit the clinic through one of the exits in the waiting room or at the end of the corridor.
+Colored icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
     \label{fig:floor-plan}
@@
 
-\subsection{Data}
+\subsection*{Data collection}
 
-\subsubsection{Environmental data}
-
-Four devices monitored indoor CO$_2$ levels (Digital CO$_2$ Monitor Carbon Dioxide Meter XE-2000, XEAST, Guangdong, China) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}).
-CO$_2$ levels were recorded in parts per million (ppm) at 1min intervals.
-Missing values were linearly imputed.
-On October~13, the corridor's CO$_2$ levels were missing and imputed with the waiting room's CO$_2$ levels because their CO$_2$ measurements were similar on the other days.
-
-\subsubsection{Clinical data}
-
+We collected environmental and clinical data and data on the movements of patients in the clinic.
+Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
+On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days.
 We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study.
@@
 We defined infectious patients as individuals who tested bacteriologically positive for TB during the study or who had received TB treatment for less than four weeks.
-Patients on treatment for more than four weeks or who had recently completed treatment were considered neither infectious nor susceptible.
-We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were ultimately not diagnosed with TB.
+Patients on treatment for more than four weeks or with recently completed treatment were considered neither infectious nor susceptible.
+We also defined suspected TB patients as individuals who were tested for TB following symptom screening, but who were not diagnosed with TB.
 
-\subsubsection{Person tracking data}
+We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}).
+The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
+People in the clinic could contribute multiple records if they moved out of the range of a sensor or were not recognized as a person.
+To address this, we created an R Shiny tool (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}.
+We also labeled the tracks to identify the movements of clinic staff.
+\supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks.
 
-We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1~second intervals (\Cref{fig:floor-plan}).
-The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-\supp~Figure~\zref{fig:tracking-examples} shows some sample tracks from our study.
-Note that people in the clinic could contribute multiple tracks if they moved out of the range of a sensor or if they were briefly unrecognized as a person.
-To address these issues, we created an R Shiny tool to manually reconnect interrupted tracks back together (\supp~Text~\zref{sec:setting-and-data}).
-We also labeled the tracks to identify movements of the clinic staff.
+\subsection*{Spatiotemporal modeling approach}
 
-\subsection{The Wells-Riley equation}
-
-Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1978AJE}, which estimates the risk (probability in \%) of infection $P$ as
+Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of infection $P$ as
 \begin{align}
-    P = \frac{D}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
+    P = \frac{C}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
 \end{align}
-where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
+where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
 The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
-Instead, our study considers a mixed airspace with spatial variation in the quanta concentration.
-This is accomplished using patient movement data combined with clinical records to identify the spatiotemporal location of infectious patients who generate quanta.
 
-\subsection{Spatiotemporal modeling approach}
-
-A detailed description of the Wells-Riley model and our spatiotemporal extension is provided in \supp~Text~\zref{sec:spattemp-model}.
-In the following, we describe the main components of our modeling approach, which are summarized in \Cref{fig:modeling-flow}.
+We extended the Wells-Riley model, allowing spatial variation in the quanta concentration by combining patient-tracking data with clinical records.
+This way, the spatiotemporal location of infectious patients who generate quanta could be determined.
+A detailed description of our extension is provided in \supp~Section~\zref{sec:spattemp-model}, together with an illustrative example using simulated data (\supp~Fig.~\zref{fig:toy-example}).
+In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).
 
@@
     \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
-(1)~Clinical and person tracking data were combined to determine the spatiotemporal location of diagnosed (infectious) TB patients.
-(2)~Both diagnosed and randomly selected undiagnosed  TB among clinic attendees generate infectious quanta.
-(3)~Tracking and environmental data were combined to estimate the air change rate.
+(1)~Clinical records and person tracks were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients.
+(2)~Both diagnosed and randomly selected undiagnosed TB patients among clinic attendees generate infectious quanta.
+(3)~Person-tracking and environmental data were combined to estimate the air change rate.
 (4)~Air change rate was related to the speed at which quanta diffuses indoors.
@@
 
-(1)~We matched clinical and patient movement data to identify the spatiotemporal location of infectious TB patients within the clinic.
-We recorded the timestamp when the patient movement was detected in the registration area and matched this with the registration time in the clinical records.
-We excluded patient movements of less than five minutes and with less than five seconds in the registration area from the matching.
-Data were matched if there was a patient movement in the registration area followed by a entry in the clinical database, considering a maximum delay of 15 minutes.
-All diagnosed TB patients could be matched with patient movements this way.
-(2)~In addition to diagnosed TB patients, we also consider a number of undiagnosed (subclinical) TB patients.
-Prior studies suggest that the number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we model the number of undiagnosed TB patients with a multinomial distribution based on the counts of the daily number of diagnosed patients in the clinical data in October/November 2021 (\supp~Figure~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four TB patients visiting the clinic.
+(1)~We matched clinical and person-tracking data to identify the spatiotemporal location of infectious TB patients within the clinic.
+We recorded the timestamp when a person was detected in the registration area and matched this with the registration time in the clinical records.
+We excluded tracks with a total duration of less than five minutes in the clinic and a duration of less than five seconds in the registration area from the matching.
+We allowed for a maximum delay of 15\,min between the track in the registration area and the entry in the clinical database.
+All diagnosed TB patients could be matched with person tracks this way.
+
+(2)~In addition to diagnosed TB patients, we also assumed a number of undiagnosed (subclinical) TB patients.
+The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
+Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, which was modeled using a lognormal prior distribution considering two different activity levels (\supp~Figure~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
-Since mask wearing was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
 
@@
 
-(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rates by daytime using a transient mass balance model\cite{Batterman2017IJERPH}.
+(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
+
 (4)~The air change rate was related to the diffusion of quanta using, as an approximation, the empirical relationship between the eddy diffusion coefficient and the air change rate in CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}.
-Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated (\supp~Figure~\zref{fig:toy-example}).
+Since airflow was not monitored in our study, we assumed that the quanta diffused radially from where it had been generated.
 
@@
 
-(5)~Quanta removal occurs through replacement of contaminated indoor air with fresh outdoor air, inactivation of bacteria in airborne particles, and gravitational settling.
+(5)~Quanta removal occurs through the replacement of contaminated indoor air with outdoor air, inactivation of bacteria in airborne particles, and gravitational settling.
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-The bacterial inactivation rate of \emph{Mtb} was modeled using a lognormal prior distribution (\supp~Figure~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-We set the gravitational settling rate to zero, thereby ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+We set the gravitational settling rate to zero, ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
 
@@
 where $N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
-We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the distance between two patient tracks was more or less than 0.25m.
+We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m.
 
-\subsection{Monte Carlo simulation}
+\subsection*{Monte Carlo simulation}
 
-Our setup and simulation approach are described in detail in Text~\zref{sec:estimation}, including a comprehensive review of our modeling assumptions and choices of prior distributions.
-To summarize, we simulated the spatiotemporal quanta concentration separately for the waiting room, corridor, and TB room of the clinic.
-Each room was rasterized into a grid of cubic cells with a squared area of 0.25m$^2$.
-The quanta concentration was modeled during clinic hours from 8am to 4pm and updated every second.
-We used a Monte-Carlo simulation approach to consider uncertainty in key modeling parameters.
-In each simulation, we sampled uncertain modeling parameters from their prior distributions and, based on that, computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
+Simulations are described in detail in \supp~Section~\zref{sec:estimation}.
+In brief, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
+Each room was rasterized into a grid of cubic cells with an area of 0.25\,m$^2$.
+The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
+We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters.
+In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
 
-\subsection{Hypothetical scenarios}
+\subsection*{Hypothetical scenarios}
 
-Our study was set during the COVID-19 pandemic when infection control measures were in place.
-To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie a fixed air change rate of 16 air changes per hour, corresponding to the average in our pre-pandemic pilot study in 2019\cite{Zurcher2022JID}.
-Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~instead of undiagnosed TB patients, we considered patients clinically suspected of having TB as infectious, because these patients may still generate aerosolized \emph{Mtb} even when they were not diagnosed with TB at first presentation\cite{Patterson2024PNAS}; 2)~randomly selecting infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}, thereby pretending that the clinical data were not available.
-Except for the scenario-specific modifications, all other modeling assumptions and parameters were the same in each Monte Carlo simulation, ensuring comparability of the results.
+Data were collected during the COVID-19 pandemic when infection control measures were in place.
+To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask-wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie assuming an air change rate of 6~air changes per hour, corresponding to the average rate in the pilot study in 2019 in the same clinic\cite{Zurcher2022JID}.
+Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
+Except for the scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.
 
-\subsection{Statistical analysis}
+\subsection*{Statistical analysis}
 
-Clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and patient movement data (spatiotemporal number of tracks, visit time, and close-contact encounters) were assessed descriptively and summarized with the median and interquartile range (IQR).
+We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
 Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8am to 12am, afternoon: 12am to 4pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).
-A Bayesian Beta regression model was used to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact, and frequency of close-contact encounters.
-All analyses were performed in R software (version 4.3.1) and the Bayesian regression model was estimated in the probabilistic programming language Stan (version 2.26.1).
+The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).
+The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
+We used a Bayesian Beta regression model to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact with, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
+Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1).
+Additional results not reported in the manuscript are shown in \supp~Section~\zref{sec:additional-results}.
 
-\subsection{Ethics statement}
+\subsection*{Ethics statement}
 
-The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 228/2019), the City of Cape Town (Project ID: 8139), South Africa, and the Ethics Committee of the Canton of Bern (KEK/REF: 2019-02131), Switzerland, approved the study.
+The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 536/2020), the City of Cape Town (Project ID: 8139), South Africa, and the Ethics Committee of the Canton of Bern (KEK/REF: 2019-02131), Switzerland, approved the study.
 
@@
 
-\section{Results}
+\section*{Results}
 
-During one week corresponding to five open clinic days, 894 patients were registered in the clinical database, of whom four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening (\supp~Figure~\zref{fig:diagnosed-tb-patients}).
-Overall, 1,563 unique movements were detected in the clinic after processing the video sensor data, of which 125 were labeled as healthcare worker, identifying 1,438 clinic attendees (patients or other visitors).
+During one week (five open clinic days), 894 patients were registered in the clinical database: four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening.
+Overall, 1,563 person tracks were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors).
 Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
-Clinic attendees spent about half an hour in the clinic (median 25\,min, IQR 13\,min$-$46\,min), most of it in the waiting room (\Cref{fig:input-data-descriptives}c).
-Close-contact encounters (proximity <1\,m and duration >1\,min) were frequent (median 5, IQR 2$-$9) and clinic attendees were in close contact with at least one other person for more than half of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm (\supp~Figure~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 29 air changes per hour (IQR 9\,h$^{-1}$$-$36\,h$^{-1}$) in the densely occupied waiting room.
+Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c).
+Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
     \includegraphics{results/data/air-changes-per-hour.png}
-    \caption{\textbf{Patient movements and environmental conditions in the clinic.} Across clinic days and for the waiting room, corridor and TB room of the clinic: \textbf{(a)}~total number of tracks recorded per unit space, \textbf{(b)}~number of people in the clinic over time, \textbf{(c)}~time per clinic visit, and \textbf{(d)}~air change rates.
-Lines and ribbons show the mean plus/minus standard deviation, respectively.
-Boxplots show the median as lines, IQR as boxes, range as whiskers, and outliers as dots.}
+    \caption{\textbf{Person tracks and environmental conditions by clinic area.} Across clinic days and for the clinic waiting room, corridor and TB room: \textbf{(a)}~total number of tracks recorded per unit space, \textbf{(b)}~number of people in the clinic over time, \textbf{(c)}~time per clinic visit, and \textbf{(d)}~air change rates.
+Lines and ribbons show the mean plus/minus standard deviation.
+Boxplots show the medians as lines, interquartile ranges as boxes, ranges as whiskers, and outliers as dots.}
     \label{fig:input-data-descriptives}
@@
 
-\supp~File~1 shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
-It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
-The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
-Across clinic days, the quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
-An exposure of one hour to the mean quanta concentration in the middle of the waiting room would correspond to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
-The mean risk of infection per clinic visit was 0.06\% (80\%-CrI 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
-To put this into perspective, a patient with twelve hourly visits per year would have a cumulative annual risk of infection of 1.37\% (80\%-CrI 0.14\%$-$2.74\%).
+The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translated to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
+The mean risk of infection per clinic visit was 0.06\% (80\%-credible interval (CrI) 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
 
-Our modeling approach allowed us to estimate the individual risk of infection with uncertainty, for example, to identify patients with a comparably higher mean risk of infection and low uncertainty (\supp~Figure\zref{fig:patient-risk-focus}).
 The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
-When the number of close contacts doubled, the odds of infection increased to 1.17 (95\%-CrI 1.04$-$1.30).
-Similarly, doubling the time spent in the clinic increased the odds of infection to 1.09 (95\%-CrI 1.01$-$1.18).
-The time spent in close contact was not significantly associated with the individual risk of infection (1.05, 95\%-CrI, 0.96$-$1.14).
-Finally, a comparison of our spatiotemporal modeling results with those of a temporal model assuming a well-mixed airspace showed a notable difference in the individual risk of infection with a median of 8\% (IQR 4\%$-$15\%).
+A doubling of close-contact encounters increased the odds of infection by 17\% (OR 1.17, 95\%-CrI 1.04$-$1.30).
+Doubling the time spent in the clinic increased the odds of infection by 9\% (OR 1.09, 95\%-CrI 1.01$-$1.18).
+The time spent in close contact was not significantly associated with the individual risk of infection (OR 1.05, 95\%-CrI, 0.96$-$1.14).
 
@@
     \includegraphics{results/modeling/mean-quanta-concentration.png}
-    \caption{\textbf{Spatiotemporal quanta concentration in the clinic.} Model-estimated average quanta concentration across clinic days in the morning and afternoon in the waiting room, corridor and TB room of the clinic.}
+    \caption{\textbf{Spatiotemporal quanta concentration by clinic area.} Model-estimated average quanta concentrations during clinic days in the morning and afternoon in the clinic's waiting room, corridor and TB room of the clinic.}
     \label{fig:main-modeling-results}
@@
 
-To assess the impact of infection control measures during the COVID-19 pandemic, we ran hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates (\Cref{fig:scenario-results}).
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.07\% (80\%-CrI 0.01\%$-$0.09\%), a 1.3-fold increase (80\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been (0.10\%, 80\%-CrI 0.01\%$-$0.09\%), a 1.7-fold increase (80\%-CrI 0.5$-$5.3).
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been (0.13\%, 95\%-CrI 0.01\%$-$0.13\%), a 2.5-fold increase (80\%-CrI 0.6$-$8.8).
-The scenarios show an increase in the number of clinic attendees with a high risk of infection, suggesting that universal use of mask and better ventilation in the clinic had a considerable impact on the tail of the risk distribution, which can also be seen in the smaller upper limit of the credible intervals.
+The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.07\% (80\%-CrI 0.01\%$-$0.09\%), a 30\% increase (OR 1.3, 80\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.10\% (80\%-CrI 0.01\%$-$0.09\%), a 70\% increase (OR 1.7, 80\%-CrI 0.5$-$5.3).
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.12\% (80\%-CrI 0.01\%$-$0.12\%), a 120\% increase (OR 2.2, 80\%-CrI 0.6$-$6.7).
+The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
 
-To assess the impact of modeling assumptions, we ran scenarios making different assumptions about the number of infectious people in the clinic.
-Assuming the number of infectious TB patients had been in proportion to the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.08\%, 80\%-CrI 0.02\%$-$0.13\%).
-Assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean and tail risk of infection would have been considerably higher (0.14\%, 80\%-CrI <0.01\%$-$0.26\%).
+In sensitivity analyses, we varied assumptions about the number of infectious people in the clinic.
+Assuming the number of infectious TB patients reflects the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.08\%, 80\%-CrI 0.02\%$-$0.13\%).
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean risk of infection would have been considerably higher (0.14\%, 80\%-CrI <0.01\%$-$0.26\%).
 
@@
     \includegraphics{results/modeling/mean-roi-comparison.png}
-    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic visit: \textbf{(a)}~with mask use and high ventilation as experienced during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as experienced before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
-Low: <5\%, Medium: 5-10\%, High: >10\% risk of infection.}
+    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic attendee (median visit time 25\,min, IQR 13\,min$-$46\,min): \textbf{(a)}~with mask use and high ventilation as observed during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as observed before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
+Low risk: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}
     \label{fig:scenario-results}
 \end{figure}
+
+Finally, \supp~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
+It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
+The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
 % summary
-We developed a spatiotemporal model to estimate the indoor concentration of infectious quanta and consider patient movement data from a South African primary care clinic in 2021 to accurately simulate the risk of Mtb infection for clinic attendees.
-We estimated an average risk of infection of 0.059\% (95\%-CrI 0.001\%$-$0.402\%), which would correspond to a cumulative annual risk of 0.7\% (95\%-CrI 0.01\%$-$4.72\%) if each patient visited the clinic once per month.
-Our model can identify patients with a higher risk of infection and low uncertainty.
-We also found that clinic attendees who stayed longer in the clinic and had more close-contact encounters  had a higher risk of infection, and we estimated the potential impact of infection control intervention.
-The modeled risk of infection would have been higher had infection control measures not been implemented during the COVID-19 pandemic such as compulsory face masks and improved ventilation conditions.
-Our spatiotemporal model can be readily adapted to model to other airborne transmission risk of other respiratory infections such as SARS-CoV-2.
+We expanded the Wells-Riley equation to a spatiotemporal model based on environmental, clinical, and patient movement data.
+We applied the model to data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
+We estimated an average risk of infection per clinic attendee of 0.06\%.
+We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
+Without infection control measures (masks and high natural ventilation), the risk of infection would have been more than twice as high.
+Our spatiotemporal model can be readily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
+
+% close contacts
+The COVID-19 pandemic has renewed interest in airborne transmission of respiratory infections in crowded indoor settings\cite{Wang2021Science,Morawska2021,Greenhalgh2022BMJ} and its modeling\cite{Shen2021STBE,Buonanno2020EI,Edwards2024RiskAnalysis}.
+Many studies\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR} rely on the Wells-Riley transmission model\cite{Riley1961Book}, based on the pioneering work of William Wells, who proposed a role of droplets in airborne transmission in the 1930s\cite{Wells1934AJE}, and Richard Riley, who developed the equation in the 1960s\cite{Riley1961Book}.
+The Wells-Riley model assumes that the airspace is well-mixed, ignoring that closer proximity to infectious individuals most likely increases the risk of infection.
+Our spatiotemporal model is an extension of the Wells-Riley model, which allows the quanta concentration to vary both over time and space, and considers that the density of infectious particles is initially higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
+We found a significant association between the modeled risk of infection and the number of close-contact encounters, and the time spent in the clinic.
+More close contacts and extended visits make exposure to high doses of infectious particles more likely.
+Several previous studies have already indicated that prolonged close contact may be required to transmit respiratory infections\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
 
 % critical appraisal of our results 
-South Africa has a high burden of TB\cite{WHO2022TBReport}.
-Despite that, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
-In our previous pre-pandemic study before the COVID-19 pandemic, we estimated a cumulative annual risk of \emph{Mtb} transmission of between 9\% and 29\% for a patient coming to the clinic once per month for one hour.
-Although the assumed visit time is longer than what we observed in this study, the risk difference is still noteworthy and most likely due to the infection control measures that were in place during the SARS-CoV-2 pandemic.
-We estimated that without mask wearing or with pre-pandemic ventilation rates, the risk of infection would on average have been 1.28-fold or 1.74-fold higher, respectively.
-But mask wearing and natural ventilation were not the only measures in place during the pandemic, patient triage and fixed appointments probably also contributed to a further reduction in the risk of \emph{Mtb} transmission in our study.
-Finally, the risk difference could, to a small extent, be related to other factors.
-The risk estimate before the COVID-19 pandemic was based on TB prevalence in the South African population.
-We also estimated a higher mean risk of infection when modeling the risk based on TB prevalence or when considering both diagnosed and suspected TB patients as infectious.
+South Africa has a high burden of TB and multi-drug-resistant TB\cite{WHO2023TBReport}.
+Nevertheless, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
+Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 1.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
+The difference can be partly attributed to the infection control measures that were in place during the SARS-CoV-2 pandemic.
+We estimated that without mask use and with pre-pandemic ventilation conditions, the risk of infection would have been more than twice as high.
+Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic.
+Further, different modeling assumptions could have contributed.
+The risk estimate from our previous study before the COVID-19 pandemic\cite{Zurcher2022JID} was based on TB prevalence in the South African population, rather than the TB status documented in the clinical data.
+Modeling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modeled risk of transmission.
 
 % relation to infection control
-A previous modeling study estimated that between 4\% and 14\% of TB transmission in adults in a high TB burden, high HIV prevalence community in South Africa occurred in primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
-Further studies suggest that compulsory face mask wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
-Our findings also suggest that the risk of infection is lower with mask wearing, although the risk reduction estimated in our study is less striking.
-The reason is that the effect of mask wearing is dominated by the impact of very high ventilation rates, which determine both the diffusion and removal of infectious quanta in the indoor space.
-The air change rates in the clinic typically ranged from 10 to 40 air changes per hour, which result in a quick diffusion and removal of infectious quanta from the indoor air.
-To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.48\,h$^{-1}$), Switzerland (0.71\,h$^{-1}$), and Tanzania (13.74\,h$^{-1}$)\cite{Banholzer2024PGPH}.
-Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, which was largely due to better ventilation, and yet its mean air change rate was less than half the rate measured in our study.
-Thus, we note that our study results do \emph{not} indicate that masks are less effective as found in other studies, rather their effectiveness is overshadowed by excellent ventilation conditions in the clinic during the COVID-19 pandemic.
-
-% close contacts
-Our spatiotemporal model can be viewed as a modified version of the Wells-Riley transmission model\cite{Riley1978AJE}, which assumes a well-mixed airspace.
-By contrast, our model allows the quanta concentration to vary both over time and space, considering that density of infectious particles is initially higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
-The exposure to infectious quanta therefore depends on the proximity to infectious individuals, which aligns with previous findings suggesting that prolonged close contact may be required for transmission of respiratory infections\cite{Leung2020NatMed,Brankston2007LancetID,Narasimhan2013PulmonaryMed}.
-In line with this, our modeled risk of infection was significantly associated with the number of close-contact encounters and the time spent in the clinic.
-This is because more close contacts and longer visits make exposure to high doses of infectious particles more likely.
-However, we note that this association is based on modeled results and it is difficult to definitively prove close contact respiratory transmission.
-Empirical studies investigating the potential link between contact data and respiratory transmission are so far scant\cite{Voirin2015ICHE,Vanhems2013PONE}.
+A previous study estimated that between 4\% and 14\% of TB transmission in adults in a high TB burden, high HIV prevalence community in South Africa occurred in primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+Other studies suggest that mask-wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less striking.
+The reason is that the effect of mask-wearing is compounded by the impact of very high ventilation rates, which determine both the diffusion and removal of infectious quanta in the indoor space.
+The air change rates in the clinic typically ranged from 10 to 20 air changes per hour, which resulted in a quick diffusion and removal of infectious quanta from the indoor air.
+To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
+Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely due to a better ventilation, with an air change rate comparable to the one measured in the primary care clinic of our study.
+We stress that our study results do \emph{not} suggest that masks are ineffective but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 
 % more complex models and our relation
-Airborne respiratory transmission depends on common environmental and patient-specific factors.
-Ventilation conditions and patient movements were considered by our model.
-However, other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see \supp~Text~\zref{sec:depth-discussion} for a detailed discussion of these factors).
-For example, poor ventilation settings can impede the removal of infectious particles\cite{Li2021BuildEnv}, and computational fluid dynamics (CFD) models have been developed to consider airflow as an important factor influencing the spatial spread of airborne pathogens\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
-However, CFD models are computationally expensive and primarily used in experimental settings and static environments, which makes their application to dynamic settings such as ours difficult.
-Moreover, our modeling approach necessitates computational efficiency because a large number of Monte Carlo simulations is required to reflect uncertainty in important modeling parameters.
+Airborne respiratory transmission depends on environmental and patient-specific factors.
+Our model considered ventilation conditions and patient movements.
+However, other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see \supp~Section~\zref{sec:depth-discussion} for a detailed discussion of these factors).
+Furthermore, computational fluid dynamics (CFD) models have been developed to examine airflow as a factor influencing the spatial spread of airborne pathogens\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
+However, CFD models are computationally expensive and primarily used in experimental and static environments, making their application to real-world settings such as clinics difficult.
+Moreover, our modeling approach required computational efficiency because a many Monte Carlo simulations are needed to reflect uncertainty in modeling parameters.
 
 % limitations
-Our modeling study has several limitations.
-First, tracking patient movements with video sensors is challenging and required considerable data processing and many patient movements were interrupted and could not be connected back together.
+Our study has several limitations.
+First, tracking patient movements with video sensors was challenging.
+It required considerable data processing and many tracks were interrupted and could not be reconnected.
 Therefore, some clinic attendees may have had longer visits and the risk of infection may have been underestimated.
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-Uncertainty in these modeling parameters was often large and could potentially be reduced in future workby incorporating , for example, additional information on patient characteristics from clinical records that are known to be associated with patient-specific infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
+Contrary to our assumptions, infectious particles may not diffuse radially from the source or approach a steady-state with uniform contamination of the airspace, although we have observed this behavior in experiments (see Text~A and Figure~S1 in S1~Appendix of Ref.
+\cite{Banholzer2023PLoSMed}).
+Furthermore, uncertainty in modeling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
 Third, we could not validate our modeling results with empirical findings.
-Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be infected.
+Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be diagnosed with TB.
 
 % conclusions
-In conclusion, our novel spatio-temporal transmission model study using clinical, environmental, and patient movement data showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa under pandemic conditions with strict infection control measures in place.
-The modeled risk of infection was significantly associated with the number of close-contact encounters by clinic attendees during their visit.
-Our spatiotemporal model could thus be used to assess the impact of interventions targeting patient flows that aim to reduce the number of close-contact encounters.
+In conclusion, our spatiotemporal transmission model considers clinical, environmental, and movement data to estimate the risk of infection per individual in the indoor space. We
+showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa was related to the number of close contact encounters and time spent in the clinic.
+The risk of infection was low under pandemic conditions with strict infection control measures in place.
+Our spatiotemporal model could be used in future work to assess the impact of infection control measures, specifically interventions targeting patient flows to reduce prolonged close contact.
+Finally, our study underscores the paramount importance of effective ventilation in primary care clinical settings.
 
@@
 \section*{Acknowledgements}
-...
+We are grateful to the clinical manager and the staff members at the primary care clinic.
+We also would like to thank the City of Cape Town, South Africa for the use of one of their clinic facilities and their support.
+Our research findings and recommendations do not represent an official view of the city of Cape Town.
+
+\section*{Data availability}
+Preprocessed data and code are available from GitHub via \url{https://github.com/nbanho/clinic-transmission}.
+
+\section*{Funding}
+The study was supported by a grant from the Swiss National Science Foundation (grant no.
+CRSK-3\textunderscore190781).
+NB and LF are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through grant no.
+5U01-AI069924-05.
+ME is supported by special project funding from the Swiss National Science Foundation [grant no.
+32FP30-189498].
+The funders had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
+
+\section*{Author contributions}
+xxx.
+
+\newpage
 

### 2024-05-16 doc/paper/supplements.tex

--- 2024-04-10 doc/paper/supplements.tex
@@
 
-\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
-Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}}
+\title{\LARGE\singlespacing{\textbf{Supplementary Information} \\ \medskip
+A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
 
@@
 
-\section{Processing of tracking data}\label{sec:setting-and-data}
+\section{Processing of person-tracking data}\label{sec:setting-and-data}
 
-Tracking data required processing before it could be fully used for modeling.
-The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database.
-Visitors of the clinic most likely contributed multiple tracks for several reasons:
-\begin{itemize}
-    \item A person moved outside the range of the video sensors and then re-entered. 
-    \item An individual may temporarily not have been recognized as a person by the tracking software because the person bowed downwards.
-    \item A person may have been temporarily hidden behind another person.
-    \item Moving objects such as opening doors may have been recognized as a person, introducing noise.
-\end{itemize}
-We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person.
-In the first step, we filtered noise (\ie IDs of less than 10\,s and a total moving distance of less than 2\,m), and then automatically matched the remaining tracks that were less than 10\,s and 3\,m apart while moving or less than 60\,s and 0.5\,m apart while sitting.
-In the second step, we build on the automatically matched tracking data and, through manual inspection, linked tracks that were further apart but still most likely of the same person.
-To facilitate the manual linkage, we developed a Shiny application and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) that assisted the students who were performing this linkage.
+Person-tracking data required processing because many tracks were interrupted, for example, if the person i)~moved outside the range of the video sensors and then re-entered, ii)~bowed heavily downwards, or iii)~was temporarily hidden behind another person or obstacle.
+In addition, moving objects such as opening doors may have been recognized as a person, introducing many extremely short tracks (total duration of less than 10\,s and distance less than 2\,m), which were removed.
+
+We reconnected tracks that most likely belonged to the same person.
+First, we automatically matched tracks that were less than 10\,s and 3\,m apart while moving or less than 60\,s and 0.5\,m apart while sitting.
+Second, we manually reconnected the remaining tracks if they were likely from the same person.
+For that purpose, we developed a Shiny app and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) to assist the students performing the matching.
 
@@
 Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
-Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track.
-The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey.
+Tracks are shown as lines with a open and crossed circles indicating the tracks' start and end, respectively.
+The currently selected track is shown in black, possible links in colors, and alternatives for the selected possible link in light grey.
 
-Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link.
+Below the floor plan, a table is showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID of the selected track (Pid) and the observation ID of possible links (Oid), (2)~the height difference at the end point of Pid and the start point of Oid, (3)~ the standing height (\ie maximum height during the entire track) difference, (4)~the duration of Oid, (5)~the time difference, and (6)~the distance between the end point of Pid and the start of point of Oid.
 Possible links are sorted in ascending order by the time difference.
 
-In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information: the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages.
-It then shows the current Pid with shortcuts to move to the next or previous Pid, or the next or previous Pid at the entrance.
+The Shiny app's left panel shows descriptive information about the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages.
+It further shows the current Pid with shortcuts to move to the next or previous Pid, or the next or previous Pid at the entrance.
 Linkages can be done either forward or backward in time.
-The duration of the current Pid (time in the clinic) is shown together with the number of linked raw tracks of which the current Pid consists.
-The sliding bars allow filtering of possible links according to difference in time, distance, or height, with pre-set filters above to quickly switch between common thresholds.
+The total duration of the Pid track (time in the clinic) is shown together with the number of connected tracks within Pid.
+The sliding bars allow filtering of possible links (Oid) according to difference in time, distance, or height, with quick filters to switch between common thresholds.
 The first dropdown menu selects the Oid for which alternative matches should be shown in the floor plan.
-The idea of this feature is to prevent false linkages when there are more plausible alternatives.
-A linkage is made by selecting the Oid in the "Link with IDs" dropdown menu and clicking "Link IDs".
-After that, the app updates the floor plan, table, and information in the panel.
-If an erroneous link was made, it can be revoked by clicking "Unlink last ID" if the last link in time should be revoked or by clicking "Unlink first ID" if the first link in time should be revoked.
-If no further linkages could be made, the track can be finished by entering a label (see below).
-The manually linked dataset is automatically saved to the file shown in the bottom of the left panel.
+The idea is to prevent false linkages when there are more plausible alternatives for the Oid.
+A connection is made by selecting the Oid in the "Link with IDs" dropdown menu and then clicking "Link IDs".
+After that, the app updates the floor plan, matching table, and descriptive information.
+Erroneous link can be revoked by clicking "Unlink last ID", if the last link in time should be revoked, or by clicking "Unlink first ID", if the first link in time should be revoked.
+If the Pid could not be connected further, its track can be finished by entering a label and clicking "Label/End track".
+The person-tracking data is then updated with the reconnected track and saved automatically to the file shown at the bottom.
 
-We created labels to denote finished tracks: 
+We created the following labels to denote finished tracks: 
 \begin{itemize}
@@
 
-After manual processing of the tracking data, we arrived at 3,319 tracks.
-We then filtered noise and short tracks <5\,min to obtain 1,563 used for analysis: 209 (13\%) "Clean", 125 (8\%) "Staff", 213 (14\%) "Lost enter", 246 (16\%) "Lost exit", and 770 (49\%) "Lost both".
-A sample of our tracks used for analysis, excluding staff, are shown in \Cref{fig:tracking-examples}.
+We obtained 3,319 tracks after preprocessing.
+Further exclusion of short tracks <5\,min resulted in 1,563 tracks used for analysis: 209 (13\%) labeled as "Clean", 125 (8\%) as "Staff", 213 (14\%) as "Lost enter", 246 (16\%) as "Lost exit", and 770 (49\%) as "Lost both".
+A sample of person tracks are shown in \Cref{fig:tracking-examples}.
 
@@
     \includegraphics{results/data/example-patient-tracks.png}
-    \caption{Sample tracks from our video sensors of this study.}
+    \caption{Sample of person tracks from our video sensors after preprocessing.}
     \label{fig:tracking-examples}
@@
 
-\section{Spatiotemporal model}\label{sec:spattemp-model}
+\section{Spatiotemporal modeling approach}\label{sec:spattemp-model}
 
@@
 
-We build upon the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection.
+We build upon the Wells-Riley model\cite{Riley1961Book}, which is an established epidemiological model to estimate the risk of infection.
 The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation:
 \begin{align}
-    P = \frac{D}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
+    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
 \end{align}
-where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
+where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 
@@
 While considering the stochastic behaviour of airborne transmission, the Wells-Riley model makes two simplifying assumptions.
-First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and evenly in the indoor airspace.
+First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and uniformly in the indoor airspace.
 In other words, the exposure to infectious quanta is the same regardless of the location of the infectious and susceptible individual.
 Second, the Wells-Riley model assumes steady-state conditions, which means that the quanta concentration and the outdoor air supply rate are constant over time.
-In other words, the risk of infection for each susceptible only varies with the duration of exposure (the time they spent in the indoor space).
+In other words, the risk of infection for each susceptible only varies with the duration of exposure (the time spent in the indoor space).
 
@@
 
-While the extension by Rudnick and Milton relaxes the steady-state assumption, it still assumes a well-mixed airspace.
-We relax this assumption by modeling the concentration of infectious quanta over space time.
-Then, we compute the cumulative risk of infection for each susceptible individual over space and time as
+The extension by Rudnick and Milton still assumes a well-mixed airspace.
+We relax this assumption by modeling spatial variation in the concentration of infectious quanta.
+We then compute the risk of infection for susceptible individuals using the Wells-Riley equation considering their cumulative exposure to infectious quanta over space and time as
 \begin{equation}\label{eq:spattemp-P}
@@
 \end{equation}
-where $N$ is the quanta concentration in cell airspace $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in cell $c$ at time $t$, and $p$ is the breathing rate.
+where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate.
 
-Modeling the spatiotemporal quanta concentration involves three processes
+Modeling of the spatiotemporal quanta concentration can be divided into three processes
 \begin{itemize}
-    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Quanta is generated at the infectious individual's location. 
-    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Quanta diffuses until we approach a well-mixed airspace. 
-    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Quanta is removed from the air or becomes inactive (not infectious). 
+    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Infectious particles are generated at the infectious individual's location. 
+    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Pathogen-carrying particles spread in the indoor space. 
+    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Contaminated indoor air is replaced or pathogens become inactive. 
 \end{itemize}
@@
 
-We divide the area of the airspace in a 2-dimensional grid of cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$.
-The grid provides a discrete approximation to the continuously varying quanta concentration across the space.
-The number of grid cells $C = n_x \cdot n_y$ determines the spatial resolution and computation time, \ie more cells increase resolution at the expense of more computation time.
-We model the quanta concentration at time $t = 1, \dots T$ (in seconds).
-
-We define the following variables and modeling parameters:
+We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$.
+The grid provides a discrete approximation to the continuously varying quanta concentration in the indoor space, with the number of grid cells $C = n_x \cdot n_y$ determining the spatial resolution.
+We denote the following variables and modeling parameters:
 \begin{itemize}
-    \item $N_{c,t}$: quanta concentration in airspace unit $c$ at time $t$
+    \item $N_{c,t}$: quanta concentration in airspace unit $c = 1, \dots, C$ at time $t = 1, \dots T$ (in seconds)
     \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
@@
     \item $D$: diffusion constant (m$^2$ s$^{-1}$)
-    \item $AER$: air change rate (air changes s$^{-1}$ / quanta s$^{-1}$)
+    \item $AER$: air change rate (air changes s$^{-1}$)
     \item $V$: volume of the airspace (m$^3$)
-    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air (ppm)
+    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air (parts per million, ppm)
     \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air (ppm)
     \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
-    \item $n_t$: number of individuals in airspace at time $t$
+    \item $n_t$: number of individuals in the airspace at time $t$
     \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
@@
 \end{align}
-For now, we assume that the initially generated quanta is confined to the cell where the infectious individuals are located, which is not an unreasonable assumption.
-The concentration of virus-laden aerosols is typically highest near the infectious source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}.
-Although coughing and sneezing can expel the aerosols further away from the individual, these activities are much less frequent than breathing, which may thus contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}.
+We assume that the initially generated quanta is confined to the infectious individual's cell, reflecting previous findings that the concentration of pathogen-laden aerosols is typically highest near the infectious source\cite{Wang2021Science,Morawska2021,Vuorinen2020SafSci,Chen2020BuildEnv}.
+Although coughing and sneezing can expel the aerosols further away from the individual, these activities are much less frequent than breathing, which may thus contribute more infectious particles\cite{Dinkele2022AJRCCM}.
 
@@
 
-Absent measurements of airflow, we assume that aerosols diffuse uniformly in the indoor air, approaching a well-mixed airspace over time.
-We assume that the quanta disperse radially in (x,y)-direction towards locations where the concentration is lower.
-The diffusion equation is
+Absent measurements of airflow, we assume that aerosols diffuse uniformly in the indoor air, approaching a well-mixed airspace over time (the steady state).
+We assume that the quanta diffuse radially in (x,y)-direction towards locations where the concentration is lower.
+The diffusion equation is given by
 \begin{align}\label{eq:diffusion}
@@
 \end{align}
-where $V$ is the volume of the airspace (in m$^3$). While this is only an approximation, it nicely relates the air change rate to quanta diffusion.
+where $V$ is the volume of the airspace (in m$^3$).
+We thus assume that pathogen-carrying particles spread with a similar speed than CO$_2$ particles, which likely overestimates the diffusion constant, because CO$_2$ particles are smaller than \emph{Mtb}-carrying particles.
 
-The $AER$ is not directly observed, but can be estimated from indoor CO$_2$ levels, room occupancy, and room volume.
-Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
+The $AER$ is not directly observed but can be estimated from indoor CO$_2$ levels, room occupancy, and volume.
+Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state concentration), the $AER$ can be computed as
 \begin{align}
@@
 However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration.
-Therefore, we determine $AER$ using a transient mass balance model\cite{Batterman2017IJERPH}
+Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
 \begin{align}
@@
 The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
-We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [300,600]$\cite{Batterman2017IJERPH}.
-We can compute $AER$ for different time windows, hence $AER_t$.
+We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints on $CO^{\text{O}} = [300,600]$\cite{Batterman2017IJERPH}.
+We can compute $AER$ over different time windows, hence the index in $AER_t$.
 
@@
 Infectious quanta may be removed from the air through outdoor air exchange, bacterial inactivation, or gravitational settling.
-We will ignore gravitational settling because \emph{Mtb} is carried in small particles between 2\,$\mu$m and 5\,$\mu$m\cite{Fennelly2020Lancet}, which become airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+We will ignore gravitational settling because \emph{Mtb} is carried in small particles between 1\,$\mu$m and 7\,$\mu$m\cite{Fennelly2020Lancet}, which become airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
 The removal rate is thus the sum of the outdoor air exchange rate $AER_t$ and the bacterial inactivation rate $\lambda$ for \emph{Mtb}, \ie
@@
 
-Considering the generation (\Cref{eq:generation}), diffusion (\Cref{eq:diffusion}), and removal (\Cref{eq:removal}) of quanta together, the quanta concentration at time $t$ in the airspace is computed as
+Combining \Cref{eq:generation}, \Cref{eq:diffusion}, and \Cref{eq:removal}, the quanta concentration at time $t$ in the indoor space can be computed as
 \begin{align}\label{eq:spattemp-N}
@@
 We illustrate our modeling approach with an example.
-Consider a room with a volume $V$=150m$^3$ (length = 10m, breadth = 5m, height = 3m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$ cells with an area of 0.25m$^2$.
+Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of 0.25\,m$^2$.
 We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
-We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle, \ie cell (20,10), which means that the initially generated quanta is $(x_c = 20\pm1, y_c = 10\pm1)$, \ie $N_{c,t} = \frac{1}{9} \cdot I \cdot q$ in cells $c \in (x_s = 20\pm1, y_s = 10\pm1)$, and zero in all other cells.
-We further assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$, a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$ (removal rate of 1\,quanta h$^{-1}$ and corresponding to a diffusion coefficient of 0.007\,m$^2$ s$^{-1}$), and a bacterial inactivation rate of $\lambda = 0.5$\,quanta h$^{-1}$.
+We monitor the room from 8\,am to 12\,am and consider one infectious individual generating quanta continuously for one hour between 8\,am and 9\,am in the middle of the room (cell (20,10)), which means that quanta is generated in cells $(x_c = 20\pm1, y_c = 10\pm1)$.
+We assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0.5$\,quanta h$^{-1}$.
+We further assume a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$, translating into a quanta removal rate of $1 + 0.5 = 1.5$\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.007\,m$^2$ s$^{-1}$.
 
 \Cref{fig:toy-example} shows the quanta concentration at different time points.
-The quanta concentration increases over time and is higher near the infectious individual.
-After the individual has left the space, the quanta diffuses fairly quickly and to approach a well-mixed already at 9:05am.
-Over time, the removal through outdoor air exchange and bacterial inactivation brings the concentration in the room back towards zero.
+The quanta concentration increases over time and is higher near the infectious individual in the room's center.
+After the individual has left the room, the quanta diffuses fairly quickly and a well-mixed airspace is already at 9:05\,am.
+Over time, the removal through outdoor air exchange and bacterial inactivation brings the concentration in the entire room back towards zero.
 
@@
     \includegraphics{tests/stm_v2-toy_example.png}
-    \caption{Spatiotemporal quanta concentration between 8am and 10am following the generation of quanta by an infectious individual between 8am and 9am in the middle of the room.}
+    \caption{Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's center.}
     \label{fig:toy-example}
@@
 
-\section{Simulation approach}\label{sec:estimation}
+\section{Simulations}\label{sec:estimation}
 
-\subsection{Setup}\label{prep:building}
+\subsection{Setup and setting}\label{prep:building}
 
-We divide the clinic into three spaces: the waiting room (length = 10.55m $\times$ breadth = 5.50m $\times$ height = 3.00m), the corridor (length = 12.45m $\times$ breadth = 2.20m $\times$ height = 2.50m), and the TB room (length = 4.75m $\times$ breadth = 3.50m $\times$ height = 3.00m).
-We separate the waiting room and the corridor because separate rectangular rooms simplify modeling and because we observe small differences in the time-varying CO$_2$ levels, suggesting these rooms are somewhat different.
-However, we note that the doors to the corridor were often most of the time, so that the corridor and waiting room were not completely enclosed.
-As such, it is possible that aerosolized \emph{Mtb} (quanta) spreads from the waiting room to the corridor, and vice versa, but this is not considered in our modeling approach.
-We rasterize each room into a grid of cubic cells with a squared area of 0.25m$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
-Since the rooms have different height, the quanta concentration in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
+We divide the clinic into three spaces: the waiting room (length = 10.55\,m $\times$ breadth = 5.50\,m $\times$ height = 3.00\,m), the corridor (length = 12.45\,m $\times$ breadth = 2.20\,m $\times$ height = 2.50\,m), and the TB room (length = 4.75\,m $\times$ breadth = 3.50\,m $\times$ height = 3.00\,m).
+We rasterize each room into a grid of cubic cells with a squared area of 0.25\,m$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
 
-We model the spatiotemporal quanta concentration during clinic hours from 8am to 4pm and analyze the modeled concentration by daytime, \ie morning (8am to 12am) and afternoon (12am to 4pm).
-Correspondingly, we estimate the outdoor air exchange rate by daytime, thereby considering potential differences in ventilation patterns between the morning and afternoon.
-We update the quanta concentration every second at $t = 1, \dots, T$, which aligns with the frequency of the tracking data.
-We assume that the quanta concentration at $t=0$ is zero anywhere in the airspace, \ie $N_{c,0} = 0 ~~ \forall c$.
-In other words, we assume that all quanta from the previous day has been removed from the air before the start of the following clinical day.
+Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate spaces to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
+Nevertheless, we acknowledge the possibility for aerosolized \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
 
-\subsection{Monte-Carlo simulation}
+We model the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and compare the modeled concentration by daytime, \ie morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm).
+Correspondingly, we estimate the outdoor air exchange rate by daytime, considering potential differences in ventilation conditions between the morning and afternoon.
+We update the quanta concentration every second from $t = 1, \dots, T$, in line with the frequency of the person-tracking data.
+We set the quanta concentration at $t=0$ to zero, \ie $N_{c,0} = 0 ~~ \forall c$, assuming that all quanta from the previous day has been removed from the air before the start of the following clinical day.
 
-We use Monte-Carlo simulation to estimate the risk of infection, thereby considering uncertainty in several modeling parameters. Each simulation consists of the following steps:
+\subsection{Monte Carlo simulation}
+
+We use Monte Carlo simulation to estimate the risk of infection, considering uncertainty in several modeling parameters. Each simulation consists of the following steps:
 \begin{itemize}
     \item[1.] \textbf{Sample uncertain modeling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, the mask reduction rate, and the bacterial inactivation rate.
-    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, compute the spatiotemporal quanta concentration in each cell of the rooms over time.
-    \item[3.] \textbf{Estimate risk of infection}: Based on their spatiotemporal quanta exposure, estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
+    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room per unit space (cell) over time.
+    \item[3.] \textbf{Estimate risk of infection}: Based on their spatiotemporal quanta exposure, we estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
 \end{itemize}
 
-Most clinic attendees could be undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, and bacterial inactivation rate is large.
-To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean or median across these simulations.
+Many clinic attendees are potential undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, and bacterial inactivation rate is large.
+To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean across these simulations.
 
-\subsection{Modeling assumptions and prior distributions}\label{sec:priors}
+\subsection{Modeling assumptions}\label{sec:priors}
 
@@
 
-We model the amount of quanta that is generated and how it diffuses.
-What we have not yet discussed is the initial spread of infectious quanta, referred to as the propagation distance, which is the maximum distance from the source to the end of the exhaled air before dispersion.
-An analysis of exhalation activities suggest a propagation distance of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity (\ie breathing, speaking, coughing).
+The initial spread, referred to as the propagation distance, is the maximum distance from the source to the end of the exhaled air before dispersion.
+An analysis of exhalation activities suggest a propagation distance of 0.7\,m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity (\ie breathing, speaking, coughing).
 Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting (leakage of particles on the sides of the mask) particles\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}.
 Therefore, we assume that mask wearing confines the quanta generation only to the cell where the infectious individual is located.
-In our setup, this corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.
-In a hypothetical scenario, we estimate the risk of infection if mask wearing would not have been compulsory.
-Without masks, we assume double the propagation propagation distance, \ie 0.7m (the diagonal lengths of two cells)\cite{Tang2013PLoSOne}.
-Since we could not observe the flow direction of the initially generated quanta, we assume that all neighbouring cells are affected equally.
-That is, we divide the generated quanta equally among the nine affected cells, \ie the cell where the infectious person is located as well as all first-neighbouring cells.
+In our setup, this corresponds to a propagation distance of 0.35\,m (the diagonal cell length) with an initial spread to an area of 0.25$^2$\,m$^2$.
+Without masks, we assume two times that propagation distance, \ie 0.7\,m (the diagonal lengths of two cells)\cite{Tang2013PLoSOne}.
+Since the direction of propagation is not observed, we assume uniform spread to all neighbouring cells, \ie the generated quanta is distributed evenly among the first-neighbouring cells from where the infectious individual is located.
 
@@
 On the other end, Escombe et al.\cite{Escombe2008PLoSMed} arrived at an average estimate of 8.2\,quanta h$^{-1}$, which, however, could be as high as 226\,quanta h$^{-1}$.
-Lower rates were observed in lower for non-multidrug-resistant (MDR), smear-negative patients and considerably higher for MDR, smear-positive TB patients\cite{Escombe2008PLoSMed}.
-Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM} also estimated a high average rate of 138\,quanta h$^{-1}$ in MDR TB patients.
-Nardell et al.\cite{Nardell1991ARRD} also reported a higher estimate of 13\,quanta h$^{-1}$, but only for a single index case.
+Lower rates were observed in non-multidrug-resistant (MDR), smear-negative patients and considerably higher ones in MDR, smear-positive TB patients\cite{Escombe2008PLoSMed}.
+Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM} also estimated a high average rate of 138\,quanta h$^{-1}$ in MDR TB patients, and Nardell et al.\cite{Nardell1991ARRD} reported an estimate of 13\,quanta h$^{-1}$ for a single index case.
 The wide range of estimates within and between studies suggest that the emitted quanta varies considerably with the infectiousness of \emph{Mtb} patients\cite{Wurie2016BMJ}.
@@
 Here, we follow the same approach to derive a quanta generation rate distribution for sitting and walking activities, assuming 80\% breathing and 20\% speaking during both activities.
-The resulting prior distributions are shown in \Cref{fig:quanta-distribution}.
+The resulting prior distributions are shown in \Cref{fig:quanta-distribution} along with the estimates from the literature.
 The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking.
-The activity is determined based on whether the distance between two patient tracks was more (walking) or less (sitting) than 0.25m.
+The activity is determined based on whether the the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).
 
@@
 McCreesh et al.
-assumed a reduction by masks of 75\% (95\%-confidence interval (56\% to 85\%) based on the findings of Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM}.
-Based on that, we model the reduction by masks with $\mathrm{Beta}(24.9, 8.3)$ distribution.
-We assume that masks were correctly worn the whole time by all clinic attendees and staff during the COVID-19 pandemic.
+assumed a reduction by mask-wearing of 75\% (95\%-confidence interval (56\% to 85\%) based on the findings of Dharmadhikari et al.\cite{Dharmadhikari2012AJRCCM}, which we translate into a $\mathrm{Beta}(24.9, 8.3)$ prior distribution for the mask reduction rate.
+We assume that masks were always correctly worn by all clinic attendees and staff during the study.
 
@@
 The bacterial inactivation rate for \emph{Mtb} has not been precisely estimated.
-Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6h ($\lambda$ = 0.12).
-By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5min ($\lambda$ = 8.3).
-The difference could be explained by differences in humidity levels during the study\cite{Lever2000LettersAppliedMicrobio}: Loudon (1969) studied survival during a relative humidity of about 50\% whereas Lever (2000) studied survival during a relative humidity of about 70\%.
-Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) seems more representative of the our study conditions.
-Moreover, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5 hours ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}.
-In contrast to that, Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30min ($\lambda$ = 3.2), which would be closer to Lever (2000).
-Considering the variation in these estimates, we model $\lambda$ with a $\mathrm{Lognormal}(\mu = \log 1, \sigma = 1)$ distribution, which has a median rate of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$).
+Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6\,h ($\lambda$ = 0.12).
+By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5\,min ($\lambda$ = 8.3).
+The difference could be explained by differences in humidity levels during the study\cite{Lever2000LettersAppliedMicrobio}: Loudon (1969) and Lever (2000) studied survival in an environment with relative humidity of about 50\% and 70\%, respectively.
+Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) may be more suitable for our study.
+Furthermore, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5\,h ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}, and Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30\,min ($\lambda$ = 3.2), which is closer to the estimate by Lever (2000).
+Considering the variation in these estimates, we model $\lambda$ with a $\mathrm{Lognormal}(\mu = \log 1, \sigma = 1)$ distribution, with a median rate of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$).
 The distribution is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
@@
 
-Undiagnosed (subclinical) TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats), or the diagnosis could not be made at first presentation based on existing tests\cite{Patterson2023Preprint}.
+Undiagnosed (subclinical) TB patients may go undetected because they do not show symptoms according to the WHO 4-question symptom screen (cough, fever, weight loss, and night sweats)\cite{Berhanu2023CID}, or the diagnosis could not be made at first presentation based on existing tests\cite{Patterson2024PNAS}.
 Previous studies suggest that the number of undiagnosed (subclinical) TB patients is roughly similar to the number of diagnosed TB patients\cite{Berhanu2023CID,Moyo2022LancetID}.
-Therefore, we model the number of undiagnosed TB patients with a Multinomial distribution based on the counts of the daily number of diagnosed TB patients during our study period from October to November 2021 (\Cref{fig:undiagnosed-distribution}.
-We assume that all clinic attendees can be potential undiagnosed TB patients and sample among them with equal probability.
+Therefore, we model the number of undiagnosed TB patients with a Multinomial distribution based on the counts of the daily number of diagnosed TB patients from October to November 2021 (\Cref{fig:undiagnosed-distribution}.
+We consider all clinic attendees as potential undiagnosed TB patients and sample among them with equal probability.
 We further assume that undiagnosed and diagnosed TB patients have the same quanta generation rate.
@@
 
-Adam et al.\cite{Adams1993} reports breathing (inhalation) rates for different activity levels and by sex.
-We use the average breathing rate for two activity levels: sitting with $p = 0.51$\,m$^3$ h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$ h$^{-1}$.
-Again, the inhalation rate for each susceptible clinic attendee is determined based on whether the distance between two tracks was more (walking) or less (sitting) than 0.25m.
+Adams (1993)\cite{Adams1993} reports breathing rates for different activity levels.
+We use the average breathing rate for the following activity levels: sitting with $p = 0.51$\,m$^3$ h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$ h$^{-1}$.
+
+\clearpage
+
+\section{Additional results}\label{sec:additional-results}
+
+% \begin{figure}[!htpb]
+%     \centering
+%     \includegraphics{results/data/diagnosed-tb-patients.png}
+%     \caption{Number of diagnosed TB patients (dark bars) and those suspected of TB (light bars) following symptom screening.
+Modeled study days in color, other clinic days in October and November in grey.}
+%     \label{fig:diagnosed-tb-patients}
+% \end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/co2-levels-over-time.png}
+    \caption{CO$_2$ levels by clinic area over time for each study day.}
+    \label{fig:co2-levels}
+\end{figure}
+
+% \begin{figure}
+%     \centering
+%     \includegraphics{results/modeling/patient-risk-focus-plot.png}
+%     \caption{Mean and 80\%-CrI (uncertainty) of the risk of infection per clinic visit.
+Highlighted in red is a visit from a clinic attendee with moderate risk of infection and low estimated uncertainty.}
+%     \label{fig:patient-risk-focus}
+% \end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/modeling/mean-roi-assumptions.png}
+    \caption{Mean risk of infection per clinic attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
+Low: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}
+    \label{fig:assumptions-results}
+\end{figure}
 
@@
 
-Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate quanta diffusion and removal.
-The combination of clinical and video sensor data further allows the spatial modeling of quanta generation.
-This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1978AJE,Rudnick2003IndoorAir}.
+Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate the air exchange rate, based on which we model quanta diffusion and removal.
+The combination of clinical and person-tracking data further allows the spatial modeling of quanta generation.
+This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1961Book,Rudnick2003IndoorAir}.
 In our spatiotemporal model, we assume a radial diffusion of quanta, which may not reflect the actual airflow at the primary care clinic.
@@
 For example, measuring airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh outdoor air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered, especially when it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
+However, airflow is rarely considered in modeling because it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
 
@@
 Survival and dispersion of aerosols are further determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
-Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modeling framework via the quanta generation rate parameter\cite{Riley1978AJE,Rudnick2003IndoorAir}.
+Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modeling framework via the quanta generation rate parameter\cite{Riley1961Book,Rudnick2003IndoorAir}.
 It is challenging to consider these properties specifically, especially because they can be modified by environmental factors with often unclear, pathogen-specific effects\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.
@@
 On the other hand, airborne \emph{Mtb} was more likely detected in healthcare facilitates with higher relative humidity\cite{Sornboot2019IJTLD,Matuka2021IJERP}.
+Ambient carbon dioxide levels may further influence aerostability\cite{Haddrell2024NatCommun}.
 
@@
 Infectiousness could be determined based on the most recent sputum test result.
-Sputum smear positive patients are more likely to transmit \emph{Mtb} than sputum smear negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.
+Sputum smear-positive patients are more likely to transmit \emph{Mtb} than sputum smear-negative patients\cite{Shaw1954ART,Brindle1993AMRRD,Grzybowski1975BIUT}.
 HIV status could further inform both infectiousness and susceptibility.
-On one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}.
-On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative patients.
+On the one hand, HIV-positive patients are more likely to have a negative sputum smear microscopy result than HIV-negative patients, due to the reduced lung cavitation and the fewer \emph{Mtb} bacilli in their sputum\cite{Brindle1993AMRRD,Telzak1997CID}.
+On the other hand, HIV-positive patients are less prone to control the \emph{Mtb} infection than HIV-negative patients, due to their reduced immune status\cite{Forte1992AIDS,Kwan2011CMR,Shen1988CEI}, and therefore TB disease progresses faster and is often more severe in HIV-positive TB patients compared to HIV-negative ones.
 

### 2024-05-21 analysis/qctr.Rmd
Revised illustrative example

@@
     title = "Quanta concentration in the waiting room on Oct 25 between 9:41 and 09:45am", # nolint
-    fill = expression("Quanta/m"^3 * " x 10e"^-3)
+    fill = expression("Quanta/m"^3 * " x 10"^-3)
   )
@@
     labs(
-      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10e"^-3 * ")")
+      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10"^-3 * ")")
     ) +

### 2024-05-21 tests/test_stm_v2.R

@@ -7,7 +7,7 @@ source("utils/plotting.R")
 dz <- 3
 gridCellLength <- 0.25
 nx <- 40
-ny <- 20
+ny <- 40
 dy <- gridCellLength * ny
 dx <- gridCellLength * nx
 roomDim <- c(dx, dy, dz)
@@ -26,16 +26,16 @@ time <- seq(
 )
 
 # assumptions
-aer <- 3 / 3600
-lambda <- 0 / 3600
+aer <- 1 / 3600
+lambda <- 0
 deposit <- 0
 
 # one infectious person in room center for 1h
 inf <- data.frame(
   t = 1:(length(time) / 2),
   x = "20",
-  y = "10",
-  q = 10 / 3600
+  y = "20",
+  q = 2 / 3600
 )
 
 
@@ -53,44 +53,31 @@ ct <- stm(
 )
 
 # plot results
-ct_to_df <- function(t) {
-  data.frame(ct[, , t]) %>%
-    add_rownames() %>%
-    reshape2::melt("rowname") %>%
-    rename(x = rowname, y = variable) %>%
-    mutate(
-      y = gsub("X", "", y),
-      across(c(x, y), as.numeric),
-      time = t
-    )
-}
-t_names <- c(
-  "08:01:00",
-  "08:30:00",
-  "09:00:00",
-  "09:01:00",
-  "09:05:00",
-  "10:00:00"
-)
-t_sel <- c(60, 1800, 3600, 3660, 3900, 7200)
-names(t_names) <- t_sel
-ctDF <- do.call(rbind, lapply(t_sel, ct_to_df)) %>%
-  mutate(
-    time = dplyr::recode(time, !!!t_names),
-    value = value * (1 / (gridCellLength^2 * 3))
-  )
+ct_df <- ct %>%
+  reshape2::melt() %>%
+  set_names(c("y", "x", "t", "q")) %>%
+  mutate(xy = paste0(x, ",", y))
 
-quanta_concn_pl <- ctDF %>%
-  ggplot(aes(x = y, y = x, fill = value)) +
-  facet_wrap(~time, ncol = 3) +
+quanta_concn_pl <- ct_df %>%
+  filter(t %in% c(1, 1800, 3600, 3660, 4200, 7200)) %>%
+  mutate(t = case_when(
+    t == 1 ~ "08:00",
+    t == 1800 ~ "08:30",
+    t == 3600 ~ "09:00",
[…]

### 2024-05-21 doc/paper/supplements.tex
Updates from Overleaf

@@
 \title{\LARGE\singlespacing{\textbf{Supplementary Information} \\ \medskip
-A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
+Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model}}
 
@@
 The Shiny app is shown in \Cref{fig:shiny-app}.
-In the center panel, a floor plan of the clinic is shown in (x,y)-space.
-Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
+In the centre panel, a floor plan of the clinic is shown in (x,y)-space.
+Different areas of the clinic are delineated by colours, \eg seating areas are marked in purple and exit areas are marked in blue.
 Tracks are shown as lines with a open and crossed circles indicating the tracks' start and end, respectively.
-The currently selected track is shown in black, possible links in colors, and alternatives for the selected possible link in light grey.
+The currently selected track is shown in black, possible links in colours, and alternatives for the selected possible link in light grey.
 
@@
 
-\section{Spatiotemporal modeling approach}\label{sec:spattemp-model}
+\section{Spatiotemporal modelling approach}\label{sec:spattemp-model}
 
-\subsection{The Wells-Riley modeling framework}
+\subsection{The Wells-Riley modelling framework}
 
@@
 It rather represents the dose of infectious particles that corresponds to a certain probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection\cite{Rudnick2003IndoorAir}.
-It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown.
+It is an attempt to consider the stochastic behaviour of airborne transmission, since the number of infectious particles that cause an infection is unknown.
 Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual.
@@
 The extension by Rudnick and Milton still assumes a well-mixed airspace.
-We relax this assumption by modeling spatial variation in the concentration of infectious quanta.
+We relax this assumption by modelling spatial variation in the concentration of infectious quanta.
 We then compute the risk of infection for susceptible individuals using the Wells-Riley equation considering their cumulative exposure to infectious quanta over space and time as
@@
 
-Modeling of the spatiotemporal quanta concentration can be divided into three processes
+Modelling of the spatiotemporal quanta concentration can be divided into three processes
 \begin{itemize}
@@
 
-We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$.
+We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the centre of cell $c$.
 The grid provides a discrete approximation to the continuously varying quanta concentration in the indoor space, with the number of grid cells $C = n_x \cdot n_y$ determining the spatial resolution.
-We denote the following variables and modeling parameters:
+We denote the following variables and modelling parameters:
 \begin{itemize}
@@
 The diffusion constant $D$ is unknown.
-We approximate it using an empirical relationship between the eddy diffusion coefficient $K$ ($K \approx D$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
+We approximate it using an empirical relationship between the eddy diffusion coefficient $K$ ($K \approx D$ in m$^2$\,s$^{-1}$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
 \begin{align}
-    K = (0.52 \cdot AER + 8.61\cdot10^{-5}) \cdot V^{\frac{2}{3}},
+    K = (0.52 \cdot AER + 8.61\cdot10^{-5} \cdot \text{s}^{-1}) \cdot V^{\frac{2}{3}},
 \end{align}
@@
 \end{align}
-where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$.
+where $Q$ is the outdoor air supply rate (in m$^3$\,s$^{-1}$ $\cdot \frac{1}{3600}$ ) and $AER = Q/V$.
 We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS), a quasi-Newton method, as implemented in the statistical software R\cite{Byrd1995SIAM}.
 The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
-We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints on $CO^{\text{O}} = [300,600]$\cite{Batterman2017IJERPH}.
+We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints on $CO^{\text{O}} = [300,600]$\,ppm\cite{Batterman2017IJERPH}.
 We can compute $AER$ over different time windows, hence the index in $AER_t$.
@@
     \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
-concn.}} + \underbrace{I_t \cdot q}_{\text{generation}})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda)\right)}_{\text{removal}} ~.
+concn.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda) \cdot 1\text{s}\right)}_{\text{removal}} ~.
 \end{align}
@@
 
-We illustrate our modeling approach with an example.
-Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of 0.25\,m$^2$.
+We illustrate our modelling approach with an example.
+Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of (0.25\,m)$^2$.
 We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
 We monitor the room from 8\,am to 12\,am and consider one infectious individual generating quanta continuously for one hour between 8\,am and 9\,am in the middle of the room (cell (20,10)), which means that quanta is generated in cells $(x_c = 20\pm1, y_c = 10\pm1)$.
-We assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0.5$\,quanta h$^{-1}$.
-We further assume a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$, translating into a quanta removal rate of $1 + 0.5 = 1.5$\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.007\,m$^2$ s$^{-1}$.
+We assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0$\,quanta h$^{-1}$.
+We further assume a constant outdoor air exchange rate of $AER = 3$\,air changes h$^{-1}$, translating into a quanta removal rate of 3\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.007\,m$^2$ s$^{-1}$.
 
 \Cref{fig:toy-example} shows the quanta concentration at different time points.
-The quanta concentration increases over time and is higher near the infectious individual in the room's center.
+The quanta concentration increases over time and is higher near the infectious individual in the room's centre.
 After the individual has left the room, the quanta diffuses fairly quickly and a well-mixed airspace is already at 9:05\,am.
@@
     \includegraphics{tests/stm_v2-toy_example.png}
-    \caption{Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's center.}
+    \includegraphics[tests/]{}
+    \caption{Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre.}
     \label{fig:toy-example}
@@
 We divide the clinic into three spaces: the waiting room (length = 10.55\,m $\times$ breadth = 5.50\,m $\times$ height = 3.00\,m), the corridor (length = 12.45\,m $\times$ breadth = 2.20\,m $\times$ height = 2.50\,m), and the TB room (length = 4.75\,m $\times$ breadth = 3.50\,m $\times$ height = 3.00\,m).
-We rasterize each room into a grid of cubic cells with a squared area of 0.25\,m$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+We rasterize each room into a grid of cubic cells with a squared area of (0.25\,m)$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
 
-Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate spaces to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
+Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate spaces to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
 Nevertheless, we acknowledge the possibility for aerosolized \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
@@
 
-We use Monte Carlo simulation to estimate the risk of infection, considering uncertainty in several modeling parameters. Each simulation consists of the following steps:
+We use Monte Carlo simulation to estimate the risk of infection, considering uncertainty in several modelling parameters. Each simulation consists of the following steps:
 \begin{itemize}
-    \item[1.] \textbf{Sample uncertain modeling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, the mask reduction rate, and the bacterial inactivation rate.
+    \item[1.] \textbf{Sample uncertain modelling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, the mask reduction rate, and the bacterial inactivation rate.
     \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room per unit space (cell) over time.
@@
 
-\subsection{Modeling assumptions}\label{sec:priors}
+\subsection{Modelling assumptions}\label{sec:priors}
 
@@
 Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate the air exchange rate, based on which we model quanta diffusion and removal.
-The combination of clinical and person-tracking data further allows the spatial modeling of quanta generation.
-This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1961Book,Rudnick2003IndoorAir}.
+The combination of clinical and person-tracking data further allows the spatial modelling of quanta generation.
+This makes our approach unique compared to existing modelling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1961Book,Rudnick2003IndoorAir}.
 In our spatiotemporal model, we assume a radial diffusion of quanta, which may not reflect the actual airflow at the primary care clinic.
@@
 For example, measuring airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh outdoor air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered in modeling because it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
+However, airflow is rarely considered in modelling because it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
 
@@
 Survival and dispersion of aerosols are further determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
-Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modeling framework via the quanta generation rate parameter\cite{Riley1961Book,Rudnick2003IndoorAir}.
+Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modelling framework via the quanta generation rate parameter\cite{Riley1961Book,Rudnick2003IndoorAir}.
 It is challenging to consider these properties specifically, especially because they can be modified by environmental factors with often unclear, pathogen-specific effects\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.

### 2024-05-21 doc/paper/manuscript.tex

@@
 
-\title{\bfseries\LARGE\singlespacing{A novel spatiotemporal model to estimate \emph{Mycobacterium tuberculosis} transmission risks in crowded indoor settings using environmental, clinical, and patient movement data}}
+\title{\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model}}
 % author list
@@
 They exit the clinic through one of the exits in the waiting room or at the end of the corridor.
-Colored icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
+Coloured icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
     \label{fig:floor-plan}
@@
 \begin{align}
-    P = \frac{C}{S} = 1 - \exp\left(\frac{Ipqt}{Q}\right),
+    P = \frac{C}{S} = 1 - \exp\left(-\frac{Ipqt}{Q}\right),
 \end{align}
 where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
-The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
+The probability of infection is estimated with a Poisson relation, considering the stochastic behaviour of airborne infection, where one quantum corresponds to a 67\% risk of infection.
 The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
@@
 A detailed description of our extension is provided in \supp~Section~\zref{sec:spattemp-model}, together with an illustrative example using simulated data (\supp~Fig.~\zref{fig:toy-example}).
-In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).
+In the following, we describe the six main components of our modelling approach (\Cref{fig:modeling-flow}).
 
@@
     \includegraphics{doc/model-flow-chart.pdf}
-    \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
+    \caption{\textbf{Visual summary of the spatiotemporal modelling approach}.
 (1)~Clinical records and person tracks were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients.
@@
 In brief, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
-Each room was rasterized into a grid of cubic cells with an area of 0.25\,m$^2$.
+Each room was rasterized into a grid of cubic cells with an area of (0.25\,m)$^2$.
 The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
-We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters.
-In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
+We used a Monte-Carlo simulation approach to consider uncertainty in modelling parameters.
+In each simulation, we sampled uncertain modelling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
 
@@
 Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
-Except for the scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.
+Except for the scenario-specific modifications, all other modelling assumptions and parameters were identical in each Monte Carlo simulation.
 
@@
 The mean risk of infection per clinic visit was 0.06\% (80\%-credible interval (CrI) 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
+\supp~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
+It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
+The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
 
@@
 
-Finally, \supp~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
-It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
-The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
-
 \FloatBarrier
@@
 % close contacts
-The COVID-19 pandemic has renewed interest in airborne transmission of respiratory infections in crowded indoor settings\cite{Wang2021Science,Morawska2021,Greenhalgh2022BMJ} and its modeling\cite{Shen2021STBE,Buonanno2020EI,Edwards2024RiskAnalysis}.
+The COVID-19 pandemic has renewed interest in airborne transmission of respiratory infections in crowded indoor settings\cite{Wang2021Science,Morawska2021,Greenhalgh2022BMJ} and its modelling\cite{Shen2021STBE,Buonanno2020EI,Edwards2024RiskAnalysis}.
 Many studies\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR} rely on the Wells-Riley transmission model\cite{Riley1961Book}, based on the pioneering work of William Wells, who proposed a role of droplets in airborne transmission in the 1930s\cite{Wells1934AJE}, and Richard Riley, who developed the equation in the 1960s\cite{Riley1961Book}.
@@
 Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic.
-Further, different modeling assumptions could have contributed.
+Further, different modelling assumptions could have contributed.
 The risk estimate from our previous study before the COVID-19 pandemic\cite{Zurcher2022JID} was based on TB prevalence in the South African population, rather than the TB status documented in the clinical data.
-Modeling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modeled risk of transmission.
+Modelling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modeled risk of transmission.
 
@@
 However, CFD models are computationally expensive and primarily used in experimental and static environments, making their application to real-world settings such as clinics difficult.
-Moreover, our modeling approach required computational efficiency because a many Monte Carlo simulations are needed to reflect uncertainty in modeling parameters.
+Moreover, our modelling approach required computational efficiency because a many Monte Carlo simulations are needed to reflect uncertainty in modelling parameters.
 
@@
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-Contrary to our assumptions, infectious particles may not diffuse radially from the source or approach a steady-state with uniform contamination of the airspace, although we have observed this behavior in experiments (see Text~A and Figure~S1 in S1~Appendix of Ref.
-\cite{Banholzer2023PLoSMed}).
-Furthermore, uncertainty in modeling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
-Third, we could not validate our modeling results with empirical findings.
+We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
+However, airflow is very difficult to measure.
+Furthermore, uncertainty in modelling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
+Third, we could not validate our modelling results with empirical findings.
 Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be diagnosed with TB.
@@
 \section*{Data availability}
-Preprocessed data and code are available from GitHub via \url{https://github.com/nbanho/clinic-transmission}.
+Restrictions on the availability of personal data apply, but are necessary to maintain the confidentiality of participants.
+The data is available upon reasonable request, contact: University of Bern, info.ispm@unibe.ch.
+All other data are included in the manuscript and accompanying supporting information.
+Code used for preprocessing, modelling and analysis, and descriptive and simulation results are available from GitHub via \url{https://github.com/nbanho/clinic-transmission}.
 
@@
 \section*{Author contributions}
-xxx.
+Conception and design: NB, KM, KZ, LF.
+Data collection: KM, JL, RS, KZ, LF.
+Modelling and statistical analysis: NB.
+Paper draft: NB, LF, ME.
+All authors reviewed and approved the final version of the manuscript.
 

### 2024-05-21 tests/test_stm_v2.R
Change labels of toy examples

@@ -73,7 +73,7 @@ quanta_concn_pl <- ct_df %>%
   geom_tile() +
   labs(
     fill = expression(Quanta %.% 10^-3 * " / 0.1875m"^3),
-    x = "x length in m", y = "y width in m", title = "A"
+    x = "x length in m", y = "y width in m", title = "a"
   ) +
   scale_fill_gradientn(
     colours = RColorBrewer::brewer.pal(n = 9, name = "YlOrRd"),
@@ -117,7 +117,7 @@ time_pl <- ct_df %>%
   scale_y_continuous(labels = function(x) round(1e3 * x, 1)) +
   labs(
     y = expression(Quanta %.% 10^-3 * " / 0.1875m"^3),
-    title = "B"
+    title = "b"
   ) +
   coord_cartesian(expand = FALSE) +
   theme_custom() +

### 2024-05-21 doc/paper/supplements.tex
Updates from Overleaf

@@
     \item $V$: volume of the airspace (m$^3$)
-    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air (parts per million, ppm)
-    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air (ppm)
+    \item $CO^{\text{I}}$: CO$_2$ mixing ratio in indoor air (parts per million, ppm)
+    \item $CO^{\text{O}}$: CO$_2$ mixing ratio in outdoor air (ppm)
     \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
@@
 The $AER$ is not directly observed but can be estimated from indoor CO$_2$ levels, room occupancy, and volume.
-Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state concentration), the $AER$ can be computed as
+Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state mixing ratio), the $AER$ can be computed as
 \begin{align}
@@
 \end{align}
-where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person, $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air\cite{Batterman2017IJERPH}.
-However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration.
+where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person, $CO^{\text{I}}$ is the CO$_2$ mixing ratio in indoor and $CO^{\text{O}}$ is the mixing ratio in outdoor air\cite{Batterman2017IJERPH}.
+However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ mixing ratio.
 Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
@@
 The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
-We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints on $CO^{\text{O}} = [300,600]$\,ppm\cite{Batterman2017IJERPH}.
+We also follow recommendations to fit the outdoor CO$_2$ mixing ratio with reasonable constraints on $CO^{\text{O}} = [300,600]$\,ppm\cite{Batterman2017IJERPH}.
 We can compute $AER$ over different time windows, hence the index in $AER_t$.
@@
 We monitor the room from 8\,am to 12\,am and consider one infectious individual generating quanta continuously for one hour between 8\,am and 9\,am in the middle of the room (cell (20,10)), which means that quanta is generated in cells $(x_c = 20\pm1, y_c = 10\pm1)$.
-We assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0$\,quanta h$^{-1}$.
-We further assume a constant outdoor air exchange rate of $AER = 3$\,air changes h$^{-1}$, translating into a quanta removal rate of 3\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.007\,m$^2$ s$^{-1}$.
+We assume a quanta generation rate of $q = 2$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0$\,quanta h$^{-1}$.
+We further assume a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$, translating into a quanta removal rate of 1\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.015\,m$^2$ s$^{-1}$.
 
-\Cref{fig:toy-example} shows the quanta concentration at different time points.
-The quanta concentration increases over time and is higher near the infectious individual in the room's centre.
-After the individual has left the room, the quanta diffuses fairly quickly and a well-mixed airspace is already at 9:05\,am.
-Over time, the removal through outdoor air exchange and bacterial inactivation brings the concentration in the entire room back towards zero.
+\Cref{fig:toy-example} shows the spatiotemporal quanta concentration.
+The quanta concentration increases over time and is higher near the infectious individual in the room's centre (\Cref{fig:toy-example}a, top panel).
+After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am (\Cref{fig:toy-example}a, bottom panel) through outdoor air exchange.
+Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until the a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).
 
@@
     \includegraphics{tests/stm_v2-toy_example.png}
-    \includegraphics[tests/]{}
-    \caption{Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre.}
+    \includegraphics{tests/stm_v2-toy_example-B.png}
+    \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre}.
+\textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time per cell of the indoor space.}
     \label{fig:toy-example}

### 2024-05-21 analysis/patient-tracking.Rmd
Examples person movements manuscript

@@
 ```{r}
+tracking_dates <- c(
+  "2021-10-13", "2021-10-15", "2021-10-25", "2021-11-04", "2021-11-05"
+)
+
 # spatial track counts
@@
 # unique id
-df$day_pid <- paste(df$date, df$patient_id)
+df$day_pid <- paste("Date:", df$date, " |  ID:", df$patient_id)
 
@@
 ex_pid_man <- c(
-  "2021-10-25 4649", # Entered + Exited, including TB room stay
-  "2021-10-13 6297", # Lost Exit, only waiting room
-  "2021-11-04 12860", # (Entered) + Exited, waiting room and corridor
-  "2021-10-13 9346", # Entered + Exited, including TB room
-  "2021-11-05 16625", # Entered + (Exited), only waiting room
-  "2021-10-25 9526", # Lost Exit, waiting room and corridor
-  "2021-11-04 9841", # Entered = (Exited), waiting room and corridor
-  "2021-10-15 8644", # Entered + Exited, including TB room
-  "2021-10-15 9419" # Lost Enter
+  "Date: 2021-10-25  |  ID: 4649", # Entered + Exited, including TB room stay
+  # "2021-10-13 6297", # Lost Exit, only waiting room
+  # "2021-11-04 12860", # (Entered) + Exited, waiting room and corridor
+  # "2021-10-13 9346", # Entered + Exited, including TB room
+  # "2021-11-05 16625", # Entered + (Exited), only waiting room
+  "Date: 2021-10-25  |  ID: 9526", # Lost Exit, waiting room and corridor
+  "Date: 2021-11-04  |  ID: 9841", # Entered = (Exited), waiting room and corridor
+  "Date: 2021-10-15  |  ID: 8644" # Entered + Exited, including TB room
+  # "2021-10-15 9419" # Lost Enter
 )
@@
     ) +
-    facet_wrap(~day_pid, ncol = 3) +
+    facet_wrap(~day_pid, ncol = 2) +
     scale_shape_manual(values = c(1, 13)),

### 2024-05-21 doc/paper/manuscript.tex
Updates from Overleaf

@@
 
-\title{\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model}}
+\title{\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}}
 % author list
@@
 We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased.
-We matched person tracks with clinical records to identify the spatiotemporal location of infectious TB patients.
+We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
 We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
@@
 
-Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person tracks) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
+Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa.
 The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm.
@@
 They exit the clinic through one of the exits in the waiting room or at the end of the corridor.
-Coloured icons show the placement of the video sensors to track persons and the devices monitoring CO$_2$ levels.}
+Coloured icons show the placement of the video sensors to track person movements and the devices monitoring CO$_2$ levels.}
     \label{fig:floor-plan}
@@
 The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-People in the clinic could contribute multiple records if they moved out of the range of a sensor or were not recognized as a person.
-To address this, we created an R Shiny tool (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}.
+People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person.
+To address this, we created an R Shiny application (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}.
 We also labeled the tracks to identify the movements of clinic staff.
-\supp~Fig.~\zref{fig:tracking-examples} shows some sample tracks.
+\supp~Fig.~\ref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/example-patient-tracks.png}
+    \caption{Examples of person movements detected with our video sensors.}
+    \label{fig:tracking-examples}
+\end{figure}
 
@@
     \caption{\textbf{Visual summary of the spatiotemporal modelling approach}.
-(1)~Clinical records and person tracks were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients.
+(1)~Clinical records and person movements were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients.
 (2)~Both diagnosed and randomly selected undiagnosed TB patients among clinic attendees generate infectious quanta.
@@
 
-(1)~We matched clinical and person-tracking data to identify the spatiotemporal location of infectious TB patients within the clinic.
+(1)~We matched clinical records with person movements to identify the spatiotemporal location of infectious TB patients within the clinic.
 We recorded the timestamp when a person was detected in the registration area and matched this with the registration time in the clinical records.
 We excluded tracks with a total duration of less than five minutes in the clinic and a duration of less than five seconds in the registration area from the matching.
-We allowed for a maximum delay of 15\,min between the track in the registration area and the entry in the clinical database.
-All diagnosed TB patients could be matched with person tracks this way.
+We allowed for a maximum delay of 15\,min between the track's timestamp in the registration area and the entry in the clinical database.
+All diagnosed TB patients could be matched with person movements this way.
 
@@
 During one week (five open clinic days), 894 patients were registered in the clinical database: four patients were bacteriologically diagnosed with TB and 16 patients were suspected of having TB after symptom screening.
-Overall, 1,563 person tracks were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors).
+Overall, 1,563 person movements were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors).
 Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).

### 2024-05-21 doc/paper/supplements.tex

@@
 \title{\LARGE\singlespacing{\textbf{Supplementary Information} \\ \medskip
-Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model}}
+Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}}
 
@@
 Second, we manually reconnected the remaining tracks if they were likely from the same person.
-For that purpose, we developed a Shiny app and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) to assist the students performing the matching.
+For that purpose, we developed a Shiny application and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) to assist the students performing the matching.
 
-The Shiny app is shown in \Cref{fig:shiny-app}.
+The Shiny application is shown in \Cref{fig:shiny-app}.
 In the centre panel, a floor plan of the clinic is shown in (x,y)-space.
@@
 
-The Shiny app's left panel shows descriptive information about the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages.
+The Shiny application's left panel shows descriptive information about the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages.
 It further shows the current Pid with shortcuts to move to the next or previous Pid, or the next or previous Pid at the entrance.
@@
     \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
-    \caption{Screenshot of the Shiny app used for manual linking tracking IDs that probably belong together.}
+    \caption{Screenshot of the Shiny application used for reconnecting interrupted tracking IDs.}
     \label{fig:shiny-app}
@@
 Further exclusion of short tracks <5\,min resulted in 1,563 tracks used for analysis: 209 (13\%) labeled as "Clean", 125 (8\%) as "Staff", 213 (14\%) as "Lost enter", 246 (16\%) as "Lost exit", and 770 (49\%) as "Lost both".
-A sample of person tracks are shown in \Cref{fig:tracking-examples}.
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/data/example-patient-tracks.png}
-    \caption{Sample of person tracks from our video sensors after preprocessing.}
-    \label{fig:tracking-examples}
-\end{figure}
 
@@
 where $V$ is the volume of the airspace (in m$^3$).
-We thus assume that pathogen-carrying particles spread with a similar speed than CO$_2$ particles, which likely overestimates the diffusion constant, because CO$_2$ particles are smaller than \emph{Mtb}-carrying particles.
+We thus assume that pathogen-carrying particles spread with a similar speed than CO$_2$ particles, which may overestimate the diffusion constant because CO$_2$ particles are smaller than \emph{Mtb}-carrying particles.
 
@@
     \includegraphics{results/inputs/quanta-generation-rate.png}
-    \caption{Prior distribution (log scale) for the quanta generation rate (quanta h$^{-1}$) by activity level (sitting and walking), along with reported average estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}.}
+    \caption[Prior distribution for the quanta generation rate]{Prior distribution (log scale) for the quanta generation rate (quanta h$^{-1}$) by activity level (sitting and walking), along with reported average estimates from the literature\cite{Andrews2014JID,Riley1962ARRD,Escombe2008PLoSMed,Nardell1991ARRD,Dharmadhikari2012AJRCCM}.}
     \label{fig:quanta-distribution}
@@
     \includegraphics{results/inputs/bacterial-inactivation-rate.png}
-    \caption{Prior distribution (log scale) for the bacterial inactivation rate (quanta h$^{-1}$) of \emph{Mtb} along with reported estimates from the literature\cite{Loudon1969AMRRD,Gannon2007ResVetSci,Klein2014IJMyco,Lever2000LettersAppliedMicrobio}.}
+    \caption[Prior distribution for the bacterial inactivation rate]{Prior distribution (log scale) for the bacterial inactivation rate (quanta h$^{-1}$) of \emph{Mtb} along with reported estimates from the literature\cite{Loudon1969AMRRD,Gannon2007ResVetSci,Klein2014IJMyco,Lever2000LettersAppliedMicrobio}.}
     \label{fig:lambda-distribution}
@@
     \includegraphics{results/inputs/undiagnosed-tb-patients.png}
-    \caption{Counts (\%) of the daily number of diagnosed TB patients visiting the clinic according to clinical records.
+    \caption[Counts of the daily number of diagnosed TB patients visiting the clinic]{Counts (\%) of the daily number of diagnosed TB patients visiting the clinic.
 The number of diagnosed patients is modeled with a Multinomial distribution based on these counts.}
@@
     \includegraphics{results/modeling/mean-roi-assumptions.png}
-    \caption{Mean risk of infection per clinic attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
+    \caption[Mean risk of infection per clinic attendee when varying assumptions about the number of infectious people in the clinic]{Mean risk of infection per clinic attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
 Low: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}

### 2024-06-03 analysis/qctr.Rmd
Update modeling results

@@
 dy <- 25
-sim <- 1
+sim <- 2
 param_1025 <- readRDS(
@@
 
-```{r}
-main_tr <- main_tr %>%
-  mutate(P_unc = P80 - P20)
-
-hr_lu_pt <- main_tr %>%
-  filter(P_unc < .25 / 100, P_mean > 0.25 / 100) %>%
-  head(1)
-
-main_focus_pl <- main_tr %>%
-  ggplot(aes(x = P_unc, y = P_mean)) +
-  geom_point(shape = 4) +
-  annotate(
-    "point",
-    x = hr_lu_pt$P_unc,
-    y = hr_lu_pt$P_mean,
-    size = 7,
-    shape = 1,
-    color = "red"
-  ) +
-  annotate(
-    geom = "segment",
-    x = hr_lu_pt$P_unc + .00002,
-    y = hr_lu_pt$P_mean + .0004,
-    xend = hr_lu_pt$P_unc + .00008,
-    yend = hr_lu_pt$P_mean + .004,
-    arrow = arrow(length = unit(0.1, "cm")),
-    color = "red"
-  ) +
-  annotate(
-    geom = "text",
-    x = hr_lu_pt$P_unc + .00008,
-    y = hr_lu_pt$P_mean + .004,
-    label = "Moderate risk,\nlow uncertainty",
-    hjust = -.1,
-    size = 6.5 / cm(1),
-    color = "red"
-  ) +
-  scale_y_continuous(
-    expand = expansion(mult = c(0.01, 0.01)),
-    limits = c(0, .01),
-    breaks = seq(0, .01, .0025),
-    labels = function(x) round_k(100 * x, 3)
-  ) +
-  scale_x_continuous(
-    expand = expansion(mult = c(0.01, 0.01)),
-    limits = c(0, .002),
-    breaks = seq(0, .002, .0005),
-    labels = function(x) round_k(100 * x, 3)
-  ) +
-  labs(
-    x = "Uncertainty (80%-CrI, pp)",
-    y = "Mean risk of infection (%)"
-  ) +
-  theme_bw2() +
-  theme(
-    plot.title.position = "plot",
-    plot.title = element_text(size = 10, face = 2),
-    plot.margin = margin(.5, .5, .5, .5, "cm")
-  )
-
-save_plot(
-  main_focus_pl,
-  "results/modeling/patient-risk-focus-plot.png",
-  w = 12,
-  h = 10
-)
-```
-
 ### Association
@@
 assoc_df <- track_ft %>%
-  # left_join(
-  #   aer %>%
-  #     filter(room == "Waiting room") %>%
-  #     dplyr::select(date, daytime, aer)
-  # ) %>%
   mutate(patient_id = paste(
@@
   ) %>%
-  mutate(across(c(lower, mean, upper), ~ round_k(.x, 2)))
-```
-
-## Comparison with temporal model
-
-```{r}
-# files
-files_temp <- list.files("simulations/temporal/",
-  full.names = TRUE, pattern = "trans-risk",
-  recursive = TRUE
-)
-
-temp_tr <- map_dfr(files_temp, average_trans_risk)
-temp_annon <- annotate_average_trans_risk(temp_tr)
-
-# comparison with temporal model
-mod_comp <- left_join(
-  temp_tr %>%
-    dplyr::select(patient_id, P_mean) %>%
-    rename(`Temporal model` = P_mean),
-  main_tr %>%
-    dplyr::select(patient_id, P_mean) %>%
-    mutate(`Spatiotemporal model` = P_mean)
-) %>%
-  mutate(
-    ratio = abs(`Spatiotemporal model` - `Temporal model`) / `Temporal model` * 100
-  )
-
-summary(mod_comp$ratio)
+  mutate(across(c(lower, mean, upper), ~ round_k(.x, 2))) %>%
+  dplyr::select(variable, mean, lower, upper)
 ```
@@
   ) +
-  facet_wrap(~scenario, ncol = 2) +
+  facet_wrap(~scenario, ncol = 2, scales = "free_x") +
   scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
@@
     legend.key.width = unit(.4, "cm"),
-    strip.text = element_text(hjust = 0)
+    strip.text = element_text(hjust = 0),
+    panel.spacing = unit(.5, "cm")
   )

### 2024-06-03 doc/paper/manuscript.tex
Updates from Overleaf

@@
 CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was 0.06\% (95\%-credible interval [CrI] 0.001\%$-$0.40\%).
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.28-fold (95\%-CrI 1.19$-$1.36) higher in scenarios without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with pre-pandemic ventilation rates.
+The estimated risk of infection per clinic attendee was 0.1\% (95\%-credible interval [CrI] 0.01\%$-$0.1\%).
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) with pre-pandemic ventilation rates.
 Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
@@
     \includegraphics{results/data/example-patient-tracks.png}
-    \caption{Examples of person movements detected with our video sensors.}
+    \caption{Examples of person movements in the clinic.}
     \label{fig:tracking-examples}
@@
 We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-We set the gravitational settling rate to zero, ignoring sedimentation because the particles carrying \emph{Mtb}, which are in the size range of $1-7\mu$m\cite{Fennelly2020Lancet}, are airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
+We set the gravitational settling rate to zero and ignore sedimentation because particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of just $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}.
 
@@
 \begin{align}
-    P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a\right),
+    P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),
 \end{align}
@@
 The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
-An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translated to a 0.21\% risk of infection in the morning compared to a 0.05\% risk in the afternoon.
-The mean risk of infection per clinic visit was 0.06\% (80\%-credible interval (CrI) 0.01\%$-$0.07\%), corresponding to a risk of 0.12\% per hour (80\%-CrI 0.01\%$-$0.23\%).
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a 0.40\% risk of infection in the morning compared to a 0.07\% risk in the afternoon.
+The mean risk of infection per clinic visit was 0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%), corresponding to a risk of 0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%).
 \supp~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
@@
 The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
-A doubling of close-contact encounters increased the odds of infection by 17\% (OR 1.17, 95\%-CrI 1.04$-$1.30).
-Doubling the time spent in the clinic increased the odds of infection by 9\% (OR 1.09, 95\%-CrI 1.01$-$1.18).
-The time spent in close contact was not significantly associated with the individual risk of infection (OR 1.05, 95\%-CrI, 0.96$-$1.14).
+A doubling of close-contact encounters increased the odds of infection by 23\% (OR 1.23, 95\%-CrI 1.09$-$1.37).
+Doubling the time spent in the clinic increased the odds of infection by 12\% (OR 1.12, 95\%-CrI 1.03$-$1.21).
+The time spent in close contact was not significantly associated with the individual risk of infection (OR 1.03, 95\%-CrI, 0.94$-$1.13).
 
@@
 The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.07\% (80\%-CrI 0.01\%$-$0.09\%), a 30\% increase (OR 1.3, 80\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.10\% (80\%-CrI 0.01\%$-$0.09\%), a 70\% increase (OR 1.7, 80\%-CrI 0.5$-$5.3).
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.12\% (80\%-CrI 0.01\%$-$0.12\%), a 120\% increase (OR 2.2, 80\%-CrI 0.6$-$6.7).
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.12\% (80\%-CrI 0.02\%$-$0.13\%), a 30\% increase (OR 1.3, 80\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.19\% (80\%-CrI 0.02\%$-$0.20\%), a 120\% increase (OR 2.2, 80\%-CrI 0.9$-$5.2).
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.24\% (80\%-CrI 0.03\%$-$0.24\%), a 180\% increase (OR 2.8, 80\%-CrI 1.1$-$6.6).
 The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
@@
 In sensitivity analyses, we varied assumptions about the number of infectious people in the clinic.
-Assuming the number of infectious TB patients reflects the prevalence of TB in the South African population, the mean risk of infection would have been higher but the tail risk smaller (0.08\%, 80\%-CrI 0.02\%$-$0.13\%).
-Further, assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, both the mean risk of infection would have been considerably higher (0.14\%, 80\%-CrI <0.01\%$-$0.26\%).
+Assuming the number of infectious TB patients reflects the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher (0.12\%, 80\%-CrI 0.03\%$-$0.19\%).
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, the mean risk of infection would have been two times higher (0.22\%, 80\%-CrI <0.01\%$-$0.43\%).
 
@@
 We applied the model to data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
-We estimated an average risk of infection per clinic attendee of 0.06\%.
+We estimated an average risk of infection per clinic attendee of 0.1\%.
 We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
-Without infection control measures (masks and high natural ventilation), the risk of infection would have been more than twice as high.
+Without infection control measures (masks and high natural ventilation), the risk of infection would have been almost three times as high.
 Our spatiotemporal model can be readily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
@@
 Nevertheless, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
-Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 1.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
+Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 2.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
 The difference can be partly attributed to the infection control measures that were in place during the SARS-CoV-2 pandemic.
@@
 Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely due to a better ventilation, with an air change rate comparable to the one measured in the primary care clinic of our study.
-We stress that our study results do \emph{not} suggest that masks are ineffective but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
+We stress that our results do \emph{not} suggest that masks are ineffective but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 

### 2024-11-13 preprocessing/prep_simulation-data.R
include gravitational settling rate

@@ -4,6 +4,7 @@ library(tidyverse)
 library(parallel)
 library(lubridate)
 
+source("utils/distr.r")
 source("utils/spatial.r")
 source("utils/trans_risk.r")
 
@@ -148,13 +149,17 @@ for (d in dates) {
     # viral inactivation rate
     lambda <- rlambda(n)
 
+    # gravitational settling rate
+    dpr <- rk(n)
+
     # combine into tibble
     sim_param <- tibble(
       sim = 1:n,
       q_wait = q_wait,
       q_walk = q_walk,
       mask_red = q_masking_red,
-      viral_inact = lambda
+      viral_inact = lambda,
+      settling_rate = dpr
     )
   }
 

### 2024-11-13 preprocessing/simulate_qctr.R

@@ -285,7 +285,7 @@ for (sim in cont_n:term_n) {
         vol = prod(roomDim),
         aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Morning"] / 3600,
         lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
-        deposit = 0
+        deposit = sim_param$settling_rate[sim_param$sim == sim] / 3600
       )
     } else if (mod == "temp") {
       ct_morn <- tm(
@@ -294,7 +294,7 @@ for (sim in cont_n:term_n) {
         inf = dplyr::select(track_tb_morn, -time, -activity),
         aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Morning"] / 3600,
         lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
-        deposit = 0
+        deposit = sim_param$settling_rate[sim_param$sim == sim] / 3600
       )
     } else {
       stop("Model does not exist")
@@ -348,7 +348,7 @@ for (sim in cont_n:term_n) {
         vol = prod(roomDim),
         aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Afternoon"] / 3600,
         lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
-        deposit = 0
+        deposit = sim_param$settling_rate[sim_param$sim == sim] / 3600
       )
     } else if (mod == "temp") {
       ct_noon <- tm(
@@ -357,7 +357,7 @@ for (sim in cont_n:term_n) {
         inf = dplyr::select(track_tb_noon, -time, -activity),
         aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Afternoon"] / 3600,
         lambda = sim_param$viral_inact[sim_param$sim == sim] / 3600,
-        deposit = 0
+        deposit = sim_param$settling_rate[sim_param$sim == sim] / 3600
       )
     } else {
       stop("Model does not exist")

### 2024-11-13 utils/trans_risk.r

@@ -127,6 +127,8 @@ IR <- function(activity, sex) {
 #' IR ratios for different physical activity levels and weight the VER by the proportion
 #' of breathing and speaking. We use the average IRs across men and women.
 #'
+#' @return quanta emission rate in quanta h-1
+#'
 #' @references
 #' - Mikszewski: Mikszewski et al. (2022) in Geoscience frontiers (doi.org/10.1016/j.gsf.2021.101285)
 #' - Adams: Adams (1993) [Report] Measurement of Breathing Rate and Volume in Routinely Performed Daily Activities.
@@ -173,6 +175,9 @@ bact_load.mtb <- function(n) rlnorm(n, log(10^5.5), log(10^1.3)) # CFU mL-1
 #' Viral inactivation rate
 #'
 #' @param n number of samples
+#' @param disease default and currently only option is TB
+#'
+#' @return inactivation rate in h-1
 #'
 #' @details
 #' Based on a crude approximation to various estimates in the literature (see references).
@@ -187,6 +192,32 @@ rlambda <- function(n, disease = "TB") {
   rlnorm(n, meanlog = log(1), sdlog = 1)
 }
 
+#' Deposition rate due to gravitational settling
+#'
+#' @param n number of samples
+#' @param disease default and currently only option is TB
+#' @param height default 1.7m
+#'
+#' @details
+#' Assuming a settling velocity of 3.5 x 10^−5 − 1.5 x 10^−3 m s−1
+#' for Mtb particles in the size range of 1 to 7mum,
+#' which is modelled with a Gamma distribution,
+#' using the central estimate of the settling velocity and
+#' computing the standard deviation from the .
+#'
+#' @return deposition rate in h-1
+#'
+#' @references
+#' - Vuorinen et al. (2020) in Saf Sci. (doi:10.1016/j.ssci.2020.104866).
+
+rk <- function(n, disease = "TB", height = 1.7) {
+  u <- 1.5e-3
+  l <- 3.5e-5
+  m <- (u + l) / 2
+  s <- (u - l) / (2 * qnorm(0.975))
+  gamma_par <- inv_gamma_moments(m, s)
+  (rgamma(n, shape = gamma_par$shape, rate = gamma_par$rate)) / height * 3600
+}
 
 #' Compute air change rate using transient mass balance method
 #'

### 2024-11-13 doc/paper/supplements.tex

@@
 \end{align}
-where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
+where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$\,h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta\,h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$\,h$^{-1}$).
 
@@
 \end{equation}
-where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate.
+where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta\,$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate (in m$^3$\,h$^{-1}$).
 
@@
     \item $k$: gravitational settling rate (quanta s$^{-1}$)
+    \item $p$: breathing rate (m$^3$\,h$^{-1}$)
 \end{itemize}
@@
 After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am (\Cref{fig:toy-example}a, bottom panel) through outdoor air exchange.
-Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until the a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).
+Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).
 
@@
     \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre}.
-\textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time per cell of the indoor space.}
+\textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time for the 800~cells into which the indoor space was divided.}
     \label{fig:toy-example}
@@
 The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking.
-The activity is determined based on whether the the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).
+The activity is determined based on whether the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).
 
@@
 Adams (1993)\cite{Adams1993} reports breathing rates for different activity levels.
-We use the average breathing rate for the following activity levels: sitting with $p = 0.51$\,m$^3$ h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$ h$^{-1}$.
+We use the average breathing rate for the following activity levels: sitting with $p = 0.51$\,m$^3$\,h$^{-1}$ and walking (slowly at 2.5 mph) with $p = 1.33$\,m$^3$\,h$^{-1}$.
 
@@
 \section{Additional results}\label{sec:additional-results}
-
-% \begin{figure}[!htpb]
-%     \centering
-%     \includegraphics{results/data/diagnosed-tb-patients.png}
-%     \caption{Number of diagnosed TB patients (dark bars) and those suspected of TB (light bars) following symptom screening.
-Modeled study days in color, other clinic days in October and November in grey.}
-%     \label{fig:diagnosed-tb-patients}
-% \end{figure}
 
@@
 \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics{results/modeling/patient-risk-focus-plot.png}
-%     \caption{Mean and 80\%-CrI (uncertainty) of the risk of infection per clinic visit.
-Highlighted in red is a visit from a clinic attendee with moderate risk of infection and low estimated uncertainty.}
-%     \label{fig:patient-risk-focus}
-% \end{figure}
 

### 2024-11-19 analysis/environmental-data.Rmd
CO2 and occupancy time series plots

@@
   geom_step() +
-  facet_wrap(~room) +
-  coord_cartesian(expand = FALSE) +
-  scale_x_datetime(date_breaks = "4 hours", date_labels = "%H:%M") +
+  facet_wrap(~room, ncol = 1) +
+  scale_y_continuous(
+    limits = c(350, 700),
+    breaks = seq(350, 700, 100),
+    expand = expansion(add = c(0, 0))
+  ) +
+  scale_x_datetime(
+    expand = expansion(add = c(0, 1200)),
+    date_breaks = "4 hours",
+    date_labels = "%H:%M"
+  ) +
   scale_color_manual(values = wes_palette("Moonrise3")) +
@@
   pdf_file = "results/data/co2-levels-over-time.png",
-  w = 16, h = 10
+  w = 16, h = 18
 )

### 2024-11-19 analysis/patient-tracking.Rmd
CO2 and occupancy time series plots

@@
 source("utils/spatial.r")
-cellSize <- 250
 source("preprocessing/prep_building-rasterize.R")
@@
 )
+abbr_dates <- format(as.Date(tracking_dates), format = "%b %d")
 
@@
 )
-```
-
-### Building
-
-```{r}
-cellSize <- 250
-source("preprocessing/prep_building-rasterize.R")
 ```
@@
     ),
-    time = as.POSIXct(format(date_time, format = "%H:%M:%S"), format = "%H:%M:%S")
+    time = as.POSIXct(format(
+      date_time,
+      format = "%H:%M:%S"
+    ), format = "%H:%M:%S")
   )
@@
   w = 8, h = 8
+)
+
+time_by_day_pl <- track_count_time %>%
+  mutate(
+    time = as.POSIXct(
+      format(date_time, format = "%H:%M:%S"),
+      format = "%H:%M:%S"
+    ),
+    mthdy = factor(
+      as.character(format(date, format = "%b %d")),
+      levels = abbr_dates
+    ),
+    time = floor_date(time, "10 minutes")
+  ) %>%
+  group_by(mthdy, time, room) %>%
+  summarize(n = mean(n)) %>%
+  ungroup() %>%
+  mutate(room = factor(
+    room,
+    levels = c("Waiting room", "Corridor", "TB room")
+  )) %>%
+  ggplot(aes(x = time, y = n, color = mthdy)) +
+  geom_step() +
+  facet_wrap(~room, ncol = 1) +
+  scale_y_continuous(
+    limits = c(0, NA),
+    expand = expansion(add = c(0, 1))
+  ) +
+  scale_x_datetime(
+    expand = expansion(add = c(0, 600)),
+    date_breaks = "4 hours",
+    date_labels = "%H:%M"
+  ) +
+  scale_color_manual(values = wes_palette("Moonrise3")) +
+  labs(
+    y = "Number of people"
+  ) +
+  theme_custom(8) +
+  theme(
+    axis.title.x = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom",
+    legend.key.width = unit(0.275, "cm"),
+    legend.key.height = unit(0.2, "cm")
+  )
+
+save_plot(
+  time_by_day_pl,
+  pdf_file = "results/data/no-people-over-time-by-day.png",
+  w = 16, h = 18
 )

### 2024-11-19 doc/paper/supplements.tex
Updates from Overleaf

@@
 %\titleformat{\section}{\normalfont\Large\bfseries}{Text \thesection.~#1}{1em}{}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{\Alph{figure}}
+\renewcommand{\thetable}{\Alph{table}}
 
@@
 
-\title{\LARGE\singlespacing{\textbf{Supplementary Information} \\ \medskip
+\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
 Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}}
@@
 Person-tracking data required processing because many tracks were interrupted, for example, if the person i)~moved outside the range of the video sensors and then re-entered, ii)~bowed heavily downwards, or iii)~was temporarily hidden behind another person or obstacle.
-In addition, moving objects such as opening doors may have been recognized as a person, introducing many extremely short tracks (total duration of less than 10\,s and distance less than 2\,m), which were removed.
+In addition, moving objects such as opening doors may have been recognised as a person, introducing many extremely short tracks (total duration of less than 10\,s and distance less than 2\,m), which were removed.
 
@@
 We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS), a quasi-Newton method, as implemented in the statistical software R\cite{Byrd1995SIAM}.
-The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
+The optimal solution is the outdoor air supply rate $Q$ that minimises the root mean square error.
 We also follow recommendations to fit the outdoor CO$_2$ mixing ratio with reasonable constraints on $CO^{\text{O}} = [300,600]$\,ppm\cite{Batterman2017IJERPH}.
@@
 We illustrate our modelling approach with an example.
-Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of (0.25\,m)$^2$.
+Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterise into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of (0.25\,m)$^2$.
 We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
@@
 We divide the clinic into three spaces: the waiting room (length = 10.55\,m $\times$ breadth = 5.50\,m $\times$ height = 3.00\,m), the corridor (length = 12.45\,m $\times$ breadth = 2.20\,m $\times$ height = 2.50\,m), and the TB room (length = 4.75\,m $\times$ breadth = 3.50\,m $\times$ height = 3.00\,m).
-We rasterize each room into a grid of cubic cells with a squared area of (0.25\,m)$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+We rasterise each room into a grid of cubic cells with a squared area of (0.25\,m)$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
 
 Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate spaces to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
-Nevertheless, we acknowledge the possibility for aerosolized \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
+Nevertheless, we acknowledge the possibility for aerosolised \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
 
@@
 The bacterial inactivation rate for \emph{Mtb} has not been precisely estimated.
-Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6\,h ($\lambda$ = 0.12).
+Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolised TB bacilli of 6\,h ($\lambda$ = 0.12).
 By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5\,min ($\lambda$ = 8.3).
@@
     \includegraphics{results/data/co2-levels-over-time.png}
-    \caption{CO$_2$ levels by clinic area over time for each study day.}
+    \caption{CO$_2$ levels by clinic area over time for each study day. Raw data at 1~min intervals averaged over a ten minute time window to reduce noise.}
     \label{fig:co2-levels}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/no-people-over-time-by-day.png}
+    \caption{Number of people by clinic area over time for each study day. Raw data at 1~second intervals  averaged over a ten minute time window to reduce noise.}
+    \label{fig:people-over-time-by-day}
 \end{figure}
@@
 The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
-Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~physicochemical properties of the pathogen, and the (3)~infectiousness of the source and susceptibility of the host.
+Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorised into (1)~environmental factors, (2)~physicochemical properties of the pathogen, and the (3)~infectiousness of the source and susceptibility of the host.
 

### 2024-11-19 doc/paper/manuscript-revised.tex

--- 2024-06-03 doc/paper/manuscript.tex
@@
 
-\noindent\noindent\textbf{Supplementary Information:} Includes supplementary text, tables and figures.
+\noindent\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
 
-\noindent\textbf{Supplementary Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
+\noindent\textbf{S2 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.
+
+\clearpage
+
+\noindent\large{\textbf{Author Summary}}
+
+\noindent In crowded indoor settings in high tuberculosis burden countries, there is a high risk of tuberculosis transmission.
+We developed a model to study how tuberculosis is spread in a South African clinic during the COVID-19 pandemic when masks were mandatory and ventilation was improved.
+We tracked the movements of clinic attendees with video sensors and matched them with clinical records to find where and when infectious tuberculosis patients were present.
+We used this information to estimate the concentration of infectious particles in the air and the risk of infection for individuals.
+Over five days, the sensors tracked nearly 1,500 people in the clinic.
+Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room.
+The average risk of infection per person was found to be 0.1\%.
+The risk increased with more close contact with infectious patients and longer time spent in the clinic.
+Without masks, the risk was 1.3 times higher.
+With lower ventilation rates, similar to pre-pandemic levels, the risk was 2.2 times higher.
+This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.
 
@@
 
-% \begin{information}\normalfont
-% \noindent\textbf{Running head}: SARS-COV-2 transmission in schools and effect of air cleaners
-
-% %\noindent\textbf{Subject categorization}: 6.20  Indoor Air; 10.11 Pediatrics: Respiratory Infections
-
-% \noindent\textbf{Word count}: \maincount{manuscript}words, abstract \abstractcount{manuscript}words (max. 500), title 157 chars (max. 200)
-
-% %\noindent\textbf{Inserts:} 2 tables, 6 figures, 44 references
-
-% \vspace{1em}
-
-% \noindent\textbf{Funding}
-
-% \noindent This study is funded by the Multidisciplinary Center for Infectious Diseases, University of Bern, Bern, Switzerland.
-NB, LF, and ME are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through cooperative agreement 5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation (grant 32FP30-189498).
-\medskip
-
-% \noindent\textbf{Contributions}
-
-% \noindent Conception and design: NB, LF.
-Epidemiological and environmental data collection: NB, PJ, TS, LF.
-Laboratory data collection: PB, LFu.
-Additional data collection: TH.
-Statistical analysis: NB, KZ.
-Genomic analysis: LB, LFu.
-Paper draft: NB, LF, ME.
-All authors reviewed and approved the final version of the manuscript.
-
-% \par
-% \end{information}
-
-%TC:newcounter abst Words in abstract
-%TC:envir abstract [] abst
 \begin{abstract}\normalfont
-The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics.
-We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation.
-We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased.
+The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics.
+We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation.
+We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation was increased.
 We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
-We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
-Over five days, video sensors detected 1,438 clinic attendees.
+We modelled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
+Over five days, video sensors tracked 1,438 clinic attendees.
 CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was 0.1\% (95\%-credible interval [CrI] 0.01\%$-$0.1\%).
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) with pre-pandemic ventilation rates.
-Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
-
-% \noindent\textbf{Background:} Tuberculosis (TB) caused by \emph{Mycobacterium tuberculosis (Mtb)} is strictly airborne and thus the risk of transmission is higher in crowded indoor environments, especially healthcare facilities.
-Existing approaches to modeling the risk of airborne transmission typically assume that the airspace is well mixed, not considering that proximity to infectious individuals may increase the risk of transmission.
-We aimed to develop a  spatiotemporal model of airborne \emph{Mtb} transmission using environmental, clinical, and patient movement data to estimate TB transmission during the COVID-19 pandemic in healthcare facilities and to evaluate the impact of infection control measures.
-
-% \noindent\textbf{Methods:} We collected environmental (CO$_2$ levels), clinical (patient visits and TB disease status), and patient movement data (anonymous patient tracking from video sensors) in a primary care clinic in Cape Town, South Africa, for five days in October/November 2021 during the COVID-19 pandemic when universal use of masks was compulsory and ventilation was increased.
-We matched patient movements with clinical records to identify the spatiotemporal location of  infectious TB patients.
-We modeled the spatiotemporal concentration of infectious doses (quanta) and estimated the patient-specific risk of infection using the Wells-Riley equation.
-
-% \noindent\textbf{Results:} Clinical records registered 894 patients, four of whom were diagnosed with TB.
-Video sensors detected 1,438 clinic attendees with a median stay of 25\,min (interquartile range [IQR] 13\,min$-$46\,min).
-CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm), compatible with high ventilation rates.
-The modeled risk of infection per clinic attendee was 0.059\% (95\%-credible interval [CrI] 0.001\%$-$0.402\%).
-The risk would have increased 1.28-fold (95\%-CrI 1.19$-$1.36) without mask use and 1.74-fold (95\%-CrI 0.47$-$5.34) with lower, pre-pandemic ventilation rates.
-Furthermore, the individual risk was significantly associated with the number of close-contact encounters and time spent in the clinic.
-
-% \noindent\textbf{Conclusions:} Our novel spatiotemporal transmission model estimated the risk of \emph{Mtb} transmission in a South African clinic as low during the COVID-19 pandemic, mainly due to infection control measures.
-Our spatiotemporal modeling could be used to identify high-risk areas and evaluate the impact of infection control measures in clinics.
+The estimated risk of infection per clinic attendee was 0.1\% (80\%-credible interval [CrI] 0.01\%$-$0.1\%).
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) higher with pre-pandemic ventilation rates.
+Spatiotemporal modelling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
 
@@
 Tuberculosis (TB) remains one of the leading causes of death globally, with the South African region bearing a disproportionate share of the TB burden.
-After two years of COVID-19-related disruption of TB prevention and control measures, the negative impact of the COVID-19 pandemic has been receding, as documented in the World Health Organization (WHO) 2023 global TB report\cite{WHO2023TBReport}.
+After two years of COVID-19-related disruption of TB prevention and control measures, the negative impact of the COVID-19 pandemic is receding, as documented in the World Health Organization (WHO) 2023 global TB report\cite{WHO2023TBReport}.
 However, progress was insufficient to meet global TB targets\cite{WHO2023TBReport}.
@@
 Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Healthcare facilities are high-risk setting because infectious individuals are more likely to visit a clinic\cite{McCreesh2020IJTLD}.
-A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+Healthcare facilities are high-risk settings \deleted{because infectious individuals are more likely to visit clinics}\added{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}.
+A modelling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
@@
 The Wells-Riley model and its variations\cite{Rudnick2003IndoorAir} assume a well-mixed airspace, meaning that the quanta concentration is the same throughout the room.
-However, this concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to these individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
+However, the concentration is typically higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}, and previous studies have shown that the risk of TB infection is associated with proximity to infectious individuals\cite{Ko2004RiskAnal,Kenyon1996NEJM}.
 
@@
 We measured indoor CO$_2$ levels to model the spatial diffusion and removal of infectious quanta over time and estimated the risk of infection per clinic attendee.
-Finally, we assessed the impact of infection control measures during the study in response to the COVID-19 pandemic.
+Finally, we evaluated the impact of infection control measures during the study in response to the COVID-19 pandemic.
 
@@
 The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
+\added{The clinic was naturally ventilated by opening all doors and windows throughout the day.}
 
@@
 Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
-On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days.
+On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days (\added{Fig.~\zref{fig:co2-levels} in \supp}).
 We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study.
@@
 
-We used an anonymous personal tracking system using video sensors (Xovis, Zollikofen, Switzerland) to monitor the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}).
+We used an anonymous person tracking system using video sensors (Xovis, Zollikofen, Switzerland) to track the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}).
 The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person.
-To address this, we created an R Shiny application (\supp~Fig.~\zref{fig:shiny-app}) to rejoin interrupted tracks as described in \supp~Section~\zref{sec:setting-and-data}.
+People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognised as a person.
+To address this, we created an R Shiny application (Fig.~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp.
 We also labeled the tracks to identify the movements of clinic staff.
-\supp~Fig.~\ref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.
+\Cref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.
 
@@
 
-\subsection*{Spatiotemporal modeling approach}
+\subsection*{Spatiotemporal modelling approach}
 
-Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of infection $P$ as
+Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of \emph{Mtb} infection $P$ as
 \begin{align}
@@
 This way, the spatiotemporal location of infectious patients who generate quanta could be determined.
-A detailed description of our extension is provided in \supp~Section~\zref{sec:spattemp-model}, together with an illustrative example using simulated data (\supp~Fig.~\zref{fig:toy-example}).
+A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig.~\zref{fig:toy-example} in \supp).
 In the following, we describe the six main components of our modelling approach (\Cref{fig:modeling-flow}).
@@
 We allowed for a maximum delay of 15\,min between the track's timestamp in the registration area and the entry in the clinical database.
-All diagnosed TB patients could be matched with person movements this way.
+This way, all diagnosed TB patients could be matched with person movements.
 
@@
 The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
+Therefore, we modelled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modelled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
 Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
@@
 
-(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
+(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-We set the gravitational settling rate to zero and ignore sedimentation because particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of just $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}.
+We modelled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\deleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\deleted{quanta }h$^{-1}$\,$-$\,7.1\,\deleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+\deleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \deleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \added{which we modelled using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$.
+Assuming a drop height of 1.7\,m, this results in a gravitational settling rate of 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
 
@@
 
-(6)~Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as 
+(6)~\deleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\added{The quanta concentration was computed separately for the waiting room, corridor, and TB room.
+Each room was discretised into a grid of cubic cells, with each cell covering a squared area of 0.0625\,m$^2$.
+We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as}
 \begin{align}\label{eq:spattemp-N}
-    \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
-concn.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda) \cdot 1\text{s}\right)}_{\text{removal}} ~.
+    % \deleted{\underbrace{N_{t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
+conc.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda\cdot 1\text{s}\right)}_{\text{removal}} ~.}
+     \added{\underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q}_{\text{generation}} \cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k\right)}_{\text{removal}} ~.}
 \end{align}
-where $D$ is the diffusion constant (m$^2$ s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta\,s$^{-1}$).
+where $D$ is the diffusion constant (m$^2$\,s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), \deleted{$I_t$ is the number of infectious individuals in space, } \added{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \deleted{$AER$ is the}\added{$AER_{r,t}$ is the room-specific} air change rate (\deleted{quanta}s$^{-1}$), $\lambda$ is bacterial inactivation rate (\deleted{quanta}s$^{-1}$), \added{and $k$ is the gravitational settling rate (s$^{-1}$)}.
+
 The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
 \begin{align}
-    P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),
+    % \deleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
+    \added{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\right),}
 \end{align}
-where $N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
-We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m.
+where \deleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\added{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
 
-\subsection*{Monte Carlo simulation}
+\subsection*{\deleted{Monte Carlo simulation}}
 
-Simulations are described in detail in \supp~Section~\zref{sec:estimation}.
-In brief, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
-Each room was rasterized into a grid of cubic cells with an area of (0.25\,m)$^2$.
-The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
+\deleted{Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
+Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
+Each room was rasterised into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
+The quanta concentration was modelled during clinic hours from 8\,am to 4\,pm and updated every second.
 We used a Monte-Carlo simulation approach to consider uncertainty in modelling parameters.
-In each simulation, we sampled uncertain modelling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
+In each simulation, we sampled uncertain modelling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.}
 
@@
 To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask-wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie assuming an air change rate of 6~air changes per hour, corresponding to the average rate in the pilot study in 2019 in the same clinic\cite{Zurcher2022JID}.
-Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
+Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolised \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
 Except for the scenario-specific modifications, all other modelling assumptions and parameters were identical in each Monte Carlo simulation.
@@
 
-We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
-Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).
-The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
-We used a Bayesian Beta regression model to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact with, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
+We summarised clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
+\added{We simulated the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modelling parameters (Section~\zref{sec:estimation} in \supp).
+In each simulation, we computed the risk of infection based on the cumulative quanta exposure during each susceptible individual's clinic visit.
+We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations.
+The scenario-specific infection risk distribution was shown with a histogram and summarised with the mean and 80\%-credible interval (CrI).} \deleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
+The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarised with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorised as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
+We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modelled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
 Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1).
-Additional results not reported in the manuscript are shown in \supp~Section~\zref{sec:additional-results}.
+Additional results not reported in the manuscript are shown in Section~\zref{sec:additional-results} in \supp.
 
@@
 Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
-Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm \deleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+\added{CO$_2$ levels slightly varied within and between clinic days (Fig.~\zref{fig:co2-levels} in \supp), largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:people-over-time-by-day} in \supp) and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
 The mean risk of infection per clinic visit was 0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%), corresponding to a risk of 0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%).
-\supp~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
+S2~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
 It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
@@
 
-The modeled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
+The modelled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
 A doubling of close-contact encounters increased the odds of infection by 23\% (OR 1.23, 95\%-CrI 1.09$-$1.37).
@@
 The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.12\% (80\%-CrI 0.02\%$-$0.13\%), a 30\% increase (OR 1.3, 80\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.19\% (80\%-CrI 0.02\%$-$0.20\%), a 120\% increase (OR 2.2, 80\%-CrI 0.9$-$5.2).
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.24\% (80\%-CrI 0.03\%$-$0.24\%), a 180\% increase (OR 2.8, 80\%-CrI 1.1$-$6.6).
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.12\% (80\%-CrI 0.02\%$-$0.13\%), an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.19\% (80\%-CrI 0.02\%$-$0.20\%), an increase of 120\% (OR 2.2, 95\%-CrI 0.9$-$5.2).
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.24\% (80\%-CrI 0.03\%$-$0.24\%), an increase of 180\% (OR 2.8, 95\%-CrI 1.1$-$6.6).
 The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
 
-In sensitivity analyses, we varied assumptions about the number of infectious people in the clinic.
-Assuming the number of infectious TB patients reflects the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher (0.12\%, 80\%-CrI 0.03\%$-$0.19\%).
-Further, assuming that both diagnosed TB patients and clinic attendees suspected of TB infection had been infectious, the mean risk of infection would have been two times higher (0.22\%, 80\%-CrI <0.01\%$-$0.43\%).
+In sensitivity analyses, we varied assumptions about the number of infectious persons in the clinic.
+Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher (0.12\%, 80\%-CrI 0.03\%$-$0.19\%).
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high (0.22\%, 80\%-CrI <0.01\%$-$0.43\%).
 
@@
 % summary
-We expanded the Wells-Riley equation to a spatiotemporal model based on environmental, clinical, and patient movement data.
-We applied the model to data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
+We expanded the Wells-Riley equation to a novel spatiotemporal model based on environmental, clinical, and patient movement data.
+We applied the model to real-world data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
 We estimated an average risk of infection per clinic attendee of 0.1\%.
 We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
-Without infection control measures (masks and high natural ventilation), the risk of infection would have been almost three times as high.
-Our spatiotemporal model can be readily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
+Without infection control measures (masks and high natural ventilation), the risk of infection would have been almost three times higher.
+Our spatiotemporal model can be easily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
 
@@
 Our spatiotemporal model is an extension of the Wells-Riley model, which allows the quanta concentration to vary both over time and space, and considers that the density of infectious particles is initially higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
-We found a significant association between the modeled risk of infection and the number of close-contact encounters, and the time spent in the clinic.
+We found a significant association between the modelled risk of infection and the number of close-contact encounters, and the time spent in the clinic.
 More close contacts and extended visits make exposure to high doses of infectious particles more likely.
@@
 South Africa has a high burden of TB and multi-drug-resistant TB\cite{WHO2023TBReport}.
-Nevertheless, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
+Nevertheless, we estimated a lower risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
 Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 2.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
-The difference can be partly attributed to the infection control measures that were in place during the SARS-CoV-2 pandemic.
+The difference can be mainly attributed to the infection control measures that were in place during the COVID-19 pandemic.
 We estimated that without mask use and with pre-pandemic ventilation conditions, the risk of infection would have been more than twice as high.
-Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic.
-Further, different modelling assumptions could have contributed.
+Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in the model.
+Other modelling assumptions may also have contributed.
 The risk estimate from our previous study before the COVID-19 pandemic\cite{Zurcher2022JID} was based on TB prevalence in the South African population, rather than the TB status documented in the clinical data.
-Modelling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modeled risk of transmission.
+Modelling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modelled risk of transmission.
 
@@
 Other studies suggest that mask-wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
-Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less striking.
-The reason is that the effect of mask-wearing is compounded by the impact of very high ventilation rates, which determine both the diffusion and removal of infectious quanta in the indoor space.
+Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less pronounced.
+The reason for this is that the effect of mask use is compounded by the effect of very high natural ventilation rates, which determine both the diffusion and the removal of infectious quanta in the indoor space.
 The air change rates in the clinic typically ranged from 10 to 20 air changes per hour, which resulted in a quick diffusion and removal of infectious quanta from the indoor air.
-To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
-Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely due to a better ventilation, with an air change rate comparable to the one measured in the primary care clinic of our study.
-We stress that our results do \emph{not} suggest that masks are ineffective but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
+To put this into perspective, we recently modelled the risk of \emph{Mtb} transmission in schools by comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
+Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely because of better natural ventilation, with an air change rate comparable to the one measured in the primary care clinic in our study.
+We emphasise that our results do \emph{not} suggest that masks are ineffective -- there is a large body of evidence for their effectiveness -- but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 
@@
 Our model considered ventilation conditions and patient movements.
-However, other factors such as variation in infectiousness and susceptibility, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see \supp~Section~\zref{sec:depth-discussion} for a detailed discussion of these factors).
+However, other factors such as variation in infectiousness, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see Section~\zref{sec:depth-discussion} in \supp for a detailed discussion of these factors).
 Furthermore, computational fluid dynamics (CFD) models have been developed to examine airflow as a factor influencing the spatial spread of airborne pathogens\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
 However, CFD models are computationally expensive and primarily used in experimental and static environments, making their application to real-world settings such as clinics difficult.
-Moreover, our modelling approach required computational efficiency because a many Monte Carlo simulations are needed to reflect uncertainty in modelling parameters.
+Moreover, our modelling approach required computational efficiency because many Monte Carlo simulations are needed to reflect uncertainty in modelling parameters.
 
@@
 First, tracking patient movements with video sensors was challenging.
-It required considerable data processing and many tracks were interrupted and could not be reconnected.
+It required considerable data processing and many interrupted tracks could not be reconnected.
 Therefore, some clinic attendees may have had longer visits and the risk of infection may have been underestimated.
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
+\added{For example, the waiting room and corridor were separated to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels, yet the door between these areas was typically open, so that aerosolised \emph{Mtb} (quanta) could spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
 However, airflow is very difficult to measure.
@@
 % conclusions
-In conclusion, our spatiotemporal transmission model considers clinical, environmental, and movement data to estimate the risk of infection per individual in the indoor space. We
-showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa was related to the number of close contact encounters and time spent in the clinic.
+In conclusion, our spatiotemporal transmission model incorporates clinical, environmental, and person movement data to estimate the risk of infection per person in the indoor space.
+It can be easily adapted to model airborne transmission risk of other respiratory infections such as SARS-CoV-2.
+We showed that the risk of \emph{Mtb} transmission in a primary care clinic in South Africa was related to the number of close contact encounters and time spent in the clinic.
 The risk of infection was low under pandemic conditions with strict infection control measures in place.
-Our spatiotemporal model could be used in future work to assess the impact of infection control measures, specifically interventions targeting patient flows to reduce prolonged close contact.
-Finally, our study underscores the paramount importance of effective ventilation in primary care clinical settings.
+Our spatiotemporal model could be used in future work to evaluate the impact of infection control measures, specifically interventions targeting patient flows to reduce prolonged close contact.
+Finally, our study highlights the paramount importance of effective ventilation in primary care clinical settings.
 
@@
 All other data are included in the manuscript and accompanying supporting information.
-Code used for preprocessing, modelling and analysis, and descriptive and simulation results are available from GitHub via \url{https://github.com/nbanho/clinic-transmission}.
+Code used for preprocessing, modelling and analysis, descriptive and simulation results are available on GitHub via \url{https://github.com/nbanho/clinic-transmission}.
 
@@
 The funders had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
+
+\section*{Competing interests}
+The authors have declared that no competing interests exist.
 

### 2024-11-21 analysis/patient-tracking.Rmd
gitkeep simulation directories

@@
   labs(
-    y = "Number of tracks",
+    y = "Number of people",
     title = "b"

### 2024-11-21 analysis/qctr.Rmd
gitkeep simulation directories

@@
 
-```{r}
+```{r libraries}
 library(tidyverse)
@@
 # building
-cellSize <- 250
 source("preprocessing/prep_building-rasterize.R")
@@
 
+Generate an animated ggplot showing the quanta concentration in the waiting room 
+between 9:41am and 9:45am on October 25 as one infectious individual enters.
+
 ```{r illustration}
-# data
+# date
 mth <- 10
 dy <- 25
+
+# simulation parameters
 sim <- 2
@@
 )
+
+# tracking data
 track_1025 <- readRDS(
@@
 )
-aer_df <- readRDS("data-clean/environmental/air-change-rate.rds") %>%
+
+# air exchange rate
+aer_df <- aer %>%
   filter(
@@
 
-# simulation
+# simulate quanta concentration in clinic
 ct_by_room <- list()
 for (room in c("Waiting room", "Corridor", "TB room")) {
-  print(paste0("- ", room))
-
-  # room setup
   if (room == "Waiting room") {
-    # initial quanta concentration matrix
-    c0 <- sP_to_matrix(waiting_room)
-
-    # (no) TB patients in room
+    c0 <- wrMat
     track_tb <- filter(track_tb_all, is_waitingroom)
-
-    # room dimension
-    roomDim <- dimWR
+    room_dim <- dimWR
   } else if (room == "Corridor") {
-    # initial quanta concentration matrix
-    c0 <- sP_to_matrix(corridor)
-
-    # TB patients in room
+    c0 <- cdMat
     track_tb <- filter(track_tb_all, is_passage)
-
-    # room dimension
-    roomDim <- dimCD
+    room_dim <- dimCD
   } else if (room == "TB room") {
-    # initial quanta concentration matrix
-    c0 <- sP_to_matrix(tb_room)
-
-    # TB patients in room
+    c0 <- tbMat
     track_tb <- filter(track_tb_all, is_tbroom)
-
-    # room dimension
-    roomDim <- dimTB
-  } else {
-    stop("Error: Room invalid.")
+    room_dim <- dimTB
   }
-
-  # add quanta generation rate per second by activity level
   track_tb <- track_tb %>%
@@
     )
-
-  # quanta conc in the morning
-  print("-- morning")
   track_tb_morn <- filter(track_tb, time <= midday)
@@
   ))
-
   ct_by_room[[room]] <- stm(
@@
     pd = 0,
-    vol = prod(roomDim),
+    vol = prod(room_dim),
     aer = aer_df$aer[aer_df$room == room & aer_df$daytime == "Morning"] / 3600,
     lambda = param_1025$viral_inact[param_1025$sim == sim] / 3600,
-    deposit = 0
+    deposit = param_1025$settling_rate[param_1025$sim == sim] / 3600
   )
@@
 
-# get subset
+# subset quanta concentration of interval
 subset_qc <- function(t) {
@@
 
-track_tb_all %>%
-  filter(between(
-    time,
-    as.POSIXct(paste(date, "09:41:00")),
-    as.POSIXct(paste(date, "09:45:00"))
-  )) %>%
-  ggplot(aes(x = x, y = y, color = factor(patient_id))) +
-  geom_path()
-
 tt_selected <- seq.POSIXt(
@@
 
-# animated plot
-qc_anim_pl <- ggplot(
-  data = left_join(
+sub_quanta <- left_join(
   fortify(waiting_room),
@@
   by = c("id")
-  ),
+) %>%
+  dplyr::select(t, long, lat, group, value)
+sub_quanta$time <- as.POSIXct(paste(date, "09:41:00")) +
+  seconds(sub_quanta$t)
+
+# subset tracking data of interval
+sub_track <- track_tb_all %>%
+  filter(between(
+    time,
+    as.POSIXct(paste(date, "09:41:00")),
+    as.POSIXct(paste(date, "09:45:00"))
+  )) %>%
+  mutate(t = t - min(tt_selected_idx))
+sub_track_rep <- tibble()
+for (t_i in 0:max(sub_track$t)) {
+  new_frame <- sub_track %>%
+    filter(t <= t_i) %>%
+    arrange(t) %>%
+    mutate(t = t_i) %>%
+    dplyr::select(t, x, y)
+  if (nrow(new_frame) == 0) {
+    new_frame <- data.frame(t = t_i, x = NA, y = NA)
+  }
+  sub_track_rep <- rbind(sub_track_rep, new_frame)
+}
+sub_track_rep$time <- as.POSIXct(paste(date, "09:41:00")) +
+  seconds(sub_track_rep$t)
+
+# specify animated ggplot
+qc_anim_pl <- ggplot() +
+  geom_polygon(
+    data = sub_quanta,
     mapping = aes(x = long, y = lat, group = group, fill = value)
   ) +
-  geom_polygon() +
+  geom_path(
+    data = sub_track_rep,
+    mapping = aes(x = x, y = y),
+    color = "blue",
+    linewidth = 2
+  ) +
   coord_cartesian(expand = FALSE) +
@@
     axis.text = element_blank(),
-    axis.title = element_blank(),
     axis.ticks = element_blank(),
@@
     ),
-    panel.grid.major = element_blank(), panel.grid.minor = element_blank()
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    legend.key.width = unit(2, "cm")
   ) +
-  transition_time(t) +
+  gganimate::transition_time(time) +
   labs(
-    subtitle = "Time (secs): {frame_time}",
+    subtitle = "Time: {frame_time}",
     title = "Quanta concentration in the waiting room on Oct 25 between 9:41 and 09:45am", # nolint
-    fill = expression("Quanta/m"^3 * " x 10"^-3)
+    caption = "Notes: Patient track in blue; jumps due to track interruptions.",
+    fill = expression("Quanta x 10"^-3 * " m"^-3),
+    x = "x length in m",
+    y = "y width in m"
   )
 
+# animate plot
 gganimate::animate(
@@
   duration = 30,
-  width = 1.850877 * 600,
+  width = (dimWR[1] / dimWR[2]) * 600,
   height = 600,
@@
 
-gganimate::anim_save("results/modeling/illustrated-quanta-conc-over-time.gif")
+# save animation
+gganimate::anim_save("results/modeling/STM - Supplementary Video.gif")
 ```
@@
     labs(
-      fill = expression("Mean quanta concentration (quanta m"^-3 * " " * "x 10"^-3 * ")")
+      fill = expression("Mean quanta concentration (quanta x 10"^-3 * " " * "m"^-3 * ")")
     ) +

### 2024-11-21 preprocessing/match_clinical-tracking.R

@@ -5,32 +5,33 @@ library(tidyverse)
 #### Matching function ####
 
 #' Matching clinical records to patient tracking data
-#' 
+#'
 #' @param sel_date date of tracking data
 #' @param min_time_in_cl minimum time spent in clinic in order for the id to be considered a clinical attendee
 #' @param min_time_at_re minimum time a track should have stayed at the reception
 #' @param max_tolerance minimum and maximum delay of clinical record after a patient was at the reception
-#' @param prio_diagnosed should only diagnosed TB patients be matched?  
-#' @param prio_suspected should only suspected TB patients be matched? 
+#' @param prio_diagnosed should only diagnosed TB patients be matched?
+#' @param prio_suspected should only suspected TB patients be matched?
 
 match_clinical_tracking <- function(
-    sel_date, 
-    min_time_in_cl = 5 * 60, 
-    min_time_at_re = 5, 
+    sel_date,
+    min_time_in_cl = 5 * 60,
+    min_time_at_re = 5,
     max_tolerance = c(0, 15 * 60),
     prio_diagnosed = T,
     prio_suspected = T) {
-  
   # clinical data
   clinical <- readRDS(paste0("data-clean/clinical/clinical-data.rds")) %>%
     filter(date == sel_date) %>%
-    rename(clinical_patient_id = patient_id,
-           clinical_record_time = date_time) %>%
+    rename(
+      clinical_patient_id = patient_id,
+      clinical_record_time = date_time
+    ) %>%
     group_by(clinical_patient_id) %>%
     arrange(clinical_record_time) %>%
     slice(1) %>%
     ungroup()
-  
+
   if (prio_diagnosed & prio_suspected) {
     clinical <- filter(clinical, is_infectious | was_suspected)
   } else if (prio_diagnosed) {
@@ -38,13 +39,13 @@ match_clinical_tracking <- function(
   } else if (prio_suspected) {
     clinical <- filter(clinical, was_suspected)
   }
-  
+
   # tracking data
   tracking <- readRDS(paste0("data-clean/patient-tracking/linked-and-imputed/", sel_date, ".rds")) %>%
     group_by(patient_id) %>%
     filter(as.numeric(difftime(last(time), first(time), units = "secs")) > min_time_in_cl) %>%
     ungroup()
-  
+
   # tracking: time at reception
[…]

### 2024-11-22 analysis/simulation-results.Rmd
cleanup simulation results

--- 2024-11-21 analysis/qctr.Rmd
@@
 library(tidyverse)
+library(parallel)
+library(data.table)
 library(reshape2)
+library(betareg)
 library(rstanarm)
@@
 source("models/stm-v2.R")
+source("utils/trans_risk.r")
 
@@
 
-## Main
+## Quanta concentration
 
-### Quanta conc
+Show the average quanta concentration of the main simulation by daytime in the clinic 
+and determine the corresponding transmission risk.
 
 ```{r}
+# main quanta simulation files
 main_qc <- data.frame(
-  files = list.files("simulations/main/",
+  files = list.files("simulations/main",
     full.names = TRUE, pattern = "quanta_conc-.*.txt$",
@@
 
-mean_qc <- function(file) {
-  df <- read.table(file)
-  colnames(df) <- c("sim", "x", "y", "value")
-  df <- df %>%
-    group_by(x, y) %>%
-    summarize(value = mean(value)) %>%
-    ungroup()
+# read average quanta concentration by daytime, room and study day
+read_quanta_table <- function(file, n_sim = 5000) {
+  df <- data.table::fread(
+    file,
+    col.names = c("sim", "x", "y", "value")
+  )
+  n_origin <- nrow(df)
+  df[, sim := as.integer(sim)]
+  df[!sim %in% 1:n_sim, sim := NA]
+  df <- na.omit(df)
+  df <- df[, .SD[1], by = .(sim, x, y)]
+  n_new <- nrow(df)
+
+  if (n_origin > n_new) {
+    warning(sprintf("File %s: %i defect rows removed", file, n_origin - n_new))
+  }
+
+  # check if simulations are complete
+  sims <- unique(df$sim)
+  missing_sims <- (1:n_sim)[!((1:n_sim) %in% sims)]
+  n_missing <- length(missing_sims)
+  if (n_missing > 0) {
+    warning(sprintf("File %s: %i simulations missing.", file, n_missing))
+  }
+
   return(df)
@@
 
-main_qc$value <- lapply(main_qc$files, mean_qc)
+main_qc$data <- mclapply(main_qc$files, read_quanta_table, mc.cores = 15)
 
+# compute daily mean quanta concentrations
+main_qc$mean_conc <- lapply(main_qc$data, function(df) {
+  df %>%
+    group_by(x, y) %>%
+    summarise(value = mean(value)) %>%
+    ungroup()
+})
+
+# compte overall average quanta concentration
 main_mean_qc <- main_qc %>%
-  unnest(value) %>%
+  unnest(mean_conc) %>%
   group_by(room, daytime, x, y) %>%
@@
 
-# example translation
-source("utils/trans_risk.r")
-IR_sit <- (.5 * IR("sitting", "male") + .5 * IR("sitting", "female")) / 3600
-WR_vol <- convert_dist(cellSize)^2 * dimWR[3]
+# corresponding transmission risk
+ir_sit <- (.5 * IR("sitting", "male") + .5 * IR("sitting", "female")) / 3600
+wr_vol <- (cellSize / 1e3)^2 * dimWR[3]
 wr_morn <- main_mean_qc %>%
@@
   max()
-round((1 - exp(-wr_morn * 3600 * IR_sit / WR_vol)) * 100, 3)
-round((1 - exp(-wr_noon * 3600 * IR_sit / WR_vol)) * 100, 3)
+round((1 - exp(-wr_morn * 3600 * ir_sit / wr_vol)) * 100, 3)
+round((1 - exp(-wr_noon * 3600 * ir_sit / wr_vol)) * 100, 3)
 
-# link to building
+# link quanta concentration to building
 main_mean_qc <- left_join(roomCoord,
@@
 
-# plot
+# plot quanta concentration
 mean_quanta_pl <- plot_spatial(
@@
     labs(
-      fill = expression("Mean quanta concentration (quanta x 10"^-3 * " " * "m"^-3 * ")")
+      fill = expression(
+        "Mean quanta concentration (quanta x 10"^-3 * " " * "m"^-3 * ")"
+      )
     ) +
@@
 
-### Trans risk
+## Transmission risk
+
+Compute the patient-specific mean risk of infection 
+and summarise the risk across patients for the main simulation.
 
 ```{r}
-average_trans_risk <- function(file) {
-  average_tr <- read.table(file) %>%
-    set_names(c("sim", "patient_id", "P")) %>%
+# main simulation files
+main_tr <- data.frame(
+  files = list.files("simulations/main",
+    full.names = TRUE, pattern = "trans-risk.txt$",
+    recursive = TRUE
+  )
+)
+
+# read transmission risk data
+read_risk_table <- function(file, n_sim = 5000) {
+  # load data and remove duplicates
+  df <- data.table::fread(
+    file,
+    select = 1:3,
+    col.names = c("sim", "patient_id", "P"),
+    fill = Inf
+  )
+  n_origin <- nrow(df)
+  df[, sim := as.integer(sim)]
+  df[!sim %in% 1:n_sim, sim := NA]
+  df[, patient_id := as.integer(patient_id)]
+  df[is.na(patient_id) | patient_id <= 0, patient_id := NA]
+  df[, P := as.numeric(P)]
+  df <- df[between(P, 0, 1)]
+  df <- na.omit(df)
+  df <- df[, .SD[1], by = .(sim, patient_id)]
+  n_new <- nrow(df)
+
+  if (n_origin > n_new) {
+    warning(sprintf("File %s: %i defect rows removed", file, n_origin - n_new))
+  }
+
+  # check if simulations are complete
+  sims <- unique(df$sim)
+  missing_sims <- (1:n_sim)[!((1:n_sim) %in% sims)]
+  n_missing <- length(missing_sims)
+  if (n_missing > 0) {
+    warning(sprintf("File %s: %i simulations missing.", file, n_missing))
+  }
+
+  # add unique mthdy and patient_id
+  df <- df %>%
     mutate(
@@
       patient_id = paste(mthdy, patient_id, sep = "-")
-    ) %>%
+    )
+
+  return(df)
+}
+
+main_tr$data <- lapply(main_tr$files, read_risk_table)
+
+# compute mean risk per patient across simulations
+patient_risk <- function(dat) {
+  dat %>%
     group_by(patient_id) %>%
@@
     )
-  return(average_tr)
 }
 
-# average risk
-main_tr <- data.frame(
-  files = list.files("simulations/main/",
-    full.names = TRUE, pattern = "trans-risk.txt$",
-    recursive = TRUE
-  )
-)
-main_tr$value <- lapply(main_tr$files, average_trans_risk)
+main_tr$value <- lapply(main_tr$dat, patient_risk)
 
@@
 
-# annotation
-annotate_average_trans_risk <- function(tr) {
+# compute overall risk across patients
+overall_risk <- function(tr) {
   roi_annon <- tr %>%
@@
 
-main_annon <- annotate_average_trans_risk(main_tr)
-
+main_annon <- overall_risk(main_tr)
 main_annon
@@
 
-### Association
+## Association with patient risk factors
+
+Estimate the association of the patient-specific risk of infection with 
+time in clinic, number of close contacts, and time spent in close contact.
 
@@
 
+Compare the modeled transmission risk with hypothetical scenarios 
+to assess the effecs of implemented IPC measures (mask wearing and ventilation).
+
 ```{r additional-analyses}
-# files
-files_nm <- list.files("simulations/nomasks/",
+# IPC scenario files
+files_nm <- list.files("simulations/nomasks",
   full.names = TRUE, pattern = "trans-risk",
@@
 )
-files_vent <- list.files("simulations/vent/",
+files_vent <- list.files("simulations/vent",
   full.names = TRUE, pattern = "trans-risk",
@@
 )
-files_nmvent <- list.files("simulations/nomasksvent/",
+files_nmvent <- list.files("simulations/nomasksvent",
   full.names = TRUE, pattern = "trans-risk",
@@
 
-# average risk
-nm_tr <- map_dfr(files_nm, average_trans_risk)
-vent_tr <- map_dfr(files_vent, average_trans_risk)
-nmvent_tr <- map_dfr(files_nmvent, average_trans_risk)
+# mean risk per patient
+nm_tr <- map_dfr(files_nm, function(f) patient_risk(read_risk_table(f)))
+vent_tr <- map_dfr(files_vent, function(f) patient_risk(read_risk_table(f)))
+nmvent_tr <- map_dfr(files_nmvent, function(f) patient_risk(read_risk_table(f)))
 
+# overall risk
+nm_annon <- overall_risk(nm_tr)
+vent_annon <- overall_risk(vent_tr)
+nmvent_annon <- overall_risk(nmvent_tr)
+nm_annon
+vent_annon
+nmvent_annon
+
+# patient risk by category
 nm_tr %>%
@@
 
+# factor difference between scenarios
 main_tr %>%
@@
 
-# combine
+# combine scenario-specific simulation results
 tr_comp <- rbind(
-  main_tr %>% mutate(scenario = "a   Mask use and high ventilation"),
-  nm_tr %>% mutate(scenario = "b   No mask use and high ventilation"),
-  vent_tr %>% mutate(scenario = "c   Mask use and normal ventilation"),
-  nmvent_tr %>% mutate(scenario = "d   No mask use and normal ventilation")
+  main_tr %>% dplyr::select(-data) %>%
+    mutate(scenario = "a   Mask use and high ventilation"),
+  nm_tr %>%
+    mutate(scenario = "b   No mask use and high ventilation"),
+  vent_tr %>%
+    mutate(scenario = "c   Mask use and normal ventilation"),
+  nmvent_tr %>%
+    mutate(scenario = "d   No mask use and normal ventilation")
 ) %>%
@@
 
-# annotation
-nm_annon <- annotate_average_trans_risk(nm_tr)
-vent_annon <- annotate_average_trans_risk(vent_tr)
-nmvent_annon <- annotate_average_trans_risk(nmvent_tr)
-
+# combine scenario-specific annotations (overall risk)
 annon_comp <- rbind(
@@
 
-# plot mean
+# plot distribution with annotation
 mean_roi_comp_pl <- ggplot(
@@
 
-## Additional analysis
+## Prevalence comparison
+
+Compare the modelled transmission risk with hypothetical scenarios 
+to assess the impact of assumptions regarding TB prevalence (suspected patients, national prevalence).
 
 ```{r} 
-files_prev <- list.files("simulations/prevalence/",
+# prevalence scenario files
+files_prev <- list.files("simulations/prevalence",
   full.names = TRUE, pattern = "trans-risk",
@@
 )
-files_susp <- list.files("simulations/suspected/",
+files_susp <- list.files("simulations/suspected",
   full.names = TRUE, pattern = "trans-risk",
@@
 
-prev_tr <- map_dfr(files_prev, average_trans_risk)
-susp_tr <- map_dfr(files_susp, average_trans_risk)
+# mean risk per patient
+prev_tr <- map_dfr(files_prev, function(f) patient_risk(read_risk_table(f)))
+susp_tr <- map_dfr(files_susp, function(f) patient_risk(read_risk_table(f)))
 
+# overall risk
+prev_annon <- overall_risk(prev_tr)
+susp_annon <- overall_risk(susp_tr)
+prev_annon
+susp_annon
+
+# combine scenario-specific simulation results
 tr_assum <- rbind(
@@
 
-prev_annon <- annotate_average_trans_risk(prev_tr)
-susp_annon <- annotate_average_trans_risk(susp_tr)
-
+# combine scenario-specific annotations (overall risk)
 annon_assum <- rbind(
@@
 
-# plot mean
+# plot distribution with annotations
 mean_roi_assum_pl <- ggplot(
@@
     legend.key.width = unit(.4, "cm"),
-    panel.spacing = unit(2, "cm"),
+    panel.spacing = unit(.5, "cm"),
     strip.text = element_text(hjust = 0)

### 2024-11-22 tests/test_stm_v2-with-spatial-prop-dist.R

@@ -1,124 +1,0 @@
-#' Room:
-#' We assume a room with length = 10m, breadth = 6m, and height = 3m.
-#' We divide the room into a grid with cell length = 0.1m, i.e. 100 x 60 matrix.
-#' 
-#' Time:
-#' We run our simulation for 1h with a time step of 1s.
-#' 
-#' Quanta emission:
-#' There are two quanta emission sources:
-#' 1. One in the center in the first five minutes.
-#' 2. One near the top-right edge between minute 30 and 35.
-#' 
-#' The quanta emission rate is 4 quanta/hour.
-#' We assume an exponential decrease with distance to source of emission.
-#' 
-#' Quanta removal:
-#' Quanta is removed only through air exchange.
-#' We assume a constant air exchange rate of 3 air changes per hour.
-#' 
-#' Quanta diffusion:
-#' We assume radial diffusion at a constant rate of 0.05m2/min. 
-#' 
-
-
-#### Libraries ####
-library(tidyverse)
-source("utils/trans_risk.r")
-source("utils/spatial.R")
-
-
-#### Setup ####
-
-# room
-Lx <- 10
-Ly <- 6
-Lz <- 3
-V <- 10 * 6 * 3
-Vxy <- 1
-dx <- dy <- sqrt(Vxy/Lz)
-nx <- Lx / dx
-ny <- Ly / dy
-
-# initial quanta concentration
-C0 <- matrix(0, nrow = ny, ncol = nx)
-
-# time
-TT <- 60 * 60
-dt <- 1
-
-# quanta emission rate
-q <- 4 / 3600
-
-# weights to source
-W1 <- matrix(NA, nrow = ny, ncol = nx)
-for (i in 1:ny) {
-  for (j in 1:nx) {
-    W1[i,j] <- euclidean(floor(ny/2), i, floor(nx/2), j) * dx
-  }
-}
-W1 <- apply(W1, c(1,2), dIc, mu = 1)
-W1 <- W1 / sum(W1)
-
-W2 <- matrix(NA, nrow = ny, ncol = nx)
-for (i in 1:ny) {
-  for (j in 1:nx) {
-    W2[i,j] <- euclidean(ny, i, nx, j) * dx
-  }
-}
-
-# weights
-W2 <- apply(W2, c(1,2), dIc, mu = 1)
-W2 <- W2 / sum(W1)
-
-# quanta removal rate
-aer <- 1 / 3600
-lambda <- 0.84 / 3600
-
[…]

### 2024-11-22 tests/test_stm_v2.R

@@ -101,7 +101,7 @@ quanta_concn_pl <- ct_df %>%
   )
 
 save_plot(quanta_concn_pl,
-  pdf_file = "tests/stm_v2-toy_example.png",
+  pdf_file = "tests/stm_v2-toy_example-A.png",
   w = 16, h = 13
 )
 

### 2024-11-22 doc/paper/manuscript-revised.tex

@@
 Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
-On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days (\added{Fig.~\zref{fig:co2-levels} in \supp}).
+On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days \added{(see time-varying CO$_2$ levels and room occupancy by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
+
 We extracted clinical data from the electronic patient registry for all patients who attended the clinic during the study.
@@
 
-(3)~We calculated room occupancy from the video sensor data and combined it with the CO$_2$ measurements to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
+(3)~We calculated \added{time-varying} room occupancy from the video sensor data and combined it with the CO$_2$ measurements \added{(Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp)} to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
 
@@
 We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
+\added{We assumed no effect of mask use on the inhalation rate of quanta\cite{McCreesh2021BMJGlobalHealth}.}
 
@@
 Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c).
-Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm \deleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
-\added{CO$_2$ levels slightly varied within and between clinic days (Fig.~\zref{fig:co2-levels} in \supp), largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:people-over-time-by-day} in \supp) and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
+\added{Time spent in the corridor and TB room varied considerably as many patients just walked through the corridor and the TB room was reserved primarily for diagnosed or suspected TB patients.} Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) \deleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+\added{CO$_2$ levels slightly varied within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp) and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
 
+There is broad consensus that tuberculosis is a strictly airborne disease.
+Therefore, public health authorities, including CDC and WHO, recommend FFP2 respirators for health care workers to prevent transmission and surgical face masks for patients with TB as source control to reduce the spread of infectious aerosols.
+Although surgical masks and community masks may also have a small effect in preventing transmission on inhalation, this effect is likely to be negligible and, to our knowledge, no studies have defined this modelling parameter.
+Similar modelling studies of the effect of infection and control measures on Mycobacterium tuberculosis transmission in healthcare settings (e.g., McCreesh et al., BMJ Global Health 2021, doi:10.1136/bmjgh-2021-007124) did not consider the impact of face masks on inhalation.
+We believe it is reasonable to include facemasks in the transmission model only as a factor in reducing the spread of infectious aerosols in people with tuberculosis.
+
 % more complex models and our relation
@@
 Furthermore, uncertainty in modelling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
-Third, we could not validate our modelling results with empirical findings.
+Third, \added{we assumed that mask use only had an effect on the exhalation of \emph{Mtb} particles.
+Although the same assumption was made in a related modelling study\cite{McCreesh2021BMJGlobalHealth}, we note that our modelled risk of infection would be lower if we assumed an additional effect of mask use on inhalation.} we could not validate our modelling results with empirical findings.
 Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be diagnosed with TB.

### 2024-11-24 simulations/simulate_risk.R
continue existing simulations with corrupt trans-risk file

@@ -308,7 +308,7 @@ if (!dir.exists(sim_path_to_file)) {
 simulations <- 1:nsim
 if (file.exists(paste0(sim_path_to_file, "trans-risk.txt"))) {
   existing_sim <- unique(as.integer(
-    read.table(paste0(sim_path_to_file, "trans-risk.txt"))[, 1]
+    read.table(paste0(sim_path_to_file, "trans-risk.txt"), fill = TRUE)[, 1]
   ))
   existing_sim <- existing_sim[!is.na(existing_sim)]
   simulations <- simulations[!(simulations %in% existing_sim)]

### 2024-11-24 doc/paper/supplements.tex

@@
 
+\section{Environmental and tracking data by study day}\label{sec:add-descriptives}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/co2-levels-over-time.png}
+    \caption{CO$_2$ levels by clinic area over time for each study day. Raw data at 1~min intervals averaged over a ten minute time window to reduce noise.}
+    \label{fig:co2-levels}
+\end{figure}
+
+\begin{figure}[!htpb]
+    \centering
+    \includegraphics{results/data/no-people-over-time-by-day.png}
+    \caption{Number of people by clinic area over time for each study day. Raw data at 1~second intervals  averaged over a ten minute time window to reduce noise.}
+    \label{fig:people-over-time-by-day}
+\end{figure}
+
+\clearpage
+
 \section{Processing of person-tracking data}\label{sec:setting-and-data}
@@
     \item $k$: gravitational settling rate (quanta s$^{-1}$)
-    \item $p$: breathing rate (m$^3$\,h$^{-1}$)
+    \item $p$: breathing rate (m$^3$\,s$^{-1}$)
 \end{itemize}
@@
 Infectious quanta may be removed from the air through outdoor air exchange, bacterial inactivation, or gravitational settling.
-We will ignore gravitational settling because \emph{Mtb} is carried in small particles between 1\,$\mu$m and 7\,$\mu$m\cite{Fennelly2020Lancet}, which become airborne immediately or within seconds\cite{Vuorinen2020SafSci}.
-The removal rate is thus the sum of the outdoor air exchange rate $AER_t$ and the bacterial inactivation rate $\lambda$ for \emph{Mtb}, \ie
+The removal rate is thus the sum of the outdoor air exchange rate $AER_t$, the bacterial inactivation rate $\lambda$ for \emph{Mtb}, and the gravitational settling rate $k$ for \emph{Mtb}-carrying particles, \ie
 \begin{align}\label{eq:removal}
@@
 \begin{align}\label{eq:spattemp-N}
-    \underbrace{N_{t}}_{\text{new concn.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
-concn.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda) \cdot 1\text{s}\right)}_{\text{removal}} ~.
+    \underbrace{N_{c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{c,t-1}}_{\text{old conc.}} + \underbrace{I_{c,t} \cdot q}_{\text{generation}} \cdot V_{c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{t} + \lambda + k\right)}_{\text{removal}},
 \end{align}
+where $V_c$ is the volume of cell $c$, assuming that the quanta concentration is vertically well-mixed.
 
@@
 We divide the clinic into three spaces: the waiting room (length = 10.55\,m $\times$ breadth = 5.50\,m $\times$ height = 3.00\,m), the corridor (length = 12.45\,m $\times$ breadth = 2.20\,m $\times$ height = 2.50\,m), and the TB room (length = 4.75\,m $\times$ breadth = 3.50\,m $\times$ height = 3.00\,m).
-We rasterise each room into a grid of cubic cells with a squared area of (0.25\,m)$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+Each room was discretised into a grid of cubic cells, with each cell covering a squared area of (0.25\,m)$^2$, corresponding to a diagonal cell length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+The volume of the cell depends on the room height, and we assume that the quanta concentration was vertically well-mixed.
 
-Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate spaces to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
+Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate areas to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
 Nevertheless, we acknowledge the possibility for aerosolised \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
@@
 We update the quanta concentration every second from $t = 1, \dots, T$, in line with the frequency of the person-tracking data.
-We set the quanta concentration at $t=0$ to zero, \ie $N_{c,0} = 0 ~~ \forall c$, assuming that all quanta from the previous day has been removed from the air before the start of the following clinical day.
+We set the quanta concentration at $t=0$ to zero, \ie $N_{r,c,0} = 0 ~~ \forall r, c$.
+That is, we assume that all quanta from the previous day has been removed from the air before the start of the following clinic day.
 
@@
 \begin{itemize}
-    \item[1.] \textbf{Sample uncertain modelling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, the mask reduction rate, and the bacterial inactivation rate.
-    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room per unit space (cell) over time.
+    \item[1.] \textbf{Sample uncertain modelling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, mask reduction rate, bacterial inactivation rate, and gravitational settling rate.
+    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room and cell over time.
     \item[3.] \textbf{Estimate risk of infection}: Based on their spatiotemporal quanta exposure, we estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
@@
 
-Many clinic attendees are potential undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, and bacterial inactivation rate is large.
+Many clinic attendees are potential undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, bacterial inactivation, and gravitational settling rate is large.
 To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean across these simulations.
@@
 
+\subsubsection{Gravitational settling rate}
+
+Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}.
+Accordingly, we model the settling velocity using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$, which is slightly skewed towards slower settling velocities, reflecting a higher proportion of \emph{Mtb} at the lower end of the size range\cite{Fennelly2020Lancet}.
+Assuming a drop height of 1.7\,m, the distribution for the gravitational settling rate has a median of 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).
+
 \subsubsection{Number of undiagnosed TB patients}
@@
 
-\section{Additional results}\label{sec:additional-results}
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/data/co2-levels-over-time.png}
-    \caption{CO$_2$ levels by clinic area over time for each study day. Raw data at 1~min intervals averaged over a ten minute time window to reduce noise.}
-    \label{fig:co2-levels}
-\end{figure}
-
-\begin{figure}[!htpb]
-    \centering
-    \includegraphics{results/data/no-people-over-time-by-day.png}
-    \caption{Number of people by clinic area over time for each study day. Raw data at 1~second intervals  averaged over a ten minute time window to reduce noise.}
-    \label{fig:people-over-time-by-day}
-\end{figure}
+\section{Additional simulation results}\label{sec:additional-results}
 

### 2024-11-25 analysis/simulation-results.Rmd
updated simulation results

@@
 
-```{r}
+```{r quanta-concentration}
 # main quanta simulation files
@@
   max()
-round((1 - exp(-wr_morn * 3600 * ir_sit / wr_vol)) * 100, 3)
-round((1 - exp(-wr_noon * 3600 * ir_sit / wr_vol)) * 100, 3)
+round((1 - exp(-wr_morn * 3600 * ir_sit / wr_vol)) * 100, 2)
+round((1 - exp(-wr_noon * 3600 * ir_sit / wr_vol)) * 100, 2)
 
@@
 
-```{r}
+```{r main-transmission-risk}
 # main simulation files
@@
 
-```{r}
+```{r association-risk-factors}
 # process tracking data
@@
 
-```{r additional-analyses}
+```{r ipc-comparison}
 # IPC scenario files
@@
 
-```{r} 
+```{r sensitivity-prevalence} 
 # prevalence scenario files

### 2024-11-25 doc/paper/manuscript-revised.tex

@@
 Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room.
-The average risk of infection per person was found to be 0.1\%.
+The average risk of infection per person was found to be \deleted{0.1\%}\added{0.05\%}.
 The risk increased with more close contact with infectious patients and longer time spent in the clinic.
 Without masks, the risk was 1.3 times higher.
-With lower ventilation rates, similar to pre-pandemic levels, the risk was 2.2 times higher.
+With lower ventilation rates, similar to pre-pandemic levels, the risk was \deleted{2.2}\added{2.1} times higher.
 This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.
@@
 We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
-We modelled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
+We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
 Over five days, video sensors tracked 1,438 clinic attendees.
 CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was 0.1\% (80\%-credible interval [CrI] 0.01\%$-$0.1\%).
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) higher with pre-pandemic ventilation rates.
-Spatiotemporal modelling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
+The estimated risk of infection per clinic attendee was \deleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\added{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}.
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and \deleted{2.2-fold (95\%-CrI 0.9$-$5.2)}\added{2.1-fold (95\%-CrI 0.9$-$5.0)} higher with pre-pandemic ventilation rates.
+Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
 
@@
 Healthcare facilities are high-risk settings \deleted{because infectious individuals are more likely to visit clinics}\added{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}.
-A modelling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
+A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
 
@@
 They exit the clinic through one of the exits in the waiting room or at the end of the corridor.
-Coloured icons show the placement of the video sensors to track person movements and the devices monitoring CO$_2$ levels.}
+Colored icons show the placement of the video sensors to track person movements and the devices monitoring CO$_2$ levels.}
     \label{fig:floor-plan}
@@
 Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
-On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days \added{(see time-varying CO$_2$ levels and room occupancy by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
+On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days \added{(see time-varying CO$_2$ levels and number of people per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
 
@@
 The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
-People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognised as a person.
+People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person.
 To address this, we created an R Shiny application (Fig.~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp.
@@
 
-\subsection*{Spatiotemporal modelling approach}
+\subsection*{Spatiotemporal modeling approach}
 
@@
 where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
-The probability of infection is estimated with a Poisson relation, considering the stochastic behaviour of airborne infection, where one quantum corresponds to a 67\% risk of infection.
+The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
 The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.
@@
 A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig.~\zref{fig:toy-example} in \supp).
-In the following, we describe the six main components of our modelling approach (\Cref{fig:modeling-flow}).
+In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).
 
@@
     \includegraphics{doc/model-flow-chart.pdf}
-    \caption{\textbf{Visual summary of the spatiotemporal modelling approach}.
+    \caption{\textbf{Visual summary of the spatiotemporal modeling approach}.
 (1)~Clinical records and person movements were linked based on their timestamps at the registration to determine the spatiotemporal location of diagnosed (infectious) TB patients.
@@
 The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we modelled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
+Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modelled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
 Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-We modelled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\deleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\deleted{quanta }h$^{-1}$\,$-$\,7.1\,\deleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-\deleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \deleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \added{which we modelled using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$.
-Assuming a drop height of 1.7\,m, this results in a gravitational settling rate of 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
+We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\deleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\deleted{quanta }h$^{-1}$\,$-$\,7.1\,\deleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+\deleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \deleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \added{which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$.
+Considering a drop height of 1.7\,m, the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
 
@@
 (6)~\deleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\added{The quanta concentration was computed separately for the waiting room, corridor, and TB room.
-Each room was discretised into a grid of cubic cells, with each cell covering a squared area of 0.0625\,m$^2$.
+Each room was discretized into a grid of cubic cells, with each cell covering a squared area of 0.0625\,m$^2$.
 We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as}
@@
 conc.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda\cdot 1\text{s}\right)}_{\text{removal}} ~.}
-     \added{\underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q}_{\text{generation}} \cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k\right)}_{\text{removal}} ~.}
+     \added{\underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.}
 \end{align}
-where $D$ is the diffusion constant (m$^2$\,s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), \deleted{$I_t$ is the number of infectious individuals in space, } \added{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \deleted{$AER$ is the}\added{$AER_{r,t}$ is the room-specific} air change rate (\deleted{quanta}s$^{-1}$), $\lambda$ is bacterial inactivation rate (\deleted{quanta}s$^{-1}$), \added{and $k$ is the gravitational settling rate (s$^{-1}$)}.
+where \deleted{$D$}\added{$D_t$} is the diffusion constant (m$^2$\,s$^{-1}$) \added{varying by daytime}, $\Delta$ is the Laplace operator (second-order differential operator), \deleted{$I_t$ is the number of infectious individuals in space, } \added{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \deleted{$AER$ is the}\added{$AER_{r,t}$ is the room-specific} air change rate \added{varying by daytime}(\deleted{quanta}s$^{-1}$), $\lambda$ is the bacterial inactivation rate (\deleted{quanta}s$^{-1}$), \added{and $k$ is the gravitational settling rate (s$^{-1}$)}.
+\added{We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.}
 
@@
     % \deleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
-    \added{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\right),}
+    \added{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),}
 \end{align}
-where \deleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\added{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+where \deleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\added{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
 We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
@@
 Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
-Each room was rasterised into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
-The quanta concentration was modelled during clinic hours from 8\,am to 4\,pm and updated every second.
-We used a Monte-Carlo simulation approach to consider uncertainty in modelling parameters.
-In each simulation, we sampled uncertain modelling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.}
+Each room was rasterized into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
+The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
+We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters.
+In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.}
 
@@
 To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask-wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie assuming an air change rate of 6~air changes per hour, corresponding to the average rate in the pilot study in 2019 in the same clinic\cite{Zurcher2022JID}.
-Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolised \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
-Except for the scenario-specific modifications, all other modelling assumptions and parameters were identical in each Monte Carlo simulation.
+Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
+Except for the scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.
 
@@
 
-We summarised clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
-\added{We simulated the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modelling parameters (Section~\zref{sec:estimation} in \supp).
-In each simulation, we computed the risk of infection based on the cumulative quanta exposure during each susceptible individual's clinic visit.
+We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
+\added{We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp).
+In each simulation, we computed the individual risk of infection based on the cumulative quanta exposure during the clinic visit.
 We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations.
-The scenario-specific infection risk distribution was shown with a histogram and summarised with the mean and 80\%-credible interval (CrI).} \deleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarised with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorised as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
-We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modelled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
+The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).} \deleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
+The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
+We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modeled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
 Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1).
@@
 The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) \deleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
-\added{CO$_2$ levels slightly varied within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp) and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
+\added{CO$_2$ levels slightly varied within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp), and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
 The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
-An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a 0.40\% risk of infection in the morning compared to a 0.07\% risk in the afternoon.
-The mean risk of infection per clinic visit was 0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%), corresponding to a risk of 0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%).
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a \deleted{0.40\%}\added{0.20\%} risk of infection in the morning compared to a \deleted{0.07\%}\added{0.04\%} risk in the afternoon.
+The mean risk of infection per clinic visit was \deleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\added{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}, corresponding to a risk of \deleted{0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%)}\added{0.1\% per hour (80\%-CrI 0.01\%$-$0.21\%)}.
 S2~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
-It can be seen that the quanta concentration was higher near the infectious patient, indicating his position and movement.
+It can be seen that the quanta concentration was higher near the infectious patient, following their position and movement.
 The quanta diffused quickly because the diffusion constant was related to the air change rate, which was generally high because all doors and windows were typically open during clinic hours.
 
-The modelled individual risk of infection was further associated with the number of close contacts and the time spent in the clinic.
-A doubling of close-contact encounters increased the odds of infection by 23\% (OR 1.23, 95\%-CrI 1.09$-$1.37).
-Doubling the time spent in the clinic increased the odds of infection by 12\% (OR 1.12, 95\%-CrI 1.03$-$1.21).
-The time spent in close contact was not significantly associated with the individual risk of infection (OR 1.03, 95\%-CrI, 0.94$-$1.13).
+The modeled individual risk of infection was associated with the number of close contacts and the time spent in the clinic.
+A doubling of close-contact encounters increased the odds of infection by \deleted{23\% (OR 1.23, 95\%-CrI 1.09$-$1.37)}\added{17\% (OR 1.17, 95\%-CrI 1.04$-$1.30)}.
+Doubling the time spent in the clinic increased the odds of infection by \deleted{12\% (OR 1.12, 95\%-CrI 1.03$-$1.21)}\added{12\% (OR 1.12, 95\%-CrI 1.03$-$1.22)}.
+The time spent in close contact was not significantly associated with the individual risk of infection \deleted{(OR 1.03, 95\%-CrI, 0.94$-$1.13)}\added{(OR 1.02, 95\%-CrI, 0.94$-$1.11)}.
 
@@
 The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.12\% (80\%-CrI 0.02\%$-$0.13\%), an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.19\% (80\%-CrI 0.02\%$-$0.20\%), an increase of 120\% (OR 2.2, 95\%-CrI 0.9$-$5.2).
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.24\% (80\%-CrI 0.03\%$-$0.24\%), an increase of 180\% (OR 2.8, 95\%-CrI 1.1$-$6.6).
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been \deleted{0.12\% (80\%-CrI 0.02\%$-$0.13\%)}\added{0.07\% (80\%-CrI 0.01\%$-$0.08\%)}, an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been \deleted{0.19\% (80\%-CrI 0.02\%$-$0.20\%)}\added{0.09\% (80\%-CrI 0.01\%$-$0.10\%)}, an increase of \deleted{120\% (OR 2.2, 95\%-CrI 0.9$-$5.2)}\added{110\% (OR 2.1, 95\%-CrI 0.9$-$5.0)}.
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been \deleted{0.24\% (80\%-CrI 0.03\%$-$0.24\%)}\added{0.12\% (80\%-CrI 0.02\%$-$0.12\%)}, an increase of \deleted{180\% (OR 2.8, 95\%-CrI 1.1$-$6.6)}\added{170\% (OR 2.7, 95\%-CrI 1.1$-$6.6)}.
 The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
@@
 In sensitivity analyses, we varied assumptions about the number of infectious persons in the clinic.
-Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher (0.12\%, 80\%-CrI 0.03\%$-$0.19\%).
-Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high (0.22\%, 80\%-CrI <0.01\%$-$0.43\%).
+Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher \deleted{(0.12\%, 80\%-CrI 0.03\%$-$0.19\%)}\added{(0.06\%, 80\%-CrI 0.01\%$-$0.10\%)}.
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high \deleted{(0.22\%, 80\%-CrI <0.01\%$-$0.43\%)}\added{(0.11\%, 80\%-CrI <0.01\%$-$0.18\%)}.
 
@@
 We applied the model to real-world data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
-We estimated an average risk of infection per clinic attendee of 0.1\%.
+We estimated an average risk of infection per clinic attendee of \deleted{0.1\%}\added{0.05\%}.
 We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
@@
 % close contacts
-The COVID-19 pandemic has renewed interest in airborne transmission of respiratory infections in crowded indoor settings\cite{Wang2021Science,Morawska2021,Greenhalgh2022BMJ} and its modelling\cite{Shen2021STBE,Buonanno2020EI,Edwards2024RiskAnalysis}.
+The COVID-19 pandemic has renewed interest in airborne transmission of respiratory infections in crowded indoor settings\cite{Wang2021Science,Morawska2021,Greenhalgh2022BMJ} and its modeling\cite{Shen2021STBE,Buonanno2020EI,Edwards2024RiskAnalysis}.
 Many studies\cite{Andrews2014JID,Taylor2016IJTLD,Hella2017JInfect,Zemouri2020JDR} rely on the Wells-Riley transmission model\cite{Riley1961Book}, based on the pioneering work of William Wells, who proposed a role of droplets in airborne transmission in the 1930s\cite{Wells1934AJE}, and Richard Riley, who developed the equation in the 1960s\cite{Riley1961Book}.
@@
 Our spatiotemporal model is an extension of the Wells-Riley model, which allows the quanta concentration to vary both over time and space, and considers that the density of infectious particles is initially higher near the infectious source\cite{Wang2021Science,Vuorinen2020SafSci,Chen2020BuildEnv}.
-We found a significant association between the modelled risk of infection and the number of close-contact encounters, and the time spent in the clinic.
+We found a significant association between the modeled risk of infection and the number of close-contact encounters, and the time spent in the clinic.
 More close contacts and extended visits make exposure to high doses of infectious particles more likely.
@@
 Nevertheless, we estimated a lower risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
-Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 2.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
+Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about \deleted{2.4\%}\added{1.2\%}, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
 The difference can be mainly attributed to the infection control measures that were in place during the COVID-19 pandemic.
@@
 Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in the model.
-Other modelling assumptions may also have contributed.
+Other modeling assumptions may also have contributed.
 The risk estimate from our previous study before the COVID-19 pandemic\cite{Zurcher2022JID} was based on TB prevalence in the South African population, rather than the TB status documented in the clinical data.
-Modelling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modelled risk of transmission.
+modeling the risk based on TB prevalence or assuming that both diagnosed and suspected TB patients were infectious also resulted in a higher modeled risk of transmission.
 
@@
 The air change rates in the clinic typically ranged from 10 to 20 air changes per hour, which resulted in a quick diffusion and removal of infectious quanta from the indoor air.
-To put this into perspective, we recently modelled the risk of \emph{Mtb} transmission in schools by comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
+To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools by comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
 Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely because of better natural ventilation, with an air change rate comparable to the one measured in the primary care clinic in our study.
-We emphasise that our results do \emph{not} suggest that masks are ineffective -- there is a large body of evidence for their effectiveness -- but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
+We emphasize that our results do \emph{not} suggest that masks are ineffective -- there is a large body of evidence for their effectiveness -- but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 
-There is broad consensus that tuberculosis is a strictly airborne disease.
+% \added{There is broad consensus that tuberculosis is a strictly airborne disease.
 Therefore, public health authorities, including CDC and WHO, recommend FFP2 respirators for health care workers to prevent transmission and surgical face masks for patients with TB as source control to reduce the spread of infectious aerosols.
-Although surgical masks and community masks may also have a small effect in preventing transmission on inhalation, this effect is likely to be negligible and, to our knowledge, no studies have defined this modelling parameter.
-Similar modelling studies of the effect of infection and control measures on Mycobacterium tuberculosis transmission in healthcare settings (e.g., McCreesh et al., BMJ Global Health 2021, doi:10.1136/bmjgh-2021-007124) did not consider the impact of face masks on inhalation.
-We believe it is reasonable to include facemasks in the transmission model only as a factor in reducing the spread of infectious aerosols in people with tuberculosis.
+Although surgical masks and community masks may also have a small effect in preventing transmission on inhalation, this effect is likely to be negligible and, to our knowledge, no studies have defined this modeling parameter.
+Similar modeling studies of the effect of infection and control measures on Mycobacterium tuberculosis transmission in healthcare settings (e.g., McCreesh et al., BMJ Global Health 2021, doi:10.1136/bmjgh-2021-007124) did not consider the impact of face masks on inhalation.
+We believe it is reasonable to include facemasks in the transmission model only as a factor in reducing the spread of infectious aerosols in people with tuberculosis.}
 
@@
 Our model considered ventilation conditions and patient movements.
-However, other factors such as variation in infectiousness, temperature and humidity, or physiochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (see Section~\zref{sec:depth-discussion} in \supp for a detailed discussion of these factors).
+However, other factors such as variation in infectiousness, temperature and humidity, or physicochemical properties of the infectious particles can further influence the generation, diffusion, and removal of infectious quanta (for a detailed discussion of these factors, see Section~\zref{sec:depth-discussion} in \supp).
 Furthermore, computational fluid dynamics (CFD) models have been developed to examine airflow as a factor influencing the spatial spread of airborne pathogens\cite{Vuorinen2020SafSci,Jung2021InfectChemo,Li2021BuildEnv,Yan2023BE,Qian2009BE,Li2022SOTTE}.
 However, CFD models are computationally expensive and primarily used in experimental and static environments, making their application to real-world settings such as clinics difficult.
-Moreover, our modelling approach required computational efficiency because many Monte Carlo simulations are needed to reflect uncertainty in modelling parameters.
+Moreover, our modeling approach required computational efficiency because many Monte Carlo simulations are needed to reflect uncertainty in modeling parameters.
 
@@
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-\added{For example, the waiting room and corridor were separated to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels, yet the door between these areas was typically open, so that aerosolised \emph{Mtb} (quanta) could spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
+\added{For example, the waiting room and corridor were separated to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels, yet the door between these areas was typically open, so that aerosolized \emph{Mtb} (quanta) could have spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
 However, airflow is very difficult to measure.
-Furthermore, uncertainty in modelling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
+Furthermore, uncertainty in modeling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
 Third, \added{we assumed that mask use only had an effect on the exhalation of \emph{Mtb} particles.
-Although the same assumption was made in a related modelling study\cite{McCreesh2021BMJGlobalHealth}, we note that our modelled risk of infection would be lower if we assumed an additional effect of mask use on inhalation.} we could not validate our modelling results with empirical findings.
+Although the same assumption was made in a related modeling study\cite{McCreesh2021BMJGlobalHealth}, we note that our modeled risk of infection would be lower if we assumed an additional effect of mask use on inhalation.
+Fourth,} we could not validate our modeling results with empirical findings.
 Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be diagnosed with TB.
@@
 All other data are included in the manuscript and accompanying supporting information.
-Code used for preprocessing, modelling and analysis, descriptive and simulation results are available on GitHub via \url{https://github.com/nbanho/clinic-transmission}.
+Code used for preprocessing, modeling and analysis, descriptive and simulation results are available on GitHub via \url{https://github.com/nbanho/clinic-transmission}.
 
@@
 Data collection: KM, JL, RS, KZ, LF.
-Modelling and statistical analysis: NB.
+Modeling and statistical analysis: NB.
 Paper draft: NB, LF, ME.

### 2024-11-25 doc/paper/supplements.tex

@@
 
-\listoftables
-
 %\listoffigures
@@
 Person-tracking data required processing because many tracks were interrupted, for example, if the person i)~moved outside the range of the video sensors and then re-entered, ii)~bowed heavily downwards, or iii)~was temporarily hidden behind another person or obstacle.
-In addition, moving objects such as opening doors may have been recognised as a person, introducing many extremely short tracks (total duration of less than 10\,s and distance less than 2\,m), which were removed.
+In addition, moving objects such as opening doors may have been recognized as a person, introducing many extremely short tracks (total duration of less than 10\,s and distance less than 2\,m), which were removed.
 
@@
 The Shiny application is shown in \Cref{fig:shiny-app}.
-In the centre panel, a floor plan of the clinic is shown in (x,y)-space.
-Different areas of the clinic are delineated by colours, \eg seating areas are marked in purple and exit areas are marked in blue.
+In the center panel, a floor plan of the clinic is shown in (x,y)-space.
+Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue.
 Tracks are shown as lines with a open and crossed circles indicating the tracks' start and end, respectively.
-The currently selected track is shown in black, possible links in colours, and alternatives for the selected possible link in light grey.
+The currently selected track is shown in black, possible links in colors, and alternatives for the selected possible link in light gray.
 
-Below the floor plan, a table is showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID of the selected track (Pid) and the observation ID of possible links (Oid), (2)~the height difference at the end point of Pid and the start point of Oid, (3)~ the standing height (\ie maximum height during the entire track) difference, (4)~the duration of Oid, (5)~the time difference, and (6)~the distance between the end point of Pid and the start of point of Oid.
+Below the floor plan, a table is showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID of the selected track (Pid) and the observation ID of possible links (Oid), (2)~the height difference at the end point of Pid and the start point of Oid, (3)~ the standing height (\ie maximum height during the entire track) difference, (4)~the duration of Oid, (5)~the time difference, and (6)~the distance between the end point of Pid and the start point of Oid.
 Possible links are sorted in ascending order by the time difference.
@@
 After that, the app updates the floor plan, matching table, and descriptive information.
-Erroneous link can be revoked by clicking "Unlink last ID", if the last link in time should be revoked, or by clicking "Unlink first ID", if the first link in time should be revoked.
+Erroneous links can be revoked by clicking "Unlink last ID", if the last link in time should be revoked, or by clicking "Unlink first ID", if the first link in time should be revoked.
 If the Pid could not be connected further, its track can be finished by entering a label and clicking "Label/End track".
@@
     \item \textbf{Clean}: A clean track throughout the clinic with known entrance and exit.
-    \item \textbf{Staff}: IDs with moving in- and out of the staff room.
+    \item \textbf{Staff}: A track moving in- and out of the staff room.
     \item \textbf{Noise}: A short track ($\leq$1\,min) without entrance or exit.
@@
 We obtained 3,319 tracks after preprocessing.
-Further exclusion of short tracks <5\,min resulted in 1,563 tracks used for analysis: 209 (13\%) labeled as "Clean", 125 (8\%) as "Staff", 213 (14\%) as "Lost enter", 246 (16\%) as "Lost exit", and 770 (49\%) as "Lost both".
+Further exclusion of relatively short tracks <5\,min resulted in 1,563 tracks used for analysis: 209 (13\%) labeled as "Clean", 125 (8\%) as "Staff", 213 (14\%) as "Lost enter", 246 (16\%) as "Lost exit", and 770 (49\%) as "Lost both".
 
@@
 
-\section{Spatiotemporal modelling approach}\label{sec:spattemp-model}
+\section{Spatiotemporal modeling approach}\label{sec:spattemp-model}
 
-\subsection{The Wells-Riley modelling framework}
+\subsection{The Wells-Riley modeling framework}
 
 We build upon the Wells-Riley model\cite{Riley1961Book}, which is an established epidemiological model to estimate the risk of infection.
-The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation:
+The model estimates the risk of airborne indoor transmission as
 \begin{align}
@@
 \end{align}
-where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$\,h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta\,h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$\,h$^{-1}$).
+where $P$ is the probability of infection, $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$\,h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta\,h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$\,h$^{-1}$).
 
@@
 It rather represents the dose of infectious particles that corresponds to a certain probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection\cite{Rudnick2003IndoorAir}.
-It is an attempt to consider the stochastic behaviour of airborne transmission, since the number of infectious particles that cause an infection is unknown.
+It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown.
 Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual.
@@
 
-While considering the stochastic behaviour of airborne transmission, the Wells-Riley model makes two simplifying assumptions.
+While considering the stochastic behavior of airborne transmission, the Wells-Riley model makes two simplifying assumptions.
 First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and uniformly in the indoor airspace.
@@
 The extension by Rudnick and Milton still assumes a well-mixed airspace.
-We relax this assumption by modelling spatial variation in the concentration of infectious quanta.
+We relax this assumption by modeling spatial variation in the concentration of infectious quanta.
 We then compute the risk of infection for susceptible individuals using the Wells-Riley equation considering their cumulative exposure to infectious quanta over space and time as
 \begin{equation}\label{eq:spattemp-P}
-    P = \sum_s \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
+    P = \sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
 \end{equation}
-where $N$ is the quanta concentration in the indoor space area $c$ at time $t$ (in quanta\,$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in $c$ at $t$, and $p$ is the breathing rate (in m$^3$\,h$^{-1}$).
-
-Modelling of the spatiotemporal quanta concentration can be divided into three processes
+where $N$ is the quanta concentration in a specific area of the indoor space $c$ at time $t$ (quanta\,m$^{-3}$), $\mathbb{I}$ indicates whether the individual was in $c$ at $t$, and $p$ is the breathing rate (m$^3$\,h$^{-1}$).
+Modeling of the spatiotemporal quanta concentration can subsequently be divided into three processes
 \begin{itemize}
-    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Infectious particles are generated at the infectious individual's location. 
-    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Pathogen-carrying particles spread in the indoor space. 
-    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Contaminated indoor air is replaced or pathogens become inactive. 
+    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Infectious particles generated at the infectious individual's location. 
+    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Pathogen-carrying particles disperse in the indoor space. 
+    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Contaminated indoor air is replaced with fresh outdoor air. 
 \end{itemize}
-We describe each process in more detail in the following, but first we introduce some notation.
+We describe each process in greater detail in the following, but first we introduce some notation.
 
@@
 
-We divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the centre of cell $c$.
+We assume that quanta concentration is well-mixed vertically and divide the airspace into a 2-dimensional grid with cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$.
 The grid provides a discrete approximation to the continuously varying quanta concentration in the indoor space, with the number of grid cells $C = n_x \cdot n_y$ determining the spatial resolution.
-We denote the following variables and modelling parameters:
+We denote the following variables and modeling parameters:
 \begin{itemize}
-    \item $N_{c,t}$: quanta concentration in airspace unit $c = 1, \dots, C$ at time $t = 1, \dots T$ (in seconds)
-    \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
+    \item $N_{c,t}$: quanta concentration in $c = 1, \dots, C$ at $t = 1, \dots T$ (in seconds)
+    \item $I_{c,t}$: number of infectious individuals in $c$ at $t$
     \item $q$: quanta generation rate (quanta s$^{-1}$)
     \item $D$: diffusion constant (m$^2$ s$^{-1}$)
-    \item $AER$: air change rate (air changes s$^{-1}$)
-    \item $V$: volume of the airspace (m$^3$)
+    \item $AER$: air change rate (s$^{-1}$)
+    \item $V_c$: volume of the cell (m$^3$)
     \item $CO^{\text{I}}$: CO$_2$ mixing ratio in indoor air (parts per million, ppm)
@@
     \item $n_t$: number of individuals in the airspace at time $t$
-    \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
-    \item $k$: gravitational settling rate (quanta s$^{-1}$)
+    \item $\lambda$: bacterial inactivation rate (s$^{-1}$)
+    \item $k$: gravitational settling rate (s$^{-1}$)
     \item $p$: breathing rate (m$^3$\,s$^{-1}$)
@@
 \end{align}
-We assume that the initially generated quanta is confined to the infectious individual's cell, reflecting previous findings that the concentration of pathogen-laden aerosols is typically highest near the infectious source\cite{Wang2021Science,Morawska2021,Vuorinen2020SafSci,Chen2020BuildEnv}.
+For now, we assume that the initially generated quanta is confined to the infectious individual's cell, reflecting previous findings that the concentration of pathogen-laden aerosols is typically highest near the infectious source\cite{Wang2021Science,Morawska2021,Vuorinen2020SafSci,Chen2020BuildEnv}.
 Although coughing and sneezing can expel the aerosols further away from the individual, these activities are much less frequent than breathing, which may thus contribute more infectious particles\cite{Dinkele2022AJRCCM}.
+Note, however, that the model can also accommodate assumptions about a larger initial spread of quanta.
 
@@
 The diffusion constant $D$ is unknown.
-We approximate it using an empirical relationship between the eddy diffusion coefficient $K$ ($K \approx D$ in m$^2$\,s$^{-1}$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
+We approximate it using an empirical relationship between the eddy diffusion coefficient $K \approx D$ (m$^2$\,s$^{-1}$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
 \begin{align}
@@
 \end{align}
-where $V$ is the volume of the airspace (in m$^3$).
-We thus assume that pathogen-carrying particles spread with a similar speed than CO$_2$ particles, which may overestimate the diffusion constant because CO$_2$ particles are smaller than \emph{Mtb}-carrying particles.
+where $V$ is the volume of the entire space (m$^3$).
+Implicitly, we assume that pathogen-carrying particles spread with a similar speed than CO$_2$ particles, which may overestimate the diffusion constant because CO$_2$ particles are smaller than \emph{Mtb}-carrying particles.
 
-The $AER$ is not directly observed but can be estimated from indoor CO$_2$ levels, room occupancy, and volume.
+The $AER$ is not directly observed but can be estimated from indoor CO$_2$ levels and room occupancy.
 Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state mixing ratio), the $AER$ can be computed as
@@
 However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ mixing ratio.
-Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
+Therefore, we determine $AER$ from time-varying CO$_2$ levels and room occupancy using a transient mass balance model of the form\cite{Batterman2017IJERPH}
 \begin{align}
@@
 \end{align}
-where $Q$ is the outdoor air supply rate (in m$^3$\,s$^{-1}$ $\cdot \frac{1}{3600}$ ) and $AER = Q/V$.
+where $Q$ is the outdoor air supply rate (m$^3$\,s$^{-1}$ $\cdot \frac{1}{3600}$ ) and $AER = Q/V$.
 We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS), a quasi-Newton method, as implemented in the statistical software R\cite{Byrd1995SIAM}.
-The optimal solution is the outdoor air supply rate $Q$ that minimises the root mean square error.
+The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error.
 We also follow recommendations to fit the outdoor CO$_2$ mixing ratio with reasonable constraints on $CO^{\text{O}} = [300,600]$\,ppm\cite{Batterman2017IJERPH}.
-We can compute $AER$ over different time windows, hence the index in $AER_t$.
 
@@
 Infectious quanta may be removed from the air through outdoor air exchange, bacterial inactivation, or gravitational settling.
-The removal rate is thus the sum of the outdoor air exchange rate $AER_t$, the bacterial inactivation rate $\lambda$ for \emph{Mtb}, and the gravitational settling rate $k$ for \emph{Mtb}-carrying particles, \ie
+The removal rate is thus the sum of the outdoor air exchange rate $AER$, the bacterial inactivation rate $\lambda$ for \emph{Mtb}, and the gravitational settling rate $k$ for \emph{Mtb}-carrying particles, \ie
 \begin{align}\label{eq:removal}
-    AER_t + \lambda ~.
+    AER + \lambda + k~.
 \end{align}
@@
 
-Combining \Cref{eq:generation}, \Cref{eq:diffusion}, and \Cref{eq:removal}, the quanta concentration at time $t$ in the indoor space can be computed as
+Combining Equations \ref{eq:generation}, \ref{eq:diffusion}, and \ref{eq:removal}, the quanta concentration can be computed as
 \begin{align}\label{eq:spattemp-N}
@@
 \end{align}
-where $V_c$ is the volume of cell $c$, assuming that the quanta concentration is vertically well-mixed.
+where $V_c$ is the volume of cell $c$.
 
@@
 
-We illustrate our modelling approach with an example.
-Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterise into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of (0.25\,m)$^2$.
-We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells.
-We monitor the room from 8\,am to 12\,am and consider one infectious individual generating quanta continuously for one hour between 8\,am and 9\,am in the middle of the room (cell (20,10)), which means that quanta is generated in cells $(x_c = 20\pm1, y_c = 10\pm1)$.
-We assume a quanta generation rate of $q = 2$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0$\,quanta h$^{-1}$.
-We further assume a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$, translating into a quanta removal rate of 1\,quanta h$^{-1}$ and corresponding to a diffusion constant of 0.015\,m$^2$ s$^{-1}$.
+We illustrate our modeling approach with an example.
+Consider a room with a volume $V$=150\,m$^3$ (length = 10\,m, breadth = 5\,m, height = 3\,m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$\,cells with an area of (0.25\,m)$^2$.
+We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighboring cells.
+We monitor the room from 8\,am to 12\,am and consider one infectious individual generating quanta continuously for one hour between 8\,am and 9\,am in the middle of the room in cell (20,10), which means that quanta is generated in cells $(x_c = 20\pm1, y_c = 10\pm1)$.
+We assume a quanta generation rate of $q = 2$\,quanta h$^{-1}$ and a bacterial inactivation rate of $\lambda = 0$\,h$^{-1}$.
+We further assume a constant outdoor air exchange rate of $AER = 1$\,h$^{-1}$, corresponding to a diffusion constant of 0.015\,m$^2$ s$^{-1}$.
 
 \Cref{fig:toy-example} shows the spatiotemporal quanta concentration.
-The quanta concentration increases over time and is higher near the infectious individual in the room's centre (\Cref{fig:toy-example}a, top panel).
-After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am (\Cref{fig:toy-example}a, bottom panel) through outdoor air exchange.
-Spatial variation in the quanta concentration peaks at 9\,am, right before the infectious individual leaves the space, and then it takes about 15\,min until a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).
+The quanta concentration increases over time and is higher near the infectious individual in the room's center (\Cref{fig:toy-example}a, top panel).
+After the individual has left the room, the quanta diffuses within minutes and most quanta is removed by 10\,am through outdoor air exchange (\Cref{fig:toy-example}a, bottom panel).
+Spatial variation in the quanta concentration peaks at 9\,am, right before the individual leaves the space, and 15\,min after that a well-mixed airspace has been reached through quanta diffusion (\Cref{fig:toy-example}b).
 
@@
     \centering
-    \includegraphics{tests/stm_v2-toy_example.png}
+    \includegraphics{tests/stm_v2-toy_example-A.png}
     \includegraphics{tests/stm_v2-toy_example-B.png}
-    \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the room's centre}.
-\textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time for the 800~cells into which the indoor space was divided.}
+    \caption[Example: Spatiotemporal quanta concentration in an indoor space]{\textbf{Example: Spatiotemporal quanta concentration in an indoor space between 8\,am and 10\,am following the generation of quanta by an infectious individual between 8\,am and 9\,am in the center of the room}.
+\textbf{(a)}~Spatial quanta concentration at selected time points and \textbf{(b)}~quanta concentration over time for the 800~cells into which the indoor space was rasterized.}
     \label{fig:toy-example}
@@
 We divide the clinic into three spaces: the waiting room (length = 10.55\,m $\times$ breadth = 5.50\,m $\times$ height = 3.00\,m), the corridor (length = 12.45\,m $\times$ breadth = 2.20\,m $\times$ height = 2.50\,m), and the TB room (length = 4.75\,m $\times$ breadth = 3.50\,m $\times$ height = 3.00\,m).
-Each room was discretised into a grid of cubic cells, with each cell covering a squared area of (0.25\,m)$^2$, corresponding to a diagonal cell length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
-The volume of the cell depends on the room height, and we assume that the quanta concentration was vertically well-mixed.
+Each room is discretized into a grid of cubic cells, with each cell covering a squared area of (0.25\,m)$^2$, corresponding to a diagonal cell length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m.
+The volume of the cell depends on the room height.
 
-Although the door between the waiting room and corridor was usually not closed, we consider these rooms as separate areas to facilitate modelling and because we observed differences in room occupancy and time-varying CO$_2$ levels.
-Nevertheless, we acknowledge the possibility for aerosolised \emph{Mtb} (quanta) to spread from the waiting room to the corridor, and vice versa, which is not considered in our simulations.
+Although the door between the waiting room and corridor was usually not closed, we consider these rooms to be separate to facilitate modeling.
+We acknowledge the possibility for aerosolized \emph{Mtb} to spread from the waiting room to the corridor, and vice versa, which is not considered.
 
@@
 Correspondingly, we estimate the outdoor air exchange rate by daytime, considering potential differences in ventilation conditions between the morning and afternoon.
-We update the quanta concentration every second from $t = 1, \dots, T$, in line with the frequency of the person-tracking data.
+We update the quanta concentration every second from $t = 1, \dots, T$, corresponding to the frequency of the person-tracking data.
 We set the quanta concentration at $t=0$ to zero, \ie $N_{r,c,0} = 0 ~~ \forall r, c$.
@@
 
-We use Monte Carlo simulation to estimate the risk of infection, considering uncertainty in several modelling parameters. Each simulation consists of the following steps:
+We use Monte Carlo simulation to model the quanta concentration and estimate the individual risk of infection, considering uncertainty in several modeling parameters.
+Each simulation consists of the following steps:
 \begin{itemize}
-    \item[1.] \textbf{Sample uncertain modelling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, mask reduction rate, bacterial inactivation rate, and gravitational settling rate.
-    \item[2.] \textbf{Compute quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room and cell over time.
-    \item[3.] \textbf{Estimate risk of infection}: Based on their spatiotemporal quanta exposure, we estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
+    \item[1.] \textbf{Sample uncertain modeling parameters}: Undiagnosed TB patients among clinic attendees, quanta generation rate, mask reduction rate, bacterial inactivation rate, and gravitational settling rate.
+    \item[2.] \textbf{Model quanta concentration}: Based on the generation of infectious quanta by diagnosed and undiagnosed TB patients, we compute the quanta concentration in each room and cell over time.
+    \item[3.] \textbf{Estimate risk of infection}: Based on their cumulative quanta exposure, we estimate the risk of infection for each clinic attendee using the Wells-Riley equation.
 \end{itemize}
 
-Many clinic attendees are potential undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, bacterial inactivation, and gravitational settling rate is large.
-To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean across these simulations.
+Many clinic attendees can be considered potential undiagnosed TB patients and prior uncertainty in the quanta generation, mask reduction, bacterial inactivation, and gravitational settling rate is large.
+To reflect this uncertainty, we perform 5,000 Monte Carlo simulations and compute the individual risk of infection as the mean across simulations.
 
-\subsection{Modelling assumptions}\label{sec:priors}
+\subsection{Modeling assumptions}\label{sec:priors}
 
@@
 Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting (leakage of particles on the sides of the mask) particles\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}.
-Therefore, we assume that mask wearing confines the quanta generation only to the cell where the infectious individual is located.
-In our setup, this corresponds to a propagation distance of 0.35\,m (the diagonal cell length) with an initial spread to an area of 0.25$^2$\,m$^2$.
-Without masks, we assume two times that propagation distance, \ie 0.7\,m (the diagonal lengths of two cells)\cite{Tang2013PLoSOne}.
-Since the direction of propagation is not observed, we assume uniform spread to all neighbouring cells, \ie the generated quanta is distributed evenly among the first-neighbouring cells from where the infectious individual is located.
+Therefore, we assume that mask wearing confines the quanta generation only to the cell where the infectious individual is located, hence our choice of the cell length corresponding to a propagation distance of 0.35\,m.
+Without masks, we assume the unmitigated propagation distance of 0.7\,m (the diagonal lengths of two cells).
+Since the direction of propagation is not observed, we assume uniform spread to all neighboring cells, \ie the generated quanta is distributed evenly among the neighboring cells from where the infectious individual is present.
 
@@
 The quanta generation rate follows a lognormal prior distribution with median 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking.
-The activity is determined based on whether the person had moved by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).
+The activity is determined based on whether the person moves by more or less than 0.25\,m (spatial distance between two recorded tracks of the same person).
 
@@
 The bacterial inactivation rate for \emph{Mtb} has not been precisely estimated.
-Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolised TB bacilli of 6\,h ($\lambda$ = 0.12).
-By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5\,min ($\lambda$ = 8.3).
+Loudon (1969)\cite{Loudon1969AMRRD} report a half-life for aerosolized TB bacilli of 6\,h ($\lambda$ = 0.12\,h$^{-1}$).
+By contrast, Lever (2000)\cite{Lever2000LettersAppliedMicrobio} reported a half-life of just 5\,min ($\lambda$ = 8.3\,h$^{-1}$).
 The difference could be explained by differences in humidity levels during the study\cite{Lever2000LettersAppliedMicrobio}: Loudon (1969) and Lever (2000) studied survival in an environment with relative humidity of about 50\% and 70\%, respectively.
 Relative humidity in our study is closer to 50\%, thus the estimate by Loudon (1969) may be more suitable for our study.
-Furthermore, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5\,h ($\lambda$ = 0.46) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}, and Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30\,min ($\lambda$ = 3.2), which is closer to the estimate by Lever (2000).
-Considering the variation in these estimates, we model $\lambda$ with a $\mathrm{Lognormal}(\mu = \log 1, \sigma = 1)$ distribution, with a median rate of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$).
+Furthermore, Gannon (2007)\cite{Gannon2007ResVetSci} report a half-life of 1.5\,h ($\lambda$ = 0.46\,h$^{-1}$) at relative humidity of over 75\%, yet the estimate is for \emph{Mtb bovis}, and Klein (2014)\cite{Klein2014IJMyco} observe a 80\% reduction within 30\,min ($\lambda$ = 3.2\,h$^{-1}$), which is closer to the estimate by Lever (2000).
+Considering the variation in these estimates, we model $\lambda$ with a $\mathrm{Lognormal}(\mu = \log 1, \sigma = 1)$ distribution, with a median rate of 1\,h$^{-1}$ (95\%-CrI 0.1\,h$^{-1}$\,$-$\,7.1\,h$^{-1}$).
 The distribution is shown in \Cref{fig:lambda-distribution} along with the estimates from the literature.
@@
 Previous studies suggest that the number of undiagnosed (subclinical) TB patients is roughly similar to the number of diagnosed TB patients\cite{Berhanu2023CID,Moyo2022LancetID}.
-Therefore, we model the number of undiagnosed TB patients with a Multinomial distribution based on the counts of the daily number of diagnosed TB patients from October to November 2021 (\Cref{fig:undiagnosed-distribution}.
+Therefore, we model the number of undiagnosed TB patients with a Multinomial distribution based on the counts of the daily number of diagnosed TB patients from October to November 2021 (\Cref{fig:undiagnosed-distribution}).
 We consider all clinic attendees as potential undiagnosed TB patients and sample among them with equal probability.
@@
     \includegraphics{results/modeling/mean-roi-assumptions.png}
-    \caption[Mean risk of infection per clinic attendee when varying assumptions about the number of infectious people in the clinic]{Mean risk of infection per clinic attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
+    \caption[Mean risk of infection per clinic attendee when varying assumptions about the number of infectious people in the clinic]{Mean risk of infection per clinic attendee if \textbf{(a)}~the number of infectious people in the clinic corresponded to the prevalence of TB in the South African population, or \textbf{(b)}~the number of infectious people in the clinic corresponded to the number of bacteriologically confirmed TB patients and those suspected of having TB due to respiratory symptoms when attending the clinic.
 Low: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}
@@
 
-\section{Discussion of factors affecting the risk of airborne infection}\label{sec:depth-discussion}
+\section{Discussion of factors influencing transmission risks}\label{sec:depth-discussion}
 
@@
 The probability of infection is modeled with a Poisson relation and the Wells-Riley model considers multiple determinants for the generation and survival of infectious quanta.
-Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorised into (1)~environmental factors, (2)~physicochemical properties of the pathogen, and the (3)~infectiousness of the source and susceptibility of the host.
+Nevertheless, transmission events are difficult to predict as the occurrence depends on multiple other factors that can broadly be categorized into (1)~environmental factors, (2)~physicochemical properties of the pathogen, and the (3)~infectiousness of the source and susceptibility of the host.
 
@@
 Environmental factors are somewhat considered by our model using CO$_2$ as a tracer gas to estimate the air exchange rate, based on which we model quanta diffusion and removal.
-The combination of clinical and person-tracking data further allows the spatial modelling of quanta generation.
-This makes our approach unique compared to existing modelling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1961Book,Rudnick2003IndoorAir}.
+The combination of clinical and person-tracking data further allows the spatial modeling of quanta generation.
+This makes our approach unique compared to existing modeling based on the Wells-Riley equation assuming a well-mixed airspace\cite{Riley1961Book,Rudnick2003IndoorAir}.
 In our spatiotemporal model, we assume a radial diffusion of quanta, which may not reflect the actual airflow at the primary care clinic.
@@
 For example, measuring airflow can identify poorly ventilated indoor spaces where infectious particles circulate inside the room rather than getting replaced with fresh outdoor air\cite{Li2021BuildEnv}.
-However, airflow is rarely considered in modelling because it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
+However, airflow is rarely considered in modeling because it is difficult to assess in a naturally ventilated space, such as our primary care clinic, where the airflow depends on the outdoor wind direction and which doors and windows are open.
 
@@
 Survival and dispersion of aerosols are further determined by the physicochemical properties of virus-laden aerosols such as particle size, viral load, and other chemical components\cite{Wang2021Science}.
-Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modelling framework via the quanta generation rate parameter\cite{Riley1961Book,Rudnick2003IndoorAir}.
+Currently, these stochastic properties are only indirectly incorporated in the Wells-Riley modeling framework via the quanta generation rate parameter\cite{Riley1961Book,Rudnick2003IndoorAir}.
 It is challenging to consider these properties specifically, especially because they can be modified by environmental factors with often unclear, pathogen-specific effects\cite{Songer1967,Chan2011AdvVir,Fernstrom2013JoP,Cox1995Book,Fernstrom2013JoP,Tang2009Interface}.

### 2024-12-09 analysis/patient-tracking.Rmd
final cosmetic changes

@@
         d, ".rds"
-      ))
+      )) %>% dplyr::select(patient_id, time)
     }
@@
     plot.title.position = "plot"
+  )
+
+saveRDS(
+  time_in_clin_pl,
+  "results/data/time-in-clinic.rds"
 )
@@
 
+saveRDS(
+  time_pl,
+  "results/data/no-people-over-time.rds"
+)
+
 save_plot(
@@
 
+saveRDS(
+  spatial_pl,
+  "results/data/no-people-spatial.rds"
+)
+
 save_plot(

### 2024-12-09 preprocessing/combine_data_plots.r

@@ -1,0 +1,16 @@
+library(tidyverse)
+library(patchwork)
+source("utils/plotting.R")
+
+a <- readRDS("results/data/no-people-spatial.rds")
+b <- readRDS("results/data/no-people-over-time.rds")
+c <- readRDS("results/data/time-in-clinic.rds")
+d <- readRDS("results/data/air-changes-per-hour.rds")
+
+pl <- a + b + c + d + plot_layout(nrow = 2, heights = c(8, 6))
+
+save_plot(
+  pl,
+  pdf_file = "results/data/data-summary.png",
+  w = 16, h = 14
+)

### 2024-12-09 doc/paper/manuscript-revised.tex
Updates from Overleaf

@@
 \renewcommand{\paragraph}[1]{\vspace{0.3cm}\noindent\underline{\emph{#1}}\hfill\noindent}
-
-% word count
-% \newcommand{\maincount}[1]{%
-%   \immediate\write18{texcount -1 -sum=1 -merge -q -nobib #1.tex > #1-words.sum}%
-%   \input{#1-words.sum}%
-% }
-
-% \newcommand{\abstractcount}[1]{%
-%   \immediate\write18{texcount -template="{abst}" #1.tex > #1-words.sum}%
-%   \input{#1-words.sum}%
-% }
 
@@
 Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room.
-The average risk of infection per person was found to be \deleted{0.1\%}\added{0.05\%}.
+The average risk of infection per person was found to be \chdeleted{0.1\%}\chadded{0.05\%}.
 The risk increased with more close contact with infectious patients and longer time spent in the clinic.
 Without masks, the risk was 1.3 times higher.
-With lower ventilation rates, similar to pre-pandemic levels, the risk was \deleted{2.2}\added{2.1} times higher.
+With lower ventilation rates, similar to pre-pandemic levels, the risk was \chdeleted{2.2}\chadded{2.1} times higher.
 This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.
@@
 We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation.
-We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation was increased.
+We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when \chadded{community or surgical} masks were compulsory and ventilation was increased.
 We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
@@
 CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was \deleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\added{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}.
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and \deleted{2.2-fold (95\%-CrI 0.9$-$5.2)}\added{2.1-fold (95\%-CrI 0.9$-$5.0)} higher with pre-pandemic ventilation rates.
+The estimated risk of infection per clinic attendee was \chdeleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\chadded{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}.
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and \chdeleted{2.2-fold (95\%-CrI 0.9$-$5.2)}\chadded{2.1-fold (95\%-CrI 0.9$-$5.0)} higher with pre-pandemic ventilation rates\chadded{, emphasizing the importance of ventilation}.
 Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
@@
 Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Healthcare facilities are high-risk settings \deleted{because infectious individuals are more likely to visit clinics}\added{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}.
+Healthcare facilities are high-risk settings \chdeleted{because infectious individuals are more likely to visit clinics}\chadded{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}.
 A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
@@
 The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
-\added{The clinic was naturally ventilated by opening all doors and windows throughout the day.}
+\chadded{The clinic was naturally ventilated by opening all doors and windows throughout the day.}
 
@@
 Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
-On October~13, the corridor's CO$_2$ levels were missing due to a malfunctioning device and the waiting room's CO$_2$ levels were used for the corridor because their CO$_2$ measurements were similar on the other days \added{(see time-varying CO$_2$ levels and number of people per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
+On October~13, the corridor's CO$_2$ \chdeleted{levels}\chadded{measurements} were missing due to a malfunctioning device and the waiting room's CO$_2$ \chdeleted{levels}\chadded{measurements} were used for the corridor because their CO$_2$ \chdeleted{levels}\chadded{measurements} were similar on the other days \chadded{(see time-varying CO$_2$ levels and number of persons per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
 
@@
 \end{align}
-where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
+where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate \chadded{per person} (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
@@
 We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
-Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+Since mask use \chadded{(community or surgical masks)} was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
 
@@
 
-(3)~We calculated \added{time-varying} room occupancy from the video sensor data and combined it with the CO$_2$ measurements \added{(Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp)} to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
+(3)~We calculated \chadded{time-varying} room occupancy from the video sensor data and combined it with the CO$_2$ measurements \chadded{(Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp)} to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\deleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\deleted{quanta }h$^{-1}$\,$-$\,7.1\,\deleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-\deleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \deleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \added{which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$.
-Considering a drop height of 1.7\,m, the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
+We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\chdeleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\chdeleted{quanta }h$^{-1}$\,$-$\,7.1\,\chdeleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+\chdeleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \chdeleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \chadded{which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$.
+Considering a drop height of 1.7\,m \chadded{(average height of visitors)}, the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
 
@@
 
-(6)~\deleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\added{The quanta concentration was computed separately for the waiting room, corridor, and TB room.
-Each room was discretized into a grid of cubic cells, with each cell covering a squared area of 0.0625\,m$^2$.
+(6)~\chdeleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\chadded{The quanta concentration was computed separately for the waiting room, corridor, and TB room.
+Each room was discretized into a grid of cubic cells, each cell covering a squared area of 0.0625\,m$^2$.
 We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as}
+\chadded{
 \begin{align}\label{eq:spattemp-N}
-    % \deleted{\underbrace{N_{t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
+    % \chdeleted{\underbrace{N_{t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
 conc.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda\cdot 1\text{s}\right)}_{\text{removal}} ~.}
-     \added{\underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.}
+     \underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.
 \end{align}
-where \deleted{$D$}\added{$D_t$} is the diffusion constant (m$^2$\,s$^{-1}$) \added{varying by daytime}, $\Delta$ is the Laplace operator (second-order differential operator), \deleted{$I_t$ is the number of infectious individuals in space, } \added{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \deleted{$AER$ is the}\added{$AER_{r,t}$ is the room-specific} air change rate \added{varying by daytime}(\deleted{quanta}s$^{-1}$), $\lambda$ is the bacterial inactivation rate (\deleted{quanta}s$^{-1}$), \added{and $k$ is the gravitational settling rate (s$^{-1}$)}.
-\added{We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.}
+}
+where \chdeleted{$D$}\chadded{$D_t$} is the diffusion constant (m$^2$\,s$^{-1}$) \chadded{varying by daytime}, $\Delta$ is the Laplace operator (second-order differential operator), \chdeleted{$I_t$ is the number of infectious individuals in space, } \chadded{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \chdeleted{$AER$ is the}\chadded{$AER_{r,t}$ is the room-specific} air change rate \chadded{varying by daytime}(\chdeleted{quanta}s$^{-1}$), $\lambda$ is the bacterial inactivation rate (\chdeleted{quanta}s$^{-1}$), \chadded{and $k$ is the gravitational settling rate (s$^{-1}$)}.
+\chadded{We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.}
 
@@
 \begin{align}
-    % \deleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
-    \added{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),}
+    % \chdeleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
+    \chadded{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),}
 \end{align}
-where \deleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\added{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+where \chdeleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\chadded{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
 We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
-\added{We assumed no effect of mask use on the inhalation rate of quanta\cite{McCreesh2021BMJGlobalHealth}.}
+\chadded{In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.}
 
-\subsection*{\deleted{Monte Carlo simulation}}
+\subsection*{\chdeleted{Monte Carlo simulation}}
 
-\deleted{Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
-Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
+\chdeleted{Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
+Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and the TB room.
 Each room was rasterized into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
@@
 Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
-Except for the scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.
+\chdeleted{Except for scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.}
 
@@
 We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
-\added{We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp).
+\chadded{We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp).
 In each simulation, we computed the individual risk of infection based on the cumulative quanta exposure during the clinic visit.
 We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations.
-The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).} \deleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
+The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).} \chdeleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
 The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
@@
 Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
-Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mostly in the waiting room (\Cref{fig:input-data-descriptives}c).
-\added{Time spent in the corridor and TB room varied considerably as many patients just walked through the corridor and the TB room was reserved primarily for diagnosed or suspected TB patients.} Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) \deleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
-\added{CO$_2$ levels slightly varied within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp), and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
+Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mainly in the waiting room (\Cref{fig:input-data-descriptives}c).
+\chadded{Time spent in the corridor and TB room varied considerably as many patients just walked through the corridor and the TB room was reserved primarily for diagnosed or suspected TB patients.} Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) \chdeleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+\chadded{CO$_2$ levels varied slightly within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp), and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
 The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
-An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a \deleted{0.40\%}\added{0.20\%} risk of infection in the morning compared to a \deleted{0.07\%}\added{0.04\%} risk in the afternoon.
-The mean risk of infection per clinic visit was \deleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\added{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}, corresponding to a risk of \deleted{0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%)}\added{0.1\% per hour (80\%-CrI 0.01\%$-$0.21\%)}.
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a \chdeleted{0.40\%}\chadded{0.20\%} risk of infection in the morning compared to a \chdeleted{0.07\%}\chadded{0.04\%} risk in the afternoon.
+The mean risk of infection per clinic visit was \chdeleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\chadded{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}, corresponding to a risk of \chdeleted{0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%)}\chadded{0.1\% per hour (80\%-CrI 0.01\%$-$0.21\%)}.
 S2~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
@@
 The modeled individual risk of infection was associated with the number of close contacts and the time spent in the clinic.
-A doubling of close-contact encounters increased the odds of infection by \deleted{23\% (OR 1.23, 95\%-CrI 1.09$-$1.37)}\added{17\% (OR 1.17, 95\%-CrI 1.04$-$1.30)}.
-Doubling the time spent in the clinic increased the odds of infection by \deleted{12\% (OR 1.12, 95\%-CrI 1.03$-$1.21)}\added{12\% (OR 1.12, 95\%-CrI 1.03$-$1.22)}.
-The time spent in close contact was not significantly associated with the individual risk of infection \deleted{(OR 1.03, 95\%-CrI, 0.94$-$1.13)}\added{(OR 1.02, 95\%-CrI, 0.94$-$1.11)}.
+A doubling of close-contact encounters increased the odds of infection by \chdeleted{23\% (OR 1.23, 95\%-CrI 1.09$-$1.37)}\chadded{17\% (OR 1.17, 95\%-CrI 1.04$-$1.30)}.
+Doubling the time spent in the clinic increased the odds of infection by \chdeleted{12\% (OR 1.12, 95\%-CrI 1.03$-$1.21)}\chadded{12\% (OR 1.12, 95\%-CrI 1.03$-$1.22)}.
+The time spent in close contact was not significantly associated with the individual risk of infection \chdeleted{(OR 1.03, 95\%-CrI, 0.94$-$1.13)}\chadded{(OR 1.02, 95\%-CrI, 0.94$-$1.11)}.
 
@@
 The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been \deleted{0.12\% (80\%-CrI 0.02\%$-$0.13\%)}\added{0.07\% (80\%-CrI 0.01\%$-$0.08\%)}, an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been \deleted{0.19\% (80\%-CrI 0.02\%$-$0.20\%)}\added{0.09\% (80\%-CrI 0.01\%$-$0.10\%)}, an increase of \deleted{120\% (OR 2.2, 95\%-CrI 0.9$-$5.2)}\added{110\% (OR 2.1, 95\%-CrI 0.9$-$5.0)}.
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been \deleted{0.24\% (80\%-CrI 0.03\%$-$0.24\%)}\added{0.12\% (80\%-CrI 0.02\%$-$0.12\%)}, an increase of \deleted{180\% (OR 2.8, 95\%-CrI 1.1$-$6.6)}\added{170\% (OR 2.7, 95\%-CrI 1.1$-$6.6)}.
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been \chdeleted{0.12\% (80\%-CrI 0.02\%$-$0.13\%)}\chadded{0.07\% (80\%-CrI 0.01\%$-$0.08\%)}, an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been \chdeleted{0.19\% (80\%-CrI 0.02\%$-$0.20\%)}\chadded{0.09\% (80\%-CrI 0.01\%$-$0.10\%)}, an increase of \chdeleted{120\% (OR 2.2, 95\%-CrI 0.9$-$5.2)}\chadded{110\% (OR 2.1, 95\%-CrI 0.9$-$5.0)}.
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been \chdeleted{0.24\% (80\%-CrI 0.03\%$-$0.24\%)}\chadded{0.12\% (80\%-CrI 0.02\%$-$0.12\%)}, an increase of \chdeleted{180\% (OR 2.8, 95\%-CrI 1.1$-$6.6)}\chadded{170\% (OR 2.7, 95\%-CrI 1.1$-$6.6)}.
 The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
@@
 In sensitivity analyses, we varied assumptions about the number of infectious persons in the clinic.
-Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher \deleted{(0.12\%, 80\%-CrI 0.03\%$-$0.19\%)}\added{(0.06\%, 80\%-CrI 0.01\%$-$0.10\%)}.
-Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high \deleted{(0.22\%, 80\%-CrI <0.01\%$-$0.43\%)}\added{(0.11\%, 80\%-CrI <0.01\%$-$0.18\%)}.
+Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher \chdeleted{(0.12\%, 80\%-CrI 0.03\%$-$0.19\%)}\chadded{(0.06\%, 80\%-CrI 0.01\%$-$0.10\%)}.
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high \chdeleted{(0.22\%, 80\%-CrI <0.01\%$-$0.43\%)}\chadded{(0.11\%, 80\%-CrI <0.01\%$-$0.18\%)}.
 
@@
     \includegraphics{results/modeling/mean-roi-comparison.png}
-    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic attendee (median visit time 25\,min, IQR 13\,min$-$46\,min): \textbf{(a)}~with mask use and high ventilation as observed during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as observed before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
+    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic attendee (median visit time 25\,min, IQR 13\,min$-$46\,min): \textbf{(a)}~with mask use \chadded{(community or surgical masks)} and high ventilation as observed during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as observed before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
 Low risk: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}
@@
 We applied the model to real-world data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
-We estimated an average risk of infection per clinic attendee of \deleted{0.1\%}\added{0.05\%}.
+We estimated an average risk of infection per clinic attendee of \chdeleted{0.1\%}\chadded{0.05\%}.
 We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
-Without infection control measures (masks and high natural ventilation), the risk of infection would have been almost three times higher.
+Without infection control measures (\chadded{community/surgical} masks and high natural ventilation), the risk of infection would have been almost three times higher.
 Our spatiotemporal model can be easily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
@@
 Nevertheless, we estimated a lower risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
-Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about \deleted{2.4\%}\added{1.2\%}, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
+Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about \chdeleted{2.4\%}\chadded{1.2\%}, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
 The difference can be mainly attributed to the infection control measures that were in place during the COVID-19 pandemic.
 We estimated that without mask use and with pre-pandemic ventilation conditions, the risk of infection would have been more than twice as high.
-Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in the model.
+Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in \chdeleted{the model}\chadded{our hypothetical scenarios}.
 Other modeling assumptions may also have contributed.
@@
 Other studies suggest that mask-wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
-Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less pronounced.
+Our findings also show that the risk of infection is lower with mask use \chadded{(community or surgical masks)}, although the risk reduction estimated in our study is less pronounced.
 The reason for this is that the effect of mask use is compounded by the effect of very high natural ventilation rates, which determine both the diffusion and the removal of infectious quanta in the indoor space.
@@
 
-% \added{There is broad consensus that tuberculosis is a strictly airborne disease.
+% \chadded{There is broad consensus that tuberculosis is a strictly airborne disease.
 Therefore, public health authorities, including CDC and WHO, recommend FFP2 respirators for health care workers to prevent transmission and surgical face masks for patients with TB as source control to reduce the spread of infectious aerosols.
@@
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-\added{For example, the waiting room and corridor were separated to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels, yet the door between these areas was typically open, so that aerosolized \emph{Mtb} (quanta) could have spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
+\chadded{For example, the waiting room and corridor were treated as separate rooms to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels, but the door between these areas was typically open, allowing aerosolized \emph{Mtb} (quanta) to spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
 However, airflow is very difficult to measure.
 Furthermore, uncertainty in modeling parameters determining quanta generation and removal was often large and could be reduced in future work by incorporating, for example, additional information on patient characteristics from clinical records that are known to be associated with infectiousness\cite{Escombe2008PLoSMed} or susceptibility\cite{Furin2019Lancet}.
-Third, \added{we assumed that mask use only had an effect on the exhalation of \emph{Mtb} particles.
-Although the same assumption was made in a related modeling study\cite{McCreesh2021BMJGlobalHealth}, we note that our modeled risk of infection would be lower if we assumed an additional effect of mask use on inhalation.
-Fourth,} we could not validate our modeling results with empirical findings.
+Third, we could not validate our modeling results with empirical findings.
 Future research could test the practical implementation of our model by examining whether clinic attendees with a high risk of infection were also subsequently more likely to be diagnosed with TB.

### 2024-12-09 doc/paper/supplements.tex
Updates from Overleaf

@@
 Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}.
-Accordingly, we model the settling velocity using a Gamma distribution with mean $7.7\cdot 10^{-3}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-3}$m\,s$^{-1}$, which is slightly skewed towards slower settling velocities, reflecting a higher proportion of \emph{Mtb} at the lower end of the size range\cite{Fennelly2020Lancet}.
+Accordingly, we model the settling velocity using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$, which is slightly skewed towards slower settling velocities, reflecting a higher proportion of \emph{Mtb} at the lower end of the size range\cite{Fennelly2020Lancet}.
 Assuming a drop height of 1.7\,m, the distribution for the gravitational settling rate has a median of 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).

### 2024-12-09 doc/paper/manuscript.tex

@@
 
-\noindent\noindent\textbf{Supplementary Information:} Includes supplementary text, tables and figures.
+\noindent\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
 
-\noindent\textbf{Supplementary Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October 25 and moves through the waiting room.
+\noindent\textbf{S2 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.
+
+\clearpage
+
+\noindent\large{\textbf{Author Summary}}
+
+\noindent In crowded indoor settings in high tuberculosis burden countries, there is a high risk of tuberculosis transmission.
+We developed a model to study how tuberculosis is spread in a South African healthcare clinic during the COVID-19 pandemic when masks were mandatory and ventilation was improved.
+We tracked the movements of clinic attendees with video sensors and matched them with clinical records to find where and when infectious tuberculosis patients were present.
+We used this information to estimate the concentration of infectious particles in the air and the risk of infection for individuals.
+Over five days, the sensors tracked nearly 1,500 people in the clinic.
+Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room.
+The average risk of infection per person was found to be 0.1\%.
+The risk increased with more close contact with infectious patients and longer time spent in the clinic.
+Without masks, the risk was 1.3 times higher.
+With lower ventilation rates, similar to pre-pandemic levels, the risk was 2.2 times higher.
+This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.
 
@@
 \begin{abstract}\normalfont
-The risk of Mycobacterium tuberculosis (\emph{Mtb}) transmission can be high in crowded clinics.
-We developed a spatiotemporal model of airborne \emph{Mtb} transmission, based on the Wells-Riley equation.
-We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation increased.
+The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics.
+We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation.
+We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when masks were compulsory and ventilation was increased.
 We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
 We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
-Over five days, video sensors detected 1,438 clinic attendees.
+Over five days, video sensors tracked 1,438 clinic attendees.
 CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was 0.1\% (95\%-credible interval [CrI] 0.01\%$-$0.1\%).
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) with pre-pandemic ventilation rates.
+The estimated risk of infection per clinic attendee was 0.1\% (80\%-credible interval [CrI] 0.01\%$-$0.1\%).
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.2-fold (95\%-CrI 0.9$-$5.2) higher with pre-pandemic ventilation rates.
 Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
@@
 Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Healthcare facilities are high-risk setting because infectious individuals are more likely to visit a clinic\cite{McCreesh2020IJTLD}.
+Healthcare facilities are high-risk settings because infectious individuals are more likely to visit clinics\cite{McCreesh2020IJTLD}.
 A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
@@
 
-Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of infection $P$ as
+Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of \emph{Mtb} infection $P$ as
 \begin{align}
@@
 The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (\supp~Fig.~\zref{fig:undiagnosed-distribution}): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
+Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (\supp~Fig.~\zref{fig:quanta-distribution}): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
 Since masks were mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (\supp~Fig.~\zref{fig:undiagnosed-distribution}) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,quanta h$^{-1}$ (95\%-CrI 0.1\,quanta h$^{-1}$\,$-$\,7.1\,quanta h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
 We set the gravitational settling rate to zero and ignore sedimentation because particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of just $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}.
@@
 \end{align}
-where $D$ is the diffusion constant (m$^2$ s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta\,s$^{-1}$).
+where $D$ is the diffusion constant (m$^2$\,s$^{-1}$), $\Delta$ is the Laplace operator (second-order differential operator), $I_t$ is the number of infectious individuals in space, $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER$ is the air change rate (quanta\,s$^{-1}$), and $\lambda$ is bacterial inactivation rate (quanta\,s$^{-1}$).
 The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
@@
 where $N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
-We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$ h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$ h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m.
+We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
 
@@
 
-Simulations are described in detail in \supp~Section~\zref{sec:estimation}.
-In brief, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
-Each room was rasterized into a grid of cubic cells with an area of (0.25\,m)$^2$.
+Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
+Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and TB room.
+Each room was rasterised into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
 The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
@@
 
-We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
+We summarised clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
 Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).
-The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
-We used a Bayesian Beta regression model to estimate the odds ratio (OR) of the modeled risk of TB infection in association with the visit time, duration of close contact with, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
+The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarised with the mean and 80\%-credible interval (CrI).
+The mean risk of infection per clinic attendee was categorised as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
+We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modeled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
 Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1).
-Additional results not reported in the manuscript are shown in \supp~Section~\zref{sec:additional-results}.
+Additional results not reported in the manuscript are shown in Section~\zref{sec:additional-results} in \supp.
 
@@
 Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (\supp~Fig.~\zref{fig:co2-levels}) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}g), with a \eg median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm (Fig.~\zref{fig:co2-levels} in \supp) and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
 Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
@@
 South Africa has a high burden of TB and multi-drug-resistant TB\cite{WHO2023TBReport}.
-Nevertheless, we estimated a low risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
+Nevertheless, we estimated a lower risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
 Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 2.4\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
-The difference can be partly attributed to the infection control measures that were in place during the SARS-CoV-2 pandemic.
+The difference can be mainly attributed to the infection control measures that were in place during the COVID-19 pandemic.
 We estimated that without mask use and with pre-pandemic ventilation conditions, the risk of infection would have been more than twice as high.
-Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic.
-Further, different modelling assumptions could have contributed.
+Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in the model.
+Other modelling assumptions may also have contributed.
 The risk estimate from our previous study before the COVID-19 pandemic\cite{Zurcher2022JID} was based on TB prevalence in the South African population, rather than the TB status documented in the clinical data.
@@
 Other studies suggest that mask-wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
-Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less striking.
-The reason is that the effect of mask-wearing is compounded by the impact of very high ventilation rates, which determine both the diffusion and removal of infectious quanta in the indoor space.
+Our findings also show that the risk of infection is lower with mask use, although the risk reduction estimated in our study is less pronounced.
+The reason for this is that the effect of mask use is compounded by the effect of very high natural ventilation rates, which determine both the diffusion and the removal of infectious quanta in the indoor space.
 The air change rates in the clinic typically ranged from 10 to 20 air changes per hour, which resulted in a quick diffusion and removal of infectious quanta from the indoor air.
-To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
-Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely due to a better ventilation, with an air change rate comparable to the one measured in the primary care clinic of our study.
-We stress that our results do \emph{not} suggest that masks are ineffective but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
+To put this into perspective, we recently modeled the risk of \emph{Mtb} transmission in schools by comparing ventilation conditions in South Africa (1.5 air changes per hour), Switzerland (0.7), and Tanzania (13.7)\cite{Banholzer2024PGPH}.
+Tanzania had a much lower estimated risk of \emph{Mtb} transmission in schools than the other countries, largely because of better natural ventilation, with an air change rate comparable to the one measured in the primary care clinic in our study.
+We emphasise that our results do \emph{not} suggest that masks are ineffective -- there is a large body of evidence for their effectiveness -- but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
 
@@
 First, tracking patient movements with video sensors was challenging.
-It required considerable data processing and many tracks were interrupted and could not be reconnected.
+It required considerable data processing and many interrupted tracks could not be reconnected.
 Therefore, some clinic attendees may have had longer visits and the risk of infection may have been underestimated.

### 2025-02-01 doc/paper/manuscript-revised-clean.tex
Updates from Overleaf

--- 2024-12-09 doc/paper/manuscript-revised.tex
@@
 
-\begin{information}
+\begin{abstract}\normalfont
+The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics.
+We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation.
+We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when community or surgical masks were compulsory and ventilation was increased.
+We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
+We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
+Over five days, video sensors tracked 1,438 clinic attendees.
+CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
+The estimated risk of infection per clinic attendee was 0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%).
+It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.1-fold (95\%-CrI 0.9$-$5.0) higher with pre-pandemic ventilation rates, emphasizing the importance of ventilation.
+Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
 
-\noindent\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
+\par
+\end{abstract}
 
-\noindent\textbf{S2 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.
+\flushbottom
+\maketitle
+\setcounter{page}{1}
+\thispagestyle{fancy}
 
-\clearpage
+\newpage
+
+\sloppy
+\raggedbottom
+
+\newpage
 
@@
 Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room.
-The average risk of infection per person was found to be \chdeleted{0.1\%}\chadded{0.05\%}.
+The average risk of infection per person was found to be 0.05\%.
 The risk increased with more close contact with infectious patients and longer time spent in the clinic.
 Without masks, the risk was 1.3 times higher.
-With lower ventilation rates, similar to pre-pandemic levels, the risk was \chdeleted{2.2}\chadded{2.1} times higher.
+With lower ventilation rates, similar to pre-pandemic levels, the risk was 2.1 times higher.
 This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.
-
-\end{information}
-
-\begin{abstract}\normalfont
-The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics.
-We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation.
-We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when \chadded{community or surgical} masks were compulsory and ventilation was increased.
-We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients.
-We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection.
-Over five days, video sensors tracked 1,438 clinic attendees.
-CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room.
-The estimated risk of infection per clinic attendee was \chdeleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\chadded{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}.
-It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and \chdeleted{2.2-fold (95\%-CrI 0.9$-$5.2)}\chadded{2.1-fold (95\%-CrI 0.9$-$5.0)} higher with pre-pandemic ventilation rates\chadded{, emphasizing the importance of ventilation}.
-Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.
-
-\par
-\end{abstract}
-
-%TC:ignore
-
-\flushbottom
-\maketitle
-\setcounter{page}{1}
-\thispagestyle{fancy}
-
-\vspace{2em}
-
-%\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)
-
-\vspace{0.5em}
-
-\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, person tracking, spatiotemporal model
-% maximum of 3-5 keywords
-\newpage
-
-\sloppy
-\raggedbottom
-%TC:endignore
 
@@
 Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}.
-Healthcare facilities are high-risk settings \chdeleted{because infectious individuals are more likely to visit clinics}\chadded{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}.
+Healthcare facilities are high-risk settings where infectious individuals are more likely to visit\cite{McCreesh2020IJTLD}.
 A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.
@@
 
+\subsection*{Ethics statement}
+
+The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 536/2020), the City of Cape Town (Project ID: 8139), South Africa, and the Ethics Committee of the Canton of Bern (KEK/REF: 2019-02131), Switzerland, approved the study.
+
 \subsection*{Study design and setting}
@@
 It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}.
-We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
+We defined three areas within the clinic (Fig~\ref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m).
 The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB.
-\chadded{The clinic was naturally ventilated by opening all doors and windows throughout the day.}
+The clinic was naturally ventilated by opening all doors and windows throughout the day.
 
@@
 We collected environmental and clinical data and data on the movements of patients in the clinic.
-Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
-On October~13, the corridor's CO$_2$ \chdeleted{levels}\chadded{measurements} were missing due to a malfunctioning device and the waiting room's CO$_2$ \chdeleted{levels}\chadded{measurements} were used for the corridor because their CO$_2$ \chdeleted{levels}\chadded{measurements} were similar on the other days \chadded{(see time-varying CO$_2$ levels and number of persons per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.
+Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (Fig \ref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals.
+On October~13, the corridor's CO$_2$ measurements were missing due to a malfunctioning device and the waiting room's CO$_2$ measurements were used for the corridor because their CO$_2$ measurements were similar on the other days (see time-varying CO$_2$ levels and number of persons per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives}).
 
@@
 
-We used an anonymous person tracking system using video sensors (Xovis, Zollikofen, Switzerland) to track the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}).
+We used an anonymous person tracking system using video sensors (Xovis, Zollikofen, Switzerland) to track the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (Fig~\ref{fig:floor-plan}).
 The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic.
 People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person.
-To address this, we created an R Shiny application (Fig.~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp.
+To address this, we created an R Shiny application (Fig~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp.
 We also labeled the tracks to identify the movements of clinic staff.
-\Cref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.
+Fig~\ref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.
 
@@
 \end{align}
-where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate \chadded{per person} (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
+where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate per person (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).
 The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection.
@@
 This way, the spatiotemporal location of infectious patients who generate quanta could be determined.
-A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig.~\zref{fig:toy-example} in \supp).
-In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).
+A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig~\zref{fig:toy-example} in \supp).
+In the following, we describe the six main components of our modeling approach (Fig~\ref{fig:modeling-flow}).
 
@@
 The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}.
-Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
+Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients.
 Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious.
-We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
-Since mask use \chadded{(community or surgical masks)} was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
+We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}.
+Since mask use (community or surgical masks) was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
 
@@
 
-(3)~We calculated \chadded{time-varying} room occupancy from the video sensor data and combined it with the CO$_2$ measurements \chadded{(Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp)} to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
+(3)~We calculated time-varying room occupancy from the video sensor data and combined it with the CO$_2$ measurements (Figs~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp) to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.
 
@@
 The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate.
-We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\chdeleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\chdeleted{quanta }h$^{-1}$\,$-$\,7.1\,\chdeleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
-\chdeleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \chdeleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \chadded{which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$.
-Considering a drop height of 1.7\,m \chadded{(average height of visitors)}, the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}
+We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,h$^{-1}$ (95\%-CrI 0.1\, h$^{-1}$\,$-$\,7.1\,h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}.
+Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$.
+Considering a drop height of 1.7\,m (average height of visitors), the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).
 
@@
 
-(6)~\chdeleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\chadded{The quanta concentration was computed separately for the waiting room, corridor, and TB room.
+(6)~The quanta concentration was computed separately for the waiting room, corridor, and TB room.
 Each room was discretized into a grid of cubic cells, each cell covering a squared area of 0.0625\,m$^2$.
-We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as}
-\chadded{
+We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as
 \begin{align}\label{eq:spattemp-N}
-    % \chdeleted{\underbrace{N_{t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev.
-conc.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda\cdot 1\text{s}\right)}_{\text{removal}} ~.}
      \underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.
 \end{align}
-}
-where \chdeleted{$D$}\chadded{$D_t$} is the diffusion constant (m$^2$\,s$^{-1}$) \chadded{varying by daytime}, $\Delta$ is the Laplace operator (second-order differential operator), \chdeleted{$I_t$ is the number of infectious individuals in space, } \chadded{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \chdeleted{$AER$ is the}\chadded{$AER_{r,t}$ is the room-specific} air change rate \chadded{varying by daytime}(\chdeleted{quanta}s$^{-1}$), $\lambda$ is the bacterial inactivation rate (\chdeleted{quanta}s$^{-1}$), \chadded{and $k$ is the gravitational settling rate (s$^{-1}$)}.
-\chadded{We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.}
+where $D_t$ is the diffusion constant (m$^2$\,s$^{-1}$) varying by daytime, $\Delta$ is the Laplace operator (second-order differential operator), $I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$), $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER_{r,t}$ is the room-specific air change rate varying by daytime(s$^{-1}$), $\lambda$ is the bacterial inactivation rate (s$^{-1}$), and $k$ is the gravitational settling rate (s$^{-1}$).
+We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.
 
@@
 \begin{align}
-    % \chdeleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
-    \chadded{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),}
+    P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),
 \end{align}
-where \chdeleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\chadded{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
+where $\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$.
 We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$.
-\chadded{In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.}
+In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.
 
-\subsection*{\chdeleted{Monte Carlo simulation}}
+\subsection*{Monte Carlo simulation}
 
-\chdeleted{Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
+Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
 Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and the TB room.
@@
 We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters.
-In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.}
+In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
 
@@
 Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.
-\chdeleted{Except for scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.}
 
@@
 We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics.
-\chadded{We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp).
+We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp).
 In each simulation, we computed the individual risk of infection based on the cumulative quanta exposure during the clinic visit.
 We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations.
-The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).} \chdeleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations.
-The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
+The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).
+The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%).
 We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modeled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min).
@@
 Additional results not reported in the manuscript are shown in Section~\zref{sec:additional-results} in \supp.
-
-\subsection*{Ethics statement}
-
-The University of Cape Town Faculty of Health Sciences Human Research Ethics Committee (HREC/REF: 536/2020), the City of Cape Town (Project ID: 8139), South Africa, and the Ethics Committee of the Canton of Bern (KEK/REF: 2019-02131), Switzerland, approved the study.
 
@@
 Overall, 1,563 person movements were detected based on the video sensor data; 125 likely from clinic staff (movements from and to staff rooms) and 1,438 from clinic attendees (registered patients and other visitors).
-Most movements were detected in the waiting room (\Cref{fig:input-data-descriptives}a) and during the morning (\Cref{fig:input-data-descriptives}b).
-Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mainly in the waiting room (\Cref{fig:input-data-descriptives}c).
-\chadded{Time spent in the corridor and TB room varied considerably as many patients just walked through the corridor and the TB room was reserved primarily for diagnosed or suspected TB patients.} Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
-The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) \chdeleted{(Fig.~\zref{fig:co2-levels} in \supp)} and the corresponding air change rates were high throughout the study (\Cref{fig:input-data-descriptives}d), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
-\chadded{CO$_2$ levels varied slightly within and between clinic days, largely corresponding to changes in room occupancy during the day (Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp), and possibly small variations in ventilation conditions between days.} Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
+Most movements were detected in the waiting room (Fig~\ref{fig:input-data-descriptives}A) and during the morning (Fig~\ref{fig:input-data-descriptives}B).
+Attendees spent about half an hour in the clinic (median 25\,min, interquartile range (IQR) 13\,min$-$46\,min), mainly in the waiting room (Fig~\ref{fig:input-data-descriptives}C).
+Time spent in the corridor and TB room varied considerably as many patients just walked through the corridor and the TB room was reserved primarily for diagnosed or suspected TB patients.
+Close-contact encounters were frequent (median 5, IQR 2$-$9 per attendee) and attendees were in close contact with at least one other person for more than half of the duration of their visit (median 68\%, IQR 41\%$-$87\%).
+The CO$_2$ levels in the clinic were low (median 436\,ppm, IQR 410\,ppm$-$462\,ppm) and the corresponding air change rates were high throughout the study (Fig~\ref{fig:input-data-descriptives}D), with a median of 17 air changes per hour (IQR 9\,h$^{-1}$$-$23\,h$^{-1}$) in the densely occupied waiting room.
+CO$_2$ levels varied slightly within and between clinic days, largely corresponding to changes in room occupancy during the day (Figs~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp), and possibly small variations in ventilation conditions between days.
+Median temperature in the clinic was 22$^{\circ}$C (IQR 21$^{\circ}$C$-$23$^{\circ}$C) and relative humidity was 55\% (IQR 47\%$-$61\%).
 
@@
 
-The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (\Cref{fig:main-modeling-results}).
-An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a \chdeleted{0.40\%}\chadded{0.20\%} risk of infection in the morning compared to a \chdeleted{0.07\%}\chadded{0.04\%} risk in the afternoon.
-The mean risk of infection per clinic visit was \chdeleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\chadded{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}, corresponding to a risk of \chdeleted{0.21\% per hour (80\%-CrI 0.03\%$-$0.36\%)}\chadded{0.1\% per hour (80\%-CrI 0.01\%$-$0.21\%)}.
-S2~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
+The quanta concentration was higher in the morning than in the afternoon, and higher in the waiting room than in the corridor or TB room of the clinic (Fig~\ref{fig:main-modeling-results}).
+An exposure of one hour to the mean quanta concentration in the middle of the waiting room would translate to a 0.20\% risk of infection in the morning compared to a 0.04\% risk in the afternoon.
+The mean risk of infection per clinic visit was 0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%), corresponding to a risk of 0.1\% per hour (80\%-CrI 0.01\%$-$0.21\%).
+S1~Video shows an animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic on the morning of October~25 and moves through the waiting room.
 It can be seen that the quanta concentration was higher near the infectious patient, following their position and movement.
@@
 The modeled individual risk of infection was associated with the number of close contacts and the time spent in the clinic.
-A doubling of close-contact encounters increased the odds of infection by \chdeleted{23\% (OR 1.23, 95\%-CrI 1.09$-$1.37)}\chadded{17\% (OR 1.17, 95\%-CrI 1.04$-$1.30)}.
-Doubling the time spent in the clinic increased the odds of infection by \chdeleted{12\% (OR 1.12, 95\%-CrI 1.03$-$1.21)}\chadded{12\% (OR 1.12, 95\%-CrI 1.03$-$1.22)}.
-The time spent in close contact was not significantly associated with the individual risk of infection \chdeleted{(OR 1.03, 95\%-CrI, 0.94$-$1.13)}\chadded{(OR 1.02, 95\%-CrI, 0.94$-$1.11)}.
+A doubling of close-contact encounters increased the odds of infection by 17\% (OR 1.17, 95\%-CrI 1.04$-$1.30).
+Doubling the time spent in the clinic increased the odds of infection by 12\% (OR 1.12, 95\%-CrI 1.03$-$1.22).
+The time spent in close contact was not significantly associated with the individual risk of infection (OR 1.02, 95\%-CrI, 0.94$-$1.11).
 
@@
 
-The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  \Cref{fig:scenario-results}.
-If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been \chdeleted{0.12\% (80\%-CrI 0.02\%$-$0.13\%)}\chadded{0.07\% (80\%-CrI 0.01\%$-$0.08\%)}, an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
-If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been \chdeleted{0.19\% (80\%-CrI 0.02\%$-$0.20\%)}\chadded{0.09\% (80\%-CrI 0.01\%$-$0.10\%)}, an increase of \chdeleted{120\% (OR 2.2, 95\%-CrI 0.9$-$5.2)}\chadded{110\% (OR 2.1, 95\%-CrI 0.9$-$5.0)}.
-Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been \chdeleted{0.24\% (80\%-CrI 0.03\%$-$0.24\%)}\chadded{0.12\% (80\%-CrI 0.02\%$-$0.12\%)}, an increase of \chdeleted{180\% (OR 2.8, 95\%-CrI 1.1$-$6.6)}\chadded{170\% (OR 2.7, 95\%-CrI 1.1$-$6.6)}.
-The scenarios show an increase in the number of clinic attendees with a high risk of infection (\Cref{fig:scenario-results}).
+The hypothetical scenarios assuming no mask use and pre-pandemic ventilation rates are shown in  Fig~\ref{fig:scenario-results}.
+If none of the clinic attendees or staff had worn masks, the mean risk of infection per visit would have been 0.07\% (80\%-CrI 0.01\%$-$0.08\%), an increase of 30\% (OR 1.3, 95\%-CrI 1.2$-$1.4).
+If the air change rate had been the same as before the COVID-19 pandemic, the mean risk of infection would have been 0.09\% (80\%-CrI 0.01\%$-$0.10\%), an increase of 110\% (OR 2.1, 95\%-CrI 0.9$-$5.0).
+Without mask use and with pre-pandemic ventilation rates, the mean risk of infection would have been 0.12\% (80\%-CrI 0.02\%$-$0.12\%), an increase of 170\% (OR 2.7, 95\%-CrI 1.1$-$6.6).
+The scenarios show an increase in the number of clinic attendees with a high risk of infection (Fig~\ref{fig:scenario-results}).
 
 In sensitivity analyses, we varied assumptions about the number of infectious persons in the clinic.
-Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher \chdeleted{(0.12\%, 80\%-CrI 0.03\%$-$0.19\%)}\chadded{(0.06\%, 80\%-CrI 0.01\%$-$0.10\%)}.
-Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high \chdeleted{(0.22\%, 80\%-CrI <0.01\%$-$0.43\%)}\chadded{(0.11\%, 80\%-CrI <0.01\%$-$0.18\%)}.
+Assuming that the number of infectious TB patients reflected the prevalence of TB in the South African population, the mean risk of infection would have been slightly higher (0.06\%, 80\%-CrI 0.01\%$-$0.10\%).
+Further, assuming that both diagnosed TB patients and clinic attendees suspected of having TB were infectious, the mean risk of infection would have been twice as high (0.11\%, 80\%-CrI <0.01\%$-$0.18\%).
 
@@
     \includegraphics{results/modeling/mean-roi-comparison.png}
-    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic attendee (median visit time 25\,min, IQR 13\,min$-$46\,min): \textbf{(a)}~with mask use \chadded{(community or surgical masks)} and high ventilation as observed during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as observed before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
+    \caption{\textbf{Impact of infection control measures.} Mean risk of infection per clinic attendee (median visit time 25\,min, IQR 13\,min$-$46\,min): \textbf{(a)}~with mask use (community or surgical masks) and high ventilation as observed during the COVID-19 pandemic, \textbf{(b)}~without mask use and high ventilation, \textbf{(c)}~with mask use and normal ventilation as observed before the COVID-19 pandemic  \textbf{(d)}~without mask use and normal ventilation.
 Low risk: <0.1\%, Medium: 0.1-1\%, High: >1\% risk of infection.}
@@
 We applied the model to real-world data from a South African primary care clinic collected in 2021 during the COVID-19 pandemic to examine the risk of \emph{Mtb} infection among clinic attendees.
-We estimated an average risk of infection per clinic attendee of \chdeleted{0.1\%}\chadded{0.05\%}.
+We estimated an average risk of infection per clinic attendee of 0.05\%.
 We found that clinic attendees who stayed longer and had more close contact encounters had a higher risk of infection.
-Without infection control measures (\chadded{community/surgical} masks and high natural ventilation), the risk of infection would have been almost three times higher.
+Without infection control measures (community/surgical masks and high natural ventilation), the risk of infection would have been almost three times higher.
 Our spatiotemporal model can be easily adapted to model airborne transmission risks of other respiratory infections such as SARS-CoV-2.
@@
 Nevertheless, we estimated a lower risk of infection per clinic attendee in our primary care clinic under pandemic conditions with strict infection control measures in place.
-Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about \chdeleted{2.4\%}\chadded{1.2\%}, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
+Assuming clinic visits amounting to 12\,hours per year, we estimated a cumulative annual risk of infection of about 1.2\%, which contrasts with an estimated risk of 9\% to 29\% from before the pandemic in the same clinic\cite{Zurcher2022JID}.
 The difference can be mainly attributed to the infection control measures that were in place during the COVID-19 pandemic.
 We estimated that without mask use and with pre-pandemic ventilation conditions, the risk of infection would have been more than twice as high.
-Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in \chdeleted{the model}\chadded{our hypothetical scenarios}.
+Patient triage and fixed appointments probably also contributed to a lower risk of \emph{Mtb} transmission during the COVID-19 pandemic, which was not considered in our hypothetical scenarios.
 Other modeling assumptions may also have contributed.
@@
 Other studies suggest that mask-wearing could reduce the risk of \emph{Mtb} transmission by more than 50\%\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.
-Our findings also show that the risk of infection is lower with mask use \chadded{(community or surgical masks)}, although the risk reduction estimated in our study is less pronounced.
+Our findings also show that the risk of infection is lower with mask use (community or surgical masks), although the risk reduction estimated in our study is less pronounced.
 The reason for this is that the effect of mask use is compounded by the effect of very high natural ventilation rates, which determine both the diffusion and the removal of infectious quanta in the indoor space.
@@
 We emphasize that our results do \emph{not} suggest that masks are ineffective -- there is a large body of evidence for their effectiveness -- but that their effect is outweighed by the excellent ventilation conditions in the clinic during the COVID-19 pandemic.
-
-% \chadded{There is broad consensus that tuberculosis is a strictly airborne disease.
-Therefore, public health authorities, including CDC and WHO, recommend FFP2 respirators for health care workers to prevent transmission and surgical face masks for patients with TB as source control to reduce the spread of infectious aerosols.
-Although surgical masks and community masks may also have a small effect in preventing transmission on inhalation, this effect is likely to be negligible and, to our knowledge, no studies have defined this modeling parameter.
-Similar modeling studies of the effect of infection and control measures on Mycobacterium tuberculosis transmission in healthcare settings (e.g., McCreesh et al., BMJ Global Health 2021, doi:10.1136/bmjgh-2021-007124) did not consider the impact of face masks on inhalation.
-We believe it is reasonable to include facemasks in the transmission model only as a factor in reducing the spread of infectious aerosols in people with tuberculosis.}
 
@@
 Second, several important assumptions had to be made regarding the generation, diffusion, and removal of infectious quanta.
-\chadded{For example, the waiting room and corridor were treated as separate rooms to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels, but the door between these areas was typically open, allowing aerosolized \emph{Mtb} (quanta) to spread from one area to the other.} We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
+For example, the waiting room and corridor were treated as separate rooms to facilitate modeling and because we observed differences in room occupancy and time-varying CO$_2$ levels, but the door between these areas was typically open, allowing aerosolized \emph{Mtb} (quanta) to spread from one area to the other.
+We also assumed that infectious particles move radially from the source and approach a steady state with uniform contamination of the airspace, which may depend on airflow.
 However, airflow is very difficult to measure.
@@
 
-%TC:ignore
-\section*{Acknowledgements}
+\section*{Supplementary Information}
+
+\noindent\noindent\textbf{S1 Appendix:} Includes supplementary text, tables and figures.
+
+\noindent\textbf{S1 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.
+
+\section*{Acknowledgments}
 We are grateful to the clinical manager and the staff members at the primary care clinic.
@@
 
-\section*{Data availability}
-Restrictions on the availability of personal data apply, but are necessary to maintain the confidentiality of participants.
-The data is available upon reasonable request, contact: University of Bern, info.ispm@unibe.ch.
-All other data are included in the manuscript and accompanying supporting information.
-Code used for preprocessing, modeling and analysis, descriptive and simulation results are available on GitHub via \url{https://github.com/nbanho/clinic-transmission}.
-
-\section*{Funding}
-The study was supported by a grant from the Swiss National Science Foundation (grant no.
-CRSK-3\textunderscore190781).
-NB and LF are supported by the National Institute of Allergy and Infectious Diseases (NIAID) through grant no.
-5U01-AI069924-05.
-ME is supported by special project funding from the Swiss National Science Foundation [grant no.
-32FP30-189498].
-The funders had no role in study design, data collection, data analysis, data interpretation, or writing of the report.
-
-\section*{Competing interests}
-The authors have declared that no competing interests exist.
-
-\section*{Author contributions}
-Conception and design: NB, KM, KZ, LF.
-Data collection: KM, JL, RS, KZ, LF.
-Modeling and statistical analysis: NB.
-Paper draft: NB, LF, ME.
-All authors reviewed and approved the final version of the manuscript.
-
 \newpage
 
-\bibliography{references.bib}
-%TC:endignore
+\begin{thebibliography}{10}
+
+\bibitem{WHO2023TBReport}
+{World Health Organization}.
+\newblock Global {Tuberculosis} {Report} 2023; 2023.
+\newblock Available from:
+  \url{https://www.who.int/teams/global-tuberculosis-programme/tb-reports/global-tuberculosis-report-2023}.
+
+\bibitem{Rieder1999}
+Rieder HL. Disease IUATaL, editor. Epidemiologic basis of tuberculosis control.
+\newblock International Union Against Tuberculosis and Lung Disease; 1999.
+
+\bibitem{Patterson2021Tuberculosis}
+Patterson B, Bryden W, Call C, McKerry A, Leonard B, Seldon R, et~al.
+\newblock Cough-independent production of viable {Mycobacterium} tuberculosis
+  in bioaerosol.
+\newblock Tuberculosis. 2021;126:102038.
+\newblock doi:10.1016/j.tube.2020.102038.
+
+\bibitem{Fennelly2020Lancet}
+Fennelly KP.
+\newblock Particle sizes of infectious aerosols: implications for infection
+  control.
+\newblock Lancet Respir Med. 2020;8(9):914--924.
+\newblock doi:10.1016/S2213-2600(20)30323-4.
+
+\bibitem{Loudon1969AMRRD}
+Loudon RG, Bumgarner LR, Lacy J, Coffman GK.
+\newblock Aerial transmission of mycobacteria.
+\newblock Am Rev Respir Dis. 1969;100(2):165--171.
+\newblock doi:10.1164/arrd.1969.100.2.165.
+
+\bibitem{CPS2013Book}
+{Center for Prevention Services  Division of Tuberculosis Elimination}.
+\newblock Core {curriculum} on {Tuberculosis}: {What} the {clinician} {should}
+  {know}.
+\newblock 6th ed. U.S. Department of Health \& Human Services, Centers for
+  Disease Control and Prevention, National Center for HIV, STD, and TB
+  Prevention, Division of Tuberculosis Elimination; 2013.
+
+\bibitem{Nardell1991ARRD}
+Nardell EA, Keegan J, Cheney SA, Etkind SC.
+\newblock Airborne infection: {Theoretical} limits of protection achievable by
+  building ventilation.
+\newblock Am Rev Respir Dis. 1991;144(2):302--306.
+\newblock doi:10.1164/ajrccm/144.2.302.
+
+\bibitem{Wang2021Science}
+Wang CC, Prather KA, Sznitman J, Jimenez JL, Lakdawala SS, Tufekci Z, et~al.
+\newblock Airborne transmission of respiratory viruses.
+\newblock Science. 2021;373(6558):eabd9149.
+\newblock doi:10.1126/science.abd9149.
+
+\bibitem{Morawska2021}
+Morawska L, Allen J, Bahnfleth W, Bluyssen PM, Boerstra A, Buonanno G, et~al.
+\newblock A paradigm shift to combat indoor respiratory infection.
+\newblock Science. 2021;372(6543):689--691.
+\newblock doi:10.1126/science.abg2025.
+
+\bibitem{McCreesh2020IJTLD}
+McCreesh N, Grant AD, Yates TA, Karat AS, White RG.
+\newblock Tuberculosis from transmission in clinics in high {HIV} settings may
+  be far higher than contact data suggest.
+\newblock Int J Tuberc Lung Dis. 2020;24(4):403--408.
+\newblock doi:10.5588/ijtld.19.0410.
+
+\bibitem{McCreesh2022BMJGlobalHealth}
+McCreesh N, Karat AS, Govender I, Baisley K, Diaconu K, Yates TA, et~al.
+\newblock Estimating the contribution of transmission in primary healthcare
+  clinics to community-wide {TB} disease incidence, and the impact of infection
+  prevention and control interventions, in {KwaZulu}-{Natal}, {South} {Africa}.
+\newblock BMJ Glob Health. 2022;7(4):e007136.
+\newblock doi:10.1136/bmjgh-2021-007136.
+
+\bibitem{Riley1961Book}
+Riley RL, O'Grady F.
+\newblock Airborne {infection}: {Transmission} and {control}.
+\newblock Macmillan; 1961.
+
+\bibitem{Haddrell2024NatCommun}
+Haddrell A, Oswin H, Otero-Fernandez M, Robinson JF, Cogan T, Alexander R,
+  et~al.
+\newblock Ambient carbon dioxide concentration correlates with {SARS}-{CoV}-2
+  aerostability and infection risk.
+\newblock Nat Commun. 2024;15:3487.
+\newblock doi:10.1038/s41467-024-47777-5.
+
+\bibitem{Andrews2014JID}
+Andrews JR, Morrow C, Walensky RP, Wood R.
+\newblock Integrating social contact and environmental data in evaluating
+  {Tuberculosis} transmission in a {South} {African} township.
+\newblock J Infect Dis. 2014;210(4):597--603.
+\newblock doi:10.1093/infdis/jiu138.
+
+\bibitem{Taylor2016IJTLD}
+Taylor JG, Yates TA, Mthethwa M, Tanser F, Abubakar I, Altamirano H.
+\newblock Measuring ventilation and modelling {M}. tuberculosis transmission in
+  indoor congregate settings, rural {KwaZulu}-{Natal}.
+\newblock Int J Tuberc Lung Dis. 2016;20(9):1155--1161.
+\newblock doi:10.5588/ijtld.16.0085.
+
+\bibitem{Hella2017JInfect}
+Hella J, Morrow C, Mhimbira F, Ginsberg S, Chitnis N, Gagneux S, et~al.
+\newblock Tuberculosis transmission in public locations in {Tanzania}: {A}
+  novel approach to studying airborne disease transmission.
+\newblock J Infect. 2017;75(3):191--197.
+\newblock doi:10.1016/j.jinf.2017.06.009.
+
+\bibitem{Zemouri2020JDR}
+Zemouri C, Awad SF, Volgenant CMC, Crielaard W, Laheij AMGA, de~Soet JJ.
+\newblock Modeling of the transmission of Coronaviruses, Measles virus,
+  Influenza virus, Mycobacterium tuberculosis, and Legionella pneumophila in
+  dental clinics.
+\newblock J Dent Res. 2020;99(10):1192--1198.
+\newblock doi:10.1177/0022034520940288.
+
+\bibitem{Zurcher2022JID}
+Zürcher K, Riou J, Morrow C, Ballif M, Koch A, Bertschinger S, et~al.
+\newblock Estimating tuberculosis transmission risks in a primary care clinic
+  in {South} {Africa}: {Modeling} of environmental and clinical data.
+\newblock J Infect Dis. 2022;225(9):1642--1652.
+\newblock doi:10.1093/infdis/jiab534.
+
+\bibitem{McCreesh2021BMJGlobalHealth}
+McCreesh N, Karat AS, Baisley K, Diaconu K, Bozzani F, Govender I, et~al.
+\newblock Modelling the effect of infection prevention and control measures on
+  rate of {Mycobacterium} tuberculosis transmission to clinic attendees in
+  primary health clinics in {South} {Africa}.
+\newblock BMJ Glob Health. 2021;6(10):e007124.
+\newblock doi:10.1136/bmjgh-2021-007124.
+
+\bibitem{Rudnick2003IndoorAir}
+Rudnick SN, Milton DK.
+\newblock Risk of indoor airborne infection transmission estimated from carbon
+  dioxide concentration.
+\newblock Indoor Air. 2003;13(3):237--245.
+\newblock doi:10.1034/j.1600-0668.2003.00189.x.
+
+\bibitem{Richardson2014PONE}
+Richardson ET, Morrow CD, Kalil DB, Bekker LG, Wood R.
+\newblock Shared air: {A} renewed focus on ventilation for the prevention of
+  {Tuberculosis} transmission.
+\newblock PLOS ONE. 2014;9(5):e96334.
+\newblock doi:10.1371/journal.pone.0096334.
+
+\bibitem{Vuorinen2020SafSci}
+Vuorinen V, Aarnio M, Alava M, Alopaeus V, Atanasova N, Auvinen M, et~al.
+\newblock Modelling aerosol transport and virus exposure with numerical
+  simulations in relation to {SARS}-{CoV}-2 transmission by inhalation indoors.
+\newblock Saf Sci. 2020;130:104866.
+\newblock doi:10.1016/j.ssci.2020.104866.
+
+\bibitem{Chen2020BuildEnv}
+Chen W, Zhang N, Wei J, Yen HL, Li Y.
+\newblock Short-range airborne route dominates exposure of respiratory
+  infection during close contact.
+\newblock Build Environ. 2020;176:106859.
+\newblock doi:10.1016/j.buildenv.2020.106859.
+
+\bibitem{Ko2004RiskAnal}
+Ko G, Thompson KM, Nardell EA.
+\newblock Estimation of tuberculosis risk on a commercial airliner.
+\newblock Risk Anal. 2004;24(2):379--388.
+\newblock doi:10.1111/j.0272-4332.2004.00439.x.
+
+\bibitem{Kenyon1996NEJM}
+Kenyon TA, Valway SE, Ihle WW, Onorato IM, Castro KG.
+\newblock Transmission of multidrug-resistant {Mycobacterium} tuberculosis
+  during a long airplane flight.
+\newblock N Engl J Med. 1996;334(15):933--938.
+\newblock doi:10.1056/NEJM199604113341501.
+
+\bibitem{Zurcher2020BMJ}
+Zürcher K, Morrow C, Riou J, Ballif M, Koch AS, Bertschinger S, et~al.
+\newblock Novel approach to estimate {Tuberculosis} transmission in primary
+  care clinics in sub-{Saharan} {Africa}: Protocol of a prospective study.
+\newblock BMJ Open. 2020;10(8):e036214.
+\newblock doi:10.1136/bmjopen-2019-036214.
+
+\bibitem{Wood2007AMJRCCD}
+Wood R, Middelkoop K, Myer L, Grant AD, Whitelaw A, Lawn SD, et~al.
+\newblock Undiagnosed {Tuberculosis} in a {community} with {high} {HIV}
+  {prevalence}.
+\newblock Am J Respir Crit Care Med. 2007;175(1):87--93.
+\newblock doi:10.1164/rccm.200606-759OC.
+
+\bibitem{Middelkoop2011JAIDS}
+Middelkoop K, Bekker LG, Myer L, Johnson LF, Kloos M, Morrow C, et~al.
+\newblock Antiretroviral therapy and {TB} notification rates in a high {HIV}
+  prevalence {South} {African} community.
+\newblock J Acquir Immune Defic Syndr. 2011;56(3):263--269.
+\newblock doi:10.1097/QAI.0b013e31820413b3.
+
+\bibitem{Berhanu2023CID}
+Berhanu RH, Lebina L, Nonyane BAS, Milovanovic M, Kinghorn A, Connell L, et~al.
+\newblock Yield of facility-based targeted universal testing for {Tuberculosis}
+  with {Xpert} and {Mycobacterial} culture in high-risk groups attending
+  primary care facilities in {South} {Africa}.
+\newblock Clin Infect Dis. 2023;p. ciac965.
+\newblock doi:10.1093/cid/ciac965.
+
+\bibitem{Moyo2022LancetID}
+Moyo S, Ismail F, Walt MVd, Ismail N, Mkhondo N, Dlamini S, et~al.
+\newblock Prevalence of bacteriologically confirmed pulmonary tuberculosis in
+  {South} {Africa}, 2017–19: {A} multistage, cluster-based, cross-sectional
+  survey.
+\newblock Lancet Infect Dis. 2022;22(8):1172--1180.
+\newblock doi:10.1016/S1473-3099(22)00149-9.
+
+\bibitem{Patterson2024PNAS}
+Patterson B, Dinkele R, Gessner S, Koch A, Hoosen Z, January V, et~al.
+\newblock Aerosolization of viable \textit{{Mycobacterium} tuberculosis}
+  bacilli by tuberculosis clinic attendees independent of sputum-{Xpert}
+  {Ultra} status.
+\newblock Proc Natl Acad Sci. 2024;121(12):e2314813121.
+\newblock doi:10.1073/pnas.2314813121.
+
+\bibitem{Mikszewski2021GF}
+Mikszewski A, Stabile L, Buonanno G, Morawska L.
+\newblock The airborne contagiousness of respiratory viruses: {A} comparative
+  analysis and implications for mitigation.
+\newblock Geosci Front. 2021;13(6):101285.
+\newblock doi:10.1016/j.gsf.2021.101285.
+
+\bibitem{Buonanno2020EI}
+Buonanno G, Morawska L, Stabile L.
+\newblock Quantitative assessment of the risk of airborne transmission of
+  {SARS}-{CoV}-2 infection: {Prospective} and retrospective applications.
+\newblock Environ Int. 2020;145:106112.
+\newblock doi:10.1016/j.envint.2020.106112.
+
+\bibitem{Banholzer2024PGPH}
+Banholzer N, Schmutz R, Middelkoop K, Hella J, Egger M, Wood R, et~al.
+\newblock Airborne transmission risks of tuberculosis and {COVID}-19 in schools
+  in {South} {Africa}, {Switzerland}, and {Tanzania}: {Modeling} of
+  environmental data.
+\newblock PLOS Glob Public Health. 2024;4(1):e0002800.
+\newblock doi:10.1371/journal.pgph.0002800.
+
+\bibitem{Dharmadhikari2012AJRCCM}
+Dharmadhikari AS, Mphahlele M, Stoltz A, Venter K, Mathebula R, Masotla T,
+  et~al.
+\newblock Surgical face masks worn by patients with multidrug-resistant
+  tuberculosis: {Impact} on infectivity of air on a hospital ward.
+\newblock Am J Respir Crit Care Med. 2012;185(10):1104--1109.
+\newblock doi:10.1164/rccm.201107-1190OC.
+
+\bibitem{Batterman2017IJERPH}
+Batterman S.
+\newblock Review and extension of {CO2}-based methods to determine ventilation
+  rates with application to school classrooms.
+\newblock Int J Environ Res Public Health. 2017;14(2):145.
+\newblock doi:10.3390/ijerph14020145.
+
+\bibitem{Cheng2011EnvSciTech}
+Cheng KC, Acevedo-Bolton V, Jiang RT, Klepeis NE, Ott WR, Fringer OB, et~al.
+\newblock Modeling exposure close to air pollution sources in naturally
+  ventilated residences: {Association} of turbulent diffusion coefficient with
+  air change rate.
+\newblock Environ Sci Technol. 2011;45(9):4016--4022.
+\newblock doi:10.1021/es103080p.
+
+\bibitem{Foat2020BE}
+Foat T, Drodge J, Nally J, Parker S.
+\newblock A relationship for the diffusion coefficient in eddy diffusion based
+  indoor dispersion modelling.
+\newblock Build Environ. 2020;169:106591.
+\newblock doi:10.1016/j.buildenv.2019.106591.
+
+\bibitem{Lever2000LettersAppliedMicrobio}
+Lever Ms, Williams A, Bennett Am.
+\newblock Survival of mycobacterial species in aerosols generated from
+  artificial saliva.
+\newblock Lett Appl Microbiol. 2000;31(3):238--241.
+\newblock doi:10.1046/j.1365-2672.2000.00807.x.
+
+\bibitem{Gannon2007ResVetSci}
+Gannon BW, Hayes CM, Roe JM.
+\newblock Survival rate of airborne {Mycobacterium} bovis.
+\newblock Res J Vet Sci. 2007;82(2):169--172.
+\newblock doi:10.1016/j.rvsc.2006.07.011.
+
+\bibitem{Klein2014IJMyco}
+Klein K, Yang Z.
+\newblock Comparison of ambient air survival of {Mycobacterium} tuberculosis
+  clinical strains associated with different epidemiological phenotypes.
+\newblock Int J Mycobacteriol. 2014;3(3):211--213.
+\newblock doi:10.1016/j.ijmyco.2014.04.002.
+
+\bibitem{Adams1993}
+Adams WC.
+\newblock Measurement of breathing rate and volume in routinely performed daily
+  activities [final report].
+\newblock Human Performance Laboratory, Physical Education Department,
+  University of California, Davis; 1993. A033-205.
+\newblock Available from:
+  \url{https://ww2.arb.ca.gov/sites/default/files/classic/research/apr/past/a033-205.pdf}.
+
+\bibitem{Greenhalgh2022BMJ}
+Greenhalgh T, Peng Z, Jimenez JL, Bahnfleth W, Dancer SJ, Bourouiba L.
+\newblock Quantifying transmission risk of {SARS}-{CoV}-2 in different
+  situations.
+\newblock BMJ. 2022;376:o106.
+\newblock doi:10.1136/bmj.o106.
+
+\bibitem{Shen2021STBE}
+Shen J, Kong M, Dong B, Birnkrant MJ, Zhang J.
+\newblock Airborne transmission of {SARS}-{CoV}-2 in indoor environments: {A}
+  comprehensive review.
+\newblock Sci Technol Built Environ. 2021;27(10):1331--1367.
+\newblock doi:10.1080/23744731.2021.1977693.
+
+\bibitem{Edwards2024RiskAnalysis}
+Edwards AJ, King MF, Noakes CJ, Peckham D, López-García M.
+\newblock The {Wells}–{Riley} model revisited: {Randomness}, heterogeneity,
+  and transient behaviours.
+\newblock Risk Anal. 2024;p. 1--22.
+\newblock doi:10.1111/risa.14295.
+
+\bibitem{Wells1934AJE}
+Wells W.
+\newblock On air-borne infection: {Study} {II}. {Droplets} and droplet nuclei.
+\newblock Am J Epidemiol. 1934;20(3):611--618.
+\newblock doi:10.1093/oxfordjournals.aje.a118097.
+
+\bibitem{Leung2020NatMed}
+Leung NHL, Chu DKW, Shiu EYC, Chan KH, McDevitt JJ, Hau BJP, et~al.
+\newblock Respiratory virus shedding in exhaled breath and efficacy of face
+  masks.
+\newblock Nat Med. 2020;26(5):676--680.
+\newblock doi:10.1038/s41591-020-0843-2.
+
+\bibitem{Brankston2007LancetID}
+Brankston G, Gitterman L, Hirji Z, Lemieux C, Gardam M.
+\newblock Transmission of influenza {A} in human beings.
+\newblock Lancet Infect Dis. 2007;7(4):257--265.
+\newblock doi:10.1016/S1473-3099(07)70029-4.
+
+\bibitem{Narasimhan2013PulmonaryMed}
+Narasimhan P, Wood J, MacIntyre CR, Mathai D.
+\newblock Risk factors for tuberculosis.
+\newblock Pulm Med. 2013;2013:e828939.
+\newblock doi:10.1155/2013/828939.
+
+\bibitem{Jung2021InfectChemo}
+Jung J, Lee J, Jo S, Bae S, Kim JY, Cha HH, et~al.
+\newblock Nosocomial outbreak of {COVID}-19 in a hematologic ward.
+\newblock Infect Chemother. 2021;53(2):332.
+\newblock doi:10.3947/ic.2021.0046.
+
+\bibitem{Li2021BuildEnv}
+Li Y, Qian H, Hang J, Chen X, Cheng P, Ling H, et~al.
+\newblock Probable airborne transmission of {SARS}-{CoV}-2 in a poorly
+  ventilated restaurant.
+\newblock Build Environ. 2021;196:107788.
+\newblock doi:10.1016/j.buildenv.2021.107788.
+
+\bibitem{Yan2023BE}
+Yan Y, Li X, Fang X, Tao Y, Tu J.
+\newblock A spatiotemporal assessment of occupants’ infection risks in a
+  multi-occupants space using modified {Wells}–{Riley} model.
+\newblock Build Environ. 2023;230:110007.
+\newblock doi:10.1016/j.buildenv.2023.110007.
+
+\bibitem{Qian2009BE}
+Qian H, Li Y, Nielsen PV, Huang X.
+\newblock Spatial distribution of infection risk of {SARS} transmission in a
+  hospital ward.
+\newblock Build Environ. 2009;44(8):1651--1658.
+\newblock doi:10.1016/j.buildenv.2008.11.002.
+
+\bibitem{Li2022SOTTE}
+Li X, Lester D, Rosengarten G, Aboltins C, Patel M, Cole I.
+\newblock A spatiotemporally resolved infection risk model for airborne
+  transmission of {COVID}-19 variants in indoor spaces.
+\newblock Sci Total Environ. 2022;812:152592.
+\newblock doi:10.1016/j.scitotenv.2021.152592.
+
+\bibitem{Escombe2008PLoSMed}
+Escombe AR, Moore DAJ, Gilman RH, Pan W, Navincopa M, Ticona E, et~al.
+\newblock The infectiousness of tuberculosis patients coinfected with {HIV}.
+\newblock PLOS Med. 2008;5(9):e188.
+\newblock doi:10.1371/journal.pmed.0050188.
+
+\bibitem{Furin2019Lancet}
+Furin J, Cox H, Pai M.
+\newblock Tuberculosis.
+\newblock Lancet. 2019;393(10181):1642--1656.
+\newblock doi:10.1016/S0140-6736(19)30308-3.
+
+\end{thebibliography}
 

### 2025-02-14 doc/paper/manuscript-revised-clean.tex
Updates from Overleaf

@@
 In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.
-
-\subsection*{Monte Carlo simulation}
-
-Simulations are described in detail in Section~\zref{sec:estimation} in \supp.
-Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and the TB room.
-Each room was rasterized into a grid of cubic cells with an area of 0.25$^2$\,m$^2$.
-The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second.
-We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters.
-In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.
 

### 2025-02-14 analysis/environmental-data.Rmd
final cosmetic changes
consistent colors in fig 4d

@@
   # ) +
-  scale_fill_manual(values = wes_palette("Royal2")[-1]) +
+  scale_fill_manual(values = wes_palette("Royal2")) +
   scale_y_continuous(expand = expansion(mult = c(0, .05)), limits = c(0, NA)) +
@@
 
+saveRDS(
+  aer_pl,
+  "results/data/air-changes-per-hour.rds"
+)
+
 save_plot(
